### 2023-03-25 doc/_bayes_freq.qmd
working on the method explanation on Quarto

---
editor:
    render-on-save: true
---

### Frequency uncertainty $\pi(\underline{\underline{F}} \mid \underline{\underline{R}})$ {#sec-bayes_freq}

We begin with the probability of the frequency values given the raw barcode
reads. The first assumption we make is that the inference of the frequency
values for time $t$ is independent of any other time. Therefore, we can write
the joint probability distribution as a product of independent distributions of
the form
$$
\pi(\underline{\underline{F}} \mid \underline{\underline{R}}) =
\prod_{t=1}^T \pi(\underline{f}_t \mid \underline{r}_t).
$${#eq-freq_indep}

Writing Bayes' theorem for this series of independent terms results in
$$
\pi(\underline{f}_t \mid \underline{r}_t) \propto
\pi(\underline{r}_t \mid \underline{f}_t)
\pi(\underline{f}_t).
$${#eq-bayes_freq}
To model the prior uncertainty in the value of the frequencies, we assume a very
simple model. Notice that the frequencies at any time point must satisfy the
normalization constraint
$$
\sum_{b = 1}^B f_t^{(b)} = 1, \;\; \forall \; f_t^{(b)} \in [0, 1].
$${#eq-freq_norm}
This means that the probability distribution of the $B$ frequency values lies on
the $B$-simplex, i.e., the $B$ dimensional space constrained such that the sum
of all the dimensions must add up to one. A natural selection for a probability
that satisfies the constraint in @eq-freq_norm is the so-called Dirichlet 
distribution. The Dirichlet distribution is parametrized by a vector
$$
\underline{\alpha}_t = (\alpha^{(1)}_t, \cdots, \alpha^{(B)}_t)^\dagger, 
\;\; \alpha_t^{(k)} \geq 0 \;\; \forall \; k.
$${#eq-alpha_dir}
Given these parameters, we choose as a prior distribution for the barcode
frequency values $\pi(\underline{f}_t)$ a Dirichlet distribution of the form
$$
\pi(\underline{f}_t) = \frac{
    \Gamma\left(\sum_{b=1}^B\alpha_t^{(b)}\right)
}{
    \prod_{b=1}^B\Gamma(\alpha_t^{(b)})
} 
\prod_{b=1}^B (f_t^{(b)})^{\alpha_t^{(b)}-1}.
$${#eq-dirichlet}
@eq-dirichlet can be much more compactly written as
$$
\underline{f}_t \sim \operatorname{Dirichlet}(\underline{\alpha}_t),
$${#eq-freq_prior}
where the symbol "$\sim$" is read "distributed as". We are free to choose the
value of the parameters $\underline{\alpha}_t$ according to our prior
information. A reasonable choice is to assume that before observing any data at
all, we have no reason to believe any barcode is in higher abundance compared to
any other. This might not be entirely true at the beginning of the experiment,
where we expect the neutral barcodes to be in much higher frequency compared to
the mutant barcodes. Nevertheless, this might not be true for the following time
points, making our "uniform prior over the simplex" a reasonable choice. 
Therefore, we define
$$
\alpha_t^{(b)} = 1 \;\; \forall \; b.
$${#eq-alpha_choice}

For the likelihood term $\pi(\underline{r}_t \mid \underline{f}_t)$, we assume a
very simple multinomial model. The story behind this choice goes as follows:
Given a "true" underlying barcode frequency, when the sequencing machine reads a
particular barcode, this is equivalent to having thrown a $B$-sided die, where
the face of the die indicates the barcode that is read. This die is then rolled
$R_t = \sum_b r_t^{(b)}$ times, giving the values of the reads we observe at
time $t$. Notice that this model does not account for noise that might come from
sources such as genetic drift, DNA extraction, and PCR amplification. Although
these sources of noise are obviously important, it is not clear that we have
enough statistical power to make meaningful statements about them given our data
(more on this later on). We can then compactly write the likelihood function as
$$
\underline{r}_t \mid \underline{f}_t \sim
\operatorname{Multinomial}(\underline{f}_t).
$${#eq-freq_likelihood}

This simple Multinomial-Dirichlet model for the probability of the frequency
values is analytically tractable. Specifically, the Dirichlet distribution is
the so-called *conjugate prior* to the multinomial distribution. This means the
posterior distribution $\pi(\underline{f}_t \mid \underline{r}_t)$ has a
closed-form solution of the same form as the prior $\pi(\underline{f}_t)$, but
with updated parameters. In other words, we can write (without proof) the
posterior distribution as
$$
\underline{f}_t \mid \underline{r}_t \sim
\operatorname{Dirichlet}(\underline{\beta}_t),
$${#eq-freq_posterior}
where the new parameters are of the form
$$
\beta_t^{(b)} = \alpha_t^{(b)} + r_t^{(b)}.
$$
Simply put, when the likelihood function is multinomial, and the prior 
distribution is Dirichlet, the posterior distribution is also Dirichlet, where
the parameters of the posterior are the parameters of the prior plus the number
of observations of the corresponding output. This fact will become handy
as we build the entire posterior distribution for our inference as defined in
@eq-split_posterior.

### 2023-03-25 doc/_bayes_meanfit.qmd
working on the method explanation on Quarto

---
editor:
    render-on-save: true
---

### Population mean fitness uncertainty $\pi(\underline{\bar{s}}_T \mid \underline{\underline{F}}, \underline{\underline{R}})$ {#sec-bayes_meanfit}

### 2023-03-25 doc/_bayesian_inference.qmd
working on the method explanation on Quarto

---
editor:
    render-on-save: true
---
## Bayesian inference {#sec-bayesian_inference}

As defined in @sec-fitness_model, our ultimate objective is to infer the vector
of relative fitness values
$$
\underline{s}^M = (s^{(1)}, s^{(2)}, \ldots, s^{(M)})^\dagger,
$${#eq-fitness_mut_vec}
where $^\dagger$ indicates the transpose. Our data consists of an $B \times T$
matrix $\underline{\underline{R}}$, where $B = M + N$ is the number of unique
barcodes given by the sum of the number of unique mutant barcodes $M$ and the
number of unique neutral barcodes $N$, and $T$ is the number of time points
where measurements were taken. The data matrix is then of the form
$$
\underline{\underline{R}} = \begin{bmatrix}
\lvert & \lvert & \cdots & \lvert\\
\underline{r}_1 & \underline{r}_2 & \cdots & \underline{r}_T\\
\lvert & \lvert & \cdots & \lvert
\end{bmatrix},
$${#eq-R_mat}
where each column $\underline{r}_t$ is a $B$-dimensional array containing the
raw barcode counts at time $t$. We can further split each vector
$\underline{r}_t$ into two vectors of the form
$$
\underline{r}_t = \begin{bmatrix}
\underline{r}_t^{M} \\
\underline{r}_t^{N}
\end{bmatrix},
$${#eq-r-vec_split}
i.e., the vector containing the mutant barcode counts $\underline{r}_t^{M}$ and
the corresponding vector containing the neutral barcode counts
$\underline{r}_t^{N}$. Our task is to compute the joint probability distribution
of all relative fitness values given our data. Writing Bayes theorem this is
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) = \frac{
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M)}
{\pi(\underline{\underline{R}})},
$${#eq-bayes_obj}
where hereafter $\pi(\cdot)$ defines the probability distribution  of the input
unless otherwise stated. We can ignore the denominator on the right-hand side
of @eq-bayes_obj, since it serves as a normalization constant that does not
depend on our parameter of interest.

Although @eq-bayes_obj seems simple enough, recall that @eq-fitness relates
frequency values and the population mean fitness to the mutant relative fitness.
Therefore, we must include these nuisance parameters as part of our inference
problem. At the end of the day, we can always marginalize the joint distribution
to obtain the distribution we are interested in. To include these parameters,
let 
$$
\underline{\bar{s}}_T = (\bar{s}_1, \bar{s}_2, \ldots, \bar{s}_{T-1})^\dagger,
$${#eq-pop_fitness_vec}
be the vector containing the $T-1$ population mean fitness we compute from the
$T$ time points where measurements were taken. We have $T-1$ since the value
of any $\bar{s}_t$ requires time points $t$ and $t+1$. Furthermore, let the
matrix $\underline{\underline{F}}$ be a $B \times T$ matrix containing all
frequency values. As with @eq-R_mat, each column of $\underline{\underline{F}}$
can be split into two components of the form
$$
\underline{f}_t = \begin{bmatrix}
\underline{f}_t^M \\
\underline{f}_t^N
\end{bmatrix},
$${#eq-f_vec}
to split the corresponding mutant and neutral barcode frequencies. With these
variables in hand, the full inference problem we must solve takes the form
$$
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) \propto
\pi(
    \underline{\underline{R}} \mid
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
)
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
).
$${#eq-bayes_full}
As a reminder, we need to include these extra nuisance parameters, but we can
always integrate them out to recover the marginal distribution we care about,
i.e.,
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) =
\int d^{T-1}\underline{\bar{s}}_T
\int d^{B}\underline{f}_1 \cdots
\int d^{B}\underline{f}_T
\;
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
).
$${#eq-bayes_full_marginal}
Although daunting at first sight, @eq-bayes_full_marginal will be solved 
numerically, so we do not have to worry about this integration.

### Splitting the posterior distribution {#seq-split_posterior}

As written, the left-hand side of @eq-bayes_full is really difficult to work
with. However, we can take advantage of the structure of our inference problem
to rewrite it in a more manageable form. Specifically, we rewrite the joint
probability distribution as a product of conditional distributions of the form
$$
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) =
\pi(
    \underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}},
    \underline{\underline{R}}
)
\pi(
    \underline{\bar{s}}_T \mid \underline{\underline{F}}, 
    \underline{\underline{R}}
)
\pi(\underline{\underline{F}} \mid \underline{\underline{R}}).
$${#eq-split_posterior}
Written in this form, @eq-split_posterior captures the three sources of 
uncertainty listed in @sec-fitness_model in each of the terms. Starting from
right to left, the first term on the right-hand side of @eq-split_posterior
accounts for the uncertainty when inferring the frequency values given the
barcode reads. The second term accounts for the uncertainty in the values of
the mean population fitness at different time points. The last term accounts for
the uncertainty in the parameter we care about---the mutants' relative
fitnesses. In the following sections, we work through each of these terms to
build up the posterior distribution we will sample from piece by piece.

### 2023-03-25 doc/_experiment.qmd
working on the method explanation on Quarto

## Experimental setup

`[add details on expeirmental setup]`

### 2023-03-25 doc/_fitness_model.qmd
working on the method explanation on Quarto

---
editor:
    render-on-save: true
---

## Preliminaries on mathematical notation
Before jumping directly into what concerns us here, let us establish the
mathematical notation to be used throughout this document. We define (column)
vectors as underlined lowercase symbols such as
$$
\underline{x} = \begin{bmatrix}
    x_1\\
    x_2\\
    \vdots\\
    x_N
\end{bmatrix}.
$${#eq-vec_def}

In the same way, we define matrices as double-underline uppercase symbols such
as $\underline{\underline{A}}$.

## Fitness model {#sec-fitness_model}

Empirically, we have seen that the barcode trajectories follow an exponential
function of the form
$$
f_{t+1}^{(m)} = f_{t}^{(m)} \mathrm{e}^{(s^{(m)} - \bar{s}_t)\tau},
$${#eq-fitness}
where $f_{t}^{(m)}$ is the frequency of barcode $m$ at time $t$, $s^{(m)}$ is
the relative fitness with respect to the ancestral strain---the quantity we want
to infer from the data---$\bar{s}_t$ is the mean fitness of the culture at time
$t$, and $\tau$ is the time pass between time $t$ and $t+1$. It is important to
notice that the frequencies $f_{t}^{(m)}$ are not the allele frequencies in the
population (most of the culture is not sequenced since it is not barcoded), but
rather the inferred frequencies from the total number of reads.

Our ultimate objective is to infer the relative fitness $s^{(m)}$ for each of
the $M$ mutants in the experiment. To do so, we must take into account the three
main sources of uncertainty in our model:

1. Uncertainty in the determination of frequencies. Our model relates
frequencies between adjacent time points to the fitness of the corresponding
strain. However, we do not measure frequencies directly. Instead, our data
consists of a vector of barcode counts $\underline{r}_t$ for each of the $T$
time points in which the measurements were taken.
2. Uncertainty in the value of the population mean fitness. As we will see later
on, we use neutral lineages barcodes, which, by definition have mean fitness
$s^{(n)}$ to determine the value of the population mean fitness $\bar{s}_t$.
This inference will come with its own uncertainty that must be propagated to the
value of the relative fitness for the mutants.
3. Uncertainty in each of the mutant's fitness values.

To account for all these sources of uncertainty in a principled way, we develop
a Bayesian inference pipeline.

### 2023-03-25 doc/_bayes_freq.qmd
continue working on the explanation of the model

@@
 
-Writing Bayes' theorem for this series of independent terms results in
+Writing Bayes' theorem for one of these independent terms results in
 $$
@@
 \underline{\alpha}_t = (\alpha^{(1)}_t, \cdots, \alpha^{(B)}_t)^\dagger, 
-\;\; \alpha_t^{(k)} \geq 0 \;\; \forall \; k.
+\;\; \alpha_t^{(k)} > 0 \;\; \forall \; k.
 $${#eq-alpha_dir}
-Given these parameters, we choose as a prior distribution for the barcode
-frequency values $\pi(\underline{f}_t)$ a Dirichlet distribution of the form
+We then choose as a prior distribution for the barcode frequency values
+$\pi(\underline{f}_t)$ a Dirichlet distribution of the form
 $$
@@
 } 
-\prod_{b=1}^B (f_t^{(b)})^{\alpha_t^{(b)}-1}.
+\prod_{b=1}^B (f_t^{(b)})^{\alpha_t^{(b)}-1},
 $${#eq-dirichlet}
-@eq-dirichlet can be much more compactly written as
+where $\Gamma(\cdot)$ is the gamma function. @eq-dirichlet can be much more
+compactly written as
 $$
@@
 $${#eq-freq_prior}
-where the symbol "$\sim$" is read "distributed as". We are free to choose the
-value of the parameters $\underline{\alpha}_t$ according to our prior
-information. A reasonable choice is to assume that before observing any data at
-all, we have no reason to believe any barcode is in higher abundance compared to
-any other. This might not be entirely true at the beginning of the experiment,
-where we expect the neutral barcodes to be in much higher frequency compared to
-the mutant barcodes. Nevertheless, this might not be true for the following time
-points, making our "uniform prior over the simplex" a reasonable choice. 
-Therefore, we define
+where the symbol "$\sim$" is read "distributed as". Again, since the ultimate
+calculation will be performed numerically, we can hide all of the mathematical
+details in @eq-dirichlet with the simple expression in @eq-freq_prior. We are
+free to choose the value of the parameters $\underline{\alpha}_t$ according to
+our prior information. A reasonable choice is to assume that before observing
+any data at all, we have no reason to believe any barcode is in higher abundance
+compared to any other. This might not be entirely true at the beginning of the
+experiment, where we expect the neutral barcodes to be in much higher frequency
+compared to the mutant barcodes. Nevertheless, we have less certainty about the
+following time points, making our "uniform prior over the simplex" a reasonable
+choice. Therefore, we define
 $$
@@
 sources such as genetic drift, DNA extraction, and PCR amplification. Although
-these sources of noise are obviously important, it is not clear that we have
-enough statistical power to make meaningful statements about them given our data
-(more on this later on). We can then compactly write the likelihood function as
+these sources of noise are important, it is not clear that we have enough
+statistical power to make meaningful statements about them given our data (more
+on this later on). We can then compactly write the likelihood function as
 $$
 \underline{r}_t \mid \underline{f}_t \sim
-\operatorname{Multinomial}(\underline{f}_t).
+\operatorname{Multinomial}(R_t, \underline{f}_t).
 $${#eq-freq_likelihood}
@@
 This simple Multinomial-Dirichlet model for the probability of the frequency
-values is analytically tractable. Specifically, the Dirichlet distribution is
-the so-called *conjugate prior* to the multinomial distribution. This means the
-posterior distribution $\pi(\underline{f}_t \mid \underline{r}_t)$ has a
-closed-form solution of the same form as the prior $\pi(\underline{f}_t)$, but
-with updated parameters. In other words, we can write (without proof) the
-posterior distribution as
+values is convenient because it is analytically tractable. Specifically, the
+Dirichlet distribution is the so-called *conjugate prior* to the multinomial
+distribution. This means the posterior distribution $\pi(\underline{f}_t \mid
+\underline{r}_t)$ has a closed-form solution of the same form as the prior
+$\pi(\underline{f}_t)$, but with updated parameters. In other words, we can
+write (without proof) the posterior distribution as
 $$
@@
 $${#eq-freq_posterior}
-where the new parameters are of the form
+where the $\underline{\alpha}_t$ parameters are updated to $\underline{\beta}$
+after observing the data. These updated parameters are of the form
 $$

### 2023-03-25 doc/_bayes_meanfit.qmd
continue working on the explanation of the model

@@
 ### Population mean fitness uncertainty $\pi(\underline{\bar{s}}_T \mid \underline{\underline{F}}, \underline{\underline{R}})$ {#sec-bayes_meanfit}
+
+Next, we turn our attention to the problem of determining the population mean
+fitnesses $\underline{\bar{s}}_T$. First, we notice that our fitness model in
+@eq-fitness does not include the value of the raw reads. They enter the
+calculation indirectly through the inference of the frequency values we
+developed in @sec-bayes_freq. This means that we can remove the conditioning of
+the value of $\underline{\bar{s}}_T$ on the number of reads, obtaining a simpler
+probability function
+$$
+\pi(
+    \underline{\bar{s}}_T \mid 
+    \underline{\underline{F}}, \underline{\underline{R}}
+) = 
+\pi(
+    \underline{\bar{s}}_T \mid 
+    \underline{\underline{F}}
+).
+$${#eq-meanfit_noreads}
+Moreover, our fitness model does not directly explain how the population mean
+fitness evolves over time. In other words, our model cannot explicitly compute
+the population mean fitness at time $t+1$ from the information we have about
+time $t$. Given this model limitation, we are led to assume that we must need
+to infer each $\bar{s}_t$ independently. Expressing this for our inference
+results in
+$$
+\pi(
+    \underline{\bar{s}}_T \mid 
+    \underline{\underline{F}}
+) =
+\prod_{t=1}^{T-1} \pi(\bar{s}_t \mid \underline{f}_t, \underline{f}_{t+1}),
+$${#eq-meanfit_indep}
+where we split our matrix $\underline{\underline{F}}$ for each time point and
+only kept the conditioning on the relevant frequencies needed to compute the
+mean fitness at time $t$.
+
+Although our fitness model in @eq-fitness also includes the relative fitness
+$s^{(m)}$, to infer the population mean fitness we only utilize data from the
+neutral lineages that, by definition, have a relative fitness $s^{(n)} = 0$.
+Therefore, the conditioning on @eq-meanfit_indep can be further simplified by
+only keeping the frequencies of the neutral lineages, i.e.,
+$$
+\pi(\bar{s}_t \mid \underline{f}_t, \underline{f}_{t+1}) =
+\pi(\bar{s}_t \mid \underline{f}_t^N, \underline{f}_{t+1}^N).
+$${#eq-meanfit_indep_neutral}
+
+Recall that in @sec-fitness_model we emphasized that the frequencies $f_t^{(n)}$
+do not represent the true frequency of a particular lineage in the population
+but rather a "normalized number of cells." Therefore, it is safe to assume each
+of the $N$ neutral lineages' frequencies is changing independently. Their
+correlation of how increasing the frequency of one lineage will decrease the
+frequency of others is already captured in the model presented in
+@sec-bayes_freq. Thus, we write
+$$
+\pi(\bar{s}_t \mid \underline{f}_t^N, \underline{f}_{t+1}^N) =
+\prod_{n=1}^N \pi(\bar{s}_t \mid f_t^{(n)}, f_{t+1}^{(n)}).
+$${#eq-meanfit_indep_lineages}
+
+Now, we can focus on one of the terms on the right-hand side of
+@eq-meanfit_indep_lineages. Writing Bayes theorem results in
+$$
+\pi(\bar{s}_t \mid f_t^{(n)}, f_{t+1}^{(n)}) \propto
+\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t) \pi(\bar{s}_t).
+$${#eq-meanfit_bayes}
+Notice the likelihood defines the joint distribution of neutral barcode
+frequencies conditioned on the population mean fitness. However, rewriting our
+fitness model in @eq-fitness for a neutral lineage to leave frequencies on one
+hand and fitness on the other results in
+$$
+\frac{f_{t+1}^{(n)}}{f_t^{(n)}} = \mathrm{e}^{- \bar{s}_t}.
+$${#eq-fitness_ratio_neutral}
+@eq-fitness_ratio_neutral implies that our fitness model only relates **the
+ratio** of frequencies and not the individual values. To get around this 
+complication, we define
+$$
+\gamma_t^{(b)} \equiv \frac{f_{t+1}^{(b)}}{f_t^{(b)}},
+$${#eq-gamma_def}
+as the ratio of frequencies between two adjacent time points for any barcode
+$b$. This allows us to rewrite the joint distribution
+$\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t)$ as
+$$
+\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t) =
+\pi(f_t^{(n)}, \gamma_{t}^{(n)} \mid \bar{s}_t).
+$${#eq-joint_freq_gamma}
+Let us rephrase this subtle but necessary change of variables since it is a key
+part of the inference problem: our series of independence assumptions lead us to
+@eq-meanfit_bayes that relates the value of the population mean fitness
+$\bar{s}_t$ to the frequency of a neutral barcode at times $t$ and $t+1$. 
+However, as shown in @eq-fitness_ratio_neutral, our model functionally relates
+the ratio of frequencies---that we defined as $\gamma_t^{(n)}$---and not the
+independent frequencies to the mean fitness. Therefore, instead of writing the
+joint distribution of the frequency values at times $t$ and $t+1$ conditioned on
+the mean fitness for the likelihood, we write the joint distribution of the
+barcode frequency at time $t$ and the ratio of the frequencies. These **must
+be** equivalent joint distributions since there is a one-to-one mapping between
+$\gamma_t^{(n)}$ and $f_{t+1}^{(n)}$ for a given value of $f_t^{(n)}$.
+
+The real advantage of rewriting the joint distribution as in
+@eq-joint_freq_gamma comes from splitting this joint distribution as a product
+of conditional distributions of the form
+$$
+\pi(f_t^{(n)}, \gamma_{t}^{(n)} \mid \bar{s}_t) =
+\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t)
+\pi(\gamma_{t}^{(n)} \mid \bar{s}_t).
+$${#eq-joint_to_prod_gamma}
+Written in this form, we can finally propose a probabilistic model for how the
+mean fitness relates to the frequency ratios we determine in our experiments.
+The second term on the right-hand side of @eq-joint_to_prod_gamma relates how
+the determined frequency ratio $\gamma_t^{(b)}$ relates to the mean fitness
+$\bar{s}_t$. From @eq-fitness_ratio_neutral and @eq-gamma_def, we can write
+$$
+\ln \gamma_t^{(n)} = - \bar{s}_t + \varepsilon_t^{(n)},
+$${#eq-log_ratio_neutral}
+where we added an extra term, $\varepsilon_t^{(n)}$, characterizing the 
+deviations of the measurements from the theoretical model. It is common to 
+assume that these errors are normally distributed with mean zero and some 
+standard deviation $\sigma_t$, implying that
+$$
+\ln \gamma_t^{(n)} \mid \bar{s}_t, \sigma_t  \sim 
+\mathcal{N}\left(-\bar{s}_t, \sigma_t \right),
+$${#eq-log_gamma_normal}
+where we include the nuisance parameter $\sigma_t$ to be determined. If we 
+assume the log frequency ratio is normally distributed, this implies the 
+frequency ratio itself is distributed log-normal. This means that
+$$
+\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t  \sim 
+\log \mathcal{N}\left(-\bar{s}_t, \sigma_t \right).
+$${#eq-meanfit_likelihood}
+Having added the nuisance parameter $\sigma_t$ implies that we must update
+@eq-meanfit_bayes to
+$$
+\pi(\bar{s}_t, \sigma_t \mid f_t^{(n)}, f_{t+1}^{(n)}) \propto
+\pi(f_t^{(n)}, \gamma_t^{(n)} \mid \bar{s}_t, \sigma_t) 
+\pi(\bar{s}_t) \pi(\sigma_t),
+$${#eq-meanfit_bayes_full}
+where we assume the prior for each parameter is independent, i.e.,
+$$
+\pi(\bar{s}_t, \sigma_t) = \pi(\bar{s}_t) \pi(\sigma_t).
+$${#eq-meanfit_indep_prior}
+
+For the first term on the right-hand side of @eq-joint_to_prod_gamma,
+$\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t)$, we remove the conditioning
+on the population mean fitness since it does not add any information on top of
+what the frequency ratio $\gamma_t^{(n)}$ already gives. Therefore, we have
+$$
+\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t) =
+\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}).
+$${#eq-freq_cond_gamma}
+The right-hand side of @eq-freq_cond_gamma asks us to compute the probability
+of observing a frequency value $f_t^{(n)}$ given that we get to observe the
+ratio $\gamma_{t}^{(n)}$. If the ratio happened to be $\gamma_{t}^{(n)} = 2$,
+we could have $f_{t+1}^{(n)} = 1$ and $f_{t+1}^{(n)} = 0.5$, for example.
+Although, it would be equally likely that $f_{t+1}^{(n)} = 0.6$ and
+$f_{t+1}^{(n)} = 0.3$ or $f_{t+1}^{(n)} = 0.1$ and $f_{t+1}^{(n)} = 0.05$ for
+that matter. As a matter of fact, given that we only get to observe the
+frequency ratio $\gamma_t^{(n)}$, we can only restrict the values of the
+frequency $f_{t}^{(n)}$ to be
+$$
+f_{t}^{(n)} \in \left(0, \frac{1}{\gamma_t^{(n)}} \right].
+$${#eq-freq_given_gamma_range}
+A priori, we do not have any reason to favor any value over any other, therefore
+it is natural to write
+$$
+f_t^{(n)} \mid \gamma_t^{(n)} \sim 
+\operatorname{Uniform}\left( 0, \frac{1}{\gamma_t^{(n)}} \right).
+$${#eq-f_given_gamma_uniform}
+
+#### Summary
+
+Putting all the pieces we have developed in this section together results in an
+inference for the population mean fitness values of the form
+$$
+\pi(\underline{\bar{s}}_T \mid \underline{\underline{F}}) \propto
+\prod_{t=1}^{T-1} \left\{
+    \prod_{n=1}^N \left[
+        \pi(f_t^{(n)} \mid \gamma_t^{(n)}) 
+        \pi(\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t)
+    \right]
+    \pi(\bar{s}_t) \pi(\sigma_t)
+\right\},
+$${#eq-meanfit_full_inference}
+where we have
+$$
+f_t^{(n)} \mid \gamma_t^{(n)} \sim 
+\operatorname{Uniform} \left(0, \frac{1}{\gamma_t^{(n)}} \right),
+$${#eq-meanfit_eq1}
+$$
+\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t \sim 
+\log\mathcal{N}(\bar{s}_t, \sigma_t),
+$${#eq-meanfit_eq2}
+$$
+\bar{s}_t \sim \mathcal{N}(0, \sigma_{\bar{s}_t}),
+$$
+and
+$$
+sigma_t \sim \operatorname{Half}-\mathcal{N}(0, \sigma_{\sigma_t}),
+$$
+where $\sigma_{\bar{s}_t}$ and $\sigma_{\sigma_t}$ are free parameters we can
+choose.

### 2023-03-25 doc/_fitness_model.qmd
continue working on the explanation of the model

@@
 $$
-f_{t+1}^{(m)} = f_{t}^{(m)} \mathrm{e}^{(s^{(m)} - \bar{s}_t)\tau},
+f_{t+1}^{(b)} = f_{t}^{(b)} \mathrm{e}^{(s^{(b)} - \bar{s}_t)\tau},
 $${#eq-fitness}
-where $f_{t}^{(m)}$ is the frequency of barcode $m$ at time $t$, $s^{(m)}$ is
+where $f_{t}^{(b)}$ is the frequency of barcode $b$ at time $t$, $s^{(b)}$ is
 the relative fitness with respect to the ancestral strain---the quantity we want
@@
 $t$, and $\tau$ is the time pass between time $t$ and $t+1$. It is important to
-notice that the frequencies $f_{t}^{(m)}$ are not the allele frequencies in the
+notice that the frequencies $f_{t}^{(b)}$ are not the allele frequencies in the
 population (most of the culture is not sequenced since it is not barcoded), but
-rather the inferred frequencies from the total number of reads.
+rather the inferred frequencies from the total number of reads. A way to 
+conceptualize this is that we assume exponential growth in the *number of cells*
+$N_t^{(b)}$ of the form
+$$
+N_{t+1}^{(b)} = N_{t}^{(b)} \mathrm{e}^{(s^{(b)} - \bar{s}_t)\tau},
+$${#eq-ncells}
+for every barcode $b$. However, when we sequence barcodes, we do not directly
+measure the number of cells, but some number of reads $r_t^{(b)}$ that map to
+barcode $b$. In the simplest possible scenario, we assume
+$$
+r_{t}^{(b)} \propto N_{t}^{(b)},
+$${#eq-r_to_n}
+where, importantly, the proportionality constant depends on the total number of
+reads for the time $t$ library, which might vary from library to library.
+Therefore, to compare the number of reads between libraries at different time
+points, we must somehow normalize them to the same scale. The simplest form is
+to define a relative abundance, i.e., a frequency with respect to the total
+number of reads,
+$$
+f_{t}^{(b)} \equiv \frac{r_{t}^{(b)}}{\sum_{b'} r_{t}^{(b')}}.
+$${#eq-r_to_n_norm}
+This is the frequency @eq-fitness describes.
 

### 2023-03-27 doc/_bayes_meanfit.qmd
finish writing model section

@@
 $${#eq-log_ratio_neutral}
-where we added an extra term, $\varepsilon_t^{(n)}$, characterizing the 
-deviations of the measurements from the theoretical model. It is common to 
-assume that these errors are normally distributed with mean zero and some 
-standard deviation $\sigma_t$, implying that
+where, for simplicity, we set $\tau = 1$. Note that we added an extra term,
+$\varepsilon_t^{(n)}$, characterizing the deviations of the measurements from
+the theoretical model. It is common to assume that these errors are normally
+distributed with mean zero and some standard deviation $\sigma_t$, implying that
 $$
@@
 $${#eq-meanfit_indep_prior}
+For numerical stability, we will select weakly-informative priors for both of
+these parameters. In the case of the nuisance parameter $\sigma_t$, the prior
+must be restricted to positive values only, since standard deviations cannot be
+negative.
 
@@
 $f_{t+1}^{(n)} = 0.3$ or $f_{t+1}^{(n)} = 0.1$ and $f_{t+1}^{(n)} = 0.05$ for
-that matter. As a matter of fact, given that we only get to observe the
-frequency ratio $\gamma_t^{(n)}$, we can only restrict the values of the
-frequency $f_{t}^{(n)}$ to be
+that matter. If we only get to observe the frequency ratio $\gamma_t^{(n)}$, we
+know that the numerator $f_{t+1}^{(n)}$ can only take values between zero and
+one, all of them being equally likely given only the information on the ratio.
+As a consequence, the value of the frequency in the denominator $f_{t}^{(n)}$ is
+restricted to fall in the range
 $$
@@
 \bar{s}_t \sim \mathcal{N}(0, \sigma_{\bar{s}_t}),
-$$
+$${#eq-meanfit_eq3}
 and
 $$
-sigma_t \sim \operatorname{Half}-\mathcal{N}(0, \sigma_{\sigma_t}),
-$$
+\sigma_t \sim \operatorname{Half}-\mathcal{N}(0, \sigma_{\sigma_t}),
+$${#eq-meanfit_eq4}
 where $\sigma_{\bar{s}_t}$ and $\sigma_{\sigma_t}$ are free parameters we can

### 2023-03-27 doc/_bayes_mutfit.qmd
finish writing model section

---
editor:
    render-on-save: true
---

### Mutant relative fitness uncertainty $\pi(\underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}}, \underline{\underline{R}})$ {#seq-bayes_mutfit}

The last piece of our inference is the piece that we care about the most: the
probability distribution of all the mutants' relative fitness, given the
inferred population mean fitness and the frequencies. First, we assume that all
fitness values are independent of each other. This allows us to write
$$
\pi(
    \underline{s}^M \mid 
    \underline{\bar{s}}_T, \underline{\underline{F}}, \underline{\underline{R}}
) = 
\prod_{m=1}^M \pi(
    s^{(m)} \mid
    \underline{\bar{s}}_T, \underline{\underline{F}}, \underline{\underline{R}}
).
$${#eq-mutfit_indep}
Furthermore, as was the case with the population mean fitness, our fitness
model relates frequencies, not raw reads. Moreover, the fitness value of mutant
$m$ only depends on the frequencies of such mutant. Therefore, we can simplify
the conditioning to
$$
\pi(
    s^{(m)} \mid
    \underline{\bar{s}}_T, \underline{\underline{F}}, \underline{\underline{R}}
) = 
\pi(s^{(m)} \mid \underline{\bar{s}}_T, \underline{f}^{(m)}),
$${#eq-mutfit_cond_simple}
where
$$
\underline{f}^{(m)} = (f_0^{(m)}, f_1^{(m)}, \ldots, f_T^{(m)})^\dagger,
$${#eq-mutfit_timeseries_def}
is the vector containing the frequency time series for mutant $m$. Writing 
Bayes' theorem for the right-hand side of @eq-mutfit_cond_simple results in
$$
\pi(s^{(m)} \mid \underline{\bar{s}}_T, \underline{f}^{(m)}) \propto
\pi(\underline{f}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)})
\pi(s^{(m)} \mid \underline{\bar{s}}_T).
$${#eq-mutfit_bayes}
Notice the conditioning on the mean fitness values $\underline{\bar{s}}_T$ is 
not inverted since we already inferred these values.

Following the logic used in @sec-bayes_meanfit, let us define
$$
\underline{\gamma}^{(m)} = 
(\gamma_0^{(m)}, \gamma_1^{(m)}, \ldots, \gamma_{T-1}^{m})^\dagger,
$${#eq-gamma_vec_def}
where each entry $\gamma_t^{(m)}$ is defined by @eq-gamma_def. In the same way
we rewrote the joint distribution between two adjacent time point frequencies 
to the joint distribution between one of the frequencies and the ratio of both 
frequencies in @eq-joint_freq_gamma, we can rewrite the joint distribution of
the frequency time series for mutant $m$ as
$$
\pi(\underline{f}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}) =
\pi(f_0^{(m)}, \underline{\gamma}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}).
$${#eq-joint_freq_gammas_mutfit}
One can think about @eq-joint_freq_gammas_mutfit as saying that knowing the
individual frequencies at each time point contain equivalent information as
knowing the initial frequency and the subsequent ratios of frequencies. This is
because if we want to know the value of $f_1^{(m)}$ given the ratios, we only
need to compute
$$
f_1^{(m)} = \gamma_0^{(m)} f_0^{(m)}.
$${#eq-f1_from_ratio}
Moreover, if we want to know $f_2^{(m)}$, we have
$$
f_2^{(m)} = \gamma_1^{(m)} f_1^{(m)} =
\gamma_1^{(m)} \left(\gamma_0^{(m)} f_0^{(m)}\right),
$${#eq-f2_from_ratio}
and so on. We can then write the joint distribution on the right-hand side of
@eq-joint_freq_gammas_mutfit as a product of conditional distributions of the
form
$$
\begin{aligned}
\pi(f_0^{(m)}, \underline{\gamma}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}) =
&\pi(
    f_0^{(m)} \mid 
    \gamma_0^{(m)}, \ldots, \gamma_{T-1}^{(m)}, \underline{\bar{s}}_T, s^{(m)}
) \times \\
&\pi(
    \gamma_0^{(m)} \mid 
    \gamma_1^{(m)}, \ldots, \gamma_{T-1}^{(m)}, \underline{\bar{s}}_T, s^{(m)}
) \times \\
&\pi(
    \gamma_1^{(m)} \mid 
    \gamma_2^{(m)}, \ldots, \gamma_{T-1}^{(m)}, \underline{\bar{s}}_T, s^{(m)}
) \times \\
&\vdots \\
&\pi(
    \gamma_{T-2}^{(m)} \mid \gamma_{T-1}^{(m)}, \underline{\bar{s}}_T, s^{(m)}
) \times \\
&\pi(\gamma_{T-1}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}).
\end{aligned}
$${#eq-mutfit_joint_to_cond}
Writing the fitness model in @eq-fitness as
$$
\gamma_t^{(m)} = \frac{f_{t+1}^{(m)}}{f_t^{(m)}} = 
\mathrm{e}^{(s^{(m)} - s_t)\tau},
$$
reveals that the value of each of the ratios $\gamma_t^{(m)}$ only depends on
the corresponding fitness value $\bar{s}_t$ and the relative fitness $s^{(m)}$.
Therefore, we can remove most of the conditioning on the right-hand side of 
@eq-mutfit_joint_to_cond, resulting in a much simpler joint distribution of the
form
$$
\begin{aligned}
\pi(f_0^{(m)}, \underline{\gamma}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}) =
&\pi(f_0^{(m)} \mid \gamma_0^{(m)}) \times \\
&\pi(\gamma_0^{(m)} \mid \bar{s}_0, s^{(m)}) \times \\
&\pi(\gamma_1^{(m)} \mid \bar{s}_1, s^{(m)}) \times \\
&\vdots \\
&\pi(\gamma_{T-2}^{(m)} \mid \bar{s}_{T-2}, s^{(m)}) \times \\
&\pi(\gamma_{T-1}^{(m)} \mid \bar{s}_{T-1}, s^{(m)}),
\end{aligned}
$${#eq-mutfit_joint_to_cond_simple}
where for the first term on the right-hand side of
@eq-mutfit_joint_to_cond_simple we apply the same logic as in
@eq-freq_cond_gamma to remove all other dependencies. We emphasize that although
@eq-mutfit_joint_to_cond_simple looks like a series of independent inferences,
the value of the relative fitness $s^{(m)}$ is shared among all of them. This
means that the parameter is not inferred individually for each time point,
resulting in different estimates of the parameter, but each time point
contributes independently to the inference of a single estimate of $s^{(m)}$.

Using equivalent arguments to those in @sec-bayes_meanfit, we assume
$$
f_0^{(m)} \mid \gamma_0^{(m)} \sim 
\operatorname{Uniform}\left(0, \frac{1}{\gamma_0^{(m)}} \right),
$$
and
$$
\gamma_t^{(m)} \mid \bar{s}_t, s^{(m)}, \sigma^{(m)} \sim 
\log\mathcal{N}\left(s^{(m)} - \bar{s}_t, \sigma^{(m)} \right),
$${#eq-mutfit_lognormal}
where we add the nuisance parameter $\sigma^{(m)}$ to the inference. Notice that
this parameter is not indexed by time. This means that we assume the deviations
from the theoretical prediction do not depend on time, but only on the mutant.
Adding the nuisance parameter demands us to update @eq-mutfit_bayes to
$$
\pi(
    s^{(m)}, \sigma^{(m)} \mid \underline{\bar{s}}_T, \underline{f}^{(m)}
) \propto
\pi(\underline{f}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}, \sigma^{(m)})
\pi(s^{(m)}) \pi(\sigma^{(m)}),
$${#eq-mutfit_bayes_full}
where we assume independent priors for both parameters. We also removed the
conditioning on the values of the mean fitness as knowing such values does not
change our prior information about the possible range of values that the
parameters can take. As with the priors on @sec-bayes_meanfit, we will assign
weakly-informative priors to these parameters.

#### Summary

With all pieces in place, we write the full inference of the relative fitness
values as
$$
\pi(
    \underline{s}^M ,\underline{\sigma}^M \mid 
    \underline{\bar{s}}_T, \underline{\underline{F}}
) \propto
\prod_{m=1}^M \left\{ 
    \pi(f_0^{(m)} \mid \gamma_0^{(m)})
    \prod_{t=0}^{T-1} \left[
        \pi(\gamma_t^{(m)} \mid \bar{s}_t, s^{(m)}, \sigma^{(m)})
    \right]
    \pi(s^{(m)}) \pi(\sigma^{(m)})
\right\},
$${#eq-mutfit_full_inference}
where
$$
f_0^{(m)} \mid \gamma_0^{(m)} \sim 
\operatorname{Uniform}\left(0, \frac{1}{\gamma_0^{(m)}} \right),
$${#eq-mutfit_eq1}
$$
\gamma_t^{(m)} \mid \bar{s}_t, s^{(m)}, \sigma^{(m)} \sim 
\log\mathcal{N}\left(s^{(m)} - \bar{s}_t, \sigma^{(m)} \right),
$${#eq-mutfit_eq2}
$$
s^{(m)} \sim \mathcal{N}(0, \sigma_{s^{(m)}}),
$${#eq-mutfit_eq3}
and
$$
\sigma^{(m)} \sim \operatorname{Half}-\mathcal{N}(0, \sigma_{\sigma^{(m)}}),
$${#eq-mutfit_eq4}
where $\sigma_{s^{(m)}}$ and $\sigma_{\sigma^{(m)}}$ are free parameters we can
choose.

### 2023-03-30 doc/_bayes_freq.qmd
save unstaged changes

@@
 the $B$-simplex, i.e., the $B$ dimensional space constrained such that the sum
-of all the dimensions must add up to one. A natural selection for a probability
+of all the dimensions must add up to one. A natural choice for a probability
 that satisfies the constraint in @eq-freq_norm is the so-called Dirichlet
@@
 \underline{r}_t \mid \underline{f}_t \sim
-\operatorname{Multinomial}(R_t, \underline{f}_t).
+\operatorname{Multinomial}(R_t, \underline{f}_t),
 $${#eq-freq_likelihood}
+which again is simply read as the number of reads at time $t$ $\underline{r}_t$
+given the frequencies at the same time $\underline{f}_t$ is distributed with
+a multinomial distribution parametrized by $R_t$, the total number of reads, and
+the frequency vector.
 
@@
 $${#eq-freq_posterior}
-where the $\underline{\alpha}_t$ parameters are updated to $\underline{\beta}$
+where the $\underline{\alpha}_t$ parameters are updated to $\underline{\beta}_t$
 after observing the data. These updated parameters are of the form

### 2023-03-30 doc/_bayes_meanfit.qmd
save unstaged changes

@@
 the population mean fitness at time $t+1$ from the information we have about
-time $t$. Given this model limitation, we are led to assume that we must need
-to infer each $\bar{s}_t$ independently. Expressing this for our inference
+time $t$. Given this model limitation, we are led to assume that we must infer
+each $\bar{s}_t$ independently. Expressing this for our inference
 results in
@@
 but rather a "normalized number of cells." Therefore, it is safe to assume each
-of the $N$ neutral lineages' frequencies is changing independently. Their
+of the $N$ neutral lineages' frequencies is changing independently. The
 correlation of how increasing the frequency of one lineage will decrease the
@@
 fitness model in @eq-fitness for a neutral lineage to leave frequencies on one
-hand and fitness on the other results in
+side and fitness on the other results in
 $$
-\frac{f_{t+1}^{(n)}}{f_t^{(n)}} = \mathrm{e}^{- \bar{s}_t}.
+\frac{f_{t+1}^{(n)}}{f_t^{(n)}} = \mathrm{e}^{- \bar{s}_t}\tau.
 $${#eq-fitness_ratio_neutral}
@@
 the ratio of frequencies---that we defined as $\gamma_t^{(n)}$---and not the
-independent frequencies to the mean fitness. Therefore, instead of writing the
-joint distribution of the frequency values at times $t$ and $t+1$ conditioned on
-the mean fitness for the likelihood, we write the joint distribution of the
+independent frequencies to the mean fitness. Therefore, instead of writing for
+the likelihood the joint distribution of the frequency values at times $t$ and
+$t+1$ conditioned on the mean fitness, we write the joint distribution of the
 barcode frequency at time $t$ and the ratio of the frequencies. These **must
 be** equivalent joint distributions since there is a one-to-one mapping between
-$\gamma_t^{(n)}$ and $f_{t+1}^{(n)}$ for a given value of $f_t^{(n)}$.
+$\gamma_t^{(n)}$ and $f_{t+1}^{(n)}$ for a given value of $f_t^{(n)}$. Another
+way to phrase this is to say that knowing the frequency at time $t$ and at time
+$t+1$ provides the same amount of information as knowing the frequency at time
+$t$ and the ratio of the frequencies. This is because if we want to obtain
+$f_{t+1}^{(n)}$ given this information, we simply compute
+$$
+f_{t+1}^{(n)} = \gamma_t^{(n)} f_t^{(n)}.
+$${#eq-meanfit_ft+1_gamma}
 
@@
 $$
-\pi(\underline{\bar{s}}_T \mid \underline{\underline{F}}) \propto
+\pi(
+    \underline{\bar{s}}_T, \underline{\sigma}_T \mid \underline{\underline{F}}
+) \propto
 \prod_{t=1}^{T-1} \left\{

### 2023-03-30 doc/_fitness_model.qmd
save unstaged changes

@@
 on, we use neutral lineages barcodes, which, by definition have mean fitness
-$s^{(n)}$ to determine the value of the population mean fitness $\bar{s}_t$.
+$s^{(n)} = 0$ to determine the value of the population mean fitness $\bar{s}_t$.
 This inference will come with its own uncertainty that must be propagated to the

### 2023-03-30 doc/_mcmc_sampling.qmd
save unstaged changes

---
editor:
    render-on-save: true
---
## Markov Chain Monte Carlo sampling of posterior {#sec-mcmc_sampling}

### 2023-04-12 doc/_mcmc_sampling.qmd
working on the mcmc_sampling section

@@
 ## Markov Chain Monte Carlo sampling of posterior {#sec-mcmc_sampling}
+
+The model proposed in @sec-bayesian_inference presents a daunting computational
+task as it demands the simultaneous inference of > $10^4$ parameters. Sampling
+such high-dimensional posterior distribution can be extremely slow, even for
+state-of-the-art Markov Chain Monte Carlo samplers. Nevertheless, given the
+structure of our problem, we can get around this issue by focusing on the
+marginal individual mutants' relative fitness posterior distributions. This is
+perfectly justified as we are only interested in these marginal distributions
+$\pi(s^{(m)} \mid \underline{\underline{R}})$, rather than the full joint
+posterior distribution.
+
+To make the inference task computationally feasible, we divide the process into
+two steps:
+
+1. Sample out of the posterior distribution of the population mean fitness using
+the information of the neutral lineages only.
+2. Sample out of each mutant's relative fitness posterior distribution given the
+knowledge acquired in step 1.
+
+In what follows, we describe the modification of the model presented in
+@eq-mutfit_full_inference to make Markov Chain Monte Carlo sampling of the
+posterior distribution computationally feasible.
+
+### Sample population mean fitness posterior distribution
+
+The full joint posterior distribution for our experiment $\pi(\underline{s}^M,
+\underline{\sigma}^M, \underline{\bar{s}}_T, \underline{\sigma}_T,
+\underline{\underline{F}} \mid \underline{\underline{R}})$ demands the
+simultaneous inference of all of the frequencies $\underline{\underline{F}}$,
+the population mean fitness values for every pair of time points
+$\underline{\bar{s}}_T$, the mutants relative fitness $\underline{s}^M$, and the
+nuisance parameters inherited from the likelihood functions
+$\underline{\sigma}^M$ and $\underline{\sigma}_T$. Throughout
+@sec-bayesian_inference, we describe a series of independence assumptions that
+enormously simplified the problem. Nevertheless, these assumptions do not remove
+all of the computational complexity, as we still need to infer all of the
+frequency values for each of the $~ 10^3-10^4$ unique barcodes.
+
+The first step towards beating the full posterior distribution to a
+computationally tractable form is to focus on the inference of the population
+mean fitness, ignoring the mutants' relative fitness values. This is
+mathematically sound, given the structure of our problem, as described in
+@sec-bayesian_inference. More specifically, @eq-split_posterior shows that we
+can split the posterior into the product of three terms, where the term
+involving the mean fitness values $\underline{\bar{s}}_T$ do not depend on the
+mutants relative fitness. Thus, we are left with the challenge of sampling from
+a "simpler" posterior distribution with less parameters,
+$$
+\pi(
+    \underline{\bar{s}}_T, \underline{\sigma}_T, \underline{\underline{F}} \mid
+    \underline{\underline{R}}
+) = 
+\pi(\underline{\bar{s}}_T, \underline{\sigma}_T \mid \underline{\underline{F}})
+\pi(\underline{\underline{F}} \mid \underline{\underline{R}}),
+$${#eq-simpler_meanfit_posterior}
+where the right-hand side splits the joint distribution into the product of
+conditional distributions, as in @eq-split_posterior.
+@eq-simpler_meanfit_posterior already removed ~$10^3$ parameters from the 
+inference problem, but we still have the simultaneous inference of all 
+frequencies $\underline{\underline{F}}$. To make further progress, we can follow
+the independence assumptions described in @sec-bayes_meanfit, which led to
+@eq-meanfit_indep, to write the posterior distribution as a product of the form
+$$
+\pi(\underline{\bar{s}}_T, \underline{\sigma}_T \mid \underline{\underline{F}})
+\pi(\underline{\underline{F}} \mid \underline{\underline{R}}) 
+= \prod_{t=1}^{T-1} 
+\pi(\bar{s}_t, \sigma_t, \mid \underline{f}_t, \underline{f}_{t+1})
+\pi(
+    \underline{f}_t, \underline{f}_{t+1} \mid 
+    \underline{r}_t, \underline{r}_{t+1}
+).
+$${#eq-meanfit_posterior_indep}
+If we sample from the posterior distribution of one of the $T-1$ terms on the
+right-hand side of @eq-meanfit_posterior_indep, we again enormously reduce the
+number of parameters that need to be inferred simultaneously, as we only focus
+on the frequencies $\underline{f}_t$, and $\underline{f}_{t+1}$, relevant for a
+single mean fitness value $\bar{s}_t$ rather than all frequencies
+$\underline{\underline{F}}$. Nevertheless, the large number of unique barcodes
+makes this still a challenging inference problem.
+
+However, the frequencies $\underline{f}_t$, and $\underline{f}_{t+1}$, and the
+standard deviation $\sigma_t$ represent nuisance parameters we ignore at the end
+of the inference procedure. In other words, what we care about is the marginal
+distribution of the mean fitness given the data, i.e.,
+$$
+\pi(\bar{s}_t \mid \underline{r}_t, \underline{r}_{t+1}) =
+\int d\sigma_t\;
+\int d^B \underline{f}_t\;
+\int d^B \underline{f}_{t+1}\;
+\pi(\bar{s}_t, \sigma_t, \mid \underline{f}_t, \underline{f}_{t+1})
+\pi(
+    \underline{f}_t, \underline{f}_{t+1} \mid 
+    \underline{r}_t, \underline{r}_{t+1}
+),
+$${#eq-meanfit_posterior_marginal}
+where we remind the reader that $B$ is the total number of unique barcodes. 
+These integrals with respect to all barcode frequencies pose the ultimate
+computational challenge we must overcome.
+
+Since the
+inference of the mean fitness $\bar{s}_t$ makes use only of the neutral lineages
+frequencies, it is convenient to split the $\underline{f}_t$ vectors into the
+neutral and mutant-only frequency arrays and write
+$$
+\pi(
+    \bar{s}_t, \sigma_t, \underline{f}_t, \underline{f}_{t+1} \mid
+    \underline{r}_t, \underline{r}_{t+1}
+) =
+\pi(
+    \bar{s}_t, \sigma_t, 
+    \underline{f}_t^N, \underline{f}_t^M, 
+    \underline{f}_{t+1}^N, \underline{f}_{t+1}^M \mid
+    \underline{r}_t, \underline{r}_{t+1}
+)
+$${#eq-meanfit_split_f}

### 2023-04-12 doc/_mcmc_sampling.qmd
keep working on _mcmc_sampling section

@@
 \pi(\bar{s}_t, \sigma_t, \mid \underline{f}_t, \underline{f}_{t+1})
-\pi(
-    \underline{f}_t, \underline{f}_{t+1} \mid 
-    \underline{r}_t, \underline{r}_{t+1}
-).
+\pi(\underline{f}_t \mid \underline{r}_t)
+\pi(\underline{f}_{t+1} \mid \underline{r}_{t+1}).
 $${#eq-meanfit_posterior_indep}
@@
 
-However, the frequencies $\underline{f}_t$, and $\underline{f}_{t+1}$, and the
-standard deviation $\sigma_t$ represent nuisance parameters we ignore at the end
-of the inference procedure. In other words, what we care about is the marginal
+The frequencies $\underline{f}_t$, and $\underline{f}_{t+1}$, and the standard
+deviation $\sigma_t$ represent nuisance parameters we ignore at the end of the
+inference procedure. In other words, what we care about is the marginal
 distribution of the mean fitness given the data, i.e.,
@@
 \pi(\bar{s}_t, \sigma_t, \mid \underline{f}_t, \underline{f}_{t+1})
-\pi(
-    \underline{f}_t, \underline{f}_{t+1} \mid 
-    \underline{r}_t, \underline{r}_{t+1}
-),
+\pi(\underline{f}_t \mid \underline{r}_t)
+\pi(\underline{f}_{t+1} \mid \underline{r}_{t+1}),
 $${#eq-meanfit_posterior_marginal}
@@
 These integrals with respect to all barcode frequencies pose the ultimate
-computational challenge we must overcome.
-
-Since the
+computational challenge we must overcome. To this goal, notice that the
 inference of the mean fitness $\bar{s}_t$ makes use only of the neutral lineages
-frequencies, it is convenient to split the $\underline{f}_t$ vectors into the
-neutral and mutant-only frequency arrays and write
+frequencies. Thus, it is convenient to split the $\underline{f}_t$ vectors into
+the neutral and mutant-only frequency arrays and write
 $$
+\pi(\bar{s}_t, \sigma_t, \mid \underline{f}_t, \underline{f}_{t+1})
+\pi(\underline{f}_t \mid \underline{r}_t)
+\pi(\underline{f}_{t+1} \mid \underline{r}_{t+1}) = 
 \pi(
-    \bar{s}_t, \sigma_t, \underline{f}_t, \underline{f}_{t+1} \mid
-    \underline{r}_t, \underline{r}_{t+1}
-) =
+    \bar{s}_t, \sigma_t, \mid 
+    \underline{f}_t^N, \underline{f}_{t+1}^N
+)
 \pi(
-    \bar{s}_t, \sigma_t, 
-    \underline{f}_t^N, \underline{f}_t^M, 
-    \underline{f}_{t+1}^N, \underline{f}_{t+1}^M \mid
-    \underline{r}_t, \underline{r}_{t+1}
+    \underline{f}_t^N, \underline{f}_t^M \mid \underline{r}_t
 )
+\pi(
+    \underline{f}_{t+1}^N, \underline{f}_{t+1}^M \mid \underline{r}_{t+1}
+),
 $${#eq-meanfit_split_f}
+where the first term on the right-hand side of @eq-meanfit_split_f only includes
+the neutral frequencies needed to infer the population mean fitness. The
+integral in @eq-meanfit_posterior_marginal can then be split for neutral and
+mutant lineages, resulting in a rather convoluted equation of the form 
+$$
+\begin{aligned}
+\pi(\bar{s}_t \mid \underline{r}_t, \underline{r}_{t+1}) =
+&\int d\sigma_t\;
+\int d^N \underline{f}_t^N \;
+\int d^N \underline{f}_{t+1}^N 
+\int d^M \underline{f}_t^M \;
+\int d^M \underline{f}_{t+1}^M \times \\
+&\pi(
+    \bar{s}_t, \sigma_t, \mid 
+    \underline{f}_t^N, \underline{f}_{t+1}^N
+)
+\pi(
+    \underline{f}_t^N, \underline{f}_t^M \mid \underline{r}_t
+)
+\pi(
+    \underline{f}_{t+1}^N, \underline{f}_{t+1}^M \mid \underline{r}_{t+1}
+),
+\end{aligned}
+$${#eq-meanfit_split_f_integral}
+where, again, we remind the reader that $N$ is the number of neutral barcodes
+and $M$ the number of mutant barcodes. Despite the convoluted form, we can
+rewrite @eq-meanfit_split_f_integral to get clarity on how to reduce the
+dimensionality of the problem even further. Distributing terms in
+@eq-meanfit_split_f_integral results in
+$$
+\begin{aligned}
+\pi(\bar{s}_t \mid \underline{r}_t, \underline{r}_{t+1}) =
+&\int d\sigma_t\;
+\int d^N \underline{f}_t^N \;
+\int d^N \underline{f}_{t+1}^N 
+\pi(
+    \bar{s}_t, \sigma_t, \mid 
+    \underline{f}_t^N, \underline{f}_{t+1}^N
+) \times \\
+&\int d^M \underline{f}_t^M \;
+\pi(
+    \underline{f}_t^N, \underline{f}_t^M \mid \underline{r}_t
+) \times \\
+&\int d^M \underline{f}_{t+1}^M \;
+\pi(
+    \underline{f}_{t+1}^N, \underline{f}_{t+1}^M \mid \underline{r}_{t+1}
+).
+\end{aligned}
+$$

### 2023-07-20 doc/_bayesian_inference.qmd
making progress on writing the paper

@@
     render-on-save: true
+bibliography: references.bib
+csl: ieee.csl
 ---
@@
 $${#eq-fitness_mut_vec}
-where $^\dagger$ indicates the transpose. Our data consists of an $B \times T$
+where $^\dagger$ indicates the transpose. Our data consists of an $T \times B$
 matrix $\underline{\underline{R}}$, where $B = M + N$ is the number of unique
@@
 \underline{\underline{R}} = \begin{bmatrix}
-\lvert & \lvert & \cdots & \lvert\\
-\underline{r}_1 & \underline{r}_2 & \cdots & \underline{r}_T\\
-\lvert & \lvert & \cdots & \lvert
+- & \underline{r}_1 & - \\
+- & \underline{r}_2 & - \\
+ & \vdots & \\
+- & \underline{r}_T & - \\
 \end{bmatrix},
 $${#eq-R_mat}
-where each column $\underline{r}_t$ is a $B$-dimensional array containing the
+where each row $\underline{r}_t$ is a $B$-dimensional array containing the
 raw barcode counts at time $t$. We can further split each vector
@@
 \underline{r}_t = \begin{bmatrix}
-\underline{r}_t^{M} \\
-\underline{r}_t^{N}
+\underline{r}_t^{N} \;
+\underline{r}_t^{M}
 \end{bmatrix},
 $${#eq-r-vec_split}
-i.e., the vector containing the mutant barcode counts $\underline{r}_t^{M}$ and
-the corresponding vector containing the neutral barcode counts
-$\underline{r}_t^{N}$. Our task is to compute the joint probability distribution
-of all relative fitness values given our data. Writing Bayes theorem this is
+i.e., the vector containing the neutral lineages barcode counts
+$\underline{r}_t^{N}$ and the corresponding vector containing the mutant barcode
+counts $\underline{r}_t^{M}$. Our objective is to compute the joint probability
+distribution of all relative fitness values given our data. We can express this 
+joint distribution using Bayes theorem as
 $$
@@
 $${#eq-bayes_obj}
-where hereafter $\pi(\cdot)$ defines the probability distribution  of the input
-unless otherwise stated. We can ignore the denominator on the right-hand side
-of @eq-bayes_obj, since it serves as a normalization constant that does not
-depend on our parameter of interest.
+where hereafter $\pi(\cdot)$ defines a probability density function unless
+otherwise stated. We can ignore the denominator on the right-hand side of
+@eq-bayes_obj, since it serves as a normalization constant that does not depend
+on our parameter of interest, simply writing
+$$
+\pi(\underline{s}^M \mid \underline{\underline{R}}) \propto
+\pi(\underline{\underline{R}} \mid \underline{s}^M) 
+\pi(\underline{s}^M)
+$${#eq-bayes_obj_propto}
 
-Although @eq-bayes_obj seems simple enough, recall that @eq-fitness relates
-frequency values and the population mean fitness to the mutant relative fitness.
-Therefore, we must include these nuisance parameters as part of our inference
-problem. At the end of the day, we can always marginalize the joint distribution
-to obtain the distribution we are interested in. To include these parameters,
-let 
+Although @eq-bayes_obj and @eq-bayes_obj_propto seem simple enough, recall that
+@eq-fitness relates barcode frequency values and the population mean fitness to
+the mutant relative fitness. Therefore, we must include these nuisance
+parameters as part of our inference problem. We can always marginalize the joint
+distribution to obtain the distribution we are interested in. To include these
+parameters, let 
 $$
@@
 of any $\bar{s}_t$ requires time points $t$ and $t+1$. Furthermore, let the
-matrix $\underline{\underline{F}}$ be a $B \times T$ matrix containing all
-frequency values. As with @eq-R_mat, each column of $\underline{\underline{F}}$
+matrix $\underline{\underline{F}}$ be a $T \times B$ matrix containing all
+frequency values. As with @eq-R_mat, each row of $\underline{\underline{F}}$
 can be split into two components of the form
@@
 \underline{f}_t = \begin{bmatrix}
-\underline{f}_t^M \\
+\underline{f}_t^M \;
 \underline{f}_t^N
@@
 $${#eq-f_vec}
-to split the corresponding mutant and neutral barcode frequencies. With these
+to separate the corresponding mutant and neutral barcode frequencies. With these
 variables in hand, the full inference problem we must solve takes the form
@@
 $${#eq-bayes_full}
-As a reminder, we need to include these extra nuisance parameters, but we can
-always integrate them out to recover the marginal distribution we care about,
-i.e.,
+We can numerically integrate out all nuisance parameters to recover the marginal
+distribution we care about, i.e.,
 $$
@@
 $${#eq-bayes_full_marginal}
-Although daunting at first sight, @eq-bayes_full_marginal will be solved 
-numerically, so we do not have to worry about this integration.
 
-### Splitting the posterior distribution {#seq-split_posterior}
+### Factorizing the posterior distribution {#seq-split_posterior}
 
-As written, the left-hand side of @eq-bayes_full is really difficult to work
-with. However, we can take advantage of the structure of our inference problem
-to rewrite it in a more manageable form. Specifically, we rewrite the joint
+The left-hand side of @eq-bayes_full is extremely difficult to work with.
+However, we can take advantage of the structure of our inference problem to
+rewrite it in a more manageable form. Specifically, we rewrite the joint
 probability distribution as a product of conditional distributions of the form
@@
 the uncertainty in the parameter we care about---the mutants' relative
-fitnesses. In the following sections, we work through each of these terms to
-build up the posterior distribution we will sample from piece by piece.
+fitnesses. We invite the reader to check the supplementary material where we
+work through each of these terms to build the posterior distribution to sample
+from.

### 2023-07-20 doc/_experiment.qmd
making progress on writing the paper

@@
+---
+editor:
+    render-on-save: true
+bibliography: references.bib
+csl: ieee.csl
+---
+
 ## Experimental setup
 
-`[add details on expeirmental setup]`
+Our inference pipeline is designed for the type of competition assays presented
+in @kinsler2020. In these experiments, an ensemble of genotypes is pooled
+together with an unlabeled reference strain that, initially, represents the vast
+majority ($\geq 90\%$) of the cells in the culture. Furthermore, a subfraction
+of labeled genotypes equivalent to the unlabeled reference strain---hereafter
+defined as *neutral* lineages---are spiked in at a relatively high abundance
+($\approx 3-5\%$), the rest of the culture being left for the ensemble of
+genotypes of interest.
+
+To determine the relative fitness of the ensemble of genotypes, a series of
+growth-dilution cycles are performed on either a single or multiple
+environments. In other words, the cultures are grown for some time to then
+dilute the culture back into fresh media and repeat for roughly 4-6 cycles,
+depending on the initial abundances of the mutants, and their relative growth
+rates. The DNA barcodes are sequenced at the end of each growth cycle to
+quantify the relative abundance of each of the barcodes. We point the reader to
+@kinsler2020 for specific details on these assays for *S. cerevisiae*. Figure
+[`ref here`] presents a typical barcode trajectory where the darker barcode
+trajectories represent the neutral lineages.

### 2023-07-20 doc/_fitness_model.qmd
making progress on writing the paper

--- 2023-03-25 doc/_fitness_model.qmd
@@
 ## Preliminaries on mathematical notation
-Before jumping directly into what concerns us here, let us establish the
-mathematical notation to be used throughout this document. We define (column)
-vectors as underlined lowercase symbols such as
+Before jumping directly into the Bayesian inference pipeline, let us establish
+the mathematical notation to be used throughout this paper. We define
+(column) vectors as underlined lowercase symbols such as
 $$
@@
 
-Empirically, we have seen that the barcode trajectories follow an exponential
+Empirically, the barcode relative frequency trajectories follow an exponential
 function of the form
@@
 where $f_{t}^{(b)}$ is the frequency of barcode $b$ at time $t$, $s^{(b)}$ is
-the relative fitness with respect to the ancestral strain---the quantity we want
+the relative fitness with respect to the reference strain---the quantity we want
 to infer from the data---$\bar{s}_t$ is the mean fitness of the culture at time
-$t$, and $\tau$ is the time pass between time $t$ and $t+1$. It is important to
-notice that the frequencies $f_{t}^{(b)}$ are not the allele frequencies in the
-population (most of the culture is not sequenced since it is not barcoded), but
-rather the inferred frequencies from the total number of reads. A way to 
-conceptualize this is that we assume exponential growth in the *number of cells*
+$t$, and $\tau$ is the time pass between time $t$ and $t+1$. We can rewrite
+@eq-fitness as
+$$
+\frac{1}{\tau}\ln \frac{f_{t+1}^{(b)}}{f_{t}^{(b)}} = (s^{(b)} - \bar{s}_t).
+$${#eq-logfreq}
+@eq-logfreq separates the observables---the barcode frequencies---from the
+latent parameters we want to infer from the data---the population mean fitness
+and the barcode relative fitness. This is ultimately the functional form used in
+our inference pipeline. Therefore, the relative fitness is computed by knowing
+the barcode log frequency ratio of each barcode throughout the growth-dilution
+cycles.
+
+It is important to notice that the frequencies $f_{t}^{(b)}$ are not the allele
+frequencies in the population (most of the culture is not sequenced since the
+reference strain is not barcoded), but rather the relative frequencies in the
+total number of sequencing reads. A way to conceptualize this subtle but
+important point is to assume exponential growth in the *number of cells*
 $N_t^{(b)}$ of the form
 $$
-N_{t+1}^{(b)} = N_{t}^{(b)} \mathrm{e}^{(s^{(b)} - \bar{s}_t)\tau},
+N_{t+1}^{(b)} = N_{t}^{(b)} \mathrm{e}^{\lambda^{(b)}\tau},
 $${#eq-ncells}
-for every barcode $b$. However, when we sequence barcodes, we do not directly
-measure the number of cells, but some number of reads $r_t^{(b)}$ that map to
-barcode $b$. In the simplest possible scenario, we assume
+for every barcode $b$ with growth rate $\lambda^{(b)}$. However, when we
+sequence barcodes, we do not directly measure the number of cells, but some
+number of reads $r_t^{(b)}$ that map to barcode $b$. In the simplest possible
+scenario, we assume
 $$
@@
 Therefore, to compare the number of reads between libraries at different time
-points, we must somehow normalize them to the same scale. The simplest form is
-to define a relative abundance, i.e., a frequency with respect to the total
-number of reads,
+points, we must normalize the number of reads to the same scale. The simplest
+form is to define a relative abundance, i.e., a frequency with respect to the
+total number of reads,
 $$
@@
 Our ultimate objective is to infer the relative fitness $s^{(m)}$ for each of
-the $M$ mutants in the experiment. To do so, we must take into account the three
-main sources of uncertainty in our model:
+the $M$ mutants in the experiment. To do so, we account the three main sources
+of uncertainty in our model:
 
@@
 frequencies between adjacent time points to the fitness of the corresponding
-strain. However, we do not measure frequencies directly. Instead, our data
-consists of a vector of barcode counts $\underline{r}_t$ for each of the $T$
-time points in which the measurements were taken.
-2. Uncertainty in the value of the population mean fitness. As we will see later
-on, we use neutral lineages barcodes, which, by definition have mean fitness
-$s^{(n)}$ to determine the value of the population mean fitness $\bar{s}_t$.
-This inference will come with its own uncertainty that must be propagated to the
-value of the relative fitness for the mutants.
-3. Uncertainty in each of the mutant's fitness values.
+strain. However, we do not directly measure frequencies. Instead, our data for
+each barcode consists of a vector of counts $\underline{r}^{(b)}_t$ for each of
+the $T$ time points in which the measurements were taken.
+2. Uncertainty in the value of the population mean fitness. We define neutral
+lineages to have fitness $s^{(n)} = 0$, helping us anchor the value of the
+population mean fitness $\bar{s}_t$ for each pair of adjacent time points. 
+Moreover, we take this parameter as an empirical observable from the data, 
+meaning that we do not have a functional form that relates $\bar{s}_t$ to
+$\bar{s}_{t+1}$. Thus, we must infer the $T-1$ values of this population mean 
+fitness with their uncertainty that must be propagated to the value of
+the mutants' relative fitness.
+3. Uncertainty in each of the mutants' fitness values.
 

### 2023-07-20 doc/_intro.qmd
making progress on writing the paper

---
editor:
    render-on-save: true
bibliography: references.bib
csl: ieee.csl
---

The advent of DNA barcoding and high-throughput sequencing has opened new venues
to understand microbial evolutionary dynamics with an unprecedented level of
temporal resolution [@levy2015; @nguyenba2019a; @ascensao2023]. These
experimental efforts rely on our ability to reliably infer the relative fitness
of an ensemble of diverse genotypes. Moreover, inferring these fitness values
over an ensemble of environmental conditions can help us determine the
phenotypic diversity of a rapid adaptation process @kinsler2020.

As with any other sequencing-based quantification, tracking lineages via DNA
barcode sequencing is inexorably accompanied by noise sources coming from
experimental manipulation of the microbial cultures, DNA extraction and
sequencing library preparation that involves multiple rounds of PCR
amplification, and the sequencing process itself. Thus, accounting for the
uncertainty when inferring the parameters we care about from the data is a key
step to drawing reliable conclusions. Bayesian statistics presents a paradigm
with which to account for all known sources of uncertainty in a principled way
@eddy2004a. This, combined with the development of modern Markov Chain Monte
Carlo sampling algorithms have boosted a resurgence of Bayesian methods in
different fields @efron2013a.

We present a Bayesian inference pipeline to properly quantify our uncertainty
about the parametric information we can extract from high-throughput competitive
fitness assays. In these assays, the fitness of an ensemble of genotypes is
determined relative to a reference genotype [@kinsler2020; @ascensao2023] [`add
reference to fig`]. Small changes to our model allow us to account for relevant
experimental variations of these competition assays: determination of multiple
fitness values per genotype for growth-dilution cycles on different
environments, and a hierarchical model to account for information from
independent experimental replicates. We benchmark the ability of our pipeline to
infer fitness parameters against synthetic data with known ground truth and
re-analyze published data from @kinsler2020 to assess the method's performance
on real experimental measurements. The method is accompanied with a `Julia`
package to readily implement the inference pipeline with state-of-the-art
scientific computing software. [`here mention something about the Python wrapper
if appropriate`].

### 2023-07-25 code/figs/fig02.jl
script to generate fig02

@@ -1,0 +1,483 @@
+##
+println("Loading packages...")
+
+# Load project package
+@load_pkg BayesFitUtils
+
+import Revise
+# Import project package
+import BayesFitUtils
+# Import package for Bayesian inference
+import BayesFitness
+
+# Import libraries to manipulate data
+import DataFrames as DF
+import CSV
+import MCMCChains
+
+# Import basic math
+import StatsBase
+
+# Import library to read MCMC chains
+import JLD2
+
+# Import plotting libraries
+using CairoMakie
+import ColorSchemes
+import ColorTypes
+
+# Activate backend
+CairoMakie.activate!()
+
+# Set PBoC Plotting style
+BayesFitUtils.viz.pboc_makie!()
+
+# Import library to set random seed
+import Random
+
+Random.seed!(42)
+
+##
+
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Loading data
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
+println("Loading data...")
+
+# Import data
+data = CSV.read(
+    "$(git_root())/data/logistic_growth/data_001/tidy_data.csv", DF.DataFrame
+)
+
+# Define file
+file = "$(git_root())/code/processing/mcmc_logistic_data001/output/" *
+       "chain_joint_fitness_1000steps_04walkers.jld2"
+
+# Load chain
+ids, chn = values(JLD2.load(file))
+# Remove the string "mut" from mutant names
+mut_num = replace.(ids, "mut" => "")
+
+# Find columns with mutant fitness values and error
+s_names = MCMCChains.namesingroup(chn, :s̲⁽ᵐ⁾)
+σ_names = MCMCChains.namesingroup(chn, :σ̲⁽ᵐ⁾)
+
+##
+
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Compute summary statistics for fitness values
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
[…]

### 2023-07-25 doc/_experiment.qmd
add references to fig01

@@
 together with an unlabeled reference strain that, initially, represents the vast
-majority ($\geq 90\%$) of the cells in the culture. Furthermore, a subfraction
-of labeled genotypes equivalent to the unlabeled reference strain---hereafter
-defined as *neutral* lineages---are spiked in at a relatively high abundance
-($\approx 3-5\%$), the rest of the culture being left for the ensemble of
-genotypes of interest.
+majority ($\geq 90\%$) of the cells in the culture (see schematic in
+@fig-fig01(A)). Furthermore, a subfraction of labeled genotypes equivalent to
+the unlabeled reference strain---hereafter defined as *neutral* lineages---are
+spiked in at a relatively high abundance ($\approx 3-5\%$), the rest of the
+culture being left for the ensemble of genotypes of interest.
 
@@
 quantify the relative abundance of each of the barcodes. We point the reader to
-@kinsler2020 for specific details on these assays for *S. cerevisiae*. Figure
-[`ref here`] presents a typical barcode trajectory where the darker barcode
+@kinsler2020 for specific details on these assays for *S. cerevisiae*.
+@fig-fig01(B) presents a typical barcode trajectory where the darker barcode
 trajectories represent the neutral lineages.
+
+![**Typical competitive fitness experiment**. (A) Schematic of the typical
+experimental design to determine the competitive fitness of an ensemble of
+barcoded genotypes. Genotypes are pooled together and grown over multiple
+growth-dilution cycles. At the end of each cycle, a sample is processed to
+generate a library for amplicon sequencing. (B) Typical barcode trajectory
+dataset. From each time point, the relative frequency of each barcode is
+determined from the total number of reads. Different shades of blue represent
+different relative fitness. Darker gray lines define the typical trajectory of
+neutral lineages.](./figs/fig01){#fig-fig01}

### 2023-07-26 doc/_bayesian_inference.qmd
continue to make progress on main results

@@
 from.
+
+### Markov Chain Monter Carlo
+
+One of the technical challenges to the adoption of Bayesian methods is the
+analytical intractability of integrals such as that of @eq-bayes_full_marginal.
+However, modern algorithms such as Hamiltonian Monte Carlo (HMC) allow for the
+efficient integration of arbitrary posterior distributions @betancourt2017. In
+particular, this work is accompanied by the Julia library `BayesFitness.jl` that
+makes use of the implementation of HMC samplers to numerically approximate the
+solution of @eq-bayes_full_marginal within the Julia ecosystem @ge2018.

### 2023-07-26 doc/_experiment.qmd
continue to make progress on main results

@@
 majority ($\geq 90\%$) of the cells in the culture (see schematic in
-@fig-fig01(A)). Furthermore, a subfraction of labeled genotypes equivalent to
+@fig-01(A)). Furthermore, a subfraction of labeled genotypes equivalent to
 the unlabeled reference strain---hereafter defined as *neutral* lineages---are
@@
 @kinsler2020 for specific details on these assays for *S. cerevisiae*.
-@fig-fig01(B) presents a typical barcode trajectory where the darker barcode
+@fig-01(B) presents a typical barcode trajectory where the darker barcode
 trajectories represent the neutral lineages.
@@
 different relative fitness. Darker gray lines define the typical trajectory of
-neutral lineages.](./figs/fig01){#fig-fig01}
+neutral lineages.](./figs/fig01){#fig-01}

### 2023-07-26 doc/_single_results.qmd
continue to make progress on main results

---
editor:
    render-on-save: true
bibliography: references.bib
csl: ieee.csl
---

### Inference on a single dataset

To assess the inference pipeline performance we applied it to a simulated
dataset with known ground truth (See `[ref to SI]` for details on simulation).
@fig-02(A) shows the structure of the synthetic dataset. The majority of
barcodes of interest (dim color lines) are adaptive compared to the neutral
barcodes ($s^{(m)} > 0$). This experimental condition is required to reliably
measure the relative fitness since all barcodes of interest occupy a small
fraction of the population at the beginning of the experiment `[ref SI section
on detrimental mutants]`. Although the barcode frequency trajectories look
relatively smooth, our fitness model requires the computation of the log
frequency ratio between adjacent time points. @fig-02(B) shows such data
transformation. Here, we can better appreciate the variability in the input that
goes into our model. This is especially clear for the darker lines representing
the neutral barcodes since they are supposed to be identically distributed under
the log frequency ratio transformation.

A way to visualize the performance of our inference pipeline in fitting our
fitness model to the observed data is to compute the so-called posterior
predictive checks (PPC). In short, the PPC consists of bootstrapping the
generation of synthetic datasets in agreement with the results from the
inference results. In other words, we use the resulting parameter values from
the MCMC chain to generate possible datasets in agreement with the inferred
values. @fig-02(C) shows these results for all neutral lineages (upper left
corner plot) and a few representative non–neutral barcodes. The dark color line
shows the median value we would expect the data to be found, while the shaded
regions show the 68\% and 95\% credible region, i.e. the regions where we expect
to find the data with the corresponding probability---or in terms of our
parameter, the $X\%$ credible region is the interval where we expect the true
parameter to lie with $X\%$ probability. The main advantage of our method is the
natural interpretability of these credible regions. A common mistake in the
literature is to confuse frequentist confidence intervals with Bayesian credible
regions when they are not equivalent @morey2016.

To capture the global performance of the model, @fig-02(D) compares the known
groud truth with the inferred relative fitness value for all barcodes of
interest. There is a great degree of correspondence between these values, with
the error bars, representing the 68\% credible region for the parameter value
crossing the identity line for most barcodes. This latter point is made clear
with @fig-02(E) where the empirical cumulative distribution function (ECDF) of
the absolute differences between the median inferred value and the true value
shows that all median inferred values are within 0.3 units of relative fitness
from the ground truth. Furthermore, because of the Bayesian scope of our method,
we do not need to limit ourselves to a point estimate of this quantity, but
report (asymmetric) error bars that fully quantify our degree of belief in the
inferred value as shown on top of the panels in @fig-02(C).

![**Single dataset inference**. (A) Frequency trajectories that represent the
raw data going into the inference. (B) Log frequency ratio between two adjacent
time points used by the inference pipeline. Darker lines represent the neutral
barcodes. Notice that these transformed data are much more noisy than the
seemingly smooth frequency trajectories. (B) Examples of the posterior
predictive checks for all neutral lineages (upper left panel) and a subset of
representative mutant lineages. Asymmetric error bars represent the 68\%
credible region on the mutant relative fitness marginal distribution. (C)
Comparison between the ground truth fitness value from the logistic-growth
simulation and the inferred fitness value. Gray error bars represent the 68\%
posterior credible region. (D) The empirical cumulative distribution function of
the absolute inference between the ground truth fitness value and the median of
the marginal distributions for each mutant. This plot reveals that all median
inferred values are within 0.3 units of relative fitness from the ground
truth.](./figs/fig02){#fig-02}

### 2023-09-06 doc/_bayesian_inference.qmd
making further progress on several sections

--- 2023-07-20 doc/_bayesian_inference.qmd
@@
 matrix $\underline{\underline{R}}$, where $B = M + N$ is the number of unique
-barcodes given by the sum of the number of unique mutant barcodes $M$ and the
-number of unique neutral barcodes $N$, and $T$ is the number of time points
-where measurements were taken. The data matrix is then of the form
+barcodes given by the sum of the number of unique relevant barcodes we care
+about (hereafter referred to as mutant barcodes), $M$, and the number of unique
+neutral barcodes, $N$, and $T$ is the number of time points where measurements
+were taken. The data matrix is then of the form
 $$
@@
 $\underline{r}_t^{N}$ and the corresponding vector containing the mutant barcode
-counts $\underline{r}_t^{M}$. Our objective is to compute the joint probability
-distribution of all relative fitness values given our data. We can express this 
-joint distribution using Bayes theorem as
+counts $\underline{r}_t^{M}$. Following the same logic, matrix
+$\underline{\underline{R}}$ can be split into two matrices as
+$$
+\underline{\underline{R}} = \left[ 
+\underline{\underline{R}}^N \; \underline{\underline{R}}^M
+\right],
+$${#eq-R-mat_split}
+where $\underline{\underline{R}}^N$ is a $T \times N$ matrix with the
+barcode reads time series for each neutral lineage and
+$\underline{\underline{R}}^M$ is the equivalent $T \times M$ matrix for the
+mutant lineages.
+
+Our objective is to compute the joint probability
+distribution for all relative fitness values given our data. We can express this 
+joint posterior distribution using Bayes theorem as
 $$
@@
 $${#eq-bayes_obj}
-where hereafter $\pi(\cdot)$ defines a probability density function unless
-otherwise stated. We can ignore the denominator on the right-hand side of
-@eq-bayes_obj, since it serves as a normalization constant that does not depend
-on our parameter of interest, simply writing
+where hereafter $\pi(\cdot)$ defines a probability density function, unless
+otherwise stated. To develop our model, we can ignore the denominator on the
+right-hand side of @eq-bayes_obj, since it serves as a normalization constant
+that does not depend on our parameter of interest. However, computing this 
+normalization constant represents a major challenge we will have to overcome. 
+For now, we can simply write
 $$
@@
 matrix $\underline{\underline{F}}$ be a $T \times B$ matrix containing all
-frequency values. As with @eq-R_mat, each row of $\underline{\underline{F}}$
-can be split into two components of the form
+frequency values. As with @eq-R_mat, we can split $\underline{\underline{F}}$
+into two matrices of the form
 $$
-\underline{f}_t = \begin{bmatrix}
-\underline{f}_t^M \;
-\underline{f}_t^N
-\end{bmatrix},
-$${#eq-f_vec}
+\underline{\underline{F}} = \left[ 
+\underline{\underline{F}}^N \; \underline{\underline{F}}^M
+\right],
+$${#eq-F-mat_split}
 to separate the corresponding mutant and neutral barcode frequencies. With these
-variables in hand, the full inference problem we must solve takes the form
+nuisance variables in hand, the full inference problem we must solve takes the
+form
 $$
@@
 However, we can take advantage of the structure of our inference problem to
-rewrite it in a more manageable form. Specifically, we rewrite the joint
-probability distribution as a product of conditional distributions of the form
+rewrite it in a more manageable form. Specifically, as depicted in @fig-01(C),
+the statistical dependencies of our observations and latent variables allows us
+to rewrite the joint probability distribution as a product of conditional
+distributions of the form
 $$
@@
 \pi(
-    \underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}},
-    \underline{\underline{R}}
+    \underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}}^M,
+    \underline{\underline{R}}^M
 )
 \pi(
-    \underline{\bar{s}}_T \mid \underline{\underline{F}}, 
-    \underline{\underline{R}}
+    \underline{\bar{s}}_T \mid \underline{\underline{F}}^N,
+    \underline{\underline{R}}^N
 )
@@
 work through each of these terms to build the posterior distribution to sample
-from.
+from. We refer the reader to `[ref SI section]` for an extended description of
+the model with specific functional forms for each term on the left-hand side of
+@eq-split_posterior as well as `[ref SI on hierarchical]` where the model is
+extended to account for multiple experimental replicates.
+
+### Variational Inference
+
+One of the technical challenges to the adoption of Bayesian methods is the
+analytical intractability of integrals such as that of @eq-bayes_full_marginal.
+Furthermore, even though efficient Markov Chain Monte Carlo (MCMC) algorithms
+such as Hamiltonian Montecarlo can numerically perform this integration
+[@betancourt2017], the dimensionality of the problem in @eq-split_posterior makes
+an MCMC-based approach prohibitively slow.
+
+To overcome this computational limitation, in this work, we rely on the recent
+development of the automatic differentiation variational inference algorithm
+(ADVI) [@kucukelbir2016]. Briefly, when performing ADVI, our target posterior
+distribution $\pi(\theta \mid \underline{\underline{R}})$, where $\theta =
+(\underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}})$, is
+replaced by an approximate posterior distribution $q_\phi(\theta \mid
+\underline{\underline{R}})$, where $\phi$ fully parametrizes the approximate
+distribution. As further explained in `[point to SI section]`, the numerical
+integration problem is replaced by an optimization problem of the form
+$$
+q^*_\phi(\theta \mid \underline{\underline{R}}) = \min _\phi
+D_{KL}(
+    q_\phi(\theta \mid \underline{\underline{R}}) \lvert \lvert
+    \pi(\theta \mid \underline{\underline{R}})
+),
+$${#eq-vi_objective}
+where $D_{KL}$ is the Kulback-Leibler divergence. In other words, the 
+complicated high-dimensional numerical integration problem is transformed into
+a gradient descent over the parameters of the approximate posterior distribution
+$\phi$ such that it minimizes the KL divergence with respect to the true 
+posterior distribution. Although to compute @eq-vi_objective we require the
+posterior distribution we are trying to approximate $\pi(\theta \mid
+\underline{\underline{R}})$, it can be shown that maximizing the so-called
+evidence lower bound (ELBO) [@kingma2014]---equivalent to minimizing the 
+variational free energy [@buckley2017]---is mathematically equivalent to 
+performing the optimization prescribed by @eq-vi_objective. We direct the reader
+to `[point to SI section]` for a short primer on variational inference.
+
+This work is accompanied by the Julia library `BayesFitness.jl` that makes use
+of the implementation of both MCMC-based integration as well as ADVI
+optimization to numerically approximate the solution of @eq-bayes_full_marginal
+within the Julia ecosystem [@ge2018].

### 2023-09-06 doc/_experiment.qmd
making further progress on several sections

@@
 environments. In other words, the cultures are grown for some time to then
-dilute the culture back into fresh media and repeat for roughly 4-6 cycles,
+dilute the culture back into fresh media and repeat for roughly 4-7 cycles,
 depending on the initial abundances of the mutants, and their relative growth
@@
 different relative fitness. Darker gray lines define the typical trajectory of
-neutral lineages.](./figs/fig01){#fig-01}
+neutral lineages. (C) Bayesian network defining the statistical dependence
+between observations (filled circles) and latent variables (empty circles). The
+dotted line highlights the set of variables we are interested in. See
+@sec-bayesian_inference for specific definitions of each
+variable.](./figs/fig01){#fig-01}

### 2023-09-06 doc/_fitness_model.qmd
making further progress on several sections

@@
 Before jumping directly into the Bayesian inference pipeline, let us establish
-the mathematical notation to be used throughout this paper. We define
+the mathematical notation used throughout this paper. We define
 (column) vectors as underlined lowercase symbols such as
@@
 $${#eq-vec_def}
-
 In the same way, we define matrices as double-underline uppercase symbols such
-as $\underline{\underline{A}}$.
+as
+$$
+\underline{\underline{A}} =
+\begin{bmatrix}
+    A_{11} & A_{12} & \cdots & A_{1N}\\
+    A_{21} & A_{22} & \cdots & A_{2N}\\
+    \vdots & \vdots & \ddots & \vdots\\
+    A_{M1} & A_{M2} & \cdots & A_{MN}\\
+\end{bmatrix}.
+$${#eq-mat_def}
 
@@
 our inference pipeline. Therefore, the relative fitness is computed by knowing
-the barcode log frequency ratio of each barcode throughout the growth-dilution
-cycles.
+the log frequency ratio of each barcode throughout the growth-dilution cycles.
 
@@
 Our ultimate objective is to infer the relative fitness $s^{(m)}$ for each of
-the $M$ mutants in the experiment. To do so, we account the three main sources
-of uncertainty in our model:
+the $M$ relevant barcodes in the experiment. To do so, we account for the three
+primary sources of uncertainty in our model:
 
@@
 strain. However, we do not directly measure frequencies. Instead, our data for
-each barcode consists of a vector of counts $\underline{r}^{(b)}_t$ for each of
-the $T$ time points in which the measurements were taken.
+each barcode consists of a length $T$ vector of counts $\underline{r}^{(b)}$ for
+each of the $T$ time points in which the measurements were taken.
 2. Uncertainty in the value of the population mean fitness. We define neutral
@@
 
-To account for all these sources of uncertainty in a principled way, we develop
-a Bayesian inference pipeline.
+To account for all these sources of uncertainty in a principled way, in the next
+section, we develop a Bayesian inference pipeline.

### 2023-09-06 doc/_hierarchical_results.qmd
making further progress on several sections

---
editor:
    render-on-save: true
bibliography: references.bib
csl: ieee.csl
---

### Accounting for experimental replicates via hierarchical models

Our inference pipeline can be extended to account for multiple experimental
replicates via Bayesian hierarchical models [@betancourt2013]. Briefly, when
accounting for multiple repeated measurements of the same phenomena, there are
two extreme cases one can use to perform the data analysis: On the one hand, we
can treat each measurement as completely independent, losing the power to
utilize multiple measurements when trying to learn a single parameter. This can
hurt the inference since, in principle, the value of our parameter of interest
should not depend on the particular experimental replicate in question. However,
this approach does not allow us to properly "combine" the uncertainties in both
experiments when performing the inference. On the other hand, we can pool all
data together and treat our different experiments as a single measurement with
higher coverage. This loses the subtle differences that come from biotic and
abiotic batch effects, effectively halving the data that goes into our inference
problem.

Hierarchical models present a middle ground between these extremes. First,
hierarchical models rely on the definition of so-called *hyper-parameters*, that
capture the ultimate parameter we are interested in---in our case we have a
hyper-fitness value $\theta^{(m)}$ for each mutant. Second, each experiment
draws randomly from the distribution of this hyper-parameter, allowing for
subtle variability between experiments to be accounted for---for our inference
pipeline, each experimental replicate gets assigned a fitness value $s^{(m,i)}$,
where the extra index indicates the $i$-th experimental replicate.

![**Hierarchical model on experimental replicates**.](./figs/fig03){#fig-03}

### 2023-09-06 doc/_intro.qmd
making further progress on several sections

@@
 over an ensemble of environmental conditions can help us determine the
-phenotypic diversity of a rapid adaptation process @kinsler2020.
+phenotypic diversity of a rapid adaptation process [@kinsler2020].
 
@@
 with which to account for all known sources of uncertainty in a principled way
-@eddy2004a. This, combined with the development of modern Markov Chain Monte
-Carlo sampling algorithms have boosted a resurgence of Bayesian methods in
-different fields @efron2013a.
+[@eddy2004a]. This, combined with the development of modern Markov Chain Monte
+Carlo sampling algorithms [@betancourt2017] and approximate variational
+approaches [@kucukelbir2016] have boosted a resurgence of Bayesian methods in
+different fields [@efron2013a].
 
@@
 fitness assays. In these assays, the fitness of an ensemble of genotypes is
-determined relative to a reference genotype [@kinsler2020; @ascensao2023] [`add
-reference to fig`]. Small changes to our model allow us to account for relevant
-experimental variations of these competition assays: determination of multiple
-fitness values per genotype for growth-dilution cycles on different
-environments, and a hierarchical model to account for information from
-independent experimental replicates. We benchmark the ability of our pipeline to
+determined relative to a reference genotype [@kinsler2020; @ascensao2023] (See
+@fig-01 for a schematic depiction of the experimental setup). Small changes to
+our model allows us to account for relevant experimental variations of these
+competition assays: determination of multiple fitness values per genotype for
+growth-dilution cycles on different environments, a hierarchical model to
+account for information from independent experimental replicates, and a
+hierarchical model for a within-experiment structure where multiple independent
+barcodes map to a specific genotype. We benchmark the ability of our pipeline to
 infer fitness parameters against synthetic data with known ground truth and
 re-analyze published data from @kinsler2020 to assess the method's performance
-on real experimental measurements. The method is accompanied with a `Julia`
+on real experimental measurements. The method is accompanied by a `Julia`
 package to readily implement the inference pipeline with state-of-the-art

### 2023-09-06 doc/_single_results.qmd
making further progress on several sections

@@
 To assess the inference pipeline performance we applied it to a simulated
-dataset with known ground truth (See `[ref to SI]` for details on simulation).
-@fig-02(A) shows the structure of the synthetic dataset. The majority of
-barcodes of interest (dim color lines) are adaptive compared to the neutral
-barcodes ($s^{(m)} > 0$). This experimental condition is required to reliably
-measure the relative fitness since all barcodes of interest occupy a small
-fraction of the population at the beginning of the experiment `[ref SI section
-on detrimental mutants]`. Although the barcode frequency trajectories look
-relatively smooth, our fitness model requires the computation of the log
-frequency ratio between adjacent time points. @fig-02(B) shows such data
-transformation. Here, we can better appreciate the variability in the input that
-goes into our model. This is especially clear for the darker lines representing
-the neutral barcodes since they are supposed to be identically distributed under
-the log frequency ratio transformation.
+dataset with known ground truth relative fitness values (See `[ref to SI]` for
+details on simulation). @fig-02(A) shows the structure of the synthetic dataset.
+The majority of barcodes of interest (dim color lines) are adaptive compared to
+the neutral barcodes ($s^{(m)} > 0$). This experimental condition is required to
+reliably measure the relative fitness since all barcodes of interest occupy a
+small fraction of the population at the beginning of the experiment `[ref SI
+section on detrimental mutants]`. Although the barcode frequency trajectories
+look relatively smooth, our fitness model requires the computation of the log
+frequency ratio between adjacent time points as derived in @eq-logfreq.
+@fig-02(B) shows such data transformation where we can better appreciate the
+observational noise input into our statistical model. This noise is
+evident for the darker lines representing the neutral barcodes since
+all of these lineages are assumed to be identically distributed under this log
+frequency ratio transformation.
 
@@
 inference results. In other words, we use the resulting parameter values from
-the MCMC chain to generate possible datasets in agreement with the inferred
+the ADVI inference to generate possible datasets in agreement with the inferred
 values. @fig-02(C) shows these results for all neutral lineages (upper left
-corner plot) and a few representative non–neutral barcodes. The dark color line
+corner plot) and a few representative non-neutral barcodes. The dark color line
 shows the median value we would expect the data to be found, while the shaded
-regions show the 68\% and 95\% credible region, i.e. the regions where we expect
-to find the data with the corresponding probability---or in terms of our
+regions show the 68\% and 95\% credible regions, i.e., the regions where we
+expect to find the data with the corresponding probability---or in terms of our
 parameter, the $X\%$ credible region is the interval where we expect the true
-parameter to lie with $X\%$ probability. The main advantage of our method is the
-natural interpretability of these credible regions. A common mistake in the
-literature is to confuse frequentist confidence intervals with Bayesian credible
-regions when they are not equivalent @morey2016.
+parameter to lie with $X\%$ probability. The main advantage of our method is
+this natural interpretability of these credible regions since a common mistake
+in the literature is to interpret frequentist confidence intervals as Bayesian
+credible regions when they are not equivalent [@morey2016].
 
 To capture the global performance of the model, @fig-02(D) compares the known
-groud truth with the inferred relative fitness value for all barcodes of
-interest. There is a great degree of correspondence between these values, with
-the error bars, representing the 68\% credible region for the parameter value
-crossing the identity line for most barcodes. This latter point is made clear
-with @fig-02(E) where the empirical cumulative distribution function (ECDF) of
-the absolute differences between the median inferred value and the true value
-shows that all median inferred values are within 0.3 units of relative fitness
-from the ground truth. Furthermore, because of the Bayesian scope of our method,
-we do not need to limit ourselves to a point estimate of this quantity, but
-report (asymmetric) error bars that fully quantify our degree of belief in the
-inferred value as shown on top of the panels in @fig-02(C).
+ground truth with the inferred relative fitness value for all barcodes of
+interest. There is an excellent degree of correspondence between these values,
+with the error bars, representing the 68\% credible region for the parameter
+value crossing the identity line for most barcodes. This latter point is made
+clear with @fig-02(E) where the empirical cumulative distribution function
+(ECDF) of the absolute differences between the median inferred value and the
+actual value shows that all median inferred values are within 0.3 units of
+relative fitness from the ground truth.
 

### 2023-09-07 doc/_single_results.qmd
making further progress on results section

@@
 
-### Inference on a single dataset
+## Inference on a single dataset
 
@@
 value crossing the identity line for most barcodes. This latter point is made
-clear with @fig-02(E) where the empirical cumulative distribution function
-(ECDF) of the absolute differences between the median inferred value and the
-actual value shows that all median inferred values are within 0.3 units of
-relative fitness from the ground truth.
+clear with @fig-02(E) where $\approx 90\%$ of ground truth fitness values fall
+within one standard deviation of the mean in the inferred posterior
+distributions.
 
@@
 predictive checks for all neutral lineages (upper left panel) and a subset of
-representative mutant lineages. Asymmetric error bars represent the 68\%
-credible region on the mutant relative fitness marginal distribution. (C)
-Comparison between the ground truth fitness value from the logistic-growth
-simulation and the inferred fitness value. Gray error bars represent the 68\%
-posterior credible region. (D) The empirical cumulative distribution function of
-the absolute inference between the ground truth fitness value and the median of
-the marginal distributions for each mutant. This plot reveals that all median
-inferred values are within 0.3 units of relative fitness from the ground
-truth.](./figs/fig02){#fig-02}
+representative mutant lineages. Error bars represent the 68\% credible region on
+the mutant relative fitness marginal distribution. (C) Comparison between the
+ground truth fitness value from the logistic-growth simulation and the inferred
+fitness value. Gray error bars represent the 68\% posterior credible region. (D)
+The empirical cumulative distribution function (ECDF) for the absolute z-score
+value of the ground truth parameter value within the inferred fitness
+posterior distribution.](./figs/fig02){#fig-02}

### 2023-09-08 doc/_hierarchical_replicates_results.qmd
rename section

---
editor:
    render-on-save: true
bibliography: references.bib
csl: ieee.csl
---

## Accounting for experimental replicates via hierarchical models

Our inference pipeline can be extended to account for multiple experimental
replicates via Bayesian hierarchical models [@betancourt2013]. Briefly, when
accounting for multiple repeated measurements of the same phenomena, there are
two extreme cases one can use to perform the data analysis: On the one hand, we
can treat each measurement as completely independent, losing the power to
utilize multiple measurements when trying to learn a single parameter. This can
hurt the inference since, in principle, the value of our parameter of interest
should not depend on the particular experimental replicate in question. However,
this approach does not allow us to properly "combine" the uncertainties in both
experiments when performing the inference. On the other hand, we can pool all
data together and treat our different experiments as a single measurement with
higher coverage. This loses the subtle differences that come from biotic and
abiotic batch effects, effectively halving the data that goes into our inference
problem.

Hierarchical models present a middle ground between these extremes. First,
hierarchical models rely on the definition of so-called *hyper-parameters*, that
capture the ultimate parameter we are interested in---for our case we have a
hyper-fitness value $\theta^{(m)}$ for each mutant. Second, each experiment
draws randomly from the distribution of this hyper-parameter, allowing for
subtle variability between experiments to be accounted for---for our inference
pipeline, each experimental replicate gets assigned a fitness value $s^{(m,i)}$,
where the extra index indicates the $i$-th experimental replicate. As
schematized by the Bayesian network on @fig-03(A) and further detailed in
`[point to SI section]`, our statistical model can easily be extended to a
hierarchical structure, where the fitness of each relevant barcode on each
experimental replicate is a sample from a global hyper-fitness parameter.

To test the performance of this model, we simulated two experimental replicates
with 1000 unique barcodes (see @fig-03(B-C)) where we randomly sampled a ground
truth hyper-fitness value $\theta^{(m)}$ for each barcode. To capture
experimental batch effects, we sampled a variation from this hyper-fitness value
for each experimental replicate $s^{(m, i)}$. @fig-03(D) shows the relationship
between hyper-fitness and replicate fitness values for this simulation. As shown
by the posterior predictive checks examples in @fig-03(E), the hierarchical
model can correctly fit the data for each experimental replicate. Furthermore,
@fig-03(F-G) show a high degree of correlation between the ground truth and the
inferred fitness values. The ECDFs shown in @fig-03(H-I) reveal that for
$\approx 75\%$ of the non-neutral barcodes, the ground truth hyper-fitness
values fall within one standard deviation from the mean value in the posterior
distributions.

![**Hierarchical model on experimental replicates**. (A) Bayesian network
schematizing the statistical dependence between observations (filled circles)
and latent variables (empty circles). The replicates are connected via the
hyper-parameters $\underline{\theta}^M$. See `point to SI` for detailed
definitions of each variable. (B-C) Simulated replicate datasets with 900
barcodes of interest and 100 neutral lineages. (D) Comparison between the
simulation ground truth hyper-fitness and each replicate ground truth fitness.
The scatter between parameters captures experimental batch effects. (E) Examples
of the posterior predictive checks for all neutral lineages (upper left panels)
and a subset of representative mutant lineages. Shaded regions from light to
dark represent the 95\%, 68\% and 5\% credible regions. (F-G) Comparison between
the ground truth hyperfitness (F) and replicate fitness (G) values from the
simulation and the inferred parameters. Gray error bars represent the 68\%
posterior credible region. (H-I) The empirical cumulative distribution function
(ECDF) for the absolute z-score value of the ground truth parameter value within
the inferred hyper-fitness posterior distribution (H) and replicate fitness
(I).](./figs/fig03){#fig-03}

As shown in @fig-04, the structure imposed by the hierarchical model, where we
explicitly account for the connection between experimental replicates can
improve the quality of the inference.  Inferred fitness values between
experimental replicate exhibit a stronger degree of correlation in the
hierarchical model (@fig-04(A)) compared to conducting inference on each
replicate independently (@fig-04(B)). Moreover, when comparing the inferred
hyper-fitness values---the objective parameter when performing multiple
experimental measurements---the hierarchical model outperforms averaging the
independent experimental replicates as shown in @fig-04(C) and (D).

![**Comparison between hierarchical model and single dataset model**. (A-B)
comparison of inferred fitness values between experimental replicates when
fitting a hierarchical model (A) or independently fitting each dataset (B). Gray
error bars represent the 68\% credible regions. (C) Comparison between the
ground truth hyper-fitness value and the inferred parameters. The blue dots show
the inferred hyper-fitness values when assuming a hierarchical model. Gray error
bars show the 68\% credible region for this inference. The yellow dots show the
average of the mean inferred fitness values for the two experimental replicates.
No error bars are shown for these as it is inappropriate to compute one with two
data points. (D) Empirical cumulative distribution function (ECDF) of the
absolute difference between the inferred mean and the ground truth
hyper-fitness.](./figs/fig04){#fig-04}

### 2023-09-11 doc/_hierarchical_genotypes_results.qmd
continue to make progress on paper

---
editor:
    render-on-save: true
bibliography: references.bib
csl: ieee.csl
---

## Accounting for multiple barcodes per genotype via hierarchical models

![**Hierarchical model for multiple barcodes per
genotype**](./figs/fig05){#fig-05}

### 2023-09-11 doc/_variational_inference_primer.qmd
continue to make progress on paper

---
editor:
    render-on-save: true
bibliography: references.bib
csl: ieee.csl
---

## Primer on Variational Inference

In this section, we will briefly introduce the idea behind variational
inference. Recall that any Bayesian inference problem deals with the joint
distribution between observations $\underline{x}$ and unobserved latent
variables $\underline{\theta}$. This joint distribution can be written as the
product of a distribution of the observations $\underline{x}$ conditioned on the
$\underline{\theta}$ and the marginal distribution of these latent variables,
i.e.,
$$
\pi(\underline{x}, \underline{\theta}) =
\pi(\underline{x} \mid \underline{\theta}) \pi(\underline{\theta}).
$${#eq-vi_joint_dist}
A Bayesian inference pipeline's objective is to compute the latent variables'
posterior probability given a set of observations. This computation is
equivalent to updating our prior beliefs about the set of values that the latent
variables take after taking in new data. We write this as Bayes theorem
$$
\pi(\underline{\theta} \mid \underline{x}) = 
\frac{
        \pi(\underline{x} \mid \underline{\theta})\pi(\underline{\theta})
    }{
        \pi(\underline{x})
    }.
$${#eq-vi_bayes_theorem}
The main technical challenge for working with @eq-vi_bayes_theorem comes from
the computation of the denominator, also known as the *evidence* or the
*marginalized likelihood*. The reason computing this term is challenging is
because it involves a (potentially) high-dimensional integral of the form
$$
\pi(\underline{x}) = 
\int\cdots\int d\underline{\theta}^K \pi(\underline{x}, \underline{\theta}) = 
\int\cdots\int d\underline{\theta}^K \pi(\underline{x} \mid \underline{\theta})
\pi(\underline{\theta}),
$${#eq-vi_evidence_int}
where $K$ is the dimesionality of the $\underline{\theta}$ vector. Here, the
integrals are taken over the support---the set of values valid for the
distribution---of $\pi(\underline{\theta})$. However, only a few selected
distributions have a closed analytical form; thus, in most cases
@eq-vi_evidence_int must be solved numerically.

Integration in high-dimensional spaces can be computationally extremely
challenging. For a naive numerical quadrature procedure, integrating over a grid
of values for each dimension of $\underline{\theta}$ comes with an exponential
explosion of the number of required grid point evaluations, most of which do not
contribute significantly to the integration. To gain visual intuition about this
challenge, imagine integrating the function depicted in @fig-SI01. If the
location of the high-density region (dark peak) is unknown, numerical quadrature
requires many grid points to ensure we capture this peak. However, most of the
numerical evaluations of the function on the grid points do not contribute
significantly to the integral. Therefore, our computational resources are wasted
on insignificant evaluations. This only gets worse as the number of dimensions
increases since the number of grid point evaluation scales exponentially.

![**High-dimensional numerical quadrature does not scale with dimensionality.**
Schematic depiction of the problem with naive numerical quadrature to integrate
over an unknown density. While the density is concentrated on the dark peak,
most of the evaluations over the $x_1-x_2$ grid do not contribute to the value
of the integral](./figs/figSIX_gaussian_peak.png){#fig-SI01}

Modern Markov Chain Monte Carlo algorithms, such as Hamiltonian Monte Carlo, can
efficiently perform this high-dimensional integration by utilizing gradient
information from the target density @betancourt2017. Nevertheless, these
sampling-based methods become prohibitively slow for the number of dimensions
our present inference problem presents. Thus, there is a need to find scalable
methods for the inference problem in @eq-vi_bayes_theorem.

Variational inference circumvents these technical challenges by proposing an
approximate solution to the problem. Instead of working with the posterior
distribution in its full glory $\pi(\underline{\theta} \mid \underline{x})$, let
us propose an approximate posterior distribution $q_\phi$ that belongs to a
distribution family fully parametrized by $\phi$. For example, let us say that
the distribution $q_\phi$ belongs to the family of multivariate Normal
distributions such that $\phi = (\underline{\mu},
\underline{\underline{\Sigma}})$, where $\underline{\mu}$ is the vector of means
and $\underline{\underline{\Sigma}}$ is the covariance matrix. If we replace
$\pi$ by $q_\phi$, we want $q_\phi$ to resemble the original posterior as much
as possible. Mathematically, this can be expressed as minimizing a "*distance
metric*"---the Kullback-Leibler (KL) divergence, for example---between the
distributions. Note that we use quotation marks because, formally, the KL
divergence is not a distance metric since it is not symmetric. Nevertheless, the
variational objective is set to find a distribution $q_\phi^*$ such that
$$
q_\phi^*(\underline{\theta} \mid \underline{x}) =
\min_\phi D_{KL}\left(
    q_\phi(\underline{\theta} \mid \underline{x}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right),
$${#eq-vi_objective}
where $D_{KL}$ is the KL divergence. We alert the reader that in the literature
$q_\phi$ is usually  written not conditioned on the observations, i.e.,
$q_\phi(\underline{\theta})$. But we think that writing the conditioning
reflects that the selection of the distribution $q_\phi^*$ out of the parametric
family depends on having the observations $\underline{x}$ in hand. Furthermore,
we highlight that the KL divergence is a strictly positive number, i.e.,
$$
D_{KL}\left(
    q_\phi(\underline{\theta} \mid \underline{x}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) \geq 0,
$${#eq-vi_kl_pos}
as this property will become important later on.

At first sight, @eq-vi_objective does not improve the situation but only
introduces further technical complications. After all, the definition of the KL
divergence
$$
D_{KL}\left(
    q_\phi(\underline{\theta} \mid \underline{x}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) \equiv 
\int \cdots \int d\underline{\theta}^K
q_\phi(\underline{\theta} \mid \underline{x})
\ln \frac{
    q_\phi(\underline{\theta} \mid \underline{x})
}{
    \pi(\underline{\theta} \mid \underline{x})
},
$${#eq-vi_kl_div}
includes the posterior distribution $\pi(\underline{\theta} \mid \underline{x})$
we are trying to get around. However, let us manipulate @eq-vi_kl_div to beat it
to a more reasonable form. First, we can use the properties of the logarithms to
write
$$
D_{KL}\left(
    q_\phi(\underline{\theta} \mid \underline{x}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) = 
\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
\ln q_\phi(\underline{\theta} \mid \underline{x}) -
\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
\ln \pi(\underline{\theta} \mid \underline{x}),
$${#eq-vi_step01}
where, for convenience, we write a single integration sign
($d\underline{\theta}^K$ still represents a multi-dimensional differential). For
the second term in @eq-vi_step01, we can substitute the term inside the
logarithm using @eq-vi_bayes_theorem. This results in
$$
\begin{aligned}
D_{KL}\left(
    q_\phi(\underline{\theta} \mid \underline{x}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) &= 
\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
\ln q_\phi(\underline{\theta} \mid \underline{x}) \\
&- \int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
\ln \left( 
    \frac{
        \pi(\underline{x} \mid \underline{\theta})\pi(\underline{\theta})
    }{
        \pi(\underline{x})
    }
\right).
\end{aligned}
$${#eq-vi_step02}
Again, using the properties of logarithms, we can split @eq-vi_step02, obtaining
$$
\begin{aligned}
D_{KL}\left(
    q_\phi(\underline{\theta} \mid \underline{x}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) &= 
\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
\ln q_\phi(\underline{\theta} \mid \underline{x}) \\
&-\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
\ln \pi(\underline{x} \mid \underline{\theta}) \\
&-\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
\ln \pi(\underline{\theta}) \\
&+\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
\ln \pi(\underline{x}).
\end{aligned}
$${#eq-vi_step03}
It is convenient to write @eq-vi_step03 as
$$
\begin{aligned}
D_{KL}\left(
    q_\phi(\underline{\theta} \mid \underline{x}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) &= 
\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
\ln \frac{
    q_\phi(\underline{\theta} \mid \underline{x})
    }{
        \pi(\underline{\theta})
    } \\
&-\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
\ln \pi(\underline{x} \mid \underline{\theta}) \\
&+ \ln \pi(\underline{x}) 
\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x}),
\end{aligned}
$${#eq-vi_step04}
where for the last term, we can take $\ln \pi(\underline{x})$ out of the
integral since it does not depend on $\underline{\theta}$. Lastly, we utilize
two properties:

1. The proposed approximate distribution must be normalized, i.e.,
$$
\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x}) = 1.
$${#eq-vi_q_norm}
2. The law of the unconscious statistician (LOTUS) establishes that for any
probability density function, it must be true that
$$
\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
f(\underline{\theta}) = \left\langle 
    f(\underline{\theta}) 
\right\rangle_{q_\phi},
$${#eq-vi_lotus}
where $\left\langle\cdot\right\rangle_{q_\phi}$ is the expected value over the
$q_\phi$ distribution.

Using these two properties, the positivity constraint on the KL divergence in
@eq-vi_kl_pos, and the definition of the KL divergence in @eq-vi_kl_div we can
rewrite @eq-vi_step04 as
$$
D_{KL}\left( 
    q_\phi(\underline{\theta} \mid \underline{x}) \vert \vert
    \pi(\underline{\theta}) 
\right) -
\left\langle
    \ln \pi(\underline{x} \mid \underline{\theta})
\right\rangle_{q_\phi}
\geq - \ln \pi(\underline{x}).
$${#eq-vi_step05}
Multiplying by a minus one, we have the functional form of the so-called
evidence lower bound (ELBO) @kingma2014,
$$
\underbrace{
    \ln \pi(\underline{x})
}_{\text{log evidence}} \geq
\underbrace{
    \left\langle
        \ln \pi(\underline{x} \mid \underline{\theta})
    \right\rangle_{q_\phi} -
    D_{KL}\left( 
        q_\phi(\underline{\theta} \mid \underline{x}) \vert \vert
        \pi(\underline{\theta}) 
    \right)
}_{\text{ELBO}}.
$${#eq-vi_elbo}

Let us recapitulate where we are. We started by presenting the challenge of
working with Bayes' theorem, as it requires a high-dimensional integral of the
form in @eq-vi_evidence_int. As an alternative, variational inference posits to
approximate the posterior distribution $\pi(\underline{\theta} \mid
\underline{x})$ with a parametric distribution $q_\phi(\underline{\theta} \mid
\underline{x})$. By minimizing the KL divergence between these distributions, we
arrive at the result in @eq-vi_elbo, where the left-hand side---the log
marginalized likelihood or log evidence---we cannot compute for
technical/computational reasons. However, the right-hand side is composed of
things we can easily evaluate. We can easily evaluate the log-likelihood $\ln
\pi(\underline{x} \mid \underline{\theta})$ and the KL divergence between our
proposed approximate distribution $q_\phi(\underline{\theta} \mid
\underline{x})$ and the prior distribution $\pi(\underline{\theta})$. Moreover,
we can compute the gradients of these functions with respect to the parameters
of our proposed distribution. This last point implies that we can change the
parameters of the proposed distribution to maximize the ELBO. And, although we
cannot compute the left-hand side of @eq-vi_elbo, we know that however large we
make the ELBO, it will always be smaller than (or equal) the log-marginal
likelihood. Therefore, the larger we can make the ELBO by modifying the
parameters $\phi$, the closer it gets to the log-marginal likelihood, and, as a
consequence, the better our proposed distribution $q_\phi(\underline{\theta}
\mid \underline{x})$ gets to the true posterior distribution
$\pi(\underline{\theta} \mid \underline{x})$.

In this sense, variational inference turns the intractable numerical integration
problem to an optimization routine, for which there are several algorithms 
available.

### ADVI algorithm

To maximize the right-hand side of @eq-vi_elbo, the Automatic Differentiation
Variational Inference (ADVI) algorithm developed in [@kucukelbir2016] takes
advantage of advances in probabilistic programming languages to generate a
robust method to perform this optimization. Without going into the details of
the algorithm implementation, for our purposes, it suffices to say that we define
our joint distribution $\pi(\underline{\theta}, \underline{x})$ as the product
defined in @eq-vi_joint_dist. ADVI then proposes an approximate variational
distribution $q_\phi$ that can either be a multivariate Normal distribution with
a diagonal covariance matrix, i.e.,
$$
\phi = (\underline{\mu}, \underline{\underline{I}}\;\underline{\sigma}^2),
$${#eq-vi_meanfield}
where $\underline{\underline{I}}$ is the identity matrix, and
$\underline{\sigma}$ is the vector of standard deviations or a full-rank 
multivariate Normal distribution
$$
\phi = (\underline{\mu}, \underline{\underline{\Sigma}}).
$${#eq-vi_full}

Then, the parameters are initialized in some value $\phi_o$. These parameters
are iteratively updated by computing the gradient of the ELBO (right-hand side
of @eq-vi_elbo), hereafter defined as $\mathcal{L}$, with respect to the
parameters, 
$$
\nabla_\phi \mathcal{L} = \nabla_{\underline{\mu}} \mathcal{L} + 
\nabla_{\underline{\sigma}}\mathcal{L},
$${#eq-vi_elbo_gradient}
and then computing
$$
\phi_{t+1} = \phi_{t} + \eta \nabla_\phi \mathcal{L},
$$
where $\eta$ defines the step size.

This short explanation behind the ADVI algorithm is intended only to gain
intuition on how the optimal variational distribution $q_\phi$ be computed.
There are many nuances in the implementation of the ADVI algorithm. We invite
the reader to look at the original reference for further details.

### 2023-09-11 doc/_hierarchical_genotypes_results.qmd
first draft of hierarchical genotypes section

@@
 
-![**Hierarchical model for multiple barcodes per
-genotype**](./figs/fig05){#fig-05}
+Hierarchical models can also capture another experimental design in which
+multiple barcodes map to the same or an equivalent genotype. As we will show,
+this many-to-one mapping can improve the inference compared to the extreme cases
+of inferring the fitness of each barcode independently or pooling the data of
+all barcodes mapping to a single genotype. As shown in @fig-05(A), a small
+modification to our model allows us to map the structure of our original model
+to that of a hierarchical model with a fitness hyperparameter vector
+$\underline{\theta}^G$, where $G$ is the number of genotypes in the datsets.
+
+@fig-05(B) shows a single experimental replicate in which 90 genotypes were
+assigned a random number of barcodes (a multinomial distribution with a mean of
+ten barcodes per genotype) for a total of 900 non-neutral barcodes. To assess
+the performance of the hierarchical model proposed in @fig-05(A), we performed
+inference using this hierarchical model, as well as the two extreme cases of
+ignoring the connection between the barcodes belonging to the same genotype
+---equivalent to performing inference using the model presented in @fig-02(A)
+over the barcodes---or pooling the data of all barcodes belonging to the same
+genotype into a single count---equivalent to performing inference using the
+model presented in @fig-02(A) over the pooled barcodes. @fig-05(C-D) shows the
+comparison between the simulation ground truth and the inferred values for these
+three cases. Not only do the hierarchical model results show higher degrees of
+correlation with the ground truth, but the error bars (representing the 64\%
+credible regions) are smaller, meaning that the uncertainty in the estimate
+decreases when using the hierarchical model. The improvement in the prediction
+can be seen in @fig-05(F) where the empirical cumulative distribution function
+of the absolute difference between the mean inferred value and the simulation
+ground truth is shown for all three inference models. The hierarchical model's
+curve ascends more rapidly, showing that, in general, the inferred values are
+closer to the ground truth. For completeness, @fig-05(G) shows some examples of
+how the hierarchical model can capture the raw log-frequency count observations.
+
+![**Hierarchical model for multiple barcodes per genotype.** (A) Bayesian
+network schematizing the statistical dependence between observations (filled
+circles) and latent variables (empty circles). The barcodes belonging to
+equivalent genotypes are connected via the hyper-parameters
+$\underline{\theta}^G$. See `point to SI` for detailed definitions of each
+variable. (B) Simulated dataset with 100 neutral lineages and 900 barcodes of
+interest distributed among 90 genotypes. (C-E) Comparison between the inferred
+and ground truth fitness values for a hierarchical model (C), a model where each
+barcode is inferred independently (D), and a model where barcodes belonging to
+the same genotype are pooled together (E). Gray error bars represent the 68\%
+credible regions. (F) Empirical cumulative distribution function (ECDF) of the
+absolute difference between the inferred mean and the ground truth fitness
+values for all three models. (G) Examples of the posterior predictive checks for
+all neutral lineages (upper left panels) and a subset of representative mutant
+lineages. Shaded regions from light to dark represent the 95\%, 68\%, and 5\%
+credible regions.](./figs/fig05){#fig-05}

### 2023-09-13 doc/_bayesian_inference.qmd
several updates, including some of the comments by Griffin

@@
 matrix $\underline{\underline{R}}$, where $B = M + N$ is the number of unique
-barcodes given by the sum of the number of unique relevant barcodes we care
+barcodes given by the sum of the number of unique, relevant barcodes we care
 about (hereafter referred to as mutant barcodes), $M$, and the number of unique
@@
 $${#eq-r-vec_split}
-i.e., the vector containing the neutral lineages barcode counts
+i.e., the vector containing the neutral lineage barcode counts
 $\underline{r}_t^{N}$ and the corresponding vector containing the mutant barcode
@@
 
-Our objective is to compute the joint probability
-distribution for all relative fitness values given our data. We can express this 
-joint posterior distribution using Bayes theorem as
+Our objective is to compute the joint probability distribution for all relative
+fitness values given our data. We can express this joint posterior distribution
+using Bayes theorem as
 $$
@@
 where hereafter $\pi(\cdot)$ defines a probability density function, unless
-otherwise stated. To develop our model, we can ignore the denominator on the
-right-hand side of @eq-bayes_obj, since it serves as a normalization constant
-that does not depend on our parameter of interest. However, computing this 
-normalization constant represents a major challenge we will have to overcome. 
-For now, we can simply write
+otherwise stated. When defining our statistical model, we need not to focus on
+the denominator on the right-hand side of @eq-bayes_obj, and write
 $$
@@
 $${#eq-bayes_obj_propto}
+However, when implementing the model computationally, the normalization constant
+on the right-hand side of @eq-bayes_obj must be computed via an integral of the
+form
+$$
+\pi(\underline{\underline{R}}) = \int d\underline{s}^M
+\pi(\underline{\underline{R}} \mid \underline{s}^M) 
+\pi(\underline{s}^M),
+$${#eq-bayes_evidence}
+also known as a marginalization integral.
 
@@
 the mutant relative fitness. Therefore, we must include these nuisance
-parameters as part of our inference problem. We can always marginalize the joint
-distribution to obtain the distribution we are interested in. To include these
+parameters as part of our inference problem. To include these nuisance
 parameters, let 
@@
 rewrite it in a more manageable form. Specifically, as depicted in @fig-01(C),
-the statistical dependencies of our observations and latent variables allows us
-to rewrite the joint probability distribution as a product of conditional
+the statistical dependencies of our observations and latent variables allow us
+to factorize the joint distribution into the product of multiple conditional
+distributions. To exemplify this, let us focus on the variable
+$\underline{\underline{F}}^M$ within the diagram in @fig-01(C). The incoming
+arrows from $\underline{\bar{s}}_T$ and $\underline{s}^M$
+
+rewrite the joint probability distribution as a product of conditional
 distributions of the form
@@
 Written in this form, @eq-split_posterior captures the three sources of 
-uncertainty listed in @sec-fitness_model in each of the terms. Starting from
+uncertainty listed in @sec-fitness_model in each term. Starting from
 right to left, the first term on the right-hand side of @eq-split_posterior
@@
 the uncertainty in the parameter we care about---the mutants' relative
-fitnesses. We invite the reader to check the supplementary material where we
-work through each of these terms to build the posterior distribution to sample
+fitnesses. We invite the reader to check the supplementary material, where we
+work through these terms to build the posterior distribution to sample
 from. We refer the reader to `[ref SI section]` for an extended description of
@@
 $\phi$ such that it minimizes the KL divergence with respect to the true 
-posterior distribution. Although to compute @eq-vi_objective we require the
+posterior distribution. Although to compute @eq-vi_objective, we require the
 posterior distribution we are trying to approximate $\pi(\theta \mid

### 2023-09-13 doc/_experiment.qmd
several updates, including some of the comments by Griffin

@@
 
-## Experimental setup
+## Experimental setup{#sec-experiment}
 
-Our inference pipeline is designed for the type of competition assays presented
-in @kinsler2020. In these experiments, an ensemble of genotypes is pooled
-together with an unlabeled reference strain that, initially, represents the vast
-majority ($\geq 90\%$) of the cells in the culture (see schematic in
-@fig-01(A)). Furthermore, a subfraction of labeled genotypes equivalent to
-the unlabeled reference strain---hereafter defined as *neutral* lineages---are
-spiked in at a relatively high abundance ($\approx 3-5\%$), the rest of the
-culture being left for the ensemble of genotypes of interest.
+The present work is designed to analyze time-series data of relative abundance
+of multiple microbial lineages uniquely identified by a DNA barcode
+[@kinsler2020; @ascensao2023]. In these competition assays, an ensemble of
+genotypes is pooled together with an unlabeled reference strain that, initially,
+represents the vast majority ($\geq 90\%$) of the cells in the culture (see
+schematic in @fig-01(A)). Furthermore, a fraction of labeled genotypes
+equivalent to the unlabeled reference strain---hereafter defined as *neutral*
+lineages---are spiked in at a relatively high abundance ($\approx 3-5\%$). The
+rest of the culture is left for the ensemble of genotypes of interest.
 
@@
 quantify the relative abundance of each of the barcodes. We point the reader to
-@kinsler2020 for specific details on these assays for *S. cerevisiae*.
-@fig-01(B) presents a typical barcode trajectory where the darker barcode
-trajectories represent the neutral lineages.
+@kinsler2020 for specific details on these assays for *S. cerevisiae* and to
+@ascensao2023 for equivalent assays for *E. coli*. @fig-01(B) presents a typical
+barcode trajectory where the darker barcode trajectories represent the so-called
+*neutral lineages* that are genetically equivalent to the untagged ancestor
+strain that initially dominates the culture. These neutral lineages simplify the
+inference problem since the fitness of all relevant barcodes is quantified with
+respect to these barcodes.
 
@@
 between observations (filled circles) and latent variables (empty circles). The
-dotted line highlights the set of variables we are interested in. See
-@sec-bayesian_inference for specific definitions of each
-variable.](./figs/fig01){#fig-01}
+dotted circle highlights the set of variables we are interested in. One-way
+arrows depict the statistical dependence between variables. For example, the
+value of $\underline{\underline{F}}^N$ only depends on the value of
+$\underline{\bar{s}}_T$, while $\underline{\underline{F}}^M$ depends on
+$\underline{\bar{s}}_T$ and $\underline{s}^M$. See @sec-bayesian_inference for
+specific definitions of each variable.](./figs/fig01){#fig-01}

### 2023-09-13 doc/_fitness_model.qmd
several updates, including some of the comments by Griffin

@@
 Empirically, the barcode relative frequency trajectories follow an exponential
-function of the form
+function of the form [@levy2015; @kinsler2020; @ascensao2023]
 $$
@@
 $${#eq-logfreq}
-@eq-logfreq separates the observables---the barcode frequencies---from the
-latent parameters we want to infer from the data---the population mean fitness
-and the barcode relative fitness. This is ultimately the functional form used in
-our inference pipeline. Therefore, the relative fitness is computed by knowing
-the log frequency ratio of each barcode throughout the growth-dilution cycles.
+@eq-logfreq separates the measurements---the barcode frequencies---from the
+unobserved---sometimes referred to as latent---parameters we want to infer from
+the data---the population mean fitness and the barcode relative fitness. This is
+ultimately the functional form used in our inference pipeline. Therefore, the
+relative fitness is computed by knowing the log frequency ratio of each barcode
+throughout the growth-dilution cycles.
 
@@
 where, importantly, the proportionality constant depends on the total number of
-reads for the time $t$ library, which might vary from library to library.
+reads for the library at time $t$, which might vary from library to library.
 Therefore, to compare the number of reads between libraries at different time

### 2023-09-13 doc/_hierarchical_replicates_results.qmd
several updates, including some of the comments by Griffin

@@
 
-## Accounting for experimental replicates via hierarchical models
+## Accounting for experimental replicates via hierarchical models{#sec-replicates}
 
@@
 two extreme cases one can use to perform the data analysis: On the one hand, we
-can treat each measurement as completely independent, losing the power to
-utilize multiple measurements when trying to learn a single parameter. This can
-hurt the inference since, in principle, the value of our parameter of interest
-should not depend on the particular experimental replicate in question. However,
-this approach does not allow us to properly "combine" the uncertainties in both
+can treat each measurement as entirely independent, losing the power to utilize
+multiple measurements when trying to learn a single parameter. This can hurt the
+inference since, in principle, the value of our parameter of interest should not
+depend on the particular experimental replicate in question. However, this
+approach does not allow us to properly "combine" the uncertainties in both
 experiments when performing the inference. On the other hand, we can pool all
 data together and treat our different experiments as a single measurement with
-higher coverage. This loses the subtle differences that come from biotic and
-abiotic batch effects, effectively halving the data that goes into our inference
-problem.
+higher coverage. This loses the subtle differences from biotic and abiotic batch
+effects, effectively halving the data that goes into our inference problem.
 
@@
 hierarchical models rely on the definition of so-called *hyper-parameters*, that
-capture the ultimate parameter we are interested in---for our case we have a
-hyper-fitness value $\theta^{(m)}$ for each mutant. Second, each experiment
-draws randomly from the distribution of this hyper-parameter, allowing for
-subtle variability between experiments to be accounted for---for our inference
-pipeline, each experimental replicate gets assigned a fitness value $s^{(m,i)}$,
-where the extra index indicates the $i$-th experimental replicate. As
-schematized by the Bayesian network on @fig-03(A) and further detailed in
-`[point to SI section]`, our statistical model can easily be extended to a
-hierarchical structure, where the fitness of each relevant barcode on each
-experimental replicate is a sample from a global hyper-fitness parameter.
+capture the parametric inference we are interested in---for this inference
+problem, we have a hyper-fitness value $\theta^{(m)}$ for each mutant. Second,
+each experiment draws randomly from the distribution of this hyper-parameter,
+allowing for subtle variability between experiments to be accounted for---in the
+present inference pipeline, each experimental replicate gets assigned a *local*
+fitness value $s^{(m)}_i$, where the extra sub-index indicates the $i$-th
+experimental replicate. For our hierarchical model, these local fitness values
+interact via the common dependency on the global hyper-fitness value. As
+schematized in @fig-03(A), this interaction runs both ways. First, the data from
+one experimental replicate ($\underline{\underline{R}}^M_k$ in the diagram)
+informs all local fitness values via the global hyper-fitness (upper panel in
+@fig-03(A)). Second, the local fitness value is informed by the data from all
+experimental replicates via the same global hyper-fitness parameter (lower panel
+in @fig-03(A)).
 
@@
 with 1000 unique barcodes (see @fig-03(B-C)) where we randomly sampled a ground
-truth hyper-fitness value $\theta^{(m)}$ for each barcode. To capture
-experimental batch effects, we sampled a variation from this hyper-fitness value
-for each experimental replicate $s^{(m, i)}$. @fig-03(D) shows the relationship
-between hyper-fitness and replicate fitness values for this simulation. As shown
-by the posterior predictive checks examples in @fig-03(E), the hierarchical
+truth hyper-fitness value $\theta^{(m)}$ for each barcode. We sampled a
+variation from this hyper-fitness value for each experimental replicate $s^{(m,
+i)}$ to capture experimental batch effects. @fig-03(D) shows the relationship
+between hyper-fitness and replicate fitness values for this simulation. The
+posterior predictive checks examples in @fig-03(E) show that the hierarchical
 model can correctly fit the data for each experimental replicate. Furthermore,
-@fig-03(F-G) show a high degree of correlation between the ground truth and the
-inferred fitness values. The ECDFs shown in @fig-03(H-I) reveal that for
-$\approx 75\%$ of the non-neutral barcodes, the ground truth hyper-fitness
-values fall within one standard deviation from the mean value in the posterior
-distributions.
+@fig-03(F-G) show a high correlation between the ground truth and the inferred
+fitness values. The ECDFs shown in @fig-03(H-I) reveal that for $\approx 75\%$
+of the non-neutral barcodes, the ground truth hyper-fitness values fall within
+one standard deviation from the mean value in the posterior distributions.
 
-![**Hierarchical model on experimental replicates**. (A) Bayesian network
-schematizing the statistical dependence between observations (filled circles)
-and latent variables (empty circles). The replicates are connected via the
-hyper-parameters $\underline{\theta}^M$. See `point to SI` for detailed
-definitions of each variable. (B-C) Simulated replicate datasets with 900
-barcodes of interest and 100 neutral lineages. (D) Comparison between the
-simulation ground truth hyper-fitness and each replicate ground truth fitness.
-The scatter between parameters captures experimental batch effects. (E) Examples
-of the posterior predictive checks for all neutral lineages (upper left panels)
-and a subset of representative mutant lineages. Shaded regions from light to
-dark represent the 95\%, 68\% and 5\% credible regions. (F-G) Comparison between
-the ground truth hyperfitness (F) and replicate fitness (G) values from the
-simulation and the inferred parameters. Gray error bars represent the 68\%
-posterior credible region. (H-I) The empirical cumulative distribution function
-(ECDF) for the absolute z-score value of the ground truth parameter value within
-the inferred hyper-fitness posterior distribution (H) and replicate fitness
-(I).](./figs/fig03){#fig-03}
+![**Hierarchical model on experimental replicates**. (A) Schematic depiction of
+the interactions between local fitness values $\underline{s}_l^M$ through the
+global hyper-fitness value $\underline{\theta}^M$ for $K$ hypothetical
+experimental replicates. The upper diagram shows how the data from replicate $k$
+informs all local fitness values via the hyper-fitness parameter. The lower
+panel shows the reverse, where all datasets inform the local fitness value.
+(B-C) Simulated replicate datasets with 900 barcodes of interest and 100 neutral
+lineages. (D) Comparison between the simulation ground truth hyper-fitness and
+each replicate ground truth fitness. The scatter between parameters captures
+experimental batch effects. (E) Examples of the posterior predictive checks for
+all neutral lineages (upper left panels) and a subset of representative mutant
+lineages. Shaded regions from light to dark represent the 95\%, 68\%, and 5\%
+credible regions. (F-G) Comparison between the simulation's ground truth
+hyper-fitness (F) and replicate fitness (G) values and the inferred parameters.
+Gray error bars represent the 68\% posterior credible region. (H-I) The
+empirical cumulative distribution function (ECDF) for the absolute z-score value
+of the ground truth parameter value within the inferred hyper-fitness posterior
+distribution (H) and replicate fitness (I).](./figs/fig03){#fig-03}
 
-As shown in @fig-04, the structure imposed by the hierarchical model, where we
-explicitly account for the connection between experimental replicates can
-improve the quality of the inference.  Inferred fitness values between
-experimental replicate exhibit a stronger degree of correlation in the
-hierarchical model (@fig-04(A)) compared to conducting inference on each
-replicate independently (@fig-04(B)). Moreover, when comparing the inferred
-hyper-fitness values---the objective parameter when performing multiple
-experimental measurements---the hierarchical model outperforms averaging the
-independent experimental replicates as shown in @fig-04(C) and (D).
+As shown in @fig-04, the structure imposed by the hierarchical model schematized
+in @fig-03(A), where we explicitly account for the connection between
+experimental replicates can improve the quality of the inference. Inferred
+fitness values between experimental replicates exhibit a stronger degree of
+correlation in the hierarchical model (@fig-04(A)) compared to conducting
+inference on each replicate independently (@fig-04(B)). Moreover, when comparing
+the inferred hyper-fitness values---the objective parameter when performing
+multiple experimental measurements---the hierarchical model outperforms
+averaging the independent experimental replicates as shown in @fig-04(C) and
+(D).
 
@@
 average of the mean inferred fitness values for the two experimental replicates.
-No error bars are shown for these as it is inappropriate to compute one with two
-data points. (D) Empirical cumulative distribution function (ECDF) of the
+No error bars are shown for these, as it is inappropriate to compute one with
+two data points. (D) Empirical cumulative distribution function (ECDF) of the
 absolute difference between the inferred mean and the ground truth

### 2023-09-13 doc/_intro.qmd
several updates, including some of the comments by Griffin

@@
 
-The advent of DNA barcoding and high-throughput sequencing has opened new venues
-to understand microbial evolutionary dynamics with an unprecedented level of
-temporal resolution [@levy2015; @nguyenba2019a; @ascensao2023]. These
-experimental efforts rely on our ability to reliably infer the relative fitness
-of an ensemble of diverse genotypes. Moreover, inferring these fitness values
-over an ensemble of environmental conditions can help us determine the
-phenotypic diversity of a rapid adaptation process [@kinsler2020].
+The advent of DNA barcoding---the ability to uniquely identify cell lineages
+with DNA sequences integrated at a specific locus---and high-throughput
+sequencing has opened new venues for understanding microbial evolutionary
+dynamics with an unprecedented level of temporal resolution [@levy2015;
+@nguyenba2019a; @ascensao2023]. These experimental efforts rely on our ability
+to reliably infer the relative fitness of an ensemble of diverse genotypes.
+Moreover, inferring these fitness values over an ensemble of environmental
+conditions can help us determine the phenotypic diversity of a rapid adaptation
+process [@kinsler2020].
 
@@
 amplification, and the sequencing process itself. Thus, accounting for the
-uncertainty when inferring the parameters we care about from the data is a key
-step to drawing reliable conclusions. Bayesian statistics presents a paradigm
-with which to account for all known sources of uncertainty in a principled way
+uncertainty when inferring the relevant parameters from the data is a key step
+to draw reliable conclusions. Bayesian statistics presents a paradigm by which
+one can account for all known sources of uncertainty in a principled way
 [@eddy2004a]. This, combined with the development of modern Markov Chain Monte
@@
 
-We present a Bayesian inference pipeline to properly quantify our uncertainty
+We present a Bayesian inference pipeline to properly quantify the uncertainty
 about the parametric information we can extract from high-throughput competitive
 fitness assays. In these assays, the fitness of an ensemble of genotypes is
-determined relative to a reference genotype [@kinsler2020; @ascensao2023] (See
-@fig-01 for a schematic depiction of the experimental setup). Small changes to
-our model allows us to account for relevant experimental variations of these
-competition assays: determination of multiple fitness values per genotype for
-growth-dilution cycles on different environments, a hierarchical model to
-account for information from independent experimental replicates, and a
-hierarchical model for a within-experiment structure where multiple independent
-barcodes map to a specific genotype. We benchmark the ability of our pipeline to
-infer fitness parameters against synthetic data with known ground truth and
-re-analyze published data from @kinsler2020 to assess the method's performance
-on real experimental measurements. The method is accompanied by a `Julia`
-package to readily implement the inference pipeline with state-of-the-art
-scientific computing software. [`here mention something about the Python wrapper
-if appropriate`].
+determined relative to a reference genotype [@kinsler2020; @ascensao2023].
+@fig-01(A) shows a schematic of the experimental procedure in which an initial
+pool of barcoded strains are mixed with a reference strain and inoculated into
+fresh media. After some time---usually, enough time for the culture to
+saturate---an aliquot is transferred to fresh media, while the remaining culture
+is used for DNA sequencing of the lineage barcodes. The time-series information
+of the relative abundance of each lineage, i.e., the barcode frequency depicted
+in @fig-01(B), is used to infer the relative fitness---the growth advantage on a
+per-cycle basis---for each lineage with respect to the reference strain. The
+proposed statistical model accounts for multiple sources of uncertainty when
+inferring the lineages' relative fitness values (see @sec-experiment for details
+on sources of uncertainty accounted by the model). Furthermore, small changes to
+the core statistical model allow us to account for relevant experimental
+variations of these competition assays. More specifically, as described in
+@sec-replicates, our statistical model can account for batch-to-batch
+differences when jointly analyzing multiple experimental replicates using a
+Bayesian hierarchical model. Moreover, a variant of these hierarchical models,
+presented in @sec-genotypes, can account for variability within multiple
+barcodes mapping to equivalent genotypes within the same experiment. Finally, in
+@sec-multienv, we present a variation of the statistical model to infer fitness
+on growth-dilution cycles in multiple environments.
+
+For all the model variations presented in this paper, we benchmark the ability
+of our pipeline to infer relative fitness parameters against synthetic data with
+known ground truth. We also re-analyze published data from @kinsler2020 to
+assess the method's performance on experimental measurements. The present method
+is accompanied by a `Julia` package to readily implement the inference pipeline
+with state-of-the-art scientific computing software. [`here mention something
+about the Python wrapper if appropriate`].

### 2023-09-13 doc/_multienv_results.qmd
several updates, including some of the comments by Griffin

---
editor:
    render-on-save: true
bibliography: references.bib
csl: ieee.csl
---

## Fitness inference on multiple environments{#sec-multienv}

![**Multi-environment fitness inference.**](./figs/fig06){#fig-06}

### 2023-09-13 doc/_variational_inference_primer.qmd
remove conditioning of data since it doesnt help

@@
 $$
-q_\phi^*(\underline{\theta} \mid \underline{x}) =
+q_\phi^*(\underline{\theta}) =
 \min_\phi D_{KL}\left(
-    q_\phi(\underline{\theta} \mid \underline{x}) \vert\vert 
+    q_\phi(\underline{\theta}) \vert\vert 
     \pi(\underline{\theta} \mid \underline{x})
@@
 $${#eq-vi_objective}
-where $D_{KL}$ is the KL divergence. We alert the reader that in the literature
-$q_\phi$ is usually  written not conditioned on the observations, i.e.,
-$q_\phi(\underline{\theta})$. But we think that writing the conditioning
-reflects that the selection of the distribution $q_\phi^*$ out of the parametric
-family depends on having the observations $\underline{x}$ in hand. Furthermore,
-we highlight that the KL divergence is a strictly positive number, i.e.,
+where $D_{KL}$ is the KL divergence. Furthermore, we highlight that the KL
+divergence is a strictly positive number, i.e.,
 $$
 D_{KL}\left(
-    q_\phi(\underline{\theta} \mid \underline{x}) \vert\vert 
+    q_\phi(\underline{\theta}) \vert\vert 
     \pi(\underline{\theta} \mid \underline{x})
@@
 D_{KL}\left(
-    q_\phi(\underline{\theta} \mid \underline{x}) \vert\vert 
+    q_\phi(\underline{\theta}) \vert\vert 
     \pi(\underline{\theta} \mid \underline{x})
@@
 \int \cdots \int d\underline{\theta}^K
-q_\phi(\underline{\theta} \mid \underline{x})
+q_\phi(\underline{\theta})
 \ln \frac{
-    q_\phi(\underline{\theta} \mid \underline{x})
+    q_\phi(\underline{\theta})
 }{
@@
 D_{KL}\left(
-    q_\phi(\underline{\theta} \mid \underline{x}) \vert\vert 
+    q_\phi(\underline{\theta}) \vert\vert 
     \pi(\underline{\theta} \mid \underline{x})
 \right) = 
-\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
-\ln q_\phi(\underline{\theta} \mid \underline{x}) -
-\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
+\int d\underline{\theta}^K q_\phi(\underline{\theta})
+\ln q_\phi(\underline{\theta}) -
+\int d\underline{\theta}^K q_\phi(\underline{\theta})
 \ln \pi(\underline{\theta} \mid \underline{x}),
@@
 D_{KL}\left(
-    q_\phi(\underline{\theta} \mid \underline{x}) \vert\vert 
+    q_\phi(\underline{\theta}) \vert\vert 
     \pi(\underline{\theta} \mid \underline{x})
 \right) &= 
-\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
-\ln q_\phi(\underline{\theta} \mid \underline{x}) \\
-&- \int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
+\int d\underline{\theta}^K q_\phi(\underline{\theta})
+\ln q_\phi(\underline{\theta}) \\
+&- \int d\underline{\theta}^K q_\phi(\underline{\theta})
 \ln \left( 
@@
 D_{KL}\left(
-    q_\phi(\underline{\theta} \mid \underline{x}) \vert\vert 
+    q_\phi(\underline{\theta}) \vert\vert 
     \pi(\underline{\theta} \mid \underline{x})
 \right) &= 
-\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
-\ln q_\phi(\underline{\theta} \mid \underline{x}) \\
-&-\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
+\int d\underline{\theta}^K q_\phi(\underline{\theta})
+\ln q_\phi(\underline{\theta}) \\
+&-\int d\underline{\theta}^K q_\phi(\underline{\theta})
 \ln \pi(\underline{x} \mid \underline{\theta}) \\
-&-\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
+&-\int d\underline{\theta}^K q_\phi(\underline{\theta})
 \ln \pi(\underline{\theta}) \\
-&+\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
+&+\int d\underline{\theta}^K q_\phi(\underline{\theta})
 \ln \pi(\underline{x}).
@@
 D_{KL}\left(
-    q_\phi(\underline{\theta} \mid \underline{x}) \vert\vert 
+    q_\phi(\underline{\theta}) \vert\vert 
     \pi(\underline{\theta} \mid \underline{x})
 \right) &= 
-\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
+\int d\underline{\theta}^K q_\phi(\underline{\theta})
 \ln \frac{
-    q_\phi(\underline{\theta} \mid \underline{x})
+    q_\phi(\underline{\theta})
     }{
@@
     } \\
-&-\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
+&-\int d\underline{\theta}^K q_\phi(\underline{\theta})
 \ln \pi(\underline{x} \mid \underline{\theta}) \\
 &+ \ln \pi(\underline{x}) 
-\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x}),
+\int d\underline{\theta}^K q_\phi(\underline{\theta}),
 \end{aligned}
@@
 $$
-\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x}) = 1.
+\int d\underline{\theta}^K q_\phi(\underline{\theta}) = 1.
 $${#eq-vi_q_norm}
@@
 $$
-\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
+\int d\underline{\theta}^K q_\phi(\underline{\theta})
 f(\underline{\theta}) = \left\langle 
@@
 D_{KL}\left( 
-    q_\phi(\underline{\theta} \mid \underline{x}) \vert \vert
+    q_\phi(\underline{\theta}) \vert \vert
     \pi(\underline{\theta}) 
@@
     D_{KL}\left( 
-        q_\phi(\underline{\theta} \mid \underline{x}) \vert \vert
+        q_\phi(\underline{\theta}) \vert \vert
         \pi(\underline{\theta}) 
@@
 approximate the posterior distribution $\pi(\underline{\theta} \mid
-\underline{x})$ with a parametric distribution $q_\phi(\underline{\theta} \mid
-\underline{x})$. By minimizing the KL divergence between these distributions, we
-arrive at the result in @eq-vi_elbo, where the left-hand side---the log
-marginalized likelihood or log evidence---we cannot compute for
-technical/computational reasons. However, the right-hand side is composed of
-things we can easily evaluate. We can easily evaluate the log-likelihood $\ln
-\pi(\underline{x} \mid \underline{\theta})$ and the KL divergence between our
-proposed approximate distribution $q_\phi(\underline{\theta} \mid
-\underline{x})$ and the prior distribution $\pi(\underline{\theta})$. Moreover,
-we can compute the gradients of these functions with respect to the parameters
-of our proposed distribution. This last point implies that we can change the
-parameters of the proposed distribution to maximize the ELBO. And, although we
-cannot compute the left-hand side of @eq-vi_elbo, we know that however large we
-make the ELBO, it will always be smaller than (or equal) the log-marginal
-likelihood. Therefore, the larger we can make the ELBO by modifying the
-parameters $\phi$, the closer it gets to the log-marginal likelihood, and, as a
-consequence, the better our proposed distribution $q_\phi(\underline{\theta}
-\mid \underline{x})$ gets to the true posterior distribution
-$\pi(\underline{\theta} \mid \underline{x})$.
+\underline{x})$ with a parametric distribution $q_\phi(\underline{\theta})$. By
+minimizing the KL divergence between these distributions, we arrive at the
+result in @eq-vi_elbo, where the left-hand side---the log marginalized
+likelihood or log evidence---we cannot compute for technical/computational
+reasons. However, the right-hand side is composed of things we can easily
+evaluate. We can easily evaluate the log-likelihood $\ln \pi(\underline{x} \mid
+\underline{\theta})$ and the KL divergence between our proposed approximate
+distribution $q_\phi(\underline{\theta})$ and the prior distribution
+$\pi(\underline{\theta})$. Moreover, we can compute the gradients of these
+functions with respect to the parameters of our proposed distribution. This last
+point implies that we can change the parameters of the proposed distribution to
+maximize the ELBO. And, although we cannot compute the left-hand side of
+@eq-vi_elbo, we know that however large we make the ELBO, it will always be
+smaller than (or equal) the log-marginal likelihood. Therefore, the larger we
+can make the ELBO by modifying the parameters $\phi$, the closer it gets to the
+log-marginal likelihood, and, as a consequence, the better our proposed
+distribution $q_\phi(\underline{\theta})$ gets to the true posterior
+distribution $\pi(\underline{\theta} \mid \underline{x})$.
 

### 2023-09-13 doc/_bayesian_inference.qmd
finished adding Griffin's comments

@@
 However, we can take advantage of the structure of our inference problem to
-rewrite it in a more manageable form. Specifically, as depicted in @fig-01(C),
-the statistical dependencies of our observations and latent variables allow us
-to factorize the joint distribution into the product of multiple conditional
-distributions. To exemplify this, let us focus on the variable
-$\underline{\underline{F}}^M$ within the diagram in @fig-01(C). The incoming
-arrows from $\underline{\bar{s}}_T$ and $\underline{s}^M$
-
-rewrite the joint probability distribution as a product of conditional
-distributions of the form
+rewrite it in a more manageable form. Specifically, the statistical dependencies
+of our observations and latent variables allow us to factorize the joint
+distribution into the product of multiple conditional distributions. To gain
+some intuition about this factorization, let us focus on the inference of the
+population mean fitness values $\underline{\bar{s}}_T$. @eq-logfreq_neutral
+relates the value of the population mean fitness to the neutral lineage
+frequencies and nothing else. This suggests that when writing the posterior for
+these population mean fitness parameters, we should be able to condition it only
+on the neutral lineage frequency values, i.e., $\pi(\underline{\bar{s}}_T \mid
+\underline{\underline{F}}^N)$. We point the reader to `[SI section]` for the
+full mathematical details on this factorization. For our purpose here, it
+suffices to say we can rewrite the joint probability distribution as a product
+of conditional distributions of the form
 $$
@@
 \pi(
-    \underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}}^M,
-    \underline{\underline{R}}^M
+    \underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}}^M
 )
 \pi(
-    \underline{\bar{s}}_T \mid \underline{\underline{F}}^N,
-    \underline{\underline{R}}^N
+    \underline{\bar{s}}_T \mid \underline{\underline{F}}^N
 )
@@
 (\underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}})$, is
-replaced by an approximate posterior distribution $q_\phi(\theta \mid
-\underline{\underline{R}})$, where $\phi$ fully parametrizes the approximate
-distribution. As further explained in `[point to SI section]`, the numerical
-integration problem is replaced by an optimization problem of the form
+replaced by an approximate posterior distribution $q_\phi(\theta)$, where $\phi$
+fully parametrizes the approximate distribution. As further explained in
+@sec-vi_primer, the numerical integration problem is replaced by an optimization
+problem of the form
 $$
-q^*_\phi(\theta \mid \underline{\underline{R}}) = \min _\phi
+q^*_\phi(\theta) = \min _\phi
 D_{KL}(
-    q_\phi(\theta \mid \underline{\underline{R}}) \lvert \lvert
+    q_\phi(\theta) \lvert \lvert
     \pi(\theta \mid \underline{\underline{R}})
@@
 where $D_{KL}$ is the Kulback-Leibler divergence. In other words, the
-complicated high-dimensional numerical integration problem is transformed into
-a gradient descent over the parameters of the approximate posterior distribution
-$\phi$ such that it minimizes the KL divergence with respect to the true 
-posterior distribution. Although to compute @eq-vi_objective, we require the
-posterior distribution we are trying to approximate $\pi(\theta \mid
+complicated high-dimensional numerical integration problem is transformed into a
+much simpler problem of finding the value of the parameters $\phi$ such that
+@eq-vi_objective is satisfied as best as possible within some finite computation
+time. Although to compute @eq-vi_objective, we require the posterior
+distribution we are trying to approximate $\pi(\theta \mid
 \underline{\underline{R}})$, it can be shown that maximizing the so-called
@@
 performing the optimization prescribed by @eq-vi_objective. We direct the reader
-to `[point to SI section]` for a short primer on variational inference.
+to @sec-vi_primer for a short primer on variational inference.
 

### 2023-09-13 doc/_experiment.qmd
finished adding Griffin's comments

@@
 growth-dilution cycles are performed on either a single or multiple
-environments. In other words, the cultures are grown for some time to then
-dilute the culture back into fresh media and repeat for roughly 4-7 cycles,
-depending on the initial abundances of the mutants, and their relative growth
-rates. The DNA barcodes are sequenced at the end of each growth cycle to
-quantify the relative abundance of each of the barcodes. We point the reader to
-@kinsler2020 for specific details on these assays for *S. cerevisiae* and to
-@ascensao2023 for equivalent assays for *E. coli*. @fig-01(B) presents a typical
-barcode trajectory where the darker barcode trajectories represent the so-called
-*neutral lineages* that are genetically equivalent to the untagged ancestor
-strain that initially dominates the culture. These neutral lineages simplify the
-inference problem since the fitness of all relevant barcodes is quantified with
-respect to these barcodes.
+environments. In other words, the cultures are grown for some time; then, an
+aliquot is inoculated into fresh media for the next growth cycle. This process
+is repeated for roughly 4-7 cycles, depending on the initial abundances of the
+mutants and their relative growth rates. The DNA barcodes are sequenced at the
+end of each growth cycle to quantify the relative abundance of each of the
+barcodes. We point the reader to @kinsler2020 for specific details on these
+assays for *S. cerevisiae* and to @ascensao2023 for equivalent assays for *E.
+coli*. @fig-01(B) presents a typical barcode trajectory where the darker barcode
+trajectories represent the so-called *neutral lineages* that are genetically
+equivalent to the untagged ancestor strain that initially dominates the culture.
+These neutral lineages simplify the inference problem since the fitness of all
+relevant barcodes is quantified with respect to these barcodes.
 
@@
 different relative fitness. Darker gray lines define the typical trajectory of
-neutral lineages. (C) Bayesian network defining the statistical dependence
-between observations (filled circles) and latent variables (empty circles). The
-dotted circle highlights the set of variables we are interested in. One-way
-arrows depict the statistical dependence between variables. For example, the
-value of $\underline{\underline{F}}^N$ only depends on the value of
-$\underline{\bar{s}}_T$, while $\underline{\underline{F}}^M$ depends on
-$\underline{\bar{s}}_T$ and $\underline{s}^M$. See @sec-bayesian_inference for
-specific definitions of each variable.](./figs/fig01){#fig-01}
+neutral lineages.](./figs/fig01){#fig-01}

### 2023-09-13 doc/_fitness_model.qmd
finished adding Griffin's comments

@@
 
+The presence of the neutral lineages facilitates the determination of the 
+population mean fitness value $\bar{s}_t$. Since every relative fitness is 
+determined relative to the neutral lineage that dominates the culture, we define
+their fitness to be $s^{(n)} = 0$, where the subscript $(n)$ specifies their
+neutrality. This means that @eq-logfreq for a neutral lineage takes the simpler
+form
+$$
+\frac{1}{\tau}\ln \frac{f_{t+1}^{(n)}}{f_{t}^{(n)}} = - \bar{s}_t.
+$${#eq-logfreq_neutral}
+Therefore, we can use the data from these reference barcodes to directly infer
+the value of the population mean fitness.
+
 It is important to notice that the frequencies $f_{t}^{(b)}$ are not the allele

### 2023-09-13 doc/_single_results.qmd
finished adding Griffin's comments

@@
 
-To assess the inference pipeline performance we applied it to a simulated
+To assess the inference pipeline performance, we applied it to a simulated
 dataset with known ground truth relative fitness values (See `[ref to SI]` for
 details on simulation). @fig-02(A) shows the structure of the synthetic dataset.
-The majority of barcodes of interest (dim color lines) are adaptive compared to
-the neutral barcodes ($s^{(m)} > 0$). This experimental condition is required to
-reliably measure the relative fitness since all barcodes of interest occupy a
+The majority of barcodes of interest (faint color lines) are adaptive compared
+to the neutral barcodes ($s^{(m)} > 0$). This experimental condition is required
+to reliably measure the relative fitness since all barcodes of interest occupy a
 small fraction of the population at the beginning of the experiment `[ref SI
@@
 @fig-02(B) shows such data transformation where we can better appreciate the
-observational noise input into our statistical model. This noise is
-evident for the darker lines representing the neutral barcodes since
-all of these lineages are assumed to be identically distributed under this log
-frequency ratio transformation.
+observational noise input into our statistical model. This noise is evident for
+the darker lines representing the neutral barcodes since all of these lineages
+are assumed to be identically distributed under this log frequency ratio
+transformation.
 
-A way to visualize the performance of our inference pipeline in fitting our
-fitness model to the observed data is to compute the so-called posterior
-predictive checks (PPC). In short, the PPC consists of bootstrapping the
-generation of synthetic datasets in agreement with the results from the
-inference results. In other words, we use the resulting parameter values from
-the ADVI inference to generate possible datasets in agreement with the inferred
-values. @fig-02(C) shows these results for all neutral lineages (upper left
-corner plot) and a few representative non-neutral barcodes. The dark color line
-shows the median value we would expect the data to be found, while the shaded
-regions show the 68\% and 95\% credible regions, i.e., the regions where we
-expect to find the data with the corresponding probability---or in terms of our
-parameter, the $X\%$ credible region is the interval where we expect the true
-parameter to lie with $X\%$ probability. The main advantage of our method is
-this natural interpretability of these credible regions since a common mistake
-in the literature is to interpret frequentist confidence intervals as Bayesian
-credible regions when they are not equivalent [@morey2016].
+To visualize the performance of our inference pipeline in fitting our fitness
+model to the observed data, we compute the so-called posterior predictive checks
+(PPC). In short, the PPC consists of repeatedly generating synthetic datasets in
+agreement with the results from the inference results. In other words, we use
+the resulting parameter values from the ADVI inference to generate possible
+datasets in agreement with the inferred values. @fig-02(C) shows these results
+for all neutral lineages (upper left corner plot) and a few representative
+non-neutral barcodes. The different color shades represent the 95\%, 68\%, and
+5\% credible regions, i.e., the regions where we expect to find the data with
+the corresponding probability---or in terms of our parameter, the $X\%$ credible
+region is the interval where we expect the true parameter to lie with $X\%$
+probability.
+
+The main advantage of our method is this natural interpretability of these
+credible regions where an $X\%$ credible region indeed captures the region of
+parameter space where we expect with $X\%$ probability the actual value of the
+parameter lies given our statistical model, our prior information, and the
+observed experimental data. A common mistake in the literature is interpreting
+frequentist confidence intervals as Bayesian credible regions when they are not
+equivalent [@morey2016]. Frequentist confidence intervals and Bayesian credible
+regions are based on fundamentally different philosophical approaches to
+statistics. Frequentist confidence intervals represent the range of values that
+would contain the true population parameter with a certain probability if the
+experiment was repeated many times. The confidence level does not represent the
+probability that the interval contains the true value. According to a specific
+model and prior information, Bayesian credible regions represent the range of
+values that contain the parameter with a certain posterior probability. The
+credible region directly represents the probability that the region contains the
+true value. So, frequentist confidence intervals cannot be interpreted as
+Bayesian credible regions because they have fundamentally different meanings.
+Treating an $X\%$ confidence interval like an $X\%$  credible region is
+fallacious since confidence levels do not represent probabilistic coverage of
+the true value like credible regions. The intervals are generated through
+entirely different procedures.
 
@@
 interest. There is an excellent degree of correspondence between these values,
-with the error bars, representing the 68\% credible region for the parameter
+with the error bars representing the 68\% credible region for the parameter
 value crossing the identity line for most barcodes. This latter point is made
@@
 time points used by the inference pipeline. Darker lines represent the neutral
-barcodes. Notice that these transformed data are much more noisy than the
+barcodes. These transformed data are much more noisy than the
 seemingly smooth frequency trajectories. (B) Examples of the posterior

### 2023-09-13 doc/_variational_inference_primer.qmd
finished adding Griffin's comments

@@
 
-## Primer on Variational Inference
+## Primer on Variational Inference{#sec-vi_primer}
 
@@
 $$
-\phi = (\underline{\mu}, \underline{\underline{I}}\;\underline{\sigma}^2),
+\phi = (\underline{\mu}, \underline{\underline{D}}),
 $${#eq-vi_meanfield}
-where $\underline{\underline{I}}$ is the identity matrix, and
-$\underline{\sigma}$ is the vector of standard deviations or a full-rank 
-multivariate Normal distribution
+where $\underline{\underline{D}}$ is the identity matrix, with the diagonal
+elements given by the vector of variances $\underline{\sigma}^2$ for each
+variable or a full-rank multivariate Normal distribution
 $$

### 2023-09-13 doc/_multienv_results.qmd
first draft of multi-env section

@@
 
-![**Multi-environment fitness inference.**](./figs/fig06){#fig-06}
+One of the benefits of Bayesian methods is the intrinsic error propagation
+embedded in the mathematical framework. For example, the uncertainty on the
+value of the population mean fitness values $\underline{\bar{s}}_T$ is
+propagated to the relative fitness of a particular strain $s^{(b)}$ since we
+defined a joint posterior distribution over all parameters as fully expressed in
+@eq-bayes_full. This natural error propagation can help us with the experimental
+design schematized in @fig-06(A). Here, rather than performing growth-dilution
+cycles in the same environment, the cells are diluted into a different 
+environment. Thus, the uncertainty on the fitness estimate for the previous
+environment must be propagated to that of the next one.
+
+To validate the extension of our statistical model to this scenario, @fig-06(B)
+shows the trajectory of the log frequency ratios between adjacent time points.
+The different colored regions correspond to the different environments. For this
+simulation, the growth rate of Environment 2 was set to be, on average, half of
+the average growth rate in Environment 1. Equivalently, the growth rate in
+Environment 3 was set to be, on average, twice the average growth rate in
+Environment 1. @fig-06(C-E) show the correspondence between the simulation
+ground truth and the inferred fitness values, where the error bars represent the
+68\% credible region. @fig-06(F) summarizes the performance of our inference
+pipeline by showing the empirical cumulative distribution functions for the
+absolute value of the ground truth fitness value z-score within the posterior
+distribution. This plot shows that, overall, $\approx 75\%$ of inferred mean
+values fall within one standard deviation of the ground truth. For completeness,
+@fig-06(G) shows the posterior predictive checks for a few example barcodes.
+
+![**Multi-environment fitness inference.** (A) Schematic of the simulated
+experimental design where growth-dilution cycles were performed in different
+environments in a cycle. (B) log frequency rations between adjacent time points.
+Darker lines represent the neutral barcodes. The colors in the background demark
+the corresponding environment, matching colors in (A). (C-E) Comparison between
+the ground truth fitness value from the logistic-growth simulation and the
+inferred fitness value for each environment. Gray error bars represent the 68\%
+posterior credible region. (F) The empirical cumulative distribution function
+(ECDF) for the absolute z-score value of the ground truth parameter value within
+the inferred fitness posterior distribution for all fitness values (black line)
+and each environment individually (color lines). (G) Examples of the posterior
+predictive checks for all neutral lineages (upper left panel) and a subset of
+representative mutant lineages. Shaded regions represent the 95\%, 68\%, and 5\%
+credible regions for the data. The reported errors above the plot represent the
+68\% credible region on the mutant relative fitness marginal distribution.
+Background colors match those of (A).](./figs/fig06){#fig-06}

### 2023-09-13 doc/_single_results.qmd
fixing fig legend

@@
 time points used by the inference pipeline. Darker lines represent the neutral
-barcodes. These transformed data are much more noisy than the
-seemingly smooth frequency trajectories. (B) Examples of the posterior
-predictive checks for all neutral lineages (upper left panel) and a subset of
-representative mutant lineages. Error bars represent the 68\% credible region on
-the mutant relative fitness marginal distribution. (C) Comparison between the
+barcodes. These transformed data are much more noisy than the seemingly smooth
+frequency trajectories. (B) Examples of the posterior predictive checks for all
+neutral lineages (upper left panel) and a subset of representative mutant
+lineages. Shaded regions represent the 95\%, 68\%, and 5\% credible regions for
+the data. The reported errors above the plot represent the 68\% credible region
+on the mutant relative fitness marginal distribution. (C) Comparison between the
 ground truth fitness value from the logistic-growth simulation and the inferred
@@
 The empirical cumulative distribution function (ECDF) for the absolute z-score
-value of the ground truth parameter value within the inferred fitness
-posterior distribution.](./figs/fig02){#fig-02}
+value of the ground truth parameter value within the inferred fitness posterior
+distribution.](./figs/fig02){#fig-02}

### 2023-09-15 doc/_discussion.qmd
continuing to make progress on paper

---
editor:
    render-on-save: true
bibliography: references.bib
csl: ieee.csl
---

Experimental evolution of microbial systems has dramatically advanced our
understanding of the basic principles of biological evolution [@kussell2013].
From questions related to the optimal fine-tuning of gene expression programs
[@Dekel2005], to the dimensionality, geometry, and accessibility of the adaptive
fitness landscape explored by these rapidly adapting populations
[@kinsler2020;@maeda2020], to the emergence of eco-evolutionary dynamics in a
long-term evolution experiment [@good2017]; for all of these and other cases,
the microbial experimental platform combined with high-throughput sequencing has
been essential to tackling these questions with empirical data. This exciting
research area promises to improve as new culturing technologies [@jagdish2022]
as well as more complex lineage barcoding schemes [@nguyenba2019a;@yang2022],
are adopted.

For this data-heavy field, properly accounting for the uncertainty in quantities
inferred from experiments is vital to ensure the conclusions drawn are reliable.
Bayesian statistics presents a principled way to quantify this uncertainty
systematically [@gelman2010]. Moreover, Bayesian analysis offers a more
natural way to interpret the role that probability theory plays when performing
data analysis compared to the often-misinterpreted frequentist methods
[@vanderplas2014]. Nevertheless, the technical challenges associated with
Bayesian analysis has limited its application. This is set to change as
recognition of the misuse of frequentist concepts such as the p-value is
receiving more attention [@nuzzo2014]. Moreover, advances in numerical methods
such as Hamiltonian Monte Carlo [@betancourt2017] and variational inference
[@kucukelbir2016] allows for complex Bayesian models to be fit to empirical
data.

In this paper, we present a computational pipeline to analyze lineage-tracking
time-series data for massive-parallel competition assays. More specifically, we
fit a Bayesian model to infer the fitness of multiple genotypes relative to a
reference [@kinsler2020;@ascensao2023]. The proposed model accounts for multiple
sources of uncertainty with proper error propagation intrinsic to Bayesian
methods. To scale the inference pipeline to large datasets with $> 10,000$
barcodes, we use the ADVI algorithm [@kucukelbir2016] to fit a variational
posterior distribution. The main difference between our method and previous
inference pipelines, such as @li2023, is that the present analysis provides
interpretable errors on the inferred fitness values. The reported uncertainty
intervals---known as credible regions---can be adequately interpreted as
capturing the corresponding probability mass of finding the true value of the
parameter given the model, the prior information, and the data. Furthermore,
minor modifications to the structure of the statistical model allow for the
analysis of different experimental designs, such as growth-dilution cycles in
different environments, joint analysis of multiple experimental replicates of
the same experiment via hierarchical models, and a hierarchical model for
multiple barcodes mapping to equivalent genotypes.

We validate our analysis pipeline on simulated datasets with known ground truth,
showing that the model fits the data adequately, capturing the ground truth
parameters within the posterior distribution. Furthermore, `[talk about the
Kinsler results.]`

### 2023-09-15 doc/_hierarchical_genotypes_results.qmd
several updates, including some of the comments by Griffin
continuing to make progress on paper

@@
 
-## Accounting for multiple barcodes per genotype via hierarchical models
+## Accounting for multiple barcodes per genotype via hierarchical models{#sec-genotypes}
 
@@
 of inferring the fitness of each barcode independently or pooling the data of
-all barcodes mapping to a single genotype. As shown in @fig-05(A), a small
-modification to our model allows us to map the structure of our original model
-to that of a hierarchical model with a fitness hyperparameter vector
-$\underline{\theta}^G$, where $G$ is the number of genotypes in the datsets.
+all barcodes mapping to a single genotype. As schematized in @fig-05(A), a small
+modification allows us to map the structure of our original model to that of a
+hierarchical model with a fitness hyperparameter vector $\underline{\theta}^G$,
+where $G$ is the number of genotypes in the dataset.
 
@@
 
-![**Hierarchical model for multiple barcodes per genotype.** (A) Bayesian
-network schematizing the statistical dependence between observations (filled
-circles) and latent variables (empty circles). The barcodes belonging to
-equivalent genotypes are connected via the hyper-parameters
-$\underline{\theta}^G$. See `point to SI` for detailed definitions of each
-variable. (B) Simulated dataset with 100 neutral lineages and 900 barcodes of
-interest distributed among 90 genotypes. (C-E) Comparison between the inferred
-and ground truth fitness values for a hierarchical model (C), a model where each
-barcode is inferred independently (D), and a model where barcodes belonging to
-the same genotype are pooled together (E). Gray error bars represent the 68\%
-credible regions. (F) Empirical cumulative distribution function (ECDF) of the
-absolute difference between the inferred mean and the ground truth fitness
-values for all three models. (G) Examples of the posterior predictive checks for
-all neutral lineages (upper left panels) and a subset of representative mutant
-lineages. Shaded regions from light to dark represent the 95\%, 68\%, and 5\%
-credible regions.](./figs/fig05){#fig-05}
+![**Hierarchical model for multiple barcodes per genotype.** (A) Schematic
+depiction of the hierarchical structure for multiple barcodes mapping to a
+single genotype. A set of barcodes mapping to an equivalent genotype map to
+"local" fitness values $s^{(b)}$ that are connected via a hyper-fitness
+parameter for the genotype $\theta^{(g)}$. (B) Simulated dataset with 100
+neutral lineages and 900 barcodes of interest distributed among 90 genotypes.
+(C-E) Comparison between the inferred and ground truth fitness values for a
+hierarchical model (C), a model where each barcode is inferred independently
+(D), and a model where barcodes mapping to the same genotype are pooled
+together (E). Gray error bars represent the 68\% credible regions. (F) Empirical
+cumulative distribution function (ECDF) of the absolute difference between the
+inferred mean and the ground truth fitness values for all three models. (G)
+Examples of the posterior predictive checks for all neutral lineages (upper left
+panels) and a subset of representative mutant lineages. Shaded regions from
+light to dark represent the 95\%, 68\%, and 5\% credible
+regions.](./figs/fig05){#fig-05}

### 2023-09-18 doc/_discussion.qmd
finish first draft of discussion

@@
 interpretable errors on the inferred fitness values. The reported uncertainty
-intervals---known as credible regions---can be adequately interpreted as
-capturing the corresponding probability mass of finding the true value of the
-parameter given the model, the prior information, and the data. Furthermore,
-minor modifications to the structure of the statistical model allow for the
-analysis of different experimental designs, such as growth-dilution cycles in
-different environments, joint analysis of multiple experimental replicates of
-the same experiment via hierarchical models, and a hierarchical model for
-multiple barcodes mapping to equivalent genotypes.
+intervals---known as credible regions---can be formally interpreted as capturing
+the corresponding probability mass of finding the true value of the parameter
+given the model, the prior information, and the data. Furthermore, minor
+modifications to the structure of the statistical model presented in this work
+allow for the analysis of different experimental designs, such as
+growth-dilution cycles in different environments, joint analysis of multiple
+experimental replicates of the same experiment via hierarchical models, and a
+hierarchical model for multiple barcodes mapping to equivalent genotypes.
 
@@
 Kinsler results.]`
+
+It is important to highlight some of the consequences of the general
+experimental design and the implicit assumptions within the proposed statistical
+model to analyze the resulting data. First, the composition of the population is
+such that the initial fraction of the population occupied by the barcoded
+genotypes is small---usually >90\% of the initial population is the non-labeled
+reference strain. This constraint is important as the fitness model used to fit
+the time series data assumes that the tracked frequencies are $\ll 1$. Second,
+when computing log frequency ratios, we can run into the issue of dividing by
+zero. This is a common problem when dealing with molecular count data
+[@lovell2020]. Our model gets around this issue by assuming that the frequency
+of any barcode cannot be, but still can get arbitrarily close to, zero.
+Therefore, we implicitly assume that no lineage goes extinct during the
+experiment. Moreover, the statistical model directly accounts for the
+uncertainty associated with having zero barcode counts, increasing the
+corresponding uncertainty. Third, the models presented in this paper require the
+existence of a labeled sub-population of barcoded reference strains. These
+barcodes help determine the fitness baseline, as every fitness is quantified
+with respect to this reference genotype. This experimental design constraint
+facilitates the inference of the population mean fitness since most of the
+culture---the unlabeled reference genotype---is not tracked. Finally, the
+presented statistical model assumes that relative fitness is solely a constant
+of the environment and the genotype. Therefore, the model is not designed to
+analyze data with time-varying or frequency-dependent fitness values.

### 2023-09-18 code/processing/kinsler_2020/viz_trajectories.jl

@@ -28,7 +28,7 @@ import PDFmerger
 CairoMakie.activate!()
 
 # Set PBoC Plotting style
-BayesFitUtils.viz.pboc_makie!()
+BayesFitUtils.viz.theme_makie!()
 
 import Random
 

### 2023-09-19 doc/_bayes_freq.qmd
update SI to poisson-multinomial model

--- 2023-03-25 doc/_bayes_freq.qmd
@@
 We begin with the probability of the frequency values given the raw barcode
-reads. The first assumption we make is that the inference of the frequency
+reads. The first assumption is that the inference of the frequency
 values for time $t$ is independent of any other time. Therefore, we can write
@@
 \pi(\underline{\underline{F}} \mid \underline{\underline{R}}) =
-\prod_{t=1}^T \pi(\underline{f}_t \mid \underline{r}_t).
+\prod_{t=1}^T \pi(\underline{f}_t \mid \underline{r}_t),
 $${#eq-freq_indep}
+where $\underline{f}_t$ and $\underline{r}_t$ are the $t$-th row of the matrix
+containing all of the measurements for time $t$. We imagine that when the 
+barcode reads are obtained via sequencing, the quantified number of reads is a
+Poisson sample from the "true" underlying number of barcodes within the pool. 
+This translates to assuming that the number of reads for each barcode at any
+time point $r^{(b)}_t$ is an independent Poisson random variable, i.e.,
+$$
+r^{(b)}_t \sim \text{Poiss}(\lambda^{(b)}_t),
+$${#eq-freq_poisson_reads}
+where the symbol "$\sim$" is read "distributed as." Furthermore, for a Poisson
+distribution, we have that
+$$
+\lambda^{(b)}_t = \left\langle r^{(b)}_t \right\rangle = 
+\left\langle 
+    \left( r^{(b)}_t - \left\langle r^{(b)}_t \right\rangle \right)^2
+\right\rangle,
+$${#eq-freq_poisson_lambda}
+where $\left\langle \cdot \right\rangle$ is the expected value. In other words
+the Poisson parameter is equal to the mean and variance of the distribution. The
+Poisson distribution has the convenient property that for two Poisson 
+distributed random variables $X \sim \text{Poiss}(\lambda_x)$ and 
+$Y \sim \text{Poiss}(\lambda_y)$, we have that
+$$
+Z \equiv X + Y \sim \text{Poiss}(\lambda_x + \lambda_y).
+$${#eq-freq_additivity}
+This additivity allows us to write the total number of reads at time $t$ $n_t$
+also as a Poisson-distributed random variable of the form
+$$
+n_t \sim \text{Poiss}\left( \sum_{b=1}^B \lambda^{(b)}_t \right),
+$${#eq-freq_poisson_total}
+where the sum is taken over all $B$ barcodes.
 
-Writing Bayes' theorem for this series of independent terms results in
+If the total number of reads is given by @eq-freq_poisson_total, the array with
+the number of reads for each barcode at time $t$, $\underline{r}_t$ is then
+distributed as
 $$
-\pi(\underline{f}_t \mid \underline{r}_t) \propto
-\pi(\underline{r}_t \mid \underline{f}_t)
-\pi(\underline{f}_t).
-$${#eq-bayes_freq}
-To model the prior uncertainty in the value of the frequencies, we assume a very
-simple model. Notice that the frequencies at any time point must satisfy the
-normalization constraint
+\underline{r}_t \sim \text{Multinomial}(n_t, \underline{f}_t),
+$${#eq-freq_multinomial}
+where each of the $B$ entries of the frequency vector $\underline{f}_t$ is a
+function of the $\underline{\lambda}_t$ vector, given by
 $$
-\sum_{b = 1}^B f_t^{(b)} = 1, \;\; \forall \; f_t^{(b)} \in [0, 1].
-$${#eq-freq_norm}
-This means that the probability distribution of the $B$ frequency values lies on
-the $B$-simplex, i.e., the $B$ dimensional space constrained such that the sum
-of all the dimensions must add up to one. A natural selection for a probability
-that satisfies the constraint in @eq-freq_norm is the so-called Dirichlet 
-distribution. The Dirichlet distribution is parametrized by a vector
+f_t^{(b)} \equiv f_t^{(b)}(\underline{\lambda}_t) = 
+\frac{\lambda_t^{(b)}}{\sum_{b'=1}^B \lambda_t^{(b')}}.
+$${#eq-freq_freq_lambda}
+In other words, we can think of the $B$ barcode counts as independent Poisson
+samples or as a single multinomial draw with a random number of total draws,
+$n_t$, and the frequency vector $\underline{f}_t$ we are interested in. Notice
+that @eq-freq_freq_lambda is a deterministic function that connects the Poisson
+parameters to the frequencies. Therefore, we have the equivalence that
 $$
-\underline{\alpha}_t = (\alpha^{(1)}_t, \cdots, \alpha^{(B)}_t)^\dagger, 
-\;\; \alpha_t^{(k)} \geq 0 \;\; \forall \; k.
-$${#eq-alpha_dir}
-Given these parameters, we choose as a prior distribution for the barcode
-frequency values $\pi(\underline{f}_t)$ a Dirichlet distribution of the form
+\pi(\underline{f}_t \mid \underline{r}_t) = 
+\pi(\underline{\lambda}_t \mid \underline{r}_t),
+$${#eq-freq_f_lambda_equiv}
+meaning that the uncertainty comes from the $\underline{\lambda}_t$ vector. By
+Bayes theorem, we therefore write
 $$
-\pi(\underline{f}_t) = \frac{
-    \Gamma\left(\sum_{b=1}^B\alpha_t^{(b)}\right)
-}{
-    \prod_{b=1}^B\Gamma(\alpha_t^{(b)})
-} 
-\prod_{b=1}^B (f_t^{(b)})^{\alpha_t^{(b)}-1}.
-$${#eq-dirichlet}
-@eq-dirichlet can be much more compactly written as
+\pi(\underline{\lambda}_t \mid n_t, \underline{r}_t) \propto
+\pi(n_t, \underline{r}_t \mid \underline{\lambda}_t) \pi(\underline{\lambda}_t),
+$${#eq-freq_lambda_bayes}
+where we explicitly include the dependence on $n_t$. This does not affect the
+distribution or brings more uncertainty because $\underline{r}_t$ already 
+contains all the information to compute $n_t$ since
 $$
-\underline{f}_t \sim \operatorname{Dirichlet}(\underline{\alpha}_t),
-$${#eq-freq_prior}
-where the symbol "$\sim$" is read "distributed as". We are free to choose the
-value of the parameters $\underline{\alpha}_t$ according to our prior
-information. A reasonable choice is to assume that before observing any data at
-all, we have no reason to believe any barcode is in higher abundance compared to
-any other. This might not be entirely true at the beginning of the experiment,
-where we expect the neutral barcodes to be in much higher frequency compared to
-the mutant barcodes. Nevertheless, this might not be true for the following time
-points, making our "uniform prior over the simplex" a reasonable choice. 
-Therefore, we define
+n_t = \sum_{b=1}^B r_t^{(b)}.
+$${#eq-freq_n_sum_r}
+But adding the variable allows us to factorize @eq-freq_lambda_bayes as
 $$
-\alpha_t^{(b)} = 1 \;\; \forall \; b.
-$${#eq-alpha_choice}
-
-For the likelihood term $\pi(\underline{r}_t \mid \underline{f}_t)$, we assume a
-very simple multinomial model. The story behind this choice goes as follows:
-Given a "true" underlying barcode frequency, when the sequencing machine reads a
-particular barcode, this is equivalent to having thrown a $B$-sided die, where
-the face of the die indicates the barcode that is read. This die is then rolled
-$R_t = \sum_b r_t^{(b)}$ times, giving the values of the reads we observe at
-time $t$. Notice that this model does not account for noise that might come from
-sources such as genetic drift, DNA extraction, and PCR amplification. Although
-these sources of noise are obviously important, it is not clear that we have
-enough statistical power to make meaningful statements about them given our data
-(more on this later on). We can then compactly write the likelihood function as
+\pi(\underline{\lambda}_t \mid n_t, \underline{r}_t) \propto
+\pi(n_t, \underline{r}_t \mid \underline{\lambda}_t) 
+\pi(n_t \mid \underline{\lambda}_t) \pi(\underline{\lambda}_t).
+$${#eq-freq_lambda_bayes_factorized}
+We then have
 $$
-\underline{r}_t \mid \underline{f}_t \sim
-\operatorname{Multinomial}(\underline{f}_t).
-$${#eq-freq_likelihood}
-
-This simple Multinomial-Dirichlet model for the probability of the frequency
-values is analytically tractable. Specifically, the Dirichlet distribution is
-the so-called *conjugate prior* to the multinomial distribution. This means the
-posterior distribution $\pi(\underline{f}_t \mid \underline{r}_t)$ has a
-closed-form solution of the same form as the prior $\pi(\underline{f}_t)$, but
-with updated parameters. In other words, we can write (without proof) the
-posterior distribution as
+\underline{r}_t \mid n_t, \underline{\lambda}_t \sim
+\text{Multinomial}(n_t, \underline{f}_t(\underline{\lambda}_t)).
+$${#eq-freq_r_bayes}
+Furthermore, we have
 $$
-\underline{f}_t \mid \underline{r}_t \sim
-\operatorname{Dirichlet}(\underline{\beta}_t),
-$${#eq-freq_posterior}
-where the new parameters are of the form
+n_t \mid \underline{\lambda}_t \sim 
+\text{Poiss}(\sum_{b=1}^B \lambda_t^{(b)}).
+$${#eq=freq_n_bayes}
+Finally, for our prior $\pi(\underline{\lambda}_t)$, we first assume each 
+parameter is independent, i.e.,
 $$
-\beta_t^{(b)} = \alpha_t^{(b)} + r_t^{(b)}.
+\pi(\underline{\lambda}_t) = \prod_{b=1}^B \pi(\lambda_t^{(b)}).
 $$
-Simply put, when the likelihood function is multinomial, and the prior 
-distribution is Dirichlet, the posterior distribution is also Dirichlet, where
-the parameters of the posterior are the parameters of the prior plus the number
-of observations of the corresponding output. This fact will become handy
-as we build the entire posterior distribution for our inference as defined in
-@eq-split_posterior.
+A reasonable prior for each $\lambda_t^{(b)}$ representing the expected number
+of reads for barcode $b$ should span several orders of magnitude. Furthermore,
+we assume that no barcode in the dataset ever goes extinct. Thus, no frequency
+can equal zero, facilitating the computation of the log frequency ratios
+needed to infer the relative fitness. The log-normal distribution satisfies 
+these constraints; therefore, for the prior, we assume
+$$
+\lambda_t^{(b)} \sim 
+\log\mathcal{N}(\mu_{\lambda_t^{(b)}}, \sigma_{\lambda_t^{(b)}}),
+$$
+with $\mu_{\lambda_t^{(b)}}, \sigma_{\lambda_t^{(b)}}$ as the user-defined 
+parameters that characterize the prior distribution.

### 2023-09-19 doc/_bayes_freq.qmd
update derivation of bayesian model SI sections

@@
 ---
+
+## Defining the Bayesian model
+
+Let us now define each of the terms in @eq-split_posterior described in
+@sec-bayesian_inference of the main text. The following sections will specify 
+the functional form each of these terms takes.
 
@@
 $$
-r^{(b)}_t \sim \text{Poiss}(\lambda^{(b)}_t),
+r^{(b)}_t \sim \operatorname{Poiss}(\lambda^{(b)}_t),
 $${#eq-freq_poisson_reads}
@@
 Poisson distribution has the convenient property that for two Poisson 
-distributed random variables $X \sim \text{Poiss}(\lambda_x)$ and 
-$Y \sim \text{Poiss}(\lambda_y)$, we have that
+distributed random variables $X \sim \operatorname{Poiss}(\lambda_x)$ and 
+$Y \sim \operatorname{Poiss}(\lambda_y)$, we have that
 $$
-Z \equiv X + Y \sim \text{Poiss}(\lambda_x + \lambda_y).
+Z \equiv X + Y \sim \operatorname{Poiss}(\lambda_x + \lambda_y).
 $${#eq-freq_additivity}
@@
 $$
-n_t \sim \text{Poiss}\left( \sum_{b=1}^B \lambda^{(b)}_t \right),
+n_t \sim \operatorname{Poiss}\left( \sum_{b=1}^B \lambda^{(b)}_t \right),
 $${#eq-freq_poisson_total}
@@
 $$
-\underline{r}_t \sim \text{Multinomial}(n_t, \underline{f}_t),
+\underline{r}_t \sim \operatorname{Multinomial}(n_t, \underline{f}_t),
 $${#eq-freq_multinomial}
@@
 \pi(\underline{\lambda}_t \mid n_t, \underline{r}_t) \propto
-\pi(n_t, \underline{r}_t \mid \underline{\lambda}_t) 
-\pi(n_t \mid \underline{\lambda}_t) \pi(\underline{\lambda}_t).
+\pi(\underline{r}_t \mid n_t, \underline{\lambda}_t)
+\pi(n_t \mid \underline{\lambda}_t)
+\pi(\underline{\lambda}_t)
 $${#eq-freq_lambda_bayes_factorized}
@@
 \underline{r}_t \mid n_t, \underline{\lambda}_t \sim
-\text{Multinomial}(n_t, \underline{f}_t(\underline{\lambda}_t)).
+\operatorname{Multinomial}(n_t, \underline{f}_t(\underline{\lambda}_t)).
 $${#eq-freq_r_bayes}
@@
 n_t \mid \underline{\lambda}_t \sim 
-\text{Poiss}(\sum_{b=1}^B \lambda_t^{(b)}).
+\operatorname{Poiss}\left(\sum_{b=1}^B \lambda_t^{(b)}\right).
 $${#eq=freq_n_bayes}
@@
 \log\mathcal{N}(\mu_{\lambda_t^{(b)}}, \sigma_{\lambda_t^{(b)}}),
-$$
+$${#eq-freq_lambda_prior}
 with $\mu_{\lambda_t^{(b)}}, \sigma_{\lambda_t^{(b)}}$ as the user-defined 
 parameters that characterize the prior distribution.
+
+#### Summary
+
+Putting all the pieces developed in this section together gives a term for our
+inference of the form
+$$
+\pi(\underline{\underline{F}} \mid \underline{\underline{R}}) \propto
+\prod_{t=1}^T\left\{
+    \pi(\underline{r}_t \mid n_t, \underline{\lambda}_t)
+    \pi(n_t \mid \underline{\lambda}_t)
+    \left[ 
+        \prod_{b=1}^B \pi(\lambda_t^{(b)})
+    \right]
+\right\}
+$${#eq-freq_final_01}
+where
+$$
+\underline{r}_t \mid n_t, \underline{\lambda}_t \sim
+\operatorname{Multinomial}(n_t, \underline{f}_t(\underline{\lambda}_t)),
+$${#eq-freq_final_02}
+$$
+n_t \mid \underline{\lambda}_t \sim 
+\operatorname{Poiss}\left(\sum_{b=1}^B \lambda_t^{(b)}\right).
+$${#eq-freq_final_03}
+and
+$$
+\lambda_t^{(b)} \sim 
+\log\mathcal{N}(\mu_{\lambda_t^{(b)}}, \sigma_{\lambda_t^{(b)}}),
+$${#eq-freq_final_04}

### 2023-09-19 doc/_bayes_meanfit.qmd
update derivation of bayesian model SI sections

@@
 $$
-\frac{f_{t+1}^{(n)}}{f_t^{(n)}} = \mathrm{e}^{- \bar{s}_t}\tau.
+\frac{f_{t+1}^{(n)}}{f_t^{(n)}} = \mathrm{e}^{- \bar{s}_t\tau}.
 $${#eq-fitness_ratio_neutral}
@@
 $\varepsilon_t^{(n)}$, characterizing the deviations of the measurements from
-the theoretical model. It is common to assume that these errors are normally
-distributed with mean zero and some standard deviation $\sigma_t$, implying that
+the theoretical model. We assume these errors are normally distributed with mean
+zero and some standard deviation $\sigma_t$, implying that
 $$
@@
 $$
-\sigma_t \sim \operatorname{Half}-\mathcal{N}(0, \sigma_{\sigma_t}),
+\sigma_t \sim \log\mathcal{N}(\mu_{\sigma_t}, \sigma_{\sigma_t}),
 $${#eq-meanfit_eq4}
-where $\sigma_{\bar{s}_t}$ and $\sigma_{\sigma_t}$ are free parameters we can
-choose.
+where $\sigma_{\bar{s}_t}$, $\mu_{\sigma_t}$, and $\sigma_{\sigma_t}$ are
+user-defined parameters.

### 2023-09-19 doc/_bayes_mutfit.qmd
update derivation of bayesian model SI sections

@@
 $$
-\sigma^{(m)} \sim \operatorname{Half}-\mathcal{N}(0, \sigma_{\sigma^{(m)}}),
+\sigma^{(m)} \sim \log\mathcal{N}(\mu_{\sigma^{(m)}}, \sigma_{\sigma^{(m)}}),
 $${#eq-mutfit_eq4}
-where $\sigma_{s^{(m)}}$ and $\sigma_{\sigma^{(m)}}$ are free parameters we can
-choose.
+where $\sigma_{s^{(m)}}$, $\mu_{\sigma^{(m)}}$, and $\sigma_{\sigma^{(m)}}$ are
+user-defined parameters.

### 2023-09-19 doc/_logistic_sim.qmd
add logistic growth section

---
editor:
    render-on-save: true
bibliography: references.bib
csl: ieee.csl
---

## Logistic growth simulation

In this section, we explain the simulations used to assess the validity of our
inference pipeline. Let us begin by assuming that, since the strains are grown
for two full days in the experiment, having left behind the exponential phase
for almost an entire day, a simple exponential growth of the form
$$
\frac{dn_i}{dt} = \lambda_i n_i,
$${#eq-logistic_expo}
where $n_i$ is the number of cells of strain $i$, and $\lambda_i$ is the
corresponding growth rate is not enough. Instead, we will assume that the cells
follow the logistic growth equation of the form
$$
\frac{dn_i}{dt} = \lambda_i n_i 
\left( 1 - \frac{\sum_{j=1}^{N}n_i}{\kappa}\right),
$${#eq-logistic_fn}
where $\kappa$ is the carrying capacity, and $N$ is the total number of strains
in the culture.

The inference method is based on the model that assumes that the time passed
between dilutions $\tau \approx 8$ generations, the change in frequency for a
mutant barcode can be approximated from cycle $t$ to the next cycle $t + 1$ as
$$
f_{t+1}^{(m)} = f_t^{(m)} \mathrm{e}^{(s^{(m)} - \bar{s}_t)\tau},
$${#eq-logistic_fitness_fn}
where $s^{(m)}$ is the relative fitness for strain $i$ compared to the ancestral
strain and $\bar{s}_t$ is the mean fitness of the population at cycle $t$. We
implemented a numerical experiment following the logistic growth model described
in @eq-logistic_fn to test this assumption. @fig-SIX_logistic_growth_01 shows an
example of the deterministic trajectories for 50 labeled neutral lineages and
1000 lineages of interest. The upper red curve that dominates the culture 
represents the unlabeled ancestral strain included in the experimental design
described in @sec-experiment.

![**Logistic growth simulation over single growth cycle**. The dashed line
represents the neutral lineages, with the upper curve being the unlabeled
neutral strain. Color curves represent the genotypes of interest colored by
growth rate relative to the neutral
lineage.](./figs/figSIX_logistic_growth_01){#fig-SIX_logistic_growth_01}

To simulate multiple growth-dilution cycles, we take the population composition
at the final time point and use it to initialize a new logistic growth
simulation. @fig-SIX_logistic_growth_02 shows the resulting number of cells at
the final time point of a cycle over multiple growth-dilution cycles for the
genotypes in @@fig-SIX_logistic_growth_01. We can see that the adaptive lineages
(blue curves) increase in abundance, while detrimental lineages (red curves)
decrease.

![**Growth-dilution cycles for logistic growth simulation**. Each point
represents the final number of cells after a growth cycle for each lineage.
Colors are the same as in
@fig-SIX_logistic_growth_01.](./figs/figSIX_logistic_growth_02){#fig-SIX_logistic_growth_02}

In @sec-fitness_model we derive the functional form to infer the relative 
fitness of each lineage as
$$
\frac{1}{\tau}\ln \frac{f_{t+1}^{(b)}}{f_{t}^{(b)}} = (s^{(b)} - \bar{s}_t).
$${#eq-logistic_logfreq}
@fig-SIX_logistic_growth_03 shows the corresponding log frequency ratio curves
for the logistic growth simulation. The displacement of these curves with 
respect to the neutral lineages determines the ground truth relative fitness
value for these simulations.

![**Log frequency ratio for logistic growth simulations**. The relative distance
of the color curves from the black dashed line determines the relative fitness
of each linage.](./figs/figSIX_logistic_growth_03){#fig-SIX_logistic_growth_03}

To simulate the experimental noise, we add two types of noise:

1. Poisson noise between dilutions. For this, we take the final point of the
logistic growth simulation and sample a random Poisson number based on this
final point to set the initial condition for the next cycle.
2. Gaussian noise when performing the measurements. When translating the 
underlying population composition to the number of reads, we can add a custom 
amount of Gaussian noise.

@fig-SIX_logistic_growth_04 shows the frequency trajectories (left panels) and
log frequency ratios (right panels) for a noiseless simulation (upper panels) 
and a simulation with added noise (lower panels). The noiseless simulation is
used to determine the relative fitness for each of the lineages, which serves
as the ground truth to be compared with the resulting inference.

![**Logistic growth-dilution simulations with and without
noise**](./figs/figSIX_logistic_growth_04){#fig-SIX_logistic_growth_04}

### 2023-09-19 doc/_logistic_sim.qmd
implement grammar suggestiosn

@@
 where $s^{(m)}$ is the relative fitness for strain $i$ compared to the ancestral
-strain and $\bar{s}_t$ is the mean fitness of the population at cycle $t$. We
-implemented a numerical experiment following the logistic growth model described
-in @eq-logistic_fn to test this assumption. @fig-SIX_logistic_growth_01 shows an
-example of the deterministic trajectories for 50 labeled neutral lineages and
-1000 lineages of interest. The upper red curve that dominates the culture 
-represents the unlabeled ancestral strain included in the experimental design
-described in @sec-experiment.
+strain and $\bar{s}_t$ is the mean fitness of the population at cycle $t$. To
+test this assumption, we implemented a numerical experiment following the
+logistic growth model described in @eq-logistic_fn. @fig-SIX_logistic_growth_01
+shows an example of the deterministic trajectories for 50 labeled neutral
+lineages and 1000 lineages of interest. The upper red curve that dominates the
+culture represents the unlabeled ancestral strain included in the experimental
+design described in @sec-experiment.
 
@@
 simulation. @fig-SIX_logistic_growth_02 shows the resulting number of cells at
-the final time point of a cycle over multiple growth-dilution cycles for the
+the last time point of a cycle over multiple growth-dilution cycles for the
 genotypes in @@fig-SIX_logistic_growth_01. We can see that the adaptive lineages
@@
 
-In @sec-fitness_model we derive the functional form to infer the relative 
+In @sec-fitness_model, we derive the functional form to infer the relative 
 fitness of each lineage as
@@
 logistic growth simulation and sample a random Poisson number based on this
-final point to set the initial condition for the next cycle.
+last point to set the initial condition for the next cycle.
 2. Gaussian noise when performing the measurements. When translating the 

### 2023-09-20 doc/_bayes_hierarchical.qmd
add first draft for hierarchical model SI

---
editor:
    render-on-save: true
bibliography: references.bib
csl: ieee.csl
---

### Hierarchical models for multiple experimental replicates

As detailed in @sec-replicates of the main text, we define a Bayesian
hierarchical model to analyze data from multiple experimental replicates. The
implementation requires only slightly modifying the base model detailed in the
previous sections. The hierarchical model defines a hyper-fitness parameter
$\theta^{(m)}$ for every non-neutral barcode. We can thus collect all of the $M$
hyperparameters in an array of the form
$$
\underline{\theta}^M = (\theta^{(1)}, \ldots, \theta^{(M)})^\dagger.
$${#eq-hier_hyper_vector}
Our data now consists of a series of matrices $\underline{\underline{R}}_{[j]}$,
where the subindex $[j]$ refers to the $j$-th experimental replicate. These
matrices need not have the same number of rows, as the time points measured for
each replicate can vary. The statistical model we must define is then of the
form
$$
\begin{aligned}
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\},
    \{\underline{\underline{F}}_{[j]}\} \mid
    \{\underline{\underline{R}}_{[j]}\}
) \propto \;
&\pi(
    \{\underline{\underline{R}}_{[j]}\} \mid
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\},
    \{\underline{\underline{F}}_{[j]}\} 
) \times \\\
&\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\},
    \{\underline{\underline{F}}_{[j]}\} 
)
\end{aligned}
$${#eq-hier_bayes_full}
where the parameters within curly braces with subindex $[j]$ indicate one set of
parameters per experimental replicate. For example,
$$
\{\underline{s}^M_{[j]}\} = \{
    \underline{s}^M_{[1]}, \underline{s}^M_{[2]}, \ldots \underline{s}^M_{[E]}
\},
$${#eq-hier_array}
where $E$ is the number of experimental replicates.

Given the dependencies between the variables, we can factorize
@eq-hier_bayes_full to be of the form
$$
\begin{aligned}
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\},
    \{\underline{\underline{F}}_{[j]}\} \mid
    \{\underline{\underline{R}}_{[j]}\}
) =
&\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\} \mid
    \{\underline{\bar{s}}_{T[j]}\},
    \{\underline{\underline{F}}_{[j]}\} 
) \times \\
&\pi(
    \{\underline{\bar{s}}_{T[j]}\} \mid
    \{\underline{\underline{F}}_{[j]}\}
) \times \\
&\pi(
    \{\underline{\underline{F}}_{[j]}\} \mid
    \{\underline{\underline{R}}_{[j]}\}
)
\end{aligned}
$${#eq-hier_factorized}
Furthermore, the hierarchical structure only connects the replicates via the
relative fitness parameters. This means that the population mean fitness values
and the frequencies can be independently inferred for each dataset. This allows
us to rewrite the right-hand side of @eq-hier_factorized as
$$
\begin{aligned}
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\},
    \{\underline{\underline{F}}_{[j]}\} \mid
    \{\underline{\underline{R}}_{[j]}\}
) =
&\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\} \mid
    \{\underline{\bar{s}}_{T[j]}\},
    \{\underline{\underline{F}}_{[j]}\} 
) \times \\
&\prod_{j=1}^E \left[
    \pi(
        \underline{\bar{s}}_{T[j]} \mid
        \underline{\underline{F}}_{[j]}
    )
    \pi(
        \underline{\underline{F}}_{[j]} \mid
        \underline{\underline{R}}_{[j]}
    )
\right].
\end{aligned}
$${#eq-hier_bayes_factorized_replicate}
The terms inside the square brackets in @eq-hier_bayes_factorized_replicate take
the same functional form as those derived in @sec-bayes_meanfit and
@sec-bayes_freq. Therefore, to implement the desired hierarchical model, we only
need to focus on the first term on the right-hand side of
@eq-hier_bayes_factorized_replicate. A way to think about the structure of the
hierarchical model is as follows: imagine each genotype as a "*true*" relative
fitness value. However, every time we perform an experiment, small variations in
the biotic and abiotic conditions---also known as batch effects---might result
in small deviations from this value. We model this by defining a distribution
for the hyper-fitness parameter---the ground truth we are interested in---and
having each experimental replicate sample from this hyper-parameter distribution
to determine the "*local*" fitness value. The wider the hyper-parameter
distribution is the more variability between experimental replicates.

Writing Bayes' theorem for the first term in @eq-hier_bayes_factorized_replicate
results in
$$
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\} \mid
    \{\underline{\underline{F}}_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\}
) \propto
\pi(
    \{\underline{\underline{F}}_{[j]}\} \mid
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\}
)
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\} \mid
    \{\underline{\bar{s}}_{T[j]}\}
),
$${#eq-hier_bayes}
where we leave the conditioning on the population mean fitness as we did in
@seq-bayes_mutfit. This expression can be simplified in two ways. First, the
frequency values for each experimental replicate depend directly on the local
fitness values and the corresponding population mean fitness, as the
relationship between experimental replicates only occurs through the relative
fitness values. Therefore, we can write
$$
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\} \mid
    \{\underline{\underline{F}}_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\}
) \propto
\prod_{j=1}^E\left[
    \pi(
        \underline{\underline{F}}_{[j]} \mid
        \underline{s}^M_{[j]},
        \underline{\bar{s}}_{T[j]}
    )
\right]
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\} \mid
    \{\underline{\bar{s}}_{T[j]}\}
).
$${#eq-hier_bayes_exp_indep}
Second, the relationship between the hyper-fitness and the local fitness values
allows us to write their joint distribution as a conditional distribution where
local fitness values depend on the global hyper-fitness value, obtaining
$$
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\} \mid
    \{\underline{\underline{F}}_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\}
) \propto
\prod_{j=1}^E\left[
    \pi(
        \underline{\underline{F}}_{[j]} \mid
        \underline{s}^M_{[j]},
        \underline{\bar{s}}_{T[j]}
    )
    \pi(
        \underline{s}^M_{[j]} \mid
        \underline{\theta}^M
    )
\right] 
\pi(
    \underline{\theta}^M
).
$${#eq-hier_bayes_exp_indep}
Notice we removed the conditioning on the population mean fitness as our prior
expectations of what the global hyper-fitness or local fitness value might be do
not depend on these nuisance parameters.

The first term on the right-hand side of @eq-hier_bayes_exp_indep takes the same
functional form as the one derived in @seq-bayes_mutfit. Therefore, all we are
left with is to determine the functional forms for the hyper–prior
$\pi(\underline{\theta}^M)$, and the conditional probability
$\pi(\underline{s}^M_{[j]} \mid \underline{\theta}^M)$. In analogy to the 
assumptions used for the fitness values in @seq-bayes_mutfit, we define the 
value of each hyper-fitness as independent. This means that we have
$$
\pi(\underline{\theta}^M) = \prod_{m=1}^M \pi(\theta^{(m)}).
$${#eq-hier_hyper_prior}
Furthermore, we assume this prior is of the form
$$
\theta^{(m)} \sim \mathcal{N}(\mu_{\theta^{(m)}}, \sigma_{\theta^{(m)}}),
$${#eq-hier_hyper_prior_02}
where $\mu_{\theta^{(m)}}$ and $\sigma_{\theta^{(m)}}$ are user-defined 
parameters encoding the prior expectations on the fitness values.

For the conditional distribution $\pi(\underline{s}^M_{[j]} \mid
\underline{\theta}^M)$, we use the so-called non-centered parametrization that
avoids some of the intrinsic degeneracies associated with hierarchical models
[@betancourt2013]. We invite the reader to check [this excellent
blog](https://betanalpha.github.io/assets/case_studies/hierarchical_modeling.html)
explaining the difficulties of working with hierarchical models. This 
non-centered parameterization implies that we introduce two nuisance parameters
such that the local fitness $s_{[j]}^{(m)}$ is computed as
$$
s_{[j]}^{(m)} = \theta^{(m)} + 
(\tau_{[j]}^{(m)} \times \xi_{[j]}^{(m)}),
$${#eq-hier_noncentered}
where $\theta^{(m)}$ is the corresponding genotype hyper-fitness value,
$\xi_{[j]}^{(m)}$ is a standard normal random variable, i.e.,
$$
\xi_{[j]}^{(m)} \sim \mathcal{N}(0, 1),
$${#eq-hier_stdnormal}
that allows deviations from the hyper-fitness value to be either positive or
negative, and $\tau_{[j]}^{(m)}$ is a strictly positive random variable that
characterizes the deviation of the local fitness value from the global
hyper-fitness. We assume
$$
\tau_{[j]}^{(m)} \sim \log\mathcal{N}(
    \mu_{\tau_{[j]}^{(m)}}, \sigma_{\tau_{[j]}^{(m)}}
)
$${#eq-hier_tau}
where $\mu_{\tau_{[j]}^{(m)}}$ and $\sigma_{\tau_{[j]}^{(m)}}$ are user-defined
parameters capturing the expected magnitude of the batch effects.

### 2023-09-26 doc/_bayes_priors.qmd
add (incomplete) section on prior selection

---
editor:
    render-on-save: true
bibliography: references.bib
csl: ieee.csl
---

### Defining prior probabilities

One aspect commonly associated---in both positive and negative ways---to
Bayesian analysis is the definition of prior probabilities. On the one hand, the
naive textbook version of Bayesian analysis defines the prior as encoding the
information we have about the inference in question before acquiring any data.
This is the "ideal" use of priors that, whenever possible, should be used. On
the other hand, for most practitioners of Bayesian statistics in the age of big
data, the definition of prior becomes a tool to ensure the convergence of
sampling algorithms such as MCMC [@gelman2017]. However, for our particular
problem, although we deal with large amounts of data (inferences can be made for
> 10K barcodes over multiple time points, resulting in > 100K parameters), each
barcode has very little data, as they are measured only once per time point over
< 10 times. Furthermore, understanding the sources of noise having to do with
culturing conditions, DNA extraction, library preparation, and so on and
encoding them into reasonable prior distributions becomes incredibly
challenging.

Empirically, our approach for this work defined the priors based solely on the
neutral lineage data, as they represent the only repeated measurements of a
single genotype in our experimental design. We acknowledge that defining the
priors after observing the data might be considered an incoherent inference.
However, as expressed by @gelman2017

> Incoherence is an unavoidable aspect of much real-world data analysis; and,
indeed, one might argue that as scientists we learn the most from the anomalies
and reassessments associated with episodes of incoherence.

With this in mind, we leave it to the reader to judge the selection of priors.
Furthermore, the present package `BayesFitness.jl` is written so that users can
experiment with different prior selection criteria that fit their needs. We
strongly advocate that statistics should not be done in a black-box fit-all tool
mindset but rather as a formal way to encode the assumptions behind the
analysis, subject to constructive criticism.

With this philosophical baggage behind us, let us now focus on how the priors
used for this work were selected.

#### Naive neutral lineage-based priors

For the base model presented in this work, the user-defined prior parameters
include the following:

- Prior on population mean fitness (one per pair of adjacent time points)
$$
\bar{s}_t \sim \mathcal{N}(\mu_{\bar{s}_t}, \sigma_{{\bar{s}_t}}).
$${#eq-prior_meanfit_mean}

- Prior on standard deviation associated with neutral lineages likelihood
function (one per pair of adjacent time points)
$$
\sigma_t \sim \log\mathcal{N}(\mu_{\sigma_t}, \sigma_{\sigma_t}).
$${#eq-prior_meanfit_var}

- Prior on relative fitness (one per non-neutral barcode)
$$
s^{(m)} \sim \mathcal{N}(\mu_{s^{(m)}}, \sigma_{s^{(m)}}).
$${#eq-prior_fit_mean}

- Prior on standard deviation associated with non-neutral lineages likelihood
function (one per non-neutral barcode)
$$
\sigma^{(m)} \sim \log\mathcal{N}(\mu_{\sigma^{(m)}}, \sigma_{\sigma^{(m)}})
$${#eq-prior_fit_var}

The `BayesFitness.jl` package includes a function `naive_prior` within the
`stats` module. This function utilizes the data from the neutral lineages to
determine some of the prior parameters to facilitate the inference algorithm's
numerical convergence. In particular, it defines the population mean fitness
parameter $\mu_{\bar{s}_t}$ as
$$
\mu_{\bar{s}_t} = \frac{1}{N} \sum_{n=1}^N 
-\ln\left( \frac{r_{t+1}^{(n)}}{r_t^{(n)}} \right),
$${#eq-prior_meanfit_mean_data}
where $N$ is the number of neutral lineages and $r_t^{(n)}$ is the number of
neutral lineages. In other words, it defines the mean of the prior distribution
as the mean of what one naively would compute from the neutral lineages,
discarding cases where the ratio diverges because the denominator $r_t^{(n)} =
0$. For the variance parameter, we chose a value $\sigma_{\bar{s}_t} = 0.05$.

Furthermore, the `naive_prior` function defines the mean of the variance
parameter as the standard deviation of the log frequency ratios for the neutral
lineages, i.e.,
$$
\mu_{\sigma_t} = \sqrt{
    \operatorname{Var}\left(\frac{r_{t+1}^{(n)}}{r_t^{(n)}}\right)
}.
$${#eq-prior_meanfit_var_data}
This same value was utilized for the

### 2023-09-26 doc/_ppc.qmd
add SI section for how to compute PPCs

---
editor:
    render-on-save: true
bibliography: references.bib
csl: ieee.csl
---

## Posterior predictive checks

Throughout the main text, we allude to the concept of posterior predictive
checks as a formal way to assess the accuracy of our inference pipeline. Here,
we explain the mechanics behind the computation of these credible regions, given
the output of the inference.

Bayesian models encode what is known as a *generative model*. This statement
means that in our definition of the likelihood function and the prior
distribution, we, as modelers, propose a mathematical function that captures all
relevant relationships between unobserved (latent) variables. Therefore, when
these latent variables are input into the mathematical model, this function
*generates* data that should be, in principle, indistinguishable from the real
observations if the model is a good account of the underlying processes involved
in the phenomena of interest. This generative model implies that once we run the
inference process and update our posterior beliefs about the state of the latent
variables, we can input back the inferred values to our model and generate
synthetic data. Furthermore, we can repeat this process multiple times to
compute the range where we expect to observe our data conditioned on the
accuracy of the model.

For our specific scenario, recall that our objective is to infer the relative
fitness of a non-neutral lineage $s^{(m)}$ along with nuisance parameters
related to the population mean fitness at each point, $\underline{\bar{s}}_t$,
and the barcode frequency time series $\underline{f}^{(m)}$. All these variables
are related through our fitness model (see @sec-fitness_model in the main text)
$$
f_{t+1}^{(m)} = f_{t}^{(m)} \mathrm{e}^{(s^{(m)} - s_t)\tau}.
$${#eq-ppc_model}
As we saw, it is convenient to rewrite @eq-ppc_model as
$$
\frac{1}{\tau}\ln\frac{f_{t+1}^{(m)}}{f_{t}^{(m)}} = (s^{(m)} - s_t).
$${#eq-ppc_logratio}
Written in this way, we separate the quantities we can compute from the 
experimental observations---the left-hand side of @eq-ppc_logratio can be 
computed from the barcode reads---from the latent variables.

Although we perform the joint inference over all barcodes in the present work,
let us focus on the inference task for a single barcode as if it were computed
independently. For a non-neutral barcode, our task consists of computing the 
posterior probability
$$
\pi(\theta \mid \underline{r}^{(m)}) =
\pi(
    s^{(m)}, \sigma^{(m)}, \underline{s}_t, \underline{f}^{(m)}
    \mid \underline{r}^{(m)}
),
$${#eq-ppc_posterior}
where $\theta$ represents all parameters to be inferred and
$\underline{r}^{(m)}$ is the vector with the barcode raw counts time series. The
list of parameters are

- $s^{(m)}$: The barcode's relative fitness.
- $\sigma^{(m)}$: A nuisance parameter used in the likelihood to generate the
data. This captures the expected deviation from @eq-ppc_logratio
- $\underline{s}_t$: The vector with all population mean fitness for each pair
of adjacent time points.
- $\underline{f}^{(m)}$: The vector with the barcode frequency time series.

Furthermore, let us define a naive estimate of the barcode frequency at time $t$
as 
$$
\hat{f}_t^{(m)} = \frac{r^{(m)}_t}{\sum_{b=1}^B r^{(b)}_t}.
$${#eq-ppc_naive_freq}
We can compute this quantity from the data by normalizing the raw barcode counts
by the sum of all barcode counts. Furthermore, we can compute a naive estimate
of the log frequency ratio from the raw barcode counts as
$$
\ln \hat{\gamma}_t^{(m)} = \ln \frac{\hat{f}_{t+1}^{(m)}}{\hat{f}_t^{(m)}}
$${#eq-ppc_naive_logfreq}

In our generative model, we assumed
$$
\ln\gamma^{(m)}_t \mid \theta \sim \mathcal{N}(s^{(m)} - s_t, \sigma^{(m)}).
$${#eq-ppc_logfreq_likelihood}
This implies that once we determine the posterior distribution of our
parameters, we can generate synthetic values of $\ln \gamma^{(m)}_t$ that we can
then compare with the values obtained from applying @eq-ppc_naive_logfreq and
@eq-ppc_naive_logfreq to the raw data.

In practice, to compute the posterior predictive checks, we generate multiple samples from the posterior distribution $\pi(\theta \mid \underline{r}^{(m)})$ 
$$
\underline{\theta} = (\theta_1, \theta_2, \ldots, \theta_N).
$${#eq-ppc_theta_samples}
With these samples in hand, the `BayesFitness.jl` package includes the function
`logfreq_ratio_bc_ppc` for non-neutral barcodes that uses this set of posterior
parameter samples to generate samples from the distribution defined in
@eq-ppc_logfreq_likelihood. For a large-enough number of samples, we can then
compute the desired percentiles---5, 68, and 95 percentiles in all figures in 
the main text---that are equivalent to the corresponding credible regions. In
other words, the range of values of $\ln\gamma_t^{(m)}$ generated by this 
bootstrap process can be used to compute the region where we expect to find our
raw estimates $\ln\hat{\gamma}_t^{(m)}$ with the desired probability. The
package `BayesFitness.jl` includes an equivalent function,
`logfreq_ratio_popmean_ppc`, for neutral lineages.

### 2023-09-26 doc/_intro.qmd
adapt intro to new objective

@@
 barcode sequencing is inexorably accompanied by noise sources coming from
-experimental manipulation of the microbial cultures, DNA extraction and
+experimental manipulation of the microbial cultures, DNA extraction, and
 sequencing library preparation that involves multiple rounds of PCR
 amplification, and the sequencing process itself. Thus, accounting for the
-uncertainty when inferring the relevant parameters from the data is a key step
-to draw reliable conclusions. Bayesian statistics presents a paradigm by which
-one can account for all known sources of uncertainty in a principled way
+uncertainty when inferring the relevant parameters from the data is a crucial
+step to draw reliable conclusions. Bayesian statistics presents a paradigm by
+which one can account for all known sources of uncertainty in a principled way
 [@eddy2004a]. This, combined with the development of modern Markov Chain Monte
@@
 inferring the lineages' relative fitness values (see @sec-experiment for details
-on sources of uncertainty accounted by the model). Furthermore, small changes to
-the core statistical model allow us to account for relevant experimental
-variations of these competition assays. More specifically, as described in
-@sec-replicates, our statistical model can account for batch-to-batch
-differences when jointly analyzing multiple experimental replicates using a
-Bayesian hierarchical model. Moreover, a variant of these hierarchical models,
-presented in @sec-genotypes, can account for variability within multiple
-barcodes mapping to equivalent genotypes within the same experiment. Finally, in
+on sources of uncertainty accounted for by the model). Furthermore, minor
+changes to the core statistical model allow us to account for relevant
+experimental variations of these competition assays. More specifically, in
 @sec-multienv, we present a variation of the statistical model to infer fitness
-on growth-dilution cycles in multiple environments.
+on growth dilution cycles in multiple environments with proper error
+propagation. Furthermore, as described in @sec-replicates, our statistical model
+can account for batch-to-batch differences when jointly analyzing multiple
+experimental replicates using a Bayesian hierarchical model. Finally, a variant
+of these hierarchical models, presented in @sec-genotypes, can account for
+variability within multiple barcodes mapping to equivalent genotypes within the
+same experiment.
 
@@
 of our pipeline to infer relative fitness parameters against synthetic data with
-known ground truth. We also re-analyze published data from @kinsler2020 to
-assess the method's performance on experimental measurements. The present method
-is accompanied by a `Julia` package to readily implement the inference pipeline
-with state-of-the-art scientific computing software. [`here mention something
-about the Python wrapper if appropriate`].
+known ground truth. A `Julia` package accompanies the present method to readily
+implement the inference pipeline with state-of-the-art scientific computing
+software.

### 2023-09-26 doc/_experiment.qmd
small changes to experiment section

@@
 coli*. @fig-01(B) presents a typical barcode trajectory where the darker barcode
-trajectories represent the so-called *neutral lineages* that are genetically
-equivalent to the untagged ancestor strain that initially dominates the culture.
-These neutral lineages simplify the inference problem since the fitness of all
+trajectories represent the so-called *neutral lineages*, genetically equivalent
+to the untagged ancestor strain that initially dominates the culture. These
+neutral lineages simplify the inference problem since the fitness of all
 relevant barcodes is quantified with respect to these barcodes.
@@
 dataset. From each time point, the relative frequency of each barcode is
-determined from the total number of reads. Different shades of blue represent
-different relative fitness. Darker gray lines define the typical trajectory of
-neutral lineages.](./figs/fig01){#fig-01}
+determined from the total number of reads. Shades of blue represent different
+relative fitness. Darker gray lines define the typical trajectory of neutral
+lineages.](./figs/fig01){#fig-01}

### 2023-09-26 doc/_fitness_model.qmd
update from time to cycle number

@@
 $${#eq-fitness}
-where $f_{t}^{(b)}$ is the frequency of barcode $b$ at time $t$, $s^{(b)}$ is
-the relative fitness with respect to the reference strain---the quantity we want
-to infer from the data---$\bar{s}_t$ is the mean fitness of the culture at time
-$t$, and $\tau$ is the time pass between time $t$ and $t+1$. We can rewrite
-@eq-fitness as
+where $f_{t}^{(b)}$ is the frequency of barcode $b$ at the end of cycle number
+$t$, $s^{(b)}$ is the relative fitness with respect to the reference
+strain---the quantity we want to infer from the data---$\bar{s}_t$ is the mean
+fitness of the culture at the end of cycle number $t$, and $\tau$ is the time
+pass between cycle $t$ and $t+1$. We can rewrite @eq-fitness as
 $$
@@
 @eq-logfreq separates the measurements---the barcode frequencies---from the
-unobserved---sometimes referred to as latent---parameters we want to infer from
+unobserved (sometimes referred to as latent) parameters we want to infer from
 the data---the population mean fitness and the barcode relative fitness. This is
@@
 determined relative to the neutral lineage that dominates the culture, we define
-their fitness to be $s^{(n)} = 0$, where the subscript $(n)$ specifies their
+their fitness to be $s^{(n)} = 0$, where the superscript $(n)$ specifies their
 neutrality. This means that @eq-logfreq for a neutral lineage takes the simpler
@@
 where, importantly, the proportionality constant depends on the total number of
-reads for the library at time $t$, which might vary from library to library.
+reads for the library for cycle $t$, which might vary from library to library.
 Therefore, to compare the number of reads between libraries at different time
@@
 1. Uncertainty in the determination of frequencies. Our model relates
-frequencies between adjacent time points to the fitness of the corresponding
-strain. However, we do not directly measure frequencies. Instead, our data for
-each barcode consists of a length $T$ vector of counts $\underline{r}^{(b)}$ for
-each of the $T$ time points in which the measurements were taken.
+frequencies between adjacent growth-dilution cycles to the fitness of the
+corresponding strain. However, we do not directly measure frequencies. Instead,
+our data for each barcode consists of a length $T$ vector of counts
+$\underline{r}^{(b)}$ for each of the $T$ cycles in which the measurements were
+taken.
 2. Uncertainty in the value of the population mean fitness. We define neutral
 lineages to have fitness $s^{(n)} = 0$, helping us anchor the value of the
-population mean fitness $\bar{s}_t$ for each pair of adjacent time points. 
-Moreover, we take this parameter as an empirical observable from the data, 
-meaning that we do not have a functional form that relates $\bar{s}_t$ to
-$\bar{s}_{t+1}$. Thus, we must infer the $T-1$ values of this population mean 
-fitness with their uncertainty that must be propagated to the value of
-the mutants' relative fitness.
+population mean fitness $\bar{s}_t$ for each pair of adjacent growth cycles.
+Moreover, we take this parameter as an empirical parameter to be obtained from
+the data, meaning that we do not impose a functional form that relates
+$\bar{s}_t$ to $\bar{s}_{t+1}$. Thus, we must infer the $T-1$ values of this
+population mean fitness with their uncertainty that must be propagated to the
+value of the mutants' relative fitness.
 3. Uncertainty in each of the mutants' fitness values.

### 2023-09-26 doc/_bayesian_inference.qmd
small fixes to section

@@
 barcodes given by the sum of the number of unique, relevant barcodes we care
-about (hereafter referred to as mutant barcodes), $M$, and the number of unique
-neutral barcodes, $N$, and $T$ is the number of time points where measurements
-were taken. The data matrix is then of the form
+about, $M$, and the number of unique neutral barcodes, $N$, and $T$ is the
+number of growth cycles where measurements were taken. The data matrix is then
+of the form
 $$
@@
 $${#eq-R_mat}
-where each row $\underline{r}_t$ is a $B$-dimensional array containing the
-raw barcode counts at time $t$. We can further split each vector
-$\underline{r}_t$ into two vectors of the form
+where each row $\underline{r}_t$ is a $B$-dimensional array containing the raw
+barcode counts at cycle $t$. We can further split each vector $\underline{r}_t$
+into two vectors of the form
 $$
 \underline{r}_t = \begin{bmatrix}
-\underline{r}_t^{N} \;
+\underline{r}_t^{N} \\
 \underline{r}_t^{M}
@@
 $${#eq-R-mat_split}
-where $\underline{\underline{R}}^N$ is a $T \times N$ matrix with the
-barcode reads time series for each neutral lineage and
-$\underline{\underline{R}}^M$ is the equivalent $T \times M$ matrix for the
-mutant lineages.
+where $\underline{\underline{R}}^N$ is a $T \times N$ matrix with the barcode
+reads time series for each neutral lineage and $\underline{\underline{R}}^M$ is
+the equivalent $T \times M$ matrix for the non-neutral lineages.
 
@@
 otherwise stated. When defining our statistical model, we need not to focus on
-the denominator on the right-hand side of @eq-bayes_obj, and write
+the denominator on the right-hand side of @eq-bayes_obj. Thus, we can write
 $$
@@
 \pi(\underline{\underline{R}} \mid \underline{s}^M) 
-\pi(\underline{s}^M)
+\pi(\underline{s}^M).
 $${#eq-bayes_obj_propto}
 However, when implementing the model computationally, the normalization constant
-on the right-hand side of @eq-bayes_obj must be computed via an integral of the
-form
+on the right-hand side of @eq-bayes_obj must be computed. This can be done from
+the definition of the model via an integral of the form
 $$
-\pi(\underline{\underline{R}}) = \int d\underline{s}^M
+\pi(\underline{\underline{R}}) = \int d^M \underline{s}^M
 \pi(\underline{\underline{R}} \mid \underline{s}^M) 
@@
 $${#eq-bayes_evidence}
-also known as a marginalization integral.
+also known as a marginalization integral. Hereafter, differentials of the form
+$d^n$ imply a $n$-dimensional integral.
 
@@
 $T$ time points where measurements were taken. We have $T-1$ since the value
-of any $\bar{s}_t$ requires time points $t$ and $t+1$. Furthermore, let the
+of any $\bar{s}_t$ requires cycle numbers $t$ and $t+1$. Furthermore, let the
 matrix $\underline{\underline{F}}$ be a $T \times B$ matrix containing all
-frequency values. As with @eq-R_mat, we can split $\underline{\underline{F}}$
-into two matrices of the form
+frequency values. As with @eq-R-mat_split, we can split
+$\underline{\underline{F}}$ into two matrices of the form
 $$
@@
 $${#eq-F-mat_split}
-to separate the corresponding mutant and neutral barcode frequencies. With these
-nuisance variables in hand, the full inference problem we must solve takes the
-form
+to separate the corresponding neutral and non-neutral barcode frequencies. With
+these nuisance variables in hand, the full inference problem we must solve takes
+the form
 $$
@@
 $${#eq-bayes_full}
-We can numerically integrate out all nuisance parameters to recover the marginal
-distribution we care about, i.e.,
+To recover the marginal distribution over the non-neutral barcodes relative
+fitness values, we can numerically integrate out all nuisance parameters, i.e.,
 $$
@@
 on the neutral lineage frequency values, i.e., $\pi(\underline{\bar{s}}_T \mid
-\underline{\underline{F}}^N)$. We point the reader to `[SI section]` for the
+\underline{\underline{F}}^N)$. We point the reader to @sec-bayes_def for the
 full mathematical details on this factorization. For our purpose here, it
@@
 Written in this form, @eq-split_posterior captures the three sources of
-uncertainty listed in @sec-fitness_model in each term. Starting from
-right to left, the first term on the right-hand side of @eq-split_posterior
-accounts for the uncertainty when inferring the frequency values given the
-barcode reads. The second term accounts for the uncertainty in the values of
-the mean population fitness at different time points. The last term accounts for
-the uncertainty in the parameter we care about---the mutants' relative
-fitnesses. We invite the reader to check the supplementary material, where we
-work through these terms to build the posterior distribution to sample
-from. We refer the reader to `[ref SI section]` for an extended description of
-the model with specific functional forms for each term on the left-hand side of
-@eq-split_posterior as well as `[ref SI on hierarchical]` where the model is
-extended to account for multiple experimental replicates.
+uncertainty listed in @sec-fitness_model in each term. Starting from right to
+left, the first term on the right-hand side of @eq-split_posterior accounts for
+the uncertainty when inferring the frequency values given the barcode reads. The
+second term accounts for the uncertainty in the values of the mean population
+fitness at different time points. The last term accounts for the uncertainty in
+the parameter we care about---the mutants' relative fitnesses. We refer the
+reader to @sec-bayes_def for an extended description of the model with specific
+functional forms for each term on the left-hand side of @eq-split_posterior as
+well as the extension of the model to account for multiple experimental
+replicates or hierarchical genotypes.
 
@@
 such as Hamiltonian Montecarlo can numerically perform this integration
-[@betancourt2017], the dimensionality of the problem in @eq-split_posterior makes
-an MCMC-based approach prohibitively slow.
+[@betancourt2017], the dimensionality of the problem in @eq-split_posterior
+makes an MCMC-based approach prohibitively slow.
 

### 2023-09-26 doc/_single_results.qmd
small enhancements to section

@@
 To assess the inference pipeline performance, we applied it to a simulated
-dataset with known ground truth relative fitness values (See `[ref to SI]` for
+dataset with known ground truth relative fitness values (See @sec-logistic for
 details on simulation). @fig-02(A) shows the structure of the synthetic dataset.
 The majority of barcodes of interest (faint color lines) are adaptive compared
-to the neutral barcodes ($s^{(m)} > 0$). This experimental condition is required
-to reliably measure the relative fitness since all barcodes of interest occupy a
-small fraction of the population at the beginning of the experiment `[ref SI
-section on detrimental mutants]`. Although the barcode frequency trajectories
-look relatively smooth, our fitness model requires the computation of the log
-frequency ratio between adjacent time points as derived in @eq-logfreq.
-@fig-02(B) shows such data transformation where we can better appreciate the
-observational noise input into our statistical model. This noise is evident for
-the darker lines representing the neutral barcodes since all of these lineages
-are assumed to be identically distributed under this log frequency ratio
-transformation.
+to the neutral barcodes ($s^{(m)} > 0$). Although the barcode frequency
+trajectories look relatively smooth, our fitness model requires the computation
+of the log frequency ratio between adjacent time points as derived in
+@eq-logfreq. @fig-02(B) shows such data transformation where we can better
+appreciate the observational noise input into our statistical model. This noise
+is evident for the darker lines representing the neutral barcodes since all of
+these lineages are assumed to be identically distributed.
 
@@
 the resulting parameter values from the ADVI inference to generate possible
-datasets in agreement with the inferred values. @fig-02(C) shows these results
-for all neutral lineages (upper left corner plot) and a few representative
-non-neutral barcodes. The different color shades represent the 95\%, 68\%, and
-5\% credible regions, i.e., the regions where we expect to find the data with
-the corresponding probability---or in terms of our parameter, the $X\%$ credible
-region is the interval where we expect the true parameter to lie with $X\%$
-probability.
+datasets in agreement with the inferred values (See @sec-ppc for further details
+on these computations). @fig-02(C) shows these results for all neutral lineages
+(upper left corner plot) and a few representative non-neutral barcodes. The
+different color shades represent the 95\%, 68\%, and 5\% credible regions, i.e.,
+the regions where we expect to find the data with the corresponding
+probability---or in terms of our parameter, the $X\%$ credible region is the
+interval where we expect the true parameter to lie with $X\%$ probability.
 
@@
 barcodes. These transformed data are much more noisy than the seemingly smooth
-frequency trajectories. (B) Examples of the posterior predictive checks for all
+frequency trajectories. (C) Examples of the posterior predictive checks for all
 neutral lineages (upper left panel) and a subset of representative mutant
@@
 the data. The reported errors above the plot represent the 68\% credible region
-on the mutant relative fitness marginal distribution. (C) Comparison between the
+on the mutant relative fitness marginal distribution. (D) Comparison between the
 ground truth fitness value from the logistic-growth simulation and the inferred
-fitness value. Gray error bars represent the 68\% posterior credible region. (D)
-The empirical cumulative distribution function (ECDF) for the absolute z-score
-value of the ground truth parameter value within the inferred fitness posterior
-distribution.](./figs/fig02){#fig-02}
+fitness value. Gray error bars represent the 68\% posterior credible region for
+the relative fitness values. (E) The empirical cumulative distribution function
+(ECDF) for the absolute z-score value of the ground truth parameter value within
+the inferred fitness posterior distribution.](./figs/fig02){#fig-02}

### 2023-09-26 doc/_multienv_results.qmd
small changes to figure caption

@@
 ![**Multi-environment fitness inference.** (A) Schematic of the simulated
-experimental design where growth-dilution cycles were performed in different
-environments in a cycle. (B) log frequency rations between adjacent time points.
-Darker lines represent the neutral barcodes. The colors in the background demark
-the corresponding environment, matching colors in (A). (C-E) Comparison between
-the ground truth fitness value from the logistic-growth simulation and the
-inferred fitness value for each environment. Gray error bars represent the 68\%
-posterior credible region. (F) The empirical cumulative distribution function
-(ECDF) for the absolute z-score value of the ground truth parameter value within
-the inferred fitness posterior distribution for all fitness values (black line)
-and each environment individually (color lines). (G) Examples of the posterior
-predictive checks for all neutral lineages (upper left panel) and a subset of
-representative mutant lineages. Shaded regions represent the 95\%, 68\%, and 5\%
-credible regions for the data. The reported errors above the plot represent the
-68\% credible region on the mutant relative fitness marginal distribution.
-Background colors match those of (A).](./figs/fig06){#fig-06}
+experimental design where growth-dilution cycles are performed into different
+environments for each cycle. (B) log frequency rations between adjacent time
+points. Darker lines represent the neutral barcodes. The colors in the
+background demark the corresponding environment, matching colors in (A).
+Environment 2 is set to have, on average, half the growth rate of environment 1.
+Likewise, environment 3 is set to have, on average, twice the growth rate of
+environment 1. (C-E) Comparison between the ground truth fitness value from the
+logistic-growth simulation and the inferred fitness value for each environment.
+Gray error bars represent the 68\% posterior credible region. (F) The empirical
+cumulative distribution function (ECDF) for the absolute z-score value of the
+ground truth parameter value within the inferred fitness posterior distribution
+for all fitness values (black line) and each environment individually (color
+lines). (G) Examples of the posterior predictive checks for all neutral lineages
+(upper left panel) and a subset of representative mutant lineages. Shaded
+regions surrounding the data represent the 95\%, 68\%, and 5\% credible regions
+for the data. The reported errors above the plot represent the 68\% credible
+region on the mutant relative fitness marginal distribution. Background colors
+match those of (A).](./figs/fig06){#fig-06}

### 2023-09-26 doc/_hierarchical_replicates_results.qmd
small changes to section

@@
 capture the parametric inference we are interested in---for this inference
-problem, we have a hyper-fitness value $\theta^{(m)}$ for each mutant. Second,
-each experiment draws randomly from the distribution of this hyper-parameter,
-allowing for subtle variability between experiments to be accounted for---in the
-present inference pipeline, each experimental replicate gets assigned a *local*
-fitness value $s^{(m)}_i$, where the extra sub-index indicates the $i$-th
-experimental replicate. For our hierarchical model, these local fitness values
-interact via the common dependency on the global hyper-fitness value. As
-schematized in @fig-03(A), this interaction runs both ways. First, the data from
-one experimental replicate ($\underline{\underline{R}}^M_k$ in the diagram)
-informs all local fitness values via the global hyper-fitness (upper panel in
-@fig-03(A)). Second, the local fitness value is informed by the data from all
-experimental replicates via the same global hyper-fitness parameter (lower panel
-in @fig-03(A)).
+problem, we have a hyper-fitness value $\theta^{(m)}$ for each non-neutral
+barcode. Second, each experiment draws randomly from the distribution of this
+hyper-parameter, allowing for subtle variability between experiments to be
+accounted for---in the present inference pipeline, each experimental replicate
+gets assigned a *local* fitness value $s^{(m)}_i$, where the extra sub-index
+indicates the $i$-th experimental replicate. For our hierarchical model, these
+local fitness values interact via the common dependency on the global
+hyper-fitness value. As schematized in @fig-03(A), this interaction runs both
+ways. First, the data from one experimental replicate
+($\underline{\underline{R}}^M_k$ in the diagram) informs all local fitness
+values via the global hyper-fitness (upper panel in @fig-03(A)). Second, the
+local fitness value is informed by the data from all experimental replicates via
+the same global hyper-fitness parameter (lower panel in @fig-03(A)).
 
@@
 truth hyper-fitness value $\theta^{(m)}$ for each barcode. We sampled a
-variation from this hyper-fitness value for each experimental replicate $s^{(m,
-i)}$ to capture experimental batch effects. @fig-03(D) shows the relationship
-between hyper-fitness and replicate fitness values for this simulation. The
-posterior predictive checks examples in @fig-03(E) show that the hierarchical
-model can correctly fit the data for each experimental replicate. Furthermore,
-@fig-03(F-G) show a high correlation between the ground truth and the inferred
-fitness values. The ECDFs shown in @fig-03(H-I) reveal that for $\approx 75\%$
-of the non-neutral barcodes, the ground truth hyper-fitness values fall within
-one standard deviation from the mean value in the posterior distributions.
+variation from this hyper-fitness value for each experimental replicate
+$s^{(m)}_i$ to capture experimental batch effects. @fig-03(D) shows the
+relationship between hyper-fitness and replicate fitness values for this
+simulation. The spread around the identity line represents the expected
+batch-to-batch variation. The posterior predictive checks examples in @fig-03(E)
+show that the hierarchical model can correctly fit the data for each
+experimental replicate. Furthermore, @fig-03(F-G) show a high correlation
+between the ground truth and the inferred fitness values. The empirical
+cumulative distribution functions shown in @fig-03(H-I) reveal that for $\approx
+75\%$ of the non-neutral barcodes, the ground truth hyper-fitness values fall
+within one standard deviation from the mean value in the posterior
+distributions.
 
 ![**Hierarchical model on experimental replicates**. (A) Schematic depiction of
-the interactions between local fitness values $\underline{s}_l^M$ through the
+the interactions between local fitness values $\underline{s}_k^M$ through the
 global hyper-fitness value $\underline{\theta}^M$ for $K$ hypothetical
@@
 informs all local fitness values via the hyper-fitness parameter. The lower
-panel shows the reverse, where all datasets inform the local fitness value.
-(B-C) Simulated replicate datasets with 900 barcodes of interest and 100 neutral
-lineages. (D) Comparison between the simulation ground truth hyper-fitness and
-each replicate ground truth fitness. The scatter between parameters captures
-experimental batch effects. (E) Examples of the posterior predictive checks for
-all neutral lineages (upper left panels) and a subset of representative mutant
-lineages. Shaded regions from light to dark represent the 95\%, 68\%, and 5\%
-credible regions. (F-G) Comparison between the simulation's ground truth
-hyper-fitness (F) and replicate fitness (G) values and the inferred parameters.
-Gray error bars represent the 68\% posterior credible region. (H-I) The
-empirical cumulative distribution function (ECDF) for the absolute z-score value
-of the ground truth parameter value within the inferred hyper-fitness posterior
-distribution (H) and replicate fitness (I).](./figs/fig03){#fig-03}
+panel shows the reverse, where all other datasets inform the local fitness
+value. (B-C) Simulated replicate datasets with 900 barcodes of interest and 100
+neutral lineages. (D) Comparison between the simulation ground truth
+hyper-fitness and each replicate ground truth fitness. The scatter between
+parameters captures experimental batch effects. (E) Examples of the posterior
+predictive checks for all neutral lineages (upper left panels) and a subset of
+representative mutant lineages. Shaded regions from light to dark represent the
+95\%, 68\%, and 5\% credible regions. (F-G) Comparison between the simulation's
+ground truth hyper-fitness (F) and replicate fitness (G) values and the inferred
+parameters. Gray error bars represent the 68\% posterior credible region. (H-I)
+The empirical cumulative distribution function (ECDF) for the absolute z-score
+value of the ground truth parameter value within the inferred hyper-fitness
+posterior distribution (H) and replicate fitness (I).](./figs/fig03){#fig-03}
 
@@
 No error bars are shown for these, as it is inappropriate to compute one with
-two data points. (D) Empirical cumulative distribution function (ECDF) of the
-absolute difference between the inferred mean and the ground truth
-hyper-fitness.](./figs/fig04){#fig-04}
+two data points per non-neutral barcode. (D) Empirical cumulative distribution
+function (ECDF) of the absolute difference between the inferred mean and the
+ground truth hyper-fitness.](./figs/fig04){#fig-04}

### 2023-09-26 doc/_hierarchical_genotypes_results.qmd
minimal changes to section

@@
 all barcodes mapping to a single genotype. As schematized in @fig-05(A), a small
-modification allows us to map the structure of our original model to that of a
-hierarchical model with a fitness hyperparameter vector $\underline{\theta}^G$,
-where $G$ is the number of genotypes in the dataset.
+modification of the base model allows us to map the structure of our original
+model to that of a hierarchical model with a fitness hyperparameter vector
+$\underline{\theta}^G$, where $G$ is the number of genotypes in the dataset.
 
@@
 inference using this hierarchical model, as well as the two extreme cases of
-ignoring the connection between the barcodes belonging to the same genotype
----equivalent to performing inference using the model presented in @fig-02(A)
-over the barcodes---or pooling the data of all barcodes belonging to the same
-genotype into a single count---equivalent to performing inference using the
-model presented in @fig-02(A) over the pooled barcodes. @fig-05(C-D) shows the
-comparison between the simulation ground truth and the inferred values for these
-three cases. Not only do the hierarchical model results show higher degrees of
-correlation with the ground truth, but the error bars (representing the 64\%
-credible regions) are smaller, meaning that the uncertainty in the estimate
-decreases when using the hierarchical model. The improvement in the prediction
-can be seen in @fig-05(F) where the empirical cumulative distribution function
-of the absolute difference between the mean inferred value and the simulation
-ground truth is shown for all three inference models. The hierarchical model's
-curve ascends more rapidly, showing that, in general, the inferred values are
-closer to the ground truth. For completeness, @fig-05(G) shows some examples of
-how the hierarchical model can capture the raw log-frequency count observations.
+ignoring the connection between the barcodes belonging to the same
+genotype---equivalent to performing inference using the model presented in
+@fig-02(A) over the barcodes---or pooling the data of all barcodes belonging to
+the same genotype into a single count---equivalent to performing inference using
+the model presented in @fig-02(A) over the pooled barcodes. @fig-05(C-D) shows
+the comparison between the simulation ground truth and the inferred values for
+these three cases. Not only do the hierarchical model results show higher
+degrees of correlation with the ground truth, but the error bars (representing
+the 68\% credible regions) are smaller, meaning that the uncertainty in the
+estimate of the parameter we care about decreases when using the hierarchical
+model. The improvement in the prediction can be seen in @fig-05(F) where the
+empirical cumulative distribution function of the absolute difference between
+the mean inferred value and the simulation ground truth is shown for all three
+inference models. The hierarchical model's curve ascends more rapidly, showing
+that, in general, the inferred values are closer to the ground truth. For
+completeness, @fig-05(G) shows some examples of how the hierarchical model can
+capture the raw log-frequency count observations.
 

### 2023-09-26 doc/_bayes_priors.qmd
finish prior definition section

@@
 information we have about the inference in question before acquiring any data.
-This is the "ideal" use of priors that, whenever possible, should be used. On
-the other hand, for most practitioners of Bayesian statistics in the age of big
-data, the definition of prior becomes a tool to ensure the convergence of
-sampling algorithms such as MCMC [@gelman2017]. However, for our particular
-problem, although we deal with large amounts of data (inferences can be made for
-> 10K barcodes over multiple time points, resulting in > 100K parameters), each
-barcode has very little data, as they are measured only once per time point over
-< 10 times. Furthermore, understanding the sources of noise having to do with
-culturing conditions, DNA extraction, library preparation, and so on and
-encoding them into reasonable prior distributions becomes incredibly
-challenging.
+This is the "ideal" use of priors that, whenever possible, should be
+implemented. On the other hand, for most practitioners of Bayesian statistics in
+the age of big data, the definition of prior becomes a tool to ensure the
+convergence of sampling algorithms such as MCMC [@gelman2017]. However, for our
+particular problem, although we deal with large amounts of data (inferences can
+be made for > 10K barcodes over multiple time points, resulting in > 100K
+parameters), each barcode has very little data, as they are measured only once
+per time point over < 10 growth-dilution cycles. Furthermore, it is incredibly
+challenging to understand the noise sources related to culturing conditions, DNA
+extraction, library preparation, etc., and encode them into reasonable
+prior distributions.
 
@@
 With this in mind, we leave it to the reader to judge the selection of priors.
-Furthermore, the present package `BayesFitness.jl` is written so that users can
-experiment with different prior selection criteria that fit their needs. We
-strongly advocate that statistics should not be done in a black-box fit-all tool
-mindset but rather as a formal way to encode the assumptions behind the
-analysis, subject to constructive criticism.
-
-With this philosophical baggage behind us, let us now focus on how the priors
-used for this work were selected.
+Furthermore, the software package associated with this work, `BayesFitness.jl`,
+is written so that users can experiment with different prior selection criteria
+that fit their needs. We strongly advocate that statistics should not be done in
+a black-box fit-all tool mindset but rather as a formal way to encode the
+assumptions behind the analysis, subject to constructive criticism. With this
+philosophical baggage behind us, let us now focus on how the priors used for
+this work were selected.
 
@@
     \operatorname{Var}\left(\frac{r_{t+1}^{(n)}}{r_t^{(n)}}\right)
-}.
+},
 $${#eq-prior_meanfit_var_data}
-This same value was utilized for the
+where $\operatorname{Var}$ is the sample variance. This same value was utilized
+for the mean of the non-neutral barcode variance $\mu_{\sigma^{(m)}}$. While we
+assigne the corresponding variances to be $\sigma_{\sigma_t} =
+\sigma_{\sigma^{(m)}} = 1$.

### 2023-09-28 doc/_variational_inference_primer.qmd
fix high-dimensional integral

@@
 \pi(\underline{x}) = 
-\int\cdots\int d\underline{\theta}^K \pi(\underline{x}, \underline{\theta}) = 
-\int\cdots\int d\underline{\theta}^K \pi(\underline{x} \mid \underline{\theta})
+\int\cdots\int d^K\underline{\theta}\; \pi(\underline{x}, \underline{\theta}) = 
+\int\cdots\int d^K\underline{\theta}\; 
+\pi(\underline{x} \mid \underline{\theta})
 \pi(\underline{\theta}),
@@
 \right) \equiv 
-\int \cdots \int d\underline{\theta}^K
+\int \cdots \int d^K\underline{\theta}\;
 q_\phi(\underline{\theta})
@@
 \right) = 
-\int d\underline{\theta}^K q_\phi(\underline{\theta})
+\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
 \ln q_\phi(\underline{\theta}) -
-\int d\underline{\theta}^K q_\phi(\underline{\theta})
+\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
 \ln \pi(\underline{\theta} \mid \underline{x}),
@@
 where, for convenience, we write a single integration sign
-($d\underline{\theta}^K$ still represents a multi-dimensional differential). For
+($d^K\underline{\theta}\;$ still represents a multi-dimensional differential). For
 the second term in @eq-vi_step01, we can substitute the term inside the
@@
 \right) &= 
-\int d\underline{\theta}^K q_\phi(\underline{\theta})
+\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
 \ln q_\phi(\underline{\theta}) \\
-&- \int d\underline{\theta}^K q_\phi(\underline{\theta})
+&- \int d^K\underline{\theta}\; q_\phi(\underline{\theta})
 \ln \left( 
@@
 \right) &= 
-\int d\underline{\theta}^K q_\phi(\underline{\theta})
+\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
 \ln q_\phi(\underline{\theta}) \\
-&-\int d\underline{\theta}^K q_\phi(\underline{\theta})
+&-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
 \ln \pi(\underline{x} \mid \underline{\theta}) \\
-&-\int d\underline{\theta}^K q_\phi(\underline{\theta})
+&-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
 \ln \pi(\underline{\theta}) \\
-&+\int d\underline{\theta}^K q_\phi(\underline{\theta})
+&+\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
 \ln \pi(\underline{x}).
@@
 \right) &= 
-\int d\underline{\theta}^K q_\phi(\underline{\theta})
+\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
 \ln \frac{
@@
     } \\
-&-\int d\underline{\theta}^K q_\phi(\underline{\theta})
+&-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
 \ln \pi(\underline{x} \mid \underline{\theta}) \\
 &+ \ln \pi(\underline{x}) 
-\int d\underline{\theta}^K q_\phi(\underline{\theta}),
+\int d^K\underline{\theta}\; q_\phi(\underline{\theta}),
 \end{aligned}
@@
 $$
-\int d\underline{\theta}^K q_\phi(\underline{\theta}) = 1.
+\int d^K\underline{\theta}\; q_\phi(\underline{\theta}) = 1.
 $${#eq-vi_q_norm}
@@
 $$
-\int d\underline{\theta}^K q_\phi(\underline{\theta})
+\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
 f(\underline{\theta}) = \left\langle 

### 2023-10-02 doc/_intro.qmd
include Madhav's comments

@@
 
-We present a Bayesian inference pipeline to properly quantify the uncertainty
-about the parametric information we can extract from high-throughput competitive
-fitness assays. In these assays, the fitness of an ensemble of genotypes is
-determined relative to a reference genotype [@kinsler2020; @ascensao2023].
-@fig-01(A) shows a schematic of the experimental procedure in which an initial
-pool of barcoded strains are mixed with a reference strain and inoculated into
-fresh media. After some time---usually, enough time for the culture to
-saturate---an aliquot is transferred to fresh media, while the remaining culture
-is used for DNA sequencing of the lineage barcodes. The time-series information
-of the relative abundance of each lineage, i.e., the barcode frequency depicted
-in @fig-01(B), is used to infer the relative fitness---the growth advantage on a
-per-cycle basis---for each lineage with respect to the reference strain. The
-proposed statistical model accounts for multiple sources of uncertainty when
-inferring the lineages' relative fitness values (see @sec-experiment for details
-on sources of uncertainty accounted for by the model). Furthermore, minor
-changes to the core statistical model allow us to account for relevant
-experimental variations of these competition assays. More specifically, in
-@sec-multienv, we present a variation of the statistical model to infer fitness
-on growth dilution cycles in multiple environments with proper error
-propagation. Furthermore, as described in @sec-replicates, our statistical model
-can account for batch-to-batch differences when jointly analyzing multiple
-experimental replicates using a Bayesian hierarchical model. Finally, a variant
-of these hierarchical models, presented in @sec-genotypes, can account for
-variability within multiple barcodes mapping to equivalent genotypes within the
-same experiment.
+We present a Bayesian inference pipeline to quantify the uncertainty about the
+parametric information we can extract from high-throughput competitive fitness
+assays given a model of the data generation process and experimental data. In
+these assays, the fitness of an ensemble of genotypes is determined relative to
+a reference genotype [@kinsler2020; @ascensao2023]. @fig-01(A) shows a schematic
+of the experimental procedure in which an initial pool of barcoded strains are
+mixed with a reference strain and inoculated into fresh media. After some
+time---usually, enough time for the culture to saturate---an aliquot is
+transferred to fresh media, while the remaining culture is used for DNA
+sequencing of the lineage barcodes. The time-series information of the relative
+abundance of each lineage, i.e., the barcode frequency depicted in @fig-01(B),
+is used to infer the relative fitness---the growth advantage on a per-cycle
+basis---for each lineage with respect to the reference strain. The proposed
+statistical model accounts for multiple sources of uncertainty when inferring
+the lineages' relative fitness values (see @sec-experiment for details on
+sources of uncertainty accounted for by the model). Furthermore, minor changes
+to the core statistical model allow us to account for relevant experimental
+variations of these competition assays. More specifically, in @sec-multienv, we
+present a variation of the statistical model to infer fitness on growth dilution
+cycles in multiple environments with proper error propagation. Furthermore, as
+described in @sec-replicates, our statistical model can account for
+batch-to-batch differences when jointly analyzing multiple experimental
+replicates using a Bayesian hierarchical model. Finally, a variant of these
+hierarchical models, presented in @sec-genotypes, can account for variability
+within multiple barcodes mapping to equivalent genotypes within the same
+experiment.
 
 For all the model variations presented in this paper, we benchmark the ability
-of our pipeline to infer relative fitness parameters against synthetic data with
-known ground truth. A `Julia` package accompanies the present method to readily
-implement the inference pipeline with state-of-the-art scientific computing
-software.
+of our pipeline to infer relative fitness parameters against synthetic data
+generated from logistic growth simulations with added random noise. A `Julia`
+package accompanies the present method to readily implement the inference
+pipeline with state-of-the-art scientific computing software.

### 2023-10-02 doc/_experiment.qmd
add Madhav's comments

@@
 to the untagged ancestor strain that initially dominates the culture. These
-neutral lineages simplify the inference problem since the fitness of all
-relevant barcodes is quantified with respect to these barcodes.
+spiked-in neutral lineages simplify the inference problem since the fitness
+metric of all relevant barcodes is quantified with respect to these
+barcodes---thus referred to as *relative fitness*.
 

### 2023-10-02 doc/_fitness_model.qmd
implement Madhav's suggestions

@@
 
-Empirically, the barcode relative frequency trajectories follow an exponential
-function of the form [@levy2015; @kinsler2020; @ascensao2023]
+Empirically, each barcode frequency trajectory follows an exponential function
+of the form [@levy2015; @kinsler2020; @ascensao2023]
 $$
@@
 
-Our ultimate objective is to infer the relative fitness $s^{(m)}$ for each of
-the $M$ relevant barcodes in the experiment. To do so, we account for the three
-primary sources of uncertainty in our model:
+Our ultimate objective is to infer the relative fitness $s^{(b)}$ for each of
+the $M$ relevant barcodes in the experiment---hereafter referred to as $s^{(m)}$
+to distinguish from the general $s^{(b)}$ and the neutral lineages $s^{(n)}$
+relative fitness. To do so, we account for the three primary sources of
+uncertainty in our model:
 

### 2023-10-04 doc/_bayes_freq.qmd
small fixes
several changes, including Griffin's suggestions

@@
 
-## Defining the Bayesian model
+## Defining the Bayesian model {#sec-bayes_def}
+
+In the main text, we specify the inference problem we must solve as being of the
+form
+$$
+\pi(
+    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
+    \underline{\underline{R}}
+) \propto
+\pi(
+    \underline{\underline{R}} \mid
+    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
+)
+\pi(
+    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
+).
+$${#eq-full_inference_SI}
+Here, we briefly define the missing nuisance parameters. Let 
+$$
+\underline{\bar{s}}_T = (\bar{s}_1, \bar{s}_2, \ldots, \bar{s}_{T-1})^\dagger,
+$${#eq-pop_fitness_vec}
+be the vector containing the $T-1$ population mean fitness we compute from the
+$T$ time points where measurements were taken. We have $T-1$ since the value
+of any $\bar{s}_t$ requires cycle numbers $t$ and $t+1$. Furthermore, let the
+matrix $\underline{\underline{F}}$ be a $T \times B$ matrix containing all
+frequency values. As with @eq-R-mat_split in the main text, we can split
+$\underline{\underline{F}}$ into two matrices of the form
+$$
+\underline{\underline{F}} = \left[ 
+\underline{\underline{F}}^N \; \underline{\underline{F}}^M
+\right],
+$${#eq-F-mat_split}
+to separate the corresponding neutral and non-neutral barcode frequencies.
 

### 2023-10-04 doc/_bayesian_inference.qmd
several changes, including Griffin's suggestions

@@
 $${#eq-bayes_obj}
-where hereafter $\pi(\cdot)$ defines a probability density function, unless
-otherwise stated. When defining our statistical model, we need not to focus on
-the denominator on the right-hand side of @eq-bayes_obj. Thus, we can write
+where hereafter $\pi(\cdot)$ defines a probability density function. When
+defining our statistical model, we need not to focus on the denominator on the
+right-hand side of @eq-bayes_obj. Thus, we can write
 $$
@@
 the mutant relative fitness. Therefore, we must include these nuisance
-parameters as part of our inference problem. To include these nuisance
-parameters, let 
-$$
-\underline{\bar{s}}_T = (\bar{s}_1, \bar{s}_2, \ldots, \bar{s}_{T-1})^\dagger,
-$${#eq-pop_fitness_vec}
-be the vector containing the $T-1$ population mean fitness we compute from the
-$T$ time points where measurements were taken. We have $T-1$ since the value
-of any $\bar{s}_t$ requires cycle numbers $t$ and $t+1$. Furthermore, let the
-matrix $\underline{\underline{F}}$ be a $T \times B$ matrix containing all
-frequency values. As with @eq-R-mat_split, we can split
-$\underline{\underline{F}}$ into two matrices of the form
-$$
-\underline{\underline{F}} = \left[ 
-\underline{\underline{F}}^N \; \underline{\underline{F}}^M
-\right],
-$${#eq-F-mat_split}
-to separate the corresponding neutral and non-neutral barcode frequencies. With
-these nuisance variables in hand, the full inference problem we must solve takes
-the form
+parameters as part of our inference problem. We direct the reader to the
+supplementary materials for the exact definitions of these parameters. Here, it
+suffices to say that the inference problem must include the vector
+$\underline{\bar{s}}_T$ of all population mean fitness values and the matrix
+$\underline{\underline{F}}$ of all barcode frequencies within the sequencing
+data. With these nuisance variables in hand, the full inference problem we must
+solve takes the form
 $$
@@
 
-To overcome this computational limitation, in this work, we rely on the recent
-development of the automatic differentiation variational inference algorithm
-(ADVI) [@kucukelbir2016]. Briefly, when performing ADVI, our target posterior
+To overcome this computational limitation, we rely on the recent development of
+the automatic differentiation variational inference algorithm (ADVI)
+[@kucukelbir2016]. Briefly, when performing ADVI, our target posterior
 distribution $\pi(\theta \mid \underline{\underline{R}})$, where $\theta =

### 2023-10-04 doc/_experiment.qmd
several changes, including Griffin's suggestions

@@
 assays for *S. cerevisiae* and to @ascensao2023 for equivalent assays for *E.
-coli*. @fig-01(B) presents a typical barcode trajectory where the darker barcode
+coli*. @fig-01(B) presents a typical barcode trajectory where the black
 trajectories represent the so-called *neutral lineages*, genetically equivalent

### 2023-10-04 doc/_hierarchical_replicates_results.qmd
several changes, including Griffin's suggestions

@@
 can treat each measurement as entirely independent, losing the power to utilize
-multiple measurements when trying to learn a single parameter. This can hurt the
-inference since, in principle, the value of our parameter of interest should not
-depend on the particular experimental replicate in question. However, this
-approach does not allow us to properly "combine" the uncertainties in both
-experiments when performing the inference. On the other hand, we can pool all
-data together and treat our different experiments as a single measurement with
-higher coverage. This loses the subtle differences from biotic and abiotic batch
-effects, effectively halving the data that goes into our inference problem.
+multiple measurements when trying to learn a single parameter. This can
+negatively impact the inference since, in principle, the value of our parameter
+of interest should not depend on the particular experimental replicate in
+question. However, this approach does not allow us to properly "combine" the
+uncertainties in both experiments when performing the inference. On the other
+hand, we can pool all data together and treat our different experiments as a
+single measurement with higher coverage. This loses the subtle differences due
+to biotic and abiotic batch effects, effectively halving the data that goes into
+our inference problem.
 

### 2023-10-04 doc/_single_results.qmd
several changes, including Griffin's suggestions

@@
 probability---or in terms of our parameter, the $X\%$ credible region is the
-interval where we expect the true parameter to lie with $X\%$ probability.
+interval where we expect the true parameter value to lie with $X\%$ probability.
 
@@
 would contain the true population parameter with a certain probability if the
-experiment was repeated many times. The confidence level does not represent the
-probability that the interval contains the true value. According to a specific
-model and prior information, Bayesian credible regions represent the range of
-values that contain the parameter with a certain posterior probability. The
-credible region directly represents the probability that the region contains the
-true value. So, frequentist confidence intervals cannot be interpreted as
+experiment was repeated many times. The confidence interval does not represent
+the probability that the interval contains the true value. According to a
+specific model and prior information, Bayesian credible regions represent the
+range of values that contain the parameter with a certain posterior probability.
+The credible region directly represents the probability that the region contains
+the true value. So, frequentist confidence intervals cannot be interpreted as
 Bayesian credible regions because they have fundamentally different meanings.
 Treating an $X\%$ confidence interval like an $X\%$  credible region is
-fallacious since confidence levels do not represent probabilistic coverage of
+fallacious since confidence intervals do not represent probabilistic coverage of
 the true value like credible regions. The intervals are generated through

### 2023-10-04 doc/_bayesian_inference.qmd
update figure numbering

@@
 evidence lower bound (ELBO) [@kingma2014]---equivalent to minimizing the
-variational free energy [@buckley2017]---is mathematically equivalent to
+variational free energy [@gottwald2020]---is mathematically equivalent to
 performing the optimization prescribed by @eq-vi_objective. We direct the reader

### 2023-10-04 doc/_hierarchical_genotypes_results.qmd
update figure numbering

@@
 of inferring the fitness of each barcode independently or pooling the data of
-all barcodes mapping to a single genotype. As schematized in @fig-05(A), a small
+all barcodes mapping to a single genotype. As schematized in @fig-06(A), a small
 modification of the base model allows us to map the structure of our original
@@
 
-@fig-05(B) shows a single experimental replicate in which 90 genotypes were
+@fig-06(B) shows a single experimental replicate in which 90 genotypes were
 assigned a random number of barcodes (a multinomial distribution with a mean of
 ten barcodes per genotype) for a total of 900 non-neutral barcodes. To assess
-the performance of the hierarchical model proposed in @fig-05(A), we performed
+the performance of the hierarchical model proposed in @fig-06(A), we performed
 inference using this hierarchical model, as well as the two extreme cases of
@@
 the same genotype into a single count---equivalent to performing inference using
-the model presented in @fig-02(A) over the pooled barcodes. @fig-05(C-D) shows
+the model presented in @fig-02(A) over the pooled barcodes. @fig-06(C-D) shows
 the comparison between the simulation ground truth and the inferred values for
@@
 estimate of the parameter we care about decreases when using the hierarchical
-model. The improvement in the prediction can be seen in @fig-05(F) where the
+model. The improvement in the prediction can be seen in @fig-06(F) where the
 empirical cumulative distribution function of the absolute difference between
@@
 that, in general, the inferred values are closer to the ground truth. For
-completeness, @fig-05(G) shows some examples of how the hierarchical model can
+completeness, @fig-06(G) shows some examples of how the hierarchical model can
 capture the raw log-frequency count observations.
@@
 light to dark represent the 95\%, 68\%, and 5\% credible
-regions.](./figs/fig05){#fig-05}
+regions.](./figs/fig06){#fig-06}

### 2023-10-04 doc/_hierarchical_replicates_results.qmd
update figure numbering

@@
 local fitness values interact via the common dependency on the global
-hyper-fitness value. As schematized in @fig-03(A), this interaction runs both
+hyper-fitness value. As schematized in @fig-04(A), this interaction runs both
 ways. First, the data from one experimental replicate
 ($\underline{\underline{R}}^M_k$ in the diagram) informs all local fitness
-values via the global hyper-fitness (upper panel in @fig-03(A)). Second, the
+values via the global hyper-fitness (upper panel in @fig-04(A)). Second, the
 local fitness value is informed by the data from all experimental replicates via
-the same global hyper-fitness parameter (lower panel in @fig-03(A)).
+the same global hyper-fitness parameter (lower panel in @fig-04(A)).
 
 To test the performance of this model, we simulated two experimental replicates
-with 1000 unique barcodes (see @fig-03(B-C)) where we randomly sampled a ground
+with 1000 unique barcodes (see @fig-04(B-C)) where we randomly sampled a ground
 truth hyper-fitness value $\theta^{(m)}$ for each barcode. We sampled a
 variation from this hyper-fitness value for each experimental replicate
-$s^{(m)}_i$ to capture experimental batch effects. @fig-03(D) shows the
+$s^{(m)}_i$ to capture experimental batch effects. @fig-04(D) shows the
 relationship between hyper-fitness and replicate fitness values for this
 simulation. The spread around the identity line represents the expected
-batch-to-batch variation. The posterior predictive checks examples in @fig-03(E)
+batch-to-batch variation. The posterior predictive checks examples in @fig-04(E)
 show that the hierarchical model can correctly fit the data for each
-experimental replicate. Furthermore, @fig-03(F-G) show a high correlation
+experimental replicate. Furthermore, @fig-04(F-G) show a high correlation
 between the ground truth and the inferred fitness values. The empirical
-cumulative distribution functions shown in @fig-03(H-I) reveal that for $\approx
+cumulative distribution functions shown in @fig-04(H-I) reveal that for $\approx
 75\%$ of the non-neutral barcodes, the ground truth hyper-fitness values fall
@@
 value of the ground truth parameter value within the inferred hyper-fitness
-posterior distribution (H) and replicate fitness (I).](./figs/fig03){#fig-03}
+posterior distribution (H) and replicate fitness (I).](./figs/fig04){#fig-04}
 
-As shown in @fig-04, the structure imposed by the hierarchical model schematized
-in @fig-03(A), where we explicitly account for the connection between
+As shown in @fig-05, the structure imposed by the hierarchical model schematized
+in @fig-04(A), where we explicitly account for the connection between
 experimental replicates can improve the quality of the inference. Inferred
 fitness values between experimental replicates exhibit a stronger degree of
-correlation in the hierarchical model (@fig-04(A)) compared to conducting
-inference on each replicate independently (@fig-04(B)). Moreover, when comparing
+correlation in the hierarchical model (@fig-05(A)) compared to conducting
+inference on each replicate independently (@fig-05(B)). Moreover, when comparing
 the inferred hyper-fitness values---the objective parameter when performing
 multiple experimental measurements---the hierarchical model outperforms
-averaging the independent experimental replicates as shown in @fig-04(C) and
+averaging the independent experimental replicates as shown in @fig-05(C) and
 (D).
@@
 function (ECDF) of the absolute difference between the inferred mean and the
-ground truth hyper-fitness.](./figs/fig04){#fig-04}
+ground truth hyper-fitness.](./figs/fig05){#fig-05}

### 2023-10-04 doc/_multienv_results.qmd
update figure numbering

@@
 @eq-bayes_full. This natural error propagation can help us with the experimental
-design schematized in @fig-06(A). Here, rather than performing growth-dilution
+design schematized in @fig-03(A). Here, rather than performing growth-dilution
 cycles in the same environment, the cells are diluted into a different 
@@
 
-To validate the extension of our statistical model to this scenario, @fig-06(B)
+To validate the extension of our statistical model to this scenario, @fig-03(B)
 shows the trajectory of the log frequency ratios between adjacent time points.
@@
 Environment 3 was set to be, on average, twice the average growth rate in
-Environment 1. @fig-06(C-E) show the correspondence between the simulation
+Environment 1. @fig-03(C-E) show the correspondence between the simulation
 ground truth and the inferred fitness values, where the error bars represent the
-68\% credible region. @fig-06(F) summarizes the performance of our inference
+68\% credible region. @fig-03(F) summarizes the performance of our inference
 pipeline by showing the empirical cumulative distribution functions for the
@@
 values fall within one standard deviation of the ground truth. For completeness,
-@fig-06(G) shows the posterior predictive checks for a few example barcodes.
+@fig-03(G) shows the posterior predictive checks for a few example barcodes.
 
@@
 region on the mutant relative fitness marginal distribution. Background colors
-match those of (A).](./figs/fig06){#fig-06}
+match those of (A).](./figs/fig03){#fig-03}

### 2023-10-05 doc/_multienv_results.qmd
update section based on Madhav's comments

@@
 
-One of the benefits of Bayesian methods is the intrinsic error propagation
-embedded in the mathematical framework. For example, the uncertainty on the
-value of the population mean fitness values $\underline{\bar{s}}_T$ is
-propagated to the relative fitness of a particular strain $s^{(b)}$ since we
-defined a joint posterior distribution over all parameters as fully expressed in
-@eq-bayes_full. This natural error propagation can help us with the experimental
-design schematized in @fig-03(A). Here, rather than performing growth-dilution
-cycles in the same environment, the cells are diluted into a different 
-environment. Thus, the uncertainty on the fitness estimate for the previous
-environment must be propagated to that of the next one.
+The fitness model in @eq-fitness relates nuisance parameters such as the
+population mean fitness and the barcode frequencies to the relative fitness
+parameter we want to infer from the data. These dependencies imply that
+uncertainty on the estimates of these nuisance parameters influences the
+inference of the relevant parameters. For example, imagine a scenario where the
+neutral lineages data were incredibly noisy, leading to poor estimates of the
+population mean fitness values $\underline{\bar{s}}_T$. Since the relative
+fitness of any non-neutral barcode $s^{(m)}$ is determined with respect to these
+neutral barcodes, not accounting for the lack of precision in the value of the
+population mean fitness would result in misleading estimates of the accuracy
+with which we determine the value of the parameter we care about. Thus,
+propagating these sources of uncertainty in nuisance parameters is vital to
+generate an unbiased estimate of the relevant information we want to extract
+from the data. One of the benefits of Bayesian methods is the intrinsic error
+propagation embedded in the mathematical framework. For our previous example,
+the uncertainty on the value of the population mean fitness values is propagated
+to the relative fitness of a non-neutral barcode since we defined a joint
+posterior distribution over all parameters as fully expressed in @eq-bayes_full.
 
-To validate the extension of our statistical model to this scenario, @fig-03(B)
-shows the trajectory of the log frequency ratios between adjacent time points.
-The different colored regions correspond to the different environments. For this
-simulation, the growth rate of Environment 2 was set to be, on average, half of
-the average growth rate in Environment 1. Equivalently, the growth rate in
-Environment 3 was set to be, on average, twice the average growth rate in
-Environment 1. @fig-03(C-E) show the correspondence between the simulation
-ground truth and the inferred fitness values, where the error bars represent the
-68\% credible region. @fig-03(F) summarizes the performance of our inference
-pipeline by showing the empirical cumulative distribution functions for the
-absolute value of the ground truth fitness value z-score within the posterior
-distribution. This plot shows that, overall, $\approx 75\%$ of inferred mean
-values fall within one standard deviation of the ground truth. For completeness,
-@fig-03(G) shows the posterior predictive checks for a few example barcodes.
+This natural error propagation can help us with the experimental design
+schematized in @fig-03(A). Here, rather than performing growth-dilution cycles
+in the same environment, the cells are diluted into a different environment.
+Thus, the uncertainty on the fitness estimate for the previous environment must
+be propagated to that of the next one. To validate the extension of our
+statistical model to this scenario, @fig-03(B) shows the trajectory of the log
+frequency ratios between adjacent time points. The different colored regions
+correspond to the different environments. For this simulation, the growth rate
+of Environment 2 was set to be, on average, half of the average growth rate in
+Environment 1. Equivalently, the growth rate in Environment 3 was set to be, on
+average, twice the average growth rate in Environment 1. @fig-03(C-E) show the
+correspondence between the simulation ground truth and the inferred fitness
+values, where the error bars represent the 68\% credible region. @fig-03(F)
+summarizes the performance of our inference pipeline by showing the empirical
+cumulative distribution functions for the absolute value of the ground truth
+fitness value z-score within the posterior distribution. This plot shows that,
+overall, $\approx 75\%$ of inferred mean values fall within one standard
+deviation of the ground truth. For completeness, @fig-03(G) shows the posterior
+predictive checks for a few example barcodes.
 

### 2023-10-05 doc/_hierarchical_replicates_results.qmd
implement Madhav's comments

@@
 gets assigned a *local* fitness value $s^{(m)}_i$, where the extra sub-index
-indicates the $i$-th experimental replicate. For our hierarchical model, these
-local fitness values interact via the common dependency on the global
-hyper-fitness value. As schematized in @fig-04(A), this interaction runs both
+indicates the $i$-th experimental replicate. Conceptually, we can think of the
+local fitness for replicate $i$ as being sampled from a distribution that
+depends on the value of the global hyper-fitness value, i.e., $s^{(m)}_i \sim
+\pi_{\theta^{(m)}}$, where the subindex $\theta^{(m)}$ indicates the
+distribution's parametric dependence on the hyper-fitness value. This way of
+interpreting the connection between the distribution $\pi_{\theta^{(m)}}$ and
+the local fitness implies that a large replicate-to-replicate variability would
+lead to a broad hyper-fitness distribution---implying a large uncertainty when
+determining the parameter that characterizes the overall relative fitness. We
+point the reader to @sec-hierarchical_model for the full definition of the
+hierarchical model used in this section. Importantly, as schematized in
+@fig-04(A), the influence between different experimental replicates runs both
 ways. First, the data from one experimental replicate

### 2023-10-05 doc/_discussion.qmd
small changes to discussion
update the closing paragraph

@@
 
-For this data-heavy field, properly accounting for the uncertainty in quantities
+For this data-heavy field, properly accounting for the uncertainty in parameters
 inferred from experiments is vital to ensure the conclusions drawn are reliable.
@@
 experimental replicates of the same experiment via hierarchical models, and a
-hierarchical model for multiple barcodes mapping to equivalent genotypes.
-
-We validate our analysis pipeline on simulated datasets with known ground truth,
+hierarchical model for multiple barcodes mapping to equivalent genotypes. We
+validate our analysis pipeline on simulated datasets with known ground truth,
 showing that the model fits the data adequately, capturing the ground truth
-parameters within the posterior distribution. Furthermore, `[talk about the
-Kinsler results.]`
+parameters within the posterior distribution.
 
@@
 presented statistical model assumes that relative fitness is solely a constant
-of the environment and the genotype. Therefore, the model is not designed to
-analyze data with time-varying or frequency-dependent fitness values.
+of the environment and the genotype. Future directions of this work could extend
+the fitness model to properly analyze data with time-varying or
+frequency-dependent fitness values.
+
+In total, the statistical model presented in this work and the software package
+accompanying the paper allow for a principled way of quantifying the accuracy
+with which we can extract relevant parametric information from large-scale
+multiplexed fitness competition assays. Furthermore, the implementation of
+Bayesian models and their fitting via automatic differentiation approaches opens
+the gate to extend this type of formal analysis to the data-rich literature in
+experimental evolution and other high-throughput technologies applications.

### 2023-10-09 code/exploratory/data002_logistic_250bc_01env_02rep/advi_meanfield_hierarchical_inference.jl
update all code

@@ -13,6 +13,7 @@ import BayesFitness
 # Import libraries to manipulate data
 import DataFrames as DF
 import CSV
+import MCMCChains
 
 # Import library to save and load native julia objects
 import JLD2
@@ -22,6 +23,8 @@ import Glob
 
 # Import library to perform Bayesian inference
 import Turing
+import MCMCChains
+import DynamicHMC
 
 # Import AutoDiff backend
 using ReverseDiff
@@ -39,6 +42,10 @@ Turing.setadbackend(:reversediff)
 # Allow system to generate cache to speed up computation
 Turing.setrdcache(true)
 
+Random.seed!(42)
+
+##
+
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 # Define ADVI hyerparameters
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
@@ -46,6 +53,8 @@ Turing.setrdcache(true)
 # Define number of samples and steps
 n_samples = 1
 n_steps = 10_000
+
+##
 
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 # Generate output directories
@@ -56,6 +65,8 @@ if !isdir("./output/")
     mkdir("./output/")
 end # if
 
+##
+
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 # Loading the data
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
@@ -64,33 +75,78 @@ println("Loading data...")
 
 # Import data
 data = CSV.read(
-    "$(git_root())/data/logistic_growth/data_003/tidy_data.csv", DF.DataFrame
[…]

### 2023-10-09 code/exploratory/data002_logistic_250bc_01env_02rep/mcmc_joint_hierarchical_inference.jl

@@ -1,0 +1,183 @@
+##
+println("Loading packages...")
+
+# Load project package
+@load_pkg BayesFitUtils
+
+# Import project package
+import BayesFitUtils
+
+# Import library package
+import BayesFitness
+
+# Import libraries to manipulate data
+import DataFrames as DF
+import CSV
+import MCMCChains
+
+# Import library to save and load native julia objects
+import JLD2
+
+# Import library to list files
+import Glob
+
+# Import library to perform Bayesian inference
+import Turing
+import MCMCChains
+import DynamicHMC
+
+# Import AutoDiff backend
+using ReverseDiff
+
+# Import Memoization
+using Memoization
+
+# Impor statistical libraries
+import Random
+import StatsBase
+import Distributions
+
+Random.seed!(42)
+
+##
+
+# Set AutoDiff backend
+Turing.setadbackend(:reversediff)
+# Allow system to generate cache to speed up computation
+Turing.setrdcache(true)
+
+##
+
+# Define sampling hyperparameters
+n_steps = 1
+n_walkers = 4
+
+# Define if plots should be generated
+gen_plots = false
+
+##
+
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Generate output directories
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
+# Generate output directory 
+if !isdir("./output/")
+    mkdir("./output/")
+end # if
+
+##
+
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Loading the data
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
+println("Loading data...")
+
+# Import data
+data = CSV.read(
+    "$(git_root())/data/logistic_growth/data_002/tidy_data.csv", DF.DataFrame
+)
+
+##
+
[…]

### 2023-10-09 code/exploratory/data002_logistic_250bc_01env_02rep/mcmc_joint_inference.jl

@@ -1,0 +1,183 @@
+##
+println("Loading packages...")
+
+# Load project package
+@load_pkg BayesFitUtils
+
+# Import project package
+import BayesFitUtils
+
+# Import library package
+import BayesFitness
+
+# Import libraries to manipulate data
+import DataFrames as DF
+import CSV
+import MCMCChains
+
+# Import library to save and load native julia objects
+import JLD2
+
+# Import library to list files
+import Glob
+
+# Import library to perform Bayesian inference
+import Turing
+import MCMCChains
+import DynamicHMC
+
+# Import AutoDiff backend
+using ReverseDiff
+
+# Import Memoization
+using Memoization
+
+# Impor statistical libraries
+import Random
+import StatsBase
+import Distributions
+
+# Import plotting libraries
+using CairoMakie
+import ColorSchemes
+
+# Activate backend
+CairoMakie.activate!()
+
+# Set PBoC Plotting style
+BayesFitUtils.viz.pboc_makie!()
+##
+
+Random.seed!(42)
+
+##
+
+# Set AutoDiff backend
+Turing.setadbackend(:reversediff)
+# Allow system to generate cache to speed up computation
+Turing.setrdcache(true)
+
+##
+
+# Define sampling hyperparameters
+n_steps = 1000
+n_walkers = 4
+
+# Define if plots should be generated
+gen_plots = false
+
+##
+
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Generate output directories
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
+# Generate output directory 
+if !isdir("./output/")
+    mkdir("./output/")
+end # if
+
+##
+
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Loading the data
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
+println("Loading data...")
+
+# Import data
+df = CSV.read(
[…]

### 2023-10-09 code/exploratory/data002_logistic_250bc_01env_02rep/sim_hierarchical_data002.jl

@@ -1,0 +1,398 @@
+##
+
+println("Loading packages...\n")
+# Load project package
+@load_pkg BayesFitUtils
+
+import Revise
+# Import project package
+import BayesFitUtils
+
+# Import differential equations package
+using DifferentialEquations
+
+# Import package to handle DataFrames
+import DataFrames as DF
+import CSV
+
+# Import basic statistical functions
+import StatsBase
+import Distributions
+import Random
+
+# Load CairoMakie for plotting
+using CairoMakie
+import ColorSchemes
+import Makie
+# Activate backend
+CairoMakie.activate!()
+
+# Set PBoC Plotting style
+BayesFitUtils.viz.pboc_makie!()
+
+##
+
+# Set random seed
+Random.seed!(42)
+
+##
+
+# Define if plots should be generated
+gen_plots = true
+
+##
+
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Define simulated experiment parameters
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
+# Define number of datasets
+n_data = 2
+# Define standard deviation for between-experiment fitness distribution
+# variability
+σ_exp = 0.01
+# Define standard deviation for added Gaussian noise
+σ_lognormal = 0.3
+# Define ancestral strain growth rate
+λ_a = 1.0
+# Define carrying capacity
+κ = 10.0^10
+# Define number of generations
+n_gen = 8
+# Define number of neutral and mutants
+n_neutral, n_mut = [25, 225]
+# Define number of barcodes
+n_bc = n_neutral + n_mut
+
+# Compute initial number of cells
+n_init = κ / (2^(n_gen))
+
+# Define fracton of culture that is ancestor
+frac_anc = 0.93
+# Define fraction of culture that is neutrals
+frac_neutral = 0.02
+# Define fraction of culture that is mutants
+frac_mut = 1 - frac_anc - frac_neutral
+
+##
+
[…]

### 2023-10-09 code/exploratory/data002_logistic_250bc_01env_02rep/viz_inference.jl

@@ -1,0 +1,1160 @@
+##
+println("Loading packages...")
+
+# Load project package
+@load_pkg BayesFitUtils
+
+# Import project package
+import BayesFitUtils
+
+# Import library package
+import BayesFitness
+
+# Import basic math
+import StatsBase
+import Distributions
+import Random
+
+# Import libraries to manipulate data
+import DataFrames as DF
+import CSV
+import MCMCChains
+
+# Import library to list files
+import Glob
+
+# Import library to load files
+import JLD2
+
+# Import plotting libraries
+using CairoMakie
+import ColorSchemes
+
+# Activate backend
+CairoMakie.activate!()
+
+# Set PBoC Plotting style
+BayesFitUtils.viz.pboc_makie!()
+
+Random.seed!(42)
+
+##
+
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Loading data
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
+# Generate output directory 
+if !isdir("./output/")
+    mkdir("./output/")
+end # if
+
+# Generate figure dictionary
+if !isdir("./output/figs/")
+    mkdir("./output/figs/")
+end # if
+
+println("Loading data...")
+
+# Import data
+data = CSV.read(
+    "$(git_root())/data/logistic_growth/data_002/tidy_data.csv", DF.DataFrame
+)
+
+# Define number of experimental repeats
+n_rep = length(unique(data.rep))
+
+##
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Plotting PPC for population mean fitness
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
+# Define file
+files = Glob.glob("./output/chain_popmean_fitness_*")
+
+# Define dictionary with corresponding parameters for variables needed for
+# the posterior predictive checks
+param = Dict(
+    :population_mean_fitness => :s̲ₜ,
[…]

### 2023-10-09 code/exploratory/data004_logistic_zeronoise_250bc_01env_01rep/advi_meanfield_joint_inference.jl

@@ -14,12 +14,6 @@ import BayesFitness
 import DataFrames as DF
 import CSV
 
-# Import library to save and load native julia objects
-import JLD2
-
-# Import library to list files
-import Glob
-
 # Import library to perform Bayesian inference
 import Turing
 
@@ -34,10 +28,14 @@ import Random
 import StatsBase
 import Distributions
 
+Random.seed!(42)
+
 # Set AutoDiff backend
 Turing.setadbackend(:reversediff)
 # Allow system to generate cache to speed up computation
 Turing.setrdcache(true)
+
+##
 
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 # Define ADVI hyerparameters
@@ -46,6 +44,8 @@ Turing.setrdcache(true)
 # Define number of samples and steps
 n_samples = 1
 n_steps = 10_000
+
+##
 
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 # Generate output directories
@@ -56,6 +56,8 @@ if !isdir("./output/")
     mkdir("./output/")
 end # if
 
+##
+
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 # Loading the data
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
@@ -64,33 +66,46 @@ println("Loading data...")
 
 # Import data
 data = CSV.read(
-    "$(git_root())/data/logistic_growth/data_003/tidy_data.csv", DF.DataFrame
+    "$(git_root())/data/logistic_growth/data_004/tidy_data.csv", DF.DataFrame
 )
+
+##
 
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 # Obtain priors on expected errors from neutral measurements
[…]

### 2023-10-09 code/exploratory/data004_logistic_zeronoise_250bc_01env_01rep/mcmc_freq.jl

@@ -1,0 +1,118 @@
+##
+println("Loading packages...")
+
+# Load project package
+@load_pkg BayesFitUtils
+
+# Import package to revise package
+import Revise
+# Import project package
+import BayesFitUtils
+
+# Import library package
+import BayesFitness
+
+# Import libraries to manipulate data
+import DataFrames as DF
+import CSV
+import MCMCChains
+
+# Import library to save and load native julia objects
+import JLD2
+
+# Import library to list files
+import Glob
+
+# Import library to perform Bayesian inference
+import Turing
+import MCMCChains
+import DynamicHMC
+
+# Import AutoDiff backend
+using ReverseDiff
+
+# Import Memoization
+using Memoization
+
+# Impor library to set random seed
+import Random
+
+# Import plotting libraries
+using CairoMakie
+import ColorSchemes
+import PDFmerger
+
+# Activate backend
+CairoMakie.activate!()
+
+# Set PBoC Plotting style
+BayesFitUtils.viz.pboc_makie!()
+
+##
+
+Random.seed!(42)
+
+##
+
+# Set AutoDiff backend
+Turing.setadbackend(:reversediff)
+# Allow system to generate cache to speed up computation
+Turing.setrdcache(true)
+
+##
+
+# Define sampling hyperparameters
+n_steps = 300
+n_walkers = 3
+
+##
+
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Generate output directories
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
+# Define output directory
+outdir = "./output/bc_freq"
+
+# Generate output directory 
+if !isdir("./output/")
+    mkdir("./output/")
+end # if
+
+##
+
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Loading the data
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
[…]

### 2023-10-09 code/exploratory/data004_logistic_zeronoise_250bc_01env_01rep/mcmc_joint_inference.jl

@@ -1,0 +1,158 @@
+##
+println("Loading packages...")
+
+# Load project package
+@load_pkg BayesFitUtils
+
+# Import project package
+import BayesFitUtils
+
+# Import library package
+import BayesFitness
+
+# Import libraries to manipulate data
+import DataFrames as DF
+import CSV
+import MCMCChains
+
+# Import library to save and load native julia objects
+import JLD2
+
+# Import library to list files
+import Glob
+
+# Import library to perform Bayesian inference
+import Turing
+import MCMCChains
+import DynamicHMC
+
+# Import AutoDiff backend
+using ReverseDiff
+
+# Import Memoization
+using Memoization
+
+# Impor statistical libraries
+import Random
+import StatsBase
+import Distributions
+
+# Import plotting libraries
+using CairoMakie
+import ColorSchemes
+
+# Activate backend
+CairoMakie.activate!()
+
+# Set PBoC Plotting style
+BayesFitUtils.viz.pboc_makie!()
+##
+
+Random.seed!(42)
+
+##
+
+# Set AutoDiff backend
+Turing.setadbackend(:reversediff)
+# Allow system to generate cache to speed up computation
+Turing.setrdcache(true)
+
+##
+
+# Define sampling hyperparameters
+n_steps = 1
+n_walkers = 4
+
+# Define if plots should be generated
+gen_plots = false
+
+##
+
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Generate output directories
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
+# Generate output directory 
+if !isdir("./output/")
+    mkdir("./output/")
+end # if
+
+##
+
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Loading the data
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
+println("Loading data...")
+
+# Import data
[…]

### 2023-10-09 code/exploratory/data004_logistic_zeronoise_250bc_01env_01rep/mcmc_popmean_fitness.jl

@@ -1,0 +1,116 @@
+##
+println("Loading packages...")
+
+# Load project package
+@load_pkg BayesFitUtils
+
+# Import package to revise package
+import Revise
+# Import project package
+import BayesFitUtils
+
+# Import library package
+import BayesFitness
+
+# Import libraries to manipulate data
+import DataFrames as DF
+import CSV
+import MCMCChains
+
+# Import library to save and load native julia objects
+import JLD2
+
+# Import library to list files
+import Glob
+
+# Import library to perform Bayesian inference
+import Turing
+import MCMCChains
+import DynamicHMC
+
+# Import AutoDiff backend
+using ReverseDiff
+
+# Import Memoization
+using Memoization
+
+# Impor library to set random seed
+import Random
+
+# Import plotting libraries
+using CairoMakie
+import ColorSchemes
+
+# Activate backend
+CairoMakie.activate!()
+
+# Set PBoC Plotting style
+BayesFitUtils.viz.pboc_makie!()
+
+##
+
+Random.seed!(42)
+
+##
+
+# Set AutoDiff backend
+Turing.setadbackend(:reversediff)
+# Allow system to generate cache to speed up computation
+Turing.setrdcache(true)
+
+##
+
+# Define sampling hyperparameters
+n_steps = 1000
+n_walkers = 4
+
+# Define if plots should be generated
+gen_plots = false
+
+##
+
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Generate output directories
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
+# Generate output directory 
+if !isdir("./output/")
+    mkdir("./output/")
+end # if
+
+##
+
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Loading the data
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
[…]

### 2023-10-09 code/exploratory/data004_logistic_zeronoise_250bc_01env_01rep/sim_zeronoise_data004.jl

@@ -1,0 +1,329 @@
+##
+
+println("Loading packages...\n")
+# Load project package
+@load_pkg BayesFitUtils
+
+import Revise
+# Import project package
+import BayesFitUtils
+
+# Import differential equations package
+using DifferentialEquations
+
+# Import package to handle DataFrames
+import DataFrames as DF
+import CSV
+
+# Import basic statistical functions
+import StatsBase
+import Distributions
+import Random
+
+# Load CairoMakie for plotting
+using CairoMakie
+import ColorSchemes
+import Makie
+# Activate backend
+CairoMakie.activate!()
+
+# Set PBoC Plotting style
+BayesFitUtils.viz.pboc_makie!()
+
+##
+
+# Set random seed
+Random.seed!(42)
+
+##
+
+# Define if plots should be generated
+gen_plots = true
+
+##
+
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Define simulated experiment parameters
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
+# Define standard deviation for added Gaussian noise
+σ_lognormal = 0.0
+# Define ancestral strain growth rate
+λ_a = 1.0
+# Define carrying capacity
+κ = 10.0^10
+# Define number of generations
+n_gen = 8
+# Define number of neutral and mutants
+n_neutral, n_mut = [25, 225]
+# Define number of barcodes
+n_bc = n_neutral + n_mut
+
+# Compute initial number of cells
+n_init = κ / (2^(n_gen))
+
+# Define fracton of culture that is ancestor
+frac_anc = 0.93
+# Define fraction of culture that is neutrals
+frac_neutral = 0.02
+# Define fraction of culture that is mutants
+frac_mut = 1 - frac_anc - frac_neutral
+
+##
+
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
[…]

### 2023-10-09 code/exploratory/data004_logistic_zeronoise_250bc_01env_01rep/viz_inference.jl

@@ -1,0 +1,666 @@
+##
+println("Loading packages...")
+
+# Load project package
+@load_pkg BayesFitUtils
+
+# Import project package
+import BayesFitUtils
+
+# Import library package
+import BayesFitness
+
+# Import basic math
+import StatsBase
+import Distributions
+import Random
+
+# Import libraries to manipulate data
+import DataFrames as DF
+import CSV
+import MCMCChains
+
+# Import library to list files
+import Glob
+
+# Import library to load files
+import JLD2
+
+# Import plotting libraries
+using CairoMakie
+import ColorSchemes
+
+# Activate backend
+CairoMakie.activate!()
+
+# Set PBoC Plotting style
+BayesFitUtils.viz.pboc_makie!()
+
+
+Random.seed!(42)
+
+##
+
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Loading data
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
+# Generate output directory 
+if !isdir("./output/")
+    mkdir("./output/")
+end # if
+
+# Generate figure dictionary
+if !isdir("./output/figs/")
+    mkdir("./output/figs/")
+end # if
+
+println("Loading data...")
+
+# Import data
+data = CSV.read(
+    "$(git_root())/data/logistic_growth/data_004/tidy_data.csv", DF.DataFrame
+)
+
+##
+
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Read chain into memory 
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
+# Define file
+file = first(Glob.glob("./output/chain_joint_fitness_*"))
+
+# Load chain
+ids, chn = values(JLD2.load(file))
+
+# Convert chain to tidy dataframe
+df_chn = DF.DataFrame(chn)
+
+# Remove the string "mut" from mutant names
+mut_num = replace.(ids, "mut" => "")
+
[…]

### 2023-10-09 code/figs/figSIX_gaussian_peak.jl

@@ -68,9 +68,9 @@ fig = Figure(resolution=(400, 300))
 # Add axis for latent space
 ax = Axis3(
     fig[1, 1],
-    xlabel="x₁",
-    ylabel="x₂",
-    zlabel="π(x₁,x₂)",
+    xlabel="θ₁",
+    ylabel="θ₂",
+    zlabel="π(x̲,θ₁,θ₂)",
     xypanelcolor="#E6E6EF",
     xzpanelcolor="#E6E6EF",
     yzpanelcolor="#E6E6EF",
@@ -83,6 +83,9 @@ ax = Axis3(
     yticklabelsvisible=false,
     zticksvisible=false,
     zticklabelsvisible=false,
+    xlabeloffset=10,
+    ylabeloffset=10,
+    zlabeloffset=10,
 )
 
 # Plot surface
@@ -103,6 +106,7 @@ wireframe!(
 )
 
 save("$(git_root())/doc/figs/figSIX_gaussian_peak.pdf", fig)
+save("$(git_root())/doc/figs/figSIX_gaussian_peak.svg", fig)
 save("$(git_root())/doc/figs/figSIX_gaussian_peak.png", fig)
 
 fig

### 2023-10-09 code/processing/data010_logistic_1000bc_01env_04rep/advi_meanfield_hierarchical_inference.jl

@@ -14,9 +14,6 @@ import BayesFitness
 import DataFrames as DF
 import CSV
 
-# Import library to save and load native julia objects
-import JLD2
-
 # Import library to list files
 import Glob
 
@@ -38,6 +35,8 @@ import Distributions
 Turing.setadbackend(:reversediff)
 # Allow system to generate cache to speed up computation
 Turing.setrdcache(true)
+
+Random.seed!(42)
 
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 # Define ADVI hyerparameters
@@ -64,7 +63,7 @@ println("Loading data...")
 
 # Import data
 data = CSV.read(
-    "$(git_root())/data/logistic_growth/data_003/tidy_data.csv", DF.DataFrame
+    "$(git_root())/data/logistic_growth/data_010/tidy_data.csv", DF.DataFrame
 )
 
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
@@ -72,7 +71,7 @@ data = CSV.read(
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 
 # Compute naive priors from neutral strains
-naive_priors = BayesFitness.stats.naive_prior(data; pseudocount=1)
+naive_priors = BayesFitness.stats.naive_prior(data; rep_col=:rep, pseudocount=1)
 
 # Select standard deviation parameters
 s_pop_prior = hcat(
@@ -98,18 +97,20 @@ logλ_prior = hcat(
 
 param = Dict(
     :data => data,
-    :outputname => "./output/advi_meanfield_$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
-    :model => BayesFitness.model.multienv_fitness_normal,
+    :outputname => "./output/advi_meanfield_" *
+                   "$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
+    :model => BayesFitness.model.replicate_fitness_normal,
     :model_kwargs => Dict(
         :s_pop_prior => s_pop_prior,
         :logσ_pop_prior => logσ_pop_prior,
[…]

### 2023-10-09 code/processing/data010_logistic_1000bc_01env_04rep/sim_hierarchical_data010.jl

@@ -1,0 +1,484 @@
+##
+
+println("Loading packages...\n")
+# Load project package
+@load_pkg BayesFitUtils
+
+import Revise
+# Import project package
+import BayesFitUtils
+
+# Import differential equations package
+using DifferentialEquations
+
+# Import package to handle DataFrames
+import DataFrames as DF
+import CSV
+
+# Import basic statistical functions
+import StatsBase
+import Distributions
+import Random
+
+# Load CairoMakie for plotting
+using CairoMakie
+import ColorSchemes
+import Makie
+# Activate backend
+CairoMakie.activate!()
+
+# Set PBoC Plotting style
+BayesFitUtils.viz.theme_makie!()
+
+##
+
+# Set random seed
+Random.seed!(42)
+
+##
+
+# Define if plots should be generated
+gen_plots = true
+
+##
+
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Define simulated experiment parameters
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
+# Define number of datasets
+n_data = 4
+# Define standard deviation for between-experiment fitness distribution
+# variability
+σ_exp = 0.015
+# Define standard deviation for added Gaussian noise
+σ_lognormal = 0.3
+# Define ancestral strain growth rate
+λ_a = 1.0
+# Define carrying capacity
+κ = 10.0^10
+# Define number of generations
+n_gen = 8
+# Define number of neutral and mutants
+n_neutral, n_mut = [100, 900]
+# Define number of barcodes
+n_bc = n_neutral + n_mut
+
+# Compute initial number of cells
+n_init = κ / (2^(n_gen))
+
+# Define fracton of culture that is ancestor
+frac_anc = 0.93
+# Define fraction of culture that is neutrals
+frac_neutral = 0.02
+# Define fraction of culture that is mutants
+frac_mut = 1 - frac_anc - frac_neutral
+
+##
+
[…]

### 2023-10-09 code/processing/data011_logistic_1000bc_01env_08rep/advi_meanfield_hierarchical_inference.jl

@@ -14,9 +14,6 @@ import BayesFitness
 import DataFrames as DF
 import CSV
 
-# Import library to save and load native julia objects
-import JLD2
-
 # Import library to list files
 import Glob
 
@@ -38,6 +35,8 @@ import Distributions
 Turing.setadbackend(:reversediff)
 # Allow system to generate cache to speed up computation
 Turing.setrdcache(true)
+
+Random.seed!(42)
 
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 # Define ADVI hyerparameters
@@ -64,7 +63,7 @@ println("Loading data...")
 
 # Import data
 data = CSV.read(
-    "$(git_root())/data/logistic_growth/data_003/tidy_data.csv", DF.DataFrame
+    "$(git_root())/data/logistic_growth/data_011/tidy_data.csv", DF.DataFrame
 )
 
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
@@ -72,7 +71,7 @@ data = CSV.read(
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 
 # Compute naive priors from neutral strains
-naive_priors = BayesFitness.stats.naive_prior(data; pseudocount=1)
+naive_priors = BayesFitness.stats.naive_prior(data; rep_col=:rep, pseudocount=1)
 
 # Select standard deviation parameters
 s_pop_prior = hcat(
@@ -98,18 +97,20 @@ logλ_prior = hcat(
 
 param = Dict(
     :data => data,
-    :outputname => "./output/advi_meanfield_$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
-    :model => BayesFitness.model.multienv_fitness_normal,
+    :outputname => "./output/advi_meanfield_" *
+                   "$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
+    :model => BayesFitness.model.replicate_fitness_normal,
     :model_kwargs => Dict(
         :s_pop_prior => s_pop_prior,
         :logσ_pop_prior => logσ_pop_prior,
[…]

### 2023-10-09 code/processing/data011_logistic_1000bc_01env_08rep/viz_inference.jl

@@ -1,0 +1,530 @@
+##
+println("Loading packages...")
+
+# Load project package
+@load_pkg BayesFitUtils
+
+# Import project package
+import BayesFitUtils
+
+# Import library package
+import BayesFitness
+
+# Import basic math
+import StatsBase
+import Distributions
+import Random
+import LinearAlgebra
+
+# Import iterator tools
+import Combinatorics
+
+# Import libraries to manipulate data
+import DataFrames as DF
+import CSV
+
+# Import library to list files
+import Glob
+
+# Import library to load files
+import JLD2
+
+# Import plotting libraries
+using CairoMakie
+import ColorSchemes
+
+# Activate backend
+CairoMakie.activate!()
+
+# Set PBoC Plotting style
+BayesFitUtils.viz.theme_makie!()
+
+Random.seed!(42)
+
+##
+
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Loading data
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
+# Generate output directory 
+if !isdir("./output/")
+    mkdir("./output/")
+end # if
+
+# Generate figure dictionary
+if !isdir("./output/figs/")
+    mkdir("./output/figs/")
+end # if
+
+println("Loading data...")
+
+# Import data
+df_counts = CSV.read(
+    "$(git_root())/data/logistic_growth/data_002/tidy_data.csv", DF.DataFrame
+)
+
+# Define number of experimental repeats
+n_rep = length(unique(df_counts.rep))
+
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Load variational inference
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
+println("Loading ADVI results...")
+# Define file
+file = first(Glob.glob("./output/advi_meanfield*3000*"))
+
+# Convert results to tidy dataframe
+df_advi = CSV.read(file, DF.DataFrame)
+
[…]

### 2023-10-09 code/processing/kinsler_2020/viz_hierarchicalreplicate_inference.jl

@@ -201,7 +201,7 @@ n_ppc = 500
 qs = [0.95, 0.675, 0.05]
 
 # Define number of rows and columns
-n_row, n_col = [2, 2]
+n_row, n_col = [4, 4]
 
 # List example barcodes to plot
 bc_plot = StatsBase.sample(
@@ -360,7 +360,8 @@ save("$(fig_dir)/ppc_$(env)env.pdf", fig)
 
 println("Plot the remainder in parallel")
 
-Threads.@threads for d = 2:length(df_group)
+# Threads.@threads 
+for d = 2:length(df_group)
     # Define quantiles to compute
     qs = [0.95, 0.675, 0.05]
 
@@ -434,7 +435,7 @@ Threads.@threads for d = 2:length(df_group)
     qs = [0.95, 0.675, 0.05]
 
     # Define number of rows and columns
-    n_row, n_col = [2, 2]
+    n_row, n_col = [4, 4]
 
     # List example barcodes to plot
     bc_plot = StatsBase.sample(

### 2023-10-09 doc/_bayes_priors.qmd
change from BayesFitness to BarBay

@@
 With this in mind, we leave it to the reader to judge the selection of priors.
-Furthermore, the software package associated with this work, `BayesFitness.jl`,
+Furthermore, the software package associated with this work, `BarBay.jl`,
 is written so that users can experiment with different prior selection criteria
@@
 
-The `BayesFitness.jl` package includes a function `naive_prior` within the
+The `BarBay.jl` package includes a function `naive_prior` within the
 `stats` module. This function utilizes the data from the neutral lineages to

### 2023-10-09 doc/_bayesian_inference.qmd
change from BayesFitness to BarBay

@@
 
-This work is accompanied by the Julia library `BayesFitness.jl` that makes use
+This work is accompanied by the Julia library `BarBay.jl` that makes use
 of the implementation of both MCMC-based integration as well as ADVI

### 2023-10-09 doc/_ppc.qmd
small fixes
change from BayesFitness to BarBay

@@
 
-## Posterior predictive checks
+## Posterior predictive checks {#sec-ppc}
 
@@
 
-In practice, to compute the posterior predictive checks, we generate multiple samples from the posterior distribution $\pi(\theta \mid \underline{r}^{(m)})$ 
+In practice, to compute the posterior predictive checks, we generate multiple
+samples from the posterior distribution $\pi(\theta \mid \underline{r}^{(m)})$ 
 $$
@@
 $${#eq-ppc_theta_samples}
-With these samples in hand, the `BayesFitness.jl` package includes the function
+With these samples in hand, the `BarBay.jl` package includes the function
 `logfreq_ratio_bc_ppc` for non-neutral barcodes that uses this set of posterior
@@
 raw estimates $\ln\hat{\gamma}_t^{(m)}$ with the desired probability. The
-package `BayesFitness.jl` includes an equivalent function,
+package `BarBay.jl` includes an equivalent function,
 `logfreq_ratio_popmean_ppc`, for neutral lineages.

### 2023-10-09 BayesFitUtils/src/viz.jl

@@ -16,7 +16,7 @@ import Distributions
 import MCMCChains
 
 # Import function from stats module
-import BayesFitness.stats: matrix_quantile_range
+import BarBay.stats: matrix_quantile_range
 
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 # Color Style

### 2023-10-09 code/exploratory/data002_logistic_250bc_01env_02rep/advi_meanfield_hierarchical_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -155,7 +155,7 @@ logσ_mut_prior = vcat(
 param = Dict(
     :data => data,
     :outputname => "./output/advi_meanfield_$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
-    :model => BayesFitness.model.exprep_fitness_normal,
+    :model => BarBay.model.exprep_fitness_normal,
     :model_kwargs => Dict(
         :s_pop_prior => s_pop_prior,
         :logσ_pop_prior => logσ_pop_prior,
@@ -181,4 +181,4 @@ end # if
 
 # Run inference
 println("Running Variational Inference...")
-@time dist = BayesFitness.vi.advi(; param...)
+@time dist = BarBay.vi.advi(; param...)

### 2023-10-09 code/exploratory/data002_logistic_250bc_01env_02rep/mcmc_joint_hierarchical_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -164,7 +164,7 @@ param = Dict(
     :n_walkers => n_walkers,
     :n_steps => n_steps,
     :outputname => "./output/chain_joint_hierarchical_fitness_$(n_steps)steps_$(lpad(n_walkers, 2, "0"))walkers",
-    :model => BayesFitness.model.exprep_fitness_lognormal,
+    :model => BarBay.model.exprep_fitness_lognormal,
     :model_kwargs => Dict(
         :s_pop_prior => s_pop_prior,
         :σ_pop_prior => σ_pop_prior,
@@ -180,4 +180,4 @@ param = Dict(
 # Run inference
 println("Running Inference...")
 
-@time BayesFitness.mcmc.mcmc_sample(; param...)
+@time BarBay.mcmc.mcmc_sample(; param...)

### 2023-10-09 code/exploratory/data002_logistic_250bc_01env_02rep/mcmc_joint_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -164,7 +164,7 @@ for (i, data) in enumerate(df_group)
         :n_walkers => n_walkers,
         :n_steps => n_steps,
         :outputname => "./output/chain_joint_fitness_R$(i)rep_$(n_steps)steps_$(lpad(n_walkers, 2, "0"))walkers",
-        :model => BayesFitness.model.fitness_lognormal,
+        :model => BarBay.model.fitness_lognormal,
         :model_kwargs => Dict(
             :s_pop_prior => s_pop_prior,
             :σ_pop_prior => σ_pop_prior,
@@ -179,5 +179,5 @@ for (i, data) in enumerate(df_group)
     # Run inference
     println("Running Inference...")
 
-    @time BayesFitness.mcmc.mcmc_joint_fitness(; param...)
+    @time BarBay.mcmc.mcmc_joint_fitness(; param...)
 end # for

### 2023-10-09 code/exploratory/data002_logistic_250bc_01env_02rep/viz_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import basic math
 import StatsBase
@@ -109,7 +109,7 @@ for (i, file) in enumerate(files)
     chn = JLD2.load(file)["chain"]
 
     # Compute posterior predictive checks
-    ppc_mat = BayesFitness.stats.logfreq_ratio_mean_ppc(
+    ppc_mat = BarBay.stats.logfreq_ratio_mean_ppc(
         chn, n_ppc; param=param
     )
 
@@ -246,7 +246,7 @@ for rep = 1:n_rep
     )
 
     # Compute posterior predictive checks
-    ppc_mat = BayesFitness.stats.logfreq_ratio_mean_ppc(
+    ppc_mat = BarBay.stats.logfreq_ratio_mean_ppc(
         df_chn, n_ppc; param=param
     )
 
@@ -635,7 +635,7 @@ for row in 1:n_row
                 :population_mean_fitness => Symbol("s̲ₜ_R$rep"),
             )
             # Compute posterior predictive checks
-            local ppc_mat = BayesFitness.stats.logfreq_ratio_mutant_ppc(
+            local ppc_mat = BarBay.stats.logfreq_ratio_mutant_ppc(
                 df_chn, n_ppc; param=param
             )
             # Plot posterior predictive checks

### 2023-10-09 code/exploratory/data004_logistic_zeronoise_250bc_01env_01rep/advi_meanfield_joint_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -114,7 +114,7 @@ logσ_mut_prior = logσ_pop_prior
 param = Dict(
     :data => data,
     :outputname => "./output/advi_meanfield_$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
-    :model => BayesFitness.model.fitness_normal,
+    :model => BarBay.model.fitness_normal,
     :model_kwargs => Dict(
         :s_pop_prior => s_pop_prior,
         :logσ_pop_prior => logσ_pop_prior,
@@ -139,4 +139,4 @@ end # if
 
 # Run inference
 println("Running Variational Inference...")
-@time dist = BayesFitness.vi.advi(; param...)
+@time dist = BarBay.vi.advi(; param...)

### 2023-10-09 code/exploratory/data004_logistic_zeronoise_250bc_01env_01rep/mcmc_freq.jl

@@ -10,7 +10,7 @@ import Revise
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -106,7 +106,7 @@ param = Dict(
     :n_walkers => n_walkers,
     :n_steps => n_steps,
     :outputname => "./output/chain_freq_$(n_steps)steps_$(lpad(n_walkers, 2, "0"))walkers",
-    :model => BayesFitness.model.freq_lognormal,
+    :model => BarBay.model.freq_lognormal,
     :sampler => Turing.DynamicNUTS(),
     :ensemble => Turing.MCMCThreads(),
     :rm_T0 => false,
@@ -115,4 +115,4 @@ param = Dict(
 # Run inference
 println("Running Inference...")
 
-@time BayesFitness.mcmc.mcmc_joint_fitness(; param...)
+@time BarBay.mcmc.mcmc_joint_fitness(; param...)

### 2023-10-09 code/exploratory/data004_logistic_zeronoise_250bc_01env_01rep/mcmc_joint_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -140,7 +140,7 @@ param = Dict(
     :n_walkers => n_walkers,
     :n_steps => n_steps,
     :outputname => "./output/chain_joint_fitness_$(n_steps)steps_$(lpad(n_walkers, 2, "0"))walkers",
-    :model => BayesFitness.model.fitness_lognormal,
+    :model => BarBay.model.fitness_lognormal,
     :model_kwargs => Dict(
         :s_pop_prior => s_pop_prior,
         :σ_pop_prior => σ_pop_prior,
@@ -155,4 +155,4 @@ param = Dict(
 # Run inference
 println("Running Inference...")
 
-@time BayesFitness.mcmc.mcmc_sample(; param...)
+@time BarBay.mcmc.mcmc_sample(; param...)

### 2023-10-09 code/exploratory/data004_logistic_zeronoise_250bc_01env_01rep/mcmc_popmean_fitness.jl

@@ -10,7 +10,7 @@ import Revise
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -105,7 +105,7 @@ param = Dict(
     :n_walkers => n_walkers,
     :n_steps => n_steps,
     :outputname => "./output/chain_popmean_fitness_$(n_steps)steps_$(lpad(n_walkers, 2, "0"))walkers",
-    :model => BayesFitness.model.neutrals_lognormal,
+    :model => BarBay.model.neutrals_lognormal,
     :sampler => Turing.DynamicNUTS(),
     :ensemble => Turing.MCMCThreads(),
     :rm_T0 => false,
@@ -113,4 +113,4 @@ param = Dict(
 
 # Run inference
 println("Running Inference...")
-@time BayesFitness.mcmc.mcmc_popmean_fitness(; param...)
+@time BarBay.mcmc.mcmc_popmean_fitness(; param...)

### 2023-10-09 code/exploratory/data004_logistic_zeronoise_250bc_01env_01rep/viz_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import basic math
 import StatsBase
@@ -127,7 +127,7 @@ qs = [0.05, 0.68, 0.95]
 colors = get(ColorSchemes.Blues_9, LinRange(0.25, 1.0, length(qs)))
 
 # Compute posterior predictive checks
-ppc_mat = BayesFitness.stats.logfreq_ratio_mean_ppc(
+ppc_mat = BarBay.stats.logfreq_ratio_mean_ppc(
     chn, n_ppc; param=param
 )
 
@@ -403,7 +403,7 @@ for row in 1:n_row
             :population_mean_fitness => :s̲ₜ,
         )
         # Compute posterior predictive checks
-        local ppc_mat = BayesFitness.stats.logfreq_ratio_mutant_ppc(
+        local ppc_mat = BarBay.stats.logfreq_ratio_mutant_ppc(
             chn, n_ppc; param=param
         )
         # Plot posterior predictive checks

### 2023-10-09 code/figs/fig02advi.jl

@@ -8,7 +8,7 @@ import Revise
 # Import project package
 import BayesFitUtils
 # Import package for Bayesian inference
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -200,7 +200,7 @@ for row in 1:n_row
             )
 
             # Compute posterior predictive checks
-            local ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+            local ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
                 df_samples, n_ppc; model=:normal, param=param
             )
 
@@ -251,7 +251,7 @@ for row in 1:n_row
             :population_mean_fitness => :s̲ₜ,
         )
         # Compute posterior predictive checks
-        local ppc_mat = BayesFitness.stats.logfreq_ratio_bc_ppc(
+        local ppc_mat = BarBay.stats.logfreq_ratio_bc_ppc(
             df_samples, n_ppc; model=:normal, param=param
         )
         # Plot posterior predictive checks

### 2023-10-09 code/figs/fig02mcmc.jl

@@ -8,7 +8,7 @@ import Revise
 # Import project package
 import BayesFitUtils
 # Import package for Bayesian inference
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -263,7 +263,7 @@ for row in 1:n_row
             )
 
             # Compute posterior predictive checks
-            local ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+            local ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
                 chn, n_ppc; param=param
             )
 
@@ -314,7 +314,7 @@ for row in 1:n_row
             :population_mean_fitness => :s̲ₜ,
         )
         # Compute posterior predictive checks
-        local ppc_mat = BayesFitness.stats.logfreq_ratio_mutant_ppc(
+        local ppc_mat = BarBay.stats.logfreq_ratio_mutant_ppc(
             chn, n_ppc; param=param
         )
         # Plot posterior predictive checks

### 2023-10-09 code/figs/fig03.jl

@@ -9,7 +9,7 @@ import Revise
 import BayesFitUtils
 
 # Import library to perform Bayesian inference
-import BayesFitness
+import BarBay
 
 # Import basic math
 import Random
@@ -291,7 +291,7 @@ for row in 1:n_row
             )
 
             # Compute posterior predictive checks
-            local ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+            local ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
                 df_samples, n_ppc; model=:normal, param=param
             )
 
@@ -370,7 +370,7 @@ for row in 1:n_row
             :population_mean_fitness => :s̲ₜ,
         )
         # Compute posterior predictive checks
-        local ppc_mat = BayesFitness.stats.logfreq_ratio_multienv_ppc(
+        local ppc_mat = BarBay.stats.logfreq_ratio_multienv_ppc(
             df_bc, n_ppc, data_bc.env; model=:normal, param=param
         )
 

### 2023-10-09 code/figs/fig04.jl

@@ -8,7 +8,7 @@ import Revise
 # Import project package
 import BayesFitUtils
 # Import package for Bayesian inference
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -396,7 +396,7 @@ for row in 1:n_row
                     :population_std_fitness => :logσ̲ₜ,
                 )
                 # Compute posterior predictive checks
-                ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+                ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
                     df_samples[:, value], n_ppc; model=:normal, param=param
                 )
 
@@ -455,7 +455,7 @@ for row in 1:n_row
                     :population_mean_fitness => Symbol("s̲ₜ"),
                 )
                 # Compute posterior predictive checks
-                local ppc_mat = BayesFitness.stats.logfreq_ratio_bc_ppc(
+                local ppc_mat = BarBay.stats.logfreq_ratio_bc_ppc(
                     df_samples[:, Symbol.(vars_bc)],
                     n_ppc;
                     model=:normal,

### 2023-10-09 code/figs/fig05.jl

@@ -8,7 +8,7 @@ import Revise
 # Import project package
 import BayesFitUtils
 # Import package for Bayesian inference
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF

### 2023-10-09 code/figs/fig06.jl

@@ -9,7 +9,7 @@ import Revise
 import BayesFitUtils
 
 # Import library to perform Bayesian inference
-import BayesFitness
+import BarBay
 
 # Import basic math
 import Random
@@ -423,7 +423,7 @@ for row in 1:n_row
             )
 
             # Compute posterior predictive checks
-            local ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+            local ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
                 df_samples, n_ppc; model=:normal, param=param
             )
 
@@ -474,7 +474,7 @@ for row in 1:n_row
             :population_mean_fitness => :s̲ₜ,
         )
         # Compute posterior predictive checks
-        local ppc_mat = BayesFitness.stats.logfreq_ratio_bc_ppc(
+        local ppc_mat = BarBay.stats.logfreq_ratio_bc_ppc(
             df_samples, n_ppc; model=:normal, param=param
         )
         # Plot posterior predictive checks

### 2023-10-09 code/figs/figSIX_logistic_growth.jl

@@ -8,7 +8,7 @@ import Revise
 # Import project package
 import BayesFitUtils
 # Import package for Bayesian inference
-import BayesFitness
+import BarBay
 
 # Import differential equations package
 using DifferentialEquations

### 2023-10-09 code/figs/web01.jl

@@ -5,7 +5,7 @@
 import BayesFitUtils
 
 # Import Bayesian inference package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -67,6 +67,6 @@ BayesFitUtils.viz.bc_time_series!(
     color=ColorSchemes.Blues_9[end],
 )
 
-save("/Users/mrazo/git/BayesFitness/docs/src/figs/fig01.svg", fig)
+save("/Users/mrazo/git/BarBay/docs/src/figs/fig01.svg", fig)
 
 fig

### 2023-10-09 code/figs/web02.jl

@@ -5,7 +5,7 @@
 import BayesFitUtils
 
 # Import Bayesian inference package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -61,6 +61,6 @@ BayesFitUtils.viz.logfreq_ratio_time_series!(
     color=ColorSchemes.Blues_9[end],
 )
 
-save("/Users/mrazo/git/BayesFitness/docs/src/figs/fig02.svg", fig)
+save("/Users/mrazo/git/BarBay/docs/src/figs/fig02.svg", fig)
 
 fig

### 2023-10-09 code/figs/web03.jl

@@ -5,7 +5,7 @@
 import BayesFitUtils
 
 # Import Bayesian inference package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -133,7 +133,7 @@ for row in 1:n_row
             )
 
             # Compute posterior predictive checks
-            local ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+            local ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
                 df_samples, n_ppc; model=:normal, param=param
             )
 
@@ -184,7 +184,7 @@ for row in 1:n_row
             :population_mean_fitness => :s̲ₜ,
         )
         # Compute posterior predictive checks
-        local ppc_mat = BayesFitness.stats.logfreq_ratio_bc_ppc(
+        local ppc_mat = BarBay.stats.logfreq_ratio_bc_ppc(
             df_samples, n_ppc; model=:normal, param=param
         )
         # Plot posterior predictive checks
@@ -222,6 +222,6 @@ Label(gl_ppc[:, 1, Left()], "ln(fₜ₊₁/fₜ)", rotation=π / 2, fontsize=22)
 rowgap!(gl_ppc, 0)
 colgap!(gl_ppc, 4)
 
-save("/Users/mrazo/git/BayesFitness/docs/src/figs/fig03.svg", fig)
+save("/Users/mrazo/git/BarBay/docs/src/figs/fig03.svg", fig)
 
 fig

### 2023-10-09 code/processing/abreu_2023/advi_meanfield_hierarchical_all.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -107,7 +107,7 @@ Threads.@threads for col = 1:n_col
     # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 
     # Compute naive priors from neutral strains
-    naive_priors = BayesFitness.stats.naive_prior(
+    naive_priors = BarBay.stats.naive_prior(
         data; rep_col=:rep, pseudocount=1
     )
 
@@ -141,7 +141,7 @@ Threads.@threads for col = 1:n_col
             :data => data,
             :outputname => "$(out_dir)/advi_meanfield_hierarchical_" *
                            "$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
-            :model => BayesFitness.model.replicate_fitness_normal,
+            :model => BarBay.model.replicate_fitness_normal,
             :model_kwargs => Dict(
                 :s_pop_prior => s_pop_prior,
                 :logσ_pop_prior => logσ_pop_prior,
@@ -160,7 +160,7 @@ Threads.@threads for col = 1:n_col
             :data => data,
             :outputname => "$(out_dir)/advi_meanfield_hierarchical_" *
                            "$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
-            :model => BayesFitness.model.multienv_replicate_fitness_normal,
+            :model => BarBay.model.multienv_replicate_fitness_normal,
             :model_kwargs => Dict(
                 :s_pop_prior => s_pop_prior,
                 :logσ_pop_prior => logσ_pop_prior,
@@ -181,6 +181,6 @@ Threads.@threads for col = 1:n_col
 
     # Run inference
     println("Running Variational Inference for $(condition)...")
-    @time dist = BayesFitness.vi.advi(; param...)
+    @time dist = BarBay.vi.advi(; param...)
 
 end # for

### 2023-10-09 code/processing/abreu_2023/advi_meanfield_hierarchical_joint_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -85,7 +85,7 @@ n_env = length(unique(data.environment))
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 
 # Compute naive priors from neutral strains
-naive_priors = BayesFitness.stats.naive_prior(
+naive_priors = BarBay.stats.naive_prior(
     data; rep_col=:replicate, pseudocount=1
 )
 
@@ -121,7 +121,7 @@ envs = collect(unique(data[:, [:time, :environment]])[:, :environment])
 param = Dict(
     :data => data,
     :outputname => "./output/advi_meanfield_hierarchical_$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
-    :model => BayesFitness.model.multienv_replicate_fitness_normal,
+    :model => BarBay.model.multienv_replicate_fitness_normal,
     :model_kwargs => Dict(
         :envs => envs,
         :s_pop_prior => s_pop_prior,
@@ -147,7 +147,7 @@ end # if
 
 # Run inference
 println("Running Variational Inference...")
-@time dist = BayesFitness.vi.advi(; param...)
+@time dist = BarBay.vi.advi(; param...)
 
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 # Convert output to tidy dataframe
@@ -170,7 +170,7 @@ dist = advi_results["dist"]
 vars = advi_results["var"]
 
 # Generate tidy dataframe with distribution information
-df_advi = BayesFitness.utils.advi_to_df(
+df_advi = BarBay.utils.advi_to_df(
     dist, vars, bc_ids; envs=envs
 )
 

### 2023-10-09 code/processing/abreu_2023/advi_meanfield_joint_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -97,7 +97,7 @@ Threads.@threads for rep = 1:length(data_rep)
     )
 
     # Compute naive priors from neutral strains
-    naive_priors = BayesFitness.stats.naive_prior(
+    naive_priors = BarBay.stats.naive_prior(
         data_rep[rep]; pseudocount=1
     )
 
@@ -126,7 +126,7 @@ Threads.@threads for rep = 1:length(data_rep)
     param = Dict(
         :data => data_rep[rep],
         :outputname => "./output/advi_meanfield_R$(rep)_$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
-        :model => BayesFitness.model.multienv_fitness_normal,
+        :model => BarBay.model.multienv_fitness_normal,
         :model_kwargs => Dict(
             :s_pop_prior => s_pop_prior,
             :logσ_pop_prior => logσ_pop_prior,
@@ -146,7 +146,7 @@ Threads.@threads for rep = 1:length(data_rep)
 
     # Run inference
     println("Running Variational Inference...")
-    @time dist = BayesFitness.vi.advi(; param...)
+    @time dist = BarBay.vi.advi(; param...)
 
 end # for
 
@@ -173,7 +173,7 @@ for f in files
     vars = advi_results["var"]
 
     # Generate tidy dataframe with distribution information
-    df_advi = BayesFitness.utils.advi_to_df(
+    df_advi = BarBay.utils.advi_to_df(
         dist, vars, bc_ids; envs=envs
     )
 

### 2023-10-09 code/processing/abreu_2023/viz_inference_hierarchical.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import basic math
 import LinearAlgebra
@@ -110,7 +110,7 @@ dist = advi_results["dist"]
 vars = advi_results["var"]
 
 # Generate tidy dataframe with distribution information
-df_advi = BayesFitness.utils.advi_to_df(dist, vars, bc_ids; n_rep=3, envs=envs)
+df_advi = BarBay.utils.advi_to_df(dist, vars, bc_ids; n_rep=3, envs=envs)
 
 # Rename env column and re-add env as numeric value
 DF.rename!(df_advi, :env => :environment)
@@ -345,7 +345,7 @@ fig = Figure(resolution=(450 * n_rep, 350))
 for (i, (key, value)) in enumerate(rep_vars)
 
     # Compute posterior predictive checks
-    ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+    ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
         df_samples[:, value], n_ppc; model=:normal, param=param
     )
 
@@ -481,7 +481,7 @@ for row in 1:n_row
                 :population_mean_fitness => Symbol("s̲ₜ"),
             )
             # Compute posterior predictive checks
-            local ppc_mat = BayesFitness.stats.logfreq_ratio_multienv_ppc(
+            local ppc_mat = BarBay.stats.logfreq_ratio_multienv_ppc(
                 df_samples[:, Symbol.(vars_bc)],
                 n_ppc,
                 envs;

### 2023-10-09 code/processing/abreu_2023/viz_inference_hierarchical_all.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import basic math
 import LinearAlgebra
@@ -67,7 +67,7 @@ for col = 1:n_col
     data = df_counts[df_idx[:, col], 1:12]
 
     # Process data to arrays
-    data_dict = BayesFitness.utils.data_to_arrays(
+    data_dict = BarBay.utils.data_to_arrays(
         data; rep_col=:rep, env_col=:env
     )
 
@@ -351,7 +351,7 @@ for col = 1:n_col
     for (i, (key, value)) in enumerate(rep_vars)
 
         # Compute posterior predictive checks
-        ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+        ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
             df_samples[:, value], n_ppc; model=:normal, param=param
         )
 
@@ -503,7 +503,7 @@ for col = 1:n_col
                         :population_mean_fitness => Symbol("s̲ₜ"),
                     )
                     # Compute posterior predictive checks
-                    local ppc_mat = BayesFitness.stats.logfreq_ratio_bc_ppc(
+                    local ppc_mat = BarBay.stats.logfreq_ratio_bc_ppc(
                         df_samples[:, Symbol.(vars_bc)],
                         n_ppc;
                         model=:normal,
@@ -518,7 +518,7 @@ for col = 1:n_col
                         :population_mean_fitness => Symbol("s̲ₜ"),
                     )
                     # Compute posterior predictive checks
-                    local ppc_mat = BayesFitness.stats.logfreq_ratio_multienv_ppc(
+                    local ppc_mat = BarBay.stats.logfreq_ratio_multienv_ppc(
                         df_samples[:, Symbol.(vars_bc)],
                         n_ppc,
                         envs[rep];

### 2023-10-09 code/processing/abreu_2023/viz_inference_single.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import basic math
 import StatsBase
@@ -141,7 +141,7 @@ for file in files
     ppc_color = get(ColorSchemes.Purples_9, LinRange(0.25, 1.0, length(qs)))
 
     # Compute posterior predictive checks
-    ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+    ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
         df_samples, n_ppc; model=:normal, param=param
     )
 
@@ -255,7 +255,7 @@ for file in files
                 :population_mean_fitness => :s̲ₜ,
             )
             # Compute posterior predictive checks
-            local ppc_mat = BayesFitness.stats.logfreq_ratio_multienv_ppc(
+            local ppc_mat = BarBay.stats.logfreq_ratio_multienv_ppc(
                 df_bc, n_ppc, envs; model=:normal, param=param
             )
 

### 2023-10-09 code/processing/aguilar_2023/advi_meanfield_hierarchicalgenotype_joint_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -94,7 +94,7 @@ Threads.@threads for i in eachindex(df_group)
     data = df_group[i]
 
     # Compute naive priors from neutral strains
-    naive_priors = BayesFitness.stats.naive_prior(
+    naive_priors = BarBay.stats.naive_prior(
         data; pseudocount=1, id_col=:oligo
     )
 
@@ -125,7 +125,7 @@ Threads.@threads for i in eachindex(df_group)
     oligo_edit_dict = Dict(oligo_edit[:, :oligo] .=> oligo_edit[:, :edit])
 
     # Extract list of oligos as they will be used in the inference
-    oligo_ids = BayesFitness.utils.data_to_arrays(data; id_col=:oligo)[:bc_ids]
+    oligo_ids = BarBay.utils.data_to_arrays(data; id_col=:oligo)[:bc_ids]
 
     # Extract edits in the order they will be used in the inference
     edit_list = [oligo_edit_dict[m] for m in oligo_ids]
@@ -136,7 +136,7 @@ Threads.@threads for i in eachindex(df_group)
         :data => data,
         :outputname => "./output/advi_meanfield_hierarchicalgenotypes_R$(i)rep_" *
                        "$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
-        :model => BayesFitness.model.genotype_fitness_normal,
+        :model => BarBay.model.genotype_fitness_normal,
         :model_kwargs => Dict(
             :s_pop_prior => s_pop_prior,
             :logσ_pop_prior => logσ_pop_prior,
@@ -152,6 +152,6 @@ Threads.@threads for i in eachindex(df_group)
 
     # Run inference
     println("Running Variational Inference...")
-    @time dist = BayesFitness.vi.advi(; param...)
+    @time dist = BarBay.vi.advi(; param...)
 
 end # for

### 2023-10-09 code/processing/aguilar_2023/advi_meanfield_hierarchicalreplicate_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -88,7 +88,7 @@ df_edit[!, :freq] = df_edit.count ./ df_edit.count_sum
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 
 # Compute naive priors from neutral strains
-naive_priors = BayesFitness.stats.naive_prior(
+naive_priors = BarBay.stats.naive_prior(
     df_edit; rep_col=:rep, pseudocount=1, id_col=:edit
 )
 
@@ -118,7 +118,7 @@ param = Dict(
     :data => df_edit,
     :outputname => "./output/advi_meanfield_hierarchicalreplicate" *
                    "$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
-    :model => BayesFitness.model.replicate_fitness_normal,
+    :model => BarBay.model.replicate_fitness_normal,
     :model_kwargs => Dict(
         :s_pop_prior => s_pop_prior,
         :logσ_pop_prior => logσ_pop_prior,
@@ -144,4 +144,4 @@ end # if
 
 # Run inference
 println("Running Variational Inference...")
-@time BayesFitness.vi.advi(; param...)
+@time BarBay.vi.advi(; param...)

### 2023-10-09 code/processing/aguilar_2023/viz_hierarchicalreplicate_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import basic math
 import StatsBase
@@ -298,7 +298,7 @@ fig = Figure(resolution=(400 * n_rep, 350))
 for (i, (key, value)) in enumerate(rep_vars)
 
     # Compute posterior predictive checks
-    ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+    ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
         df_samples[:, value], n_ppc; model=:normal, param=param
     )
 
@@ -409,7 +409,7 @@ for row in 1:n_row
                 :population_mean_fitness => Symbol("s̲ₜ"),
             )
             # Compute posterior predictive checks
-            local ppc_mat = BayesFitness.stats.logfreq_ratio_bc_ppc(
+            local ppc_mat = BarBay.stats.logfreq_ratio_bc_ppc(
                 df_samples[:, Symbol.(vars_bc)],
                 n_ppc;
                 model=:normal,

### 2023-10-09 code/processing/aguilar_2023/viz_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import basic math
 import StatsBase
@@ -67,7 +67,7 @@ df_counts = df_counts[df_counts.rep.!="N/A", :]
 oligo_edit_dict = Dict(values.(keys(DF.groupby(df_counts, [:oligo, :edit]))))
 
 # Extract list of mutants as they were used in the inference
-oligo_ids = BayesFitness.utils.data_to_arrays(
+oligo_ids = BarBay.utils.data_to_arrays(
     df_counts[df_counts.rep.=="R1", :]; id_col=:oligo
 )[:bc_ids]
 
@@ -96,7 +96,7 @@ for (i, file) in enumerate(files)
     # Load distribution
     setindex!(advi_output, JLD2.load(file), "R$(i)")
     # Convert distribution to tidy dataframe
-    df_advi = BayesFitness.utils.advi_to_df(
+    df_advi = BarBay.utils.advi_to_df(
         advi_output["R$(i)"]["dist"],
         advi_output["R$(i)"]["var"],
         advi_output["R$(i)"]["ids"];
@@ -164,7 +164,7 @@ rm("./output/figs/advi_logfreqratio_ppc_neutral.pdf"; force=true)
 for rep in sort(collect(eachindex(advi_output)))
     println("Generating plot for $(rep)")
     # Compute posterior predictive checks
-    ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+    ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
         advi_output[rep]["df_samples"], n_ppc; model=:normal, param=param
     )
 
@@ -282,7 +282,7 @@ for rep in sort(collect(eachindex(advi_output)))
                 :population_mean_fitness => Symbol("s̲ₜ"),
             )
             # Compute posterior predictive checks
-            local ppc_mat = BayesFitness.stats.logfreq_ratio_mutant_ppc(
+            local ppc_mat = BarBay.stats.logfreq_ratio_mutant_ppc(
                 df_samples[:, Symbol.(vars_bc)],
                 n_ppc;
                 model=:normal,

### 2023-10-09 code/processing/bigbatch/advi_meanfield_joint_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -94,7 +94,7 @@ Threads.@threads for i in 1:length(df_keys)
     data = df_group[i]
 
     # Compute naive parameters
-    prior_param = BayesFitness.stats.naive_prior_neutral(data)
+    prior_param = BarBay.stats.naive_prior_neutral(data)
 
     # Define parameter for population mean fitness adding a standard deviation
     s_pop_prior = hcat(
@@ -108,7 +108,7 @@ Threads.@threads for i in 1:length(df_keys)
         :outputname => "$(outdir)/bigbatch_$(batch)batch_$(hub)hub_" *
                        "$(pert)perturbation_$(rep)rep_" *
                        "$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
-        :model => BayesFitness.model.fitness_normal,
+        :model => BarBay.model.fitness_normal,
         :model_kwargs => Dict(
             :s_pop_prior => s_pop_prior,
             :logσ_pop_prior => prior_param[:logσ_pop_prior],
@@ -124,5 +124,5 @@ Threads.@threads for i in 1:length(df_keys)
     # Run inference
     println("Running Inference for group $(i) / $(length(df_group))...")
 
-    @time BayesFitness.vi.advi(; param...)
+    @time BarBay.vi.advi(; param...)
 end # for

### 2023-10-09 code/processing/bigbatch/mcmc_mean_fitness_test.jl

@@ -8,8 +8,8 @@ import Revise
 import Suppressor
 # Import project package
 import BayesFitUtils
-# Import BayesFitness package
-import BayesFitness
+# Import BarBay package
+import BarBay
 # Import library to list files
 import Glob
 # Import library to save output

### 2023-10-09 code/processing/bigbatch/viz_trajectories.jl

@@ -10,7 +10,7 @@ import Revise
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF

### 2023-10-09 code/processing/data001_logistic_250bc_01env_01rep/mcmc_joint_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -75,7 +75,7 @@ data = CSV.read(
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 
 # Compute naive priors from neutral strains
-naive_priors = BayesFitness.stats.naive_prior(data)
+naive_priors = BarBay.stats.naive_prior(data)
 
 # Select standard deviation parameters
 s_pop_prior = hcat(
@@ -107,7 +107,7 @@ param = Dict(
     :n_walkers => n_walkers,
     :n_steps => n_steps,
     :outputname => "./output/chain_joint_fitness_$(n_steps)steps_$(lpad(n_walkers, 2, "0"))walkers",
-    :model => BayesFitness.model.fitness_normal,
+    :model => BarBay.model.fitness_normal,
     :model_kwargs => Dict(
         :s_pop_prior => s_pop_prior,
         :logσ_pop_prior => logσ_pop_prior,
@@ -123,4 +123,4 @@ param = Dict(
 # Run inference
 println("Running Inference...")
 
-@time BayesFitness.mcmc.mcmc_sample(; param...)
+@time BarBay.mcmc.mcmc_sample(; param...)

### 2023-10-09 code/processing/data001_logistic_250bc_01env_01rep/viz_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import basic math
 import LinearAlgebra
@@ -127,7 +127,7 @@ qs = [0.05, 0.68, 0.95]
 colors = get(ColorSchemes.Blues_9, LinRange(0.25, 1.0, length(qs)))
 
 # Compute posterior predictive checks
-ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
     chn, n_ppc; param=param, model=:normal
 )
 
@@ -398,7 +398,7 @@ for row in 1:n_row
             :population_mean_fitness => :s̲ₜ,
         )
         # Compute posterior predictive checks
-        local ppc_mat = BayesFitness.stats.logfreq_ratio_mutant_ppc(
+        local ppc_mat = BarBay.stats.logfreq_ratio_mutant_ppc(
             chn, n_ppc; param=param
         )
         # Plot posterior predictive checks
@@ -666,7 +666,7 @@ ppc_color = get(ColorSchemes.Blues_9, LinRange(0.25, 1.0, length(qs)))
 fig = Figure(resolution=(400, 350))
 
 # Compute posterior predictive checks
-ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
     df_samples, n_ppc; model=:normal, param=param
 )
 
@@ -761,7 +761,7 @@ for row in 1:n_row
             :population_mean_fitness => Symbol("s̲ₜ"),
         )
         # Compute posterior predictive checks
-        local ppc_mat = BayesFitness.stats.logfreq_ratio_mutant_ppc(
+        local ppc_mat = BarBay.stats.logfreq_ratio_mutant_ppc(
             df_samples[:, Symbol.(vars_bc)],
             n_ppc;
             model=:normal,

### 2023-10-09 code/processing/data002_logistic_1000bc_01env_02rep/advi_meanfield_hierarchical_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -71,7 +71,7 @@ data = CSV.read(
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 
 # Compute naive priors from neutral strains
-naive_priors = BayesFitness.stats.naive_prior(data; rep_col=:rep, pseudocount=1)
+naive_priors = BarBay.stats.naive_prior(data; rep_col=:rep, pseudocount=1)
 
 # Select standard deviation parameters
 s_pop_prior = hcat(
@@ -99,7 +99,7 @@ param = Dict(
     :data => data,
     :outputname => "./output/advi_meanfield_" *
                    "$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
-    :model => BayesFitness.model.replicate_fitness_normal,
+    :model => BarBay.model.replicate_fitness_normal,
     :model_kwargs => Dict(
         :s_pop_prior => s_pop_prior,
         :logσ_pop_prior => logσ_pop_prior,
@@ -125,4 +125,4 @@ end # if
 
 # Run inference
 println("Running Variational Inference...")
-@time BayesFitness.vi.advi(; param...)
+@time BarBay.vi.advi(; param...)

### 2023-10-09 code/processing/data002_logistic_1000bc_01env_02rep/advi_meanfield_single_inference.jl

@@ -7,7 +7,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -76,7 +76,7 @@ Threads.@threads for data in df_group
     # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 
     # Compute naive priors from neutral strains
-    naive_priors = BayesFitness.stats.naive_prior(data)
+    naive_priors = BarBay.stats.naive_prior(data)
 
     # Select standard deviation parameters
     s_pop_prior = hcat(
@@ -104,7 +104,7 @@ Threads.@threads for data in df_group
         :data => data,
         :outputname => "./output/advi_meanfield_$(first(data.rep))rep_" *
                        "$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
-        :model => BayesFitness.model.fitness_normal,
+        :model => BarBay.model.fitness_normal,
         :model_kwargs => Dict(
             :s_pop_prior => s_pop_prior,
             :logσ_pop_prior => logσ_pop_prior,
@@ -128,5 +128,5 @@ Threads.@threads for data in df_group
 
     # Run inference
     println("Running Variational Inference...")
-    @time BayesFitness.vi.advi(; param...)
+    @time BarBay.vi.advi(; param...)
 end # for

### 2023-10-09 code/processing/data002_logistic_1000bc_01env_02rep/viz_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import basic math
 import StatsBase
@@ -510,7 +510,7 @@ fig = Figure(resolution=(400 * n_rep, 350))
 for (i, (key, value)) in enumerate(sort(rep_vars))
 
     # Compute posterior predictive checks
-    ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+    ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
         df_samples[:, value], n_ppc; model=:normal, param=param
     )
 
@@ -621,7 +621,7 @@ for row in 1:n_row
                 :population_mean_fitness => Symbol("s̲ₜ"),
             )
             # Compute posterior predictive checks
-            local ppc_mat = BayesFitness.stats.logfreq_ratio_bc_ppc(
+            local ppc_mat = BarBay.stats.logfreq_ratio_bc_ppc(
                 df_samples[:, Symbol.(vars_bc)],
                 n_ppc;
                 model=:normal,

### 2023-10-09 code/processing/data003_logistic_1000bc_03env_01rep/advi_meanfield_joint_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -72,7 +72,7 @@ data = CSV.read(
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 
 # Compute naive priors from neutral strains
-naive_priors = BayesFitness.stats.naive_prior(data; pseudocount=1)
+naive_priors = BarBay.stats.naive_prior(data; pseudocount=1)
 
 # Select standard deviation parameters
 s_pop_prior = hcat(
@@ -99,7 +99,7 @@ logλ_prior = hcat(
 param = Dict(
     :data => data,
     :outputname => "./output/advi_meanfield_$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
-    :model => BayesFitness.model.multienv_fitness_normal,
+    :model => BarBay.model.multienv_fitness_normal,
     :model_kwargs => Dict(
         :s_pop_prior => s_pop_prior,
         :logσ_pop_prior => logσ_pop_prior,
@@ -124,4 +124,4 @@ end # if
 
 # Run inference
 println("Running Variational Inference...")
-@time BayesFitness.vi.advi(; param...)
+@time BarBay.vi.advi(; param...)

### 2023-10-09 code/processing/data003_logistic_1000bc_03env_01rep/mcmc_joint_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -159,7 +159,7 @@ param = Dict(
     :n_walkers => n_walkers,
     :n_steps => n_steps,
     :outputname => "./output/chain_joint_fitness_$(n_steps)steps_$(lpad(n_walkers, 2, "0"))walkers",
-    :model => BayesFitness.model.multienv_fitness_lognormal,
+    :model => BarBay.model.multienv_fitness_lognormal,
     :model_kwargs => Dict(
         :s_pop_prior => s_pop_prior,
         :σ_pop_prior => σ_pop_prior,
@@ -175,4 +175,4 @@ param = Dict(
 # Run inference
 println("Running Inference...")
 
-@time BayesFitness.mcmc.mcmc_joint_fitness(; param...)
+@time BarBay.mcmc.mcmc_joint_fitness(; param...)

### 2023-10-09 code/processing/data003_logistic_1000bc_03env_01rep/viz_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import basic math
 import StatsBase
@@ -119,7 +119,7 @@ qs = [0.05, 0.68, 0.95]
 ppc_color = get(ColorSchemes.Purples_9, LinRange(0.25, 1.0, length(qs)))
 
 # Compute posterior predictive checks
-ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
     df_samples, n_ppc; model=:normal, param=param
 )
 
@@ -389,7 +389,7 @@ for row in 1:n_row
             :population_mean_fitness => :s̲ₜ,
         )
         # Compute posterior predictive checks
-        local ppc_mat = BayesFitness.stats.logfreq_ratio_multienv_ppc(
+        local ppc_mat = BarBay.stats.logfreq_ratio_multienv_ppc(
             df_bc, n_ppc, data_bc.env; model=:normal, param=param
         )
 

### 2023-10-09 code/processing/data004_logistic_zeronoise_1000bc_01env_01rep/advi_meanfield_joint_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -70,7 +70,7 @@ data = CSV.read(
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 
 # Compute naive priors from neutral strains
-naive_priors = BayesFitness.stats.naive_prior(data)
+naive_priors = BarBay.stats.naive_prior(data)
 
 # Select standard deviation parameters
 s_pop_prior = hcat(
@@ -98,7 +98,7 @@ param = Dict(
     :data => data,
     :outputname => "./output/advi_meanfield_" *
                    "$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
-    :model => BayesFitness.model.fitness_normal,
+    :model => BarBay.model.fitness_normal,
     :model_kwargs => Dict(
         :s_pop_prior => s_pop_prior,
         :logσ_pop_prior => logσ_pop_prior,
@@ -122,4 +122,4 @@ end # if
 
 # Run inference
 println("Running Variational Inference...")
-@time dist = BayesFitness.vi.advi(; param...)
+@time dist = BarBay.vi.advi(; param...)

### 2023-10-09 code/processing/data004_logistic_zeronoise_1000bc_01env_01rep/viz_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import basic math
 import LinearAlgebra
@@ -164,7 +164,7 @@ ppc_color = get(ColorSchemes.Blues_9, LinRange(0.25, 1.0, length(qs)))
 fig = Figure(resolution=(400, 350))
 
 # Compute posterior predictive checks
-ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
     df_samples, n_ppc; model=:normal, param=param
 )
 
@@ -259,7 +259,7 @@ for row in 1:n_row
             :population_mean_fitness => Symbol("s̲ₜ"),
         )
         # Compute posterior predictive checks
-        local ppc_mat = BayesFitness.stats.logfreq_ratio_mutant_ppc(
+        local ppc_mat = BarBay.stats.logfreq_ratio_mutant_ppc(
             df_samples[:, Symbol.(vars_bc)],
             n_ppc;
             model=:normal,

### 2023-10-09 code/processing/data005_logistic_zeronoise_1000bc_01env_02rep/advi_meanfield_hierarchical_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -74,7 +74,7 @@ data = CSV.read(
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 
 # Compute naive priors from neutral strains
-naive_priors = BayesFitness.stats.naive_prior(data; rep_col=:rep, pseudocount=1)
+naive_priors = BarBay.stats.naive_prior(data; rep_col=:rep, pseudocount=1)
 
 # Select standard deviation parameters
 s_pop_prior = hcat(
@@ -101,7 +101,7 @@ logλ_prior = hcat(
 param = Dict(
     :data => data,
     :outputname => "./output/advi_meanfield_$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
-    :model => BayesFitness.model.replicate_fitness_normal,
+    :model => BarBay.model.replicate_fitness_normal,
     :model_kwargs => Dict(
         :s_pop_prior => s_pop_prior,
         :logσ_pop_prior => logσ_pop_prior,
@@ -126,4 +126,4 @@ end # if
 
 # Run inference
 println("Running Variational Inference...")
-@time dist = BayesFitness.vi.advi(; param...)
+@time dist = BarBay.vi.advi(; param...)

### 2023-10-09 code/processing/data005_logistic_zeronoise_1000bc_01env_02rep/viz_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import basic math
 import StatsBase
@@ -365,7 +365,7 @@ fig = Figure(resolution=(400 * n_rep, 350))
 for (i, (key, value)) in enumerate(rep_vars)
 
     # Compute posterior predictive checks
-    ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+    ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
         df_samples[:, value], n_ppc; model=:normal, param=param
     )
 
@@ -472,7 +472,7 @@ for row in 1:n_row
                 :population_mean_fitness => Symbol("s̲ₜ"),
             )
             # Compute posterior predictive checks
-            local ppc_mat = BayesFitness.stats.logfreq_ratio_mutant_ppc(
+            local ppc_mat = BarBay.stats.logfreq_ratio_mutant_ppc(
                 df_samples[:, Symbol.(vars_bc)],
                 n_ppc;
                 model=:normal,

### 2023-10-09 code/processing/data006_logistic_5000bc_01env_01rep/advi_meanfield_joint_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -70,7 +70,7 @@ data = CSV.read(
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 
 # Compute naive priors from neutral strains
-naive_priors = BayesFitness.stats.naive_prior(data; pseudocount=1)
+naive_priors = BarBay.stats.naive_prior(data; pseudocount=1)
 
 # Select standard deviation parameters
 s_pop_prior = hcat(
@@ -98,7 +98,7 @@ param = Dict(
     :data => data,
     :outputname => "./output/advi_meanfield_" *
                    "$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
-    :model => BayesFitness.model.fitness_normal,
+    :model => BarBay.model.fitness_normal,
     :model_kwargs => Dict(
         :s_pop_prior => s_pop_prior,
         :logσ_pop_prior => logσ_pop_prior,
@@ -122,4 +122,4 @@ end # if
 
 # Run inference
 println("Running Variational Inference...")
-@time dist = BayesFitness.vi.advi(; param...)
+@time dist = BarBay.vi.advi(; param...)

### 2023-10-09 code/processing/data006_logistic_5000bc_01env_01rep/viz_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import basic math
 import LinearAlgebra
@@ -164,7 +164,7 @@ ppc_color = get(ColorSchemes.Blues_9, LinRange(0.5, 1.0, length(qs)))
 fig = Figure(resolution=(400, 350))
 
 # Compute posterior predictive checks
-ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
     df_samples, n_ppc; model=:normal, param=param
 )
 
@@ -260,7 +260,7 @@ for row in 1:n_row
             :population_mean_fitness => Symbol("s̲ₜ"),
         )
         # Compute posterior predictive checks
-        local ppc_mat = BayesFitness.stats.logfreq_ratio_mutant_ppc(
+        local ppc_mat = BarBay.stats.logfreq_ratio_mutant_ppc(
             df_samples[:, Symbol.(vars_bc)],
             n_ppc;
             model=:normal,

### 2023-10-09 code/processing/data007_logistic_hierarchicalgenotype_1000bc_01env_01rep/advi_meanfield_hierarchicalgenotype_joint_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -68,7 +68,7 @@ data = CSV.read(
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 
 # Compute naive priors from neutral strains
-naive_priors = BayesFitness.stats.naive_prior(data; pseudocount=1)
+naive_priors = BarBay.stats.naive_prior(data; pseudocount=1)
 
 # Select standard deviation parameters
 s_pop_prior = hcat(
@@ -96,7 +96,7 @@ param = Dict(
     :data => data,
     :outputname => "./output/advi_meanfield_hierarchicalgenotypes_" *
                    "$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
-    :model => BayesFitness.model.genotype_fitness_normal,
+    :model => BarBay.model.genotype_fitness_normal,
     :model_kwargs => Dict(
         :s_pop_prior => s_pop_prior,
         :logσ_pop_prior => logσ_pop_prior,
@@ -121,4 +121,4 @@ end # if
 
 # Run inference
 println("Running Variational Inference...")
-@time dist = BayesFitness.vi.advi(; param...)
+@time dist = BarBay.vi.advi(; param...)

### 2023-10-09 code/processing/data007_logistic_hierarchicalgenotype_1000bc_01env_01rep/advi_meanfield_joint_inference.jl

@@ -7,7 +7,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -67,7 +67,7 @@ data = CSV.read(
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 
 # Compute naive priors from neutral strains
-naive_priors = BayesFitness.stats.naive_prior(data)
+naive_priors = BarBay.stats.naive_prior(data)
 
 # Select standard deviation parameters
 s_pop_prior = hcat(
@@ -95,7 +95,7 @@ param = Dict(
     :data => data,
     :outputname => "./output/advi_meanfield_" *
                    "$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
-    :model => BayesFitness.model.fitness_normal,
+    :model => BarBay.model.fitness_normal,
     :model_kwargs => Dict(
         :s_pop_prior => s_pop_prior,
         :logσ_pop_prior => logσ_pop_prior,
@@ -119,4 +119,4 @@ end # if
 
 # Run inference
 println("Running Variational Inference...")
-@time BayesFitness.vi.advi(; param...)
+@time BarBay.vi.advi(; param...)

### 2023-10-09 code/processing/data007_logistic_hierarchicalgenotype_1000bc_01env_01rep/advi_meanfield_pool_inference.jl

@@ -7,7 +7,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -94,7 +94,7 @@ data[!, :freq] = data.count ./ data.count_sum
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 
 # Compute naive priors from neutral strains
-naive_priors = BayesFitness.stats.naive_prior(data; id_col=:genotype)
+naive_priors = BarBay.stats.naive_prior(data; id_col=:genotype)
 
 # Select standard deviation parameters
 s_pop_prior = hcat(
@@ -122,7 +122,7 @@ param = Dict(
     :data => data,
     :outputname => "./output/advi_meanfield_genotype-group_" *
                    "$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
-    :model => BayesFitness.model.fitness_normal,
+    :model => BarBay.model.fitness_normal,
     :model_kwargs => Dict(
         :s_pop_prior => s_pop_prior,
         :logσ_pop_prior => logσ_pop_prior,
@@ -147,4 +147,4 @@ end # if
 
 # Run inference
 println("Running Variational Inference...")
-@time BayesFitness.vi.advi(; param...)
+@time BarBay.vi.advi(; param...)

### 2023-10-09 code/processing/data007_logistic_hierarchicalgenotype_1000bc_01env_01rep/viz_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import basic math
 import StatsBase
@@ -60,7 +60,7 @@ data = CSV.read(
 bc_geno_dict = Dict(values.(keys(DF.groupby(data, [:barcode, :genotype]))))
 
 # Extract list of mutants as they were used in the inference
-bc_ids = BayesFitness.utils.data_to_arrays(data)[:bc_ids]
+bc_ids = BarBay.utils.data_to_arrays(data)[:bc_ids]
 
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 # Load variational inference
@@ -113,7 +113,7 @@ qs = [0.05, 0.68, 0.95]
 colors = get(ColorSchemes.Blues_9, LinRange(0.25, 1.0, length(qs)))
 
 # Compute posterior predictive checks
-ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
     df_samples, n_ppc; model=:normal, param=param
 )
 
@@ -351,7 +351,7 @@ for row in 1:n_row
             :population_mean_fitness => Symbol("s̲ₜ"),
         )
         # Compute posterior predictive checks
-        local ppc_mat = BayesFitness.stats.logfreq_ratio_bc_ppc(
+        local ppc_mat = BarBay.stats.logfreq_ratio_bc_ppc(
             df_samples[:, Symbol.(vars_bc)],
             n_ppc;
             model=:normal,

### 2023-10-09 code/processing/data008_logistic_250bc_03env_02rep/advi_meanfield_hierarchicalgenotype_joint_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -153,7 +153,7 @@ env_idx = [1, 1, 2, 3, 1, 2, 3]
 param = Dict(
     :data => data,
     :outputname => "./output/advi_meanfield_$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
-    :model => BayesFitness.model.multienv_replicate_fitness_normal,
+    :model => BarBay.model.multienv_replicate_fitness_normal,
     :model_kwargs => Dict(
         :envs => env_idx,
         :s_pop_prior => s_pop_prior,
@@ -178,4 +178,4 @@ end # if
 
 # Run inference
 println("Running Variational Inference...")
-@time dist = BayesFitness.vi.advi(; param...)
+@time dist = BarBay.vi.advi(; param...)

### 2023-10-09 code/processing/data008_logistic_250bc_03env_02rep/viz_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import basic math
 import LinearAlgebra
@@ -92,7 +92,7 @@ dist = advi_results["dist"]
 vars = advi_results["var"]
 
 # Generate tidy dataframe with distribution information
-df_advi = BayesFitness.utils.advi_to_df(
+df_advi = BarBay.utils.advi_to_df(
     dist, vars, bc_ids; n_rep=2, envs=[1, 1, 2, 3, 1, 2, 3]
 )
 
@@ -386,7 +386,7 @@ fig = Figure(resolution=(450 * 2, 350))
 for (i, (key, value)) in enumerate(rep_vars)
 
     # Compute posterior predictive checks
-    ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+    ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
         df_samples[:, value], n_ppc; model=:normal, param=param
     )
 
@@ -521,7 +521,7 @@ for row in 1:n_row
                 :population_mean_fitness => Symbol("s̲ₜ"),
             )
             # Compute posterior predictive checks
-            local ppc_mat = BayesFitness.stats.logfreq_ratio_multienv_ppc(
+            local ppc_mat = BarBay.stats.logfreq_ratio_multienv_ppc(
                 df_samples[:, Symbol.(vars_bc)],
                 n_ppc,
                 envs;

### 2023-10-09 code/processing/data009_logistic_1200bc_03env_03rep/advi_meanfield_hierarchicalgenotype_joint_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -73,7 +73,7 @@ n_env = length(unique(data.env))
 println("Defining priors from neutral lineages data...\n")
 
 # Compute naive priors from neutral strains
-naive_priors = BayesFitness.stats.naive_prior(
+naive_priors = BarBay.stats.naive_prior(
     data; rep_col=:rep, pseudocount=1
 )
 
@@ -106,7 +106,7 @@ env_idx = [1, 1, 2, 3, 1, 2, 3]
 param = Dict(
     :data => data,
     :outputname => "./output/advi_meanfield_$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
-    :model => BayesFitness.model.multienv_replicate_fitness_normal,
+    :model => BarBay.model.multienv_replicate_fitness_normal,
     :model_kwargs => Dict(
         :envs => env_idx,
         :s_pop_prior => s_pop_prior,
@@ -132,4 +132,4 @@ end # if
 
 # Run inference
 println("Running Variational Inference...")
-@time dist = BayesFitness.vi.advi(; param...)
+@time dist = BarBay.vi.advi(; param...)

### 2023-10-09 code/processing/data009_logistic_1200bc_03env_03rep/advi_meanfield_joint_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -88,7 +88,7 @@ Threads.@threads for rep = 1:length(data_rep)
     )
 
     # Compute naive priors from neutral strains
-    naive_priors = BayesFitness.stats.naive_prior(
+    naive_priors = BarBay.stats.naive_prior(
         data_rep[rep]; pseudocount=1
     )
 
@@ -117,7 +117,7 @@ Threads.@threads for rep = 1:length(data_rep)
     param = Dict(
         :data => data_rep[rep],
         :outputname => "./output/advi_meanfield_R$(rep)_$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
-        :model => BayesFitness.model.multienv_fitness_normal,
+        :model => BarBay.model.multienv_fitness_normal,
         :model_kwargs => Dict(
             :s_pop_prior => s_pop_prior,
             :logσ_pop_prior => logσ_pop_prior,
@@ -137,7 +137,7 @@ Threads.@threads for rep = 1:length(data_rep)
 
     # Run inference
     println("Running Variational Inference...")
-    @time dist = BayesFitness.vi.advi(; param...)
+    @time dist = BarBay.vi.advi(; param...)
 
 end # for
 
@@ -164,7 +164,7 @@ for f in files
     vars = advi_results["var"]
 
     # Generate tidy dataframe with distribution information
-    global df_advi = BayesFitness.utils.advi_to_df(
+    global df_advi = BarBay.utils.advi_to_df(
         dist, vars, bc_ids; n_rep=1, envs=envs
     )
 

### 2023-10-09 code/processing/data009_logistic_1200bc_03env_03rep/viz_inference_hierarchical.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import basic math
 import LinearAlgebra
@@ -90,7 +90,7 @@ dist = advi_results["dist"]
 vars = advi_results["var"]
 
 # Generate tidy dataframe with distribution information
-df_advi = BayesFitness.utils.advi_to_df(
+df_advi = BarBay.utils.advi_to_df(
     dist, vars, bc_ids; n_rep=3, envs=[1, 1, 2, 3, 1, 2, 3]
 )
 
@@ -444,7 +444,7 @@ fig = Figure(resolution=(450 * n_rep, 350))
 for (i, (key, value)) in enumerate(rep_vars)
 
     # Compute posterior predictive checks
-    ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+    ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
         df_samples[:, value], n_ppc; model=:normal, param=param
     )
 
@@ -580,7 +580,7 @@ for row in 1:n_row
                 :population_mean_fitness => Symbol("s̲ₜ"),
             )
             # Compute posterior predictive checks
-            local ppc_mat = BayesFitness.stats.logfreq_ratio_multienv_ppc(
+            local ppc_mat = BarBay.stats.logfreq_ratio_multienv_ppc(
                 df_samples[:, Symbol.(vars_bc)],
                 n_ppc,
                 envs;

### 2023-10-09 code/processing/data009_logistic_1200bc_03env_03rep/viz_inference_single.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import basic math
 import StatsBase
@@ -128,7 +128,7 @@ for file in files
     ppc_color = get(ColorSchemes.Purples_9, LinRange(0.25, 1.0, length(qs)))
 
     # Compute posterior predictive checks
-    ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+    ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
         df_samples, n_ppc; model=:normal, param=param
     )
 
@@ -398,7 +398,7 @@ for file in files
                 :population_mean_fitness => :s̲ₜ,
             )
             # Compute posterior predictive checks
-            local ppc_mat = BayesFitness.stats.logfreq_ratio_multienv_ppc(
+            local ppc_mat = BarBay.stats.logfreq_ratio_multienv_ppc(
                 df_bc, n_ppc, data_bc.env; model=:normal, param=param
             )
 

### 2023-10-09 code/processing/data011_logistic_1000bc_01env_08rep/viz_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import basic math
 import StatsBase
@@ -366,7 +366,7 @@ fig = Figure(resolution=(400 * n_rep, 350))
 for (i, (key, value)) in enumerate(rep_vars)
 
     # Compute posterior predictive checks
-    ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+    ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
         df_samples[:, value], n_ppc; model=:normal, param=param
     )
 
@@ -473,7 +473,7 @@ for row in 1:n_row
                 :population_mean_fitness => Symbol("s̲ₜ"),
             )
             # Compute posterior predictive checks
-            local ppc_mat = BayesFitness.stats.logfreq_ratio_bc_ppc(
+            local ppc_mat = BarBay.stats.logfreq_ratio_bc_ppc(
                 df_samples[:, Symbol.(vars_bc)],
                 n_ppc;
                 model=:normal,

### 2023-10-09 code/processing/kinsler_2020/advi_meanfield_hierarchicalreplicate_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -113,7 +113,7 @@ Threads.@threads for i = 1:length(rep_envs)
     end # for
 
     # Compute naive priors from neutral strains
-    naive_priors = BayesFitness.stats.naive_prior(
+    naive_priors = BarBay.stats.naive_prior(
         data; rep_col=:rep, pseudocount=1
     )
 
@@ -140,7 +140,7 @@ Threads.@threads for i = 1:length(rep_envs)
         :data => data,
         :outputname => "$(outdir)/advi_meanfield_hierarchical_$(env)" *
                        "$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
-        :model => BayesFitness.model.replicate_fitness_normal,
+        :model => BarBay.model.replicate_fitness_normal,
         :model_kwargs => Dict(
             :s_pop_prior => s_pop_prior,
             :logσ_pop_prior => logσ_pop_prior,
@@ -157,5 +157,5 @@ Threads.@threads for i = 1:length(rep_envs)
 
     # Run inference
     println("Running Variational Inference...")
-    @time BayesFitness.vi.advi(; param...)
+    @time BarBay.vi.advi(; param...)
 end # for

### 2023-10-09 code/processing/kinsler_2020/advi_meanfield_joint_inference.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -124,7 +124,7 @@ Threads.@threads for i in axes(df_include, 1)
     data = df[(df.env.==env).&(df.rep.==rep), :]
 
     # Compute naive priors from neutral strains
-    naive_priors = BayesFitness.stats.naive_prior(data; rm_T0=rm_T0)
+    naive_priors = BarBay.stats.naive_prior(data; rm_T0=rm_T0)
 
     # Select standard deviation parameters
     s_pop_prior = hcat(
@@ -149,7 +149,7 @@ Threads.@threads for i in axes(df_include, 1)
         :data => data,
         :outputname => "$(outdir)/kinsler_$(env)env_$(rep)rep_$(rm_T0)rmT0_" *
                        "$(lpad(n_samples, 2, "0"))samples_$(n_steps)steps",
-        :model => BayesFitness.model.fitness_normal,
+        :model => BarBay.model.fitness_normal,
         :model_kwargs => Dict(
             :s_pop_prior => s_pop_prior,
             :logσ_pop_prior => logσ_pop_prior,
@@ -166,5 +166,5 @@ Threads.@threads for i in axes(df_include, 1)
     # Run inference
     println("Running Inference for group $(i)...")
 
-    @time BayesFitness.vi.advi(; param...)
+    @time BarBay.vi.advi(; param...)
 end # for

### 2023-10-09 code/processing/kinsler_2020/mcmc_freq.jl

@@ -10,7 +10,7 @@ import Revise
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -154,7 +154,7 @@ for i in axes(df_include, 1)
         :n_walkers => n_walkers,
         :n_steps => n_steps,
         :outputname => "$(outdir)/kinsler_$(env)env_$(rep)rep_$(rm_T0)rmT0_$(n_steps)steps_$(lpad(2, n_walkers, "0")walkers)",
-        :model => BayesFitness.model.freq_lognormal,
+        :model => BarBay.model.freq_lognormal,
         :sampler => Turing.DynamicNUTS(),
         :ensemble => Turing.MCMCThreads(),
         :rm_T0 => rm_T0,
@@ -164,7 +164,7 @@ for i in axes(df_include, 1)
     println("Running Inference for group $(i)...")
 
     try
-        @time BayesFitness.mcmc.mcmc_joint_fitness(; param...)
+        @time BarBay.mcmc.mcmc_joint_fitness(; param...)
     catch
         @warn "Group $(i) was already processed"
     end

### 2023-10-09 code/processing/kinsler_2020/mcmc_joint_inference_multiprocess.jl

@@ -20,7 +20,7 @@ println("Importing packages on all processess...\n")
 # Import packages everywhere
 Distributed.@everywhere begin
     # Import library package
-    import BayesFitness
+    import BarBay
 
     # Import MCMC-related packages
     import Turing
@@ -83,7 +83,7 @@ for i = 1:size(df_include, 1)
         :n_walkers => n_walkers,
         :n_steps => n_steps,
         :outputname => "./output/kinsler_$(env)env_$(rep)rep_$(rm_T0)rmT0",
-        :model => BayesFitness.model.fitness_lognormal,
+        :model => BarBay.model.fitness_lognormal,
         :sampler => Turing.NUTS(0.65),
         :ensemble => Turing.MCMCDistributed(),
         :rm_T0 => rm_T0,
@@ -94,7 +94,7 @@ for i = 1:size(df_include, 1)
     println("Running Inference for group $(i)...")
 
     try
-        @time BayesFitness.mcmc.mcmc_joint_fitness(; param...)
+        @time BarBay.mcmc.mcmc_joint_fitness(; param...)
     catch
         @warn "Group $(i) was already processed"
     end

### 2023-10-09 code/processing/kinsler_2020/mcmc_joint_inference_multiprocess_multithread.jl

@@ -17,7 +17,7 @@ println("Importing packages on all processess...\n")
 # Import packages everywhere
 Distributed.@everywhere begin
     # Import library package
-    import BayesFitness
+    import BarBay
 
     # Import MCMC-related packages
     import Turing
@@ -78,7 +78,7 @@ Distributed.@everywhere function pmap_inference(i)
         :n_walkers => n_walkers,
         :n_steps => n_steps,
         :outputname => "./output/kinsler_$(env)env_$(rep)rep_$(rm_T0)rmT0",
-        :model => BayesFitness.model.fitness_lognormal,
+        :model => BarBay.model.fitness_lognormal,
         :sampler => Turing.NUTS(0.65),
         :ensemble => Turing.MCMCThreads(),
         :rm_T0 => rm_T0,
@@ -88,7 +88,7 @@ Distributed.@everywhere function pmap_inference(i)
     println("Running Inference for group $(i)...")
 
     try
-        @time BayesFitness.mcmc.mcmc_joint_fitness(; param...)
+        @time BarBay.mcmc.mcmc_joint_fitness(; param...)
     catch
         @warn "Group $(i) was already processed"
     end

### 2023-10-09 code/processing/kinsler_2020/mcmc_joint_inference_multithread.jl

@@ -8,7 +8,7 @@ println("Loading packages...")
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -170,7 +170,7 @@ for i in axes(df_include, 1)
         :n_walkers => n_walkers,
         :n_steps => n_steps,
         :outputname => "$(outdir)/kinsler_$(env)env_$(rep)rep_$(rm_T0)rmT0_$(n_steps)steps_$(lpad(2, n_walkers, "0"))walkers",
-        :model => BayesFitness.model.fitness_lognormal,
+        :model => BarBay.model.fitness_lognormal,
         :model_kwargs => Dict(
             :s_pop_prior => s_pop_prior,
             :σ_pop_prior => σ_pop_prior,
@@ -185,7 +185,7 @@ for i in axes(df_include, 1)
     println("Running Inference for group $(i)...")
 
     try
-        @time BayesFitness.mcmc.mcmc_joint_fitness(; param...)
+        @time BarBay.mcmc.mcmc_joint_fitness(; param...)
     catch
         @warn "Group $(i) was already processed"
     end

### 2023-10-09 code/processing/kinsler_2020/mcmc_mutant_fitness.jl

@@ -10,7 +10,7 @@ import Revise
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -82,8 +82,8 @@ for i = 1:length(df_group)
     end # if
 
     # Infer mean fitness distributions
-    mean_fitness_dist = BayesFitness.stats.gaussian_prior_mean_fitness(
-        BayesFitness.utils.var_jld2_to_df(outdir, meanfit_pattern, :sₜ)
+    mean_fitness_dist = BarBay.stats.gaussian_prior_mean_fitness(
+        BarBay.utils.var_jld2_to_df(outdir, meanfit_pattern, :sₜ)
     )
 
     println("Processing $(env)-$(rep) mutant fitness ($i / $(length(df_group))) \n")
@@ -95,9 +95,9 @@ for i = 1:length(df_group)
         :n_steps => 4_000,
         :outputdir => outdir,
         :outputname => "$(env)_$(rep)_mutantfitness",
-        :model => BayesFitness.model.mutant_fitness_lognormal,
+        :model => BarBay.model.mutant_fitness_lognormal,
         :model_kwargs => Dict(
-            :α => BayesFitness.stats.beta_prior_mutant(
+            :α => BarBay.stats.beta_prior_mutant(
                 data[data.time.==minimum(data.time), :barcode],
             ),
             :μ_s̄ => mean_fitness_dist[1],
@@ -107,6 +107,6 @@ for i = 1:length(df_group)
     )
 
     # Run inference in a multithread fasshion
-    BayesFitness.mcmc.mcmc_mutant_fitness(; param...)
+    BarBay.mcmc.mcmc_mutant_fitness(; param...)
 
 end # for

### 2023-10-09 code/processing/kinsler_2020/mcmc_mutant_rmT0_fitness.jl

@@ -10,7 +10,7 @@ import Revise
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -88,8 +88,8 @@ for i = 1:length(df_group)
     end # if
 
     # Infer mean fitness distributions
-    mean_fitness_dist = BayesFitness.stats.gaussian_prior_mean_fitness(
-        BayesFitness.utils.var_jld2_to_df(sort!(mean_files), :sₜ)
+    mean_fitness_dist = BarBay.stats.gaussian_prior_mean_fitness(
+        BarBay.utils.var_jld2_to_df(sort!(mean_files), :sₜ)
     )
 
     println("Processing $(env)-$(rep) mutant fitness ($i / $(length(df_group))) \n")
@@ -101,9 +101,9 @@ for i = 1:length(df_group)
         :n_steps => 4_000,
         :outputdir => outdir,
         :outputname => "$(env)_$(rep)_mutantfitness_rmT0",
-        :model => BayesFitness.model.mutant_fitness_lognormal,
+        :model => BarBay.model.mutant_fitness_lognormal,
         :model_kwargs => Dict(
-            :α => BayesFitness.stats.beta_prior_mutant(
+            :α => BarBay.stats.beta_prior_mutant(
                 data[data.time.==minimum(data.time), :barcode],
             ),
             :μ_s̄ => mean_fitness_dist[1],
@@ -113,6 +113,6 @@ for i = 1:length(df_group)
     )
 
     # Run inference in a multithread fasshion
-    BayesFitness.mcmc.mcmc_mutant_fitness(; param...)
+    BarBay.mcmc.mcmc_mutant_fitness(; param...)
 
 end # for

### 2023-10-09 code/processing/kinsler_2020/mcmc_popmean_fitness.jl

@@ -10,7 +10,7 @@ import Revise
 import BayesFitUtils
 
 # Import library package
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -157,7 +157,7 @@ for i in axes(df_include, 1)
         :n_walkers => n_walkers,
         :n_steps => n_steps,
         :outputname => "$(outdir)/kinsler_$(env)env_$(rep)rep_$(rm_T0)rmT0_$(n_steps)steps_$(lpad(2, n_walkers, "0")walkers)",
-        :model => BayesFitness.model.neutrals_lognormal,
+        :model => BarBay.model.neutrals_lognormal,
         :sampler => Turing.DynamicNUTS(),
         :ensemble => Turing.MCMCThreads(),
         :rm_T0 => rm_T0,
@@ -167,7 +167,7 @@ for i in axes(df_include, 1)
     println("Running Inference for group $(i)...")
 
     try
-        @time BayesFitness.mcmc.mcmc_popmean_fitness(; param...)
+        @time BarBay.mcmc.mcmc_popmean_fitness(; param...)
     catch
         @warn "Group $(i) was already processed"
     end
@@ -190,7 +190,7 @@ if gen_plots
         end # if
 
         # Concatenate population mean fitness chains into single chain
-        chains = BayesFitness.utils.var_jld2_concat(od, "meanfitness", :sₜ)
+        chains = BarBay.utils.var_jld2_concat(od, "meanfitness", :sₜ)
 
         # Initialize figure
         fig = Figure(resolution=(800, 800))

### 2023-10-09 code/processing/kinsler_2020/viz_hierarchicalreplicate_inference.jl

@@ -8,7 +8,7 @@ import Revise
 # Import project package
 import BayesFitUtils
 # Import package for Bayesian inference
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -240,7 +240,7 @@ for row in 1:n_row
                     :population_std_fitness => :logσ̲ₜ,
                 )
                 # Compute posterior predictive checks
-                ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+                ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
                     df_samples[:, value], n_ppc; model=:normal, param=param
                 )
 
@@ -302,7 +302,7 @@ for row in 1:n_row
                     :population_mean_fitness => Symbol("s̲ₜ"),
                 )
                 # Compute posterior predictive checks
-                local ppc_mat = BayesFitness.stats.logfreq_ratio_bc_ppc(
+                local ppc_mat = BarBay.stats.logfreq_ratio_bc_ppc(
                     df_samples[:, Symbol.(vars_bc)],
                     n_ppc;
                     model=:normal,
@@ -472,7 +472,7 @@ for d = 2:length(df_group)
                         :population_std_fitness => :logσ̲ₜ,
                     )
                     # Compute posterior predictive checks
-                    ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+                    ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
                         df_samples[:, value], n_ppc; model=:normal, param=param
                     )
 
@@ -541,7 +541,7 @@ for d = 2:length(df_group)
                         :population_mean_fitness => Symbol("s̲ₜ"),
                     )
                     # Compute posterior predictive checks
-                    local ppc_mat = BayesFitness.stats.logfreq_ratio_bc_ppc(
+                    local ppc_mat = BarBay.stats.logfreq_ratio_bc_ppc(
                         df_samples[:, Symbol.(vars_bc)],
                         n_ppc;
                         model=:normal,

### 2023-10-09 code/processing/kinsler_2020/viz_joint_inference.jl

@@ -8,7 +8,7 @@ import Revise
 # Import project package
 import BayesFitUtils
 # Import package for Bayesian inference
-import BayesFitness
+import BarBay
 
 # Import libraries to manipulate data
 import DataFrames as DF
@@ -208,7 +208,7 @@ for row in 1:n_row
             )
 
             # Compute posterior predictive checks
-            local ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+            local ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
                 df_samples, n_ppc; model=:normal, param=param
             )
 
@@ -266,7 +266,7 @@ for row in 1:n_row
             :population_mean_fitness => :s̲ₜ,
         )
         # Compute posterior predictive checks
-        local ppc_mat = BayesFitness.stats.logfreq_ratio_bc_ppc(
+        local ppc_mat = BarBay.stats.logfreq_ratio_bc_ppc(
             df_samples, n_ppc; model=:normal, param=param
         )
         # Plot posterior predictive checks
@@ -402,7 +402,7 @@ Threads.@threads for d = 2:length(df_group)
                 )
 
                 # Compute posterior predictive checks
-                local ppc_mat = BayesFitness.stats.logfreq_ratio_popmean_ppc(
+                local ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
                     df_samples, n_ppc; model=:normal, param=param
                 )
 
@@ -460,7 +460,7 @@ Threads.@threads for d = 2:length(df_group)
                 :population_mean_fitness => :s̲ₜ,
             )
             # Compute posterior predictive checks
-            local ppc_mat = BayesFitness.stats.logfreq_ratio_bc_ppc(
+            local ppc_mat = BarBay.stats.logfreq_ratio_bc_ppc(
                 df_samples, n_ppc; model=:normal, param=param
             )
             # Plot posterior predictive checks

### 2023-10-10 doc/_bayes_hierarchical.qmd
add section name
update section naming

@@
 
-### Hierarchical models for multiple experimental replicates
+### Hierarchical models for multiple experimental replicates {#sec-hierarchical_model}
 
@@
 where we leave the conditioning on the population mean fitness as we did in
-@seq-bayes_mutfit. This expression can be simplified in two ways. First, the
+@sec-bayes_mutfit. This expression can be simplified in two ways. First, the
 frequency values for each experimental replicate depend directly on the local
@@
 The first term on the right-hand side of @eq-hier_bayes_exp_indep takes the same
-functional form as the one derived in @seq-bayes_mutfit. Therefore, all we are
+functional form as the one derived in @sec-bayes_mutfit. Therefore, all we are
 left with is to determine the functional forms for the hyper–prior
@@
 $\pi(\underline{s}^M_{[j]} \mid \underline{\theta}^M)$. In analogy to the 
-assumptions used for the fitness values in @seq-bayes_mutfit, we define the 
+assumptions used for the fitness values in @sec-bayes_mutfit, we define the 
 value of each hyper-fitness as independent. This means that we have

### 2023-10-10 doc/_bayes_mutfit.qmd
update section naming

@@
 
-### Mutant relative fitness uncertainty $\pi(\underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}}, \underline{\underline{R}})$ {#seq-bayes_mutfit}
+### Mutant relative fitness uncertainty $\pi(\underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}}, \underline{\underline{R}})$ {#sec-bayes_mutfit}
 

### 2023-10-11 doc/paper.tex
modifying tex file

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{lineno}
\linenumbers
\usepackage[font={footnotesize, sf}, labelfont=bf, margin=0.05\linewidth]{caption}
\renewcommand{\familydefault}{\sfdefault}
% Bold the 'Figure #' in the caption and separate it from the title/caption
% with a period
% Captions will be left justified
\usepackage[
	aboveskip=1pt,
	labelfont=bf,
	labelsep=period,
	justification=raggedright,
	singlelinecheck=off
]{caption}

% Add numbered lines
\usepackage{lineno}
\linenumbers

% Package to include multiple title pages
% This will allow me to add a tile to the main text and to the SI
\usepackage{titling}

% Add author affiliations
\usepackage{authblk}

\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

% References
% \usepackage[]{biblatex}
\usepackage[
	backend=bibtex,
	style=phys, 
	sorting=none,				% Do not sort bibliography
	url=false, 					% Do not show url in reference
	doi=false, 					% Do not show doi in reference
	isbn=false, 				% Do not show isbn link in reference
	eprint=false, 			% Do not show eprint link in reference
	maxbibnames=9, 			% Include up to 9 names in citation
	firstinits=true,
]{biblatex}
\addbibresource{references.bib}

\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdfauthor={Manuel Razo-Mejia; Madhav Mani; Dmitri Petrov},
  pdfkeywords={bayesian inference, variational inference, relative
fitness, open-source, reproducible research},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

% Title
\title{Bayesian inference of relative fitness on high-throughput pooled competition assays}

% Authors
\author[1]{Manuel Razo-Mejia}
\author[2,3,4]{Madhav Mani}
\author[1,5,*]{Dmitri Petrov}

% Affiliations
\affil[1]{Department of Biology, Stanford University, Stanford, CA, USA}
\affil[2]{Department of Engineering Sciences and Applied Mathematics, Northwestern University, Chicago, IL, USA}
\affil[3]{NSF-Simons Center for Quantitative Biology, Northwestern University, Chicago, IL, USA}
\affil[4]{Department of Molecular Biosciences, Northwestern University, Chicago, IL, USA}
\affil[5]{Stanford Cancer Institute, Stanford University School of Medicine, Stanford, CA, USA}
\affil[*]{Correspondence: dpetrov@stanford.edu}

\setcounter{Maxaffil}{0}
% Set affiliations in small font
\renewcommand\Affilfont{\itshape\small}

\date{}

\begin{document}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden, interior hidden, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi

\maketitle

\begin{abstract}
    The tracking of lineage frequencies via DNA barcode sequencing enables the
  quantification of microbial fitness. However, experimental noise coming from
  biotic and abiotic sources complicates the computation of a reliable
  inference. We present a Bayesian pipeline to infer relative microbial fitness
  from high-throughput lineage tracking assays. Our model accounts for multiple
  sources of noise and propagates uncertainties throughout all parameters in a
  systematic way. Furthermore, using modern variational inference methods based
  on automatic differentiation, we are able to scale the inference to a large
  number of unique barcodes. We extend this core model to analyze
  multi-environment assays, replicate experiments, and barcodes linked to
  genotypes. On simulations, our method recovers known parameters within
  posterior credible intervals. This work provides a generalizable Bayesian
  framework to analyze lineage tracking experiments. The accompanying
  open-source software library enables the adoption of principled statistical
  methods in experimental evolution.
\end{abstract}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The advent of DNA barcoding---the ability to uniquely identify cell
lineages with DNA sequences integrated at a specific locus---and
high-throughput sequencing has opened new venues for understanding
microbial evolutionary dynamics with an unprecedented level of temporal
resolution \autocite{levy2015,nguyenba2019a,ascensao2023}. These
experimental efforts rely on our ability to reliably infer the relative
fitness of an ensemble of diverse genotypes. Moreover, inferring these
fitness values over an ensemble of environmental conditions can help us
determine the phenotypic diversity of a rapid adaptation process
\autocite{kinsler2020}.

As with any other sequencing-based quantification, tracking lineages via
DNA barcode sequencing is inexorably accompanied by noise sources coming
from experimental manipulation of the microbial cultures, DNA
extraction, and sequencing library preparation that involves multiple
rounds of PCR amplification, and the sequencing process itself. Thus,
accounting for the uncertainty when inferring the relevant parameters
from the data is a crucial step to draw reliable conclusions. Bayesian
statistics presents a paradigm by which one can account for all known
sources of uncertainty in a principled way \autocite{eddy2004a}. This,
combined with the development of modern Markov Chain Monte Carlo
sampling algorithms \autocite{betancourt2017} and approximate
variational approaches \autocite{kucukelbir2016} have boosted a
resurgence of Bayesian methods in different fields
\autocite{efron2013a}.

We present a Bayesian inference pipeline to quantify the uncertainty
about the parametric information we can extract from high-throughput
competitive fitness assays given a model of the data generation process
and experimental data. In these assays, the fitness of an ensemble of
genotypes is determined relative to a reference genotype
\autocite{kinsler2020,ascensao2023}. Figure~\ref{fig-01}(A) shows a
schematic of the experimental procedure in which an initial pool of
barcoded strains are mixed with a reference strain and inoculated into
fresh media. After some time---usually, enough time for the culture to
saturate---an aliquot is transferred to fresh media, while the remaining
culture is used for DNA sequencing of the lineage barcodes. The
time-series information of the relative abundance of each lineage, i.e.,
the barcode frequency depicted in Figure~\ref{fig-01}(B), is used to
infer the relative fitness---the growth advantage on a per-cycle
basis---for each lineage with respect to the reference strain. The
proposed statistical model accounts for multiple sources of uncertainty
when inferring the lineages' relative fitness values (see
Section~\ref{sec-experiment} for details on sources of uncertainty
accounted for by the model). Furthermore, minor changes to the core
statistical model allow us to account for relevant experimental
variations of these competition assays. More specifically, in
Section~\ref{sec-multienv}, we present a variation of the statistical
model to infer fitness on growth dilution cycles in multiple
environments with proper error propagation. Furthermore, as described in
Section~\ref{sec-replicates}, our statistical model can account for
batch-to-batch differences when jointly analyzing multiple experimental
replicates using a Bayesian hierarchical model. Finally, a variant of
these hierarchical models, presented in Section~\ref{sec-genotypes}, can
account for variability within multiple barcodes mapping to equivalent
genotypes within the same experiment.

For all the model variations presented in this paper, we benchmark the
ability of our pipeline to infer relative fitness parameters against
synthetic data generated from logistic growth simulations with added
random noise. A \texttt{Julia} package accompanies the present method to
readily implement the inference pipeline with state-of-the-art
scientific computing software.

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{sec-experiment}{%
\subsection{Experimental setup}\label{sec-experiment}}

The present work is designed to analyze time-series data of relative
abundance of multiple microbial lineages uniquely identified by a DNA
barcode \autocite{kinsler2020,ascensao2023}. In these competition
assays, an ensemble of genotypes is pooled together with an unlabeled
reference strain that, initially, represents the vast majority
(\(\geq 90\%\)) of the cells in the culture (see schematic in
Figure~\ref{fig-01}(A)). Furthermore, a fraction of labeled genotypes
equivalent to the unlabeled reference strain---hereafter defined as
\emph{neutral} lineages---are spiked in at a relatively high abundance
(\(\approx 3-5\%\)). The rest of the culture is left for the ensemble of
genotypes of interest.

To determine the relative fitness of the ensemble of genotypes, a series
of growth-dilution cycles are performed on either a single or multiple
environments. In other words, the cultures are grown for some time;
then, an aliquot is inoculated into fresh media for the next growth
cycle. This process is repeated for roughly 4-7 cycles, depending on the
initial abundances of the mutants and their relative growth rates. The
DNA barcodes are sequenced at the end of each growth cycle to quantify
the relative abundance of each of the barcodes. We point the reader to
\textcite{kinsler2020} for specific details on these assays for \emph{S.
cerevisiae} and to \textcite{ascensao2023} for equivalent assays for
\emph{E. coli}. Figure~\ref{fig-01}(B) presents a typical barcode
trajectory where the black trajectories represent the so-called
\emph{neutral lineages}, genetically equivalent to the untagged ancestor
strain that initially dominates the culture. These spiked-in neutral
lineages simplify the inference problem since the fitness metric of all
relevant barcodes is quantified with respect to these barcodes---thus
referred to as \emph{relative fitness}.

\begin{figure}

{\centering \includegraphics{./figs/fig01.pdf}

}

\caption{\label{fig-01}\textbf{Typical competitive fitness experiment}.
(A) Schematic of the typical experimental design to determine the
competitive fitness of an ensemble of barcoded genotypes. Genotypes are
pooled together and grown over multiple growth-dilution cycles. At the
end of each cycle, a sample is processed to generate a library for
amplicon sequencing. (B) Typical barcode trajectory dataset. From each
time point, the relative frequency of each barcode is determined from
the total number of reads. Shades of blue represent different relative
fitness. Darker gray lines define the typical trajectory of neutral
lineages.}

\end{figure}

\hypertarget{preliminaries-on-mathematical-notation}{%
\subsection{Preliminaries on mathematical
notation}\label{preliminaries-on-mathematical-notation}}

Before jumping directly into the Bayesian inference pipeline, let us
establish the mathematical notation used throughout this paper. We
define (column) vectors as underlined lowercase symbols such as
\begin{equation}\protect\hypertarget{eq-vec_def}{}{
\underline{x} = \begin{bmatrix}
    x_1\\
    x_2\\
    \vdots\\
    x_N
\end{bmatrix}.
}\label{eq-vec_def}\end{equation} In the same way, we define matrices as
double-underline uppercase symbols such as
\begin{equation}\protect\hypertarget{eq-mat_def}{}{
\underline{\underline{A}} =
\begin{bmatrix}
    A_{11} & A_{12} & \cdots & A_{1N}\\
    A_{21} & A_{22} & \cdots & A_{2N}\\
    \vdots & \vdots & \ddots & \vdots\\
    A_{M1} & A_{M2} & \cdots & A_{MN}\\
\end{bmatrix}.
}\label{eq-mat_def}\end{equation}

\hypertarget{sec-fitness_model}{%
\subsection{Fitness model}\label{sec-fitness_model}}

Empirically, each barcode frequency trajectory follows an exponential
function of the form \autocite{levy2015,kinsler2020,ascensao2023}
\begin{equation}\protect\hypertarget{eq-fitness}{}{
f_{t+1}^{(b)} = f_{t}^{(b)} \mathrm{e}^{(s^{(b)} - \bar{s}_t)\tau},
}\label{eq-fitness}\end{equation} where \(f_{t}^{(b)}\) is the frequency
of barcode \(b\) at the end of cycle number \(t\), \(s^{(b)}\) is the
relative fitness with respect to the reference strain---the quantity we
want to infer from the data---\(\bar{s}_t\) is the mean fitness of the
culture at the end of cycle number \(t\), and \(\tau\) is the time pass
between cycle \(t\) and \(t+1\). We can rewrite
Equation~\ref{eq-fitness} as
\begin{equation}\protect\hypertarget{eq-logfreq}{}{
\frac{1}{\tau}\ln \frac{f_{t+1}^{(b)}}{f_{t}^{(b)}} = (s^{(b)} - \bar{s}_t).
}\label{eq-logfreq}\end{equation} Equation~\ref{eq-logfreq} separates
the measurements---the barcode frequencies---from the unobserved
(sometimes referred to as latent) parameters we want to infer from the
data---the population mean fitness and the barcode relative fitness.
This is ultimately the functional form used in our inference pipeline.
Therefore, the relative fitness is computed by knowing the log frequency
ratio of each barcode throughout the growth-dilution cycles.

The presence of the neutral lineages facilitates the determination of
the population mean fitness value \(\bar{s}_t\). Since every relative
fitness is determined relative to the neutral lineage that dominates the
culture, we define their fitness to be \(s^{(n)} = 0\), where the
superscript \((n)\) specifies their neutrality. This means that
Equation~\ref{eq-logfreq} for a neutral lineage takes the simpler form
\begin{equation}\protect\hypertarget{eq-logfreq_neutral}{}{
\frac{1}{\tau}\ln \frac{f_{t+1}^{(n)}}{f_{t}^{(n)}} = - \bar{s}_t.
}\label{eq-logfreq_neutral}\end{equation} Therefore, we can use the data
from these reference barcodes to directly infer the value of the
population mean fitness.

It is important to notice that the frequencies \(f_{t}^{(b)}\) are not
the allele frequencies in the population (most of the culture is not
sequenced since the reference strain is not barcoded), but rather the
relative frequencies in the total number of sequencing reads. A way to
conceptualize this subtle but important point is to assume exponential
growth in the \emph{number of cells} \(N_t^{(b)}\) of the form
\begin{equation}\protect\hypertarget{eq-ncells}{}{
N_{t+1}^{(b)} = N_{t}^{(b)} \mathrm{e}^{\lambda^{(b)}\tau},
}\label{eq-ncells}\end{equation} for every barcode \(b\) with growth
rate \(\lambda^{(b)}\). However, when we sequence barcodes, we do not
directly measure the number of cells, but some number of reads
\(r_t^{(b)}\) that map to barcode \(b\). In the simplest possible
scenario, we assume \begin{equation}\protect\hypertarget{eq-r_to_n}{}{
r_{t}^{(b)} \propto N_{t}^{(b)},
}\label{eq-r_to_n}\end{equation} where, importantly, the proportionality
constant depends on the total number of reads for the library for cycle
\(t\), which might vary from library to library. Therefore, to compare
the number of reads between libraries at different time points, we must
normalize the number of reads to the same scale. The simplest form is to
define a relative abundance, i.e., a frequency with respect to the total
number of reads, \begin{equation}\protect\hypertarget{eq-r_to_n_norm}{}{
f_{t}^{(b)} \equiv \frac{r_{t}^{(b)}}{\sum_{b'} r_{t}^{(b')}}.
}\label{eq-r_to_n_norm}\end{equation} This is the frequency
Equation~\ref{eq-fitness} describes.

Our ultimate objective is to infer the relative fitness \(s^{(b)}\) for
each of the \(M\) relevant barcodes in the experiment---hereafter
referred to as \(s^{(m)}\) to distinguish from the general \(s^{(b)}\)
and the neutral lineages \(s^{(n)}\) relative fitness. To do so, we
account for the three primary sources of uncertainty in our model:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Uncertainty in the determination of frequencies. Our model relates
  frequencies between adjacent growth-dilution cycles to the fitness of
  the corresponding strain. However, we do not directly measure
  frequencies. Instead, our data for each barcode consists of a length
  \(T\) vector of counts \(\underline{r}^{(b)}\) for each of the \(T\)
  cycles in which the measurements were taken.
\item
  Uncertainty in the value of the population mean fitness. We define
  neutral lineages to have fitness \(s^{(n)} = 0\), helping us anchor
  the value of the population mean fitness \(\bar{s}_t\) for each pair
  of adjacent growth cycles. Moreover, we take this parameter as an
  empirical parameter to be obtained from the data, meaning that we do
  not impose a functional form that relates \(\bar{s}_t\) to
  \(\bar{s}_{t+1}\). Thus, we must infer the \(T-1\) values of this
  population mean fitness with their uncertainty that must be propagated
  to the value of the mutants' relative fitness.
\item
  Uncertainty in each of the mutants' fitness values.
\end{enumerate}

To account for all these sources of uncertainty in a principled way, in
the next section, we develop a Bayesian inference pipeline.

\hypertarget{sec-bayesian_inference}{%
\subsection{Bayesian inference}\label{sec-bayesian_inference}}

As defined in Section~\ref{sec-fitness_model}, our ultimate objective is
to infer the vector of relative fitness values
\begin{equation}\protect\hypertarget{eq-fitness_mut_vec}{}{
\underline{s}^M = (s^{(1)}, s^{(2)}, \ldots, s^{(M)})^\dagger,
}\label{eq-fitness_mut_vec}\end{equation} where \(^\dagger\) indicates
the transpose. Our data consists of an \(T \times B\) matrix
\(\underline{\underline{R}}\), where \(B = M + N\) is the number of
unique barcodes given by the sum of the number of unique, relevant
barcodes we care about, \(M\), and the number of unique neutral
barcodes, \(N\), and \(T\) is the number of growth cycles where
measurements were taken. The data matrix is then of the form
\begin{equation}\protect\hypertarget{eq-R_mat}{}{
\underline{\underline{R}} = \begin{bmatrix}
- & \underline{r}_1 & - \\
- & \underline{r}_2 & - \\
 & \vdots & \\
- & \underline{r}_T & - \\
\end{bmatrix},
}\label{eq-R_mat}\end{equation} where each row \(\underline{r}_t\) is a
\(B\)-dimensional array containing the raw barcode counts at cycle
\(t\). We can further split each vector \(\underline{r}_t\) into two
vectors of the form
\begin{equation}\protect\hypertarget{eq-r-vec_split}{}{
\underline{r}_t = \begin{bmatrix}
\underline{r}_t^{N} \\
\underline{r}_t^{M}
\end{bmatrix},
}\label{eq-r-vec_split}\end{equation} i.e., the vector containing the
neutral lineage barcode counts \(\underline{r}_t^{N}\) and the
corresponding vector containing the mutant barcode counts
\(\underline{r}_t^{M}\). Following the same logic, matrix
\(\underline{\underline{R}}\) can be split into two matrices as
\begin{equation}\protect\hypertarget{eq-R-mat_split}{}{
\underline{\underline{R}} = \left[ 
\underline{\underline{R}}^N \; \underline{\underline{R}}^M
\right],
}\label{eq-R-mat_split}\end{equation} where
\(\underline{\underline{R}}^N\) is a \(T \times N\) matrix with the
barcode reads time series for each neutral lineage and
\(\underline{\underline{R}}^M\) is the equivalent \(T \times M\) matrix
for the non-neutral lineages.

Our objective is to compute the joint probability distribution for all
relative fitness values given our data. We can express this joint
posterior distribution using Bayes theorem as
\begin{equation}\protect\hypertarget{eq-bayes_obj}{}{
\pi(\underline{s}^M \mid \underline{\underline{R}}) = \frac{
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M)}
{\pi(\underline{\underline{R}})},
}\label{eq-bayes_obj}\end{equation} where hereafter \(\pi(\cdot)\)
defines a probability density function. When defining our statistical
model, we need not to focus on the denominator on the right-hand side of
Equation~\ref{eq-bayes_obj}. Thus, we can write
\begin{equation}\protect\hypertarget{eq-bayes_obj_propto}{}{
\pi(\underline{s}^M \mid \underline{\underline{R}}) \propto
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M).
}\label{eq-bayes_obj_propto}\end{equation} However, when implementing
the model computationally, the normalization constant on the right-hand
side of Equation~\ref{eq-bayes_obj} must be computed. This can be done
from the definition of the model via an integral of the form
\begin{equation}\protect\hypertarget{eq-bayes_evidence}{}{
\pi(\underline{\underline{R}}) = \int d^M \underline{s}^M
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M),
}\label{eq-bayes_evidence}\end{equation} also known as a marginalization
integral. Hereafter, differentials of the form \(d^n\) imply a
\(n\)-dimensional integral.

Although Equation~\ref{eq-bayes_obj} and
Equation~\ref{eq-bayes_obj_propto} seem simple enough, recall that
Equation~\ref{eq-fitness} relates barcode frequency values and the
population mean fitness to the mutant relative fitness. Therefore, we
must include these nuisance parameters as part of our inference problem.
We direct the reader to the supplementary materials for the exact
definitions of these parameters. Here, it suffices to say that the
inference problem must include the vector \(\underline{\bar{s}}_T\) of
all population mean fitness values and the matrix
\(\underline{\underline{F}}\) of all barcode frequencies within the
sequencing data. With these nuisance variables in hand, the full
inference problem we must solve takes the form
\begin{equation}\protect\hypertarget{eq-bayes_full}{}{
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) \propto
\pi(
    \underline{\underline{R}} \mid
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
)
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
).
}\label{eq-bayes_full}\end{equation} To recover the marginal
distribution over the non-neutral barcodes relative fitness values, we
can numerically integrate out all nuisance parameters, i.e.,
\begin{equation}\protect\hypertarget{eq-bayes_full_marginal}{}{
\pi(\underline{s}^M \mid \underline{\underline{R}}) =
\int d^{T-1}\underline{\bar{s}}_T
\int d^{B}\underline{f}_1 \cdots
\int d^{B}\underline{f}_T
\;
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
).
}\label{eq-bayes_full_marginal}\end{equation}

\hypertarget{seq-split_posterior}{%
\subsubsection{Factorizing the posterior
distribution}\label{seq-split_posterior}}

The left-hand side of Equation~\ref{eq-bayes_full} is extremely
difficult to work with. However, we can take advantage of the structure
of our inference problem to rewrite it in a more manageable form.
Specifically, the statistical dependencies of our observations and
latent variables allow us to factorize the joint distribution into the
product of multiple conditional distributions. To gain some intuition
about this factorization, let us focus on the inference of the
population mean fitness values \(\underline{\bar{s}}_T\).
Equation~\ref{eq-logfreq_neutral} relates the value of the population
mean fitness to the neutral lineage frequencies and nothing else. This
suggests that when writing the posterior for these population mean
fitness parameters, we should be able to condition it only on the
neutral lineage frequency values, i.e.,
\(\pi(\underline{\bar{s}}_T \mid \underline{\underline{F}}^N)\). We
point the reader to Section~\ref{sec-bayes_def} for the full
mathematical details on this factorization. For our purpose here, it
suffices to say we can rewrite the joint probability distribution as a
product of conditional distributions of the form
\begin{equation}\protect\hypertarget{eq-split_posterior}{}{
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) =
\pi(
    \underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}}^M
)
\pi(
    \underline{\bar{s}}_T \mid \underline{\underline{F}}^N
)
\pi(\underline{\underline{F}} \mid \underline{\underline{R}}).
}\label{eq-split_posterior}\end{equation} Written in this form,
Equation~\ref{eq-split_posterior} captures the three sources of
uncertainty listed in Section~\ref{sec-fitness_model} in each term.
Starting from right to left, the first term on the right-hand side of
Equation~\ref{eq-split_posterior} accounts for the uncertainty when
inferring the frequency values given the barcode reads. The second term
accounts for the uncertainty in the values of the mean population
fitness at different time points. The last term accounts for the
uncertainty in the parameter we care about---the mutants' relative
fitnesses. We refer the reader to Section~\ref{sec-bayes_def} for an
extended description of the model with specific functional forms for
each term on the left-hand side of Equation~\ref{eq-split_posterior} as
well as the extension of the model to account for multiple experimental
replicates or hierarchical genotypes.

\hypertarget{variational-inference}{%
\subsubsection{Variational Inference}\label{variational-inference}}

One of the technical challenges to the adoption of Bayesian methods is
the analytical intractability of integrals such as that of
Equation~\ref{eq-bayes_full_marginal}. Furthermore, even though
efficient Markov Chain Monte Carlo (MCMC) algorithms such as Hamiltonian
Montecarlo can numerically perform this integration
\autocite{betancourt2017}, the dimensionality of the problem in
Equation~\ref{eq-split_posterior} makes an MCMC-based approach
prohibitively slow.

To overcome this computational limitation, we rely on the recent
development of the automatic differentiation variational inference
algorithm (ADVI) \autocite{kucukelbir2016}. Briefly, when performing
ADVI, our target posterior distribution
\(\pi(\theta \mid \underline{\underline{R}})\), where
\(\theta = (\underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}})\),
is replaced by an approximate posterior distribution \(q_\phi(\theta)\),
where \(\phi\) fully parametrizes the approximate distribution. As
further explained in Section~\ref{sec-vi_primer}, the numerical
integration problem is replaced by an optimization problem of the form
\begin{equation}\protect\hypertarget{eq-vi_objective}{}{
q^*_\phi(\theta) = \min _\phi
D_{KL}(
    q_\phi(\theta) \lvert \lvert
    \pi(\theta \mid \underline{\underline{R}})
),
}\label{eq-vi_objective}\end{equation} where \(D_{KL}\) is the
Kulback-Leibler divergence. In other words, the complicated
high-dimensional numerical integration problem is transformed into a
much simpler problem of finding the value of the parameters \(\phi\)
such that Equation~\ref{eq-vi_objective} is satisfied as best as
possible within some finite computation time. Although to compute
Equation~\ref{eq-vi_objective}, we require the posterior distribution we
are trying to approximate
\(\pi(\theta \mid \underline{\underline{R}})\), it can be shown that
maximizing the so-called evidence lower bound (ELBO)
\autocite{kingma2014}---equivalent to minimizing the variational free
energy \autocite{gottwald2020}---is mathematically equivalent to
performing the optimization prescribed by
Equation~\ref{eq-vi_objective}. We direct the reader to
Section~\ref{sec-vi_primer} for a short primer on variational inference.

This work is accompanied by the Julia library \texttt{BarBay.jl} that
makes use of the implementation of both MCMC-based integration as well
as ADVI optimization to numerically approximate the solution of
Equation~\ref{eq-bayes_full_marginal} within the Julia ecosystem
\autocite{ge2018}.

\hypertarget{inference-on-a-single-dataset}{%
\subsection{Inference on a single
dataset}\label{inference-on-a-single-dataset}}

To assess the inference pipeline performance, we applied it to a
simulated dataset with known ground truth relative fitness values (See
Section~\ref{sec-logistic} for details on simulation).
Figure~\ref{fig-02}(A) shows the structure of the synthetic dataset. The
majority of barcodes of interest (faint color lines) are adaptive
compared to the neutral barcodes (\(s^{(m)} > 0\)). Although the barcode
frequency trajectories look relatively smooth, our fitness model
requires the computation of the log frequency ratio between adjacent
time points as derived in Equation~\ref{eq-logfreq}.
Figure~\ref{fig-02}(B) shows such data transformation where we can
better appreciate the observational noise input into our statistical
model. This noise is evident for the darker lines representing the
neutral barcodes since all of these lineages are assumed to be
identically distributed.

To visualize the performance of our inference pipeline in fitting our
fitness model to the observed data, we compute the so-called posterior
predictive checks (PPC). In short, the PPC consists of repeatedly
generating synthetic datasets in agreement with the results from the
inference results. In other words, we use the resulting parameter values
from the ADVI inference to generate possible datasets in agreement with
the inferred values (See Section~\ref{sec-ppc} for further details on
these computations). Figure~\ref{fig-02}(C) shows these results for all
neutral lineages (upper left corner plot) and a few representative
non-neutral barcodes. The different color shades represent the 95\%,
68\%, and 5\% credible regions, i.e., the regions where we expect to
find the data with the corresponding probability---or in terms of our
parameter, the \(X\%\) credible region is the interval where we expect
the true parameter value to lie with \(X\%\) probability.

The main advantage of our method is this natural interpretability of
these credible regions where an \(X\%\) credible region indeed captures
the region of parameter space where we expect with \(X\%\) probability
the actual value of the parameter lies given our statistical model, our
prior information, and the observed experimental data. A common mistake
in the literature is interpreting frequentist confidence intervals as
Bayesian credible regions when they are not equivalent
\autocite{morey2016}. Frequentist confidence intervals and Bayesian
credible regions are based on fundamentally different philosophical
approaches to statistics. Frequentist confidence intervals represent the
range of values that would contain the true population parameter with a
certain probability if the experiment was repeated many times. The
confidence interval does not represent the probability that the interval
contains the true value. According to a specific model and prior
information, Bayesian credible regions represent the range of values
that contain the parameter with a certain posterior probability. The
credible region directly represents the probability that the region
contains the true value. So, frequentist confidence intervals cannot be
interpreted as Bayesian credible regions because they have fundamentally
different meanings. Treating an \(X\%\) confidence interval like an
\(X\%\) credible region is fallacious since confidence intervals do not
represent probabilistic coverage of the true value like credible
regions. The intervals are generated through entirely different
procedures.

To capture the global performance of the model, Figure~\ref{fig-02}(D)
compares the known ground truth with the inferred relative fitness value
for all barcodes of interest. There is an excellent degree of
correspondence between these values, with the error bars representing
the 68\% credible region for the parameter value crossing the identity
line for most barcodes. This latter point is made clear with
Figure~\ref{fig-02}(E) where \(\approx 90\%\) of ground truth fitness
values fall within one standard deviation of the mean in the inferred
posterior distributions.

\begin{figure}

{\centering \includegraphics{./figs/fig02.pdf}

}

\caption{\label{fig-02}\textbf{Single dataset inference}. (A) Frequency
trajectories that represent the raw data going into the inference. (B)
Log frequency ratio between two adjacent time points used by the
inference pipeline. Darker lines represent the neutral barcodes. These
transformed data are much more noisy than the seemingly smooth frequency
trajectories. (C) Examples of the posterior predictive checks for all
neutral lineages (upper left panel) and a subset of representative
mutant lineages. Shaded regions represent the 95\%, 68\%, and 5\%
credible regions for the data. The reported errors above the plot
represent the 68\% credible region on the mutant relative fitness
marginal distribution. (D) Comparison between the ground truth fitness
value from the logistic-growth simulation and the inferred fitness
value. Gray error bars represent the 68\% posterior credible region for
the relative fitness values. (E) The empirical cumulative distribution
function (ECDF) for the absolute z-score value of the ground truth
parameter value within the inferred fitness posterior distribution.}

\end{figure}

\hypertarget{sec-multienv}{%
\subsection{Fitness inference on multiple
environments}\label{sec-multienv}}

The fitness model in Equation~\ref{eq-fitness} relates nuisance
parameters such as the population mean fitness and the barcode
frequencies to the relative fitness parameter we want to infer from the
data. These dependencies imply that uncertainty on the estimates of
these nuisance parameters influences the inference of the relevant
parameters. For example, imagine a scenario where the neutral lineages
data were incredibly noisy, leading to poor estimates of the population
mean fitness values \(\underline{\bar{s}}_T\). Since the relative
fitness of any non-neutral barcode \(s^{(m)}\) is determined with
respect to these neutral barcodes, not accounting for the lack of
precision in the value of the population mean fitness would result in
misleading estimates of the accuracy with which we determine the value
of the parameter we care about. Thus, propagating these sources of
uncertainty in nuisance parameters is vital to generate an unbiased
estimate of the relevant information we want to extract from the data.
One of the benefits of Bayesian methods is the intrinsic error
propagation embedded in the mathematical framework. For our previous
example, the uncertainty on the value of the population mean fitness
values is propagated to the relative fitness of a non-neutral barcode
since we defined a joint posterior distribution over all parameters as
fully expressed in Equation~\ref{eq-bayes_full}.

This natural error propagation can help us with the experimental design
schematized in Figure~\ref{fig-03}(A). Here, rather than performing
growth-dilution cycles in the same environment, the cells are diluted
into a different environment. Thus, the uncertainty on the fitness
estimate for the previous environment must be propagated to that of the
next one. To validate the extension of our statistical model to this
scenario, Figure~\ref{fig-03}(B) shows the trajectory of the log
frequency ratios between adjacent time points. The different colored
regions correspond to the different environments. For this simulation,
the growth rate of Environment 2 was set to be, on average, half of the
average growth rate in Environment 1. Equivalently, the growth rate in
Environment 3 was set to be, on average, twice the average growth rate
in Environment 1. Figure~\ref{fig-03}(C-E) show the correspondence
between the simulation ground truth and the inferred fitness values,
where the error bars represent the 68\% credible region.
Figure~\ref{fig-03}(F) summarizes the performance of our inference
pipeline by showing the empirical cumulative distribution functions for
the absolute value of the ground truth fitness value z-score within the
posterior distribution. This plot shows that, overall, \(\approx 75\%\)
of inferred mean values fall within one standard deviation of the ground
truth. For completeness, Figure~\ref{fig-03}(G) shows the posterior
predictive checks for a few example barcodes.

\begin{figure}

{\centering \includegraphics{./figs/fig03.pdf}

}

\caption{\label{fig-03}\textbf{Multi-environment fitness inference.} (A)
Schematic of the simulated experimental design where growth-dilution
cycles are performed into different environments for each cycle. (B) log
frequency rations between adjacent time points. Darker lines represent
the neutral barcodes. The colors in the background demark the
corresponding environment, matching colors in (A). Environment 2 is set
to have, on average, half the growth rate of environment 1. Likewise,
environment 3 is set to have, on average, twice the growth rate of
environment 1. (C-E) Comparison between the ground truth fitness value
from the logistic-growth simulation and the inferred fitness value for
each environment. Gray error bars represent the 68\% posterior credible
region. (F) The empirical cumulative distribution function (ECDF) for
the absolute z-score value of the ground truth parameter value within
the inferred fitness posterior distribution for all fitness values
(black line) and each environment individually (color lines). (G)
Examples of the posterior predictive checks for all neutral lineages
(upper left panel) and a subset of representative mutant lineages.
Shaded regions surrounding the data represent the 95\%, 68\%, and 5\%
credible regions for the data. The reported errors above the plot
represent the 68\% credible region on the mutant relative fitness
marginal distribution. Background colors match those of (A).}

\end{figure}

\hypertarget{sec-replicates}{%
\subsection{Accounting for experimental replicates via hierarchical
models}\label{sec-replicates}}

Our inference pipeline can be extended to account for multiple
experimental replicates via Bayesian hierarchical models
\autocite{betancourt2013}. Briefly, when accounting for multiple
repeated measurements of the same phenomena, there are two extreme cases
one can use to perform the data analysis: On the one hand, we can treat
each measurement as entirely independent, losing the power to utilize
multiple measurements when trying to learn a single parameter. This can
negatively impact the inference since, in principle, the value of our
parameter of interest should not depend on the particular experimental
replicate in question. However, this approach does not allow us to
properly ``combine'' the uncertainties in both experiments when
performing the inference. On the other hand, we can pool all data
together and treat our different experiments as a single measurement
with higher coverage. This loses the subtle differences due to biotic
and abiotic batch effects, effectively halving the data that goes into
our inference problem.

Hierarchical models present a middle ground between these extremes.
First, hierarchical models rely on the definition of so-called
\emph{hyper-parameters}, that capture the parametric inference we are
interested in---for this inference problem, we have a hyper-fitness
value \(\theta^{(m)}\) for each non-neutral barcode. Second, each
experiment draws randomly from the distribution of this hyper-parameter,
allowing for subtle variability between experiments to be accounted
for---in the present inference pipeline, each experimental replicate
gets assigned a \emph{local} fitness value \(s^{(m)}_i\), where the
extra sub-index indicates the \(i\)-th experimental replicate.
Conceptually, we can think of the local fitness for replicate \(i\) as
being sampled from a distribution that depends on the value of the
global hyper-fitness value, i.e., \(s^{(m)}_i \sim \pi_{\theta^{(m)}}\),
where the subindex \(\theta^{(m)}\) indicates the distribution's
parametric dependence on the hyper-fitness value. This way of
interpreting the connection between the distribution
\(\pi_{\theta^{(m)}}\) and the local fitness implies that a large
replicate-to-replicate variability would lead to a broad hyper-fitness
distribution---implying a large uncertainty when determining the
parameter that characterizes the overall relative fitness. We point the
reader to Section~\ref{sec-hierarchical_model} for the full definition
of the hierarchical model used in this section. Importantly, as
schematized in Figure~\ref{fig-04}(A), the influence between different
experimental replicates runs both ways. First, the data from one
experimental replicate (\(\underline{\underline{R}}^M_k\) in the
diagram) informs all local fitness values via the global hyper-fitness
(upper panel in Figure~\ref{fig-04}(A)). Second, the local fitness value
is informed by the data from all experimental replicates via the same
global hyper-fitness parameter (lower panel in Figure~\ref{fig-04}(A)).

To test the performance of this model, we simulated two experimental
replicates with 1000 unique barcodes (see Figure~\ref{fig-04}(B-C))
where we randomly sampled a ground truth hyper-fitness value
\(\theta^{(m)}\) for each barcode. We sampled a variation from this
hyper-fitness value for each experimental replicate \(s^{(m)}_i\) to
capture experimental batch effects. Figure~\ref{fig-04}(D) shows the
relationship between hyper-fitness and replicate fitness values for this
simulation. The spread around the identity line represents the expected
batch-to-batch variation. The posterior predictive checks examples in
Figure~\ref{fig-04}(E) show that the hierarchical model can correctly
fit the data for each experimental replicate. Furthermore,
Figure~\ref{fig-04}(F-G) show a high correlation between the ground
truth and the inferred fitness values. The empirical cumulative
distribution functions shown in Figure~\ref{fig-04}(H-I) reveal that for
\(\approx 75\%\) of the non-neutral barcodes, the ground truth
hyper-fitness values fall within one standard deviation from the mean
value in the posterior distributions.

\begin{figure}

{\centering \includegraphics{./figs/fig04.pdf}

}

\caption{\label{fig-04}\textbf{Hierarchical model on experimental
replicates}. (A) Schematic depiction of the interactions between local
fitness values \(\underline{s}_k^M\) through the global hyper-fitness
value \(\underline{\theta}^M\) for \(K\) hypothetical experimental
replicates. The upper diagram shows how the data from replicate \(k\)
informs all local fitness values via the hyper-fitness parameter. The
lower panel shows the reverse, where all other datasets inform the local
fitness value. (B-C) Simulated replicate datasets with 900 barcodes of
interest and 100 neutral lineages. (D) Comparison between the simulation
ground truth hyper-fitness and each replicate ground truth fitness. The
scatter between parameters captures experimental batch effects. (E)
Examples of the posterior predictive checks for all neutral lineages
(upper left panels) and a subset of representative mutant lineages.
Shaded regions from light to dark represent the 95\%, 68\%, and 5\%
credible regions. (F-G) Comparison between the simulation's ground truth
hyper-fitness (F) and replicate fitness (G) values and the inferred
parameters. Gray error bars represent the 68\% posterior credible
region. (H-I) The empirical cumulative distribution function (ECDF) for
the absolute z-score value of the ground truth parameter value within
the inferred hyper-fitness posterior distribution (H) and replicate
fitness (I).}

\end{figure}

As shown in Figure~\ref{fig-05}, the structure imposed by the
hierarchical model schematized in Figure~\ref{fig-04}(A), where we
explicitly account for the connection between experimental replicates
can improve the quality of the inference. Inferred fitness values
between experimental replicates exhibit a stronger degree of correlation
in the hierarchical model (Figure~\ref{fig-05}(A)) compared to
conducting inference on each replicate independently
(Figure~\ref{fig-05}(B)). Moreover, when comparing the inferred
hyper-fitness values---the objective parameter when performing multiple
experimental measurements---the hierarchical model outperforms averaging
the independent experimental replicates as shown in
Figure~\ref{fig-05}(C) and (D).

\begin{figure}

{\centering \includegraphics{./figs/fig05.pdf}

}

\caption{\label{fig-05}\textbf{Comparison between hierarchical model and
single dataset model}. (A-B) comparison of inferred fitness values
between experimental replicates when fitting a hierarchical model (A) or
independently fitting each dataset (B). Gray error bars represent the
68\% credible regions. (C) Comparison between the ground truth
hyper-fitness value and the inferred parameters. The blue dots show the
inferred hyper-fitness values when assuming a hierarchical model. Gray
error bars show the 68\% credible region for this inference. The yellow
dots show the average of the mean inferred fitness values for the two
experimental replicates. No error bars are shown for these, as it is
inappropriate to compute one with two data points per non-neutral
barcode. (D) Empirical cumulative distribution function (ECDF) of the
absolute difference between the inferred mean and the ground truth
hyper-fitness.}

\end{figure}

\hypertarget{sec-genotypes}{%
\subsection{Accounting for multiple barcodes per genotype via
hierarchical models}\label{sec-genotypes}}

Hierarchical models can also capture another experimental design in
which multiple barcodes map to the same or an equivalent genotype. As we
will show, this many-to-one mapping can improve the inference compared
to the extreme cases of inferring the fitness of each barcode
independently or pooling the data of all barcodes mapping to a single
genotype. As schematized in Figure~\ref{fig-06}(A), a small modification
of the base model allows us to map the structure of our original model
to that of a hierarchical model with a fitness hyperparameter vector
\(\underline{\theta}^G\), where \(G\) is the number of genotypes in the
dataset.

Figure~\ref{fig-06}(B) shows a single experimental replicate in which 90
genotypes were assigned a random number of barcodes (a multinomial
distribution with a mean of ten barcodes per genotype) for a total of
900 non-neutral barcodes. To assess the performance of the hierarchical
model proposed in Figure~\ref{fig-06}(A), we performed inference using
this hierarchical model, as well as the two extreme cases of ignoring
the connection between the barcodes belonging to the same
genotype---equivalent to performing inference using the model presented
in Figure~\ref{fig-02}(A) over the barcodes---or pooling the data of all
barcodes belonging to the same genotype into a single count---equivalent
to performing inference using the model presented in
Figure~\ref{fig-02}(A) over the pooled barcodes.
Figure~\ref{fig-06}(C-D) shows the comparison between the simulation
ground truth and the inferred values for these three cases. Not only do
the hierarchical model results show higher degrees of correlation with
the ground truth, but the error bars (representing the 68\% credible
regions) are smaller, meaning that the uncertainty in the estimate of
the parameter we care about decreases when using the hierarchical model.
The improvement in the prediction can be seen in Figure~\ref{fig-06}(F)
where the empirical cumulative distribution function of the absolute
difference between the mean inferred value and the simulation ground
truth is shown for all three inference models. The hierarchical model's
curve ascends more rapidly, showing that, in general, the inferred
values are closer to the ground truth. For completeness,
Figure~\ref{fig-06}(G) shows some examples of how the hierarchical model
can capture the raw log-frequency count observations.

\begin{figure}

{\centering \includegraphics{./figs/fig06.pdf}

}

\caption{\label{fig-06}\textbf{Hierarchical model for multiple barcodes
per genotype.} (A) Schematic depiction of the hierarchical structure for
multiple barcodes mapping to a single genotype. A set of barcodes
mapping to an equivalent genotype map to ``local'' fitness values
\(s^{(b)}\) that are connected via a hyper-fitness parameter for the
genotype \(\theta^{(g)}\). (B) Simulated dataset with 100 neutral
lineages and 900 barcodes of interest distributed among 90 genotypes.
(C-E) Comparison between the inferred and ground truth fitness values
for a hierarchical model (C), a model where each barcode is inferred
independently (D), and a model where barcodes mapping to the same
genotype are pooled together (E). Gray error bars represent the 68\%
credible regions. (F) Empirical cumulative distribution function (ECDF)
of the absolute difference between the inferred mean and the ground
truth fitness values for all three models. (G) Examples of the posterior
predictive checks for all neutral lineages (upper left panels) and a
subset of representative mutant lineages. Shaded regions from light to
dark represent the 95\%, 68\%, and 5\% credible regions.}

\end{figure}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

Experimental evolution of microbial systems has dramatically advanced
our understanding of the basic principles of biological evolution
\autocite{kussell2013}. From questions related to the optimal
fine-tuning of gene expression programs \autocite{Dekel2005}, to the
dimensionality, geometry, and accessibility of the adaptive fitness
landscape explored by these rapidly adapting populations
\autocite{kinsler2020,maeda2020}, to the emergence of eco-evolutionary
dynamics in a long-term evolution experiment \autocite{good2017}; for
all of these and other cases, the microbial experimental platform
combined with high-throughput sequencing has been essential to tackling
these questions with empirical data. This exciting research area
promises to improve as new culturing technologies \autocite{jagdish2022}
as well as more complex lineage barcoding schemes
\autocite{nguyenba2019a,yang2022}, are adopted.

For this data-heavy field, properly accounting for the uncertainty in
parameters inferred from experiments is vital to ensure the conclusions
drawn are reliable. Bayesian statistics presents a principled way to
quantify this uncertainty systematically \autocite{gelman2010}.
Moreover, Bayesian analysis offers a more natural way to interpret the
role that probability theory plays when performing data analysis
compared to the often-misinterpreted frequentist methods
\autocite{vanderplas2014}. Nevertheless, the technical challenges
associated with Bayesian analysis has limited its application. This is
set to change as recognition of the misuse of frequentist concepts such
as the p-value is receiving more attention \autocite{nuzzo2014}.
Moreover, advances in numerical methods such as Hamiltonian Monte Carlo
\autocite{betancourt2017} and variational inference
\autocite{kucukelbir2016} allows for complex Bayesian models to be fit
to empirical data.

In this paper, we present a computational pipeline to analyze
lineage-tracking time-series data for massive-parallel competition
assays. More specifically, we fit a Bayesian model to infer the fitness
of multiple genotypes relative to a reference
\autocite{kinsler2020,ascensao2023}. The proposed model accounts for
multiple sources of uncertainty with proper error propagation intrinsic
to Bayesian methods. To scale the inference pipeline to large datasets
with \(> 10,000\) barcodes, we use the ADVI algorithm
\autocite{kucukelbir2016} to fit a variational posterior distribution.
The main difference between our method and previous inference pipelines,
such as \textcite{li2023}, is that the present analysis provides
interpretable errors on the inferred fitness values. The reported
uncertainty intervals---known as credible regions---can be formally
interpreted as capturing the corresponding probability mass of finding
the true value of the parameter given the model, the prior information,
and the data. Furthermore, minor modifications to the structure of the
statistical model presented in this work allow for the analysis of
different experimental designs, such as growth-dilution cycles in
different environments, joint analysis of multiple experimental
replicates of the same experiment via hierarchical models, and a
hierarchical model for multiple barcodes mapping to equivalent
genotypes. We validate our analysis pipeline on simulated datasets with
known ground truth, showing that the model fits the data adequately,
capturing the ground truth parameters within the posterior distribution.

It is important to highlight some of the consequences of the general
experimental design and the implicit assumptions within the proposed
statistical model to analyze the resulting data. First, the composition
of the population is such that the initial fraction of the population
occupied by the barcoded genotypes is small---usually \textgreater90\%
of the initial population is the non-labeled reference strain. This
constraint is important as the fitness model used to fit the time series
data assumes that the tracked frequencies are \(\ll 1\). Second, when
computing log frequency ratios, we can run into the issue of dividing by
zero. This is a common problem when dealing with molecular count data
\autocite{lovell2020}. Our model gets around this issue by assuming that
the frequency of any barcode cannot be, but still can get arbitrarily
close to, zero. Therefore, we implicitly assume that no lineage goes
extinct during the experiment. Moreover, the statistical model directly
accounts for the uncertainty associated with having zero barcode counts,
increasing the corresponding uncertainty. Third, the models presented in
this paper require the existence of a labeled sub-population of barcoded
reference strains. These barcodes help determine the fitness baseline,
as every fitness is quantified with respect to this reference genotype.
This experimental design constraint facilitates the inference of the
population mean fitness since most of the culture---the unlabeled
reference genotype---is not tracked. Finally, the presented statistical
model assumes that relative fitness is solely a constant of the
environment and the genotype. Future directions of this work could
extend the fitness model to properly analyze data with time-varying or
frequency-dependent fitness values.

In total, the statistical model presented in this work and the software
package accompanying the paper allow for a principled way of quantifying
the accuracy with which we can extract relevant parametric information
from large-scale multiplexed fitness competition assays. Furthermore,
the implementation of Bayesian models and their fitting via automatic
differentiation approaches opens the gate to extend this type of formal
analysis to the data-rich literature in experimental evolution and other
high-throughput technologies applications.

\hypertarget{supplementary-materials}{%
\section{Supplementary Materials}\label{supplementary-materials}}

\hypertarget{sec-vi_primer}{%
\subsection{Primer on Variational Inference}\label{sec-vi_primer}}

In this section, we will briefly introduce the idea behind variational
inference. Recall that any Bayesian inference problem deals with the
joint distribution between observations \(\underline{x}\) and unobserved
latent variables \(\underline{\theta}\). This joint distribution can be
written as the product of a distribution of the observations
\(\underline{x}\) conditioned on the \(\underline{\theta}\) and the
marginal distribution of these latent variables, i.e.,
\begin{equation}\protect\hypertarget{eq-vi_joint_dist}{}{
\pi(\underline{x}, \underline{\theta}) =
\pi(\underline{x} \mid \underline{\theta}) \pi(\underline{\theta}).
}\label{eq-vi_joint_dist}\end{equation} A Bayesian inference pipeline's
objective is to compute the latent variables' posterior probability
given a set of observations. This computation is equivalent to updating
our prior beliefs about the set of values that the latent variables take
after taking in new data. We write this as Bayes theorem
\begin{equation}\protect\hypertarget{eq-vi_bayes_theorem}{}{
\pi(\underline{\theta} \mid \underline{x}) = 
\frac{
        \pi(\underline{x} \mid \underline{\theta})\pi(\underline{\theta})
    }{
        \pi(\underline{x})
    }.
}\label{eq-vi_bayes_theorem}\end{equation} The main technical challenge
for working with Equation~\ref{eq-vi_bayes_theorem} comes from the
computation of the denominator, also known as the \emph{evidence} or the
\emph{marginalized likelihood}. The reason computing this term is
challenging is because it involves a (potentially) high-dimensional
integral of the form
\begin{equation}\protect\hypertarget{eq-vi_evidence_int}{}{
\pi(\underline{x}) = 
\int\cdots\int d^K\underline{\theta}\; \pi(\underline{x}, \underline{\theta}) = 
\int\cdots\int d^K\underline{\theta}\; 
\pi(\underline{x} \mid \underline{\theta})
\pi(\underline{\theta}),
}\label{eq-vi_evidence_int}\end{equation} where \(K\) is the
dimesionality of the \(\underline{\theta}\) vector. Here, the integrals
are taken over the support---the set of values valid for the
distribution---of \(\pi(\underline{\theta})\). However, only a few
selected distributions have a closed analytical form; thus, in most
cases Equation~\ref{eq-vi_evidence_int} must be solved numerically.

Integration in high-dimensional spaces can be computationally extremely
challenging. For a naive numerical quadrature procedure, integrating
over a grid of values for each dimension of \(\underline{\theta}\) comes
with an exponential explosion of the number of required grid point
evaluations, most of which do not contribute significantly to the
integration. To gain visual intuition about this challenge, imagine
integrating the function depicted in Figure~\ref{fig-SI01}. If the
location of the high-density region (dark peak) is unknown, numerical
quadrature requires many grid points to ensure we capture this peak.
However, most of the numerical evaluations of the function on the grid
points do not contribute significantly to the integral. Therefore, our
computational resources are wasted on insignificant evaluations. This
only gets worse as the number of dimensions increases since the number
of grid point evaluation scales exponentially.

\begin{figure}

{\centering \includegraphics{./figs/figSIX_gaussian_peak.png}

}

\caption{\label{fig-SI01}\textbf{High-dimensional numerical quadrature
does not scale with dimensionality.} Schematic depiction of the problem
with naive numerical quadrature to integrate over an unknown density.
While the density is concentrated on the dark peak, most of the
evaluations over the \(x_1-x_2\) grid do not contribute to the value of
the integral}

\end{figure}

Modern Markov Chain Monte Carlo algorithms, such as Hamiltonian Monte
Carlo, can efficiently perform this high-dimensional integration by
utilizing gradient information from the target density
\textcite{betancourt2017}. Nevertheless, these sampling-based methods
become prohibitively slow for the number of dimensions our present
inference problem presents. Thus, there is a need to find scalable
methods for the inference problem in Equation~\ref{eq-vi_bayes_theorem}.

Variational inference circumvents these technical challenges by
proposing an approximate solution to the problem. Instead of working
with the posterior distribution in its full glory
\(\pi(\underline{\theta} \mid \underline{x})\), let us propose an
approximate posterior distribution \(q_\phi\) that belongs to a
distribution family fully parametrized by \(\phi\). For example, let us
say that the distribution \(q_\phi\) belongs to the family of
multivariate Normal distributions such that
\(\phi = (\underline{\mu}, \underline{\underline{\Sigma}})\), where
\(\underline{\mu}\) is the vector of means and
\(\underline{\underline{\Sigma}}\) is the covariance matrix. If we
replace \(\pi\) by \(q_\phi\), we want \(q_\phi\) to resemble the
original posterior as much as possible. Mathematically, this can be
expressed as minimizing a ``\emph{distance metric}''---the
Kullback-Leibler (KL) divergence, for example---between the
distributions. Note that we use quotation marks because, formally, the
KL divergence is not a distance metric since it is not symmetric.
Nevertheless, the variational objective is set to find a distribution
\(q_\phi^*\) such that
\begin{equation}\protect\hypertarget{eq-vi_objective}{}{
q_\phi^*(\underline{\theta}) =
\min_\phi D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right),
}\label{eq-vi_objective}\end{equation} where \(D_{KL}\) is the KL
divergence. Furthermore, we highlight that the KL divergence is a
strictly positive number, i.e.,
\begin{equation}\protect\hypertarget{eq-vi_kl_pos}{}{
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) \geq 0,
}\label{eq-vi_kl_pos}\end{equation} as this property will become
important later on.

At first sight, Equation~\ref{eq-vi_objective} does not improve the
situation but only introduces further technical complications. After
all, the definition of the KL divergence
\begin{equation}\protect\hypertarget{eq-vi_kl_div}{}{
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) \equiv 
\int \cdots \int d^K\underline{\theta}\;
q_\phi(\underline{\theta})
\ln \frac{
    q_\phi(\underline{\theta})
}{
    \pi(\underline{\theta} \mid \underline{x})
},
}\label{eq-vi_kl_div}\end{equation} includes the posterior distribution
\(\pi(\underline{\theta} \mid \underline{x})\) we are trying to get
around. However, let us manipulate Equation~\ref{eq-vi_kl_div} to beat
it to a more reasonable form. First, we can use the properties of the
logarithms to write
\begin{equation}\protect\hypertarget{eq-vi_step01}{}{
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) = 
\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln q_\phi(\underline{\theta}) -
\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \pi(\underline{\theta} \mid \underline{x}),
}\label{eq-vi_step01}\end{equation} where, for convenience, we write a
single integration sign (\(d^K\underline{\theta}\;\) still represents a
multi-dimensional differential). For the second term in
Equation~\ref{eq-vi_step01}, we can substitute the term inside the
logarithm using Equation~\ref{eq-vi_bayes_theorem}. This results in
\begin{equation}\protect\hypertarget{eq-vi_step02}{}{
\begin{aligned}
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) &= 
\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln q_\phi(\underline{\theta}) \\
&- \int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \left( 
    \frac{
        \pi(\underline{x} \mid \underline{\theta})\pi(\underline{\theta})
    }{
        \pi(\underline{x})
    }
\right).
\end{aligned}
}\label{eq-vi_step02}\end{equation} Again, using the properties of
logarithms, we can split Equation~\ref{eq-vi_step02}, obtaining
\begin{equation}\protect\hypertarget{eq-vi_step03}{}{
\begin{aligned}
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) &= 
\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln q_\phi(\underline{\theta}) \\
&-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \pi(\underline{x} \mid \underline{\theta}) \\
&-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \pi(\underline{\theta}) \\
&+\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \pi(\underline{x}).
\end{aligned}
}\label{eq-vi_step03}\end{equation} It is convenient to write
Equation~\ref{eq-vi_step03} as
\begin{equation}\protect\hypertarget{eq-vi_step04}{}{
\begin{aligned}
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) &= 
\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \frac{
    q_\phi(\underline{\theta})
    }{
        \pi(\underline{\theta})
    } \\
&-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \pi(\underline{x} \mid \underline{\theta}) \\
&+ \ln \pi(\underline{x}) 
\int d^K\underline{\theta}\; q_\phi(\underline{\theta}),
\end{aligned}
}\label{eq-vi_step04}\end{equation} where for the last term, we can take
\(\ln \pi(\underline{x})\) out of the integral since it does not depend
on \(\underline{\theta}\). Lastly, we utilize two properties:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The proposed approximate distribution must be normalized, i.e.,
  \begin{equation}\protect\hypertarget{eq-vi_q_norm}{}{
  \int d^K\underline{\theta}\; q_\phi(\underline{\theta}) = 1.
  }\label{eq-vi_q_norm}\end{equation}
\item
  The law of the unconscious statistician (LOTUS) establishes that for
  any probability density function, it must be true that
  \begin{equation}\protect\hypertarget{eq-vi_lotus}{}{
  \int d^K\underline{\theta}\; q_\phi(\underline{\theta})
  f(\underline{\theta}) = \left\langle 
   f(\underline{\theta}) 
  \right\rangle_{q_\phi},
  }\label{eq-vi_lotus}\end{equation} where
  \(\left\langle\cdot\right\rangle_{q_\phi}\) is the expected value over
  the \(q_\phi\) distribution.
\end{enumerate}

Using these two properties, the positivity constraint on the KL
divergence in Equation~\ref{eq-vi_kl_pos}, and the definition of the KL
divergence in Equation~\ref{eq-vi_kl_div} we can rewrite
Equation~\ref{eq-vi_step04} as
\begin{equation}\protect\hypertarget{eq-vi_step05}{}{
D_{KL}\left( 
    q_\phi(\underline{\theta}) \vert \vert
    \pi(\underline{\theta}) 
\right) -
\left\langle
    \ln \pi(\underline{x} \mid \underline{\theta})
\right\rangle_{q_\phi}
\geq - \ln \pi(\underline{x}).
}\label{eq-vi_step05}\end{equation} Multiplying by a minus one, we have
the functional form of the so-called evidence lower bound (ELBO)
\textcite{kingma2014},
\begin{equation}\protect\hypertarget{eq-vi_elbo}{}{
\underbrace{
    \ln \pi(\underline{x})
}_{\text{log evidence}} \geq
\underbrace{
    \left\langle
        \ln \pi(\underline{x} \mid \underline{\theta})
    \right\rangle_{q_\phi} -
    D_{KL}\left( 
        q_\phi(\underline{\theta}) \vert \vert
        \pi(\underline{\theta}) 
    \right)
}_{\text{ELBO}}.
}\label{eq-vi_elbo}\end{equation}

Let us recapitulate where we are. We started by presenting the challenge
of working with Bayes' theorem, as it requires a high-dimensional
integral of the form in Equation~\ref{eq-vi_evidence_int}. As an
alternative, variational inference posits to approximate the posterior
distribution \(\pi(\underline{\theta} \mid \underline{x})\) with a
parametric distribution \(q_\phi(\underline{\theta})\). By minimizing
the KL divergence between these distributions, we arrive at the result
in Equation~\ref{eq-vi_elbo}, where the left-hand side---the log
marginalized likelihood or log evidence---we cannot compute for
technical/computational reasons. However, the right-hand side is
composed of things we can easily evaluate. We can easily evaluate the
log-likelihood \(\ln \pi(\underline{x} \mid \underline{\theta})\) and
the KL divergence between our proposed approximate distribution
\(q_\phi(\underline{\theta})\) and the prior distribution
\(\pi(\underline{\theta})\). Moreover, we can compute the gradients of
these functions with respect to the parameters of our proposed
distribution. This last point implies that we can change the parameters
of the proposed distribution to maximize the ELBO. And, although we
cannot compute the left-hand side of Equation~\ref{eq-vi_elbo}, we know
that however large we make the ELBO, it will always be smaller than (or
equal) the log-marginal likelihood. Therefore, the larger we can make
the ELBO by modifying the parameters \(\phi\), the closer it gets to the
log-marginal likelihood, and, as a consequence, the better our proposed
distribution \(q_\phi(\underline{\theta})\) gets to the true posterior
distribution \(\pi(\underline{\theta} \mid \underline{x})\).

In this sense, variational inference turns the intractable numerical
integration problem to an optimization routine, for which there are
several algorithms available.

\hypertarget{advi-algorithm}{%
\subsubsection{ADVI algorithm}\label{advi-algorithm}}

To maximize the right-hand side of Equation~\ref{eq-vi_elbo}, the
Automatic Differentiation Variational Inference (ADVI) algorithm
developed in \autocite{kucukelbir2016} takes advantage of advances in
probabilistic programming languages to generate a robust method to
perform this optimization. Without going into the details of the
algorithm implementation, for our purposes, it suffices to say that we
define our joint distribution \(\pi(\underline{\theta}, \underline{x})\)
as the product defined in Equation~\ref{eq-vi_joint_dist}. ADVI then
proposes an approximate variational distribution \(q_\phi\) that can
either be a multivariate Normal distribution with a diagonal covariance
matrix, i.e., \begin{equation}\protect\hypertarget{eq-vi_meanfield}{}{
\phi = (\underline{\mu}, \underline{\underline{D}}),
}\label{eq-vi_meanfield}\end{equation} where
\(\underline{\underline{D}}\) is the identity matrix, with the diagonal
elements given by the vector of variances \(\underline{\sigma}^2\) for
each variable or a full-rank multivariate Normal distribution
\begin{equation}\protect\hypertarget{eq-vi_full}{}{
\phi = (\underline{\mu}, \underline{\underline{\Sigma}}).
}\label{eq-vi_full}\end{equation}

Then, the parameters are initialized in some value \(\phi_o\). These
parameters are iteratively updated by computing the gradient of the ELBO
(right-hand side of Equation~\ref{eq-vi_elbo}), hereafter defined as
\(\mathcal{L}\), with respect to the parameters,
\begin{equation}\protect\hypertarget{eq-vi_elbo_gradient}{}{
\nabla_\phi \mathcal{L} = \nabla_{\underline{\mu}} \mathcal{L} + 
\nabla_{\underline{\sigma}}\mathcal{L},
}\label{eq-vi_elbo_gradient}\end{equation} and then computing \[
\phi_{t+1} = \phi_{t} + \eta \nabla_\phi \mathcal{L},
\] where \(\eta\) defines the step size.

This short explanation behind the ADVI algorithm is intended only to
gain intuition on how the optimal variational distribution \(q_\phi\) be
computed. There are many nuances in the implementation of the ADVI
algorithm. We invite the reader to look at the original reference for
further details.

\hypertarget{sec-bayes_def}{%
\subsection{Defining the Bayesian model}\label{sec-bayes_def}}

In the main text, we specify the inference problem we must solve as
being of the form
\begin{equation}\protect\hypertarget{eq-full_inference_SI}{}{
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) \propto
\pi(
    \underline{\underline{R}} \mid
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
)
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
).
}\label{eq-full_inference_SI}\end{equation} Here, we briefly define the
missing nuisance parameters. Let
\begin{equation}\protect\hypertarget{eq-pop_fitness_vec}{}{
\underline{\bar{s}}_T = (\bar{s}_1, \bar{s}_2, \ldots, \bar{s}_{T-1})^\dagger,
}\label{eq-pop_fitness_vec}\end{equation} be the vector containing the
\(T-1\) population mean fitness we compute from the \(T\) time points
where measurements were taken. We have \(T-1\) since the value of any
\(\bar{s}_t\) requires cycle numbers \(t\) and \(t+1\). Furthermore, let
the matrix \(\underline{\underline{F}}\) be a \(T \times B\) matrix
containing all frequency values. As with Equation~\ref{eq-R-mat_split}
in the main text, we can split \(\underline{\underline{F}}\) into two
matrices of the form
\begin{equation}\protect\hypertarget{eq-F-mat_split}{}{
\underline{\underline{F}} = \left[ 
\underline{\underline{F}}^N \; \underline{\underline{F}}^M
\right],
}\label{eq-F-mat_split}\end{equation} to separate the corresponding
neutral and non-neutral barcode frequencies.

Let us now define each of the terms in Equation~\ref{eq-split_posterior}
described in Section~\ref{sec-bayesian_inference} of the main text. The
following sections will specify the functional form each of these terms
takes.

\hypertarget{sec-bayes_freq}{%
\subsubsection{\texorpdfstring{Frequency uncertainty
\(\pi(\underline{\underline{F}} \mid \underline{\underline{R}})\)}{Frequency uncertainty \textbackslash pi(\textbackslash underline\{\textbackslash underline\{F\}\} \textbackslash mid \textbackslash underline\{\textbackslash underline\{R\}\})}}\label{sec-bayes_freq}}

We begin with the probability of the frequency values given the raw
barcode reads. The first assumption is that the inference of the
frequency values for time \(t\) is independent of any other time.
Therefore, we can write the joint probability distribution as a product
of independent distributions of the form
\begin{equation}\protect\hypertarget{eq-freq_indep}{}{
\pi(\underline{\underline{F}} \mid \underline{\underline{R}}) =
\prod_{t=1}^T \pi(\underline{f}_t \mid \underline{r}_t),
}\label{eq-freq_indep}\end{equation} where \(\underline{f}_t\) and
\(\underline{r}_t\) are the \(t\)-th row of the matrix containing all of
the measurements for time \(t\). We imagine that when the barcode reads
are obtained via sequencing, the quantified number of reads is a Poisson
sample from the ``true'' underlying number of barcodes within the pool.
This translates to assuming that the number of reads for each barcode at
any time point \(r^{(b)}_t\) is an independent Poisson random variable,
i.e., \begin{equation}\protect\hypertarget{eq-freq_poisson_reads}{}{
r^{(b)}_t \sim \operatorname{Poiss}(\lambda^{(b)}_t),
}\label{eq-freq_poisson_reads}\end{equation} where the symbol
``\(\sim\)'' is read ``distributed as.'' Furthermore, for a Poisson
distribution, we have that
\begin{equation}\protect\hypertarget{eq-freq_poisson_lambda}{}{
\lambda^{(b)}_t = \left\langle r^{(b)}_t \right\rangle = 
\left\langle 
    \left( r^{(b)}_t - \left\langle r^{(b)}_t \right\rangle \right)^2
\right\rangle,
}\label{eq-freq_poisson_lambda}\end{equation} where
\(\left\langle \cdot \right\rangle\) is the expected value. In other
words the Poisson parameter is equal to the mean and variance of the
distribution. The Poisson distribution has the convenient property that
for two Poisson distributed random variables
\(X \sim \operatorname{Poiss}(\lambda_x)\) and
\(Y \sim \operatorname{Poiss}(\lambda_y)\), we have that
\begin{equation}\protect\hypertarget{eq-freq_additivity}{}{
Z \equiv X + Y \sim \operatorname{Poiss}(\lambda_x + \lambda_y).
}\label{eq-freq_additivity}\end{equation} This additivity allows us to
write the total number of reads at time \(t\) \(n_t\) also as a
Poisson-distributed random variable of the form
\begin{equation}\protect\hypertarget{eq-freq_poisson_total}{}{
n_t \sim \operatorname{Poiss}\left( \sum_{b=1}^B \lambda^{(b)}_t \right),
}\label{eq-freq_poisson_total}\end{equation} where the sum is taken over
all \(B\) barcodes.

If the total number of reads is given by
Equation~\ref{eq-freq_poisson_total}, the array with the number of reads
for each barcode at time \(t\), \(\underline{r}_t\) is then distributed
as \begin{equation}\protect\hypertarget{eq-freq_multinomial}{}{
\underline{r}_t \sim \operatorname{Multinomial}(n_t, \underline{f}_t),
}\label{eq-freq_multinomial}\end{equation} where each of the \(B\)
entries of the frequency vector \(\underline{f}_t\) is a function of the
\(\underline{\lambda}_t\) vector, given by
\begin{equation}\protect\hypertarget{eq-freq_freq_lambda}{}{
f_t^{(b)} \equiv f_t^{(b)}(\underline{\lambda}_t) = 
\frac{\lambda_t^{(b)}}{\sum_{b'=1}^B \lambda_t^{(b')}}.
}\label{eq-freq_freq_lambda}\end{equation} In other words, we can think
of the \(B\) barcode counts as independent Poisson samples or as a
single multinomial draw with a random number of total draws, \(n_t\),
and the frequency vector \(\underline{f}_t\) we are interested in.
Notice that Equation~\ref{eq-freq_freq_lambda} is a deterministic
function that connects the Poisson parameters to the frequencies.
Therefore, we have the equivalence that
\begin{equation}\protect\hypertarget{eq-freq_f_lambda_equiv}{}{
\pi(\underline{f}_t \mid \underline{r}_t) = 
\pi(\underline{\lambda}_t \mid \underline{r}_t),
}\label{eq-freq_f_lambda_equiv}\end{equation} meaning that the
uncertainty comes from the \(\underline{\lambda}_t\) vector. By Bayes
theorem, we therefore write
\begin{equation}\protect\hypertarget{eq-freq_lambda_bayes}{}{
\pi(\underline{\lambda}_t \mid n_t, \underline{r}_t) \propto
\pi(n_t, \underline{r}_t \mid \underline{\lambda}_t) \pi(\underline{\lambda}_t),
}\label{eq-freq_lambda_bayes}\end{equation} where we explicitly include
the dependence on \(n_t\). This does not affect the distribution or
brings more uncertainty because \(\underline{r}_t\) already contains all
the information to compute \(n_t\) since
\begin{equation}\protect\hypertarget{eq-freq_n_sum_r}{}{
n_t = \sum_{b=1}^B r_t^{(b)}.
}\label{eq-freq_n_sum_r}\end{equation} But adding the variable allows us
to factorize Equation~\ref{eq-freq_lambda_bayes} as
\begin{equation}\protect\hypertarget{eq-freq_lambda_bayes_factorized}{}{
\pi(\underline{\lambda}_t \mid n_t, \underline{r}_t) \propto
\pi(\underline{r}_t \mid n_t, \underline{\lambda}_t)
\pi(n_t \mid \underline{\lambda}_t)
\pi(\underline{\lambda}_t)
}\label{eq-freq_lambda_bayes_factorized}\end{equation} We then have
\begin{equation}\protect\hypertarget{eq-freq_r_bayes}{}{
\underline{r}_t \mid n_t, \underline{\lambda}_t \sim
\operatorname{Multinomial}(n_t, \underline{f}_t(\underline{\lambda}_t)).
}\label{eq-freq_r_bayes}\end{equation} Furthermore, we have \[
n_t \mid \underline{\lambda}_t \sim 
\operatorname{Poiss}\left(\sum_{b=1}^B \lambda_t^{(b)}\right).
\]\{\#eq=freq\_n\_bayes\} Finally, for our prior
\(\pi(\underline{\lambda}_t)\), we first assume each parameter is
independent, i.e., \[
\pi(\underline{\lambda}_t) = \prod_{b=1}^B \pi(\lambda_t^{(b)}).
\] A reasonable prior for each \(\lambda_t^{(b)}\) representing the
expected number of reads for barcode \(b\) should span several orders of
magnitude. Furthermore, we assume that no barcode in the dataset ever
goes extinct. Thus, no frequency can equal zero, facilitating the
computation of the log frequency ratios needed to infer the relative
fitness. The log-normal distribution satisfies these constraints;
therefore, for the prior, we assume
\begin{equation}\protect\hypertarget{eq-freq_lambda_prior}{}{
\lambda_t^{(b)} \sim 
\log\mathcal{N}(\mu_{\lambda_t^{(b)}}, \sigma_{\lambda_t^{(b)}}),
}\label{eq-freq_lambda_prior}\end{equation} with
\(\mu_{\lambda_t^{(b)}}, \sigma_{\lambda_t^{(b)}}\) as the user-defined
parameters that characterize the prior distribution.

\hypertarget{summary}{%
\paragraph{Summary}\label{summary}}

Putting all the pieces developed in this section together gives a term
for our inference of the form
\begin{equation}\protect\hypertarget{eq-freq_final_01}{}{
\pi(\underline{\underline{F}} \mid \underline{\underline{R}}) \propto
\prod_{t=1}^T\left\{
    \pi(\underline{r}_t \mid n_t, \underline{\lambda}_t)
    \pi(n_t \mid \underline{\lambda}_t)
    \left[ 
        \prod_{b=1}^B \pi(\lambda_t^{(b)})
    \right]
\right\}
}\label{eq-freq_final_01}\end{equation} where
\begin{equation}\protect\hypertarget{eq-freq_final_02}{}{
\underline{r}_t \mid n_t, \underline{\lambda}_t \sim
\operatorname{Multinomial}(n_t, \underline{f}_t(\underline{\lambda}_t)),
}\label{eq-freq_final_02}\end{equation}
\begin{equation}\protect\hypertarget{eq-freq_final_03}{}{
n_t \mid \underline{\lambda}_t \sim 
\operatorname{Poiss}\left(\sum_{b=1}^B \lambda_t^{(b)}\right).
}\label{eq-freq_final_03}\end{equation} and
\begin{equation}\protect\hypertarget{eq-freq_final_04}{}{
\lambda_t^{(b)} \sim 
\log\mathcal{N}(\mu_{\lambda_t^{(b)}}, \sigma_{\lambda_t^{(b)}}),
}\label{eq-freq_final_04}\end{equation}

\hypertarget{sec-bayes_meanfit}{%
\subsubsection{\texorpdfstring{Population mean fitness uncertainty
\(\pi(\underline{\bar{s}}_T \mid \underline{\underline{F}}, \underline{\underline{R}})\)}{Population mean fitness uncertainty \textbackslash pi(\textbackslash underline\{\textbackslash bar\{s\}\}\_T \textbackslash mid \textbackslash underline\{\textbackslash underline\{F\}\}, \textbackslash underline\{\textbackslash underline\{R\}\})}}\label{sec-bayes_meanfit}}

Next, we turn our attention to the problem of determining the population
mean fitnesses \(\underline{\bar{s}}_T\). First, we notice that our
fitness model in Equation~\ref{eq-fitness} does not include the value of
the raw reads. They enter the calculation indirectly through the
inference of the frequency values we developed in
Section~\ref{sec-bayes_freq}. This means that we can remove the
conditioning of the value of \(\underline{\bar{s}}_T\) on the number of
reads, obtaining a simpler probability function
\begin{equation}\protect\hypertarget{eq-meanfit_noreads}{}{
\pi(
    \underline{\bar{s}}_T \mid 
    \underline{\underline{F}}, \underline{\underline{R}}
) = 
\pi(
    \underline{\bar{s}}_T \mid 
    \underline{\underline{F}}
).
}\label{eq-meanfit_noreads}\end{equation} Moreover, our fitness model
does not directly explain how the population mean fitness evolves over
time. In other words, our model cannot explicitly compute the population
mean fitness at time \(t+1\) from the information we have about time
\(t\). Given this model limitation, we are led to assume that we must
infer each \(\bar{s}_t\) independently. Expressing this for our
inference results in
\begin{equation}\protect\hypertarget{eq-meanfit_indep}{}{
\pi(
    \underline{\bar{s}}_T \mid 
    \underline{\underline{F}}
) =
\prod_{t=1}^{T-1} \pi(\bar{s}_t \mid \underline{f}_t, \underline{f}_{t+1}),
}\label{eq-meanfit_indep}\end{equation} where we split our matrix
\(\underline{\underline{F}}\) for each time point and only kept the
conditioning on the relevant frequencies needed to compute the mean
fitness at time \(t\).

Although our fitness model in Equation~\ref{eq-fitness} also includes
the relative fitness \(s^{(m)}\), to infer the population mean fitness
we only utilize data from the neutral lineages that, by definition, have
a relative fitness \(s^{(n)} = 0\). Therefore, the conditioning on
Equation~\ref{eq-meanfit_indep} can be further simplified by only
keeping the frequencies of the neutral lineages, i.e.,
\begin{equation}\protect\hypertarget{eq-meanfit_indep_neutral}{}{
\pi(\bar{s}_t \mid \underline{f}_t, \underline{f}_{t+1}) =
\pi(\bar{s}_t \mid \underline{f}_t^N, \underline{f}_{t+1}^N).
}\label{eq-meanfit_indep_neutral}\end{equation}

Recall that in Section~\ref{sec-fitness_model} we emphasized that the
frequencies \(f_t^{(n)}\) do not represent the true frequency of a
particular lineage in the population but rather a ``normalized number of
cells.'' Therefore, it is safe to assume each of the \(N\) neutral
lineages' frequencies is changing independently. The correlation of how
increasing the frequency of one lineage will decrease the frequency of
others is already captured in the model presented in
Section~\ref{sec-bayes_freq}. Thus, we write
\begin{equation}\protect\hypertarget{eq-meanfit_indep_lineages}{}{
\pi(\bar{s}_t \mid \underline{f}_t^N, \underline{f}_{t+1}^N) =
\prod_{n=1}^N \pi(\bar{s}_t \mid f_t^{(n)}, f_{t+1}^{(n)}).
}\label{eq-meanfit_indep_lineages}\end{equation}

Now, we can focus on one of the terms on the right-hand side of
Equation~\ref{eq-meanfit_indep_lineages}. Writing Bayes theorem results
in \begin{equation}\protect\hypertarget{eq-meanfit_bayes}{}{
\pi(\bar{s}_t \mid f_t^{(n)}, f_{t+1}^{(n)}) \propto
\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t) \pi(\bar{s}_t).
}\label{eq-meanfit_bayes}\end{equation} Notice the likelihood defines
the joint distribution of neutral barcode frequencies conditioned on the
population mean fitness. However, rewriting our fitness model in
Equation~\ref{eq-fitness} for a neutral lineage to leave frequencies on
one side and fitness on the other results in
\begin{equation}\protect\hypertarget{eq-fitness_ratio_neutral}{}{
\frac{f_{t+1}^{(n)}}{f_t^{(n)}} = \mathrm{e}^{- \bar{s}_t\tau}.
}\label{eq-fitness_ratio_neutral}\end{equation}
Equation~\ref{eq-fitness_ratio_neutral} implies that our fitness model
only relates \textbf{the ratio} of frequencies and not the individual
values. To get around this complication, we define
\begin{equation}\protect\hypertarget{eq-gamma_def}{}{
\gamma_t^{(b)} \equiv \frac{f_{t+1}^{(b)}}{f_t^{(b)}},
}\label{eq-gamma_def}\end{equation} as the ratio of frequencies between
two adjacent time points for any barcode \(b\). This allows us to
rewrite the joint distribution
\(\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t)\) as
\begin{equation}\protect\hypertarget{eq-joint_freq_gamma}{}{
\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t) =
\pi(f_t^{(n)}, \gamma_{t}^{(n)} \mid \bar{s}_t).
}\label{eq-joint_freq_gamma}\end{equation} Let us rephrase this subtle
but necessary change of variables since it is a key part of the
inference problem: our series of independence assumptions lead us to
Equation~\ref{eq-meanfit_bayes} that relates the value of the population
mean fitness \(\bar{s}_t\) to the frequency of a neutral barcode at
times \(t\) and \(t+1\). However, as shown in
Equation~\ref{eq-fitness_ratio_neutral}, our model functionally relates
the ratio of frequencies---that we defined as \(\gamma_t^{(n)}\)---and
not the independent frequencies to the mean fitness. Therefore, instead
of writing for the likelihood the joint distribution of the frequency
values at times \(t\) and \(t+1\) conditioned on the mean fitness, we
write the joint distribution of the barcode frequency at time \(t\) and
the ratio of the frequencies. These \textbf{must be} equivalent joint
distributions since there is a one-to-one mapping between
\(\gamma_t^{(n)}\) and \(f_{t+1}^{(n)}\) for a given value of
\(f_t^{(n)}\). Another way to phrase this is to say that knowing the
frequency at time \(t\) and at time \(t+1\) provides the same amount of
information as knowing the frequency at time \(t\) and the ratio of the
frequencies. This is because if we want to obtain \(f_{t+1}^{(n)}\)
given this information, we simply compute
\begin{equation}\protect\hypertarget{eq-meanfit_ft+1_gamma}{}{
f_{t+1}^{(n)} = \gamma_t^{(n)} f_t^{(n)}.
}\label{eq-meanfit_ft+1_gamma}\end{equation}

The real advantage of rewriting the joint distribution as in
Equation~\ref{eq-joint_freq_gamma} comes from splitting this joint
distribution as a product of conditional distributions of the form
\begin{equation}\protect\hypertarget{eq-joint_to_prod_gamma}{}{
\pi(f_t^{(n)}, \gamma_{t}^{(n)} \mid \bar{s}_t) =
\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t)
\pi(\gamma_{t}^{(n)} \mid \bar{s}_t).
}\label{eq-joint_to_prod_gamma}\end{equation} Written in this form, we
can finally propose a probabilistic model for how the mean fitness
relates to the frequency ratios we determine in our experiments. The
second term on the right-hand side of
Equation~\ref{eq-joint_to_prod_gamma} relates how the determined
frequency ratio \(\gamma_t^{(b)}\) relates to the mean fitness
\(\bar{s}_t\). From Equation~\ref{eq-fitness_ratio_neutral} and
Equation~\ref{eq-gamma_def}, we can write
\begin{equation}\protect\hypertarget{eq-log_ratio_neutral}{}{
\ln \gamma_t^{(n)} = - \bar{s}_t + \varepsilon_t^{(n)},
}\label{eq-log_ratio_neutral}\end{equation} where, for simplicity, we
set \(\tau = 1\). Note that we added an extra term,
\(\varepsilon_t^{(n)}\), characterizing the deviations of the
measurements from the theoretical model. We assume these errors are
normally distributed with mean zero and some standard deviation
\(\sigma_t\), implying that
\begin{equation}\protect\hypertarget{eq-log_gamma_normal}{}{
\ln \gamma_t^{(n)} \mid \bar{s}_t, \sigma_t  \sim 
\mathcal{N}\left(-\bar{s}_t, \sigma_t \right),
}\label{eq-log_gamma_normal}\end{equation} where we include the nuisance
parameter \(\sigma_t\) to be determined. If we assume the log frequency
ratio is normally distributed, this implies the frequency ratio itself
is distributed log-normal. This means that
\begin{equation}\protect\hypertarget{eq-meanfit_likelihood}{}{
\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t  \sim 
\log \mathcal{N}\left(-\bar{s}_t, \sigma_t \right).
}\label{eq-meanfit_likelihood}\end{equation} Having added the nuisance
parameter \(\sigma_t\) implies that we must update
Equation~\ref{eq-meanfit_bayes} to
\begin{equation}\protect\hypertarget{eq-meanfit_bayes_full}{}{
\pi(\bar{s}_t, \sigma_t \mid f_t^{(n)}, f_{t+1}^{(n)}) \propto
\pi(f_t^{(n)}, \gamma_t^{(n)} \mid \bar{s}_t, \sigma_t) 
\pi(\bar{s}_t) \pi(\sigma_t),
}\label{eq-meanfit_bayes_full}\end{equation} where we assume the prior
for each parameter is independent, i.e.,
\begin{equation}\protect\hypertarget{eq-meanfit_indep_prior}{}{
\pi(\bar{s}_t, \sigma_t) = \pi(\bar{s}_t) \pi(\sigma_t).
}\label{eq-meanfit_indep_prior}\end{equation} For numerical stability,
we will select weakly-informative priors for both of these parameters.
In the case of the nuisance parameter \(\sigma_t\), the prior must be
restricted to positive values only, since standard deviations cannot be
negative.

For the first term on the right-hand side of
Equation~\ref{eq-joint_to_prod_gamma},
\(\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t)\), we remove the
conditioning on the population mean fitness since it does not add any
information on top of what the frequency ratio \(\gamma_t^{(n)}\)
already gives. Therefore, we have
\begin{equation}\protect\hypertarget{eq-freq_cond_gamma}{}{
\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t) =
\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}).
}\label{eq-freq_cond_gamma}\end{equation} The right-hand side of
Equation~\ref{eq-freq_cond_gamma} asks us to compute the probability of
observing a frequency value \(f_t^{(n)}\) given that we get to observe
the ratio \(\gamma_{t}^{(n)}\). If the ratio happened to be
\(\gamma_{t}^{(n)} = 2\), we could have \(f_{t+1}^{(n)} = 1\) and
\(f_{t+1}^{(n)} = 0.5\), for example. Although, it would be equally
likely that \(f_{t+1}^{(n)} = 0.6\) and \(f_{t+1}^{(n)} = 0.3\) or
\(f_{t+1}^{(n)} = 0.1\) and \(f_{t+1}^{(n)} = 0.05\) for that matter. If
we only get to observe the frequency ratio \(\gamma_t^{(n)}\), we know
that the numerator \(f_{t+1}^{(n)}\) can only take values between zero
and one, all of them being equally likely given only the information on
the ratio. As a consequence, the value of the frequency in the
denominator \(f_{t}^{(n)}\) is restricted to fall in the range
\begin{equation}\protect\hypertarget{eq-freq_given_gamma_range}{}{
f_{t}^{(n)} \in \left(0, \frac{1}{\gamma_t^{(n)}} \right].
}\label{eq-freq_given_gamma_range}\end{equation} A priori, we do not
have any reason to favor any value over any other, therefore it is
natural to write
\begin{equation}\protect\hypertarget{eq-f_given_gamma_uniform}{}{
f_t^{(n)} \mid \gamma_t^{(n)} \sim 
\operatorname{Uniform}\left( 0, \frac{1}{\gamma_t^{(n)}} \right).
}\label{eq-f_given_gamma_uniform}\end{equation}

\hypertarget{summary-1}{%
\paragraph{Summary}\label{summary-1}}

Putting all the pieces we have developed in this section together
results in an inference for the population mean fitness values of the
form \begin{equation}\protect\hypertarget{eq-meanfit_full_inference}{}{
\pi(
    \underline{\bar{s}}_T, \underline{\sigma}_T \mid \underline{\underline{F}}
) \propto
\prod_{t=1}^{T-1} \left\{
    \prod_{n=1}^N \left[
        \pi(f_t^{(n)} \mid \gamma_t^{(n)}) 
        \pi(\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t)
    \right]
    \pi(\bar{s}_t) \pi(\sigma_t)
\right\},
}\label{eq-meanfit_full_inference}\end{equation} where we have
\begin{equation}\protect\hypertarget{eq-meanfit_eq1}{}{
f_t^{(n)} \mid \gamma_t^{(n)} \sim 
\operatorname{Uniform} \left(0, \frac{1}{\gamma_t^{(n)}} \right),
}\label{eq-meanfit_eq1}\end{equation}
\begin{equation}\protect\hypertarget{eq-meanfit_eq2}{}{
\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t \sim 
\log\mathcal{N}(\bar{s}_t, \sigma_t),
}\label{eq-meanfit_eq2}\end{equation}
\begin{equation}\protect\hypertarget{eq-meanfit_eq3}{}{
\bar{s}_t \sim \mathcal{N}(0, \sigma_{\bar{s}_t}),
}\label{eq-meanfit_eq3}\end{equation} and
\begin{equation}\protect\hypertarget{eq-meanfit_eq4}{}{
\sigma_t \sim \log\mathcal{N}(\mu_{\sigma_t}, \sigma_{\sigma_t}),
}\label{eq-meanfit_eq4}\end{equation} where \(\sigma_{\bar{s}_t}\),
\(\mu_{\sigma_t}\), and \(\sigma_{\sigma_t}\) are user-defined
parameters.

\hypertarget{sec-bayes_mutfit}{%
\subsubsection{\texorpdfstring{Mutant relative fitness uncertainty
\(\pi(\underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}}, \underline{\underline{R}})\)}{Mutant relative fitness uncertainty \textbackslash pi(\textbackslash underline\{s\}\^{}M \textbackslash mid \textbackslash underline\{\textbackslash bar\{s\}\}\_T, \textbackslash underline\{\textbackslash underline\{F\}\}, \textbackslash underline\{\textbackslash underline\{R\}\})}}\label{sec-bayes_mutfit}}

The last piece of our inference is the piece that we care about the
most: the probability distribution of all the mutants' relative fitness,
given the inferred population mean fitness and the frequencies. First,
we assume that all fitness values are independent of each other. This
allows us to write
\begin{equation}\protect\hypertarget{eq-mutfit_indep}{}{
\pi(
    \underline{s}^M \mid 
    \underline{\bar{s}}_T, \underline{\underline{F}}, \underline{\underline{R}}
) = 
\prod_{m=1}^M \pi(
    s^{(m)} \mid
    \underline{\bar{s}}_T, \underline{\underline{F}}, \underline{\underline{R}}
).
}\label{eq-mutfit_indep}\end{equation} Furthermore, as was the case with
the population mean fitness, our fitness model relates frequencies, not
raw reads. Moreover, the fitness value of mutant \(m\) only depends on
the frequencies of such mutant. Therefore, we can simplify the
conditioning to
\begin{equation}\protect\hypertarget{eq-mutfit_cond_simple}{}{
\pi(
    s^{(m)} \mid
    \underline{\bar{s}}_T, \underline{\underline{F}}, \underline{\underline{R}}
) = 
\pi(s^{(m)} \mid \underline{\bar{s}}_T, \underline{f}^{(m)}),
}\label{eq-mutfit_cond_simple}\end{equation} where
\begin{equation}\protect\hypertarget{eq-mutfit_timeseries_def}{}{
\underline{f}^{(m)} = (f_0^{(m)}, f_1^{(m)}, \ldots, f_T^{(m)})^\dagger,
}\label{eq-mutfit_timeseries_def}\end{equation} is the vector containing
the frequency time series for mutant \(m\). Writing Bayes' theorem for
the right-hand side of Equation~\ref{eq-mutfit_cond_simple} results in
\begin{equation}\protect\hypertarget{eq-mutfit_bayes}{}{
\pi(s^{(m)} \mid \underline{\bar{s}}_T, \underline{f}^{(m)}) \propto
\pi(\underline{f}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)})
\pi(s^{(m)} \mid \underline{\bar{s}}_T).
}\label{eq-mutfit_bayes}\end{equation} Notice the conditioning on the
mean fitness values \(\underline{\bar{s}}_T\) is not inverted since we
already inferred these values.

Following the logic used in Section~\ref{sec-bayes_meanfit}, let us
define \begin{equation}\protect\hypertarget{eq-gamma_vec_def}{}{
\underline{\gamma}^{(m)} = 
(\gamma_0^{(m)}, \gamma_1^{(m)}, \ldots, \gamma_{T-1}^{m})^\dagger,
}\label{eq-gamma_vec_def}\end{equation} where each entry
\(\gamma_t^{(m)}\) is defined by Equation~\ref{eq-gamma_def}. In the
same way we rewrote the joint distribution between two adjacent time
point frequencies to the joint distribution between one of the
frequencies and the ratio of both frequencies in
Equation~\ref{eq-joint_freq_gamma}, we can rewrite the joint
distribution of the frequency time series for mutant \(m\) as
\begin{equation}\protect\hypertarget{eq-joint_freq_gammas_mutfit}{}{
\pi(\underline{f}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}) =
\pi(f_0^{(m)}, \underline{\gamma}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}).
}\label{eq-joint_freq_gammas_mutfit}\end{equation} One can think about
Equation~\ref{eq-joint_freq_gammas_mutfit} as saying that knowing the
individual frequencies at each time point contain equivalent information
as knowing the initial frequency and the subsequent ratios of
frequencies. This is because if we want to know the value of
\(f_1^{(m)}\) given the ratios, we only need to compute
\begin{equation}\protect\hypertarget{eq-f1_from_ratio}{}{
f_1^{(m)} = \gamma_0^{(m)} f_0^{(m)}.
}\label{eq-f1_from_ratio}\end{equation} Moreover, if we want to know
\(f_2^{(m)}\), we have
\begin{equation}\protect\hypertarget{eq-f2_from_ratio}{}{
f_2^{(m)} = \gamma_1^{(m)} f_1^{(m)} =
\gamma_1^{(m)} \left(\gamma_0^{(m)} f_0^{(m)}\right),
}\label{eq-f2_from_ratio}\end{equation} and so on. We can then write the
joint distribution on the right-hand side of
Equation~\ref{eq-joint_freq_gammas_mutfit} as a product of conditional
distributions of the form
\begin{equation}\protect\hypertarget{eq-mutfit_joint_to_cond}{}{
\begin{aligned}
\pi(f_0^{(m)}, \underline{\gamma}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}) =
&\pi(
    f_0^{(m)} \mid 
    \gamma_0^{(m)}, \ldots, \gamma_{T-1}^{(m)}, \underline{\bar{s}}_T, s^{(m)}
) \times \\
&\pi(
    \gamma_0^{(m)} \mid 
    \gamma_1^{(m)}, \ldots, \gamma_{T-1}^{(m)}, \underline{\bar{s}}_T, s^{(m)}
) \times \\
&\pi(
    \gamma_1^{(m)} \mid 
    \gamma_2^{(m)}, \ldots, \gamma_{T-1}^{(m)}, \underline{\bar{s}}_T, s^{(m)}
) \times \\
&\vdots \\
&\pi(
    \gamma_{T-2}^{(m)} \mid \gamma_{T-1}^{(m)}, \underline{\bar{s}}_T, s^{(m)}
) \times \\
&\pi(\gamma_{T-1}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}).
\end{aligned}
}\label{eq-mutfit_joint_to_cond}\end{equation} Writing the fitness model
in Equation~\ref{eq-fitness} as \[
\gamma_t^{(m)} = \frac{f_{t+1}^{(m)}}{f_t^{(m)}} = 
\mathrm{e}^{(s^{(m)} - s_t)\tau},
\] reveals that the value of each of the ratios \(\gamma_t^{(m)}\) only
depends on the corresponding fitness value \(\bar{s}_t\) and the
relative fitness \(s^{(m)}\). Therefore, we can remove most of the
conditioning on the right-hand side of
Equation~\ref{eq-mutfit_joint_to_cond}, resulting in a much simpler
joint distribution of the form
\begin{equation}\protect\hypertarget{eq-mutfit_joint_to_cond_simple}{}{
\begin{aligned}
\pi(f_0^{(m)}, \underline{\gamma}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}) =
&\pi(f_0^{(m)} \mid \gamma_0^{(m)}) \times \\
&\pi(\gamma_0^{(m)} \mid \bar{s}_0, s^{(m)}) \times \\
&\pi(\gamma_1^{(m)} \mid \bar{s}_1, s^{(m)}) \times \\
&\vdots \\
&\pi(\gamma_{T-2}^{(m)} \mid \bar{s}_{T-2}, s^{(m)}) \times \\
&\pi(\gamma_{T-1}^{(m)} \mid \bar{s}_{T-1}, s^{(m)}),
\end{aligned}
}\label{eq-mutfit_joint_to_cond_simple}\end{equation} where for the
first term on the right-hand side of
Equation~\ref{eq-mutfit_joint_to_cond_simple} we apply the same logic as
in Equation~\ref{eq-freq_cond_gamma} to remove all other dependencies.
We emphasize that although Equation~\ref{eq-mutfit_joint_to_cond_simple}
looks like a series of independent inferences, the value of the relative
fitness \(s^{(m)}\) is shared among all of them. This means that the
parameter is not inferred individually for each time point, resulting in
different estimates of the parameter, but each time point contributes
independently to the inference of a single estimate of \(s^{(m)}\).

Using equivalent arguments to those in Section~\ref{sec-bayes_meanfit},
we assume \[
f_0^{(m)} \mid \gamma_0^{(m)} \sim 
\operatorname{Uniform}\left(0, \frac{1}{\gamma_0^{(m)}} \right),
\] and \begin{equation}\protect\hypertarget{eq-mutfit_lognormal}{}{
\gamma_t^{(m)} \mid \bar{s}_t, s^{(m)}, \sigma^{(m)} \sim 
\log\mathcal{N}\left(s^{(m)} - \bar{s}_t, \sigma^{(m)} \right),
}\label{eq-mutfit_lognormal}\end{equation} where we add the nuisance
parameter \(\sigma^{(m)}\) to the inference. Notice that this parameter
is not indexed by time. This means that we assume the deviations from
the theoretical prediction do not depend on time, but only on the
mutant. Adding the nuisance parameter demands us to update
Equation~\ref{eq-mutfit_bayes} to
\begin{equation}\protect\hypertarget{eq-mutfit_bayes_full}{}{
\pi(
    s^{(m)}, \sigma^{(m)} \mid \underline{\bar{s}}_T, \underline{f}^{(m)}
) \propto
\pi(\underline{f}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}, \sigma^{(m)})
\pi(s^{(m)}) \pi(\sigma^{(m)}),
}\label{eq-mutfit_bayes_full}\end{equation} where we assume independent
priors for both parameters. We also removed the conditioning on the
values of the mean fitness as knowing such values does not change our
prior information about the possible range of values that the parameters
can take. As with the priors on Section~\ref{sec-bayes_meanfit}, we will
assign weakly-informative priors to these parameters.

\hypertarget{summary-2}{%
\paragraph{Summary}\label{summary-2}}

With all pieces in place, we write the full inference of the relative
fitness values as
\begin{equation}\protect\hypertarget{eq-mutfit_full_inference}{}{
\pi(
    \underline{s}^M ,\underline{\sigma}^M \mid 
    \underline{\bar{s}}_T, \underline{\underline{F}}
) \propto
\prod_{m=1}^M \left\{ 
    \pi(f_0^{(m)} \mid \gamma_0^{(m)})
    \prod_{t=0}^{T-1} \left[
        \pi(\gamma_t^{(m)} \mid \bar{s}_t, s^{(m)}, \sigma^{(m)})
    \right]
    \pi(s^{(m)}) \pi(\sigma^{(m)})
\right\},
}\label{eq-mutfit_full_inference}\end{equation} where
\begin{equation}\protect\hypertarget{eq-mutfit_eq1}{}{
f_0^{(m)} \mid \gamma_0^{(m)} \sim 
\operatorname{Uniform}\left(0, \frac{1}{\gamma_0^{(m)}} \right),
}\label{eq-mutfit_eq1}\end{equation}
\begin{equation}\protect\hypertarget{eq-mutfit_eq2}{}{
\gamma_t^{(m)} \mid \bar{s}_t, s^{(m)}, \sigma^{(m)} \sim 
\log\mathcal{N}\left(s^{(m)} - \bar{s}_t, \sigma^{(m)} \right),
}\label{eq-mutfit_eq2}\end{equation}
\begin{equation}\protect\hypertarget{eq-mutfit_eq3}{}{
s^{(m)} \sim \mathcal{N}(0, \sigma_{s^{(m)}}),
}\label{eq-mutfit_eq3}\end{equation} and
\begin{equation}\protect\hypertarget{eq-mutfit_eq4}{}{
\sigma^{(m)} \sim \log\mathcal{N}(\mu_{\sigma^{(m)}}, \sigma_{\sigma^{(m)}}),
}\label{eq-mutfit_eq4}\end{equation} where \(\sigma_{s^{(m)}}\),
\(\mu_{\sigma^{(m)}}\), and \(\sigma_{\sigma^{(m)}}\) are user-defined
parameters.

\hypertarget{sec-hierarchical_model}{%
\subsubsection{Hierarchical models for multiple experimental
replicates}\label{sec-hierarchical_model}}

As detailed in Section~\ref{sec-replicates} of the main text, we define
a Bayesian hierarchical model to analyze data from multiple experimental
replicates. The implementation requires only slightly modifying the base
model detailed in the previous sections. The hierarchical model defines
a hyper-fitness parameter \(\theta^{(m)}\) for every non-neutral
barcode. We can thus collect all of the \(M\) hyperparameters in an
array of the form
\begin{equation}\protect\hypertarget{eq-hier_hyper_vector}{}{
\underline{\theta}^M = (\theta^{(1)}, \ldots, \theta^{(M)})^\dagger.
}\label{eq-hier_hyper_vector}\end{equation} Our data now consists of a
series of matrices \(\underline{\underline{R}}_{[j]}\), where the
subindex \([j]\) refers to the \(j\)-th experimental replicate. These
matrices need not have the same number of rows, as the time points
measured for each replicate can vary. The statistical model we must
define is then of the form
\begin{equation}\protect\hypertarget{eq-hier_bayes_full}{}{
\begin{aligned}
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\},
    \{\underline{\underline{F}}_{[j]}\} \mid
    \{\underline{\underline{R}}_{[j]}\}
) \propto \;
&\pi(
    \{\underline{\underline{R}}_{[j]}\} \mid
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\},
    \{\underline{\underline{F}}_{[j]}\} 
) \times \\\
&\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\},
    \{\underline{\underline{F}}_{[j]}\} 
)
\end{aligned}
}\label{eq-hier_bayes_full}\end{equation} where the parameters within
curly braces with subindex \([j]\) indicate one set of parameters per
experimental replicate. For example,
\begin{equation}\protect\hypertarget{eq-hier_array}{}{
\{\underline{s}^M_{[j]}\} = \{
    \underline{s}^M_{[1]}, \underline{s}^M_{[2]}, \ldots \underline{s}^M_{[E]}
\},
}\label{eq-hier_array}\end{equation} where \(E\) is the number of
experimental replicates.

Given the dependencies between the variables, we can factorize
Equation~\ref{eq-hier_bayes_full} to be of the form
\begin{equation}\protect\hypertarget{eq-hier_factorized}{}{
\begin{aligned}
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\},
    \{\underline{\underline{F}}_{[j]}\} \mid
    \{\underline{\underline{R}}_{[j]}\}
) =
&\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\} \mid
    \{\underline{\bar{s}}_{T[j]}\},
    \{\underline{\underline{F}}_{[j]}\} 
) \times \\
&\pi(
    \{\underline{\bar{s}}_{T[j]}\} \mid
    \{\underline{\underline{F}}_{[j]}\}
) \times \\
&\pi(
    \{\underline{\underline{F}}_{[j]}\} \mid
    \{\underline{\underline{R}}_{[j]}\}
)
\end{aligned}
}\label{eq-hier_factorized}\end{equation} Furthermore, the hierarchical
structure only connects the replicates via the relative fitness
parameters. This means that the population mean fitness values and the
frequencies can be independently inferred for each dataset. This allows
us to rewrite the right-hand side of Equation~\ref{eq-hier_factorized}
as
\begin{equation}\protect\hypertarget{eq-hier_bayes_factorized_replicate}{}{
\begin{aligned}
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\},
    \{\underline{\underline{F}}_{[j]}\} \mid
    \{\underline{\underline{R}}_{[j]}\}
) =
&\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\} \mid
    \{\underline{\bar{s}}_{T[j]}\},
    \{\underline{\underline{F}}_{[j]}\} 
) \times \\
&\prod_{j=1}^E \left[
    \pi(
        \underline{\bar{s}}_{T[j]} \mid
        \underline{\underline{F}}_{[j]}
    )
    \pi(
        \underline{\underline{F}}_{[j]} \mid
        \underline{\underline{R}}_{[j]}
    )
\right].
\end{aligned}
}\label{eq-hier_bayes_factorized_replicate}\end{equation} The terms
inside the square brackets in
Equation~\ref{eq-hier_bayes_factorized_replicate} take the same
functional form as those derived in Section~\ref{sec-bayes_meanfit} and
Section~\ref{sec-bayes_freq}. Therefore, to implement the desired
hierarchical model, we only need to focus on the first term on the
right-hand side of Equation~\ref{eq-hier_bayes_factorized_replicate}. A
way to think about the structure of the hierarchical model is as
follows: imagine each genotype as a ``\emph{true}'' relative fitness
value. However, every time we perform an experiment, small variations in
the biotic and abiotic conditions---also known as batch effects---might
result in small deviations from this value. We model this by defining a
distribution for the hyper-fitness parameter---the ground truth we are
interested in---and having each experimental replicate sample from this
hyper-parameter distribution to determine the ``\emph{local}'' fitness
value. The wider the hyper-parameter distribution is the more
variability between experimental replicates.

Writing Bayes' theorem for the first term in
Equation~\ref{eq-hier_bayes_factorized_replicate} results in
\begin{equation}\protect\hypertarget{eq-hier_bayes}{}{
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\} \mid
    \{\underline{\underline{F}}_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\}
) \propto
\pi(
    \{\underline{\underline{F}}_{[j]}\} \mid
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\}
)
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\} \mid
    \{\underline{\bar{s}}_{T[j]}\}
),
}\label{eq-hier_bayes}\end{equation} where we leave the conditioning on
the population mean fitness as we did in Section~\ref{sec-bayes_mutfit}.
This expression can be simplified in two ways. First, the frequency
values for each experimental replicate depend directly on the local
fitness values and the corresponding population mean fitness, as the
relationship between experimental replicates only occurs through the
relative fitness values. Therefore, we can write
\begin{equation}\protect\hypertarget{eq-hier_bayes_exp_indep}{}{
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\} \mid
    \{\underline{\underline{F}}_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\}
) \propto
\prod_{j=1}^E\left[
    \pi(
        \underline{\underline{F}}_{[j]} \mid
        \underline{s}^M_{[j]},
        \underline{\bar{s}}_{T[j]}
    )
\right]
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\} \mid
    \{\underline{\bar{s}}_{T[j]}\}
).
}\label{eq-hier_bayes_exp_indep}\end{equation} Second, the relationship
between the hyper-fitness and the local fitness values allows us to
write their joint distribution as a conditional distribution where local
fitness values depend on the global hyper-fitness value, obtaining
\begin{equation}\protect\hypertarget{eq-hier_bayes_exp_indep}{}{
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\} \mid
    \{\underline{\underline{F}}_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\}
) \propto
\prod_{j=1}^E\left[
    \pi(
        \underline{\underline{F}}_{[j]} \mid
        \underline{s}^M_{[j]},
        \underline{\bar{s}}_{T[j]}
    )
    \pi(
        \underline{s}^M_{[j]} \mid
        \underline{\theta}^M
    )
\right] 
\pi(
    \underline{\theta}^M
).
}\label{eq-hier_bayes_exp_indep}\end{equation} Notice we removed the
conditioning on the population mean fitness as our prior expectations of
what the global hyper-fitness or local fitness value might be do not
depend on these nuisance parameters.

The first term on the right-hand side of
Equation~\ref{eq-hier_bayes_exp_indep} takes the same functional form as
the one derived in Section~\ref{sec-bayes_mutfit}. Therefore, all we are
left with is to determine the functional forms for the hyper--prior
\(\pi(\underline{\theta}^M)\), and the conditional probability
\(\pi(\underline{s}^M_{[j]} \mid \underline{\theta}^M)\). In analogy to
the assumptions used for the fitness values in
Section~\ref{sec-bayes_mutfit}, we define the value of each
hyper-fitness as independent. This means that we have
\begin{equation}\protect\hypertarget{eq-hier_hyper_prior}{}{
\pi(\underline{\theta}^M) = \prod_{m=1}^M \pi(\theta^{(m)}).
}\label{eq-hier_hyper_prior}\end{equation} Furthermore, we assume this
prior is of the form
\begin{equation}\protect\hypertarget{eq-hier_hyper_prior_02}{}{
\theta^{(m)} \sim \mathcal{N}(\mu_{\theta^{(m)}}, \sigma_{\theta^{(m)}}),
}\label{eq-hier_hyper_prior_02}\end{equation} where
\(\mu_{\theta^{(m)}}\) and \(\sigma_{\theta^{(m)}}\) are user-defined
parameters encoding the prior expectations on the fitness values.

For the conditional distribution
\(\pi(\underline{s}^M_{[j]} \mid \underline{\theta}^M)\), we use the
so-called non-centered parametrization that avoids some of the intrinsic
degeneracies associated with hierarchical models
\autocite{betancourt2013}. We invite the reader to check
\href{https://betanalpha.github.io/assets/case_studies/hierarchical_modeling.html}{this
excellent blog} explaining the difficulties of working with hierarchical
models. This non-centered parameterization implies that we introduce two
nuisance parameters such that the local fitness \(s_{[j]}^{(m)}\) is
computed as \begin{equation}\protect\hypertarget{eq-hier_noncentered}{}{
s_{[j]}^{(m)} = \theta^{(m)} + 
(\tau_{[j]}^{(m)} \times \xi_{[j]}^{(m)}),
}\label{eq-hier_noncentered}\end{equation} where \(\theta^{(m)}\) is the
corresponding genotype hyper-fitness value, \(\xi_{[j]}^{(m)}\) is a
standard normal random variable, i.e.,
\begin{equation}\protect\hypertarget{eq-hier_stdnormal}{}{
\xi_{[j]}^{(m)} \sim \mathcal{N}(0, 1),
}\label{eq-hier_stdnormal}\end{equation} that allows deviations from the
hyper-fitness value to be either positive or negative, and
\(\tau_{[j]}^{(m)}\) is a strictly positive random variable that
characterizes the deviation of the local fitness value from the global
hyper-fitness. We assume
\begin{equation}\protect\hypertarget{eq-hier_tau}{}{
\tau_{[j]}^{(m)} \sim \log\mathcal{N}(
    \mu_{\tau_{[j]}^{(m)}}, \sigma_{\tau_{[j]}^{(m)}}
)
}\label{eq-hier_tau}\end{equation} where \(\mu_{\tau_{[j]}^{(m)}}\) and
\(\sigma_{\tau_{[j]}^{(m)}}\) are user-defined parameters capturing the
expected magnitude of the batch effects.

\hypertarget{defining-prior-probabilities}{%
\subsubsection{Defining prior
probabilities}\label{defining-prior-probabilities}}

One aspect commonly associated---in both positive and negative ways---to
Bayesian analysis is the definition of prior probabilities. On the one
hand, the naive textbook version of Bayesian analysis defines the prior
as encoding the information we have about the inference in question
before acquiring any data. This is the ``ideal'' use of priors that,
whenever possible, should be implemented. On the other hand, for most
practitioners of Bayesian statistics in the age of big data, the
definition of prior becomes a tool to ensure the convergence of sampling
algorithms such as MCMC \autocite{gelman2017}. However, for our
particular problem, although we deal with large amounts of data
(inferences can be made for \textgreater{} 10K barcodes over multiple
time points, resulting in \textgreater{} 100K parameters), each barcode
has very little data, as they are measured only once per time point over
\textless{} 10 growth-dilution cycles. Furthermore, it is incredibly
challenging to understand the noise sources related to culturing
conditions, DNA extraction, library preparation, etc., and encode them
into reasonable prior distributions.

Empirically, our approach for this work defined the priors based solely
on the neutral lineage data, as they represent the only repeated
measurements of a single genotype in our experimental design. We
acknowledge that defining the priors after observing the data might be
considered an incoherent inference. However, as expressed by
\textcite{gelman2017}

\begin{quote}
Incoherence is an unavoidable aspect of much real-world data analysis;
and, indeed, one might argue that as scientists we learn the most from
the anomalies and reassessments associated with episodes of incoherence.
\end{quote}

With this in mind, we leave it to the reader to judge the selection of
priors. Furthermore, the software package associated with this work,
\texttt{BarBay.jl}, is written so that users can experiment with
different prior selection criteria that fit their needs. We strongly
advocate that statistics should not be done in a black-box fit-all tool
mindset but rather as a formal way to encode the assumptions behind the
analysis, subject to constructive criticism. With this philosophical
baggage behind us, let us now focus on how the priors used for this work
were selected.

\hypertarget{naive-neutral-lineage-based-priors}{%
\paragraph{Naive neutral lineage-based
priors}\label{naive-neutral-lineage-based-priors}}

For the base model presented in this work, the user-defined prior
parameters include the following:

\begin{itemize}
\item
  Prior on population mean fitness (one per pair of adjacent time
  points) \begin{equation}\protect\hypertarget{eq-prior_meanfit_mean}{}{
  \bar{s}_t \sim \mathcal{N}(\mu_{\bar{s}_t}, \sigma_{{\bar{s}_t}}).
  }\label{eq-prior_meanfit_mean}\end{equation}
\item
  Prior on standard deviation associated with neutral lineages
  likelihood function (one per pair of adjacent time points)
  \begin{equation}\protect\hypertarget{eq-prior_meanfit_var}{}{
  \sigma_t \sim \log\mathcal{N}(\mu_{\sigma_t}, \sigma_{\sigma_t}).
  }\label{eq-prior_meanfit_var}\end{equation}
\item
  Prior on relative fitness (one per non-neutral barcode)
  \begin{equation}\protect\hypertarget{eq-prior_fit_mean}{}{
  s^{(m)} \sim \mathcal{N}(\mu_{s^{(m)}}, \sigma_{s^{(m)}}).
  }\label{eq-prior_fit_mean}\end{equation}
\item
  Prior on standard deviation associated with non-neutral lineages
  likelihood function (one per non-neutral barcode)
  \begin{equation}\protect\hypertarget{eq-prior_fit_var}{}{
  \sigma^{(m)} \sim \log\mathcal{N}(\mu_{\sigma^{(m)}}, \sigma_{\sigma^{(m)}})
  }\label{eq-prior_fit_var}\end{equation}
\end{itemize}

The \texttt{BarBay.jl} package includes a function \texttt{naive\_prior}
within the \texttt{stats} module. This function utilizes the data from
the neutral lineages to determine some of the prior parameters to
facilitate the inference algorithm's numerical convergence. In
particular, it defines the population mean fitness parameter
\(\mu_{\bar{s}_t}\) as
\begin{equation}\protect\hypertarget{eq-prior_meanfit_mean_data}{}{
\mu_{\bar{s}_t} = \frac{1}{N} \sum_{n=1}^N 
-\ln\left( \frac{r_{t+1}^{(n)}}{r_t^{(n)}} \right),
}\label{eq-prior_meanfit_mean_data}\end{equation} where \(N\) is the
number of neutral lineages and \(r_t^{(n)}\) is the number of neutral
lineages. In other words, it defines the mean of the prior distribution
as the mean of what one naively would compute from the neutral lineages,
discarding cases where the ratio diverges because the denominator
\(r_t^{(n)} = 0\). For the variance parameter, we chose a value
\(\sigma_{\bar{s}_t} = 0.05\).

Furthermore, the \texttt{naive\_prior} function defines the mean of the
variance parameter as the standard deviation of the log frequency ratios
for the neutral lineages, i.e.,
\begin{equation}\protect\hypertarget{eq-prior_meanfit_var_data}{}{
\mu_{\sigma_t} = \sqrt{
    \operatorname{Var}\left(\frac{r_{t+1}^{(n)}}{r_t^{(n)}}\right)
},
}\label{eq-prior_meanfit_var_data}\end{equation} where
\(\operatorname{Var}\) is the sample variance. This same value was
utilized for the mean of the non-neutral barcode variance
\(\mu_{\sigma^{(m)}}\). While we assigne the corresponding variances to
be \(\sigma_{\sigma_t} = \sigma_{\sigma^{(m)}} = 1\).

\hypertarget{sec-ppc}{%
\subsection{Posterior predictive checks}\label{sec-ppc}}

Throughout the main text, we allude to the concept of posterior
predictive checks as a formal way to assess the accuracy of our
inference pipeline. Here, we explain the mechanics behind the
computation of these credible regions, given the output of the
inference.

Bayesian models encode what is known as a \emph{generative model}. This
statement means that in our definition of the likelihood function and
the prior distribution, we, as modelers, propose a mathematical function
that captures all relevant relationships between unobserved (latent)
variables. Therefore, when these latent variables are input into the
mathematical model, this function \emph{generates} data that should be,
in principle, indistinguishable from the real observations if the model
is a good account of the underlying processes involved in the phenomena
of interest. This generative model implies that once we run the
inference process and update our posterior beliefs about the state of
the latent variables, we can input back the inferred values to our model
and generate synthetic data. Furthermore, we can repeat this process
multiple times to compute the range where we expect to observe our data
conditioned on the accuracy of the model.

For our specific scenario, recall that our objective is to infer the
relative fitness of a non-neutral lineage \(s^{(m)}\) along with
nuisance parameters related to the population mean fitness at each
point, \(\underline{\bar{s}}_t\), and the barcode frequency time series
\(\underline{f}^{(m)}\). All these variables are related through our
fitness model (see Section~\ref{sec-fitness_model} in the main text)
\begin{equation}\protect\hypertarget{eq-ppc_model}{}{
f_{t+1}^{(m)} = f_{t}^{(m)} \mathrm{e}^{(s^{(m)} - s_t)\tau}.
}\label{eq-ppc_model}\end{equation} As we saw, it is convenient to
rewrite Equation~\ref{eq-ppc_model} as
\begin{equation}\protect\hypertarget{eq-ppc_logratio}{}{
\frac{1}{\tau}\ln\frac{f_{t+1}^{(m)}}{f_{t}^{(m)}} = (s^{(m)} - s_t).
}\label{eq-ppc_logratio}\end{equation} Written in this way, we separate
the quantities we can compute from the experimental observations---the
left-hand side of Equation~\ref{eq-ppc_logratio} can be computed from
the barcode reads---from the latent variables.

Although we perform the joint inference over all barcodes in the present
work, let us focus on the inference task for a single barcode as if it
were computed independently. For a non-neutral barcode, our task
consists of computing the posterior probability
\begin{equation}\protect\hypertarget{eq-ppc_posterior}{}{
\pi(\theta \mid \underline{r}^{(m)}) =
\pi(
    s^{(m)}, \sigma^{(m)}, \underline{s}_t, \underline{f}^{(m)}
    \mid \underline{r}^{(m)}
),
}\label{eq-ppc_posterior}\end{equation} where \(\theta\) represents all
parameters to be inferred and \(\underline{r}^{(m)}\) is the vector with
the barcode raw counts time series. The list of parameters are

\begin{itemize}
\tightlist
\item
  \(s^{(m)}\): The barcode's relative fitness.
\item
  \(\sigma^{(m)}\): A nuisance parameter used in the likelihood to
  generate the data. This captures the expected deviation from
  Equation~\ref{eq-ppc_logratio}
\item
  \(\underline{s}_t\): The vector with all population mean fitness for
  each pair of adjacent time points.
\item
  \(\underline{f}^{(m)}\): The vector with the barcode frequency time
  series.
\end{itemize}

Furthermore, let us define a naive estimate of the barcode frequency at
time \(t\) as \begin{equation}\protect\hypertarget{eq-ppc_naive_freq}{}{
\hat{f}_t^{(m)} = \frac{r^{(m)}_t}{\sum_{b=1}^B r^{(b)}_t}.
}\label{eq-ppc_naive_freq}\end{equation} We can compute this quantity
from the data by normalizing the raw barcode counts by the sum of all
barcode counts. Furthermore, we can compute a naive estimate of the log
frequency ratio from the raw barcode counts as
\begin{equation}\protect\hypertarget{eq-ppc_naive_logfreq}{}{
\ln \hat{\gamma}_t^{(m)} = \ln \frac{\hat{f}_{t+1}^{(m)}}{\hat{f}_t^{(m)}}
}\label{eq-ppc_naive_logfreq}\end{equation}

In our generative model, we assumed
\begin{equation}\protect\hypertarget{eq-ppc_logfreq_likelihood}{}{
\ln\gamma^{(m)}_t \mid \theta \sim \mathcal{N}(s^{(m)} - s_t, \sigma^{(m)}).
}\label{eq-ppc_logfreq_likelihood}\end{equation} This implies that once
we determine the posterior distribution of our parameters, we can
generate synthetic values of \(\ln \gamma^{(m)}_t\) that we can then
compare with the values obtained from applying
Equation~\ref{eq-ppc_naive_logfreq} and
Equation~\ref{eq-ppc_naive_logfreq} to the raw data.

In practice, to compute the posterior predictive checks, we generate
multiple samples from the posterior distribution
\(\pi(\theta \mid \underline{r}^{(m)})\)
\begin{equation}\protect\hypertarget{eq-ppc_theta_samples}{}{
\underline{\theta} = (\theta_1, \theta_2, \ldots, \theta_N).
}\label{eq-ppc_theta_samples}\end{equation} With these samples in hand,
the \texttt{BarBay.jl} package includes the function
\texttt{logfreq\_ratio\_bc\_ppc} for non-neutral barcodes that uses this
set of posterior parameter samples to generate samples from the
distribution defined in Equation~\ref{eq-ppc_logfreq_likelihood}. For a
large-enough number of samples, we can then compute the desired
percentiles---5, 68, and 95 percentiles in all figures in the main
text---that are equivalent to the corresponding credible regions. In
other words, the range of values of \(\ln\gamma_t^{(m)}\) generated by
this bootstrap process can be used to compute the region where we expect
to find our raw estimates \(\ln\hat{\gamma}_t^{(m)}\) with the desired
probability. The package \texttt{BarBay.jl} includes an equivalent
function, \texttt{logfreq\_ratio\_popmean\_ppc}, for neutral lineages.

\hypertarget{sec-logistic}{%
\subsection{Logistic growth simulation}\label{sec-logistic}}

In this section, we explain the simulations used to assess the validity
of our inference pipeline. Let us begin by assuming that, since the
strains are grown for two full days in the experiment, having left
behind the exponential phase for almost an entire day, a simple
exponential growth of the form
\begin{equation}\protect\hypertarget{eq-logistic_expo}{}{
\frac{dn_i}{dt} = \lambda_i n_i,
}\label{eq-logistic_expo}\end{equation} where \(n_i\) is the number of
cells of strain \(i\), and \(\lambda_i\) is the corresponding growth
rate is not enough. Instead, we will assume that the cells follow the
logistic growth equation of the form
\begin{equation}\protect\hypertarget{eq-logistic_fn}{}{
\frac{dn_i}{dt} = \lambda_i n_i 
\left( 1 - \frac{\sum_{j=1}^{N}n_i}{\kappa}\right),
}\label{eq-logistic_fn}\end{equation} where \(\kappa\) is the carrying
capacity, and \(N\) is the total number of strains in the culture.

The inference method is based on the model that assumes that the time
passed between dilutions \(\tau \approx 8\) generations, the change in
frequency for a mutant barcode can be approximated from cycle \(t\) to
the next cycle \(t + 1\) as
\begin{equation}\protect\hypertarget{eq-logistic_fitness_fn}{}{
f_{t+1}^{(m)} = f_t^{(m)} \mathrm{e}^{(s^{(m)} - \bar{s}_t)\tau},
}\label{eq-logistic_fitness_fn}\end{equation} where \(s^{(m)}\) is the
relative fitness for strain \(i\) compared to the ancestral strain and
\(\bar{s}_t\) is the mean fitness of the population at cycle \(t\). To
test this assumption, we implemented a numerical experiment following
the logistic growth model described in Equation~\ref{eq-logistic_fn}.
Figure~\ref{fig-SIX_logistic_growth_01} shows an example of the
deterministic trajectories for 50 labeled neutral lineages and 1000
lineages of interest. The upper red curve that dominates the culture
represents the unlabeled ancestral strain included in the experimental
design described in Section~\ref{sec-experiment}.

\begin{figure}

{\centering \includegraphics{./figs/figSIX_logistic_growth_01.pdf}

}

\caption{\label{fig-SIX_logistic_growth_01}\textbf{Logistic growth
simulation over single growth cycle}. The dashed line represents the
neutral lineages, with the upper curve being the unlabeled neutral
strain. Color curves represent the genotypes of interest colored by
growth rate relative to the neutral lineage.}

\end{figure}

To simulate multiple growth-dilution cycles, we take the population
composition at the final time point and use it to initialize a new
logistic growth simulation. Figure~\ref{fig-SIX_logistic_growth_02}
shows the resulting number of cells at the last time point of a cycle
over multiple growth-dilution cycles for the genotypes in
@Figure~\ref{fig-SIX_logistic_growth_01}. We can see that the adaptive
lineages (blue curves) increase in abundance, while detrimental lineages
(red curves) decrease.

\begin{figure}

{\centering \includegraphics{./figs/figSIX_logistic_growth_02.pdf}

}

\caption{\label{fig-SIX_logistic_growth_02}\textbf{Growth-dilution
cycles for logistic growth simulation}. Each point represents the final
number of cells after a growth cycle for each lineage. Colors are the
same as in Figure~\ref{fig-SIX_logistic_growth_01}.}

\end{figure}

In Section~\ref{sec-fitness_model}, we derive the functional form to
infer the relative fitness of each lineage as
\begin{equation}\protect\hypertarget{eq-logistic_logfreq}{}{
\frac{1}{\tau}\ln \frac{f_{t+1}^{(b)}}{f_{t}^{(b)}} = (s^{(b)} - \bar{s}_t).
}\label{eq-logistic_logfreq}\end{equation}
Figure~\ref{fig-SIX_logistic_growth_03} shows the corresponding log
frequency ratio curves for the logistic growth simulation. The
displacement of these curves with respect to the neutral lineages
determines the ground truth relative fitness value for these
simulations.

\begin{figure}

{\centering \includegraphics{./figs/figSIX_logistic_growth_03.pdf}

}

\caption{\label{fig-SIX_logistic_growth_03}\textbf{Log frequency ratio
for logistic growth simulations}. The relative distance of the color
curves from the black dashed line determines the relative fitness of
each linage.}

\end{figure}

To simulate the experimental noise, we add two types of noise:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Poisson noise between dilutions. For this, we take the final point of
  the logistic growth simulation and sample a random Poisson number
  based on this last point to set the initial condition for the next
  cycle.
\item
  Gaussian noise when performing the measurements. When translating the
  underlying population composition to the number of reads, we can add a
  custom amount of Gaussian noise.
\end{enumerate}

Figure~\ref{fig-SIX_logistic_growth_04} shows the frequency trajectories
(left panels) and log frequency ratios (right panels) for a noiseless
simulation (upper panels) and a simulation with added noise (lower
panels). The noiseless simulation is used to determine the relative
fitness for each of the lineages, which serves as the ground truth to be
compared with the resulting inference.

\begin{figure}

{\centering \includegraphics{./figs/figSIX_logistic_growth_04.pdf}

}

\caption{\label{fig-SIX_logistic_growth_04}\textbf{Logistic
growth-dilution simulations with and without noise}}

\end{figure}

\printbibliography

\end{document}

### 2023-10-11 doc/paper.tex
adding acknowledments to paper

@@
 \usepackage{titling}
+
+% Define command to begin the supplementary section
+\newcommand{\beginsupplement}{
+\setcounter{section}{0} % Restart section counter
+\renewcommand{\thesection}{S\arabic{section}}%
+\setcounter{table}{0} % Restart table counter
+\renewcommand{\thetable}{S\arabic{table}}%
+\setcounter{figure}{0} % Restart figure counter
+\renewcommand{\thefigure}{S\arabic{figure}}%
+\setcounter{equation}{0} % Restart equation counter
+\renewcommand{\theequation}{S\arabic{equation}}%
+}
 
@@
 \author[1]{Manuel Razo-Mejia}
-\author[2,3,4]{Madhav Mani}
-\author[1,5,*]{Dmitri Petrov}
+\author[2,3]{Madhav Mani}
+\author[1,4,*]{Dmitri Petrov}
 
@@
 \affil[3]{NSF-Simons Center for Quantitative Biology, Northwestern University, Chicago, IL, USA}
-\affil[4]{Department of Molecular Biosciences, Northwestern University, Chicago, IL, USA}
-\affil[5]{Stanford Cancer Institute, Stanford University School of Medicine, Stanford, CA, USA}
+\affil[4]{Stanford Cancer Institute, Stanford University School of Medicine, Stanford, CA, USA}
 \affil[*]{Correspondence: dpetrov@stanford.edu}
@@
 \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden, interior hidden, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi
+
+% Remove main text from the table of contents by specifying not to include
+% any section or subsection
+\addtocontents{toc}{\protect\setcounter{tocdepth}{-1}}
+
+% Define reference segment for main text
+\begin{refsegment}
+% Generate filter to not include references from main text in the
+% supplemental references
+\defbibfilter{notother}{not segment=\therefsegment}
 
@@
 
+\hypertarget{acknowledgements}{%
+\section{Acknowledgements}\label{acknowledgements}}
+
+We would like to thank Griffin Chure and Michael Betancourt for their helpful
+advice and discussion. We would like to thank Karna Gowda, Spencer Farrell, and
+Shaili Mathur for critical observations on the manuscript. MRM was supported by
+the Schmidt Science Fellowship. MM was supported by The National Science
+Foundation-Simons Center for Quantitative Biology at Northwestern University and
+the Simons Foundation grant 597491. MM is a Simons Investigator.
+
+% Print main text references
+\printbibliography[segment=\therefsegment]
+% Close reference segment
+\end{refsegment}
+
+\clearpage
+
+% SUPPLEMENTAL MATERIAL
+
+% Indicate that now all sections and subsections should be included in the
+% table of contents so that only the SI is included.
+\addtocontents{toc}{\protect\setcounter{tocdepth}{3}}
+
+% Define reference section for the supplemental material
+\begin{refsegment}
+% Set equation, table and figure counters to begin with "S"
+\beginsupplement
+
 \hypertarget{supplementary-materials}{%
 \section{Supplementary Materials}\label{supplementary-materials}}
+
+\tableofcontents
 
@@
 
-\printbibliography
+% Print supplemental references changing the title
+\printbibliography[title={Supplemental References},
+segment=\therefsegment, filter=notother]
+\end{refsegment}
+
+% \printbibliography
 

### 2023-10-11 doc/_abstract.qmd
add missing sections to paper

--- 2023-10-11 doc/paper.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
-%
-\documentclass[
-  letterpaper,
-  DIV=11,
-  numbers=noendperiod]{scrartcl}
-
-\usepackage{amsmath,amssymb}
-\usepackage{iftex}
-\ifPDFTeX
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-\usepackage{lmodern}
-\ifPDFTeX\else  
-    % xetex/luatex font selection
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-
-\usepackage{lineno}
-\linenumbers
-\usepackage[font={footnotesize, sf}, labelfont=bf, margin=0.05\linewidth]{caption}
-\renewcommand{\familydefault}{\sfdefault}
-% Bold the 'Figure #' in the caption and separate it from the title/caption
-% with a period
-% Captions will be left justified
-\usepackage[
-	aboveskip=1pt,
-	labelfont=bf,
-	labelsep=period,
-	justification=raggedright,
-	singlelinecheck=off
-]{caption}
-
-% Add numbered lines
-\usepackage{lineno}
-\linenumbers
-
-% Package to include multiple title pages
-% This will allow me to add a tile to the main text and to the SI
-\usepackage{titling}
-
-% Add author affiliations
-\usepackage{authblk}
-
-\KOMAoption{captions}{tableheading}
-\makeatletter
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\AtBeginDocument{%
-\ifdefined\contentsname
-  \renewcommand*\contentsname{Table of contents}
-\else
-  \newcommand\contentsname{Table of contents}
-\fi
-\ifdefined\listfigurename
-  \renewcommand*\listfigurename{List of Figures}
-\else
-  \newcommand\listfigurename{List of Figures}
-\fi
-\ifdefined\listtablename
-  \renewcommand*\listtablename{List of Tables}
-\else
-  \newcommand\listtablename{List of Tables}
-\fi
-\ifdefined\figurename
-  \renewcommand*\figurename{Figure}
-\else
-  \newcommand\figurename{Figure}
-\fi
-\ifdefined\tablename
-  \renewcommand*\tablename{Table}
-\else
-  \newcommand\tablename{Table}
-\fi
-}
-\@ifpackageloaded{float}{}{\usepackage{float}}
-\floatstyle{ruled}
-\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
-\floatname{codelisting}{Listing}
-\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
-\makeatother
-\makeatletter
-\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
-\makeatother
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-
-% References
-% \usepackage[]{biblatex}
-\usepackage[
-	backend=bibtex,
-	style=phys, 
-	sorting=none,				% Do not sort bibliography
-	url=false, 					% Do not show url in reference
-	doi=false, 					% Do not show doi in reference
-	isbn=false, 				% Do not show isbn link in reference
-	eprint=false, 			% Do not show eprint link in reference
-	maxbibnames=9, 			% Include up to 9 names in citation
-	firstinits=true,
-]{biblatex}
-\addbibresource{references.bib}
-
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
-\hypersetup{
-  pdfauthor={Manuel Razo-Mejia; Madhav Mani; Dmitri Petrov},
-  pdfkeywords={bayesian inference, variational inference, relative
-fitness, open-source, reproducible research},
-  colorlinks=true,
-  linkcolor={blue},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
-  pdfcreator={LaTeX via pandoc}}
-
-% Title
-\title{Bayesian inference of relative fitness on high-throughput pooled competition assays}
-
-% Authors
-\author[1]{Manuel Razo-Mejia}
-\author[2,3,4]{Madhav Mani}
-\author[1,5,*]{Dmitri Petrov}
-
-% Affiliations
-\affil[1]{Department of Biology, Stanford University, Stanford, CA, USA}
-\affil[2]{Department of Engineering Sciences and Applied Mathematics, Northwestern University, Chicago, IL, USA}
-\affil[3]{NSF-Simons Center for Quantitative Biology, Northwestern University, Chicago, IL, USA}
-\affil[4]{Department of Molecular Biosciences, Northwestern University, Chicago, IL, USA}
-\affil[5]{Stanford Cancer Institute, Stanford University School of Medicine, Stanford, CA, USA}
-\affil[*]{Correspondence: dpetrov@stanford.edu}
-
-\setcounter{Maxaffil}{0}
-% Set affiliations in small font
-\renewcommand\Affilfont{\itshape\small}
-
-\date{}
-
-\begin{document}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden, interior hidden, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi
-
-\maketitle
-
-\begin{abstract}
 The tracking of lineage frequencies via DNA barcode sequencing enables the
 quantification of microbial fitness. However, experimental noise coming from
-  biotic and abiotic sources complicates the computation of a reliable
-  inference. We present a Bayesian pipeline to infer relative microbial fitness
-  from high-throughput lineage tracking assays. Our model accounts for multiple
-  sources of noise and propagates uncertainties throughout all parameters in a
-  systematic way. Furthermore, using modern variational inference methods based
-  on automatic differentiation, we are able to scale the inference to a large
-  number of unique barcodes. We extend this core model to analyze
-  multi-environment assays, replicate experiments, and barcodes linked to
-  genotypes. On simulations, our method recovers known parameters within
-  posterior credible intervals. This work provides a generalizable Bayesian
-  framework to analyze lineage tracking experiments. The accompanying
-  open-source software library enables the adoption of principled statistical
-  methods in experimental evolution.
-\end{abstract}
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-The advent of DNA barcoding---the ability to uniquely identify cell
-lineages with DNA sequences integrated at a specific locus---and
-high-throughput sequencing has opened new venues for understanding
-microbial evolutionary dynamics with an unprecedented level of temporal
-resolution \autocite{levy2015,nguyenba2019a,ascensao2023}. These
-experimental efforts rely on our ability to reliably infer the relative
-fitness of an ensemble of diverse genotypes. Moreover, inferring these
-fitness values over an ensemble of environmental conditions can help us
-determine the phenotypic diversity of a rapid adaptation process
-\autocite{kinsler2020}.
-
-As with any other sequencing-based quantification, tracking lineages via
-DNA barcode sequencing is inexorably accompanied by noise sources coming
-from experimental manipulation of the microbial cultures, DNA
-extraction, and sequencing library preparation that involves multiple
-rounds of PCR amplification, and the sequencing process itself. Thus,
-accounting for the uncertainty when inferring the relevant parameters
-from the data is a crucial step to draw reliable conclusions. Bayesian
-statistics presents a paradigm by which one can account for all known
-sources of uncertainty in a principled way \autocite{eddy2004a}. This,
-combined with the development of modern Markov Chain Monte Carlo
-sampling algorithms \autocite{betancourt2017} and approximate
-variational approaches \autocite{kucukelbir2016} have boosted a
-resurgence of Bayesian methods in different fields
-\autocite{efron2013a}.
-
-We present a Bayesian inference pipeline to quantify the uncertainty
-about the parametric information we can extract from high-throughput
-competitive fitness assays given a model of the data generation process
-and experimental data. In these assays, the fitness of an ensemble of
-genotypes is determined relative to a reference genotype
-\autocite{kinsler2020,ascensao2023}. Figure~\ref{fig-01}(A) shows a
-schematic of the experimental procedure in which an initial pool of
-barcoded strains are mixed with a reference strain and inoculated into
-fresh media. After some time---usually, enough time for the culture to
-saturate---an aliquot is transferred to fresh media, while the remaining
-culture is used for DNA sequencing of the lineage barcodes. The
-time-series information of the relative abundance of each lineage, i.e.,
-the barcode frequency depicted in Figure~\ref{fig-01}(B), is used to
-infer the relative fitness---the growth advantage on a per-cycle
-basis---for each lineage with respect to the reference strain. The
-proposed statistical model accounts for multiple sources of uncertainty
-when inferring the lineages' relative fitness values (see
-Section~\ref{sec-experiment} for details on sources of uncertainty
-accounted for by the model). Furthermore, minor changes to the core
-statistical model allow us to account for relevant experimental
-variations of these competition assays. More specifically, in
-Section~\ref{sec-multienv}, we present a variation of the statistical
-model to infer fitness on growth dilution cycles in multiple
-environments with proper error propagation. Furthermore, as described in
-Section~\ref{sec-replicates}, our statistical model can account for
-batch-to-batch differences when jointly analyzing multiple experimental
-replicates using a Bayesian hierarchical model. Finally, a variant of
-these hierarchical models, presented in Section~\ref{sec-genotypes}, can
-account for variability within multiple barcodes mapping to equivalent
-genotypes within the same experiment.
-
-For all the model variations presented in this paper, we benchmark the
-ability of our pipeline to infer relative fitness parameters against
-synthetic data generated from logistic growth simulations with added
-random noise. A \texttt{Julia} package accompanies the present method to
-readily implement the inference pipeline with state-of-the-art
-scientific computing software.
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-\hypertarget{sec-experiment}{%
-\subsection{Experimental setup}\label{sec-experiment}}
-
-The present work is designed to analyze time-series data of relative
-abundance of multiple microbial lineages uniquely identified by a DNA
-barcode \autocite{kinsler2020,ascensao2023}. In these competition
-assays, an ensemble of genotypes is pooled together with an unlabeled
-reference strain that, initially, represents the vast majority
-(\(\geq 90\%\)) of the cells in the culture (see schematic in
-Figure~\ref{fig-01}(A)). Furthermore, a fraction of labeled genotypes
-equivalent to the unlabeled reference strain---hereafter defined as
-\emph{neutral} lineages---are spiked in at a relatively high abundance
-(\(\approx 3-5\%\)). The rest of the culture is left for the ensemble of
-genotypes of interest.
-
-To determine the relative fitness of the ensemble of genotypes, a series
-of growth-dilution cycles are performed on either a single or multiple
-environments. In other words, the cultures are grown for some time;
-then, an aliquot is inoculated into fresh media for the next growth
-cycle. This process is repeated for roughly 4-7 cycles, depending on the
-initial abundances of the mutants and their relative growth rates. The
-DNA barcodes are sequenced at the end of each growth cycle to quantify
-the relative abundance of each of the barcodes. We point the reader to
-\textcite{kinsler2020} for specific details on these assays for \emph{S.
-cerevisiae} and to \textcite{ascensao2023} for equivalent assays for
-\emph{E. coli}. Figure~\ref{fig-01}(B) presents a typical barcode
-trajectory where the black trajectories represent the so-called
-\emph{neutral lineages}, genetically equivalent to the untagged ancestor
-strain that initially dominates the culture. These spiked-in neutral
-lineages simplify the inference problem since the fitness metric of all
-relevant barcodes is quantified with respect to these barcodes---thus
-referred to as \emph{relative fitness}.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/fig01.pdf}
-
-}
-
-\caption{\label{fig-01}\textbf{Typical competitive fitness experiment}.
-(A) Schematic of the typical experimental design to determine the
-competitive fitness of an ensemble of barcoded genotypes. Genotypes are
-pooled together and grown over multiple growth-dilution cycles. At the
-end of each cycle, a sample is processed to generate a library for
-amplicon sequencing. (B) Typical barcode trajectory dataset. From each
-time point, the relative frequency of each barcode is determined from
-the total number of reads. Shades of blue represent different relative
-fitness. Darker gray lines define the typical trajectory of neutral
-lineages.}
-
-\end{figure}
-
-\hypertarget{preliminaries-on-mathematical-notation}{%
-\subsection{Preliminaries on mathematical
-notation}\label{preliminaries-on-mathematical-notation}}
-
-Before jumping directly into the Bayesian inference pipeline, let us
-establish the mathematical notation used throughout this paper. We
-define (column) vectors as underlined lowercase symbols such as
-\begin{equation}\protect\hypertarget{eq-vec_def}{}{
-\underline{x} = \begin{bmatrix}
-    x_1\\
-    x_2\\
-    \vdots\\
-    x_N
-\end{bmatrix}.
-}\label{eq-vec_def}\end{equation} In the same way, we define matrices as
-double-underline uppercase symbols such as
-\begin{equation}\protect\hypertarget{eq-mat_def}{}{
-\underline{\underline{A}} =
-\begin{bmatrix}
-    A_{11} & A_{12} & \cdots & A_{1N}\\
-    A_{21} & A_{22} & \cdots & A_{2N}\\
-    \vdots & \vdots & \ddots & \vdots\\
-    A_{M1} & A_{M2} & \cdots & A_{MN}\\
-\end{bmatrix}.
-}\label{eq-mat_def}\end{equation}
-
-\hypertarget{sec-fitness_model}{%
-\subsection{Fitness model}\label{sec-fitness_model}}
-
-Empirically, each barcode frequency trajectory follows an exponential
-function of the form \autocite{levy2015,kinsler2020,ascensao2023}
-\begin{equation}\protect\hypertarget{eq-fitness}{}{
-f_{t+1}^{(b)} = f_{t}^{(b)} \mathrm{e}^{(s^{(b)} - \bar{s}_t)\tau},
-}\label{eq-fitness}\end{equation} where \(f_{t}^{(b)}\) is the frequency
-of barcode \(b\) at the end of cycle number \(t\), \(s^{(b)}\) is the
-relative fitness with respect to the reference strain---the quantity we
-want to infer from the data---\(\bar{s}_t\) is the mean fitness of the
-culture at the end of cycle number \(t\), and \(\tau\) is the time pass
-between cycle \(t\) and \(t+1\). We can rewrite
-Equation~\ref{eq-fitness} as
-\begin{equation}\protect\hypertarget{eq-logfreq}{}{
-\frac{1}{\tau}\ln \frac{f_{t+1}^{(b)}}{f_{t}^{(b)}} = (s^{(b)} - \bar{s}_t).
-}\label{eq-logfreq}\end{equation} Equation~\ref{eq-logfreq} separates
-the measurements---the barcode frequencies---from the unobserved
-(sometimes referred to as latent) parameters we want to infer from the
-data---the population mean fitness and the barcode relative fitness.
-This is ultimately the functional form used in our inference pipeline.
-Therefore, the relative fitness is computed by knowing the log frequency
-ratio of each barcode throughout the growth-dilution cycles.
-
-The presence of the neutral lineages facilitates the determination of
-the population mean fitness value \(\bar{s}_t\). Since every relative
-fitness is determined relative to the neutral lineage that dominates the
-culture, we define their fitness to be \(s^{(n)} = 0\), where the
-superscript \((n)\) specifies their neutrality. This means that
-Equation~\ref{eq-logfreq} for a neutral lineage takes the simpler form
-\begin{equation}\protect\hypertarget{eq-logfreq_neutral}{}{
-\frac{1}{\tau}\ln \frac{f_{t+1}^{(n)}}{f_{t}^{(n)}} = - \bar{s}_t.
-}\label{eq-logfreq_neutral}\end{equation} Therefore, we can use the data
-from these reference barcodes to directly infer the value of the
-population mean fitness.
-
-It is important to notice that the frequencies \(f_{t}^{(b)}\) are not
-the allele frequencies in the population (most of the culture is not
-sequenced since the reference strain is not barcoded), but rather the
-relative frequencies in the total number of sequencing reads. A way to
-conceptualize this subtle but important point is to assume exponential
-growth in the \emph{number of cells} \(N_t^{(b)}\) of the form
-\begin{equation}\protect\hypertarget{eq-ncells}{}{
-N_{t+1}^{(b)} = N_{t}^{(b)} \mathrm{e}^{\lambda^{(b)}\tau},
-}\label{eq-ncells}\end{equation} for every barcode \(b\) with growth
-rate \(\lambda^{(b)}\). However, when we sequence barcodes, we do not
-directly measure the number of cells, but some number of reads
-\(r_t^{(b)}\) that map to barcode \(b\). In the simplest possible
-scenario, we assume \begin{equation}\protect\hypertarget{eq-r_to_n}{}{
-r_{t}^{(b)} \propto N_{t}^{(b)},
-}\label{eq-r_to_n}\end{equation} where, importantly, the proportionality
-constant depends on the total number of reads for the library for cycle
-\(t\), which might vary from library to library. Therefore, to compare
-the number of reads between libraries at different time points, we must
-normalize the number of reads to the same scale. The simplest form is to
-define a relative abundance, i.e., a frequency with respect to the total
-number of reads, \begin{equation}\protect\hypertarget{eq-r_to_n_norm}{}{
-f_{t}^{(b)} \equiv \frac{r_{t}^{(b)}}{\sum_{b'} r_{t}^{(b')}}.
-}\label{eq-r_to_n_norm}\end{equation} This is the frequency
-Equation~\ref{eq-fitness} describes.
-
-Our ultimate objective is to infer the relative fitness \(s^{(b)}\) for
-each of the \(M\) relevant barcodes in the experiment---hereafter
-referred to as \(s^{(m)}\) to distinguish from the general \(s^{(b)}\)
-and the neutral lineages \(s^{(n)}\) relative fitness. To do so, we
-account for the three primary sources of uncertainty in our model:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Uncertainty in the determination of frequencies. Our model relates
-  frequencies between adjacent growth-dilution cycles to the fitness of
-  the corresponding strain. However, we do not directly measure
-  frequencies. Instead, our data for each barcode consists of a length
-  \(T\) vector of counts \(\underline{r}^{(b)}\) for each of the \(T\)
-  cycles in which the measurements were taken.
-\item
-  Uncertainty in the value of the population mean fitness. We define
-  neutral lineages to have fitness \(s^{(n)} = 0\), helping us anchor
-  the value of the population mean fitness \(\bar{s}_t\) for each pair
-  of adjacent growth cycles. Moreover, we take this parameter as an
-  empirical parameter to be obtained from the data, meaning that we do
-  not impose a functional form that relates \(\bar{s}_t\) to
-  \(\bar{s}_{t+1}\). Thus, we must infer the \(T-1\) values of this
-  population mean fitness with their uncertainty that must be propagated
-  to the value of the mutants' relative fitness.
-\item
-  Uncertainty in each of the mutants' fitness values.
-\end{enumerate}
-
-To account for all these sources of uncertainty in a principled way, in
-the next section, we develop a Bayesian inference pipeline.
-
-\hypertarget{sec-bayesian_inference}{%
-\subsection{Bayesian inference}\label{sec-bayesian_inference}}
-
-As defined in Section~\ref{sec-fitness_model}, our ultimate objective is
-to infer the vector of relative fitness values
-\begin{equation}\protect\hypertarget{eq-fitness_mut_vec}{}{
-\underline{s}^M = (s^{(1)}, s^{(2)}, \ldots, s^{(M)})^\dagger,
-}\label{eq-fitness_mut_vec}\end{equation} where \(^\dagger\) indicates
-the transpose. Our data consists of an \(T \times B\) matrix
-\(\underline{\underline{R}}\), where \(B = M + N\) is the number of
-unique barcodes given by the sum of the number of unique, relevant
-barcodes we care about, \(M\), and the number of unique neutral
-barcodes, \(N\), and \(T\) is the number of growth cycles where
-measurements were taken. The data matrix is then of the form
-\begin{equation}\protect\hypertarget{eq-R_mat}{}{
-\underline{\underline{R}} = \begin{bmatrix}
-- & \underline{r}_1 & - \\
-- & \underline{r}_2 & - \\
- & \vdots & \\
-- & \underline{r}_T & - \\
-\end{bmatrix},
-}\label{eq-R_mat}\end{equation} where each row \(\underline{r}_t\) is a
-\(B\)-dimensional array containing the raw barcode counts at cycle
-\(t\). We can further split each vector \(\underline{r}_t\) into two
-vectors of the form
-\begin{equation}\protect\hypertarget{eq-r-vec_split}{}{
-\underline{r}_t = \begin{bmatrix}
-\underline{r}_t^{N} \\
-\underline{r}_t^{M}
-\end{bmatrix},
-}\label{eq-r-vec_split}\end{equation} i.e., the vector containing the
-neutral lineage barcode counts \(\underline{r}_t^{N}\) and the
-corresponding vector containing the mutant barcode counts
-\(\underline{r}_t^{M}\). Following the same logic, matrix
-\(\underline{\underline{R}}\) can be split into two matrices as
-\begin{equation}\protect\hypertarget{eq-R-mat_split}{}{
-\underline{\underline{R}} = \left[ 
-\underline{\underline{R}}^N \; \underline{\underline{R}}^M
-\right],
-}\label{eq-R-mat_split}\end{equation} where
-\(\underline{\underline{R}}^N\) is a \(T \times N\) matrix with the
-barcode reads time series for each neutral lineage and
-\(\underline{\underline{R}}^M\) is the equivalent \(T \times M\) matrix
-for the non-neutral lineages.
-
-Our objective is to compute the joint probability distribution for all
-relative fitness values given our data. We can express this joint
-posterior distribution using Bayes theorem as
-\begin{equation}\protect\hypertarget{eq-bayes_obj}{}{
-\pi(\underline{s}^M \mid \underline{\underline{R}}) = \frac{
-\pi(\underline{\underline{R}} \mid \underline{s}^M) 
-\pi(\underline{s}^M)}
-{\pi(\underline{\underline{R}})},
-}\label{eq-bayes_obj}\end{equation} where hereafter \(\pi(\cdot)\)
-defines a probability density function. When defining our statistical
-model, we need not to focus on the denominator on the right-hand side of
-Equation~\ref{eq-bayes_obj}. Thus, we can write
-\begin{equation}\protect\hypertarget{eq-bayes_obj_propto}{}{
-\pi(\underline{s}^M \mid \underline{\underline{R}}) \propto
-\pi(\underline{\underline{R}} \mid \underline{s}^M) 
-\pi(\underline{s}^M).
-}\label{eq-bayes_obj_propto}\end{equation} However, when implementing
-the model computationally, the normalization constant on the right-hand
-side of Equation~\ref{eq-bayes_obj} must be computed. This can be done
-from the definition of the model via an integral of the form
-\begin{equation}\protect\hypertarget{eq-bayes_evidence}{}{
-\pi(\underline{\underline{R}}) = \int d^M \underline{s}^M
-\pi(\underline{\underline{R}} \mid \underline{s}^M) 
-\pi(\underline{s}^M),
-}\label{eq-bayes_evidence}\end{equation} also known as a marginalization
-integral. Hereafter, differentials of the form \(d^n\) imply a
-\(n\)-dimensional integral.
-
-Although Equation~\ref{eq-bayes_obj} and
-Equation~\ref{eq-bayes_obj_propto} seem simple enough, recall that
-Equation~\ref{eq-fitness} relates barcode frequency values and the
-population mean fitness to the mutant relative fitness. Therefore, we
-must include these nuisance parameters as part of our inference problem.
-We direct the reader to the supplementary materials for the exact
-definitions of these parameters. Here, it suffices to say that the
-inference problem must include the vector \(\underline{\bar{s}}_T\) of
-all population mean fitness values and the matrix
-\(\underline{\underline{F}}\) of all barcode frequencies within the
-sequencing data. With these nuisance variables in hand, the full
-inference problem we must solve takes the form
-\begin{equation}\protect\hypertarget{eq-bayes_full}{}{
-\pi(
-    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
-    \underline{\underline{R}}
-) \propto
-\pi(
-    \underline{\underline{R}} \mid
-    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
-)
-\pi(
-    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
-).
-}\label{eq-bayes_full}\end{equation} To recover the marginal
-distribution over the non-neutral barcodes relative fitness values, we
-can numerically integrate out all nuisance parameters, i.e.,
-\begin{equation}\protect\hypertarget{eq-bayes_full_marginal}{}{
-\pi(\underline{s}^M \mid \underline{\underline{R}}) =
-\int d^{T-1}\underline{\bar{s}}_T
-\int d^{B}\underline{f}_1 \cdots
-\int d^{B}\underline{f}_T
-\;
-\pi(
-    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
-    \underline{\underline{R}}
-).
-}\label{eq-bayes_full_marginal}\end{equation}
-
-\hypertarget{seq-split_posterior}{%
-\subsubsection{Factorizing the posterior
-distribution}\label{seq-split_posterior}}
-
-The left-hand side of Equation~\ref{eq-bayes_full} is extremely
-difficult to work with. However, we can take advantage of the structure
-of our inference problem to rewrite it in a more manageable form.
-Specifically, the statistical dependencies of our observations and
-latent variables allow us to factorize the joint distribution into the
-product of multiple conditional distributions. To gain some intuition
-about this factorization, let us focus on the inference of the
-population mean fitness values \(\underline{\bar{s}}_T\).
-Equation~\ref{eq-logfreq_neutral} relates the value of the population
-mean fitness to the neutral lineage frequencies and nothing else. This
-suggests that when writing the posterior for these population mean
-fitness parameters, we should be able to condition it only on the
-neutral lineage frequency values, i.e.,
-\(\pi(\underline{\bar{s}}_T \mid \underline{\underline{F}}^N)\). We
-point the reader to Section~\ref{sec-bayes_def} for the full
-mathematical details on this factorization. For our purpose here, it
-suffices to say we can rewrite the joint probability distribution as a
-product of conditional distributions of the form
-\begin{equation}\protect\hypertarget{eq-split_posterior}{}{
-\pi(
-    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
-    \underline{\underline{R}}
-) =
-\pi(
-    \underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}}^M
-)
-\pi(
-    \underline{\bar{s}}_T \mid \underline{\underline{F}}^N
-)
-\pi(\underline{\underline{F}} \mid \underline{\underline{R}}).
-}\label{eq-split_posterior}\end{equation} Written in this form,
-Equation~\ref{eq-split_posterior} captures the three sources of
-uncertainty listed in Section~\ref{sec-fitness_model} in each term.
-Starting from right to left, the first term on the right-hand side of
-Equation~\ref{eq-split_posterior} accounts for the uncertainty when
-inferring the frequency values given the barcode reads. The second term
-accounts for the uncertainty in the values of the mean population
-fitness at different time points. The last term accounts for the
-uncertainty in the parameter we care about---the mutants' relative
-fitnesses. We refer the reader to Section~\ref{sec-bayes_def} for an
-extended description of the model with specific functional forms for
-each term on the left-hand side of Equation~\ref{eq-split_posterior} as
-well as the extension of the model to account for multiple experimental
-replicates or hierarchical genotypes.
-
-\hypertarget{variational-inference}{%
-\subsubsection{Variational Inference}\label{variational-inference}}
-
-One of the technical challenges to the adoption of Bayesian methods is
-the analytical intractability of integrals such as that of
-Equation~\ref{eq-bayes_full_marginal}. Furthermore, even though
-efficient Markov Chain Monte Carlo (MCMC) algorithms such as Hamiltonian
-Montecarlo can numerically perform this integration
-\autocite{betancourt2017}, the dimensionality of the problem in
-Equation~\ref{eq-split_posterior} makes an MCMC-based approach
-prohibitively slow.
-
-To overcome this computational limitation, we rely on the recent
-development of the automatic differentiation variational inference
-algorithm (ADVI) \autocite{kucukelbir2016}. Briefly, when performing
-ADVI, our target posterior distribution
-\(\pi(\theta \mid \underline{\underline{R}})\), where
-\(\theta = (\underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}})\),
-is replaced by an approximate posterior distribution \(q_\phi(\theta)\),
-where \(\phi\) fully parametrizes the approximate distribution. As
-further explained in Section~\ref{sec-vi_primer}, the numerical
-integration problem is replaced by an optimization problem of the form
-\begin{equation}\protect\hypertarget{eq-vi_objective}{}{
-q^*_\phi(\theta) = \min _\phi
-D_{KL}(
-    q_\phi(\theta) \lvert \lvert
-    \pi(\theta \mid \underline{\underline{R}})
-),
-}\label{eq-vi_objective}\end{equation} where \(D_{KL}\) is the
-Kulback-Leibler divergence. In other words, the complicated
-high-dimensional numerical integration problem is transformed into a
-much simpler problem of finding the value of the parameters \(\phi\)
-such that Equation~\ref{eq-vi_objective} is satisfied as best as
-possible within some finite computation time. Although to compute
-Equation~\ref{eq-vi_objective}, we require the posterior distribution we
-are trying to approximate
-\(\pi(\theta \mid \underline{\underline{R}})\), it can be shown that
-maximizing the so-called evidence lower bound (ELBO)
-\autocite{kingma2014}---equivalent to minimizing the variational free
-energy \autocite{gottwald2020}---is mathematically equivalent to
-performing the optimization prescribed by
-Equation~\ref{eq-vi_objective}. We direct the reader to
-Section~\ref{sec-vi_primer} for a short primer on variational inference.
-
-This work is accompanied by the Julia library \texttt{BarBay.jl} that
-makes use of the implementation of both MCMC-based integration as well
-as ADVI optimization to numerically approximate the solution of
-Equation~\ref{eq-bayes_full_marginal} within the Julia ecosystem
-\autocite{ge2018}.
-
-\hypertarget{inference-on-a-single-dataset}{%
-\subsection{Inference on a single
-dataset}\label{inference-on-a-single-dataset}}
-
-To assess the inference pipeline performance, we applied it to a
-simulated dataset with known ground truth relative fitness values (See
-Section~\ref{sec-logistic} for details on simulation).
-Figure~\ref{fig-02}(A) shows the structure of the synthetic dataset. The
-majority of barcodes of interest (faint color lines) are adaptive
-compared to the neutral barcodes (\(s^{(m)} > 0\)). Although the barcode
-frequency trajectories look relatively smooth, our fitness model
-requires the computation of the log frequency ratio between adjacent
-time points as derived in Equation~\ref{eq-logfreq}.
-Figure~\ref{fig-02}(B) shows such data transformation where we can
-better appreciate the observational noise input into our statistical
-model. This noise is evident for the darker lines representing the
-neutral barcodes since all of these lineages are assumed to be
-identically distributed.
-
-To visualize the performance of our inference pipeline in fitting our
-fitness model to the observed data, we compute the so-called posterior
-predictive checks (PPC). In short, the PPC consists of repeatedly
-generating synthetic datasets in agreement with the results from the
-inference results. In other words, we use the resulting parameter values
-from the ADVI inference to generate possible datasets in agreement with
-the inferred values (See Section~\ref{sec-ppc} for further details on
-these computations). Figure~\ref{fig-02}(C) shows these results for all
-neutral lineages (upper left corner plot) and a few representative
-non-neutral barcodes. The different color shades represent the 95\%,
-68\%, and 5\% credible regions, i.e., the regions where we expect to
-find the data with the corresponding probability---or in terms of our
-parameter, the \(X\%\) credible region is the interval where we expect
-the true parameter value to lie with \(X\%\) probability.
-
-The main advantage of our method is this natural interpretability of
-these credible regions where an \(X\%\) credible region indeed captures
-the region of parameter space where we expect with \(X\%\) probability
-the actual value of the parameter lies given our statistical model, our
-prior information, and the observed experimental data. A common mistake
-in the literature is interpreting frequentist confidence intervals as
-Bayesian credible regions when they are not equivalent
-\autocite{morey2016}. Frequentist confidence intervals and Bayesian
-credible regions are based on fundamentally different philosophical
-approaches to statistics. Frequentist confidence intervals represent the
-range of values that would contain the true population parameter with a
-certain probability if the experiment was repeated many times. The
-confidence interval does not represent the probability that the interval
-contains the true value. According to a specific model and prior
-information, Bayesian credible regions represent the range of values
-that contain the parameter with a certain posterior probability. The
-credible region directly represents the probability that the region
-contains the true value. So, frequentist confidence intervals cannot be
-interpreted as Bayesian credible regions because they have fundamentally
-different meanings. Treating an \(X\%\) confidence interval like an
-\(X\%\) credible region is fallacious since confidence intervals do not
-represent probabilistic coverage of the true value like credible
-regions. The intervals are generated through entirely different
-procedures.
-
-To capture the global performance of the model, Figure~\ref{fig-02}(D)
-compares the known ground truth with the inferred relative fitness value
-for all barcodes of interest. There is an excellent degree of
-correspondence between these values, with the error bars representing
-the 68\% credible region for the parameter value crossing the identity
-line for most barcodes. This latter point is made clear with
-Figure~\ref{fig-02}(E) where \(\approx 90\%\) of ground truth fitness
-values fall within one standard deviation of the mean in the inferred
-posterior distributions.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/fig02.pdf}
-
-}
-
-\caption{\label{fig-02}\textbf{Single dataset inference}. (A) Frequency
-trajectories that represent the raw data going into the inference. (B)
-Log frequency ratio between two adjacent time points used by the
-inference pipeline. Darker lines represent the neutral barcodes. These
-transformed data are much more noisy than the seemingly smooth frequency
-trajectories. (C) Examples of the posterior predictive checks for all
-neutral lineages (upper left panel) and a subset of representative
-mutant lineages. Shaded regions represent the 95\%, 68\%, and 5\%
-credible regions for the data. The reported errors above the plot
-represent the 68\% credible region on the mutant relative fitness
-marginal distribution. (D) Comparison between the ground truth fitness
-value from the logistic-growth simulation and the inferred fitness
-value. Gray error bars represent the 68\% posterior credible region for
-the relative fitness values. (E) The empirical cumulative distribution
-function (ECDF) for the absolute z-score value of the ground truth
-parameter value within the inferred fitness posterior distribution.}
-
-\end{figure}
-
-\hypertarget{sec-multienv}{%
-\subsection{Fitness inference on multiple
-environments}\label{sec-multienv}}
-
-The fitness model in Equation~\ref{eq-fitness} relates nuisance
-parameters such as the population mean fitness and the barcode
-frequencies to the relative fitness parameter we want to infer from the
-data. These dependencies imply that uncertainty on the estimates of
-these nuisance parameters influences the inference of the relevant
-parameters. For example, imagine a scenario where the neutral lineages
-data were incredibly noisy, leading to poor estimates of the population
-mean fitness values \(\underline{\bar{s}}_T\). Since the relative
-fitness of any non-neutral barcode \(s^{(m)}\) is determined with
-respect to these neutral barcodes, not accounting for the lack of
-precision in the value of the population mean fitness would result in
-misleading estimates of the accuracy with which we determine the value
-of the parameter we care about. Thus, propagating these sources of
-uncertainty in nuisance parameters is vital to generate an unbiased
-estimate of the relevant information we want to extract from the data.
-One of the benefits of Bayesian methods is the intrinsic error
-propagation embedded in the mathematical framework. For our previous
-example, the uncertainty on the value of the population mean fitness
-values is propagated to the relative fitness of a non-neutral barcode
-since we defined a joint posterior distribution over all parameters as
-fully expressed in Equation~\ref{eq-bayes_full}.
-
-This natural error propagation can help us with the experimental design
-schematized in Figure~\ref{fig-03}(A). Here, rather than performing
-growth-dilution cycles in the same environment, the cells are diluted
-into a different environment. Thus, the uncertainty on the fitness
-estimate for the previous environment must be propagated to that of the
-next one. To validate the extension of our statistical model to this
-scenario, Figure~\ref{fig-03}(B) shows the trajectory of the log
-frequency ratios between adjacent time points. The different colored
-regions correspond to the different environments. For this simulation,
-the growth rate of Environment 2 was set to be, on average, half of the
-average growth rate in Environment 1. Equivalently, the growth rate in
-Environment 3 was set to be, on average, twice the average growth rate
-in Environment 1. Figure~\ref{fig-03}(C-E) show the correspondence
-between the simulation ground truth and the inferred fitness values,
-where the error bars represent the 68\% credible region.
-Figure~\ref{fig-03}(F) summarizes the performance of our inference
-pipeline by showing the empirical cumulative distribution functions for
-the absolute value of the ground truth fitness value z-score within the
-posterior distribution. This plot shows that, overall, \(\approx 75\%\)
-of inferred mean values fall within one standard deviation of the ground
-truth. For completeness, Figure~\ref{fig-03}(G) shows the posterior
-predictive checks for a few example barcodes.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/fig03.pdf}
-
-}
-
-\caption{\label{fig-03}\textbf{Multi-environment fitness inference.} (A)
-Schematic of the simulated experimental design where growth-dilution
-cycles are performed into different environments for each cycle. (B) log
-frequency rations between adjacent time points. Darker lines represent
-the neutral barcodes. The colors in the background demark the
-corresponding environment, matching colors in (A). Environment 2 is set
-to have, on average, half the growth rate of environment 1. Likewise,
-environment 3 is set to have, on average, twice the growth rate of
-environment 1. (C-E) Comparison between the ground truth fitness value
-from the logistic-growth simulation and the inferred fitness value for
-each environment. Gray error bars represent the 68\% posterior credible
-region. (F) The empirical cumulative distribution function (ECDF) for
-the absolute z-score value of the ground truth parameter value within
-the inferred fitness posterior distribution for all fitness values
-(black line) and each environment individually (color lines). (G)
-Examples of the posterior predictive checks for all neutral lineages
-(upper left panel) and a subset of representative mutant lineages.
-Shaded regions surrounding the data represent the 95\%, 68\%, and 5\%
-credible regions for the data. The reported errors above the plot
-represent the 68\% credible region on the mutant relative fitness
-marginal distribution. Background colors match those of (A).}
-
-\end{figure}
-
-\hypertarget{sec-replicates}{%
-\subsection{Accounting for experimental replicates via hierarchical
-models}\label{sec-replicates}}
-
-Our inference pipeline can be extended to account for multiple
-experimental replicates via Bayesian hierarchical models
-\autocite{betancourt2013}. Briefly, when accounting for multiple
-repeated measurements of the same phenomena, there are two extreme cases
-one can use to perform the data analysis: On the one hand, we can treat
-each measurement as entirely independent, losing the power to utilize
-multiple measurements when trying to learn a single parameter. This can
-negatively impact the inference since, in principle, the value of our
-parameter of interest should not depend on the particular experimental
-replicate in question. However, this approach does not allow us to
-properly ``combine'' the uncertainties in both experiments when
-performing the inference. On the other hand, we can pool all data
-together and treat our different experiments as a single measurement
-with higher coverage. This loses the subtle differences due to biotic
-and abiotic batch effects, effectively halving the data that goes into
-our inference problem.
-
-Hierarchical models present a middle ground between these extremes.
-First, hierarchical models rely on the definition of so-called
-\emph{hyper-parameters}, that capture the parametric inference we are
-interested in---for this inference problem, we have a hyper-fitness
-value \(\theta^{(m)}\) for each non-neutral barcode. Second, each
-experiment draws randomly from the distribution of this hyper-parameter,
-allowing for subtle variability between experiments to be accounted
-for---in the present inference pipeline, each experimental replicate
-gets assigned a \emph{local} fitness value \(s^{(m)}_i\), where the
-extra sub-index indicates the \(i\)-th experimental replicate.
-Conceptually, we can think of the local fitness for replicate \(i\) as
-being sampled from a distribution that depends on the value of the
-global hyper-fitness value, i.e., \(s^{(m)}_i \sim \pi_{\theta^{(m)}}\),
-where the subindex \(\theta^{(m)}\) indicates the distribution's
-parametric dependence on the hyper-fitness value. This way of
-interpreting the connection between the distribution
-\(\pi_{\theta^{(m)}}\) and the local fitness implies that a large
-replicate-to-replicate variability would lead to a broad hyper-fitness
-distribution---implying a large uncertainty when determining the
-parameter that characterizes the overall relative fitness. We point the
-reader to Section~\ref{sec-hierarchical_model} for the full definition
-of the hierarchical model used in this section. Importantly, as
-schematized in Figure~\ref{fig-04}(A), the influence between different
-experimental replicates runs both ways. First, the data from one
-experimental replicate (\(\underline{\underline{R}}^M_k\) in the
-diagram) informs all local fitness values via the global hyper-fitness
-(upper panel in Figure~\ref{fig-04}(A)). Second, the local fitness value
-is informed by the data from all experimental replicates via the same
-global hyper-fitness parameter (lower panel in Figure~\ref{fig-04}(A)).
-
-To test the performance of this model, we simulated two experimental
-replicates with 1000 unique barcodes (see Figure~\ref{fig-04}(B-C))
-where we randomly sampled a ground truth hyper-fitness value
-\(\theta^{(m)}\) for each barcode. We sampled a variation from this
-hyper-fitness value for each experimental replicate \(s^{(m)}_i\) to
-capture experimental batch effects. Figure~\ref{fig-04}(D) shows the
-relationship between hyper-fitness and replicate fitness values for this
-simulation. The spread around the identity line represents the expected
-batch-to-batch variation. The posterior predictive checks examples in
-Figure~\ref{fig-04}(E) show that the hierarchical model can correctly
-fit the data for each experimental replicate. Furthermore,
-Figure~\ref{fig-04}(F-G) show a high correlation between the ground
-truth and the inferred fitness values. The empirical cumulative
-distribution functions shown in Figure~\ref{fig-04}(H-I) reveal that for
-\(\approx 75\%\) of the non-neutral barcodes, the ground truth
-hyper-fitness values fall within one standard deviation from the mean
-value in the posterior distributions.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/fig04.pdf}
-
-}
-
-\caption{\label{fig-04}\textbf{Hierarchical model on experimental
-replicates}. (A) Schematic depiction of the interactions between local
-fitness values \(\underline{s}_k^M\) through the global hyper-fitness
-value \(\underline{\theta}^M\) for \(K\) hypothetical experimental
-replicates. The upper diagram shows how the data from replicate \(k\)
-informs all local fitness values via the hyper-fitness parameter. The
-lower panel shows the reverse, where all other datasets inform the local
-fitness value. (B-C) Simulated replicate datasets with 900 barcodes of
-interest and 100 neutral lineages. (D) Comparison between the simulation
-ground truth hyper-fitness and each replicate ground truth fitness. The
-scatter between parameters captures experimental batch effects. (E)
-Examples of the posterior predictive checks for all neutral lineages
-(upper left panels) and a subset of representative mutant lineages.
-Shaded regions from light to dark represent the 95\%, 68\%, and 5\%
-credible regions. (F-G) Comparison between the simulation's ground truth
-hyper-fitness (F) and replicate fitness (G) values and the inferred
-parameters. Gray error bars represent the 68\% posterior credible
-region. (H-I) The empirical cumulative distribution function (ECDF) for
-the absolute z-score value of the ground truth parameter value within
-the inferred hyper-fitness posterior distribution (H) and replicate
-fitness (I).}
-
-\end{figure}
-
-As shown in Figure~\ref{fig-05}, the structure imposed by the
-hierarchical model schematized in Figure~\ref{fig-04}(A), where we
-explicitly account for the connection between experimental replicates
-can improve the quality of the inference. Inferred fitness values
-between experimental replicates exhibit a stronger degree of correlation
-in the hierarchical model (Figure~\ref{fig-05}(A)) compared to
-conducting inference on each replicate independently
-(Figure~\ref{fig-05}(B)). Moreover, when comparing the inferred
-hyper-fitness values---the objective parameter when performing multiple
-experimental measurements---the hierarchical model outperforms averaging
-the independent experimental replicates as shown in
-Figure~\ref{fig-05}(C) and (D).
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/fig05.pdf}
-
-}
-
-\caption{\label{fig-05}\textbf{Comparison between hierarchical model and
-single dataset model}. (A-B) comparison of inferred fitness values
-between experimental replicates when fitting a hierarchical model (A) or
-independently fitting each dataset (B). Gray error bars represent the
-68\% credible regions. (C) Comparison between the ground truth
-hyper-fitness value and the inferred parameters. The blue dots show the
-inferred hyper-fitness values when assuming a hierarchical model. Gray
-error bars show the 68\% credible region for this inference. The yellow
-dots show the average of the mean inferred fitness values for the two
-experimental replicates. No error bars are shown for these, as it is
-inappropriate to compute one with two data points per non-neutral
-barcode. (D) Empirical cumulative distribution function (ECDF) of the
-absolute difference between the inferred mean and the ground truth
-hyper-fitness.}
-
-\end{figure}
-
-\hypertarget{sec-genotypes}{%
-\subsection{Accounting for multiple barcodes per genotype via
-hierarchical models}\label{sec-genotypes}}
-
-Hierarchical models can also capture another experimental design in
-which multiple barcodes map to the same or an equivalent genotype. As we
-will show, this many-to-one mapping can improve the inference compared
-to the extreme cases of inferring the fitness of each barcode
-independently or pooling the data of all barcodes mapping to a single
-genotype. As schematized in Figure~\ref{fig-06}(A), a small modification
-of the base model allows us to map the structure of our original model
-to that of a hierarchical model with a fitness hyperparameter vector
-\(\underline{\theta}^G\), where \(G\) is the number of genotypes in the
-dataset.
-
-Figure~\ref{fig-06}(B) shows a single experimental replicate in which 90
-genotypes were assigned a random number of barcodes (a multinomial
-distribution with a mean of ten barcodes per genotype) for a total of
-900 non-neutral barcodes. To assess the performance of the hierarchical
-model proposed in Figure~\ref{fig-06}(A), we performed inference using
-this hierarchical model, as well as the two extreme cases of ignoring
-the connection between the barcodes belonging to the same
-genotype---equivalent to performing inference using the model presented
-in Figure~\ref{fig-02}(A) over the barcodes---or pooling the data of all
-barcodes belonging to the same genotype into a single count---equivalent
-to performing inference using the model presented in
-Figure~\ref{fig-02}(A) over the pooled barcodes.
-Figure~\ref{fig-06}(C-D) shows the comparison between the simulation
-ground truth and the inferred values for these three cases. Not only do
-the hierarchical model results show higher degrees of correlation with
-the ground truth, but the error bars (representing the 68\% credible
-regions) are smaller, meaning that the uncertainty in the estimate of
-the parameter we care about decreases when using the hierarchical model.
-The improvement in the prediction can be seen in Figure~\ref{fig-06}(F)
-where the empirical cumulative distribution function of the absolute
-difference between the mean inferred value and the simulation ground
-truth is shown for all three inference models. The hierarchical model's
-curve ascends more rapidly, showing that, in general, the inferred
-values are closer to the ground truth. For completeness,
-Figure~\ref{fig-06}(G) shows some examples of how the hierarchical model
-can capture the raw log-frequency count observations.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/fig06.pdf}
-
-}
-
-\caption{\label{fig-06}\textbf{Hierarchical model for multiple barcodes
-per genotype.} (A) Schematic depiction of the hierarchical structure for
-multiple barcodes mapping to a single genotype. A set of barcodes
-mapping to an equivalent genotype map to ``local'' fitness values
-\(s^{(b)}\) that are connected via a hyper-fitness parameter for the
-genotype \(\theta^{(g)}\). (B) Simulated dataset with 100 neutral
-lineages and 900 barcodes of interest distributed among 90 genotypes.
-(C-E) Comparison between the inferred and ground truth fitness values
-for a hierarchical model (C), a model where each barcode is inferred
-independently (D), and a model where barcodes mapping to the same
-genotype are pooled together (E). Gray error bars represent the 68\%
-credible regions. (F) Empirical cumulative distribution function (ECDF)
-of the absolute difference between the inferred mean and the ground
-truth fitness values for all three models. (G) Examples of the posterior
-predictive checks for all neutral lineages (upper left panels) and a
-subset of representative mutant lineages. Shaded regions from light to
-dark represent the 95\%, 68\%, and 5\% credible regions.}
-
-\end{figure}
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-Experimental evolution of microbial systems has dramatically advanced
-our understanding of the basic principles of biological evolution
-\autocite{kussell2013}. From questions related to the optimal
-fine-tuning of gene expression programs \autocite{Dekel2005}, to the
-dimensionality, geometry, and accessibility of the adaptive fitness
-landscape explored by these rapidly adapting populations
-\autocite{kinsler2020,maeda2020}, to the emergence of eco-evolutionary
-dynamics in a long-term evolution experiment \autocite{good2017}; for
-all of these and other cases, the microbial experimental platform
-combined with high-throughput sequencing has been essential to tackling
-these questions with empirical data. This exciting research area
-promises to improve as new culturing technologies \autocite{jagdish2022}
-as well as more complex lineage barcoding schemes
-\autocite{nguyenba2019a,yang2022}, are adopted.
-
-For this data-heavy field, properly accounting for the uncertainty in
-parameters inferred from experiments is vital to ensure the conclusions
-drawn are reliable. Bayesian statistics presents a principled way to
-quantify this uncertainty systematically \autocite{gelman2010}.
-Moreover, Bayesian analysis offers a more natural way to interpret the
-role that probability theory plays when performing data analysis
-compared to the often-misinterpreted frequentist methods
-\autocite{vanderplas2014}. Nevertheless, the technical challenges
-associated with Bayesian analysis has limited its application. This is
-set to change as recognition of the misuse of frequentist concepts such
-as the p-value is receiving more attention \autocite{nuzzo2014}.
-Moreover, advances in numerical methods such as Hamiltonian Monte Carlo
-\autocite{betancourt2017} and variational inference
-\autocite{kucukelbir2016} allows for complex Bayesian models to be fit
-to empirical data.
-
-In this paper, we present a computational pipeline to analyze
-lineage-tracking time-series data for massive-parallel competition
-assays. More specifically, we fit a Bayesian model to infer the fitness
-of multiple genotypes relative to a reference
-\autocite{kinsler2020,ascensao2023}. The proposed model accounts for
-multiple sources of uncertainty with proper error propagation intrinsic
-to Bayesian methods. To scale the inference pipeline to large datasets
-with \(> 10,000\) barcodes, we use the ADVI algorithm
-\autocite{kucukelbir2016} to fit a variational posterior distribution.
-The main difference between our method and previous inference pipelines,
-such as \textcite{li2023}, is that the present analysis provides
-interpretable errors on the inferred fitness values. The reported
-uncertainty intervals---known as credible regions---can be formally
-interpreted as capturing the corresponding probability mass of finding
-the true value of the parameter given the model, the prior information,
-and the data. Furthermore, minor modifications to the structure of the
-statistical model presented in this work allow for the analysis of
-different experimental designs, such as growth-dilution cycles in
-different environments, joint analysis of multiple experimental
-replicates of the same experiment via hierarchical models, and a
-hierarchical model for multiple barcodes mapping to equivalent
-genotypes. We validate our analysis pipeline on simulated datasets with
-known ground truth, showing that the model fits the data adequately,
-capturing the ground truth parameters within the posterior distribution.
-
-It is important to highlight some of the consequences of the general
-experimental design and the implicit assumptions within the proposed
-statistical model to analyze the resulting data. First, the composition
-of the population is such that the initial fraction of the population
-occupied by the barcoded genotypes is small---usually \textgreater90\%
-of the initial population is the non-labeled reference strain. This
-constraint is important as the fitness model used to fit the time series
-data assumes that the tracked frequencies are \(\ll 1\). Second, when
-computing log frequency ratios, we can run into the issue of dividing by
-zero. This is a common problem when dealing with molecular count data
-\autocite{lovell2020}. Our model gets around this issue by assuming that
-the frequency of any barcode cannot be, but still can get arbitrarily
-close to, zero. Therefore, we implicitly assume that no lineage goes
-extinct during the experiment. Moreover, the statistical model directly
-accounts for the uncertainty associated with having zero barcode counts,
-increasing the corresponding uncertainty. Third, the models presented in
-this paper require the existence of a labeled sub-population of barcoded
-reference strains. These barcodes help determine the fitness baseline,
-as every fitness is quantified with respect to this reference genotype.
-This experimental design constraint facilitates the inference of the
-population mean fitness since most of the culture---the unlabeled
-reference genotype---is not tracked. Finally, the presented statistical
-model assumes that relative fitness is solely a constant of the
-environment and the genotype. Future directions of this work could
-extend the fitness model to properly analyze data with time-varying or
-frequency-dependent fitness values.
-
-In total, the statistical model presented in this work and the software
-package accompanying the paper allow for a principled way of quantifying
-the accuracy with which we can extract relevant parametric information
-from large-scale multiplexed fitness competition assays. Furthermore,
-the implementation of Bayesian models and their fitting via automatic
-differentiation approaches opens the gate to extend this type of formal
-analysis to the data-rich literature in experimental evolution and other
-high-throughput technologies applications.
-
-\hypertarget{supplementary-materials}{%
-\section{Supplementary Materials}\label{supplementary-materials}}
-
-\hypertarget{sec-vi_primer}{%
-\subsection{Primer on Variational Inference}\label{sec-vi_primer}}
-
-In this section, we will briefly introduce the idea behind variational
-inference. Recall that any Bayesian inference problem deals with the
-joint distribution between observations \(\underline{x}\) and unobserved
-latent variables \(\underline{\theta}\). This joint distribution can be
-written as the product of a distribution of the observations
-\(\underline{x}\) conditioned on the \(\underline{\theta}\) and the
-marginal distribution of these latent variables, i.e.,
-\begin{equation}\protect\hypertarget{eq-vi_joint_dist}{}{
-\pi(\underline{x}, \underline{\theta}) =
-\pi(\underline{x} \mid \underline{\theta}) \pi(\underline{\theta}).
-}\label{eq-vi_joint_dist}\end{equation} A Bayesian inference pipeline's
-objective is to compute the latent variables' posterior probability
-given a set of observations. This computation is equivalent to updating
-our prior beliefs about the set of values that the latent variables take
-after taking in new data. We write this as Bayes theorem
-\begin{equation}\protect\hypertarget{eq-vi_bayes_theorem}{}{
-\pi(\underline{\theta} \mid \underline{x}) = 
-\frac{
-        \pi(\underline{x} \mid \underline{\theta})\pi(\underline{\theta})
-    }{
-        \pi(\underline{x})
-    }.
-}\label{eq-vi_bayes_theorem}\end{equation} The main technical challenge
-for working with Equation~\ref{eq-vi_bayes_theorem} comes from the
-computation of the denominator, also known as the \emph{evidence} or the
-\emph{marginalized likelihood}. The reason computing this term is
-challenging is because it involves a (potentially) high-dimensional
-integral of the form
-\begin{equation}\protect\hypertarget{eq-vi_evidence_int}{}{
-\pi(\underline{x}) = 
-\int\cdots\int d^K\underline{\theta}\; \pi(\underline{x}, \underline{\theta}) = 
-\int\cdots\int d^K\underline{\theta}\; 
-\pi(\underline{x} \mid \underline{\theta})
-\pi(\underline{\theta}),
-}\label{eq-vi_evidence_int}\end{equation} where \(K\) is the
-dimesionality of the \(\underline{\theta}\) vector. Here, the integrals
-are taken over the support---the set of values valid for the
-distribution---of \(\pi(\underline{\theta})\). However, only a few
-selected distributions have a closed analytical form; thus, in most
-cases Equation~\ref{eq-vi_evidence_int} must be solved numerically.
-
-Integration in high-dimensional spaces can be computationally extremely
-challenging. For a naive numerical quadrature procedure, integrating
-over a grid of values for each dimension of \(\underline{\theta}\) comes
-with an exponential explosion of the number of required grid point
-evaluations, most of which do not contribute significantly to the
-integration. To gain visual intuition about this challenge, imagine
-integrating the function depicted in Figure~\ref{fig-SI01}. If the
-location of the high-density region (dark peak) is unknown, numerical
-quadrature requires many grid points to ensure we capture this peak.
-However, most of the numerical evaluations of the function on the grid
-points do not contribute significantly to the integral. Therefore, our
-computational resources are wasted on insignificant evaluations. This
-only gets worse as the number of dimensions increases since the number
-of grid point evaluation scales exponentially.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/figSIX_gaussian_peak.png}
-
-}
-
-\caption{\label{fig-SI01}\textbf{High-dimensional numerical quadrature
-does not scale with dimensionality.} Schematic depiction of the problem
-with naive numerical quadrature to integrate over an unknown density.
-While the density is concentrated on the dark peak, most of the
-evaluations over the \(x_1-x_2\) grid do not contribute to the value of
-the integral}
-
-\end{figure}
-
-Modern Markov Chain Monte Carlo algorithms, such as Hamiltonian Monte
-Carlo, can efficiently perform this high-dimensional integration by
-utilizing gradient information from the target density
-\textcite{betancourt2017}. Nevertheless, these sampling-based methods
-become prohibitively slow for the number of dimensions our present
-inference problem presents. Thus, there is a need to find scalable
-methods for the inference problem in Equation~\ref{eq-vi_bayes_theorem}.
-
-Variational inference circumvents these technical challenges by
-proposing an approximate solution to the problem. Instead of working
-with the posterior distribution in its full glory
-\(\pi(\underline{\theta} \mid \underline{x})\), let us propose an
-approximate posterior distribution \(q_\phi\) that belongs to a
-distribution family fully parametrized by \(\phi\). For example, let us
-say that the distribution \(q_\phi\) belongs to the family of
-multivariate Normal distributions such that
-\(\phi = (\underline{\mu}, \underline{\underline{\Sigma}})\), where
-\(\underline{\mu}\) is the vector of means and
-\(\underline{\underline{\Sigma}}\) is the covariance matrix. If we
-replace \(\pi\) by \(q_\phi\), we want \(q_\phi\) to resemble the
-original posterior as much as possible. Mathematically, this can be
-expressed as minimizing a ``\emph{distance metric}''---the
-Kullback-Leibler (KL) divergence, for example---between the
-distributions. Note that we use quotation marks because, formally, the
-KL divergence is not a distance metric since it is not symmetric.
-Nevertheless, the variational objective is set to find a distribution
-\(q_\phi^*\) such that
-\begin{equation}\protect\hypertarget{eq-vi_objective}{}{
-q_\phi^*(\underline{\theta}) =
-\min_\phi D_{KL}\left(
-    q_\phi(\underline{\theta}) \vert\vert 
-    \pi(\underline{\theta} \mid \underline{x})
-\right),
-}\label{eq-vi_objective}\end{equation} where \(D_{KL}\) is the KL
-divergence. Furthermore, we highlight that the KL divergence is a
-strictly positive number, i.e.,
-\begin{equation}\protect\hypertarget{eq-vi_kl_pos}{}{
-D_{KL}\left(
-    q_\phi(\underline{\theta}) \vert\vert 
-    \pi(\underline{\theta} \mid \underline{x})
-\right) \geq 0,
-}\label{eq-vi_kl_pos}\end{equation} as this property will become
-important later on.
-
-At first sight, Equation~\ref{eq-vi_objective} does not improve the
-situation but only introduces further technical complications. After
-all, the definition of the KL divergence
-\begin{equation}\protect\hypertarget{eq-vi_kl_div}{}{
-D_{KL}\left(
-    q_\phi(\underline{\theta}) \vert\vert 
-    \pi(\underline{\theta} \mid \underline{x})
-\right) \equiv 
-\int \cdots \int d^K\underline{\theta}\;
-q_\phi(\underline{\theta})
-\ln \frac{
-    q_\phi(\underline{\theta})
-}{
-    \pi(\underline{\theta} \mid \underline{x})
-},
-}\label{eq-vi_kl_div}\end{equation} includes the posterior distribution
-\(\pi(\underline{\theta} \mid \underline{x})\) we are trying to get
-around. However, let us manipulate Equation~\ref{eq-vi_kl_div} to beat
-it to a more reasonable form. First, we can use the properties of the
-logarithms to write
-\begin{equation}\protect\hypertarget{eq-vi_step01}{}{
-D_{KL}\left(
-    q_\phi(\underline{\theta}) \vert\vert 
-    \pi(\underline{\theta} \mid \underline{x})
-\right) = 
-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln q_\phi(\underline{\theta}) -
-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln \pi(\underline{\theta} \mid \underline{x}),
-}\label{eq-vi_step01}\end{equation} where, for convenience, we write a
-single integration sign (\(d^K\underline{\theta}\;\) still represents a
-multi-dimensional differential). For the second term in
-Equation~\ref{eq-vi_step01}, we can substitute the term inside the
-logarithm using Equation~\ref{eq-vi_bayes_theorem}. This results in
-\begin{equation}\protect\hypertarget{eq-vi_step02}{}{
-\begin{aligned}
-D_{KL}\left(
-    q_\phi(\underline{\theta}) \vert\vert 
-    \pi(\underline{\theta} \mid \underline{x})
-\right) &= 
-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln q_\phi(\underline{\theta}) \\
-&- \int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln \left( 
-    \frac{
-        \pi(\underline{x} \mid \underline{\theta})\pi(\underline{\theta})
-    }{
-        \pi(\underline{x})
-    }
-\right).
-\end{aligned}
-}\label{eq-vi_step02}\end{equation} Again, using the properties of
-logarithms, we can split Equation~\ref{eq-vi_step02}, obtaining
-\begin{equation}\protect\hypertarget{eq-vi_step03}{}{
-\begin{aligned}
-D_{KL}\left(
-    q_\phi(\underline{\theta}) \vert\vert 
-    \pi(\underline{\theta} \mid \underline{x})
-\right) &= 
-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln q_\phi(\underline{\theta}) \\
-&-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln \pi(\underline{x} \mid \underline{\theta}) \\
-&-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln \pi(\underline{\theta}) \\
-&+\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln \pi(\underline{x}).
-\end{aligned}
-}\label{eq-vi_step03}\end{equation} It is convenient to write
-Equation~\ref{eq-vi_step03} as
-\begin{equation}\protect\hypertarget{eq-vi_step04}{}{
-\begin{aligned}
-D_{KL}\left(
-    q_\phi(\underline{\theta}) \vert\vert 
-    \pi(\underline{\theta} \mid \underline{x})
-\right) &= 
-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln \frac{
-    q_\phi(\underline{\theta})
-    }{
-        \pi(\underline{\theta})
-    } \\
-&-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln \pi(\underline{x} \mid \underline{\theta}) \\
-&+ \ln \pi(\underline{x}) 
-\int d^K\underline{\theta}\; q_\phi(\underline{\theta}),
-\end{aligned}
-}\label{eq-vi_step04}\end{equation} where for the last term, we can take
-\(\ln \pi(\underline{x})\) out of the integral since it does not depend
-on \(\underline{\theta}\). Lastly, we utilize two properties:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  The proposed approximate distribution must be normalized, i.e.,
-  \begin{equation}\protect\hypertarget{eq-vi_q_norm}{}{
-  \int d^K\underline{\theta}\; q_\phi(\underline{\theta}) = 1.
-  }\label{eq-vi_q_norm}\end{equation}
-\item
-  The law of the unconscious statistician (LOTUS) establishes that for
-  any probability density function, it must be true that
-  \begin{equation}\protect\hypertarget{eq-vi_lotus}{}{
-  \int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-  f(\underline{\theta}) = \left\langle 
-   f(\underline{\theta}) 
-  \right\rangle_{q_\phi},
-  }\label{eq-vi_lotus}\end{equation} where
-  \(\left\langle\cdot\right\rangle_{q_\phi}\) is the expected value over
-  the \(q_\phi\) distribution.
-\end{enumerate}
-
-Using these two properties, the positivity constraint on the KL
-divergence in Equation~\ref{eq-vi_kl_pos}, and the definition of the KL
-divergence in Equation~\ref{eq-vi_kl_div} we can rewrite
-Equation~\ref{eq-vi_step04} as
-\begin{equation}\protect\hypertarget{eq-vi_step05}{}{
-D_{KL}\left( 
-    q_\phi(\underline{\theta}) \vert \vert
-    \pi(\underline{\theta}) 
-\right) -
-\left\langle
-    \ln \pi(\underline{x} \mid \underline{\theta})
-\right\rangle_{q_\phi}
-\geq - \ln \pi(\underline{x}).
-}\label{eq-vi_step05}\end{equation} Multiplying by a minus one, we have
-the functional form of the so-called evidence lower bound (ELBO)
-\textcite{kingma2014},
-\begin{equation}\protect\hypertarget{eq-vi_elbo}{}{
-\underbrace{
-    \ln \pi(\underline{x})
-}_{\text{log evidence}} \geq
-\underbrace{
-    \left\langle
-        \ln \pi(\underline{x} \mid \underline{\theta})
-    \right\rangle_{q_\phi} -
-    D_{KL}\left( 
-        q_\phi(\underline{\theta}) \vert \vert
-        \pi(\underline{\theta}) 
-    \right)
-}_{\text{ELBO}}.
-}\label{eq-vi_elbo}\end{equation}
-
-Let us recapitulate where we are. We started by presenting the challenge
-of working with Bayes' theorem, as it requires a high-dimensional
-integral of the form in Equation~\ref{eq-vi_evidence_int}. As an
-alternative, variational inference posits to approximate the posterior
-distribution \(\pi(\underline{\theta} \mid \underline{x})\) with a
-parametric distribution \(q_\phi(\underline{\theta})\). By minimizing
-the KL divergence between these distributions, we arrive at the result
-in Equation~\ref{eq-vi_elbo}, where the left-hand side---the log
-marginalized likelihood or log evidence---we cannot compute for
-technical/computational reasons. However, the right-hand side is
-composed of things we can easily evaluate. We can easily evaluate the
-log-likelihood \(\ln \pi(\underline{x} \mid \underline{\theta})\) and
-the KL divergence between our proposed approximate distribution
-\(q_\phi(\underline{\theta})\) and the prior distribution
-\(\pi(\underline{\theta})\). Moreover, we can compute the gradients of
-these functions with respect to the parameters of our proposed
-distribution. This last point implies that we can change the parameters
-of the proposed distribution to maximize the ELBO. And, although we
-cannot compute the left-hand side of Equation~\ref{eq-vi_elbo}, we know
-that however large we make the ELBO, it will always be smaller than (or
-equal) the log-marginal likelihood. Therefore, the larger we can make
-the ELBO by modifying the parameters \(\phi\), the closer it gets to the
-log-marginal likelihood, and, as a consequence, the better our proposed
-distribution \(q_\phi(\underline{\theta})\) gets to the true posterior
-distribution \(\pi(\underline{\theta} \mid \underline{x})\).
-
-In this sense, variational inference turns the intractable numerical
-integration problem to an optimization routine, for which there are
-several algorithms available.
-
-\hypertarget{advi-algorithm}{%
-\subsubsection{ADVI algorithm}\label{advi-algorithm}}
-
-To maximize the right-hand side of Equation~\ref{eq-vi_elbo}, the
-Automatic Differentiation Variational Inference (ADVI) algorithm
-developed in \autocite{kucukelbir2016} takes advantage of advances in
-probabilistic programming languages to generate a robust method to
-perform this optimization. Without going into the details of the
-algorithm implementation, for our purposes, it suffices to say that we
-define our joint distribution \(\pi(\underline{\theta}, \underline{x})\)
-as the product defined in Equation~\ref{eq-vi_joint_dist}. ADVI then
-proposes an approximate variational distribution \(q_\phi\) that can
-either be a multivariate Normal distribution with a diagonal covariance
-matrix, i.e., \begin{equation}\protect\hypertarget{eq-vi_meanfield}{}{
-\phi = (\underline{\mu}, \underline{\underline{D}}),
-}\label{eq-vi_meanfield}\end{equation} where
-\(\underline{\underline{D}}\) is the identity matrix, with the diagonal
-elements given by the vector of variances \(\underline{\sigma}^2\) for
-each variable or a full-rank multivariate Normal distribution
-\begin{equation}\protect\hypertarget{eq-vi_full}{}{
-\phi = (\underline{\mu}, \underline{\underline{\Sigma}}).
-}\label{eq-vi_full}\end{equation}
-
-Then, the parameters are initialized in some value \(\phi_o\). These
-parameters are iteratively updated by computing the gradient of the ELBO
-(right-hand side of Equation~\ref{eq-vi_elbo}), hereafter defined as
-\(\mathcal{L}\), with respect to the parameters,
-\begin{equation}\protect\hypertarget{eq-vi_elbo_gradient}{}{
-\nabla_\phi \mathcal{L} = \nabla_{\underline{\mu}} \mathcal{L} + 
-\nabla_{\underline{\sigma}}\mathcal{L},
-}\label{eq-vi_elbo_gradient}\end{equation} and then computing \[
-\phi_{t+1} = \phi_{t} + \eta \nabla_\phi \mathcal{L},
-\] where \(\eta\) defines the step size.
-
-This short explanation behind the ADVI algorithm is intended only to
-gain intuition on how the optimal variational distribution \(q_\phi\) be
-computed. There are many nuances in the implementation of the ADVI
-algorithm. We invite the reader to look at the original reference for
-further details.
-
-\hypertarget{sec-bayes_def}{%
-\subsection{Defining the Bayesian model}\label{sec-bayes_def}}
-
-In the main text, we specify the inference problem we must solve as
-being of the form
-\begin{equation}\protect\hypertarget{eq-full_inference_SI}{}{
-\pi(
-    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
-    \underline{\underline{R}}
-) \propto
-\pi(
-    \underline{\underline{R}} \mid
-    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
-)
-\pi(
-    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
-).
-}\label{eq-full_inference_SI}\end{equation} Here, we briefly define the
-missing nuisance parameters. Let
-\begin{equation}\protect\hypertarget{eq-pop_fitness_vec}{}{
-\underline{\bar{s}}_T = (\bar{s}_1, \bar{s}_2, \ldots, \bar{s}_{T-1})^\dagger,
-}\label{eq-pop_fitness_vec}\end{equation} be the vector containing the
-\(T-1\) population mean fitness we compute from the \(T\) time points
-where measurements were taken. We have \(T-1\) since the value of any
-\(\bar{s}_t\) requires cycle numbers \(t\) and \(t+1\). Furthermore, let
-the matrix \(\underline{\underline{F}}\) be a \(T \times B\) matrix
-containing all frequency values. As with Equation~\ref{eq-R-mat_split}
-in the main text, we can split \(\underline{\underline{F}}\) into two
-matrices of the form
-\begin{equation}\protect\hypertarget{eq-F-mat_split}{}{
-\underline{\underline{F}} = \left[ 
-\underline{\underline{F}}^N \; \underline{\underline{F}}^M
-\right],
-}\label{eq-F-mat_split}\end{equation} to separate the corresponding
-neutral and non-neutral barcode frequencies.
-
-Let us now define each of the terms in Equation~\ref{eq-split_posterior}
-described in Section~\ref{sec-bayesian_inference} of the main text. The
-following sections will specify the functional form each of these terms
-takes.
-
-\hypertarget{sec-bayes_freq}{%
-\subsubsection{\texorpdfstring{Frequency uncertainty
-\(\pi(\underline{\underline{F}} \mid \underline{\underline{R}})\)}{Frequency uncertainty \textbackslash pi(\textbackslash underline\{\textbackslash underline\{F\}\} \textbackslash mid \textbackslash underline\{\textbackslash underline\{R\}\})}}\label{sec-bayes_freq}}
-
-We begin with the probability of the frequency values given the raw
-barcode reads. The first assumption is that the inference of the
-frequency values for time \(t\) is independent of any other time.
-Therefore, we can write the joint probability distribution as a product
-of independent distributions of the form
-\begin{equation}\protect\hypertarget{eq-freq_indep}{}{
-\pi(\underline{\underline{F}} \mid \underline{\underline{R}}) =
-\prod_{t=1}^T \pi(\underline{f}_t \mid \underline{r}_t),
-}\label{eq-freq_indep}\end{equation} where \(\underline{f}_t\) and
-\(\underline{r}_t\) are the \(t\)-th row of the matrix containing all of
-the measurements for time \(t\). We imagine that when the barcode reads
-are obtained via sequencing, the quantified number of reads is a Poisson
-sample from the ``true'' underlying number of barcodes within the pool.
-This translates to assuming that the number of reads for each barcode at
-any time point \(r^{(b)}_t\) is an independent Poisson random variable,
-i.e., \begin{equation}\protect\hypertarget{eq-freq_poisson_reads}{}{
-r^{(b)}_t \sim \operatorname{Poiss}(\lambda^{(b)}_t),
-}\label{eq-freq_poisson_reads}\end{equation} where the symbol
-``\(\sim\)'' is read ``distributed as.'' Furthermore, for a Poisson
-distribution, we have that
-\begin{equation}\protect\hypertarget{eq-freq_poisson_lambda}{}{
-\lambda^{(b)}_t = \left\langle r^{(b)}_t \right\rangle = 
-\left\langle 
-    \left( r^{(b)}_t - \left\langle r^{(b)}_t \right\rangle \right)^2
-\right\rangle,
-}\label{eq-freq_poisson_lambda}\end{equation} where
-\(\left\langle \cdot \right\rangle\) is the expected value. In other
-words the Poisson parameter is equal to the mean and variance of the
-distribution. The Poisson distribution has the convenient property that
-for two Poisson distributed random variables
-\(X \sim \operatorname{Poiss}(\lambda_x)\) and
-\(Y \sim \operatorname{Poiss}(\lambda_y)\), we have that
-\begin{equation}\protect\hypertarget{eq-freq_additivity}{}{
-Z \equiv X + Y \sim \operatorname{Poiss}(\lambda_x + \lambda_y).
-}\label{eq-freq_additivity}\end{equation} This additivity allows us to
-write the total number of reads at time \(t\) \(n_t\) also as a
-Poisson-distributed random variable of the form
-\begin{equation}\protect\hypertarget{eq-freq_poisson_total}{}{
-n_t \sim \operatorname{Poiss}\left( \sum_{b=1}^B \lambda^{(b)}_t \right),
-}\label{eq-freq_poisson_total}\end{equation} where the sum is taken over
-all \(B\) barcodes.
-
-If the total number of reads is given by
-Equation~\ref{eq-freq_poisson_total}, the array with the number of reads
-for each barcode at time \(t\), \(\underline{r}_t\) is then distributed
-as \begin{equation}\protect\hypertarget{eq-freq_multinomial}{}{
-\underline{r}_t \sim \operatorname{Multinomial}(n_t, \underline{f}_t),
-}\label{eq-freq_multinomial}\end{equation} where each of the \(B\)
-entries of the frequency vector \(\underline{f}_t\) is a function of the
-\(\underline{\lambda}_t\) vector, given by
-\begin{equation}\protect\hypertarget{eq-freq_freq_lambda}{}{
-f_t^{(b)} \equiv f_t^{(b)}(\underline{\lambda}_t) = 
-\frac{\lambda_t^{(b)}}{\sum_{b'=1}^B \lambda_t^{(b')}}.
-}\label{eq-freq_freq_lambda}\end{equation} In other words, we can think
-of the \(B\) barcode counts as independent Poisson samples or as a
-single multinomial draw with a random number of total draws, \(n_t\),
-and the frequency vector \(\underline{f}_t\) we are interested in.
-Notice that Equation~\ref{eq-freq_freq_lambda} is a deterministic
-function that connects the Poisson parameters to the frequencies.
-Therefore, we have the equivalence that
-\begin{equation}\protect\hypertarget{eq-freq_f_lambda_equiv}{}{
-\pi(\underline{f}_t \mid \underline{r}_t) = 
-\pi(\underline{\lambda}_t \mid \underline{r}_t),
-}\label{eq-freq_f_lambda_equiv}\end{equation} meaning that the
-uncertainty comes from the \(\underline{\lambda}_t\) vector. By Bayes
-theorem, we therefore write
-\begin{equation}\protect\hypertarget{eq-freq_lambda_bayes}{}{
-\pi(\underline{\lambda}_t \mid n_t, \underline{r}_t) \propto
-\pi(n_t, \underline{r}_t \mid \underline{\lambda}_t) \pi(\underline{\lambda}_t),
-}\label{eq-freq_lambda_bayes}\end{equation} where we explicitly include
-the dependence on \(n_t\). This does not affect the distribution or
-brings more uncertainty because \(\underline{r}_t\) already contains all
-the information to compute \(n_t\) since
-\begin{equation}\protect\hypertarget{eq-freq_n_sum_r}{}{
-n_t = \sum_{b=1}^B r_t^{(b)}.
-}\label{eq-freq_n_sum_r}\end{equation} But adding the variable allows us
-to factorize Equation~\ref{eq-freq_lambda_bayes} as
-\begin{equation}\protect\hypertarget{eq-freq_lambda_bayes_factorized}{}{
-\pi(\underline{\lambda}_t \mid n_t, \underline{r}_t) \propto
-\pi(\underline{r}_t \mid n_t, \underline{\lambda}_t)
-\pi(n_t \mid \underline{\lambda}_t)
-\pi(\underline{\lambda}_t)
-}\label{eq-freq_lambda_bayes_factorized}\end{equation} We then have
-\begin{equation}\protect\hypertarget{eq-freq_r_bayes}{}{
-\underline{r}_t \mid n_t, \underline{\lambda}_t \sim
-\operatorname{Multinomial}(n_t, \underline{f}_t(\underline{\lambda}_t)).
-}\label{eq-freq_r_bayes}\end{equation} Furthermore, we have \[
-n_t \mid \underline{\lambda}_t \sim 
-\operatorname{Poiss}\left(\sum_{b=1}^B \lambda_t^{(b)}\right).
-\]\{\#eq=freq\_n\_bayes\} Finally, for our prior
-\(\pi(\underline{\lambda}_t)\), we first assume each parameter is
-independent, i.e., \[
-\pi(\underline{\lambda}_t) = \prod_{b=1}^B \pi(\lambda_t^{(b)}).
-\] A reasonable prior for each \(\lambda_t^{(b)}\) representing the
-expected number of reads for barcode \(b\) should span several orders of
-magnitude. Furthermore, we assume that no barcode in the dataset ever
-goes extinct. Thus, no frequency can equal zero, facilitating the
-computation of the log frequency ratios needed to infer the relative
-fitness. The log-normal distribution satisfies these constraints;
-therefore, for the prior, we assume
-\begin{equation}\protect\hypertarget{eq-freq_lambda_prior}{}{
-\lambda_t^{(b)} \sim 
-\log\mathcal{N}(\mu_{\lambda_t^{(b)}}, \sigma_{\lambda_t^{(b)}}),
-}\label{eq-freq_lambda_prior}\end{equation} with
-\(\mu_{\lambda_t^{(b)}}, \sigma_{\lambda_t^{(b)}}\) as the user-defined
-parameters that characterize the prior distribution.
-
-\hypertarget{summary}{%
-\paragraph{Summary}\label{summary}}
-
-Putting all the pieces developed in this section together gives a term
-for our inference of the form
-\begin{equation}\protect\hypertarget{eq-freq_final_01}{}{
-\pi(\underline{\underline{F}} \mid \underline{\underline{R}}) \propto
-\prod_{t=1}^T\left\{
-    \pi(\underline{r}_t \mid n_t, \underline{\lambda}_t)
-    \pi(n_t \mid \underline{\lambda}_t)
-    \left[ 
-        \prod_{b=1}^B \pi(\lambda_t^{(b)})
-    \right]
-\right\}
-}\label{eq-freq_final_01}\end{equation} where
-\begin{equation}\protect\hypertarget{eq-freq_final_02}{}{
-\underline{r}_t \mid n_t, \underline{\lambda}_t \sim
-\operatorname{Multinomial}(n_t, \underline{f}_t(\underline{\lambda}_t)),
-}\label{eq-freq_final_02}\end{equation}
-\begin{equation}\protect\hypertarget{eq-freq_final_03}{}{
-n_t \mid \underline{\lambda}_t \sim 
-\operatorname{Poiss}\left(\sum_{b=1}^B \lambda_t^{(b)}\right).
-}\label{eq-freq_final_03}\end{equation} and
-\begin{equation}\protect\hypertarget{eq-freq_final_04}{}{
-\lambda_t^{(b)} \sim 
-\log\mathcal{N}(\mu_{\lambda_t^{(b)}}, \sigma_{\lambda_t^{(b)}}),
-}\label{eq-freq_final_04}\end{equation}
-
-\hypertarget{sec-bayes_meanfit}{%
-\subsubsection{\texorpdfstring{Population mean fitness uncertainty
-\(\pi(\underline{\bar{s}}_T \mid \underline{\underline{F}}, \underline{\underline{R}})\)}{Population mean fitness uncertainty \textbackslash pi(\textbackslash underline\{\textbackslash bar\{s\}\}\_T \textbackslash mid \textbackslash underline\{\textbackslash underline\{F\}\}, \textbackslash underline\{\textbackslash underline\{R\}\})}}\label{sec-bayes_meanfit}}
-
-Next, we turn our attention to the problem of determining the population
-mean fitnesses \(\underline{\bar{s}}_T\). First, we notice that our
-fitness model in Equation~\ref{eq-fitness} does not include the value of
-the raw reads. They enter the calculation indirectly through the
-inference of the frequency values we developed in
-Section~\ref{sec-bayes_freq}. This means that we can remove the
-conditioning of the value of \(\underline{\bar{s}}_T\) on the number of
-reads, obtaining a simpler probability function
-\begin{equation}\protect\hypertarget{eq-meanfit_noreads}{}{
-\pi(
-    \underline{\bar{s}}_T \mid 
-    \underline{\underline{F}}, \underline{\underline{R}}
-) = 
-\pi(
-    \underline{\bar{s}}_T \mid 
-    \underline{\underline{F}}
-).
-}\label{eq-meanfit_noreads}\end{equation} Moreover, our fitness model
-does not directly explain how the population mean fitness evolves over
-time. In other words, our model cannot explicitly compute the population
-mean fitness at time \(t+1\) from the information we have about time
-\(t\). Given this model limitation, we are led to assume that we must
-infer each \(\bar{s}_t\) independently. Expressing this for our
-inference results in
-\begin{equation}\protect\hypertarget{eq-meanfit_indep}{}{
-\pi(
-    \underline{\bar{s}}_T \mid 
-    \underline{\underline{F}}
-) =
-\prod_{t=1}^{T-1} \pi(\bar{s}_t \mid \underline{f}_t, \underline{f}_{t+1}),
-}\label{eq-meanfit_indep}\end{equation} where we split our matrix
-\(\underline{\underline{F}}\) for each time point and only kept the
-conditioning on the relevant frequencies needed to compute the mean
-fitness at time \(t\).
-
-Although our fitness model in Equation~\ref{eq-fitness} also includes
-the relative fitness \(s^{(m)}\), to infer the population mean fitness
-we only utilize data from the neutral lineages that, by definition, have
-a relative fitness \(s^{(n)} = 0\). Therefore, the conditioning on
-Equation~\ref{eq-meanfit_indep} can be further simplified by only
-keeping the frequencies of the neutral lineages, i.e.,
-\begin{equation}\protect\hypertarget{eq-meanfit_indep_neutral}{}{
-\pi(\bar{s}_t \mid \underline{f}_t, \underline{f}_{t+1}) =
-\pi(\bar{s}_t \mid \underline{f}_t^N, \underline{f}_{t+1}^N).
-}\label{eq-meanfit_indep_neutral}\end{equation}
-
-Recall that in Section~\ref{sec-fitness_model} we emphasized that the
-frequencies \(f_t^{(n)}\) do not represent the true frequency of a
-particular lineage in the population but rather a ``normalized number of
-cells.'' Therefore, it is safe to assume each of the \(N\) neutral
-lineages' frequencies is changing independently. The correlation of how
-increasing the frequency of one lineage will decrease the frequency of
-others is already captured in the model presented in
-Section~\ref{sec-bayes_freq}. Thus, we write
-\begin{equation}\protect\hypertarget{eq-meanfit_indep_lineages}{}{
-\pi(\bar{s}_t \mid \underline{f}_t^N, \underline{f}_{t+1}^N) =
-\prod_{n=1}^N \pi(\bar{s}_t \mid f_t^{(n)}, f_{t+1}^{(n)}).
-}\label{eq-meanfit_indep_lineages}\end{equation}
-
-Now, we can focus on one of the terms on the right-hand side of
-Equation~\ref{eq-meanfit_indep_lineages}. Writing Bayes theorem results
-in \begin{equation}\protect\hypertarget{eq-meanfit_bayes}{}{
-\pi(\bar{s}_t \mid f_t^{(n)}, f_{t+1}^{(n)}) \propto
-\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t) \pi(\bar{s}_t).
-}\label{eq-meanfit_bayes}\end{equation} Notice the likelihood defines
-the joint distribution of neutral barcode frequencies conditioned on the
-population mean fitness. However, rewriting our fitness model in
-Equation~\ref{eq-fitness} for a neutral lineage to leave frequencies on
-one side and fitness on the other results in
-\begin{equation}\protect\hypertarget{eq-fitness_ratio_neutral}{}{
-\frac{f_{t+1}^{(n)}}{f_t^{(n)}} = \mathrm{e}^{- \bar{s}_t\tau}.
-}\label{eq-fitness_ratio_neutral}\end{equation}
-Equation~\ref{eq-fitness_ratio_neutral} implies that our fitness model
-only relates \textbf{the ratio} of frequencies and not the individual
-values. To get around this complication, we define
-\begin{equation}\protect\hypertarget{eq-gamma_def}{}{
-\gamma_t^{(b)} \equiv \frac{f_{t+1}^{(b)}}{f_t^{(b)}},
-}\label{eq-gamma_def}\end{equation} as the ratio of frequencies between
-two adjacent time points for any barcode \(b\). This allows us to
-rewrite the joint distribution
-\(\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t)\) as
-\begin{equation}\protect\hypertarget{eq-joint_freq_gamma}{}{
-\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t) =
-\pi(f_t^{(n)}, \gamma_{t}^{(n)} \mid \bar{s}_t).
-}\label{eq-joint_freq_gamma}\end{equation} Let us rephrase this subtle
-but necessary change of variables since it is a key part of the
-inference problem: our series of independence assumptions lead us to
-Equation~\ref{eq-meanfit_bayes} that relates the value of the population
-mean fitness \(\bar{s}_t\) to the frequency of a neutral barcode at
-times \(t\) and \(t+1\). However, as shown in
-Equation~\ref{eq-fitness_ratio_neutral}, our model functionally relates
-the ratio of frequencies---that we defined as \(\gamma_t^{(n)}\)---and
-not the independent frequencies to the mean fitness. Therefore, instead
-of writing for the likelihood the joint distribution of the frequency
-values at times \(t\) and \(t+1\) conditioned on the mean fitness, we
-write the joint distribution of the barcode frequency at time \(t\) and
-the ratio of the frequencies. These \textbf{must be} equivalent joint
-distributions since there is a one-to-one mapping between
-\(\gamma_t^{(n)}\) and \(f_{t+1}^{(n)}\) for a given value of
-\(f_t^{(n)}\). Another way to phrase this is to say that knowing the
-frequency at time \(t\) and at time \(t+1\) provides the same amount of
-information as knowing the frequency at time \(t\) and the ratio of the
-frequencies. This is because if we want to obtain \(f_{t+1}^{(n)}\)
-given this information, we simply compute
-\begin{equation}\protect\hypertarget{eq-meanfit_ft+1_gamma}{}{
-f_{t+1}^{(n)} = \gamma_t^{(n)} f_t^{(n)}.
-}\label{eq-meanfit_ft+1_gamma}\end{equation}
-
-The real advantage of rewriting the joint distribution as in
-Equation~\ref{eq-joint_freq_gamma} comes from splitting this joint
-distribution as a product of conditional distributions of the form
-\begin{equation}\protect\hypertarget{eq-joint_to_prod_gamma}{}{
-\pi(f_t^{(n)}, \gamma_{t}^{(n)} \mid \bar{s}_t) =
-\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t)
-\pi(\gamma_{t}^{(n)} \mid \bar{s}_t).
-}\label{eq-joint_to_prod_gamma}\end{equation} Written in this form, we
-can finally propose a probabilistic model for how the mean fitness
-relates to the frequency ratios we determine in our experiments. The
-second term on the right-hand side of
-Equation~\ref{eq-joint_to_prod_gamma} relates how the determined
-frequency ratio \(\gamma_t^{(b)}\) relates to the mean fitness
-\(\bar{s}_t\). From Equation~\ref{eq-fitness_ratio_neutral} and
-Equation~\ref{eq-gamma_def}, we can write
-\begin{equation}\protect\hypertarget{eq-log_ratio_neutral}{}{
-\ln \gamma_t^{(n)} = - \bar{s}_t + \varepsilon_t^{(n)},
-}\label{eq-log_ratio_neutral}\end{equation} where, for simplicity, we
-set \(\tau = 1\). Note that we added an extra term,
-\(\varepsilon_t^{(n)}\), characterizing the deviations of the
-measurements from the theoretical model. We assume these errors are
-normally distributed with mean zero and some standard deviation
-\(\sigma_t\), implying that
-\begin{equation}\protect\hypertarget{eq-log_gamma_normal}{}{
-\ln \gamma_t^{(n)} \mid \bar{s}_t, \sigma_t  \sim 
-\mathcal{N}\left(-\bar{s}_t, \sigma_t \right),
-}\label{eq-log_gamma_normal}\end{equation} where we include the nuisance
-parameter \(\sigma_t\) to be determined. If we assume the log frequency
-ratio is normally distributed, this implies the frequency ratio itself
-is distributed log-normal. This means that
-\begin{equation}\protect\hypertarget{eq-meanfit_likelihood}{}{
-\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t  \sim 
-\log \mathcal{N}\left(-\bar{s}_t, \sigma_t \right).
-}\label{eq-meanfit_likelihood}\end{equation} Having added the nuisance
-parameter \(\sigma_t\) implies that we must update
-Equation~\ref{eq-meanfit_bayes} to
-\begin{equation}\protect\hypertarget{eq-meanfit_bayes_full}{}{
-\pi(\bar{s}_t, \sigma_t \mid f_t^{(n)}, f_{t+1}^{(n)}) \propto
-\pi(f_t^{(n)}, \gamma_t^{(n)} \mid \bar{s}_t, \sigma_t) 
-\pi(\bar{s}_t) \pi(\sigma_t),
-}\label{eq-meanfit_bayes_full}\end{equation} where we assume the prior
-for each parameter is independent, i.e.,
-\begin{equation}\protect\hypertarget{eq-meanfit_indep_prior}{}{
-\pi(\bar{s}_t, \sigma_t) = \pi(\bar{s}_t) \pi(\sigma_t).
-}\label{eq-meanfit_indep_prior}\end{equation} For numerical stability,
-we will select weakly-informative priors for both of these parameters.
-In the case of the nuisance parameter \(\sigma_t\), the prior must be
-restricted to positive values only, since standard deviations cannot be
-negative.
-
-For the first term on the right-hand side of
-Equation~\ref{eq-joint_to_prod_gamma},
-\(\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t)\), we remove the
-conditioning on the population mean fitness since it does not add any
-information on top of what the frequency ratio \(\gamma_t^{(n)}\)
-already gives. Therefore, we have
-\begin{equation}\protect\hypertarget{eq-freq_cond_gamma}{}{
-\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t) =
-\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}).
-}\label{eq-freq_cond_gamma}\end{equation} The right-hand side of
-Equation~\ref{eq-freq_cond_gamma} asks us to compute the probability of
-observing a frequency value \(f_t^{(n)}\) given that we get to observe
-the ratio \(\gamma_{t}^{(n)}\). If the ratio happened to be
-\(\gamma_{t}^{(n)} = 2\), we could have \(f_{t+1}^{(n)} = 1\) and
-\(f_{t+1}^{(n)} = 0.5\), for example. Although, it would be equally
-likely that \(f_{t+1}^{(n)} = 0.6\) and \(f_{t+1}^{(n)} = 0.3\) or
-\(f_{t+1}^{(n)} = 0.1\) and \(f_{t+1}^{(n)} = 0.05\) for that matter. If
-we only get to observe the frequency ratio \(\gamma_t^{(n)}\), we know
-that the numerator \(f_{t+1}^{(n)}\) can only take values between zero
-and one, all of them being equally likely given only the information on
-the ratio. As a consequence, the value of the frequency in the
-denominator \(f_{t}^{(n)}\) is restricted to fall in the range
-\begin{equation}\protect\hypertarget{eq-freq_given_gamma_range}{}{
-f_{t}^{(n)} \in \left(0, \frac{1}{\gamma_t^{(n)}} \right].
-}\label{eq-freq_given_gamma_range}\end{equation} A priori, we do not
-have any reason to favor any value over any other, therefore it is
-natural to write
-\begin{equation}\protect\hypertarget{eq-f_given_gamma_uniform}{}{
-f_t^{(n)} \mid \gamma_t^{(n)} \sim 
-\operatorname{Uniform}\left( 0, \frac{1}{\gamma_t^{(n)}} \right).
-}\label{eq-f_given_gamma_uniform}\end{equation}
-
-\hypertarget{summary-1}{%
-\paragraph{Summary}\label{summary-1}}
-
-Putting all the pieces we have developed in this section together
-results in an inference for the population mean fitness values of the
-form \begin{equation}\protect\hypertarget{eq-meanfit_full_inference}{}{
-\pi(
-    \underline{\bar{s}}_T, \underline{\sigma}_T \mid \underline{\underline{F}}
-) \propto
-\prod_{t=1}^{T-1} \left\{
-    \prod_{n=1}^N \left[
-        \pi(f_t^{(n)} \mid \gamma_t^{(n)}) 
-        \pi(\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t)
-    \right]
-    \pi(\bar{s}_t) \pi(\sigma_t)
-\right\},
-}\label{eq-meanfit_full_inference}\end{equation} where we have
-\begin{equation}\protect\hypertarget{eq-meanfit_eq1}{}{
-f_t^{(n)} \mid \gamma_t^{(n)} \sim 
-\operatorname{Uniform} \left(0, \frac{1}{\gamma_t^{(n)}} \right),
-}\label{eq-meanfit_eq1}\end{equation}
-\begin{equation}\protect\hypertarget{eq-meanfit_eq2}{}{
-\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t \sim 
-\log\mathcal{N}(\bar{s}_t, \sigma_t),
-}\label{eq-meanfit_eq2}\end{equation}
-\begin{equation}\protect\hypertarget{eq-meanfit_eq3}{}{
-\bar{s}_t \sim \mathcal{N}(0, \sigma_{\bar{s}_t}),
-}\label{eq-meanfit_eq3}\end{equation} and
-\begin{equation}\protect\hypertarget{eq-meanfit_eq4}{}{
-\sigma_t \sim \log\mathcal{N}(\mu_{\sigma_t}, \sigma_{\sigma_t}),
-}\label{eq-meanfit_eq4}\end{equation} where \(\sigma_{\bar{s}_t}\),
-\(\mu_{\sigma_t}\), and \(\sigma_{\sigma_t}\) are user-defined
-parameters.
-
-\hypertarget{sec-bayes_mutfit}{%
-\subsubsection{\texorpdfstring{Mutant relative fitness uncertainty
-\(\pi(\underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}}, \underline{\underline{R}})\)}{Mutant relative fitness uncertainty \textbackslash pi(\textbackslash underline\{s\}\^{}M \textbackslash mid \textbackslash underline\{\textbackslash bar\{s\}\}\_T, \textbackslash underline\{\textbackslash underline\{F\}\}, \textbackslash underline\{\textbackslash underline\{R\}\})}}\label{sec-bayes_mutfit}}
-
-The last piece of our inference is the piece that we care about the
-most: the probability distribution of all the mutants' relative fitness,
-given the inferred population mean fitness and the frequencies. First,
-we assume that all fitness values are independent of each other. This
-allows us to write
-\begin{equation}\protect\hypertarget{eq-mutfit_indep}{}{
-\pi(
-    \underline{s}^M \mid 
-    \underline{\bar{s}}_T, \underline{\underline{F}}, \underline{\underline{R}}
-) = 
-\prod_{m=1}^M \pi(
-    s^{(m)} \mid
-    \underline{\bar{s}}_T, \underline{\underline{F}}, \underline{\underline{R}}
-).
-}\label{eq-mutfit_indep}\end{equation} Furthermore, as was the case with
-the population mean fitness, our fitness model relates frequencies, not
-raw reads. Moreover, the fitness value of mutant \(m\) only depends on
-the frequencies of such mutant. Therefore, we can simplify the
-conditioning to
-\begin{equation}\protect\hypertarget{eq-mutfit_cond_simple}{}{
-\pi(
-    s^{(m)} \mid
-    \underline{\bar{s}}_T, \underline{\underline{F}}, \underline{\underline{R}}
-) = 
-\pi(s^{(m)} \mid \underline{\bar{s}}_T, \underline{f}^{(m)}),
-}\label{eq-mutfit_cond_simple}\end{equation} where
-\begin{equation}\protect\hypertarget{eq-mutfit_timeseries_def}{}{
-\underline{f}^{(m)} = (f_0^{(m)}, f_1^{(m)}, \ldots, f_T^{(m)})^\dagger,
-}\label{eq-mutfit_timeseries_def}\end{equation} is the vector containing
-the frequency time series for mutant \(m\). Writing Bayes' theorem for
-the right-hand side of Equation~\ref{eq-mutfit_cond_simple} results in
-\begin{equation}\protect\hypertarget{eq-mutfit_bayes}{}{
-\pi(s^{(m)} \mid \underline{\bar{s}}_T, \underline{f}^{(m)}) \propto
-\pi(\underline{f}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)})
-\pi(s^{(m)} \mid \underline{\bar{s}}_T).
-}\label{eq-mutfit_bayes}\end{equation} Notice the conditioning on the
-mean fitness values \(\underline{\bar{s}}_T\) is not inverted since we
-already inferred these values.
-
-Following the logic used in Section~\ref{sec-bayes_meanfit}, let us
-define \begin{equation}\protect\hypertarget{eq-gamma_vec_def}{}{
-\underline{\gamma}^{(m)} = 
-(\gamma_0^{(m)}, \gamma_1^{(m)}, \ldots, \gamma_{T-1}^{m})^\dagger,
-}\label{eq-gamma_vec_def}\end{equation} where each entry
-\(\gamma_t^{(m)}\) is defined by Equation~\ref{eq-gamma_def}. In the
-same way we rewrote the joint distribution between two adjacent time
-point frequencies to the joint distribution between one of the
-frequencies and the ratio of both frequencies in
-Equation~\ref{eq-joint_freq_gamma}, we can rewrite the joint
-distribution of the frequency time series for mutant \(m\) as
-\begin{equation}\protect\hypertarget{eq-joint_freq_gammas_mutfit}{}{
-\pi(\underline{f}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}) =
-\pi(f_0^{(m)}, \underline{\gamma}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}).
-}\label{eq-joint_freq_gammas_mutfit}\end{equation} One can think about
-Equation~\ref{eq-joint_freq_gammas_mutfit} as saying that knowing the
-individual frequencies at each time point contain equivalent information
-as knowing the initial frequency and the subsequent ratios of
-frequencies. This is because if we want to know the value of
-\(f_1^{(m)}\) given the ratios, we only need to compute
-\begin{equation}\protect\hypertarget{eq-f1_from_ratio}{}{
-f_1^{(m)} = \gamma_0^{(m)} f_0^{(m)}.
-}\label{eq-f1_from_ratio}\end{equation} Moreover, if we want to know
-\(f_2^{(m)}\), we have
-\begin{equation}\protect\hypertarget{eq-f2_from_ratio}{}{
-f_2^{(m)} = \gamma_1^{(m)} f_1^{(m)} =
-\gamma_1^{(m)} \left(\gamma_0^{(m)} f_0^{(m)}\right),
-}\label{eq-f2_from_ratio}\end{equation} and so on. We can then write the
-joint distribution on the right-hand side of
-Equation~\ref{eq-joint_freq_gammas_mutfit} as a product of conditional
-distributions of the form
-\begin{equation}\protect\hypertarget{eq-mutfit_joint_to_cond}{}{
-\begin{aligned}
-\pi(f_0^{(m)}, \underline{\gamma}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}) =
-&\pi(
-    f_0^{(m)} \mid 
-    \gamma_0^{(m)}, \ldots, \gamma_{T-1}^{(m)}, \underline{\bar{s}}_T, s^{(m)}
-) \times \\
-&\pi(
-    \gamma_0^{(m)} \mid 
-    \gamma_1^{(m)}, \ldots, \gamma_{T-1}^{(m)}, \underline{\bar{s}}_T, s^{(m)}
-) \times \\
-&\pi(
-    \gamma_1^{(m)} \mid 
-    \gamma_2^{(m)}, \ldots, \gamma_{T-1}^{(m)}, \underline{\bar{s}}_T, s^{(m)}
-) \times \\
-&\vdots \\
-&\pi(
-    \gamma_{T-2}^{(m)} \mid \gamma_{T-1}^{(m)}, \underline{\bar{s}}_T, s^{(m)}
-) \times \\
-&\pi(\gamma_{T-1}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}).
-\end{aligned}
-}\label{eq-mutfit_joint_to_cond}\end{equation} Writing the fitness model
-in Equation~\ref{eq-fitness} as \[
-\gamma_t^{(m)} = \frac{f_{t+1}^{(m)}}{f_t^{(m)}} = 
-\mathrm{e}^{(s^{(m)} - s_t)\tau},
-\] reveals that the value of each of the ratios \(\gamma_t^{(m)}\) only
-depends on the corresponding fitness value \(\bar{s}_t\) and the
-relative fitness \(s^{(m)}\). Therefore, we can remove most of the
-conditioning on the right-hand side of
-Equation~\ref{eq-mutfit_joint_to_cond}, resulting in a much simpler
-joint distribution of the form
-\begin{equation}\protect\hypertarget{eq-mutfit_joint_to_cond_simple}{}{
-\begin{aligned}
-\pi(f_0^{(m)}, \underline{\gamma}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}) =
-&\pi(f_0^{(m)} \mid \gamma_0^{(m)}) \times \\
-&\pi(\gamma_0^{(m)} \mid \bar{s}_0, s^{(m)}) \times \\
-&\pi(\gamma_1^{(m)} \mid \bar{s}_1, s^{(m)}) \times \\
-&\vdots \\
-&\pi(\gamma_{T-2}^{(m)} \mid \bar{s}_{T-2}, s^{(m)}) \times \\
-&\pi(\gamma_{T-1}^{(m)} \mid \bar{s}_{T-1}, s^{(m)}),
-\end{aligned}
-}\label{eq-mutfit_joint_to_cond_simple}\end{equation} where for the
-first term on the right-hand side of
-Equation~\ref{eq-mutfit_joint_to_cond_simple} we apply the same logic as
-in Equation~\ref{eq-freq_cond_gamma} to remove all other dependencies.
-We emphasize that although Equation~\ref{eq-mutfit_joint_to_cond_simple}
-looks like a series of independent inferences, the value of the relative
-fitness \(s^{(m)}\) is shared among all of them. This means that the
-parameter is not inferred individually for each time point, resulting in
-different estimates of the parameter, but each time point contributes
-independently to the inference of a single estimate of \(s^{(m)}\).
-
-Using equivalent arguments to those in Section~\ref{sec-bayes_meanfit},
-we assume \[
-f_0^{(m)} \mid \gamma_0^{(m)} \sim 
-\operatorname{Uniform}\left(0, \frac{1}{\gamma_0^{(m)}} \right),
-\] and \begin{equation}\protect\hypertarget{eq-mutfit_lognormal}{}{
-\gamma_t^{(m)} \mid \bar{s}_t, s^{(m)}, \sigma^{(m)} \sim 
-\log\mathcal{N}\left(s^{(m)} - \bar{s}_t, \sigma^{(m)} \right),
-}\label{eq-mutfit_lognormal}\end{equation} where we add the nuisance
-parameter \(\sigma^{(m)}\) to the inference. Notice that this parameter
-is not indexed by time. This means that we assume the deviations from
-the theoretical prediction do not depend on time, but only on the
-mutant. Adding the nuisance parameter demands us to update
-Equation~\ref{eq-mutfit_bayes} to
-\begin{equation}\protect\hypertarget{eq-mutfit_bayes_full}{}{
-\pi(
-    s^{(m)}, \sigma^{(m)} \mid \underline{\bar{s}}_T, \underline{f}^{(m)}
-) \propto
-\pi(\underline{f}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}, \sigma^{(m)})
-\pi(s^{(m)}) \pi(\sigma^{(m)}),
-}\label{eq-mutfit_bayes_full}\end{equation} where we assume independent
-priors for both parameters. We also removed the conditioning on the
-values of the mean fitness as knowing such values does not change our
-prior information about the possible range of values that the parameters
-can take. As with the priors on Section~\ref{sec-bayes_meanfit}, we will
-assign weakly-informative priors to these parameters.
-
-\hypertarget{summary-2}{%
-\paragraph{Summary}\label{summary-2}}
-
-With all pieces in place, we write the full inference of the relative
-fitness values as
-\begin{equation}\protect\hypertarget{eq-mutfit_full_inference}{}{
-\pi(
-    \underline{s}^M ,\underline{\sigma}^M \mid 
-    \underline{\bar{s}}_T, \underline{\underline{F}}
-) \propto
-\prod_{m=1}^M \left\{ 
-    \pi(f_0^{(m)} \mid \gamma_0^{(m)})
-    \prod_{t=0}^{T-1} \left[
-        \pi(\gamma_t^{(m)} \mid \bar{s}_t, s^{(m)}, \sigma^{(m)})
-    \right]
-    \pi(s^{(m)}) \pi(\sigma^{(m)})
-\right\},
-}\label{eq-mutfit_full_inference}\end{equation} where
-\begin{equation}\protect\hypertarget{eq-mutfit_eq1}{}{
-f_0^{(m)} \mid \gamma_0^{(m)} \sim 
-\operatorname{Uniform}\left(0, \frac{1}{\gamma_0^{(m)}} \right),
-}\label{eq-mutfit_eq1}\end{equation}
-\begin{equation}\protect\hypertarget{eq-mutfit_eq2}{}{
-\gamma_t^{(m)} \mid \bar{s}_t, s^{(m)}, \sigma^{(m)} \sim 
-\log\mathcal{N}\left(s^{(m)} - \bar{s}_t, \sigma^{(m)} \right),
-}\label{eq-mutfit_eq2}\end{equation}
-\begin{equation}\protect\hypertarget{eq-mutfit_eq3}{}{
-s^{(m)} \sim \mathcal{N}(0, \sigma_{s^{(m)}}),
-}\label{eq-mutfit_eq3}\end{equation} and
-\begin{equation}\protect\hypertarget{eq-mutfit_eq4}{}{
-\sigma^{(m)} \sim \log\mathcal{N}(\mu_{\sigma^{(m)}}, \sigma_{\sigma^{(m)}}),
-}\label{eq-mutfit_eq4}\end{equation} where \(\sigma_{s^{(m)}}\),
-\(\mu_{\sigma^{(m)}}\), and \(\sigma_{\sigma^{(m)}}\) are user-defined
-parameters.
-
-\hypertarget{sec-hierarchical_model}{%
-\subsubsection{Hierarchical models for multiple experimental
-replicates}\label{sec-hierarchical_model}}
-
-As detailed in Section~\ref{sec-replicates} of the main text, we define
-a Bayesian hierarchical model to analyze data from multiple experimental
-replicates. The implementation requires only slightly modifying the base
-model detailed in the previous sections. The hierarchical model defines
-a hyper-fitness parameter \(\theta^{(m)}\) for every non-neutral
-barcode. We can thus collect all of the \(M\) hyperparameters in an
-array of the form
-\begin{equation}\protect\hypertarget{eq-hier_hyper_vector}{}{
-\underline{\theta}^M = (\theta^{(1)}, \ldots, \theta^{(M)})^\dagger.
-}\label{eq-hier_hyper_vector}\end{equation} Our data now consists of a
-series of matrices \(\underline{\underline{R}}_{[j]}\), where the
-subindex \([j]\) refers to the \(j\)-th experimental replicate. These
-matrices need not have the same number of rows, as the time points
-measured for each replicate can vary. The statistical model we must
-define is then of the form
-\begin{equation}\protect\hypertarget{eq-hier_bayes_full}{}{
-\begin{aligned}
-\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\},
-    \{\underline{\underline{F}}_{[j]}\} \mid
-    \{\underline{\underline{R}}_{[j]}\}
-) \propto \;
-&\pi(
-    \{\underline{\underline{R}}_{[j]}\} \mid
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\},
-    \{\underline{\underline{F}}_{[j]}\} 
-) \times \\\
-&\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\},
-    \{\underline{\underline{F}}_{[j]}\} 
-)
-\end{aligned}
-}\label{eq-hier_bayes_full}\end{equation} where the parameters within
-curly braces with subindex \([j]\) indicate one set of parameters per
-experimental replicate. For example,
-\begin{equation}\protect\hypertarget{eq-hier_array}{}{
-\{\underline{s}^M_{[j]}\} = \{
-    \underline{s}^M_{[1]}, \underline{s}^M_{[2]}, \ldots \underline{s}^M_{[E]}
-\},
-}\label{eq-hier_array}\end{equation} where \(E\) is the number of
-experimental replicates.
-
-Given the dependencies between the variables, we can factorize
-Equation~\ref{eq-hier_bayes_full} to be of the form
-\begin{equation}\protect\hypertarget{eq-hier_factorized}{}{
-\begin{aligned}
-\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\},
-    \{\underline{\underline{F}}_{[j]}\} \mid
-    \{\underline{\underline{R}}_{[j]}\}
-) =
-&\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\} \mid
-    \{\underline{\bar{s}}_{T[j]}\},
-    \{\underline{\underline{F}}_{[j]}\} 
-) \times \\
-&\pi(
-    \{\underline{\bar{s}}_{T[j]}\} \mid
-    \{\underline{\underline{F}}_{[j]}\}
-) \times \\
-&\pi(
-    \{\underline{\underline{F}}_{[j]}\} \mid
-    \{\underline{\underline{R}}_{[j]}\}
-)
-\end{aligned}
-}\label{eq-hier_factorized}\end{equation} Furthermore, the hierarchical
-structure only connects the replicates via the relative fitness
-parameters. This means that the population mean fitness values and the
-frequencies can be independently inferred for each dataset. This allows
-us to rewrite the right-hand side of Equation~\ref{eq-hier_factorized}
-as
-\begin{equation}\protect\hypertarget{eq-hier_bayes_factorized_replicate}{}{
-\begin{aligned}
-\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\},
-    \{\underline{\underline{F}}_{[j]}\} \mid
-    \{\underline{\underline{R}}_{[j]}\}
-) =
-&\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\} \mid
-    \{\underline{\bar{s}}_{T[j]}\},
-    \{\underline{\underline{F}}_{[j]}\} 
-) \times \\
-&\prod_{j=1}^E \left[
-    \pi(
-        \underline{\bar{s}}_{T[j]} \mid
-        \underline{\underline{F}}_{[j]}
-    )
-    \pi(
-        \underline{\underline{F}}_{[j]} \mid
-        \underline{\underline{R}}_{[j]}
-    )
-\right].
-\end{aligned}
-}\label{eq-hier_bayes_factorized_replicate}\end{equation} The terms
-inside the square brackets in
-Equation~\ref{eq-hier_bayes_factorized_replicate} take the same
-functional form as those derived in Section~\ref{sec-bayes_meanfit} and
-Section~\ref{sec-bayes_freq}. Therefore, to implement the desired
-hierarchical model, we only need to focus on the first term on the
-right-hand side of Equation~\ref{eq-hier_bayes_factorized_replicate}. A
-way to think about the structure of the hierarchical model is as
-follows: imagine each genotype as a ``\emph{true}'' relative fitness
-value. However, every time we perform an experiment, small variations in
-the biotic and abiotic conditions---also known as batch effects---might
-result in small deviations from this value. We model this by defining a
-distribution for the hyper-fitness parameter---the ground truth we are
-interested in---and having each experimental replicate sample from this
-hyper-parameter distribution to determine the ``\emph{local}'' fitness
-value. The wider the hyper-parameter distribution is the more
-variability between experimental replicates.
-
-Writing Bayes' theorem for the first term in
-Equation~\ref{eq-hier_bayes_factorized_replicate} results in
-\begin{equation}\protect\hypertarget{eq-hier_bayes}{}{
-\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\} \mid
-    \{\underline{\underline{F}}_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\}
-) \propto
-\pi(
-    \{\underline{\underline{F}}_{[j]}\} \mid
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\}
-)
-\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\} \mid
-    \{\underline{\bar{s}}_{T[j]}\}
-),
-}\label{eq-hier_bayes}\end{equation} where we leave the conditioning on
-the population mean fitness as we did in Section~\ref{sec-bayes_mutfit}.
-This expression can be simplified in two ways. First, the frequency
-values for each experimental replicate depend directly on the local
-fitness values and the corresponding population mean fitness, as the
-relationship between experimental replicates only occurs through the
-relative fitness values. Therefore, we can write
-\begin{equation}\protect\hypertarget{eq-hier_bayes_exp_indep}{}{
-\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\} \mid
-    \{\underline{\underline{F}}_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\}
-) \propto
-\prod_{j=1}^E\left[
-    \pi(
-        \underline{\underline{F}}_{[j]} \mid
-        \underline{s}^M_{[j]},
-        \underline{\bar{s}}_{T[j]}
-    )
-\right]
-\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\} \mid
-    \{\underline{\bar{s}}_{T[j]}\}
-).
-}\label{eq-hier_bayes_exp_indep}\end{equation} Second, the relationship
-between the hyper-fitness and the local fitness values allows us to
-write their joint distribution as a conditional distribution where local
-fitness values depend on the global hyper-fitness value, obtaining
-\begin{equation}\protect\hypertarget{eq-hier_bayes_exp_indep}{}{
-\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\} \mid
-    \{\underline{\underline{F}}_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\}
-) \propto
-\prod_{j=1}^E\left[
-    \pi(
-        \underline{\underline{F}}_{[j]} \mid
-        \underline{s}^M_{[j]},
-        \underline{\bar{s}}_{T[j]}
-    )
-    \pi(
-        \underline{s}^M_{[j]} \mid
-        \underline{\theta}^M
-    )
-\right] 
-\pi(
-    \underline{\theta}^M
-).
-}\label{eq-hier_bayes_exp_indep}\end{equation} Notice we removed the
-conditioning on the population mean fitness as our prior expectations of
-what the global hyper-fitness or local fitness value might be do not
-depend on these nuisance parameters.
-
-The first term on the right-hand side of
-Equation~\ref{eq-hier_bayes_exp_indep} takes the same functional form as
-the one derived in Section~\ref{sec-bayes_mutfit}. Therefore, all we are
-left with is to determine the functional forms for the hyper--prior
-\(\pi(\underline{\theta}^M)\), and the conditional probability
-\(\pi(\underline{s}^M_{[j]} \mid \underline{\theta}^M)\). In analogy to
-the assumptions used for the fitness values in
-Section~\ref{sec-bayes_mutfit}, we define the value of each
-hyper-fitness as independent. This means that we have
-\begin{equation}\protect\hypertarget{eq-hier_hyper_prior}{}{
-\pi(\underline{\theta}^M) = \prod_{m=1}^M \pi(\theta^{(m)}).
-}\label{eq-hier_hyper_prior}\end{equation} Furthermore, we assume this
-prior is of the form
-\begin{equation}\protect\hypertarget{eq-hier_hyper_prior_02}{}{
-\theta^{(m)} \sim \mathcal{N}(\mu_{\theta^{(m)}}, \sigma_{\theta^{(m)}}),
-}\label{eq-hier_hyper_prior_02}\end{equation} where
-\(\mu_{\theta^{(m)}}\) and \(\sigma_{\theta^{(m)}}\) are user-defined
-parameters encoding the prior expectations on the fitness values.
-
-For the conditional distribution
-\(\pi(\underline{s}^M_{[j]} \mid \underline{\theta}^M)\), we use the
-so-called non-centered parametrization that avoids some of the intrinsic
-degeneracies associated with hierarchical models
-\autocite{betancourt2013}. We invite the reader to check
-\href{https://betanalpha.github.io/assets/case_studies/hierarchical_modeling.html}{this
-excellent blog} explaining the difficulties of working with hierarchical
-models. This non-centered parameterization implies that we introduce two
-nuisance parameters such that the local fitness \(s_{[j]}^{(m)}\) is
-computed as \begin{equation}\protect\hypertarget{eq-hier_noncentered}{}{
-s_{[j]}^{(m)} = \theta^{(m)} + 
-(\tau_{[j]}^{(m)} \times \xi_{[j]}^{(m)}),
-}\label{eq-hier_noncentered}\end{equation} where \(\theta^{(m)}\) is the
-corresponding genotype hyper-fitness value, \(\xi_{[j]}^{(m)}\) is a
-standard normal random variable, i.e.,
-\begin{equation}\protect\hypertarget{eq-hier_stdnormal}{}{
-\xi_{[j]}^{(m)} \sim \mathcal{N}(0, 1),
-}\label{eq-hier_stdnormal}\end{equation} that allows deviations from the
-hyper-fitness value to be either positive or negative, and
-\(\tau_{[j]}^{(m)}\) is a strictly positive random variable that
-characterizes the deviation of the local fitness value from the global
-hyper-fitness. We assume
-\begin{equation}\protect\hypertarget{eq-hier_tau}{}{
-\tau_{[j]}^{(m)} \sim \log\mathcal{N}(
-    \mu_{\tau_{[j]}^{(m)}}, \sigma_{\tau_{[j]}^{(m)}}
-)
-}\label{eq-hier_tau}\end{equation} where \(\mu_{\tau_{[j]}^{(m)}}\) and
-\(\sigma_{\tau_{[j]}^{(m)}}\) are user-defined parameters capturing the
-expected magnitude of the batch effects.
-
-\hypertarget{defining-prior-probabilities}{%
-\subsubsection{Defining prior
-probabilities}\label{defining-prior-probabilities}}
-
-One aspect commonly associated---in both positive and negative ways---to
-Bayesian analysis is the definition of prior probabilities. On the one
-hand, the naive textbook version of Bayesian analysis defines the prior
-as encoding the information we have about the inference in question
-before acquiring any data. This is the ``ideal'' use of priors that,
-whenever possible, should be implemented. On the other hand, for most
-practitioners of Bayesian statistics in the age of big data, the
-definition of prior becomes a tool to ensure the convergence of sampling
-algorithms such as MCMC \autocite{gelman2017}. However, for our
-particular problem, although we deal with large amounts of data
-(inferences can be made for \textgreater{} 10K barcodes over multiple
-time points, resulting in \textgreater{} 100K parameters), each barcode
-has very little data, as they are measured only once per time point over
-\textless{} 10 growth-dilution cycles. Furthermore, it is incredibly
-challenging to understand the noise sources related to culturing
-conditions, DNA extraction, library preparation, etc., and encode them
-into reasonable prior distributions.
-
-Empirically, our approach for this work defined the priors based solely
-on the neutral lineage data, as they represent the only repeated
-measurements of a single genotype in our experimental design. We
-acknowledge that defining the priors after observing the data might be
-considered an incoherent inference. However, as expressed by
-\textcite{gelman2017}
-
-\begin{quote}
-Incoherence is an unavoidable aspect of much real-world data analysis;
-and, indeed, one might argue that as scientists we learn the most from
-the anomalies and reassessments associated with episodes of incoherence.
-\end{quote}
-
-With this in mind, we leave it to the reader to judge the selection of
-priors. Furthermore, the software package associated with this work,
-\texttt{BarBay.jl}, is written so that users can experiment with
-different prior selection criteria that fit their needs. We strongly
-advocate that statistics should not be done in a black-box fit-all tool
-mindset but rather as a formal way to encode the assumptions behind the
-analysis, subject to constructive criticism. With this philosophical
-baggage behind us, let us now focus on how the priors used for this work
-were selected.
-
-\hypertarget{naive-neutral-lineage-based-priors}{%
-\paragraph{Naive neutral lineage-based
-priors}\label{naive-neutral-lineage-based-priors}}
-
-For the base model presented in this work, the user-defined prior
-parameters include the following:
-
-\begin{itemize}
-\item
-  Prior on population mean fitness (one per pair of adjacent time
-  points) \begin{equation}\protect\hypertarget{eq-prior_meanfit_mean}{}{
-  \bar{s}_t \sim \mathcal{N}(\mu_{\bar{s}_t}, \sigma_{{\bar{s}_t}}).
-  }\label{eq-prior_meanfit_mean}\end{equation}
-\item
-  Prior on standard deviation associated with neutral lineages
-  likelihood function (one per pair of adjacent time points)
-  \begin{equation}\protect\hypertarget{eq-prior_meanfit_var}{}{
-  \sigma_t \sim \log\mathcal{N}(\mu_{\sigma_t}, \sigma_{\sigma_t}).
-  }\label{eq-prior_meanfit_var}\end{equation}
-\item
-  Prior on relative fitness (one per non-neutral barcode)
-  \begin{equation}\protect\hypertarget{eq-prior_fit_mean}{}{
-  s^{(m)} \sim \mathcal{N}(\mu_{s^{(m)}}, \sigma_{s^{(m)}}).
-  }\label{eq-prior_fit_mean}\end{equation}
-\item
-  Prior on standard deviation associated with non-neutral lineages
-  likelihood function (one per non-neutral barcode)
-  \begin{equation}\protect\hypertarget{eq-prior_fit_var}{}{
-  \sigma^{(m)} \sim \log\mathcal{N}(\mu_{\sigma^{(m)}}, \sigma_{\sigma^{(m)}})
-  }\label{eq-prior_fit_var}\end{equation}
-\end{itemize}
-
-The \texttt{BarBay.jl} package includes a function \texttt{naive\_prior}
-within the \texttt{stats} module. This function utilizes the data from
-the neutral lineages to determine some of the prior parameters to
-facilitate the inference algorithm's numerical convergence. In
-particular, it defines the population mean fitness parameter
-\(\mu_{\bar{s}_t}\) as
-\begin{equation}\protect\hypertarget{eq-prior_meanfit_mean_data}{}{
-\mu_{\bar{s}_t} = \frac{1}{N} \sum_{n=1}^N 
--\ln\left( \frac{r_{t+1}^{(n)}}{r_t^{(n)}} \right),
-}\label{eq-prior_meanfit_mean_data}\end{equation} where \(N\) is the
-number of neutral lineages and \(r_t^{(n)}\) is the number of neutral
-lineages. In other words, it defines the mean of the prior distribution
-as the mean of what one naively would compute from the neutral lineages,
-discarding cases where the ratio diverges because the denominator
-\(r_t^{(n)} = 0\). For the variance parameter, we chose a value
-\(\sigma_{\bar{s}_t} = 0.05\).
-
-Furthermore, the \texttt{naive\_prior} function defines the mean of the
-variance parameter as the standard deviation of the log frequency ratios
-for the neutral lineages, i.e.,
-\begin{equation}\protect\hypertarget{eq-prior_meanfit_var_data}{}{
-\mu_{\sigma_t} = \sqrt{
-    \operatorname{Var}\left(\frac{r_{t+1}^{(n)}}{r_t^{(n)}}\right)
-},
-}\label{eq-prior_meanfit_var_data}\end{equation} where
-\(\operatorname{Var}\) is the sample variance. This same value was
-utilized for the mean of the non-neutral barcode variance
-\(\mu_{\sigma^{(m)}}\). While we assigne the corresponding variances to
-be \(\sigma_{\sigma_t} = \sigma_{\sigma^{(m)}} = 1\).
-
-\hypertarget{sec-ppc}{%
-\subsection{Posterior predictive checks}\label{sec-ppc}}
-
-Throughout the main text, we allude to the concept of posterior
-predictive checks as a formal way to assess the accuracy of our
-inference pipeline. Here, we explain the mechanics behind the
-computation of these credible regions, given the output of the
-inference.
-
-Bayesian models encode what is known as a \emph{generative model}. This
-statement means that in our definition of the likelihood function and
-the prior distribution, we, as modelers, propose a mathematical function
-that captures all relevant relationships between unobserved (latent)
-variables. Therefore, when these latent variables are input into the
-mathematical model, this function \emph{generates} data that should be,
-in principle, indistinguishable from the real observations if the model
-is a good account of the underlying processes involved in the phenomena
-of interest. This generative model implies that once we run the
-inference process and update our posterior beliefs about the state of
-the latent variables, we can input back the inferred values to our model
-and generate synthetic data. Furthermore, we can repeat this process
-multiple times to compute the range where we expect to observe our data
-conditioned on the accuracy of the model.
-
-For our specific scenario, recall that our objective is to infer the
-relative fitness of a non-neutral lineage \(s^{(m)}\) along with
-nuisance parameters related to the population mean fitness at each
-point, \(\underline{\bar{s}}_t\), and the barcode frequency time series
-\(\underline{f}^{(m)}\). All these variables are related through our
-fitness model (see Section~\ref{sec-fitness_model} in the main text)
-\begin{equation}\protect\hypertarget{eq-ppc_model}{}{
-f_{t+1}^{(m)} = f_{t}^{(m)} \mathrm{e}^{(s^{(m)} - s_t)\tau}.
-}\label{eq-ppc_model}\end{equation} As we saw, it is convenient to
-rewrite Equation~\ref{eq-ppc_model} as
-\begin{equation}\protect\hypertarget{eq-ppc_logratio}{}{
-\frac{1}{\tau}\ln\frac{f_{t+1}^{(m)}}{f_{t}^{(m)}} = (s^{(m)} - s_t).
-}\label{eq-ppc_logratio}\end{equation} Written in this way, we separate
-the quantities we can compute from the experimental observations---the
-left-hand side of Equation~\ref{eq-ppc_logratio} can be computed from
-the barcode reads---from the latent variables.
-
-Although we perform the joint inference over all barcodes in the present
-work, let us focus on the inference task for a single barcode as if it
-were computed independently. For a non-neutral barcode, our task
-consists of computing the posterior probability
-\begin{equation}\protect\hypertarget{eq-ppc_posterior}{}{
-\pi(\theta \mid \underline{r}^{(m)}) =
-\pi(
-    s^{(m)}, \sigma^{(m)}, \underline{s}_t, \underline{f}^{(m)}
-    \mid \underline{r}^{(m)}
-),
-}\label{eq-ppc_posterior}\end{equation} where \(\theta\) represents all
-parameters to be inferred and \(\underline{r}^{(m)}\) is the vector with
-the barcode raw counts time series. The list of parameters are
-
-\begin{itemize}
-\tightlist
-\item
-  \(s^{(m)}\): The barcode's relative fitness.
-\item
-  \(\sigma^{(m)}\): A nuisance parameter used in the likelihood to
-  generate the data. This captures the expected deviation from
-  Equation~\ref{eq-ppc_logratio}
-\item
-  \(\underline{s}_t\): The vector with all population mean fitness for
-  each pair of adjacent time points.
-\item
-  \(\underline{f}^{(m)}\): The vector with the barcode frequency time
-  series.
-\end{itemize}
-
-Furthermore, let us define a naive estimate of the barcode frequency at
-time \(t\) as \begin{equation}\protect\hypertarget{eq-ppc_naive_freq}{}{
-\hat{f}_t^{(m)} = \frac{r^{(m)}_t}{\sum_{b=1}^B r^{(b)}_t}.
-}\label{eq-ppc_naive_freq}\end{equation} We can compute this quantity
-from the data by normalizing the raw barcode counts by the sum of all
-barcode counts. Furthermore, we can compute a naive estimate of the log
-frequency ratio from the raw barcode counts as
-\begin{equation}\protect\hypertarget{eq-ppc_naive_logfreq}{}{
-\ln \hat{\gamma}_t^{(m)} = \ln \frac{\hat{f}_{t+1}^{(m)}}{\hat{f}_t^{(m)}}
-}\label{eq-ppc_naive_logfreq}\end{equation}
-
-In our generative model, we assumed
-\begin{equation}\protect\hypertarget{eq-ppc_logfreq_likelihood}{}{
-\ln\gamma^{(m)}_t \mid \theta \sim \mathcal{N}(s^{(m)} - s_t, \sigma^{(m)}).
-}\label{eq-ppc_logfreq_likelihood}\end{equation} This implies that once
-we determine the posterior distribution of our parameters, we can
-generate synthetic values of \(\ln \gamma^{(m)}_t\) that we can then
-compare with the values obtained from applying
-Equation~\ref{eq-ppc_naive_logfreq} and
-Equation~\ref{eq-ppc_naive_logfreq} to the raw data.
-
-In practice, to compute the posterior predictive checks, we generate
-multiple samples from the posterior distribution
-\(\pi(\theta \mid \underline{r}^{(m)})\)
-\begin{equation}\protect\hypertarget{eq-ppc_theta_samples}{}{
-\underline{\theta} = (\theta_1, \theta_2, \ldots, \theta_N).
-}\label{eq-ppc_theta_samples}\end{equation} With these samples in hand,
-the \texttt{BarBay.jl} package includes the function
-\texttt{logfreq\_ratio\_bc\_ppc} for non-neutral barcodes that uses this
-set of posterior parameter samples to generate samples from the
-distribution defined in Equation~\ref{eq-ppc_logfreq_likelihood}. For a
-large-enough number of samples, we can then compute the desired
-percentiles---5, 68, and 95 percentiles in all figures in the main
-text---that are equivalent to the corresponding credible regions. In
-other words, the range of values of \(\ln\gamma_t^{(m)}\) generated by
-this bootstrap process can be used to compute the region where we expect
-to find our raw estimates \(\ln\hat{\gamma}_t^{(m)}\) with the desired
-probability. The package \texttt{BarBay.jl} includes an equivalent
-function, \texttt{logfreq\_ratio\_popmean\_ppc}, for neutral lineages.
-
-\hypertarget{sec-logistic}{%
-\subsection{Logistic growth simulation}\label{sec-logistic}}
-
-In this section, we explain the simulations used to assess the validity
-of our inference pipeline. Let us begin by assuming that, since the
-strains are grown for two full days in the experiment, having left
-behind the exponential phase for almost an entire day, a simple
-exponential growth of the form
-\begin{equation}\protect\hypertarget{eq-logistic_expo}{}{
-\frac{dn_i}{dt} = \lambda_i n_i,
-}\label{eq-logistic_expo}\end{equation} where \(n_i\) is the number of
-cells of strain \(i\), and \(\lambda_i\) is the corresponding growth
-rate is not enough. Instead, we will assume that the cells follow the
-logistic growth equation of the form
-\begin{equation}\protect\hypertarget{eq-logistic_fn}{}{
-\frac{dn_i}{dt} = \lambda_i n_i 
-\left( 1 - \frac{\sum_{j=1}^{N}n_i}{\kappa}\right),
-}\label{eq-logistic_fn}\end{equation} where \(\kappa\) is the carrying
-capacity, and \(N\) is the total number of strains in the culture.
-
-The inference method is based on the model that assumes that the time
-passed between dilutions \(\tau \approx 8\) generations, the change in
-frequency for a mutant barcode can be approximated from cycle \(t\) to
-the next cycle \(t + 1\) as
-\begin{equation}\protect\hypertarget{eq-logistic_fitness_fn}{}{
-f_{t+1}^{(m)} = f_t^{(m)} \mathrm{e}^{(s^{(m)} - \bar{s}_t)\tau},
-}\label{eq-logistic_fitness_fn}\end{equation} where \(s^{(m)}\) is the
-relative fitness for strain \(i\) compared to the ancestral strain and
-\(\bar{s}_t\) is the mean fitness of the population at cycle \(t\). To
-test this assumption, we implemented a numerical experiment following
-the logistic growth model described in Equation~\ref{eq-logistic_fn}.
-Figure~\ref{fig-SIX_logistic_growth_01} shows an example of the
-deterministic trajectories for 50 labeled neutral lineages and 1000
-lineages of interest. The upper red curve that dominates the culture
-represents the unlabeled ancestral strain included in the experimental
-design described in Section~\ref{sec-experiment}.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/figSIX_logistic_growth_01.pdf}
-
-}
-
-\caption{\label{fig-SIX_logistic_growth_01}\textbf{Logistic growth
-simulation over single growth cycle}. The dashed line represents the
-neutral lineages, with the upper curve being the unlabeled neutral
-strain. Color curves represent the genotypes of interest colored by
-growth rate relative to the neutral lineage.}
-
-\end{figure}
-
-To simulate multiple growth-dilution cycles, we take the population
-composition at the final time point and use it to initialize a new
-logistic growth simulation. Figure~\ref{fig-SIX_logistic_growth_02}
-shows the resulting number of cells at the last time point of a cycle
-over multiple growth-dilution cycles for the genotypes in
-@Figure~\ref{fig-SIX_logistic_growth_01}. We can see that the adaptive
-lineages (blue curves) increase in abundance, while detrimental lineages
-(red curves) decrease.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/figSIX_logistic_growth_02.pdf}
-
-}
-
-\caption{\label{fig-SIX_logistic_growth_02}\textbf{Growth-dilution
-cycles for logistic growth simulation}. Each point represents the final
-number of cells after a growth cycle for each lineage. Colors are the
-same as in Figure~\ref{fig-SIX_logistic_growth_01}.}
-
-\end{figure}
-
-In Section~\ref{sec-fitness_model}, we derive the functional form to
-infer the relative fitness of each lineage as
-\begin{equation}\protect\hypertarget{eq-logistic_logfreq}{}{
-\frac{1}{\tau}\ln \frac{f_{t+1}^{(b)}}{f_{t}^{(b)}} = (s^{(b)} - \bar{s}_t).
-}\label{eq-logistic_logfreq}\end{equation}
-Figure~\ref{fig-SIX_logistic_growth_03} shows the corresponding log
-frequency ratio curves for the logistic growth simulation. The
-displacement of these curves with respect to the neutral lineages
-determines the ground truth relative fitness value for these
-simulations.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/figSIX_logistic_growth_03.pdf}
-
-}
-
-\caption{\label{fig-SIX_logistic_growth_03}\textbf{Log frequency ratio
-for logistic growth simulations}. The relative distance of the color
-curves from the black dashed line determines the relative fitness of
-each linage.}
-
-\end{figure}
-
-To simulate the experimental noise, we add two types of noise:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Poisson noise between dilutions. For this, we take the final point of
-  the logistic growth simulation and sample a random Poisson number
-  based on this last point to set the initial condition for the next
-  cycle.
-\item
-  Gaussian noise when performing the measurements. When translating the
-  underlying population composition to the number of reads, we can add a
-  custom amount of Gaussian noise.
-\end{enumerate}
-
-Figure~\ref{fig-SIX_logistic_growth_04} shows the frequency trajectories
-(left panels) and log frequency ratios (right panels) for a noiseless
-simulation (upper panels) and a simulation with added noise (lower
-panels). The noiseless simulation is used to determine the relative
-fitness for each of the lineages, which serves as the ground truth to be
-compared with the resulting inference.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/figSIX_logistic_growth_04.pdf}
-
-}
-
-\caption{\label{fig-SIX_logistic_growth_04}\textbf{Logistic
-growth-dilution simulations with and without noise}}
-
-\end{figure}
-
-\printbibliography
-
-\end{document}
+biotic and abiotic sources complicates the computation of a reliable inference.
+We present a Bayesian pipeline to infer relative microbial fitness from
+high-throughput lineage tracking assays. Our model accounts for multiple sources
+of noise and propagates uncertainties throughout all parameters in a systematic
+way. Furthermore, using modern variational inference methods based on automatic
+differentiation, we are able to scale the inference to a large number of unique
+barcodes. We extend this core model to analyze multi-environment assays,
+replicate experiments, and barcodes linked to genotypes. On simulations, our
+method recovers known parameters within posterior credible intervals. This work
+provides a generalizable Bayesian framework to analyze lineage tracking
+experiments. The accompanying open-source software library enables the adoption
+of principled statistical methods in experimental evolution.

### 2023-10-11 doc/_acknowledgements.qmd
add missing sections to paper

--- 2023-10-11 doc/paper.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
-%
-\documentclass[
-  letterpaper,
-  DIV=11,
-  numbers=noendperiod]{scrartcl}
-
-\usepackage{amsmath,amssymb}
-\usepackage{iftex}
-\ifPDFTeX
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-\usepackage{lmodern}
-\ifPDFTeX\else  
-    % xetex/luatex font selection
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-
-\usepackage{lineno}
-\linenumbers
-\usepackage[font={footnotesize, sf}, labelfont=bf, margin=0.05\linewidth]{caption}
-\renewcommand{\familydefault}{\sfdefault}
-% Bold the 'Figure #' in the caption and separate it from the title/caption
-% with a period
-% Captions will be left justified
-\usepackage[
-	aboveskip=1pt,
-	labelfont=bf,
-	labelsep=period,
-	justification=raggedright,
-	singlelinecheck=off
-]{caption}
-
-% Add numbered lines
-\usepackage{lineno}
-\linenumbers
-
-% Package to include multiple title pages
-% This will allow me to add a tile to the main text and to the SI
-\usepackage{titling}
-
-% Add author affiliations
-\usepackage{authblk}
-
-\KOMAoption{captions}{tableheading}
-\makeatletter
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\AtBeginDocument{%
-\ifdefined\contentsname
-  \renewcommand*\contentsname{Table of contents}
-\else
-  \newcommand\contentsname{Table of contents}
-\fi
-\ifdefined\listfigurename
-  \renewcommand*\listfigurename{List of Figures}
-\else
-  \newcommand\listfigurename{List of Figures}
-\fi
-\ifdefined\listtablename
-  \renewcommand*\listtablename{List of Tables}
-\else
-  \newcommand\listtablename{List of Tables}
-\fi
-\ifdefined\figurename
-  \renewcommand*\figurename{Figure}
-\else
-  \newcommand\figurename{Figure}
-\fi
-\ifdefined\tablename
-  \renewcommand*\tablename{Table}
-\else
-  \newcommand\tablename{Table}
-\fi
-}
-\@ifpackageloaded{float}{}{\usepackage{float}}
-\floatstyle{ruled}
-\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
-\floatname{codelisting}{Listing}
-\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
-\makeatother
-\makeatletter
-\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
-\makeatother
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-
-% References
-% \usepackage[]{biblatex}
-\usepackage[
-	backend=bibtex,
-	style=phys, 
-	sorting=none,				% Do not sort bibliography
-	url=false, 					% Do not show url in reference
-	doi=false, 					% Do not show doi in reference
-	isbn=false, 				% Do not show isbn link in reference
-	eprint=false, 			% Do not show eprint link in reference
-	maxbibnames=9, 			% Include up to 9 names in citation
-	firstinits=true,
-]{biblatex}
-\addbibresource{references.bib}
-
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
-\hypersetup{
-  pdfauthor={Manuel Razo-Mejia; Madhav Mani; Dmitri Petrov},
-  pdfkeywords={bayesian inference, variational inference, relative
-fitness, open-source, reproducible research},
-  colorlinks=true,
-  linkcolor={blue},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
-  pdfcreator={LaTeX via pandoc}}
-
-% Title
-\title{Bayesian inference of relative fitness on high-throughput pooled competition assays}
-
-% Authors
-\author[1]{Manuel Razo-Mejia}
-\author[2,3,4]{Madhav Mani}
-\author[1,5,*]{Dmitri Petrov}
-
-% Affiliations
-\affil[1]{Department of Biology, Stanford University, Stanford, CA, USA}
-\affil[2]{Department of Engineering Sciences and Applied Mathematics, Northwestern University, Chicago, IL, USA}
-\affil[3]{NSF-Simons Center for Quantitative Biology, Northwestern University, Chicago, IL, USA}
-\affil[4]{Department of Molecular Biosciences, Northwestern University, Chicago, IL, USA}
-\affil[5]{Stanford Cancer Institute, Stanford University School of Medicine, Stanford, CA, USA}
-\affil[*]{Correspondence: dpetrov@stanford.edu}
-
-\setcounter{Maxaffil}{0}
-% Set affiliations in small font
-\renewcommand\Affilfont{\itshape\small}
-
-\date{}
-
-\begin{document}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden, interior hidden, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi
-
-\maketitle
-
-\begin{abstract}
-    The tracking of lineage frequencies via DNA barcode sequencing enables the
-  quantification of microbial fitness. However, experimental noise coming from
-  biotic and abiotic sources complicates the computation of a reliable
-  inference. We present a Bayesian pipeline to infer relative microbial fitness
-  from high-throughput lineage tracking assays. Our model accounts for multiple
-  sources of noise and propagates uncertainties throughout all parameters in a
-  systematic way. Furthermore, using modern variational inference methods based
-  on automatic differentiation, we are able to scale the inference to a large
-  number of unique barcodes. We extend this core model to analyze
-  multi-environment assays, replicate experiments, and barcodes linked to
-  genotypes. On simulations, our method recovers known parameters within
-  posterior credible intervals. This work provides a generalizable Bayesian
-  framework to analyze lineage tracking experiments. The accompanying
-  open-source software library enables the adoption of principled statistical
-  methods in experimental evolution.
-\end{abstract}
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-The advent of DNA barcoding---the ability to uniquely identify cell
-lineages with DNA sequences integrated at a specific locus---and
-high-throughput sequencing has opened new venues for understanding
-microbial evolutionary dynamics with an unprecedented level of temporal
-resolution \autocite{levy2015,nguyenba2019a,ascensao2023}. These
-experimental efforts rely on our ability to reliably infer the relative
-fitness of an ensemble of diverse genotypes. Moreover, inferring these
-fitness values over an ensemble of environmental conditions can help us
-determine the phenotypic diversity of a rapid adaptation process
-\autocite{kinsler2020}.
-
-As with any other sequencing-based quantification, tracking lineages via
-DNA barcode sequencing is inexorably accompanied by noise sources coming
-from experimental manipulation of the microbial cultures, DNA
-extraction, and sequencing library preparation that involves multiple
-rounds of PCR amplification, and the sequencing process itself. Thus,
-accounting for the uncertainty when inferring the relevant parameters
-from the data is a crucial step to draw reliable conclusions. Bayesian
-statistics presents a paradigm by which one can account for all known
-sources of uncertainty in a principled way \autocite{eddy2004a}. This,
-combined with the development of modern Markov Chain Monte Carlo
-sampling algorithms \autocite{betancourt2017} and approximate
-variational approaches \autocite{kucukelbir2016} have boosted a
-resurgence of Bayesian methods in different fields
-\autocite{efron2013a}.
-
-We present a Bayesian inference pipeline to quantify the uncertainty
-about the parametric information we can extract from high-throughput
-competitive fitness assays given a model of the data generation process
-and experimental data. In these assays, the fitness of an ensemble of
-genotypes is determined relative to a reference genotype
-\autocite{kinsler2020,ascensao2023}. Figure~\ref{fig-01}(A) shows a
-schematic of the experimental procedure in which an initial pool of
-barcoded strains are mixed with a reference strain and inoculated into
-fresh media. After some time---usually, enough time for the culture to
-saturate---an aliquot is transferred to fresh media, while the remaining
-culture is used for DNA sequencing of the lineage barcodes. The
-time-series information of the relative abundance of each lineage, i.e.,
-the barcode frequency depicted in Figure~\ref{fig-01}(B), is used to
-infer the relative fitness---the growth advantage on a per-cycle
-basis---for each lineage with respect to the reference strain. The
-proposed statistical model accounts for multiple sources of uncertainty
-when inferring the lineages' relative fitness values (see
-Section~\ref{sec-experiment} for details on sources of uncertainty
-accounted for by the model). Furthermore, minor changes to the core
-statistical model allow us to account for relevant experimental
-variations of these competition assays. More specifically, in
-Section~\ref{sec-multienv}, we present a variation of the statistical
-model to infer fitness on growth dilution cycles in multiple
-environments with proper error propagation. Furthermore, as described in
-Section~\ref{sec-replicates}, our statistical model can account for
-batch-to-batch differences when jointly analyzing multiple experimental
-replicates using a Bayesian hierarchical model. Finally, a variant of
-these hierarchical models, presented in Section~\ref{sec-genotypes}, can
-account for variability within multiple barcodes mapping to equivalent
-genotypes within the same experiment.
-
-For all the model variations presented in this paper, we benchmark the
-ability of our pipeline to infer relative fitness parameters against
-synthetic data generated from logistic growth simulations with added
-random noise. A \texttt{Julia} package accompanies the present method to
-readily implement the inference pipeline with state-of-the-art
-scientific computing software.
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-\hypertarget{sec-experiment}{%
-\subsection{Experimental setup}\label{sec-experiment}}
-
-The present work is designed to analyze time-series data of relative
-abundance of multiple microbial lineages uniquely identified by a DNA
-barcode \autocite{kinsler2020,ascensao2023}. In these competition
-assays, an ensemble of genotypes is pooled together with an unlabeled
-reference strain that, initially, represents the vast majority
-(\(\geq 90\%\)) of the cells in the culture (see schematic in
-Figure~\ref{fig-01}(A)). Furthermore, a fraction of labeled genotypes
-equivalent to the unlabeled reference strain---hereafter defined as
-\emph{neutral} lineages---are spiked in at a relatively high abundance
-(\(\approx 3-5\%\)). The rest of the culture is left for the ensemble of
-genotypes of interest.
-
-To determine the relative fitness of the ensemble of genotypes, a series
-of growth-dilution cycles are performed on either a single or multiple
-environments. In other words, the cultures are grown for some time;
-then, an aliquot is inoculated into fresh media for the next growth
-cycle. This process is repeated for roughly 4-7 cycles, depending on the
-initial abundances of the mutants and their relative growth rates. The
-DNA barcodes are sequenced at the end of each growth cycle to quantify
-the relative abundance of each of the barcodes. We point the reader to
-\textcite{kinsler2020} for specific details on these assays for \emph{S.
-cerevisiae} and to \textcite{ascensao2023} for equivalent assays for
-\emph{E. coli}. Figure~\ref{fig-01}(B) presents a typical barcode
-trajectory where the black trajectories represent the so-called
-\emph{neutral lineages}, genetically equivalent to the untagged ancestor
-strain that initially dominates the culture. These spiked-in neutral
-lineages simplify the inference problem since the fitness metric of all
-relevant barcodes is quantified with respect to these barcodes---thus
-referred to as \emph{relative fitness}.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/fig01.pdf}
-
-}
-
-\caption{\label{fig-01}\textbf{Typical competitive fitness experiment}.
-(A) Schematic of the typical experimental design to determine the
-competitive fitness of an ensemble of barcoded genotypes. Genotypes are
-pooled together and grown over multiple growth-dilution cycles. At the
-end of each cycle, a sample is processed to generate a library for
-amplicon sequencing. (B) Typical barcode trajectory dataset. From each
-time point, the relative frequency of each barcode is determined from
-the total number of reads. Shades of blue represent different relative
-fitness. Darker gray lines define the typical trajectory of neutral
-lineages.}
-
-\end{figure}
-
-\hypertarget{preliminaries-on-mathematical-notation}{%
-\subsection{Preliminaries on mathematical
-notation}\label{preliminaries-on-mathematical-notation}}
-
-Before jumping directly into the Bayesian inference pipeline, let us
-establish the mathematical notation used throughout this paper. We
-define (column) vectors as underlined lowercase symbols such as
-\begin{equation}\protect\hypertarget{eq-vec_def}{}{
-\underline{x} = \begin{bmatrix}
-    x_1\\
-    x_2\\
-    \vdots\\
-    x_N
-\end{bmatrix}.
-}\label{eq-vec_def}\end{equation} In the same way, we define matrices as
-double-underline uppercase symbols such as
-\begin{equation}\protect\hypertarget{eq-mat_def}{}{
-\underline{\underline{A}} =
-\begin{bmatrix}
-    A_{11} & A_{12} & \cdots & A_{1N}\\
-    A_{21} & A_{22} & \cdots & A_{2N}\\
-    \vdots & \vdots & \ddots & \vdots\\
-    A_{M1} & A_{M2} & \cdots & A_{MN}\\
-\end{bmatrix}.
-}\label{eq-mat_def}\end{equation}
-
-\hypertarget{sec-fitness_model}{%
-\subsection{Fitness model}\label{sec-fitness_model}}
-
-Empirically, each barcode frequency trajectory follows an exponential
-function of the form \autocite{levy2015,kinsler2020,ascensao2023}
-\begin{equation}\protect\hypertarget{eq-fitness}{}{
-f_{t+1}^{(b)} = f_{t}^{(b)} \mathrm{e}^{(s^{(b)} - \bar{s}_t)\tau},
-}\label{eq-fitness}\end{equation} where \(f_{t}^{(b)}\) is the frequency
-of barcode \(b\) at the end of cycle number \(t\), \(s^{(b)}\) is the
-relative fitness with respect to the reference strain---the quantity we
-want to infer from the data---\(\bar{s}_t\) is the mean fitness of the
-culture at the end of cycle number \(t\), and \(\tau\) is the time pass
-between cycle \(t\) and \(t+1\). We can rewrite
-Equation~\ref{eq-fitness} as
-\begin{equation}\protect\hypertarget{eq-logfreq}{}{
-\frac{1}{\tau}\ln \frac{f_{t+1}^{(b)}}{f_{t}^{(b)}} = (s^{(b)} - \bar{s}_t).
-}\label{eq-logfreq}\end{equation} Equation~\ref{eq-logfreq} separates
-the measurements---the barcode frequencies---from the unobserved
-(sometimes referred to as latent) parameters we want to infer from the
-data---the population mean fitness and the barcode relative fitness.
-This is ultimately the functional form used in our inference pipeline.
-Therefore, the relative fitness is computed by knowing the log frequency
-ratio of each barcode throughout the growth-dilution cycles.
-
-The presence of the neutral lineages facilitates the determination of
-the population mean fitness value \(\bar{s}_t\). Since every relative
-fitness is determined relative to the neutral lineage that dominates the
-culture, we define their fitness to be \(s^{(n)} = 0\), where the
-superscript \((n)\) specifies their neutrality. This means that
-Equation~\ref{eq-logfreq} for a neutral lineage takes the simpler form
-\begin{equation}\protect\hypertarget{eq-logfreq_neutral}{}{
-\frac{1}{\tau}\ln \frac{f_{t+1}^{(n)}}{f_{t}^{(n)}} = - \bar{s}_t.
-}\label{eq-logfreq_neutral}\end{equation} Therefore, we can use the data
-from these reference barcodes to directly infer the value of the
-population mean fitness.
-
-It is important to notice that the frequencies \(f_{t}^{(b)}\) are not
-the allele frequencies in the population (most of the culture is not
-sequenced since the reference strain is not barcoded), but rather the
-relative frequencies in the total number of sequencing reads. A way to
-conceptualize this subtle but important point is to assume exponential
-growth in the \emph{number of cells} \(N_t^{(b)}\) of the form
-\begin{equation}\protect\hypertarget{eq-ncells}{}{
-N_{t+1}^{(b)} = N_{t}^{(b)} \mathrm{e}^{\lambda^{(b)}\tau},
-}\label{eq-ncells}\end{equation} for every barcode \(b\) with growth
-rate \(\lambda^{(b)}\). However, when we sequence barcodes, we do not
-directly measure the number of cells, but some number of reads
-\(r_t^{(b)}\) that map to barcode \(b\). In the simplest possible
-scenario, we assume \begin{equation}\protect\hypertarget{eq-r_to_n}{}{
-r_{t}^{(b)} \propto N_{t}^{(b)},
-}\label{eq-r_to_n}\end{equation} where, importantly, the proportionality
-constant depends on the total number of reads for the library for cycle
-\(t\), which might vary from library to library. Therefore, to compare
-the number of reads between libraries at different time points, we must
-normalize the number of reads to the same scale. The simplest form is to
-define a relative abundance, i.e., a frequency with respect to the total
-number of reads, \begin{equation}\protect\hypertarget{eq-r_to_n_norm}{}{
-f_{t}^{(b)} \equiv \frac{r_{t}^{(b)}}{\sum_{b'} r_{t}^{(b')}}.
-}\label{eq-r_to_n_norm}\end{equation} This is the frequency
-Equation~\ref{eq-fitness} describes.
-
-Our ultimate objective is to infer the relative fitness \(s^{(b)}\) for
-each of the \(M\) relevant barcodes in the experiment---hereafter
-referred to as \(s^{(m)}\) to distinguish from the general \(s^{(b)}\)
-and the neutral lineages \(s^{(n)}\) relative fitness. To do so, we
-account for the three primary sources of uncertainty in our model:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Uncertainty in the determination of frequencies. Our model relates
-  frequencies between adjacent growth-dilution cycles to the fitness of
-  the corresponding strain. However, we do not directly measure
-  frequencies. Instead, our data for each barcode consists of a length
-  \(T\) vector of counts \(\underline{r}^{(b)}\) for each of the \(T\)
-  cycles in which the measurements were taken.
-\item
-  Uncertainty in the value of the population mean fitness. We define
-  neutral lineages to have fitness \(s^{(n)} = 0\), helping us anchor
-  the value of the population mean fitness \(\bar{s}_t\) for each pair
-  of adjacent growth cycles. Moreover, we take this parameter as an
-  empirical parameter to be obtained from the data, meaning that we do
-  not impose a functional form that relates \(\bar{s}_t\) to
-  \(\bar{s}_{t+1}\). Thus, we must infer the \(T-1\) values of this
-  population mean fitness with their uncertainty that must be propagated
-  to the value of the mutants' relative fitness.
-\item
-  Uncertainty in each of the mutants' fitness values.
-\end{enumerate}
-
-To account for all these sources of uncertainty in a principled way, in
-the next section, we develop a Bayesian inference pipeline.
-
-\hypertarget{sec-bayesian_inference}{%
-\subsection{Bayesian inference}\label{sec-bayesian_inference}}
-
-As defined in Section~\ref{sec-fitness_model}, our ultimate objective is
-to infer the vector of relative fitness values
-\begin{equation}\protect\hypertarget{eq-fitness_mut_vec}{}{
-\underline{s}^M = (s^{(1)}, s^{(2)}, \ldots, s^{(M)})^\dagger,
-}\label{eq-fitness_mut_vec}\end{equation} where \(^\dagger\) indicates
-the transpose. Our data consists of an \(T \times B\) matrix
-\(\underline{\underline{R}}\), where \(B = M + N\) is the number of
-unique barcodes given by the sum of the number of unique, relevant
-barcodes we care about, \(M\), and the number of unique neutral
-barcodes, \(N\), and \(T\) is the number of growth cycles where
-measurements were taken. The data matrix is then of the form
-\begin{equation}\protect\hypertarget{eq-R_mat}{}{
-\underline{\underline{R}} = \begin{bmatrix}
-- & \underline{r}_1 & - \\
-- & \underline{r}_2 & - \\
- & \vdots & \\
-- & \underline{r}_T & - \\
-\end{bmatrix},
-}\label{eq-R_mat}\end{equation} where each row \(\underline{r}_t\) is a
-\(B\)-dimensional array containing the raw barcode counts at cycle
-\(t\). We can further split each vector \(\underline{r}_t\) into two
-vectors of the form
-\begin{equation}\protect\hypertarget{eq-r-vec_split}{}{
-\underline{r}_t = \begin{bmatrix}
-\underline{r}_t^{N} \\
-\underline{r}_t^{M}
-\end{bmatrix},
-}\label{eq-r-vec_split}\end{equation} i.e., the vector containing the
-neutral lineage barcode counts \(\underline{r}_t^{N}\) and the
-corresponding vector containing the mutant barcode counts
-\(\underline{r}_t^{M}\). Following the same logic, matrix
-\(\underline{\underline{R}}\) can be split into two matrices as
-\begin{equation}\protect\hypertarget{eq-R-mat_split}{}{
-\underline{\underline{R}} = \left[ 
-\underline{\underline{R}}^N \; \underline{\underline{R}}^M
-\right],
-}\label{eq-R-mat_split}\end{equation} where
-\(\underline{\underline{R}}^N\) is a \(T \times N\) matrix with the
-barcode reads time series for each neutral lineage and
-\(\underline{\underline{R}}^M\) is the equivalent \(T \times M\) matrix
-for the non-neutral lineages.
-
-Our objective is to compute the joint probability distribution for all
-relative fitness values given our data. We can express this joint
-posterior distribution using Bayes theorem as
-\begin{equation}\protect\hypertarget{eq-bayes_obj}{}{
-\pi(\underline{s}^M \mid \underline{\underline{R}}) = \frac{
-\pi(\underline{\underline{R}} \mid \underline{s}^M) 
-\pi(\underline{s}^M)}
-{\pi(\underline{\underline{R}})},
-}\label{eq-bayes_obj}\end{equation} where hereafter \(\pi(\cdot)\)
-defines a probability density function. When defining our statistical
-model, we need not to focus on the denominator on the right-hand side of
-Equation~\ref{eq-bayes_obj}. Thus, we can write
-\begin{equation}\protect\hypertarget{eq-bayes_obj_propto}{}{
-\pi(\underline{s}^M \mid \underline{\underline{R}}) \propto
-\pi(\underline{\underline{R}} \mid \underline{s}^M) 
-\pi(\underline{s}^M).
-}\label{eq-bayes_obj_propto}\end{equation} However, when implementing
-the model computationally, the normalization constant on the right-hand
-side of Equation~\ref{eq-bayes_obj} must be computed. This can be done
-from the definition of the model via an integral of the form
-\begin{equation}\protect\hypertarget{eq-bayes_evidence}{}{
-\pi(\underline{\underline{R}}) = \int d^M \underline{s}^M
-\pi(\underline{\underline{R}} \mid \underline{s}^M) 
-\pi(\underline{s}^M),
-}\label{eq-bayes_evidence}\end{equation} also known as a marginalization
-integral. Hereafter, differentials of the form \(d^n\) imply a
-\(n\)-dimensional integral.
-
-Although Equation~\ref{eq-bayes_obj} and
-Equation~\ref{eq-bayes_obj_propto} seem simple enough, recall that
-Equation~\ref{eq-fitness} relates barcode frequency values and the
-population mean fitness to the mutant relative fitness. Therefore, we
-must include these nuisance parameters as part of our inference problem.
-We direct the reader to the supplementary materials for the exact
-definitions of these parameters. Here, it suffices to say that the
-inference problem must include the vector \(\underline{\bar{s}}_T\) of
-all population mean fitness values and the matrix
-\(\underline{\underline{F}}\) of all barcode frequencies within the
-sequencing data. With these nuisance variables in hand, the full
-inference problem we must solve takes the form
-\begin{equation}\protect\hypertarget{eq-bayes_full}{}{
-\pi(
-    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
-    \underline{\underline{R}}
-) \propto
-\pi(
-    \underline{\underline{R}} \mid
-    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
-)
-\pi(
-    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
-).
-}\label{eq-bayes_full}\end{equation} To recover the marginal
-distribution over the non-neutral barcodes relative fitness values, we
-can numerically integrate out all nuisance parameters, i.e.,
-\begin{equation}\protect\hypertarget{eq-bayes_full_marginal}{}{
-\pi(\underline{s}^M \mid \underline{\underline{R}}) =
-\int d^{T-1}\underline{\bar{s}}_T
-\int d^{B}\underline{f}_1 \cdots
-\int d^{B}\underline{f}_T
-\;
-\pi(
-    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
-    \underline{\underline{R}}
-).
-}\label{eq-bayes_full_marginal}\end{equation}
-
-\hypertarget{seq-split_posterior}{%
-\subsubsection{Factorizing the posterior
-distribution}\label{seq-split_posterior}}
-
-The left-hand side of Equation~\ref{eq-bayes_full} is extremely
-difficult to work with. However, we can take advantage of the structure
-of our inference problem to rewrite it in a more manageable form.
-Specifically, the statistical dependencies of our observations and
-latent variables allow us to factorize the joint distribution into the
-product of multiple conditional distributions. To gain some intuition
-about this factorization, let us focus on the inference of the
-population mean fitness values \(\underline{\bar{s}}_T\).
-Equation~\ref{eq-logfreq_neutral} relates the value of the population
-mean fitness to the neutral lineage frequencies and nothing else. This
-suggests that when writing the posterior for these population mean
-fitness parameters, we should be able to condition it only on the
-neutral lineage frequency values, i.e.,
-\(\pi(\underline{\bar{s}}_T \mid \underline{\underline{F}}^N)\). We
-point the reader to Section~\ref{sec-bayes_def} for the full
-mathematical details on this factorization. For our purpose here, it
-suffices to say we can rewrite the joint probability distribution as a
-product of conditional distributions of the form
-\begin{equation}\protect\hypertarget{eq-split_posterior}{}{
-\pi(
-    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
-    \underline{\underline{R}}
-) =
-\pi(
-    \underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}}^M
-)
-\pi(
-    \underline{\bar{s}}_T \mid \underline{\underline{F}}^N
-)
-\pi(\underline{\underline{F}} \mid \underline{\underline{R}}).
-}\label{eq-split_posterior}\end{equation} Written in this form,
-Equation~\ref{eq-split_posterior} captures the three sources of
-uncertainty listed in Section~\ref{sec-fitness_model} in each term.
-Starting from right to left, the first term on the right-hand side of
-Equation~\ref{eq-split_posterior} accounts for the uncertainty when
-inferring the frequency values given the barcode reads. The second term
-accounts for the uncertainty in the values of the mean population
-fitness at different time points. The last term accounts for the
-uncertainty in the parameter we care about---the mutants' relative
-fitnesses. We refer the reader to Section~\ref{sec-bayes_def} for an
-extended description of the model with specific functional forms for
-each term on the left-hand side of Equation~\ref{eq-split_posterior} as
-well as the extension of the model to account for multiple experimental
-replicates or hierarchical genotypes.
-
-\hypertarget{variational-inference}{%
-\subsubsection{Variational Inference}\label{variational-inference}}
-
-One of the technical challenges to the adoption of Bayesian methods is
-the analytical intractability of integrals such as that of
-Equation~\ref{eq-bayes_full_marginal}. Furthermore, even though
-efficient Markov Chain Monte Carlo (MCMC) algorithms such as Hamiltonian
-Montecarlo can numerically perform this integration
-\autocite{betancourt2017}, the dimensionality of the problem in
-Equation~\ref{eq-split_posterior} makes an MCMC-based approach
-prohibitively slow.
-
-To overcome this computational limitation, we rely on the recent
-development of the automatic differentiation variational inference
-algorithm (ADVI) \autocite{kucukelbir2016}. Briefly, when performing
-ADVI, our target posterior distribution
-\(\pi(\theta \mid \underline{\underline{R}})\), where
-\(\theta = (\underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}})\),
-is replaced by an approximate posterior distribution \(q_\phi(\theta)\),
-where \(\phi\) fully parametrizes the approximate distribution. As
-further explained in Section~\ref{sec-vi_primer}, the numerical
-integration problem is replaced by an optimization problem of the form
-\begin{equation}\protect\hypertarget{eq-vi_objective}{}{
-q^*_\phi(\theta) = \min _\phi
-D_{KL}(
-    q_\phi(\theta) \lvert \lvert
-    \pi(\theta \mid \underline{\underline{R}})
-),
-}\label{eq-vi_objective}\end{equation} where \(D_{KL}\) is the
-Kulback-Leibler divergence. In other words, the complicated
-high-dimensional numerical integration problem is transformed into a
-much simpler problem of finding the value of the parameters \(\phi\)
-such that Equation~\ref{eq-vi_objective} is satisfied as best as
-possible within some finite computation time. Although to compute
-Equation~\ref{eq-vi_objective}, we require the posterior distribution we
-are trying to approximate
-\(\pi(\theta \mid \underline{\underline{R}})\), it can be shown that
-maximizing the so-called evidence lower bound (ELBO)
-\autocite{kingma2014}---equivalent to minimizing the variational free
-energy \autocite{gottwald2020}---is mathematically equivalent to
-performing the optimization prescribed by
-Equation~\ref{eq-vi_objective}. We direct the reader to
-Section~\ref{sec-vi_primer} for a short primer on variational inference.
-
-This work is accompanied by the Julia library \texttt{BarBay.jl} that
-makes use of the implementation of both MCMC-based integration as well
-as ADVI optimization to numerically approximate the solution of
-Equation~\ref{eq-bayes_full_marginal} within the Julia ecosystem
-\autocite{ge2018}.
-
-\hypertarget{inference-on-a-single-dataset}{%
-\subsection{Inference on a single
-dataset}\label{inference-on-a-single-dataset}}
-
-To assess the inference pipeline performance, we applied it to a
-simulated dataset with known ground truth relative fitness values (See
-Section~\ref{sec-logistic} for details on simulation).
-Figure~\ref{fig-02}(A) shows the structure of the synthetic dataset. The
-majority of barcodes of interest (faint color lines) are adaptive
-compared to the neutral barcodes (\(s^{(m)} > 0\)). Although the barcode
-frequency trajectories look relatively smooth, our fitness model
-requires the computation of the log frequency ratio between adjacent
-time points as derived in Equation~\ref{eq-logfreq}.
-Figure~\ref{fig-02}(B) shows such data transformation where we can
-better appreciate the observational noise input into our statistical
-model. This noise is evident for the darker lines representing the
-neutral barcodes since all of these lineages are assumed to be
-identically distributed.
-
-To visualize the performance of our inference pipeline in fitting our
-fitness model to the observed data, we compute the so-called posterior
-predictive checks (PPC). In short, the PPC consists of repeatedly
-generating synthetic datasets in agreement with the results from the
-inference results. In other words, we use the resulting parameter values
-from the ADVI inference to generate possible datasets in agreement with
-the inferred values (See Section~\ref{sec-ppc} for further details on
-these computations). Figure~\ref{fig-02}(C) shows these results for all
-neutral lineages (upper left corner plot) and a few representative
-non-neutral barcodes. The different color shades represent the 95\%,
-68\%, and 5\% credible regions, i.e., the regions where we expect to
-find the data with the corresponding probability---or in terms of our
-parameter, the \(X\%\) credible region is the interval where we expect
-the true parameter value to lie with \(X\%\) probability.
-
-The main advantage of our method is this natural interpretability of
-these credible regions where an \(X\%\) credible region indeed captures
-the region of parameter space where we expect with \(X\%\) probability
-the actual value of the parameter lies given our statistical model, our
-prior information, and the observed experimental data. A common mistake
-in the literature is interpreting frequentist confidence intervals as
-Bayesian credible regions when they are not equivalent
-\autocite{morey2016}. Frequentist confidence intervals and Bayesian
-credible regions are based on fundamentally different philosophical
-approaches to statistics. Frequentist confidence intervals represent the
-range of values that would contain the true population parameter with a
-certain probability if the experiment was repeated many times. The
-confidence interval does not represent the probability that the interval
-contains the true value. According to a specific model and prior
-information, Bayesian credible regions represent the range of values
-that contain the parameter with a certain posterior probability. The
-credible region directly represents the probability that the region
-contains the true value. So, frequentist confidence intervals cannot be
-interpreted as Bayesian credible regions because they have fundamentally
-different meanings. Treating an \(X\%\) confidence interval like an
-\(X\%\) credible region is fallacious since confidence intervals do not
-represent probabilistic coverage of the true value like credible
-regions. The intervals are generated through entirely different
-procedures.
-
-To capture the global performance of the model, Figure~\ref{fig-02}(D)
-compares the known ground truth with the inferred relative fitness value
-for all barcodes of interest. There is an excellent degree of
-correspondence between these values, with the error bars representing
-the 68\% credible region for the parameter value crossing the identity
-line for most barcodes. This latter point is made clear with
-Figure~\ref{fig-02}(E) where \(\approx 90\%\) of ground truth fitness
-values fall within one standard deviation of the mean in the inferred
-posterior distributions.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/fig02.pdf}
-
-}
-
-\caption{\label{fig-02}\textbf{Single dataset inference}. (A) Frequency
-trajectories that represent the raw data going into the inference. (B)
-Log frequency ratio between two adjacent time points used by the
-inference pipeline. Darker lines represent the neutral barcodes. These
-transformed data are much more noisy than the seemingly smooth frequency
-trajectories. (C) Examples of the posterior predictive checks for all
-neutral lineages (upper left panel) and a subset of representative
-mutant lineages. Shaded regions represent the 95\%, 68\%, and 5\%
-credible regions for the data. The reported errors above the plot
-represent the 68\% credible region on the mutant relative fitness
-marginal distribution. (D) Comparison between the ground truth fitness
-value from the logistic-growth simulation and the inferred fitness
-value. Gray error bars represent the 68\% posterior credible region for
-the relative fitness values. (E) The empirical cumulative distribution
-function (ECDF) for the absolute z-score value of the ground truth
-parameter value within the inferred fitness posterior distribution.}
-
-\end{figure}
-
-\hypertarget{sec-multienv}{%
-\subsection{Fitness inference on multiple
-environments}\label{sec-multienv}}
-
-The fitness model in Equation~\ref{eq-fitness} relates nuisance
-parameters such as the population mean fitness and the barcode
-frequencies to the relative fitness parameter we want to infer from the
-data. These dependencies imply that uncertainty on the estimates of
-these nuisance parameters influences the inference of the relevant
-parameters. For example, imagine a scenario where the neutral lineages
-data were incredibly noisy, leading to poor estimates of the population
-mean fitness values \(\underline{\bar{s}}_T\). Since the relative
-fitness of any non-neutral barcode \(s^{(m)}\) is determined with
-respect to these neutral barcodes, not accounting for the lack of
-precision in the value of the population mean fitness would result in
-misleading estimates of the accuracy with which we determine the value
-of the parameter we care about. Thus, propagating these sources of
-uncertainty in nuisance parameters is vital to generate an unbiased
-estimate of the relevant information we want to extract from the data.
-One of the benefits of Bayesian methods is the intrinsic error
-propagation embedded in the mathematical framework. For our previous
-example, the uncertainty on the value of the population mean fitness
-values is propagated to the relative fitness of a non-neutral barcode
-since we defined a joint posterior distribution over all parameters as
-fully expressed in Equation~\ref{eq-bayes_full}.
-
-This natural error propagation can help us with the experimental design
-schematized in Figure~\ref{fig-03}(A). Here, rather than performing
-growth-dilution cycles in the same environment, the cells are diluted
-into a different environment. Thus, the uncertainty on the fitness
-estimate for the previous environment must be propagated to that of the
-next one. To validate the extension of our statistical model to this
-scenario, Figure~\ref{fig-03}(B) shows the trajectory of the log
-frequency ratios between adjacent time points. The different colored
-regions correspond to the different environments. For this simulation,
-the growth rate of Environment 2 was set to be, on average, half of the
-average growth rate in Environment 1. Equivalently, the growth rate in
-Environment 3 was set to be, on average, twice the average growth rate
-in Environment 1. Figure~\ref{fig-03}(C-E) show the correspondence
-between the simulation ground truth and the inferred fitness values,
-where the error bars represent the 68\% credible region.
-Figure~\ref{fig-03}(F) summarizes the performance of our inference
-pipeline by showing the empirical cumulative distribution functions for
-the absolute value of the ground truth fitness value z-score within the
-posterior distribution. This plot shows that, overall, \(\approx 75\%\)
-of inferred mean values fall within one standard deviation of the ground
-truth. For completeness, Figure~\ref{fig-03}(G) shows the posterior
-predictive checks for a few example barcodes.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/fig03.pdf}
-
-}
-
-\caption{\label{fig-03}\textbf{Multi-environment fitness inference.} (A)
-Schematic of the simulated experimental design where growth-dilution
-cycles are performed into different environments for each cycle. (B) log
-frequency rations between adjacent time points. Darker lines represent
-the neutral barcodes. The colors in the background demark the
-corresponding environment, matching colors in (A). Environment 2 is set
-to have, on average, half the growth rate of environment 1. Likewise,
-environment 3 is set to have, on average, twice the growth rate of
-environment 1. (C-E) Comparison between the ground truth fitness value
-from the logistic-growth simulation and the inferred fitness value for
-each environment. Gray error bars represent the 68\% posterior credible
-region. (F) The empirical cumulative distribution function (ECDF) for
-the absolute z-score value of the ground truth parameter value within
-the inferred fitness posterior distribution for all fitness values
-(black line) and each environment individually (color lines). (G)
-Examples of the posterior predictive checks for all neutral lineages
-(upper left panel) and a subset of representative mutant lineages.
-Shaded regions surrounding the data represent the 95\%, 68\%, and 5\%
-credible regions for the data. The reported errors above the plot
-represent the 68\% credible region on the mutant relative fitness
-marginal distribution. Background colors match those of (A).}
-
-\end{figure}
-
-\hypertarget{sec-replicates}{%
-\subsection{Accounting for experimental replicates via hierarchical
-models}\label{sec-replicates}}
-
-Our inference pipeline can be extended to account for multiple
-experimental replicates via Bayesian hierarchical models
-\autocite{betancourt2013}. Briefly, when accounting for multiple
-repeated measurements of the same phenomena, there are two extreme cases
-one can use to perform the data analysis: On the one hand, we can treat
-each measurement as entirely independent, losing the power to utilize
-multiple measurements when trying to learn a single parameter. This can
-negatively impact the inference since, in principle, the value of our
-parameter of interest should not depend on the particular experimental
-replicate in question. However, this approach does not allow us to
-properly ``combine'' the uncertainties in both experiments when
-performing the inference. On the other hand, we can pool all data
-together and treat our different experiments as a single measurement
-with higher coverage. This loses the subtle differences due to biotic
-and abiotic batch effects, effectively halving the data that goes into
-our inference problem.
-
-Hierarchical models present a middle ground between these extremes.
-First, hierarchical models rely on the definition of so-called
-\emph{hyper-parameters}, that capture the parametric inference we are
-interested in---for this inference problem, we have a hyper-fitness
-value \(\theta^{(m)}\) for each non-neutral barcode. Second, each
-experiment draws randomly from the distribution of this hyper-parameter,
-allowing for subtle variability between experiments to be accounted
-for---in the present inference pipeline, each experimental replicate
-gets assigned a \emph{local} fitness value \(s^{(m)}_i\), where the
-extra sub-index indicates the \(i\)-th experimental replicate.
-Conceptually, we can think of the local fitness for replicate \(i\) as
-being sampled from a distribution that depends on the value of the
-global hyper-fitness value, i.e., \(s^{(m)}_i \sim \pi_{\theta^{(m)}}\),
-where the subindex \(\theta^{(m)}\) indicates the distribution's
-parametric dependence on the hyper-fitness value. This way of
-interpreting the connection between the distribution
-\(\pi_{\theta^{(m)}}\) and the local fitness implies that a large
-replicate-to-replicate variability would lead to a broad hyper-fitness
-distribution---implying a large uncertainty when determining the
-parameter that characterizes the overall relative fitness. We point the
-reader to Section~\ref{sec-hierarchical_model} for the full definition
-of the hierarchical model used in this section. Importantly, as
-schematized in Figure~\ref{fig-04}(A), the influence between different
-experimental replicates runs both ways. First, the data from one
-experimental replicate (\(\underline{\underline{R}}^M_k\) in the
-diagram) informs all local fitness values via the global hyper-fitness
-(upper panel in Figure~\ref{fig-04}(A)). Second, the local fitness value
-is informed by the data from all experimental replicates via the same
-global hyper-fitness parameter (lower panel in Figure~\ref{fig-04}(A)).
-
-To test the performance of this model, we simulated two experimental
-replicates with 1000 unique barcodes (see Figure~\ref{fig-04}(B-C))
-where we randomly sampled a ground truth hyper-fitness value
-\(\theta^{(m)}\) for each barcode. We sampled a variation from this
-hyper-fitness value for each experimental replicate \(s^{(m)}_i\) to
-capture experimental batch effects. Figure~\ref{fig-04}(D) shows the
-relationship between hyper-fitness and replicate fitness values for this
-simulation. The spread around the identity line represents the expected
-batch-to-batch variation. The posterior predictive checks examples in
-Figure~\ref{fig-04}(E) show that the hierarchical model can correctly
-fit the data for each experimental replicate. Furthermore,
-Figure~\ref{fig-04}(F-G) show a high correlation between the ground
-truth and the inferred fitness values. The empirical cumulative
-distribution functions shown in Figure~\ref{fig-04}(H-I) reveal that for
-\(\approx 75\%\) of the non-neutral barcodes, the ground truth
-hyper-fitness values fall within one standard deviation from the mean
-value in the posterior distributions.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/fig04.pdf}
-
-}
-
-\caption{\label{fig-04}\textbf{Hierarchical model on experimental
-replicates}. (A) Schematic depiction of the interactions between local
-fitness values \(\underline{s}_k^M\) through the global hyper-fitness
-value \(\underline{\theta}^M\) for \(K\) hypothetical experimental
-replicates. The upper diagram shows how the data from replicate \(k\)
-informs all local fitness values via the hyper-fitness parameter. The
-lower panel shows the reverse, where all other datasets inform the local
-fitness value. (B-C) Simulated replicate datasets with 900 barcodes of
-interest and 100 neutral lineages. (D) Comparison between the simulation
-ground truth hyper-fitness and each replicate ground truth fitness. The
-scatter between parameters captures experimental batch effects. (E)
-Examples of the posterior predictive checks for all neutral lineages
-(upper left panels) and a subset of representative mutant lineages.
-Shaded regions from light to dark represent the 95\%, 68\%, and 5\%
-credible regions. (F-G) Comparison between the simulation's ground truth
-hyper-fitness (F) and replicate fitness (G) values and the inferred
-parameters. Gray error bars represent the 68\% posterior credible
-region. (H-I) The empirical cumulative distribution function (ECDF) for
-the absolute z-score value of the ground truth parameter value within
-the inferred hyper-fitness posterior distribution (H) and replicate
-fitness (I).}
-
-\end{figure}
-
-As shown in Figure~\ref{fig-05}, the structure imposed by the
-hierarchical model schematized in Figure~\ref{fig-04}(A), where we
-explicitly account for the connection between experimental replicates
-can improve the quality of the inference. Inferred fitness values
-between experimental replicates exhibit a stronger degree of correlation
-in the hierarchical model (Figure~\ref{fig-05}(A)) compared to
-conducting inference on each replicate independently
-(Figure~\ref{fig-05}(B)). Moreover, when comparing the inferred
-hyper-fitness values---the objective parameter when performing multiple
-experimental measurements---the hierarchical model outperforms averaging
-the independent experimental replicates as shown in
-Figure~\ref{fig-05}(C) and (D).
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/fig05.pdf}
-
-}
-
-\caption{\label{fig-05}\textbf{Comparison between hierarchical model and
-single dataset model}. (A-B) comparison of inferred fitness values
-between experimental replicates when fitting a hierarchical model (A) or
-independently fitting each dataset (B). Gray error bars represent the
-68\% credible regions. (C) Comparison between the ground truth
-hyper-fitness value and the inferred parameters. The blue dots show the
-inferred hyper-fitness values when assuming a hierarchical model. Gray
-error bars show the 68\% credible region for this inference. The yellow
-dots show the average of the mean inferred fitness values for the two
-experimental replicates. No error bars are shown for these, as it is
-inappropriate to compute one with two data points per non-neutral
-barcode. (D) Empirical cumulative distribution function (ECDF) of the
-absolute difference between the inferred mean and the ground truth
-hyper-fitness.}
-
-\end{figure}
-
-\hypertarget{sec-genotypes}{%
-\subsection{Accounting for multiple barcodes per genotype via
-hierarchical models}\label{sec-genotypes}}
-
-Hierarchical models can also capture another experimental design in
-which multiple barcodes map to the same or an equivalent genotype. As we
-will show, this many-to-one mapping can improve the inference compared
-to the extreme cases of inferring the fitness of each barcode
-independently or pooling the data of all barcodes mapping to a single
-genotype. As schematized in Figure~\ref{fig-06}(A), a small modification
-of the base model allows us to map the structure of our original model
-to that of a hierarchical model with a fitness hyperparameter vector
-\(\underline{\theta}^G\), where \(G\) is the number of genotypes in the
-dataset.
-
-Figure~\ref{fig-06}(B) shows a single experimental replicate in which 90
-genotypes were assigned a random number of barcodes (a multinomial
-distribution with a mean of ten barcodes per genotype) for a total of
-900 non-neutral barcodes. To assess the performance of the hierarchical
-model proposed in Figure~\ref{fig-06}(A), we performed inference using
-this hierarchical model, as well as the two extreme cases of ignoring
-the connection between the barcodes belonging to the same
-genotype---equivalent to performing inference using the model presented
-in Figure~\ref{fig-02}(A) over the barcodes---or pooling the data of all
-barcodes belonging to the same genotype into a single count---equivalent
-to performing inference using the model presented in
-Figure~\ref{fig-02}(A) over the pooled barcodes.
-Figure~\ref{fig-06}(C-D) shows the comparison between the simulation
-ground truth and the inferred values for these three cases. Not only do
-the hierarchical model results show higher degrees of correlation with
-the ground truth, but the error bars (representing the 68\% credible
-regions) are smaller, meaning that the uncertainty in the estimate of
-the parameter we care about decreases when using the hierarchical model.
-The improvement in the prediction can be seen in Figure~\ref{fig-06}(F)
-where the empirical cumulative distribution function of the absolute
-difference between the mean inferred value and the simulation ground
-truth is shown for all three inference models. The hierarchical model's
-curve ascends more rapidly, showing that, in general, the inferred
-values are closer to the ground truth. For completeness,
-Figure~\ref{fig-06}(G) shows some examples of how the hierarchical model
-can capture the raw log-frequency count observations.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/fig06.pdf}
-
-}
-
-\caption{\label{fig-06}\textbf{Hierarchical model for multiple barcodes
-per genotype.} (A) Schematic depiction of the hierarchical structure for
-multiple barcodes mapping to a single genotype. A set of barcodes
-mapping to an equivalent genotype map to ``local'' fitness values
-\(s^{(b)}\) that are connected via a hyper-fitness parameter for the
-genotype \(\theta^{(g)}\). (B) Simulated dataset with 100 neutral
-lineages and 900 barcodes of interest distributed among 90 genotypes.
-(C-E) Comparison between the inferred and ground truth fitness values
-for a hierarchical model (C), a model where each barcode is inferred
-independently (D), and a model where barcodes mapping to the same
-genotype are pooled together (E). Gray error bars represent the 68\%
-credible regions. (F) Empirical cumulative distribution function (ECDF)
-of the absolute difference between the inferred mean and the ground
-truth fitness values for all three models. (G) Examples of the posterior
-predictive checks for all neutral lineages (upper left panels) and a
-subset of representative mutant lineages. Shaded regions from light to
-dark represent the 95\%, 68\%, and 5\% credible regions.}
-
-\end{figure}
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-Experimental evolution of microbial systems has dramatically advanced
-our understanding of the basic principles of biological evolution
-\autocite{kussell2013}. From questions related to the optimal
-fine-tuning of gene expression programs \autocite{Dekel2005}, to the
-dimensionality, geometry, and accessibility of the adaptive fitness
-landscape explored by these rapidly adapting populations
-\autocite{kinsler2020,maeda2020}, to the emergence of eco-evolutionary
-dynamics in a long-term evolution experiment \autocite{good2017}; for
-all of these and other cases, the microbial experimental platform
-combined with high-throughput sequencing has been essential to tackling
-these questions with empirical data. This exciting research area
-promises to improve as new culturing technologies \autocite{jagdish2022}
-as well as more complex lineage barcoding schemes
-\autocite{nguyenba2019a,yang2022}, are adopted.
-
-For this data-heavy field, properly accounting for the uncertainty in
-parameters inferred from experiments is vital to ensure the conclusions
-drawn are reliable. Bayesian statistics presents a principled way to
-quantify this uncertainty systematically \autocite{gelman2010}.
-Moreover, Bayesian analysis offers a more natural way to interpret the
-role that probability theory plays when performing data analysis
-compared to the often-misinterpreted frequentist methods
-\autocite{vanderplas2014}. Nevertheless, the technical challenges
-associated with Bayesian analysis has limited its application. This is
-set to change as recognition of the misuse of frequentist concepts such
-as the p-value is receiving more attention \autocite{nuzzo2014}.
-Moreover, advances in numerical methods such as Hamiltonian Monte Carlo
-\autocite{betancourt2017} and variational inference
-\autocite{kucukelbir2016} allows for complex Bayesian models to be fit
-to empirical data.
-
-In this paper, we present a computational pipeline to analyze
-lineage-tracking time-series data for massive-parallel competition
-assays. More specifically, we fit a Bayesian model to infer the fitness
-of multiple genotypes relative to a reference
-\autocite{kinsler2020,ascensao2023}. The proposed model accounts for
-multiple sources of uncertainty with proper error propagation intrinsic
-to Bayesian methods. To scale the inference pipeline to large datasets
-with \(> 10,000\) barcodes, we use the ADVI algorithm
-\autocite{kucukelbir2016} to fit a variational posterior distribution.
-The main difference between our method and previous inference pipelines,
-such as \textcite{li2023}, is that the present analysis provides
-interpretable errors on the inferred fitness values. The reported
-uncertainty intervals---known as credible regions---can be formally
-interpreted as capturing the corresponding probability mass of finding
-the true value of the parameter given the model, the prior information,
-and the data. Furthermore, minor modifications to the structure of the
-statistical model presented in this work allow for the analysis of
-different experimental designs, such as growth-dilution cycles in
-different environments, joint analysis of multiple experimental
-replicates of the same experiment via hierarchical models, and a
-hierarchical model for multiple barcodes mapping to equivalent
-genotypes. We validate our analysis pipeline on simulated datasets with
-known ground truth, showing that the model fits the data adequately,
-capturing the ground truth parameters within the posterior distribution.
-
-It is important to highlight some of the consequences of the general
-experimental design and the implicit assumptions within the proposed
-statistical model to analyze the resulting data. First, the composition
-of the population is such that the initial fraction of the population
-occupied by the barcoded genotypes is small---usually \textgreater90\%
-of the initial population is the non-labeled reference strain. This
-constraint is important as the fitness model used to fit the time series
-data assumes that the tracked frequencies are \(\ll 1\). Second, when
-computing log frequency ratios, we can run into the issue of dividing by
-zero. This is a common problem when dealing with molecular count data
-\autocite{lovell2020}. Our model gets around this issue by assuming that
-the frequency of any barcode cannot be, but still can get arbitrarily
-close to, zero. Therefore, we implicitly assume that no lineage goes
-extinct during the experiment. Moreover, the statistical model directly
-accounts for the uncertainty associated with having zero barcode counts,
-increasing the corresponding uncertainty. Third, the models presented in
-this paper require the existence of a labeled sub-population of barcoded
-reference strains. These barcodes help determine the fitness baseline,
-as every fitness is quantified with respect to this reference genotype.
-This experimental design constraint facilitates the inference of the
-population mean fitness since most of the culture---the unlabeled
-reference genotype---is not tracked. Finally, the presented statistical
-model assumes that relative fitness is solely a constant of the
-environment and the genotype. Future directions of this work could
-extend the fitness model to properly analyze data with time-varying or
-frequency-dependent fitness values.
-
-In total, the statistical model presented in this work and the software
-package accompanying the paper allow for a principled way of quantifying
-the accuracy with which we can extract relevant parametric information
-from large-scale multiplexed fitness competition assays. Furthermore,
-the implementation of Bayesian models and their fitting via automatic
-differentiation approaches opens the gate to extend this type of formal
-analysis to the data-rich literature in experimental evolution and other
-high-throughput technologies applications.
-
-\hypertarget{supplementary-materials}{%
-\section{Supplementary Materials}\label{supplementary-materials}}
-
-\hypertarget{sec-vi_primer}{%
-\subsection{Primer on Variational Inference}\label{sec-vi_primer}}
-
-In this section, we will briefly introduce the idea behind variational
-inference. Recall that any Bayesian inference problem deals with the
-joint distribution between observations \(\underline{x}\) and unobserved
-latent variables \(\underline{\theta}\). This joint distribution can be
-written as the product of a distribution of the observations
-\(\underline{x}\) conditioned on the \(\underline{\theta}\) and the
-marginal distribution of these latent variables, i.e.,
-\begin{equation}\protect\hypertarget{eq-vi_joint_dist}{}{
-\pi(\underline{x}, \underline{\theta}) =
-\pi(\underline{x} \mid \underline{\theta}) \pi(\underline{\theta}).
-}\label{eq-vi_joint_dist}\end{equation} A Bayesian inference pipeline's
-objective is to compute the latent variables' posterior probability
-given a set of observations. This computation is equivalent to updating
-our prior beliefs about the set of values that the latent variables take
-after taking in new data. We write this as Bayes theorem
-\begin{equation}\protect\hypertarget{eq-vi_bayes_theorem}{}{
-\pi(\underline{\theta} \mid \underline{x}) = 
-\frac{
-        \pi(\underline{x} \mid \underline{\theta})\pi(\underline{\theta})
-    }{
-        \pi(\underline{x})
-    }.
-}\label{eq-vi_bayes_theorem}\end{equation} The main technical challenge
-for working with Equation~\ref{eq-vi_bayes_theorem} comes from the
-computation of the denominator, also known as the \emph{evidence} or the
-\emph{marginalized likelihood}. The reason computing this term is
-challenging is because it involves a (potentially) high-dimensional
-integral of the form
-\begin{equation}\protect\hypertarget{eq-vi_evidence_int}{}{
-\pi(\underline{x}) = 
-\int\cdots\int d^K\underline{\theta}\; \pi(\underline{x}, \underline{\theta}) = 
-\int\cdots\int d^K\underline{\theta}\; 
-\pi(\underline{x} \mid \underline{\theta})
-\pi(\underline{\theta}),
-}\label{eq-vi_evidence_int}\end{equation} where \(K\) is the
-dimesionality of the \(\underline{\theta}\) vector. Here, the integrals
-are taken over the support---the set of values valid for the
-distribution---of \(\pi(\underline{\theta})\). However, only a few
-selected distributions have a closed analytical form; thus, in most
-cases Equation~\ref{eq-vi_evidence_int} must be solved numerically.
-
-Integration in high-dimensional spaces can be computationally extremely
-challenging. For a naive numerical quadrature procedure, integrating
-over a grid of values for each dimension of \(\underline{\theta}\) comes
-with an exponential explosion of the number of required grid point
-evaluations, most of which do not contribute significantly to the
-integration. To gain visual intuition about this challenge, imagine
-integrating the function depicted in Figure~\ref{fig-SI01}. If the
-location of the high-density region (dark peak) is unknown, numerical
-quadrature requires many grid points to ensure we capture this peak.
-However, most of the numerical evaluations of the function on the grid
-points do not contribute significantly to the integral. Therefore, our
-computational resources are wasted on insignificant evaluations. This
-only gets worse as the number of dimensions increases since the number
-of grid point evaluation scales exponentially.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/figSIX_gaussian_peak.png}
-
-}
-
-\caption{\label{fig-SI01}\textbf{High-dimensional numerical quadrature
-does not scale with dimensionality.} Schematic depiction of the problem
-with naive numerical quadrature to integrate over an unknown density.
-While the density is concentrated on the dark peak, most of the
-evaluations over the \(x_1-x_2\) grid do not contribute to the value of
-the integral}
-
-\end{figure}
-
-Modern Markov Chain Monte Carlo algorithms, such as Hamiltonian Monte
-Carlo, can efficiently perform this high-dimensional integration by
-utilizing gradient information from the target density
-\textcite{betancourt2017}. Nevertheless, these sampling-based methods
-become prohibitively slow for the number of dimensions our present
-inference problem presents. Thus, there is a need to find scalable
-methods for the inference problem in Equation~\ref{eq-vi_bayes_theorem}.
-
-Variational inference circumvents these technical challenges by
-proposing an approximate solution to the problem. Instead of working
-with the posterior distribution in its full glory
-\(\pi(\underline{\theta} \mid \underline{x})\), let us propose an
-approximate posterior distribution \(q_\phi\) that belongs to a
-distribution family fully parametrized by \(\phi\). For example, let us
-say that the distribution \(q_\phi\) belongs to the family of
-multivariate Normal distributions such that
-\(\phi = (\underline{\mu}, \underline{\underline{\Sigma}})\), where
-\(\underline{\mu}\) is the vector of means and
-\(\underline{\underline{\Sigma}}\) is the covariance matrix. If we
-replace \(\pi\) by \(q_\phi\), we want \(q_\phi\) to resemble the
-original posterior as much as possible. Mathematically, this can be
-expressed as minimizing a ``\emph{distance metric}''---the
-Kullback-Leibler (KL) divergence, for example---between the
-distributions. Note that we use quotation marks because, formally, the
-KL divergence is not a distance metric since it is not symmetric.
-Nevertheless, the variational objective is set to find a distribution
-\(q_\phi^*\) such that
-\begin{equation}\protect\hypertarget{eq-vi_objective}{}{
-q_\phi^*(\underline{\theta}) =
-\min_\phi D_{KL}\left(
-    q_\phi(\underline{\theta}) \vert\vert 
-    \pi(\underline{\theta} \mid \underline{x})
-\right),
-}\label{eq-vi_objective}\end{equation} where \(D_{KL}\) is the KL
-divergence. Furthermore, we highlight that the KL divergence is a
-strictly positive number, i.e.,
-\begin{equation}\protect\hypertarget{eq-vi_kl_pos}{}{
-D_{KL}\left(
-    q_\phi(\underline{\theta}) \vert\vert 
-    \pi(\underline{\theta} \mid \underline{x})
-\right) \geq 0,
-}\label{eq-vi_kl_pos}\end{equation} as this property will become
-important later on.
-
-At first sight, Equation~\ref{eq-vi_objective} does not improve the
-situation but only introduces further technical complications. After
-all, the definition of the KL divergence
-\begin{equation}\protect\hypertarget{eq-vi_kl_div}{}{
-D_{KL}\left(
-    q_\phi(\underline{\theta}) \vert\vert 
-    \pi(\underline{\theta} \mid \underline{x})
-\right) \equiv 
-\int \cdots \int d^K\underline{\theta}\;
-q_\phi(\underline{\theta})
-\ln \frac{
-    q_\phi(\underline{\theta})
-}{
-    \pi(\underline{\theta} \mid \underline{x})
-},
-}\label{eq-vi_kl_div}\end{equation} includes the posterior distribution
-\(\pi(\underline{\theta} \mid \underline{x})\) we are trying to get
-around. However, let us manipulate Equation~\ref{eq-vi_kl_div} to beat
-it to a more reasonable form. First, we can use the properties of the
-logarithms to write
-\begin{equation}\protect\hypertarget{eq-vi_step01}{}{
-D_{KL}\left(
-    q_\phi(\underline{\theta}) \vert\vert 
-    \pi(\underline{\theta} \mid \underline{x})
-\right) = 
-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln q_\phi(\underline{\theta}) -
-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln \pi(\underline{\theta} \mid \underline{x}),
-}\label{eq-vi_step01}\end{equation} where, for convenience, we write a
-single integration sign (\(d^K\underline{\theta}\;\) still represents a
-multi-dimensional differential). For the second term in
-Equation~\ref{eq-vi_step01}, we can substitute the term inside the
-logarithm using Equation~\ref{eq-vi_bayes_theorem}. This results in
-\begin{equation}\protect\hypertarget{eq-vi_step02}{}{
-\begin{aligned}
-D_{KL}\left(
-    q_\phi(\underline{\theta}) \vert\vert 
-    \pi(\underline{\theta} \mid \underline{x})
-\right) &= 
-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln q_\phi(\underline{\theta}) \\
-&- \int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln \left( 
-    \frac{
-        \pi(\underline{x} \mid \underline{\theta})\pi(\underline{\theta})
-    }{
-        \pi(\underline{x})
-    }
-\right).
-\end{aligned}
-}\label{eq-vi_step02}\end{equation} Again, using the properties of
-logarithms, we can split Equation~\ref{eq-vi_step02}, obtaining
-\begin{equation}\protect\hypertarget{eq-vi_step03}{}{
-\begin{aligned}
-D_{KL}\left(
-    q_\phi(\underline{\theta}) \vert\vert 
-    \pi(\underline{\theta} \mid \underline{x})
-\right) &= 
-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln q_\phi(\underline{\theta}) \\
-&-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln \pi(\underline{x} \mid \underline{\theta}) \\
-&-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln \pi(\underline{\theta}) \\
-&+\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln \pi(\underline{x}).
-\end{aligned}
-}\label{eq-vi_step03}\end{equation} It is convenient to write
-Equation~\ref{eq-vi_step03} as
-\begin{equation}\protect\hypertarget{eq-vi_step04}{}{
-\begin{aligned}
-D_{KL}\left(
-    q_\phi(\underline{\theta}) \vert\vert 
-    \pi(\underline{\theta} \mid \underline{x})
-\right) &= 
-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln \frac{
-    q_\phi(\underline{\theta})
-    }{
-        \pi(\underline{\theta})
-    } \\
-&-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln \pi(\underline{x} \mid \underline{\theta}) \\
-&+ \ln \pi(\underline{x}) 
-\int d^K\underline{\theta}\; q_\phi(\underline{\theta}),
-\end{aligned}
-}\label{eq-vi_step04}\end{equation} where for the last term, we can take
-\(\ln \pi(\underline{x})\) out of the integral since it does not depend
-on \(\underline{\theta}\). Lastly, we utilize two properties:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  The proposed approximate distribution must be normalized, i.e.,
-  \begin{equation}\protect\hypertarget{eq-vi_q_norm}{}{
-  \int d^K\underline{\theta}\; q_\phi(\underline{\theta}) = 1.
-  }\label{eq-vi_q_norm}\end{equation}
-\item
-  The law of the unconscious statistician (LOTUS) establishes that for
-  any probability density function, it must be true that
-  \begin{equation}\protect\hypertarget{eq-vi_lotus}{}{
-  \int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-  f(\underline{\theta}) = \left\langle 
-   f(\underline{\theta}) 
-  \right\rangle_{q_\phi},
-  }\label{eq-vi_lotus}\end{equation} where
-  \(\left\langle\cdot\right\rangle_{q_\phi}\) is the expected value over
-  the \(q_\phi\) distribution.
-\end{enumerate}
-
-Using these two properties, the positivity constraint on the KL
-divergence in Equation~\ref{eq-vi_kl_pos}, and the definition of the KL
-divergence in Equation~\ref{eq-vi_kl_div} we can rewrite
-Equation~\ref{eq-vi_step04} as
-\begin{equation}\protect\hypertarget{eq-vi_step05}{}{
-D_{KL}\left( 
-    q_\phi(\underline{\theta}) \vert \vert
-    \pi(\underline{\theta}) 
-\right) -
-\left\langle
-    \ln \pi(\underline{x} \mid \underline{\theta})
-\right\rangle_{q_\phi}
-\geq - \ln \pi(\underline{x}).
-}\label{eq-vi_step05}\end{equation} Multiplying by a minus one, we have
-the functional form of the so-called evidence lower bound (ELBO)
-\textcite{kingma2014},
-\begin{equation}\protect\hypertarget{eq-vi_elbo}{}{
-\underbrace{
-    \ln \pi(\underline{x})
-}_{\text{log evidence}} \geq
-\underbrace{
-    \left\langle
-        \ln \pi(\underline{x} \mid \underline{\theta})
-    \right\rangle_{q_\phi} -
-    D_{KL}\left( 
-        q_\phi(\underline{\theta}) \vert \vert
-        \pi(\underline{\theta}) 
-    \right)
-}_{\text{ELBO}}.
-}\label{eq-vi_elbo}\end{equation}
-
-Let us recapitulate where we are. We started by presenting the challenge
-of working with Bayes' theorem, as it requires a high-dimensional
-integral of the form in Equation~\ref{eq-vi_evidence_int}. As an
-alternative, variational inference posits to approximate the posterior
-distribution \(\pi(\underline{\theta} \mid \underline{x})\) with a
-parametric distribution \(q_\phi(\underline{\theta})\). By minimizing
-the KL divergence between these distributions, we arrive at the result
-in Equation~\ref{eq-vi_elbo}, where the left-hand side---the log
-marginalized likelihood or log evidence---we cannot compute for
-technical/computational reasons. However, the right-hand side is
-composed of things we can easily evaluate. We can easily evaluate the
-log-likelihood \(\ln \pi(\underline{x} \mid \underline{\theta})\) and
-the KL divergence between our proposed approximate distribution
-\(q_\phi(\underline{\theta})\) and the prior distribution
-\(\pi(\underline{\theta})\). Moreover, we can compute the gradients of
-these functions with respect to the parameters of our proposed
-distribution. This last point implies that we can change the parameters
-of the proposed distribution to maximize the ELBO. And, although we
-cannot compute the left-hand side of Equation~\ref{eq-vi_elbo}, we know
-that however large we make the ELBO, it will always be smaller than (or
-equal) the log-marginal likelihood. Therefore, the larger we can make
-the ELBO by modifying the parameters \(\phi\), the closer it gets to the
-log-marginal likelihood, and, as a consequence, the better our proposed
-distribution \(q_\phi(\underline{\theta})\) gets to the true posterior
-distribution \(\pi(\underline{\theta} \mid \underline{x})\).
-
-In this sense, variational inference turns the intractable numerical
-integration problem to an optimization routine, for which there are
-several algorithms available.
-
-\hypertarget{advi-algorithm}{%
-\subsubsection{ADVI algorithm}\label{advi-algorithm}}
-
-To maximize the right-hand side of Equation~\ref{eq-vi_elbo}, the
-Automatic Differentiation Variational Inference (ADVI) algorithm
-developed in \autocite{kucukelbir2016} takes advantage of advances in
-probabilistic programming languages to generate a robust method to
-perform this optimization. Without going into the details of the
-algorithm implementation, for our purposes, it suffices to say that we
-define our joint distribution \(\pi(\underline{\theta}, \underline{x})\)
-as the product defined in Equation~\ref{eq-vi_joint_dist}. ADVI then
-proposes an approximate variational distribution \(q_\phi\) that can
-either be a multivariate Normal distribution with a diagonal covariance
-matrix, i.e., \begin{equation}\protect\hypertarget{eq-vi_meanfield}{}{
-\phi = (\underline{\mu}, \underline{\underline{D}}),
-}\label{eq-vi_meanfield}\end{equation} where
-\(\underline{\underline{D}}\) is the identity matrix, with the diagonal
-elements given by the vector of variances \(\underline{\sigma}^2\) for
-each variable or a full-rank multivariate Normal distribution
-\begin{equation}\protect\hypertarget{eq-vi_full}{}{
-\phi = (\underline{\mu}, \underline{\underline{\Sigma}}).
-}\label{eq-vi_full}\end{equation}
-
-Then, the parameters are initialized in some value \(\phi_o\). These
-parameters are iteratively updated by computing the gradient of the ELBO
-(right-hand side of Equation~\ref{eq-vi_elbo}), hereafter defined as
-\(\mathcal{L}\), with respect to the parameters,
-\begin{equation}\protect\hypertarget{eq-vi_elbo_gradient}{}{
-\nabla_\phi \mathcal{L} = \nabla_{\underline{\mu}} \mathcal{L} + 
-\nabla_{\underline{\sigma}}\mathcal{L},
-}\label{eq-vi_elbo_gradient}\end{equation} and then computing \[
-\phi_{t+1} = \phi_{t} + \eta \nabla_\phi \mathcal{L},
-\] where \(\eta\) defines the step size.
-
-This short explanation behind the ADVI algorithm is intended only to
-gain intuition on how the optimal variational distribution \(q_\phi\) be
-computed. There are many nuances in the implementation of the ADVI
-algorithm. We invite the reader to look at the original reference for
-further details.
-
-\hypertarget{sec-bayes_def}{%
-\subsection{Defining the Bayesian model}\label{sec-bayes_def}}
-
-In the main text, we specify the inference problem we must solve as
-being of the form
-\begin{equation}\protect\hypertarget{eq-full_inference_SI}{}{
-\pi(
-    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
-    \underline{\underline{R}}
-) \propto
-\pi(
-    \underline{\underline{R}} \mid
-    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
-)
-\pi(
-    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
-).
-}\label{eq-full_inference_SI}\end{equation} Here, we briefly define the
-missing nuisance parameters. Let
-\begin{equation}\protect\hypertarget{eq-pop_fitness_vec}{}{
-\underline{\bar{s}}_T = (\bar{s}_1, \bar{s}_2, \ldots, \bar{s}_{T-1})^\dagger,
-}\label{eq-pop_fitness_vec}\end{equation} be the vector containing the
-\(T-1\) population mean fitness we compute from the \(T\) time points
-where measurements were taken. We have \(T-1\) since the value of any
-\(\bar{s}_t\) requires cycle numbers \(t\) and \(t+1\). Furthermore, let
-the matrix \(\underline{\underline{F}}\) be a \(T \times B\) matrix
-containing all frequency values. As with Equation~\ref{eq-R-mat_split}
-in the main text, we can split \(\underline{\underline{F}}\) into two
-matrices of the form
-\begin{equation}\protect\hypertarget{eq-F-mat_split}{}{
-\underline{\underline{F}} = \left[ 
-\underline{\underline{F}}^N \; \underline{\underline{F}}^M
-\right],
-}\label{eq-F-mat_split}\end{equation} to separate the corresponding
-neutral and non-neutral barcode frequencies.
-
-Let us now define each of the terms in Equation~\ref{eq-split_posterior}
-described in Section~\ref{sec-bayesian_inference} of the main text. The
-following sections will specify the functional form each of these terms
-takes.
-
-\hypertarget{sec-bayes_freq}{%
-\subsubsection{\texorpdfstring{Frequency uncertainty
-\(\pi(\underline{\underline{F}} \mid \underline{\underline{R}})\)}{Frequency uncertainty \textbackslash pi(\textbackslash underline\{\textbackslash underline\{F\}\} \textbackslash mid \textbackslash underline\{\textbackslash underline\{R\}\})}}\label{sec-bayes_freq}}
-
-We begin with the probability of the frequency values given the raw
-barcode reads. The first assumption is that the inference of the
-frequency values for time \(t\) is independent of any other time.
-Therefore, we can write the joint probability distribution as a product
-of independent distributions of the form
-\begin{equation}\protect\hypertarget{eq-freq_indep}{}{
-\pi(\underline{\underline{F}} \mid \underline{\underline{R}}) =
-\prod_{t=1}^T \pi(\underline{f}_t \mid \underline{r}_t),
-}\label{eq-freq_indep}\end{equation} where \(\underline{f}_t\) and
-\(\underline{r}_t\) are the \(t\)-th row of the matrix containing all of
-the measurements for time \(t\). We imagine that when the barcode reads
-are obtained via sequencing, the quantified number of reads is a Poisson
-sample from the ``true'' underlying number of barcodes within the pool.
-This translates to assuming that the number of reads for each barcode at
-any time point \(r^{(b)}_t\) is an independent Poisson random variable,
-i.e., \begin{equation}\protect\hypertarget{eq-freq_poisson_reads}{}{
-r^{(b)}_t \sim \operatorname{Poiss}(\lambda^{(b)}_t),
-}\label{eq-freq_poisson_reads}\end{equation} where the symbol
-``\(\sim\)'' is read ``distributed as.'' Furthermore, for a Poisson
-distribution, we have that
-\begin{equation}\protect\hypertarget{eq-freq_poisson_lambda}{}{
-\lambda^{(b)}_t = \left\langle r^{(b)}_t \right\rangle = 
-\left\langle 
-    \left( r^{(b)}_t - \left\langle r^{(b)}_t \right\rangle \right)^2
-\right\rangle,
-}\label{eq-freq_poisson_lambda}\end{equation} where
-\(\left\langle \cdot \right\rangle\) is the expected value. In other
-words the Poisson parameter is equal to the mean and variance of the
-distribution. The Poisson distribution has the convenient property that
-for two Poisson distributed random variables
-\(X \sim \operatorname{Poiss}(\lambda_x)\) and
-\(Y \sim \operatorname{Poiss}(\lambda_y)\), we have that
-\begin{equation}\protect\hypertarget{eq-freq_additivity}{}{
-Z \equiv X + Y \sim \operatorname{Poiss}(\lambda_x + \lambda_y).
-}\label{eq-freq_additivity}\end{equation} This additivity allows us to
-write the total number of reads at time \(t\) \(n_t\) also as a
-Poisson-distributed random variable of the form
-\begin{equation}\protect\hypertarget{eq-freq_poisson_total}{}{
-n_t \sim \operatorname{Poiss}\left( \sum_{b=1}^B \lambda^{(b)}_t \right),
-}\label{eq-freq_poisson_total}\end{equation} where the sum is taken over
-all \(B\) barcodes.
-
-If the total number of reads is given by
-Equation~\ref{eq-freq_poisson_total}, the array with the number of reads
-for each barcode at time \(t\), \(\underline{r}_t\) is then distributed
-as \begin{equation}\protect\hypertarget{eq-freq_multinomial}{}{
-\underline{r}_t \sim \operatorname{Multinomial}(n_t, \underline{f}_t),
-}\label{eq-freq_multinomial}\end{equation} where each of the \(B\)
-entries of the frequency vector \(\underline{f}_t\) is a function of the
-\(\underline{\lambda}_t\) vector, given by
-\begin{equation}\protect\hypertarget{eq-freq_freq_lambda}{}{
-f_t^{(b)} \equiv f_t^{(b)}(\underline{\lambda}_t) = 
-\frac{\lambda_t^{(b)}}{\sum_{b'=1}^B \lambda_t^{(b')}}.
-}\label{eq-freq_freq_lambda}\end{equation} In other words, we can think
-of the \(B\) barcode counts as independent Poisson samples or as a
-single multinomial draw with a random number of total draws, \(n_t\),
-and the frequency vector \(\underline{f}_t\) we are interested in.
-Notice that Equation~\ref{eq-freq_freq_lambda} is a deterministic
-function that connects the Poisson parameters to the frequencies.
-Therefore, we have the equivalence that
-\begin{equation}\protect\hypertarget{eq-freq_f_lambda_equiv}{}{
-\pi(\underline{f}_t \mid \underline{r}_t) = 
-\pi(\underline{\lambda}_t \mid \underline{r}_t),
-}\label{eq-freq_f_lambda_equiv}\end{equation} meaning that the
-uncertainty comes from the \(\underline{\lambda}_t\) vector. By Bayes
-theorem, we therefore write
-\begin{equation}\protect\hypertarget{eq-freq_lambda_bayes}{}{
-\pi(\underline{\lambda}_t \mid n_t, \underline{r}_t) \propto
-\pi(n_t, \underline{r}_t \mid \underline{\lambda}_t) \pi(\underline{\lambda}_t),
-}\label{eq-freq_lambda_bayes}\end{equation} where we explicitly include
-the dependence on \(n_t\). This does not affect the distribution or
-brings more uncertainty because \(\underline{r}_t\) already contains all
-the information to compute \(n_t\) since
-\begin{equation}\protect\hypertarget{eq-freq_n_sum_r}{}{
-n_t = \sum_{b=1}^B r_t^{(b)}.
-}\label{eq-freq_n_sum_r}\end{equation} But adding the variable allows us
-to factorize Equation~\ref{eq-freq_lambda_bayes} as
-\begin{equation}\protect\hypertarget{eq-freq_lambda_bayes_factorized}{}{
-\pi(\underline{\lambda}_t \mid n_t, \underline{r}_t) \propto
-\pi(\underline{r}_t \mid n_t, \underline{\lambda}_t)
-\pi(n_t \mid \underline{\lambda}_t)
-\pi(\underline{\lambda}_t)
-}\label{eq-freq_lambda_bayes_factorized}\end{equation} We then have
-\begin{equation}\protect\hypertarget{eq-freq_r_bayes}{}{
-\underline{r}_t \mid n_t, \underline{\lambda}_t \sim
-\operatorname{Multinomial}(n_t, \underline{f}_t(\underline{\lambda}_t)).
-}\label{eq-freq_r_bayes}\end{equation} Furthermore, we have \[
-n_t \mid \underline{\lambda}_t \sim 
-\operatorname{Poiss}\left(\sum_{b=1}^B \lambda_t^{(b)}\right).
-\]\{\#eq=freq\_n\_bayes\} Finally, for our prior
-\(\pi(\underline{\lambda}_t)\), we first assume each parameter is
-independent, i.e., \[
-\pi(\underline{\lambda}_t) = \prod_{b=1}^B \pi(\lambda_t^{(b)}).
-\] A reasonable prior for each \(\lambda_t^{(b)}\) representing the
-expected number of reads for barcode \(b\) should span several orders of
-magnitude. Furthermore, we assume that no barcode in the dataset ever
-goes extinct. Thus, no frequency can equal zero, facilitating the
-computation of the log frequency ratios needed to infer the relative
-fitness. The log-normal distribution satisfies these constraints;
-therefore, for the prior, we assume
-\begin{equation}\protect\hypertarget{eq-freq_lambda_prior}{}{
-\lambda_t^{(b)} \sim 
-\log\mathcal{N}(\mu_{\lambda_t^{(b)}}, \sigma_{\lambda_t^{(b)}}),
-}\label{eq-freq_lambda_prior}\end{equation} with
-\(\mu_{\lambda_t^{(b)}}, \sigma_{\lambda_t^{(b)}}\) as the user-defined
-parameters that characterize the prior distribution.
-
-\hypertarget{summary}{%
-\paragraph{Summary}\label{summary}}
-
-Putting all the pieces developed in this section together gives a term
-for our inference of the form
-\begin{equation}\protect\hypertarget{eq-freq_final_01}{}{
-\pi(\underline{\underline{F}} \mid \underline{\underline{R}}) \propto
-\prod_{t=1}^T\left\{
-    \pi(\underline{r}_t \mid n_t, \underline{\lambda}_t)
-    \pi(n_t \mid \underline{\lambda}_t)
-    \left[ 
-        \prod_{b=1}^B \pi(\lambda_t^{(b)})
-    \right]
-\right\}
-}\label{eq-freq_final_01}\end{equation} where
-\begin{equation}\protect\hypertarget{eq-freq_final_02}{}{
-\underline{r}_t \mid n_t, \underline{\lambda}_t \sim
-\operatorname{Multinomial}(n_t, \underline{f}_t(\underline{\lambda}_t)),
-}\label{eq-freq_final_02}\end{equation}
-\begin{equation}\protect\hypertarget{eq-freq_final_03}{}{
-n_t \mid \underline{\lambda}_t \sim 
-\operatorname{Poiss}\left(\sum_{b=1}^B \lambda_t^{(b)}\right).
-}\label{eq-freq_final_03}\end{equation} and
-\begin{equation}\protect\hypertarget{eq-freq_final_04}{}{
-\lambda_t^{(b)} \sim 
-\log\mathcal{N}(\mu_{\lambda_t^{(b)}}, \sigma_{\lambda_t^{(b)}}),
-}\label{eq-freq_final_04}\end{equation}
-
-\hypertarget{sec-bayes_meanfit}{%
-\subsubsection{\texorpdfstring{Population mean fitness uncertainty
-\(\pi(\underline{\bar{s}}_T \mid \underline{\underline{F}}, \underline{\underline{R}})\)}{Population mean fitness uncertainty \textbackslash pi(\textbackslash underline\{\textbackslash bar\{s\}\}\_T \textbackslash mid \textbackslash underline\{\textbackslash underline\{F\}\}, \textbackslash underline\{\textbackslash underline\{R\}\})}}\label{sec-bayes_meanfit}}
-
-Next, we turn our attention to the problem of determining the population
-mean fitnesses \(\underline{\bar{s}}_T\). First, we notice that our
-fitness model in Equation~\ref{eq-fitness} does not include the value of
-the raw reads. They enter the calculation indirectly through the
-inference of the frequency values we developed in
-Section~\ref{sec-bayes_freq}. This means that we can remove the
-conditioning of the value of \(\underline{\bar{s}}_T\) on the number of
-reads, obtaining a simpler probability function
-\begin{equation}\protect\hypertarget{eq-meanfit_noreads}{}{
-\pi(
-    \underline{\bar{s}}_T \mid 
-    \underline{\underline{F}}, \underline{\underline{R}}
-) = 
-\pi(
-    \underline{\bar{s}}_T \mid 
-    \underline{\underline{F}}
-).
-}\label{eq-meanfit_noreads}\end{equation} Moreover, our fitness model
-does not directly explain how the population mean fitness evolves over
-time. In other words, our model cannot explicitly compute the population
-mean fitness at time \(t+1\) from the information we have about time
-\(t\). Given this model limitation, we are led to assume that we must
-infer each \(\bar{s}_t\) independently. Expressing this for our
-inference results in
-\begin{equation}\protect\hypertarget{eq-meanfit_indep}{}{
-\pi(
-    \underline{\bar{s}}_T \mid 
-    \underline{\underline{F}}
-) =
-\prod_{t=1}^{T-1} \pi(\bar{s}_t \mid \underline{f}_t, \underline{f}_{t+1}),
-}\label{eq-meanfit_indep}\end{equation} where we split our matrix
-\(\underline{\underline{F}}\) for each time point and only kept the
-conditioning on the relevant frequencies needed to compute the mean
-fitness at time \(t\).
-
-Although our fitness model in Equation~\ref{eq-fitness} also includes
-the relative fitness \(s^{(m)}\), to infer the population mean fitness
-we only utilize data from the neutral lineages that, by definition, have
-a relative fitness \(s^{(n)} = 0\). Therefore, the conditioning on
-Equation~\ref{eq-meanfit_indep} can be further simplified by only
-keeping the frequencies of the neutral lineages, i.e.,
-\begin{equation}\protect\hypertarget{eq-meanfit_indep_neutral}{}{
-\pi(\bar{s}_t \mid \underline{f}_t, \underline{f}_{t+1}) =
-\pi(\bar{s}_t \mid \underline{f}_t^N, \underline{f}_{t+1}^N).
-}\label{eq-meanfit_indep_neutral}\end{equation}
-
-Recall that in Section~\ref{sec-fitness_model} we emphasized that the
-frequencies \(f_t^{(n)}\) do not represent the true frequency of a
-particular lineage in the population but rather a ``normalized number of
-cells.'' Therefore, it is safe to assume each of the \(N\) neutral
-lineages' frequencies is changing independently. The correlation of how
-increasing the frequency of one lineage will decrease the frequency of
-others is already captured in the model presented in
-Section~\ref{sec-bayes_freq}. Thus, we write
-\begin{equation}\protect\hypertarget{eq-meanfit_indep_lineages}{}{
-\pi(\bar{s}_t \mid \underline{f}_t^N, \underline{f}_{t+1}^N) =
-\prod_{n=1}^N \pi(\bar{s}_t \mid f_t^{(n)}, f_{t+1}^{(n)}).
-}\label{eq-meanfit_indep_lineages}\end{equation}
-
-Now, we can focus on one of the terms on the right-hand side of
-Equation~\ref{eq-meanfit_indep_lineages}. Writing Bayes theorem results
-in \begin{equation}\protect\hypertarget{eq-meanfit_bayes}{}{
-\pi(\bar{s}_t \mid f_t^{(n)}, f_{t+1}^{(n)}) \propto
-\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t) \pi(\bar{s}_t).
-}\label{eq-meanfit_bayes}\end{equation} Notice the likelihood defines
-the joint distribution of neutral barcode frequencies conditioned on the
-population mean fitness. However, rewriting our fitness model in
-Equation~\ref{eq-fitness} for a neutral lineage to leave frequencies on
-one side and fitness on the other results in
-\begin{equation}\protect\hypertarget{eq-fitness_ratio_neutral}{}{
-\frac{f_{t+1}^{(n)}}{f_t^{(n)}} = \mathrm{e}^{- \bar{s}_t\tau}.
-}\label{eq-fitness_ratio_neutral}\end{equation}
-Equation~\ref{eq-fitness_ratio_neutral} implies that our fitness model
-only relates \textbf{the ratio} of frequencies and not the individual
-values. To get around this complication, we define
-\begin{equation}\protect\hypertarget{eq-gamma_def}{}{
-\gamma_t^{(b)} \equiv \frac{f_{t+1}^{(b)}}{f_t^{(b)}},
-}\label{eq-gamma_def}\end{equation} as the ratio of frequencies between
-two adjacent time points for any barcode \(b\). This allows us to
-rewrite the joint distribution
-\(\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t)\) as
-\begin{equation}\protect\hypertarget{eq-joint_freq_gamma}{}{
-\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t) =
-\pi(f_t^{(n)}, \gamma_{t}^{(n)} \mid \bar{s}_t).
-}\label{eq-joint_freq_gamma}\end{equation} Let us rephrase this subtle
-but necessary change of variables since it is a key part of the
-inference problem: our series of independence assumptions lead us to
-Equation~\ref{eq-meanfit_bayes} that relates the value of the population
-mean fitness \(\bar{s}_t\) to the frequency of a neutral barcode at
-times \(t\) and \(t+1\). However, as shown in
-Equation~\ref{eq-fitness_ratio_neutral}, our model functionally relates
-the ratio of frequencies---that we defined as \(\gamma_t^{(n)}\)---and
-not the independent frequencies to the mean fitness. Therefore, instead
-of writing for the likelihood the joint distribution of the frequency
-values at times \(t\) and \(t+1\) conditioned on the mean fitness, we
-write the joint distribution of the barcode frequency at time \(t\) and
-the ratio of the frequencies. These \textbf{must be} equivalent joint
-distributions since there is a one-to-one mapping between
-\(\gamma_t^{(n)}\) and \(f_{t+1}^{(n)}\) for a given value of
-\(f_t^{(n)}\). Another way to phrase this is to say that knowing the
-frequency at time \(t\) and at time \(t+1\) provides the same amount of
-information as knowing the frequency at time \(t\) and the ratio of the
-frequencies. This is because if we want to obtain \(f_{t+1}^{(n)}\)
-given this information, we simply compute
-\begin{equation}\protect\hypertarget{eq-meanfit_ft+1_gamma}{}{
-f_{t+1}^{(n)} = \gamma_t^{(n)} f_t^{(n)}.
-}\label{eq-meanfit_ft+1_gamma}\end{equation}
-
-The real advantage of rewriting the joint distribution as in
-Equation~\ref{eq-joint_freq_gamma} comes from splitting this joint
-distribution as a product of conditional distributions of the form
-\begin{equation}\protect\hypertarget{eq-joint_to_prod_gamma}{}{
-\pi(f_t^{(n)}, \gamma_{t}^{(n)} \mid \bar{s}_t) =
-\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t)
-\pi(\gamma_{t}^{(n)} \mid \bar{s}_t).
-}\label{eq-joint_to_prod_gamma}\end{equation} Written in this form, we
-can finally propose a probabilistic model for how the mean fitness
-relates to the frequency ratios we determine in our experiments. The
-second term on the right-hand side of
-Equation~\ref{eq-joint_to_prod_gamma} relates how the determined
-frequency ratio \(\gamma_t^{(b)}\) relates to the mean fitness
-\(\bar{s}_t\). From Equation~\ref{eq-fitness_ratio_neutral} and
-Equation~\ref{eq-gamma_def}, we can write
-\begin{equation}\protect\hypertarget{eq-log_ratio_neutral}{}{
-\ln \gamma_t^{(n)} = - \bar{s}_t + \varepsilon_t^{(n)},
-}\label{eq-log_ratio_neutral}\end{equation} where, for simplicity, we
-set \(\tau = 1\). Note that we added an extra term,
-\(\varepsilon_t^{(n)}\), characterizing the deviations of the
-measurements from the theoretical model. We assume these errors are
-normally distributed with mean zero and some standard deviation
-\(\sigma_t\), implying that
-\begin{equation}\protect\hypertarget{eq-log_gamma_normal}{}{
-\ln \gamma_t^{(n)} \mid \bar{s}_t, \sigma_t  \sim 
-\mathcal{N}\left(-\bar{s}_t, \sigma_t \right),
-}\label{eq-log_gamma_normal}\end{equation} where we include the nuisance
-parameter \(\sigma_t\) to be determined. If we assume the log frequency
-ratio is normally distributed, this implies the frequency ratio itself
-is distributed log-normal. This means that
-\begin{equation}\protect\hypertarget{eq-meanfit_likelihood}{}{
-\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t  \sim 
-\log \mathcal{N}\left(-\bar{s}_t, \sigma_t \right).
-}\label{eq-meanfit_likelihood}\end{equation} Having added the nuisance
-parameter \(\sigma_t\) implies that we must update
-Equation~\ref{eq-meanfit_bayes} to
-\begin{equation}\protect\hypertarget{eq-meanfit_bayes_full}{}{
-\pi(\bar{s}_t, \sigma_t \mid f_t^{(n)}, f_{t+1}^{(n)}) \propto
-\pi(f_t^{(n)}, \gamma_t^{(n)} \mid \bar{s}_t, \sigma_t) 
-\pi(\bar{s}_t) \pi(\sigma_t),
-}\label{eq-meanfit_bayes_full}\end{equation} where we assume the prior
-for each parameter is independent, i.e.,
-\begin{equation}\protect\hypertarget{eq-meanfit_indep_prior}{}{
-\pi(\bar{s}_t, \sigma_t) = \pi(\bar{s}_t) \pi(\sigma_t).
-}\label{eq-meanfit_indep_prior}\end{equation} For numerical stability,
-we will select weakly-informative priors for both of these parameters.
-In the case of the nuisance parameter \(\sigma_t\), the prior must be
-restricted to positive values only, since standard deviations cannot be
-negative.
-
-For the first term on the right-hand side of
-Equation~\ref{eq-joint_to_prod_gamma},
-\(\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t)\), we remove the
-conditioning on the population mean fitness since it does not add any
-information on top of what the frequency ratio \(\gamma_t^{(n)}\)
-already gives. Therefore, we have
-\begin{equation}\protect\hypertarget{eq-freq_cond_gamma}{}{
-\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t) =
-\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}).
-}\label{eq-freq_cond_gamma}\end{equation} The right-hand side of
-Equation~\ref{eq-freq_cond_gamma} asks us to compute the probability of
-observing a frequency value \(f_t^{(n)}\) given that we get to observe
-the ratio \(\gamma_{t}^{(n)}\). If the ratio happened to be
-\(\gamma_{t}^{(n)} = 2\), we could have \(f_{t+1}^{(n)} = 1\) and
-\(f_{t+1}^{(n)} = 0.5\), for example. Although, it would be equally
-likely that \(f_{t+1}^{(n)} = 0.6\) and \(f_{t+1}^{(n)} = 0.3\) or
-\(f_{t+1}^{(n)} = 0.1\) and \(f_{t+1}^{(n)} = 0.05\) for that matter. If
-we only get to observe the frequency ratio \(\gamma_t^{(n)}\), we know
-that the numerator \(f_{t+1}^{(n)}\) can only take values between zero
-and one, all of them being equally likely given only the information on
-the ratio. As a consequence, the value of the frequency in the
-denominator \(f_{t}^{(n)}\) is restricted to fall in the range
-\begin{equation}\protect\hypertarget{eq-freq_given_gamma_range}{}{
-f_{t}^{(n)} \in \left(0, \frac{1}{\gamma_t^{(n)}} \right].
-}\label{eq-freq_given_gamma_range}\end{equation} A priori, we do not
-have any reason to favor any value over any other, therefore it is
-natural to write
-\begin{equation}\protect\hypertarget{eq-f_given_gamma_uniform}{}{
-f_t^{(n)} \mid \gamma_t^{(n)} \sim 
-\operatorname{Uniform}\left( 0, \frac{1}{\gamma_t^{(n)}} \right).
-}\label{eq-f_given_gamma_uniform}\end{equation}
-
-\hypertarget{summary-1}{%
-\paragraph{Summary}\label{summary-1}}
-
-Putting all the pieces we have developed in this section together
-results in an inference for the population mean fitness values of the
-form \begin{equation}\protect\hypertarget{eq-meanfit_full_inference}{}{
-\pi(
-    \underline{\bar{s}}_T, \underline{\sigma}_T \mid \underline{\underline{F}}
-) \propto
-\prod_{t=1}^{T-1} \left\{
-    \prod_{n=1}^N \left[
-        \pi(f_t^{(n)} \mid \gamma_t^{(n)}) 
-        \pi(\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t)
-    \right]
-    \pi(\bar{s}_t) \pi(\sigma_t)
-\right\},
-}\label{eq-meanfit_full_inference}\end{equation} where we have
-\begin{equation}\protect\hypertarget{eq-meanfit_eq1}{}{
-f_t^{(n)} \mid \gamma_t^{(n)} \sim 
-\operatorname{Uniform} \left(0, \frac{1}{\gamma_t^{(n)}} \right),
-}\label{eq-meanfit_eq1}\end{equation}
-\begin{equation}\protect\hypertarget{eq-meanfit_eq2}{}{
-\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t \sim 
-\log\mathcal{N}(\bar{s}_t, \sigma_t),
-}\label{eq-meanfit_eq2}\end{equation}
-\begin{equation}\protect\hypertarget{eq-meanfit_eq3}{}{
-\bar{s}_t \sim \mathcal{N}(0, \sigma_{\bar{s}_t}),
-}\label{eq-meanfit_eq3}\end{equation} and
-\begin{equation}\protect\hypertarget{eq-meanfit_eq4}{}{
-\sigma_t \sim \log\mathcal{N}(\mu_{\sigma_t}, \sigma_{\sigma_t}),
-}\label{eq-meanfit_eq4}\end{equation} where \(\sigma_{\bar{s}_t}\),
-\(\mu_{\sigma_t}\), and \(\sigma_{\sigma_t}\) are user-defined
-parameters.
-
-\hypertarget{sec-bayes_mutfit}{%
-\subsubsection{\texorpdfstring{Mutant relative fitness uncertainty
-\(\pi(\underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}}, \underline{\underline{R}})\)}{Mutant relative fitness uncertainty \textbackslash pi(\textbackslash underline\{s\}\^{}M \textbackslash mid \textbackslash underline\{\textbackslash bar\{s\}\}\_T, \textbackslash underline\{\textbackslash underline\{F\}\}, \textbackslash underline\{\textbackslash underline\{R\}\})}}\label{sec-bayes_mutfit}}
-
-The last piece of our inference is the piece that we care about the
-most: the probability distribution of all the mutants' relative fitness,
-given the inferred population mean fitness and the frequencies. First,
-we assume that all fitness values are independent of each other. This
-allows us to write
-\begin{equation}\protect\hypertarget{eq-mutfit_indep}{}{
-\pi(
-    \underline{s}^M \mid 
-    \underline{\bar{s}}_T, \underline{\underline{F}}, \underline{\underline{R}}
-) = 
-\prod_{m=1}^M \pi(
-    s^{(m)} \mid
-    \underline{\bar{s}}_T, \underline{\underline{F}}, \underline{\underline{R}}
-).
-}\label{eq-mutfit_indep}\end{equation} Furthermore, as was the case with
-the population mean fitness, our fitness model relates frequencies, not
-raw reads. Moreover, the fitness value of mutant \(m\) only depends on
-the frequencies of such mutant. Therefore, we can simplify the
-conditioning to
-\begin{equation}\protect\hypertarget{eq-mutfit_cond_simple}{}{
-\pi(
-    s^{(m)} \mid
-    \underline{\bar{s}}_T, \underline{\underline{F}}, \underline{\underline{R}}
-) = 
-\pi(s^{(m)} \mid \underline{\bar{s}}_T, \underline{f}^{(m)}),
-}\label{eq-mutfit_cond_simple}\end{equation} where
-\begin{equation}\protect\hypertarget{eq-mutfit_timeseries_def}{}{
-\underline{f}^{(m)} = (f_0^{(m)}, f_1^{(m)}, \ldots, f_T^{(m)})^\dagger,
-}\label{eq-mutfit_timeseries_def}\end{equation} is the vector containing
-the frequency time series for mutant \(m\). Writing Bayes' theorem for
-the right-hand side of Equation~\ref{eq-mutfit_cond_simple} results in
-\begin{equation}\protect\hypertarget{eq-mutfit_bayes}{}{
-\pi(s^{(m)} \mid \underline{\bar{s}}_T, \underline{f}^{(m)}) \propto
-\pi(\underline{f}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)})
-\pi(s^{(m)} \mid \underline{\bar{s}}_T).
-}\label{eq-mutfit_bayes}\end{equation} Notice the conditioning on the
-mean fitness values \(\underline{\bar{s}}_T\) is not inverted since we
-already inferred these values.
-
-Following the logic used in Section~\ref{sec-bayes_meanfit}, let us
-define \begin{equation}\protect\hypertarget{eq-gamma_vec_def}{}{
-\underline{\gamma}^{(m)} = 
-(\gamma_0^{(m)}, \gamma_1^{(m)}, \ldots, \gamma_{T-1}^{m})^\dagger,
-}\label{eq-gamma_vec_def}\end{equation} where each entry
-\(\gamma_t^{(m)}\) is defined by Equation~\ref{eq-gamma_def}. In the
-same way we rewrote the joint distribution between two adjacent time
-point frequencies to the joint distribution between one of the
-frequencies and the ratio of both frequencies in
-Equation~\ref{eq-joint_freq_gamma}, we can rewrite the joint
-distribution of the frequency time series for mutant \(m\) as
-\begin{equation}\protect\hypertarget{eq-joint_freq_gammas_mutfit}{}{
-\pi(\underline{f}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}) =
-\pi(f_0^{(m)}, \underline{\gamma}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}).
-}\label{eq-joint_freq_gammas_mutfit}\end{equation} One can think about
-Equation~\ref{eq-joint_freq_gammas_mutfit} as saying that knowing the
-individual frequencies at each time point contain equivalent information
-as knowing the initial frequency and the subsequent ratios of
-frequencies. This is because if we want to know the value of
-\(f_1^{(m)}\) given the ratios, we only need to compute
-\begin{equation}\protect\hypertarget{eq-f1_from_ratio}{}{
-f_1^{(m)} = \gamma_0^{(m)} f_0^{(m)}.
-}\label{eq-f1_from_ratio}\end{equation} Moreover, if we want to know
-\(f_2^{(m)}\), we have
-\begin{equation}\protect\hypertarget{eq-f2_from_ratio}{}{
-f_2^{(m)} = \gamma_1^{(m)} f_1^{(m)} =
-\gamma_1^{(m)} \left(\gamma_0^{(m)} f_0^{(m)}\right),
-}\label{eq-f2_from_ratio}\end{equation} and so on. We can then write the
-joint distribution on the right-hand side of
-Equation~\ref{eq-joint_freq_gammas_mutfit} as a product of conditional
-distributions of the form
-\begin{equation}\protect\hypertarget{eq-mutfit_joint_to_cond}{}{
-\begin{aligned}
-\pi(f_0^{(m)}, \underline{\gamma}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}) =
-&\pi(
-    f_0^{(m)} \mid 
-    \gamma_0^{(m)}, \ldots, \gamma_{T-1}^{(m)}, \underline{\bar{s}}_T, s^{(m)}
-) \times \\
-&\pi(
-    \gamma_0^{(m)} \mid 
-    \gamma_1^{(m)}, \ldots, \gamma_{T-1}^{(m)}, \underline{\bar{s}}_T, s^{(m)}
-) \times \\
-&\pi(
-    \gamma_1^{(m)} \mid 
-    \gamma_2^{(m)}, \ldots, \gamma_{T-1}^{(m)}, \underline{\bar{s}}_T, s^{(m)}
-) \times \\
-&\vdots \\
-&\pi(
-    \gamma_{T-2}^{(m)} \mid \gamma_{T-1}^{(m)}, \underline{\bar{s}}_T, s^{(m)}
-) \times \\
-&\pi(\gamma_{T-1}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}).
-\end{aligned}
-}\label{eq-mutfit_joint_to_cond}\end{equation} Writing the fitness model
-in Equation~\ref{eq-fitness} as \[
-\gamma_t^{(m)} = \frac{f_{t+1}^{(m)}}{f_t^{(m)}} = 
-\mathrm{e}^{(s^{(m)} - s_t)\tau},
-\] reveals that the value of each of the ratios \(\gamma_t^{(m)}\) only
-depends on the corresponding fitness value \(\bar{s}_t\) and the
-relative fitness \(s^{(m)}\). Therefore, we can remove most of the
-conditioning on the right-hand side of
-Equation~\ref{eq-mutfit_joint_to_cond}, resulting in a much simpler
-joint distribution of the form
-\begin{equation}\protect\hypertarget{eq-mutfit_joint_to_cond_simple}{}{
-\begin{aligned}
-\pi(f_0^{(m)}, \underline{\gamma}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}) =
-&\pi(f_0^{(m)} \mid \gamma_0^{(m)}) \times \\
-&\pi(\gamma_0^{(m)} \mid \bar{s}_0, s^{(m)}) \times \\
-&\pi(\gamma_1^{(m)} \mid \bar{s}_1, s^{(m)}) \times \\
-&\vdots \\
-&\pi(\gamma_{T-2}^{(m)} \mid \bar{s}_{T-2}, s^{(m)}) \times \\
-&\pi(\gamma_{T-1}^{(m)} \mid \bar{s}_{T-1}, s^{(m)}),
-\end{aligned}
-}\label{eq-mutfit_joint_to_cond_simple}\end{equation} where for the
-first term on the right-hand side of
-Equation~\ref{eq-mutfit_joint_to_cond_simple} we apply the same logic as
-in Equation~\ref{eq-freq_cond_gamma} to remove all other dependencies.
-We emphasize that although Equation~\ref{eq-mutfit_joint_to_cond_simple}
-looks like a series of independent inferences, the value of the relative
-fitness \(s^{(m)}\) is shared among all of them. This means that the
-parameter is not inferred individually for each time point, resulting in
-different estimates of the parameter, but each time point contributes
-independently to the inference of a single estimate of \(s^{(m)}\).
-
-Using equivalent arguments to those in Section~\ref{sec-bayes_meanfit},
-we assume \[
-f_0^{(m)} \mid \gamma_0^{(m)} \sim 
-\operatorname{Uniform}\left(0, \frac{1}{\gamma_0^{(m)}} \right),
-\] and \begin{equation}\protect\hypertarget{eq-mutfit_lognormal}{}{
-\gamma_t^{(m)} \mid \bar{s}_t, s^{(m)}, \sigma^{(m)} \sim 
-\log\mathcal{N}\left(s^{(m)} - \bar{s}_t, \sigma^{(m)} \right),
-}\label{eq-mutfit_lognormal}\end{equation} where we add the nuisance
-parameter \(\sigma^{(m)}\) to the inference. Notice that this parameter
-is not indexed by time. This means that we assume the deviations from
-the theoretical prediction do not depend on time, but only on the
-mutant. Adding the nuisance parameter demands us to update
-Equation~\ref{eq-mutfit_bayes} to
-\begin{equation}\protect\hypertarget{eq-mutfit_bayes_full}{}{
-\pi(
-    s^{(m)}, \sigma^{(m)} \mid \underline{\bar{s}}_T, \underline{f}^{(m)}
-) \propto
-\pi(\underline{f}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}, \sigma^{(m)})
-\pi(s^{(m)}) \pi(\sigma^{(m)}),
-}\label{eq-mutfit_bayes_full}\end{equation} where we assume independent
-priors for both parameters. We also removed the conditioning on the
-values of the mean fitness as knowing such values does not change our
-prior information about the possible range of values that the parameters
-can take. As with the priors on Section~\ref{sec-bayes_meanfit}, we will
-assign weakly-informative priors to these parameters.
-
-\hypertarget{summary-2}{%
-\paragraph{Summary}\label{summary-2}}
-
-With all pieces in place, we write the full inference of the relative
-fitness values as
-\begin{equation}\protect\hypertarget{eq-mutfit_full_inference}{}{
-\pi(
-    \underline{s}^M ,\underline{\sigma}^M \mid 
-    \underline{\bar{s}}_T, \underline{\underline{F}}
-) \propto
-\prod_{m=1}^M \left\{ 
-    \pi(f_0^{(m)} \mid \gamma_0^{(m)})
-    \prod_{t=0}^{T-1} \left[
-        \pi(\gamma_t^{(m)} \mid \bar{s}_t, s^{(m)}, \sigma^{(m)})
-    \right]
-    \pi(s^{(m)}) \pi(\sigma^{(m)})
-\right\},
-}\label{eq-mutfit_full_inference}\end{equation} where
-\begin{equation}\protect\hypertarget{eq-mutfit_eq1}{}{
-f_0^{(m)} \mid \gamma_0^{(m)} \sim 
-\operatorname{Uniform}\left(0, \frac{1}{\gamma_0^{(m)}} \right),
-}\label{eq-mutfit_eq1}\end{equation}
-\begin{equation}\protect\hypertarget{eq-mutfit_eq2}{}{
-\gamma_t^{(m)} \mid \bar{s}_t, s^{(m)}, \sigma^{(m)} \sim 
-\log\mathcal{N}\left(s^{(m)} - \bar{s}_t, \sigma^{(m)} \right),
-}\label{eq-mutfit_eq2}\end{equation}
-\begin{equation}\protect\hypertarget{eq-mutfit_eq3}{}{
-s^{(m)} \sim \mathcal{N}(0, \sigma_{s^{(m)}}),
-}\label{eq-mutfit_eq3}\end{equation} and
-\begin{equation}\protect\hypertarget{eq-mutfit_eq4}{}{
-\sigma^{(m)} \sim \log\mathcal{N}(\mu_{\sigma^{(m)}}, \sigma_{\sigma^{(m)}}),
-}\label{eq-mutfit_eq4}\end{equation} where \(\sigma_{s^{(m)}}\),
-\(\mu_{\sigma^{(m)}}\), and \(\sigma_{\sigma^{(m)}}\) are user-defined
-parameters.
-
-\hypertarget{sec-hierarchical_model}{%
-\subsubsection{Hierarchical models for multiple experimental
-replicates}\label{sec-hierarchical_model}}
-
-As detailed in Section~\ref{sec-replicates} of the main text, we define
-a Bayesian hierarchical model to analyze data from multiple experimental
-replicates. The implementation requires only slightly modifying the base
-model detailed in the previous sections. The hierarchical model defines
-a hyper-fitness parameter \(\theta^{(m)}\) for every non-neutral
-barcode. We can thus collect all of the \(M\) hyperparameters in an
-array of the form
-\begin{equation}\protect\hypertarget{eq-hier_hyper_vector}{}{
-\underline{\theta}^M = (\theta^{(1)}, \ldots, \theta^{(M)})^\dagger.
-}\label{eq-hier_hyper_vector}\end{equation} Our data now consists of a
-series of matrices \(\underline{\underline{R}}_{[j]}\), where the
-subindex \([j]\) refers to the \(j\)-th experimental replicate. These
-matrices need not have the same number of rows, as the time points
-measured for each replicate can vary. The statistical model we must
-define is then of the form
-\begin{equation}\protect\hypertarget{eq-hier_bayes_full}{}{
-\begin{aligned}
-\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\},
-    \{\underline{\underline{F}}_{[j]}\} \mid
-    \{\underline{\underline{R}}_{[j]}\}
-) \propto \;
-&\pi(
-    \{\underline{\underline{R}}_{[j]}\} \mid
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\},
-    \{\underline{\underline{F}}_{[j]}\} 
-) \times \\\
-&\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\},
-    \{\underline{\underline{F}}_{[j]}\} 
-)
-\end{aligned}
-}\label{eq-hier_bayes_full}\end{equation} where the parameters within
-curly braces with subindex \([j]\) indicate one set of parameters per
-experimental replicate. For example,
-\begin{equation}\protect\hypertarget{eq-hier_array}{}{
-\{\underline{s}^M_{[j]}\} = \{
-    \underline{s}^M_{[1]}, \underline{s}^M_{[2]}, \ldots \underline{s}^M_{[E]}
-\},
-}\label{eq-hier_array}\end{equation} where \(E\) is the number of
-experimental replicates.
-
-Given the dependencies between the variables, we can factorize
-Equation~\ref{eq-hier_bayes_full} to be of the form
-\begin{equation}\protect\hypertarget{eq-hier_factorized}{}{
-\begin{aligned}
-\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\},
-    \{\underline{\underline{F}}_{[j]}\} \mid
-    \{\underline{\underline{R}}_{[j]}\}
-) =
-&\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\} \mid
-    \{\underline{\bar{s}}_{T[j]}\},
-    \{\underline{\underline{F}}_{[j]}\} 
-) \times \\
-&\pi(
-    \{\underline{\bar{s}}_{T[j]}\} \mid
-    \{\underline{\underline{F}}_{[j]}\}
-) \times \\
-&\pi(
-    \{\underline{\underline{F}}_{[j]}\} \mid
-    \{\underline{\underline{R}}_{[j]}\}
-)
-\end{aligned}
-}\label{eq-hier_factorized}\end{equation} Furthermore, the hierarchical
-structure only connects the replicates via the relative fitness
-parameters. This means that the population mean fitness values and the
-frequencies can be independently inferred for each dataset. This allows
-us to rewrite the right-hand side of Equation~\ref{eq-hier_factorized}
-as
-\begin{equation}\protect\hypertarget{eq-hier_bayes_factorized_replicate}{}{
-\begin{aligned}
-\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\},
-    \{\underline{\underline{F}}_{[j]}\} \mid
-    \{\underline{\underline{R}}_{[j]}\}
-) =
-&\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\} \mid
-    \{\underline{\bar{s}}_{T[j]}\},
-    \{\underline{\underline{F}}_{[j]}\} 
-) \times \\
-&\prod_{j=1}^E \left[
-    \pi(
-        \underline{\bar{s}}_{T[j]} \mid
-        \underline{\underline{F}}_{[j]}
-    )
-    \pi(
-        \underline{\underline{F}}_{[j]} \mid
-        \underline{\underline{R}}_{[j]}
-    )
-\right].
-\end{aligned}
-}\label{eq-hier_bayes_factorized_replicate}\end{equation} The terms
-inside the square brackets in
-Equation~\ref{eq-hier_bayes_factorized_replicate} take the same
-functional form as those derived in Section~\ref{sec-bayes_meanfit} and
-Section~\ref{sec-bayes_freq}. Therefore, to implement the desired
-hierarchical model, we only need to focus on the first term on the
-right-hand side of Equation~\ref{eq-hier_bayes_factorized_replicate}. A
-way to think about the structure of the hierarchical model is as
-follows: imagine each genotype as a ``\emph{true}'' relative fitness
-value. However, every time we perform an experiment, small variations in
-the biotic and abiotic conditions---also known as batch effects---might
-result in small deviations from this value. We model this by defining a
-distribution for the hyper-fitness parameter---the ground truth we are
-interested in---and having each experimental replicate sample from this
-hyper-parameter distribution to determine the ``\emph{local}'' fitness
-value. The wider the hyper-parameter distribution is the more
-variability between experimental replicates.
-
-Writing Bayes' theorem for the first term in
-Equation~\ref{eq-hier_bayes_factorized_replicate} results in
-\begin{equation}\protect\hypertarget{eq-hier_bayes}{}{
-\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\} \mid
-    \{\underline{\underline{F}}_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\}
-) \propto
-\pi(
-    \{\underline{\underline{F}}_{[j]}\} \mid
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\}
-)
-\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\} \mid
-    \{\underline{\bar{s}}_{T[j]}\}
-),
-}\label{eq-hier_bayes}\end{equation} where we leave the conditioning on
-the population mean fitness as we did in Section~\ref{sec-bayes_mutfit}.
-This expression can be simplified in two ways. First, the frequency
-values for each experimental replicate depend directly on the local
-fitness values and the corresponding population mean fitness, as the
-relationship between experimental replicates only occurs through the
-relative fitness values. Therefore, we can write
-\begin{equation}\protect\hypertarget{eq-hier_bayes_exp_indep}{}{
-\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\} \mid
-    \{\underline{\underline{F}}_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\}
-) \propto
-\prod_{j=1}^E\left[
-    \pi(
-        \underline{\underline{F}}_{[j]} \mid
-        \underline{s}^M_{[j]},
-        \underline{\bar{s}}_{T[j]}
-    )
-\right]
-\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\} \mid
-    \{\underline{\bar{s}}_{T[j]}\}
-).
-}\label{eq-hier_bayes_exp_indep}\end{equation} Second, the relationship
-between the hyper-fitness and the local fitness values allows us to
-write their joint distribution as a conditional distribution where local
-fitness values depend on the global hyper-fitness value, obtaining
-\begin{equation}\protect\hypertarget{eq-hier_bayes_exp_indep}{}{
-\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\} \mid
-    \{\underline{\underline{F}}_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\}
-) \propto
-\prod_{j=1}^E\left[
-    \pi(
-        \underline{\underline{F}}_{[j]} \mid
-        \underline{s}^M_{[j]},
-        \underline{\bar{s}}_{T[j]}
-    )
-    \pi(
-        \underline{s}^M_{[j]} \mid
-        \underline{\theta}^M
-    )
-\right] 
-\pi(
-    \underline{\theta}^M
-).
-}\label{eq-hier_bayes_exp_indep}\end{equation} Notice we removed the
-conditioning on the population mean fitness as our prior expectations of
-what the global hyper-fitness or local fitness value might be do not
-depend on these nuisance parameters.
-
-The first term on the right-hand side of
-Equation~\ref{eq-hier_bayes_exp_indep} takes the same functional form as
-the one derived in Section~\ref{sec-bayes_mutfit}. Therefore, all we are
-left with is to determine the functional forms for the hyper--prior
-\(\pi(\underline{\theta}^M)\), and the conditional probability
-\(\pi(\underline{s}^M_{[j]} \mid \underline{\theta}^M)\). In analogy to
-the assumptions used for the fitness values in
-Section~\ref{sec-bayes_mutfit}, we define the value of each
-hyper-fitness as independent. This means that we have
-\begin{equation}\protect\hypertarget{eq-hier_hyper_prior}{}{
-\pi(\underline{\theta}^M) = \prod_{m=1}^M \pi(\theta^{(m)}).
-}\label{eq-hier_hyper_prior}\end{equation} Furthermore, we assume this
-prior is of the form
-\begin{equation}\protect\hypertarget{eq-hier_hyper_prior_02}{}{
-\theta^{(m)} \sim \mathcal{N}(\mu_{\theta^{(m)}}, \sigma_{\theta^{(m)}}),
-}\label{eq-hier_hyper_prior_02}\end{equation} where
-\(\mu_{\theta^{(m)}}\) and \(\sigma_{\theta^{(m)}}\) are user-defined
-parameters encoding the prior expectations on the fitness values.
-
-For the conditional distribution
-\(\pi(\underline{s}^M_{[j]} \mid \underline{\theta}^M)\), we use the
-so-called non-centered parametrization that avoids some of the intrinsic
-degeneracies associated with hierarchical models
-\autocite{betancourt2013}. We invite the reader to check
-\href{https://betanalpha.github.io/assets/case_studies/hierarchical_modeling.html}{this
-excellent blog} explaining the difficulties of working with hierarchical
-models. This non-centered parameterization implies that we introduce two
-nuisance parameters such that the local fitness \(s_{[j]}^{(m)}\) is
-computed as \begin{equation}\protect\hypertarget{eq-hier_noncentered}{}{
-s_{[j]}^{(m)} = \theta^{(m)} + 
-(\tau_{[j]}^{(m)} \times \xi_{[j]}^{(m)}),
-}\label{eq-hier_noncentered}\end{equation} where \(\theta^{(m)}\) is the
-corresponding genotype hyper-fitness value, \(\xi_{[j]}^{(m)}\) is a
-standard normal random variable, i.e.,
-\begin{equation}\protect\hypertarget{eq-hier_stdnormal}{}{
-\xi_{[j]}^{(m)} \sim \mathcal{N}(0, 1),
-}\label{eq-hier_stdnormal}\end{equation} that allows deviations from the
-hyper-fitness value to be either positive or negative, and
-\(\tau_{[j]}^{(m)}\) is a strictly positive random variable that
-characterizes the deviation of the local fitness value from the global
-hyper-fitness. We assume
-\begin{equation}\protect\hypertarget{eq-hier_tau}{}{
-\tau_{[j]}^{(m)} \sim \log\mathcal{N}(
-    \mu_{\tau_{[j]}^{(m)}}, \sigma_{\tau_{[j]}^{(m)}}
-)
-}\label{eq-hier_tau}\end{equation} where \(\mu_{\tau_{[j]}^{(m)}}\) and
-\(\sigma_{\tau_{[j]}^{(m)}}\) are user-defined parameters capturing the
-expected magnitude of the batch effects.
-
-\hypertarget{defining-prior-probabilities}{%
-\subsubsection{Defining prior
-probabilities}\label{defining-prior-probabilities}}
-
-One aspect commonly associated---in both positive and negative ways---to
-Bayesian analysis is the definition of prior probabilities. On the one
-hand, the naive textbook version of Bayesian analysis defines the prior
-as encoding the information we have about the inference in question
-before acquiring any data. This is the ``ideal'' use of priors that,
-whenever possible, should be implemented. On the other hand, for most
-practitioners of Bayesian statistics in the age of big data, the
-definition of prior becomes a tool to ensure the convergence of sampling
-algorithms such as MCMC \autocite{gelman2017}. However, for our
-particular problem, although we deal with large amounts of data
-(inferences can be made for \textgreater{} 10K barcodes over multiple
-time points, resulting in \textgreater{} 100K parameters), each barcode
-has very little data, as they are measured only once per time point over
-\textless{} 10 growth-dilution cycles. Furthermore, it is incredibly
-challenging to understand the noise sources related to culturing
-conditions, DNA extraction, library preparation, etc., and encode them
-into reasonable prior distributions.
-
-Empirically, our approach for this work defined the priors based solely
-on the neutral lineage data, as they represent the only repeated
-measurements of a single genotype in our experimental design. We
-acknowledge that defining the priors after observing the data might be
-considered an incoherent inference. However, as expressed by
-\textcite{gelman2017}
-
-\begin{quote}
-Incoherence is an unavoidable aspect of much real-world data analysis;
-and, indeed, one might argue that as scientists we learn the most from
-the anomalies and reassessments associated with episodes of incoherence.
-\end{quote}
-
-With this in mind, we leave it to the reader to judge the selection of
-priors. Furthermore, the software package associated with this work,
-\texttt{BarBay.jl}, is written so that users can experiment with
-different prior selection criteria that fit their needs. We strongly
-advocate that statistics should not be done in a black-box fit-all tool
-mindset but rather as a formal way to encode the assumptions behind the
-analysis, subject to constructive criticism. With this philosophical
-baggage behind us, let us now focus on how the priors used for this work
-were selected.
-
-\hypertarget{naive-neutral-lineage-based-priors}{%
-\paragraph{Naive neutral lineage-based
-priors}\label{naive-neutral-lineage-based-priors}}
-
-For the base model presented in this work, the user-defined prior
-parameters include the following:
-
-\begin{itemize}
-\item
-  Prior on population mean fitness (one per pair of adjacent time
-  points) \begin{equation}\protect\hypertarget{eq-prior_meanfit_mean}{}{
-  \bar{s}_t \sim \mathcal{N}(\mu_{\bar{s}_t}, \sigma_{{\bar{s}_t}}).
-  }\label{eq-prior_meanfit_mean}\end{equation}
-\item
-  Prior on standard deviation associated with neutral lineages
-  likelihood function (one per pair of adjacent time points)
-  \begin{equation}\protect\hypertarget{eq-prior_meanfit_var}{}{
-  \sigma_t \sim \log\mathcal{N}(\mu_{\sigma_t}, \sigma_{\sigma_t}).
-  }\label{eq-prior_meanfit_var}\end{equation}
-\item
-  Prior on relative fitness (one per non-neutral barcode)
-  \begin{equation}\protect\hypertarget{eq-prior_fit_mean}{}{
-  s^{(m)} \sim \mathcal{N}(\mu_{s^{(m)}}, \sigma_{s^{(m)}}).
-  }\label{eq-prior_fit_mean}\end{equation}
-\item
-  Prior on standard deviation associated with non-neutral lineages
-  likelihood function (one per non-neutral barcode)
-  \begin{equation}\protect\hypertarget{eq-prior_fit_var}{}{
-  \sigma^{(m)} \sim \log\mathcal{N}(\mu_{\sigma^{(m)}}, \sigma_{\sigma^{(m)}})
-  }\label{eq-prior_fit_var}\end{equation}
-\end{itemize}
-
-The \texttt{BarBay.jl} package includes a function \texttt{naive\_prior}
-within the \texttt{stats} module. This function utilizes the data from
-the neutral lineages to determine some of the prior parameters to
-facilitate the inference algorithm's numerical convergence. In
-particular, it defines the population mean fitness parameter
-\(\mu_{\bar{s}_t}\) as
-\begin{equation}\protect\hypertarget{eq-prior_meanfit_mean_data}{}{
-\mu_{\bar{s}_t} = \frac{1}{N} \sum_{n=1}^N 
--\ln\left( \frac{r_{t+1}^{(n)}}{r_t^{(n)}} \right),
-}\label{eq-prior_meanfit_mean_data}\end{equation} where \(N\) is the
-number of neutral lineages and \(r_t^{(n)}\) is the number of neutral
-lineages. In other words, it defines the mean of the prior distribution
-as the mean of what one naively would compute from the neutral lineages,
-discarding cases where the ratio diverges because the denominator
-\(r_t^{(n)} = 0\). For the variance parameter, we chose a value
-\(\sigma_{\bar{s}_t} = 0.05\).
-
-Furthermore, the \texttt{naive\_prior} function defines the mean of the
-variance parameter as the standard deviation of the log frequency ratios
-for the neutral lineages, i.e.,
-\begin{equation}\protect\hypertarget{eq-prior_meanfit_var_data}{}{
-\mu_{\sigma_t} = \sqrt{
-    \operatorname{Var}\left(\frac{r_{t+1}^{(n)}}{r_t^{(n)}}\right)
-},
-}\label{eq-prior_meanfit_var_data}\end{equation} where
-\(\operatorname{Var}\) is the sample variance. This same value was
-utilized for the mean of the non-neutral barcode variance
-\(\mu_{\sigma^{(m)}}\). While we assigne the corresponding variances to
-be \(\sigma_{\sigma_t} = \sigma_{\sigma^{(m)}} = 1\).
-
-\hypertarget{sec-ppc}{%
-\subsection{Posterior predictive checks}\label{sec-ppc}}
-
-Throughout the main text, we allude to the concept of posterior
-predictive checks as a formal way to assess the accuracy of our
-inference pipeline. Here, we explain the mechanics behind the
-computation of these credible regions, given the output of the
-inference.
-
-Bayesian models encode what is known as a \emph{generative model}. This
-statement means that in our definition of the likelihood function and
-the prior distribution, we, as modelers, propose a mathematical function
-that captures all relevant relationships between unobserved (latent)
-variables. Therefore, when these latent variables are input into the
-mathematical model, this function \emph{generates} data that should be,
-in principle, indistinguishable from the real observations if the model
-is a good account of the underlying processes involved in the phenomena
-of interest. This generative model implies that once we run the
-inference process and update our posterior beliefs about the state of
-the latent variables, we can input back the inferred values to our model
-and generate synthetic data. Furthermore, we can repeat this process
-multiple times to compute the range where we expect to observe our data
-conditioned on the accuracy of the model.
-
-For our specific scenario, recall that our objective is to infer the
-relative fitness of a non-neutral lineage \(s^{(m)}\) along with
-nuisance parameters related to the population mean fitness at each
-point, \(\underline{\bar{s}}_t\), and the barcode frequency time series
-\(\underline{f}^{(m)}\). All these variables are related through our
-fitness model (see Section~\ref{sec-fitness_model} in the main text)
-\begin{equation}\protect\hypertarget{eq-ppc_model}{}{
-f_{t+1}^{(m)} = f_{t}^{(m)} \mathrm{e}^{(s^{(m)} - s_t)\tau}.
-}\label{eq-ppc_model}\end{equation} As we saw, it is convenient to
-rewrite Equation~\ref{eq-ppc_model} as
-\begin{equation}\protect\hypertarget{eq-ppc_logratio}{}{
-\frac{1}{\tau}\ln\frac{f_{t+1}^{(m)}}{f_{t}^{(m)}} = (s^{(m)} - s_t).
-}\label{eq-ppc_logratio}\end{equation} Written in this way, we separate
-the quantities we can compute from the experimental observations---the
-left-hand side of Equation~\ref{eq-ppc_logratio} can be computed from
-the barcode reads---from the latent variables.
-
-Although we perform the joint inference over all barcodes in the present
-work, let us focus on the inference task for a single barcode as if it
-were computed independently. For a non-neutral barcode, our task
-consists of computing the posterior probability
-\begin{equation}\protect\hypertarget{eq-ppc_posterior}{}{
-\pi(\theta \mid \underline{r}^{(m)}) =
-\pi(
-    s^{(m)}, \sigma^{(m)}, \underline{s}_t, \underline{f}^{(m)}
-    \mid \underline{r}^{(m)}
-),
-}\label{eq-ppc_posterior}\end{equation} where \(\theta\) represents all
-parameters to be inferred and \(\underline{r}^{(m)}\) is the vector with
-the barcode raw counts time series. The list of parameters are
-
-\begin{itemize}
-\tightlist
-\item
-  \(s^{(m)}\): The barcode's relative fitness.
-\item
-  \(\sigma^{(m)}\): A nuisance parameter used in the likelihood to
-  generate the data. This captures the expected deviation from
-  Equation~\ref{eq-ppc_logratio}
-\item
-  \(\underline{s}_t\): The vector with all population mean fitness for
-  each pair of adjacent time points.
-\item
-  \(\underline{f}^{(m)}\): The vector with the barcode frequency time
-  series.
-\end{itemize}
-
-Furthermore, let us define a naive estimate of the barcode frequency at
-time \(t\) as \begin{equation}\protect\hypertarget{eq-ppc_naive_freq}{}{
-\hat{f}_t^{(m)} = \frac{r^{(m)}_t}{\sum_{b=1}^B r^{(b)}_t}.
-}\label{eq-ppc_naive_freq}\end{equation} We can compute this quantity
-from the data by normalizing the raw barcode counts by the sum of all
-barcode counts. Furthermore, we can compute a naive estimate of the log
-frequency ratio from the raw barcode counts as
-\begin{equation}\protect\hypertarget{eq-ppc_naive_logfreq}{}{
-\ln \hat{\gamma}_t^{(m)} = \ln \frac{\hat{f}_{t+1}^{(m)}}{\hat{f}_t^{(m)}}
-}\label{eq-ppc_naive_logfreq}\end{equation}
-
-In our generative model, we assumed
-\begin{equation}\protect\hypertarget{eq-ppc_logfreq_likelihood}{}{
-\ln\gamma^{(m)}_t \mid \theta \sim \mathcal{N}(s^{(m)} - s_t, \sigma^{(m)}).
-}\label{eq-ppc_logfreq_likelihood}\end{equation} This implies that once
-we determine the posterior distribution of our parameters, we can
-generate synthetic values of \(\ln \gamma^{(m)}_t\) that we can then
-compare with the values obtained from applying
-Equation~\ref{eq-ppc_naive_logfreq} and
-Equation~\ref{eq-ppc_naive_logfreq} to the raw data.
-
-In practice, to compute the posterior predictive checks, we generate
-multiple samples from the posterior distribution
-\(\pi(\theta \mid \underline{r}^{(m)})\)
-\begin{equation}\protect\hypertarget{eq-ppc_theta_samples}{}{
-\underline{\theta} = (\theta_1, \theta_2, \ldots, \theta_N).
-}\label{eq-ppc_theta_samples}\end{equation} With these samples in hand,
-the \texttt{BarBay.jl} package includes the function
-\texttt{logfreq\_ratio\_bc\_ppc} for non-neutral barcodes that uses this
-set of posterior parameter samples to generate samples from the
-distribution defined in Equation~\ref{eq-ppc_logfreq_likelihood}. For a
-large-enough number of samples, we can then compute the desired
-percentiles---5, 68, and 95 percentiles in all figures in the main
-text---that are equivalent to the corresponding credible regions. In
-other words, the range of values of \(\ln\gamma_t^{(m)}\) generated by
-this bootstrap process can be used to compute the region where we expect
-to find our raw estimates \(\ln\hat{\gamma}_t^{(m)}\) with the desired
-probability. The package \texttt{BarBay.jl} includes an equivalent
-function, \texttt{logfreq\_ratio\_popmean\_ppc}, for neutral lineages.
-
-\hypertarget{sec-logistic}{%
-\subsection{Logistic growth simulation}\label{sec-logistic}}
-
-In this section, we explain the simulations used to assess the validity
-of our inference pipeline. Let us begin by assuming that, since the
-strains are grown for two full days in the experiment, having left
-behind the exponential phase for almost an entire day, a simple
-exponential growth of the form
-\begin{equation}\protect\hypertarget{eq-logistic_expo}{}{
-\frac{dn_i}{dt} = \lambda_i n_i,
-}\label{eq-logistic_expo}\end{equation} where \(n_i\) is the number of
-cells of strain \(i\), and \(\lambda_i\) is the corresponding growth
-rate is not enough. Instead, we will assume that the cells follow the
-logistic growth equation of the form
-\begin{equation}\protect\hypertarget{eq-logistic_fn}{}{
-\frac{dn_i}{dt} = \lambda_i n_i 
-\left( 1 - \frac{\sum_{j=1}^{N}n_i}{\kappa}\right),
-}\label{eq-logistic_fn}\end{equation} where \(\kappa\) is the carrying
-capacity, and \(N\) is the total number of strains in the culture.
-
-The inference method is based on the model that assumes that the time
-passed between dilutions \(\tau \approx 8\) generations, the change in
-frequency for a mutant barcode can be approximated from cycle \(t\) to
-the next cycle \(t + 1\) as
-\begin{equation}\protect\hypertarget{eq-logistic_fitness_fn}{}{
-f_{t+1}^{(m)} = f_t^{(m)} \mathrm{e}^{(s^{(m)} - \bar{s}_t)\tau},
-}\label{eq-logistic_fitness_fn}\end{equation} where \(s^{(m)}\) is the
-relative fitness for strain \(i\) compared to the ancestral strain and
-\(\bar{s}_t\) is the mean fitness of the population at cycle \(t\). To
-test this assumption, we implemented a numerical experiment following
-the logistic growth model described in Equation~\ref{eq-logistic_fn}.
-Figure~\ref{fig-SIX_logistic_growth_01} shows an example of the
-deterministic trajectories for 50 labeled neutral lineages and 1000
-lineages of interest. The upper red curve that dominates the culture
-represents the unlabeled ancestral strain included in the experimental
-design described in Section~\ref{sec-experiment}.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/figSIX_logistic_growth_01.pdf}
-
-}
-
-\caption{\label{fig-SIX_logistic_growth_01}\textbf{Logistic growth
-simulation over single growth cycle}. The dashed line represents the
-neutral lineages, with the upper curve being the unlabeled neutral
-strain. Color curves represent the genotypes of interest colored by
-growth rate relative to the neutral lineage.}
-
-\end{figure}
-
-To simulate multiple growth-dilution cycles, we take the population
-composition at the final time point and use it to initialize a new
-logistic growth simulation. Figure~\ref{fig-SIX_logistic_growth_02}
-shows the resulting number of cells at the last time point of a cycle
-over multiple growth-dilution cycles for the genotypes in
-@Figure~\ref{fig-SIX_logistic_growth_01}. We can see that the adaptive
-lineages (blue curves) increase in abundance, while detrimental lineages
-(red curves) decrease.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/figSIX_logistic_growth_02.pdf}
-
-}
-
-\caption{\label{fig-SIX_logistic_growth_02}\textbf{Growth-dilution
-cycles for logistic growth simulation}. Each point represents the final
-number of cells after a growth cycle for each lineage. Colors are the
-same as in Figure~\ref{fig-SIX_logistic_growth_01}.}
-
-\end{figure}
-
-In Section~\ref{sec-fitness_model}, we derive the functional form to
-infer the relative fitness of each lineage as
-\begin{equation}\protect\hypertarget{eq-logistic_logfreq}{}{
-\frac{1}{\tau}\ln \frac{f_{t+1}^{(b)}}{f_{t}^{(b)}} = (s^{(b)} - \bar{s}_t).
-}\label{eq-logistic_logfreq}\end{equation}
-Figure~\ref{fig-SIX_logistic_growth_03} shows the corresponding log
-frequency ratio curves for the logistic growth simulation. The
-displacement of these curves with respect to the neutral lineages
-determines the ground truth relative fitness value for these
-simulations.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/figSIX_logistic_growth_03.pdf}
-
-}
-
-\caption{\label{fig-SIX_logistic_growth_03}\textbf{Log frequency ratio
-for logistic growth simulations}. The relative distance of the color
-curves from the black dashed line determines the relative fitness of
-each linage.}
-
-\end{figure}
-
-To simulate the experimental noise, we add two types of noise:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Poisson noise between dilutions. For this, we take the final point of
-  the logistic growth simulation and sample a random Poisson number
-  based on this last point to set the initial condition for the next
-  cycle.
-\item
-  Gaussian noise when performing the measurements. When translating the
-  underlying population composition to the number of reads, we can add a
-  custom amount of Gaussian noise.
-\end{enumerate}
-
-Figure~\ref{fig-SIX_logistic_growth_04} shows the frequency trajectories
-(left panels) and log frequency ratios (right panels) for a noiseless
-simulation (upper panels) and a simulation with added noise (lower
-panels). The noiseless simulation is used to determine the relative
-fitness for each of the lineages, which serves as the ground truth to be
-compared with the resulting inference.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/figSIX_logistic_growth_04.pdf}
-
-}
-
-\caption{\label{fig-SIX_logistic_growth_04}\textbf{Logistic
-growth-dilution simulations with and without noise}}
-
-\end{figure}
-
-\printbibliography
-
-\end{document}
+We would like to thank Griffin Chure and Michael Betancourt for their helpful
+advice and discussion. We would like to thank Karna Gowda, Spencer Farrell, and
+Shaili Mathur for critical observations on the manuscript. MRM was supported by
+the Schmidt Science Fellowship. MM was supported by The National Science
+Foundation-Simons Center for Quantitative Biology at Northwestern University and
+the Simons Foundation grant 597491. MM is a Simons Investigator.

### 2023-10-11 doc/paper.qmd
latest version of paper

# Abstract

{{< include _abstract.qmd >}}

# Introduction

{{< include _intro.qmd >}}

# Results

<!-- Experimental results -->
{{< include _experiment.qmd >}}

<!-- Fitness model -->
{{< include _fitness_model.qmd >}}

<!-- Bayesian inference -->
{{< include _bayesian_inference.qmd >}}

<!-- Single dataset -->
{{< include _single_results.qmd >}}

<!-- Multi-environment results -->
{{< include _multienv_results.qmd >}}

<!-- Hierarchical model for experimental replicates -->
{{< include _hierarchical_replicates_results.qmd >}}

<!-- Hierarchical model for genotypes-->
{{< include _hierarchical_genotypes_results.qmd >}}

# Discussion

{{< include _discussion.qmd >}}

# Acknowledgements

{{< include _acknowledgements.qmd >}}

# Supplementary Materials

<!-- Primer on Variational Bayes -->
{{< include _variational_inference_primer.qmd >}}

<!-- Bayesian inference pipeline -->
{{< include _bayes_freq.qmd >}}

{{< include _bayes_meanfit.qmd >}}

{{< include _bayes_mutfit.qmd >}}

<!-- Hierarchical models -->
{{< include _bayes_hierarchical.qmd >}}

<!-- Prior selection -->
{{< include _bayes_priors.qmd >}}

<!-- Posterior predictive checks -->
{{< include _ppc.qmd >}}

<!-- Logistic growth simulations -->
{{< include _logistic_sim.qmd >}}

### 2023-10-11 doc/_acknowledgements.qmd
update acknowledgements and affiliations
finish adding all funding sources

@@
 advice and discussion. We would like to thank Karna Gowda, Spencer Farrell, and
-Shaili Mathur for critical observations on the manuscript. MRM was supported by
-the Schmidt Science Fellowship. MM was supported by The National Science
-Foundation-Simons Center for Quantitative Biology at Northwestern University and
-the Simons Foundation grant 597491. MM is a Simons Investigator.
+Shaili Mathur for critical observations on the manuscript. This work was
+supported by the NIH/NIGMS, Genomics of rapid adaptation in the lab and in the
+wild R35GM11816506 (MIRA grant), the NIH, Unravelling mechanisms of tumor
+suppression in lung cancer R01CA23434903, the NIH (PQ4), Quantitative and
+multiplexed analysis of gene function in cancer in vivo R01CA23125303, the NIH,
+Genetic Determinants of Tumor Growth and Drug Sensitivity in EGFR Mutant Lung
+Cancer R01CA263715, the NIH, Dissecting the interplay between aging, genotype,
+and the microenvironment in lung cancer U01AG077922, the NIH, Genetic dissection
+of oncogenic Kras signaling R01CA230025, and the CZ Biohub investigator program.
+MRM was supported by the Schmidt Science Fellowship. MM was supported by The
+National Science Foundation-Simons Center for Quantitative Biology at
+Northwestern University and the Simons Foundation grant 597491. MM is a Simons
+Investigator. DP is a CZ Biohub investigator.

### 2023-10-11 doc/paper.tex
change corresponding author
update acknowledgements and affiliations
finish adding all funding sources

@@
 % Authors
-\author[1]{Manuel Razo-Mejia}
+\author[1,*]{Manuel Razo-Mejia}
 \author[2,3]{Madhav Mani}
-\author[1,4,*]{Dmitri Petrov}
+\author[1,4,5]{Dmitri Petrov}
 
@@
 \affil[4]{Stanford Cancer Institute, Stanford University School of Medicine, Stanford, CA, USA}
-\affil[*]{Correspondence: dpetrov@stanford.edu}
+\affil[5]{Chan Zuckerberg Biohub}
+\affil[*]{Correspondence: mrazo@stanford.edu}
 
@@
 advice and discussion. We would like to thank Karna Gowda, Spencer Farrell, and
-Shaili Mathur for critical observations on the manuscript. MRM was supported by
-the Schmidt Science Fellowship. MM was supported by The National Science
-Foundation-Simons Center for Quantitative Biology at Northwestern University and
-the Simons Foundation grant 597491. MM is a Simons Investigator.
+Shaili Mathur for critical observations on the manuscript. This work was
+supported by the NIH/NIGMS, Genomics of rapid adaptation in the lab and in the
+wild R35GM11816506 (MIRA grant), the NIH, Unravelling mechanisms of tumor
+suppression in lung cancer R01CA23434903, the NIH (PQ4), Quantitative and
+multiplexed analysis of gene function in cancer in vivo R01CA23125303, the NIH,
+Genetic Determinants of Tumor Growth and Drug Sensitivity in EGFR Mutant Lung
+Cancer R01CA263715, the NIH, Dissecting the interplay between aging, genotype,
+and the microenvironment in lung cancer U01AG077922, the NIH, Genetic dissection
+of oncogenic Kras signaling R01CA230025, and the CZ Biohub investigator program.
+MRM was supported by the Schmidt Science Fellowship. MM was supported by The
+National Science Foundation-Simons Center for Quantitative Biology at
+Northwestern University and the Simons Foundation grant 597491. MM is a Simons
+Investigator. DP is a CZ Biohub investigator.
 

### 2023-10-13 doc/paper.tex
prepare paper for submission

@@
 \documentclass[
-  letterpaper,
-  DIV=11,
-  numbers=noendperiod]{scrartcl}
+]{scrartcl}
 
@@
 
-\usepackage{lineno}
-\linenumbers
+% Change font size
 \usepackage[font={footnotesize, sf}, labelfont=bf, margin=0.05\linewidth]{caption}
@@
 \usepackage{authblk}
-
-\KOMAoption{captions}{tableheading}
 \makeatletter
@@
 
-% References
-% \usepackage[]{biblatex}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \usepackage[
-	backend=bibtex,
     style=phys,
@@
 ]{biblatex}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \addbibresource{references.bib}
-
 \IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
@@
 \hypersetup{
+  pdftitle={Bayesian inference of relative fitness on high-throughput pooled competition assays},
   pdfauthor={Manuel Razo-Mejia; Madhav Mani; Dmitri Petrov},
@@
 
-% Title
-\title{Bayesian inference of relative fitness on high-throughput pooled competition assays}
-
-% Authors
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Add title
+\title{Bayesian inference of relative fitness on high-throughput pooled
+competition assays}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Add list of authors
+
+% Loop through authors
 \author[1,*]{Manuel Razo-Mejia}
-\author[2,3]{Madhav Mani}
-\author[1,4,5]{Dmitri Petrov}
-
-% Affiliations
-\affil[1]{Department of Biology, Stanford University, Stanford, CA, USA}
-\affil[2]{Department of Engineering Sciences and Applied Mathematics, Northwestern University, Chicago, IL, USA}
-\affil[3]{NSF-Simons Center for Quantitative Biology, Northwestern University, Chicago, IL, USA}
-\affil[4]{Stanford Cancer Institute, Stanford University School of Medicine, Stanford, CA, USA}
+\author[3,4]{Madhav Mani}
+\author[1,2,5]{Dmitri Petrov}
+
+% Loop through affiliations
+\affil[1]{Department of Biology, Stanford University}
+\affil[2]{Stanford Cancer Institute, Stanford University School of
+Medicine}
+\affil[3]{NSF-Simons Center for Quantitative Biology, Northwestern
+University}
+\affil[4]{Department of Molecular Biosciences, Northwestern University}
 \affil[5]{Chan Zuckerberg Biohub}
-\affil[*]{Correspondence: mrazo@stanford.edu}
-
-\setcounter{Maxaffil}{0}
+
+% Add corresponding author
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Set affiliations in small font
 \renewcommand\Affilfont{\itshape\small}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Remove date
 \date{}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 \begin{document}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden, interior hidden, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi
-
+\maketitle
+\begin{abstract}
+The tracking of lineage frequencies via DNA barcode sequencing enables
+the quantification of microbial fitness. However, experimental noise
+coming from biotic and abiotic sources complicates the computation of a
+reliable inference. We present a Bayesian pipeline to infer relative
+microbial fitness from high-throughput lineage tracking assays. Our
+model accounts for multiple sources of noise and propagates
+uncertainties throughout all parameters in a systematic way.
+Furthermore, using modern variational inference methods based on
+automatic differentiation, we are able to scale the inference to a large
+number of unique barcodes. We extend this core model to analyze
+multi-environment assays, replicate experiments, and barcodes linked to
+genotypes. On simulations, our method recovers known parameters within
+posterior credible intervals. This work provides a generalizable
+Bayesian framework to analyze lineage tracking experiments. The
+accompanying open-source software library enables the adoption of
+principled statistical methods in experimental evolution.
+\end{abstract}
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden]}{\end{tcolorbox}}\fi
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Remove main text from the table of contents by specifying not to include
@@
 \defbibfilter{notother}{not segment=\therefsegment}
-
-\maketitle
-
-\begin{abstract}
-    The tracking of lineage frequencies via DNA barcode sequencing enables the
-  quantification of microbial fitness. However, experimental noise coming from
-  biotic and abiotic sources complicates the computation of a reliable
-  inference. We present a Bayesian pipeline to infer relative microbial fitness
-  from high-throughput lineage tracking assays. Our model accounts for multiple
-  sources of noise and propagates uncertainties throughout all parameters in a
-  systematic way. Furthermore, using modern variational inference methods based
-  on automatic differentiation, we are able to scale the inference to a large
-  number of unique barcodes. We extend this core model to analyze
-  multi-environment assays, replicate experiments, and barcodes linked to
-  genotypes. On simulations, our method recovers known parameters within
-  posterior credible intervals. This work provides a generalizable Bayesian
-  framework to analyze lineage tracking experiments. The accompanying
-  open-source software library enables the adoption of principled statistical
-  methods in experimental evolution.
-\end{abstract}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-We would like to thank Griffin Chure and Michael Betancourt for their helpful
-advice and discussion. We would like to thank Karna Gowda, Spencer Farrell, and
-Shaili Mathur for critical observations on the manuscript. This work was
-supported by the NIH/NIGMS, Genomics of rapid adaptation in the lab and in the
-wild R35GM11816506 (MIRA grant), the NIH, Unravelling mechanisms of tumor
-suppression in lung cancer R01CA23434903, the NIH (PQ4), Quantitative and
-multiplexed analysis of gene function in cancer in vivo R01CA23125303, the NIH,
-Genetic Determinants of Tumor Growth and Drug Sensitivity in EGFR Mutant Lung
-Cancer R01CA263715, the NIH, Dissecting the interplay between aging, genotype,
-and the microenvironment in lung cancer U01AG077922, the NIH, Genetic dissection
-of oncogenic Kras signaling R01CA230025, and the CZ Biohub investigator program.
-MRM was supported by the Schmidt Science Fellowship. MM was supported by The
-National Science Foundation-Simons Center for Quantitative Biology at
-Northwestern University and the Simons Foundation grant 597491. MM is a Simons
-Investigator. DP is a CZ Biohub investigator.
-
+We would like to thank Griffin Chure and Michael Betancourt for their
+helpful advice and discussion. We would like to thank Karna Gowda,
+Spencer Farrell, and Shaili Mathur for critical observations on the
+manuscript. This work was supported by the NIH/NIGMS, Genomics of rapid
+adaptation in the lab and in the wild R35GM11816506 (MIRA grant), the
+NIH, Unravelling mechanisms of tumor suppression in lung cancer
+R01CA23434903, the NIH (PQ4), Quantitative and multiplexed analysis of
+gene function in cancer in vivo R01CA23125303, the NIH, Genetic
+Determinants of Tumor Growth and Drug Sensitivity in EGFR Mutant Lung
+Cancer R01CA263715, the NIH, Dissecting the interplay between aging,
+genotype, and the microenvironment in lung cancer U01AG077922, the NIH,
+Genetic dissection of oncogenic Kras signaling R01CA230025, and the CZ
+Biohub investigator program. MRM was supported by the Schmidt Science
+Fellowship. MM was supported by The National Science Foundation-Simons
+Center for Quantitative Biology at Northwestern University and the
+Simons Foundation grant 597491. MM is a Simons Investigator. DP is a CZ
+Biohub investigator.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Print main text references
@@
 \clearpage
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % SUPPLEMENTAL MATERIAL
 
-% Indicate that now all sections and subsections should be included in the
+% Indicate that all sections and subsections should be included in the
 % table of contents so that only the SI is included.
@@
 
-% Define reference section for the supplemental material
+% Define the reference section for the supplemental material
 \begin{refsegment}
-% Set equation, table and figure counters to begin with "S"
+% Set equation, table, and figure counters to begin with "S"
 \beginsupplement
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Add table of contents
 \tableofcontents
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% HERE IS WHERE THE SI TEXT GOES
 % Print supplemental references changing the title
@@
 \end{refsegment}
-
-% \printbibliography
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2023-11-20 doc/_bayes_hierarchical.qmd
remove repeated label

@@
 ).
-$${#eq-hier_bayes_exp_indep}
+$${#eq-hier_bayes_exp_indep_02}
 Notice we removed the conditioning on the population mean fitness as our prior
@@
 
-The first term on the right-hand side of @eq-hier_bayes_exp_indep takes the same
-functional form as the one derived in @sec-bayes_mutfit. Therefore, all we are
-left with is to determine the functional forms for the hyper–prior
+The first term on the right-hand side of @eq-hier_bayes_exp_indep_02 takes the
+same functional form as the one derived in @sec-bayes_mutfit. Therefore, all we
+are left with is to determine the functional forms for the hyper–prior
 $\pi(\underline{\theta}^M)$, and the conditional probability

### 2023-11-20 doc/_variational_inference_primer.qmd
remove repeated label

@@
 \right),
-$${#eq-vi_objective}
+$${#eq-vi_objective_si}
 where $D_{KL}$ is the KL divergence. Furthermore, we highlight that the KL
@@
 
-At first sight, @eq-vi_objective does not improve the situation but only
+At first sight, @eq-vi_objective_si does not improve the situation but only
 introduces further technical complications. After all, the definition of the KL

### 2023-12-04 doc/_fitseq2_sim.qmd
first draft of section

---
editor:
    render-on-save: true
bibliography: references.bib
csl: ieee.csl
---

## Comparison with `FitSeq2.0` {#sec-fitseq2}

For a complete comparison, we apply our inference pipeline to a simulated
dataset closely following the simulation pipeline from @li2023. This simulation
explicitly accounts for multiple noise sources captured in the added Gaussian
noise in @sec-logistic. Here, we list all the steps involved in this explicit
simulation, highlighting that the overall results are qualitatively very similar
to that of our simple logistic growth plus added Gaussian noise when computing
the relevant quantity of the log frequency ratio.

1. We initialized a population consisting of 1000 unique barcodes---50 neutral
barcodes and 950 mutant barcodes---where the initial number of cells with
barcode $i$ in the culture was sampled from a Gamma distribution as
$$
n_o^{(i)} \sim \text{Gamma}(20, 0.2)
$${#eq-fitseq_gamma_pop}

2. The Malthusian fitness (related but not exactly the relative fitness to be
inferred from the data) for non-neutral barcodes was then sampled from a skewed
normal distribution of the form
$$
\lambda^{(j)} \sim \text{Skew}\mathcal{N}(0, 0.225, 3)
$${#eq-fitseq_skewnormal}

3. To simulate the noisy growth, the number of cells underwent the following
steps:

**for** $t=2:T$ **do**

&nbsp;&nbsp;&nbsp;&nbsp; 
$n_{t,c}^{(i)} = \text{Poisson}(n_{t-1,c}^{(i)} e^{\lambda^{(i)}})$

**end for**,

where $T$ is the number of generations, and $c$ indexes the growth-dilution
cycle. We use $T=4$ for our simulation.

4. After the $T$ generations, cells were stochastically diluted back for the
next growth-dilution cycle as
$$
n_{0,c+1}^{(i)} \sim \text{Poisson}
\left(\frac{n_{T,c}^{(i)}}{2^{T}}\right).
$${#eq-fitseq_dilution}
This process was repeated $C=4$ times for a total of four growth-dilution
cycles.

5. To simulate the sampling noise when taking cells that will then be
subsequently processed to obtain barcode reads, we sampled
$n_{\text{sample},c}^{(i)}$ cells with barcode $i$ as
$$
n_{\text{sample},c}^{(i)} \sim 
\text{Poisson}(L \times N \times f_{c}^{(i)}),
$${#eq-fitseq_sample}
where $c$ indexes the growth cycle, $N$ is the total number of unique barcodes,
$f_{j}^{(i)}$ is the fraction of cells in the culture with barcode $i$ at the
end of cycle $j$, i.e.,
$$
f_{c}^{(i)} = \frac{n_{c}^{(i)}}
{\sum_{j=1}^N n_{c}^{(j)}},
$${#eq-fitseq_frac_cell}
and $L$ is the expected average number of cells per lineage. We used $L = 500$.

6. To simulate the noise due to PCR amplification, we iterate $c_{\text{PCR}} =
25$ times, stochastically doubling the number of barcodes with a Poisson sample
of the form

**for** $a=1:A$ **do**

&nbsp;&nbsp;&nbsp;&nbsp; 
$n_{\text{PCR},(a+1),c}^{(i)} \sim \text{Poisson}(2 n_{\text{PCR},(a),c}^{(i)})$

**end for**,

where $(a)$ indexes the amplification cycle, $A$ is the total number of cycles,
and $c$ indexes the growth-dilution cycle. We use $A=25$ for our simulation.For
$a = 1$, we use the number of cells sampled in @eq-fitseq_sample.

7. To simulate the noise due to the sampling step involved during the sequencing
of the amplicon library, the number of reads mapping to barcode $i$ at the end
of cycle $c$ used in the inference was sampled as 
$$
n_{\text{seq},c}^{(i)} \sim 
\text{Poisson}(R \times N \times f_{\text{PCR},c}^{(i)}),
$${#eq-fitseq_seq}
where $R$ is the expected number of reads per barcode (R=100 for our
simulation), $N$ is the number of total barcodes, and $f_{\text{PCR},c}^{(i)}$
is the fraction of barcodes of type $i$ in the final amplification cycle from
step 4.

@fig-SIX_fitseq2(A) and (B) show the simulation's frequency and log frequency
ratio trajectories, respectively. Qualitatively, these results are very similar
to those shown in Fig. 1 of @li2023. As done throughout the main text,
@fig-SIX_fitseq2(C) shows examples of the posterior predictive checks for the
neutral lineages and a few of the mutant lineages. We can see that our inference
pipeline can fit this simulated data for the examples shown. Furthermore,
@fig-SIX_fitseq2(D) compares the ground truth fitness values and the inferred
parameters for all mutant lineages. @fig-SIX_fitseq2 shows the comparison
between the inference with the software provided in @li2023 and our inference
pipeline. We note that although the values are highly correlated, the scale of
the axis is very different.

![**FitSeq2.0 simulation comparison**.(A) Frequency trajectories that represent
the raw data going into the inference. (B) Log frequency ratio between two
adjacent time points used by the inference pipeline. Darker lines represent the
neutral barcodes. These transformed data are much more noisy than the seemingly
smooth frequency trajectories. (C) Examples of the posterior predictive checks
for all neutral lineages (upper left panel) and a subset of representative
mutant lineages. Shaded regions represent the 95\%, 68\%, and 5\% credible
regions for the data. The reported errors above the plot represent the 68\%
credible region on the mutant relative fitness marginal distribution. (D)
Comparison between the ground truth fitness value from the logistic-growth
simulation and the inferred fitness value. Gray error bars represent the 68\%
posterior credible region for the relative fitness values. (E) Comparison
between fitness inferences as obtained using `FitSeq2.0` from @li2023 and our
Bayesian method. Vertical gray error bars represent the 68\% posterior credible
region for the relative fitness
values.](./figs/figSIX_fitseq2){#fig-SIX_fitseq2}

### 2023-12-05 doc/_logistic_sim.qmd
small fixes
add simulation information

@@
 
-## Logistic growth simulation
+## Logistic growth simulation {#sec-logistic}
 
@@
 noise**](./figs/figSIX_logistic_growth_04){#fig-SIX_logistic_growth_04}
+
+### Simulation details
+
+This section lists the parameters used to generate the synthetic data used
+throughout the main text. All the code to reproduce every simulation and figure
+in the paper can be found in [the GitHub
+repository](https://github.com/mrazomej/bayesian_fitness/tree/master). Unless
+stated otherwise, the simulations were initialized following these steps:
+
+1. The initial number of cells were defined as:
+- Ancestor lineage: `n_init` $\times$ `frac_anc`.
+- Neutral lineages: $n_0^{(n)} \sim \text{Poisson} 
+\left( 
+    \texttt{n\_init} \times \frac{\texttt{frac\_neutral}}{\texttt{n\_neutral}} 
+\right)$ 
+- Mutant lineages: $n_0^{(m)} \sim \text{LogNormal} 
+\left( 
+    \mu = \log(\texttt{n\_init} \times \frac{\texttt{frac\_mut}}{\texttt{n\_mut}}),
+    \sigma = 2 
+\right)$
+
+where
+
+- `n_init` is the initial number of cells.
+- `frac_anc` $\in [0, 1]$ is the initial fraction of the culture that is
+ancestor.
+- `frac_neutral` $\in [0, 1]$ is the initial fraction of the culture that is
+barcoded neutral lineages.
+- `frac_mut` $\in [0, 1]$ is the initial fraction of the culture that is
+barcoded lineages whose relative fitness we want to infer--hereafter "mutat"
+barcodes.
+- `frac_anc` + `frac_neutral` + `frac_mut` = 1
+- `n_neutral` is the number of unique neutral lineage barcodes.
+- `n_mut` is the number of unique mutant barcodes.
+
+2. The distribution of growth rates $\lambda$ were defined as
+- Ancestor lineage \& neutral lineages: $\lambda_a =$ `λ_a`. See details below
+for specific values for each simulation.
+- Mutant lineages: $\lambda_m \sim 
+\text{truncated}-\mathcal{N}(
+    \mu =$ `λ_mut_mean`,
+    $\sigma =$ `λ_bc_std`,
+    $\text{truncation limits} =$ `λ_trunc`
+$)$. See details below for specific values for each simulation.
+
+#### Fig 2.
+
+The following is a `README.md` file automatically generated when simulating the
+data for Fig. 2 in the main text:
+
+- File location
+`https://github.com/mrazomej/bayesian_fitness/code/processing/mcmc_logistic_data001/sim_data001.jl`
+- Number of mutant barcodes
+`n_mut = 225`
+- Number of neutral barcodes
+`n_neutral = 25`
+- Ancestral strain growth rate
+`λ_a = 1.0`
+- Carrying capacity
+`κ = 1.0e10`
+- Number of generations
+`n_gen = 8`
+- Initial number of cells
+`n_init = κ / (2^(n_gen)) = 3.90625e7`
+- Initial fraction of culture that is ancestor
+`frac_anc = 0.93`
+- Initial fraction of culture that is neutrals
+`frac_neutral = 0.02`
+- Initial fraction of culture that is mutants
+`frac_mut = 0.04999999999999995`
+- Mutant fitness distribution mean
+`λ_mut_mean = λ_a * 1.005 = 1.005`
+- Mutant fitness distribution standard deviation to sample growth rates
+`λ_mut_std = 0.1`
+- Mutant fitness distribution truncation ranges for growth rates
+`λ_trunc = [λ_a .* 0.9999, λ_a * 1.5] = [0.9999, 1.5]`
+- Gaussian noise distribution standard deviation
+`σ_lognormal = 0.3`
+
+#### Fig 3.
+
+The following is a `README.md` file automatically generated when simulating the
+data for Fig. 3 in the main text:
+
+- File location
+`https://github.com/mrazomej/bayesian_fitness/code/processing/data003_logistic_1000bc_03env_01rep/sim_multienv_data003.jl`
+- Number of environments
+`n_env = 3`
+- Environment transitions
+`env_idx = [1, 2, 3, 1, 2, 3]`
+- Number of mutant barcodes
+`n_mut = 900`
+- Number of neutral barcodes
+`n_neutral = 100`
+- Ancestral strain growth rate on each environment
+`λ_a = [1.0, 2.0, 0.5]`
+- Carrying capacity
+`κ = 1.0e10`
+- Number of generations
+`n_gen = 8`
+- Initial number of cells
+`n_init = κ / (2^(n_gen)) = 3.90625e7`
+- Initial fraction of culture that is ancestor
+`frac_anc = 0.93`
+- Initial fraction of culture that is neutrals
+`frac_neutral = 0.02`
+- Initial fraction of culture that is mutants
+`frac_mut = 0.04999999999999995`
+- Mutant fitness distribution mean
+`λ_mut_mean = λ_a * 1.005 = [1.005, 2.01, 0.5025]
+- Mutant fitness distribution standard deviation to sample growth rates
+`λ_mut_std = 0.1`
+- Mutant fitness distribution truncation ranges for growth rates
+`λ_trunc = [λ_a .* 0.9995, λ_a .* 1.5] = [[[0.995, 1.99, 0.4975]], [[1.5, 3.0, 0.75]]]`
+- Gaussian noise distribution standard deviation
+`σ_lognormal = 0.125`
+
+The variable `env_idx` lists the order of the environment cycles used in the
+simulation. In this case, transitions were of the form `1 -> 2`, `2 -> 3`, and
+`3 -> 1`.
+
+#### Fig 4.
+
+The following is a `README.md` file automatically generated when simulating the
+data for Fig. 4 in the main text:
+
+- File location
+`https://github.com/mrazomej/bayesian_fitness/code/processing/data002_logistic_1000bc_01env_02rep/sim_hierarchical_data002.jl`
+- Number of experimental replicates
+`n_data = 2`
+- Number of mutant barcodes
+`n_mut = 900`
+- Number of neutral barcodes
+`n_neutral = 100`
+- Ancestral strain growth rate
+`λ_a = 1.0`
+- Carrying capacity
+`κ = 1.0e10`
+- Number of generations
+`n_gen = 8`
+- Initial number of cells
+`n_init = κ / (2^(n_gen)) = 3.90625e7`
+- Initial fraction of culture that is ancestor
+`frac_anc = 0.93`
+- Initial fraction of culture that is neutrals
+`frac_neutral = 0.02`
+- Initial fraction of culture that is mutants
+`frac_mut = 0.04999999999999995`
+- Mutant fitness distribution mean
+`λ_bc_mean = λ_a * 1.005 = 1.005`
+- Mutant fitness distribution standard deviation to sample growth rates
+`λ_bc_std = 0.1`
+- Mutant fitness distribution truncation ranges for growth rates
+`λ_trunc = [λ_a .* 0.9999, λ_a * 1.5] = [0.9999, 1.5]`
+- Gaussian noise distribution standard deviation
+`σ_lognormal = 0.3`
+- Standard deviation for between-experiment fitness distribution variability
+`σ_exp = 0.015`
+
+The variable `σ_exp` defines the variance used to sample the variability between
+simulated experimental replicates. In other words, given a "hyper-fitness" value
+$\lambda$, for each experimental replicate, the growth rate was defined as
+$\lambda_j \sim \mathcal{N}(\lambda,$ `σ_exp` $)$.
+
+#### Fig 6.
+
+The following is a `README.md` file automatically generated when simulating the
+data for Fig. 6 in the main text:
+
+- File location
+`https://github.com/mrazomej/bayesian_fitness/code/processing/data007_logistic_hierarchicalgenotype_1200bc_01env_01rep/sim_data007.jl`
+- Number of mutant barcodes
+`n_mut = 900`
+- Number of neutral barcodes
+`n_neutral = 100`
+- Ancestral strain growth rate
+`λ_a = 1.0`
+- Carrying capacity
+`κ = 1.0e10`
+- Number of generations
+`n_gen = 8`
+- Initial number of cells
+`n_init = κ / (2^(n_gen)) = 3.90625e7`
+- Initial fraction of culture that is ancestor
+`frac_anc = 0.93`
+- Initial fraction of culture that is neutrals
+`frac_neutral = 0.02`
+- Initial fraction of culture that is mutants
+`frac_mut = 0.04999999999999995`
+- Mutant fitness distribution mean
+`λ_bc_mean = λ_a * 1.005 = 1.005`
+- Mutant fitness distribution standard deviation to sample growth rates
+`λ_bc_std = 0.1`
+- Mutant fitness distribution truncation ranges for growth rates
+`λ_trunc = [λ_a .* 0.9999, λ_a * 1.5] = [0.9999, 1.5]`
+- Gaussian noise distribution standard deviation
+`σ_lognormal = 0.3`
+- Number of unique genotypes
+`n_geno = Int64(n_mut / 10) = 90`
+
+The variable `n_geno` defines the number of unique genotypes in the simulation.
+In this simulation, we aimed to have ten unique barcodes on average for each
+genotype. The actual number of genotypes per barcode was sampled from a
+multinomial distribution with probability `1 / n_geno` for each genotype.

### 2023-12-05 doc/_hierarchical_replicates_results.qmd
add text color based on reviews

@@
 hand, we can pool all data together and treat our different experiments as a
-single measurement with higher coverage. This loses the subtle differences due
-to biotic and abiotic batch effects, effectively halving the data that goes into
-our inference problem.
+single measurement with higher coverage. 
+<span style="color: #912E27;">
+However, by doing so, we lose all information about experiment-to-experiment
+variability due to intrinsic biological variability and environmental
+fluctuations---such as small temperature fluctuations---that we cannot control.
+In this sense, combining the datasets defeats the purpose of performing multiple
+measurements to account for this variability when extracting the relevant
+information from the observations.
+</span>
 

### 2023-12-05 code/figs/fig06.jl
update ECDF plot to match reviewer's suggestion

@@ -361,12 +361,30 @@ ecdfplot!(
     linewidth=2.5,
 )
 
+# Add line to complete ECDF
+lines!(
+    ax,
+    [maximum(abs.(df_fitness.mean_h .- df_fitness.fitness)), 1.5],
+    [1, 1],
+    color=ColorSchemes.seaborn_colorblind[1],
+    linewidth=2.5,
+)
+
 # Plot single-barcode inference ECDF
 ecdfplot!(
     ax,
     abs.(df_fitness_single.mean .- df_fitness_single.fitness),
     color=ColorSchemes.seaborn_colorblind[2],
     label="single-barcode",
+    linewidth=2.5,
+)
+
+# Add line to complete ECDF
+lines!(
+    ax,
+    [maximum(abs.(df_fitness_single.mean .- df_fitness_single.fitness)), 1.5],
+    [1, 1],
+    color=ColorSchemes.seaborn_colorblind[2],
     linewidth=2.5,
 )
 
@@ -378,6 +396,9 @@ ecdfplot!(
     label="pooled-barcodes",
     linewidth=2.5,
 )
+
+# Defie ax limits
+xlims!(ax, [0, 1.1])
 
 axislegend(ax, position=:rb, framevisible=false)
 
@@ -424,7 +445,7 @@ for row in 1:n_row
 
             # Compute posterior predictive checks
             local ppc_mat = BarBay.stats.logfreq_ratio_popmean_ppc(
-                df_samples, n_ppc; model=:normal, param=param
+                df_samples, n_ppc; param=param
             )
 
             # Define time
@@ -475,7 +496,7 @@ for row in 1:n_row
         )
         # Compute posterior predictive checks
         local ppc_mat = BarBay.stats.logfreq_ratio_bc_ppc(
-            df_samples, n_ppc; model=:normal, param=param
+            df_samples, n_ppc; param=param
         )
         # Plot posterior predictive checks
         BayesFitUtils.viz.ppc_time_series!(

### 2023-12-05 doc/_experiment.qmd
add new section based on reviewer comment

@@
 rest of the culture is left for the ensemble of genotypes of interest.
+<span style="color: #912E27;">
+This experimental design in which the barcodes of interest represent a small
+fraction of the culture serves three main purposes: First, the model used to
+infer the relative fitness differences between genotypes is valid in the regime
+where the genotype frequency is significantly smaller than one (see
+@sec-fitness_model). Second, potential problems with frequency-dependent
+selection are minimized as long as the frequency of each genotype remains small.
+Third, having a single genotype dominating the culture standardizes the
+environment experienced by all genotypes. This is because variations in the
+chemistry of the environment are effectively dictated by the dominating
+genotype, allowing for reproducible growth-dilution cycles.
+</span>
 

### 2023-12-05 code/processing/data014_kinsler2020/advi_meanfield_hierarchicalreplicate_inference.jl
scripts to process part of Kinsler 2020 data

@@ -1,0 +1,166 @@
+##
+println("Loading packages...")
+
+# Load project package
+@load_pkg BayesFitUtils
+
+# Import project package
+import BayesFitUtils
+
+# Import library package
+import BarBay
+
+# Import libraries to manipulate data
+import DataFrames as DF
+import CSV
+
+# Import library to list files
+import Glob
+
+# Import library to perform Bayesian inference
+import Turing
+
+# Import AutoDiff backend
+using ReverseDiff
+
+# Import Memoization
+using Memoization
+
+# Impor statistical libraries
+import Random
+import StatsBase
+
+# Set AutoDiff backend
+Turing.setadbackend(:reversediff)
+# Allow system to generate cache to speed up computation
+Turing.setrdcache(true)
+
+Random.seed!(42)
+
+## -----------------------------------------------------------------------------
+
+# Define number of samples and steps
+n_samples = 1
+n_steps = 10_000
+
+## %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+# Generate output directories
+## %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+# Define output directory
+outdir = "./output/advi_meanfield_hierarchicalreplicate_inference"
+
+# Generate output directory 
+if !isdir("./output/")
+    mkdir("./output/")
+end # if
+
+if !isdir(outdir)
+    mkdir(outdir)
+end # if
+
+## %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+# Load the data
+## %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+println("Loading data...")
+
+# Import data
+df_counts = CSV.read(
+    "$(git_root())/data/kinsler_2020/tidy_counts_no_anc.csv", DF.DataFrame
+)
+
+# Read datasets visual evaluation info
+df_include = CSV.read(
+    "$(git_root())/data/kinsler_2020/exp_include.csv", DF.DataFrame
+)
+
[…]

### 2023-12-05 code/processing/data014_kinsler2020/advi_meanfield_inference.jl

@@ -1,0 +1,143 @@
+##
+println("Loading packages...")
+
+# Load project package
+@load_pkg BayesFitUtils
+
+# Import project package
+import BayesFitUtils
+
+# Import library package
+import BarBay
+
+# Import libraries to manipulate data
+import DataFrames as DF
+import CSV
+
+# Import library to list files
+import Glob
+
+# Import library to perform Bayesian inference
+import Turing
+
+# Import AutoDiff backend
+using ReverseDiff
+
+# Import Memoization
+using Memoization
+
+# Impor statistical libraries
+import Random
+import StatsBase
+
+# Set AutoDiff backend
+Turing.setadbackend(:reversediff)
+# Allow system to generate cache to speed up computation
+Turing.setrdcache(true)
+
+Random.seed!(42)
+
+## -----------------------------------------------------------------------------
+
+# Define number of samples and steps
+n_samples = 1
+n_steps = 4_500
+
+## %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+# Generate output directories
+## %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+# Define output directory
+outdir = "./output/advi_meanfield_joint_inference"
+
+# Generate output directory 
+if !isdir("./output/")
+    mkdir("./output/")
+end # if
+
+if !isdir(outdir)
+    mkdir(outdir)
+end # if
+
+## %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+# Load the data
+## %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+println("Loading data...")
+
+# Import data
+df = CSV.read(
+    "$(git_root())/data/kinsler_2020/tidy_counts_no_anc.csv", DF.DataFrame
+)
+
+# Read datasets visual evaluation info
+df_include = CSV.read(
+    "$(git_root())/data/kinsler_2020/exp_include.csv", DF.DataFrame
+)
+
+# Define datasets to process
[…]

### 2023-12-05 code/processing/data014_kinsler2020/viz_inference.jl

@@ -1,0 +1,839 @@
+##
+println("Loading packages...")
+
+# Load project package
+@load_pkg BayesFitUtils
+
+# Import project package
+import BayesFitUtils
+
+# Import library package
+import BarBay
+
+# Import basic math
+import LinearAlgebra
+import StatsBase
+import Distributions
+import Random
+
+# Import libraries to manipulate data
+import DataFrames as DF
+import CSV
+
+# Import library to list files
+import Glob
+
+# Import plotting libraries
+using CairoMakie
+import ColorSchemes
+
+# Activate backend
+CairoMakie.activate!()
+
+# Set PBoC Plotting style
+BayesFitUtils.viz.theme_makie!()
+
+Random.seed!(42)
+
+## %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+# Load the data
+## %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+println("Loading data...")
+
+# Import data
+df = CSV.read(
+    "$(git_root())/data/kinsler_2020/tidy_counts_no_anc.csv", DF.DataFrame
+)
+
+# Define datasets to process
+envs = ["1.7%", "1.8%", "2ugFlu", "8.5uMGdA"]
+
+# Define titles
+env_titles = ["1.7% glucose", "1.8% glucose", "2 µg Flu", "8.5 µM Gda"]
+
+# Keep data from environments to process
+df = df[[x ∈ envs for x in df.env], :]
+
+## %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+# Plot environment trajectories
+## %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+# Group data by environment
+df_group = DF.groupby(df, :env)
+
+# Initialize figure
+fig = Figure(resolution=(300 * 4, 350 * 2))
+
+# An array to store the GridLayout objects
+gls = []
+
+# Generating four GridLayout objects
+for i in 1:2, j in 1:2
+    # Initialize grid layout
+    gl = GridLayout()
[…]

### 2023-12-05 code/figs/figSIX_kinsler2020.jl
script to generate SI figures for Kinsler data analysis

@@ -1,0 +1,841 @@
+##
+println("Loading packages...")
+
+# Load project package
+@load_pkg BayesFitUtils
+
+# Import project package
+import BayesFitUtils
+
+# Import library package
+import BarBay
+
+# Import basic math
+import LinearAlgebra
+import StatsBase
+import Distributions
+import Random
+
+# Import libraries to manipulate data
+import DataFrames as DF
+import CSV
+
+# Import library to list files
+import Glob
+
+# Import plotting libraries
+using CairoMakie
+import ColorSchemes
+
+# Activate backend
+CairoMakie.activate!()
+
+# Set PBoC Plotting style
+BayesFitUtils.viz.theme_makie!()
+
+Random.seed!(42)
+
+## %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+# Load the data
+## %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+println("Loading data...")
+
+# Import data
+df = CSV.read(
+    "$(git_root())/data/kinsler_2020/tidy_counts_no_anc.csv", DF.DataFrame
+)
+
+# Define datasets to process
+envs = ["1.7%", "1.8%", "2ugFlu", "8.5uMGdA"]
+
+# Define titles
+env_titles = ["1.7% glucose", "1.8% glucose", "2 µg Flu", "8.5 µM Gda"]
+
+# Keep data from environments to process
+df = df[[x ∈ envs for x in df.env], :]
+
+## %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+# Plot environment trajectories
+## %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+# Group data by environment
+df_group = DF.groupby(df, :env)
+
+# Initialize figure
+fig = Figure(resolution=(300 * 4, 350 * 2))
+
+# An array to store the GridLayout objects
+gls = []
+
+# Generating four GridLayout objects
+for i in 1:2, j in 1:2
+    # Initialize grid layout
+    gl = GridLayout()
[…]

### 2023-12-05 doc/_single_results.qmd
shorten discussion about Bayes vs Freq

@@
 bibliography: references.bib
-csl: ieee.csl
 ---
@@
 
-The main advantage of our method is this natural interpretability of these
+The main advantage of our method is the natural interpretability of these
 credible regions where an $X\%$ credible region indeed captures the region of
@@
 parameter lies given our statistical model, our prior information, and the
-observed experimental data. A common mistake in the literature is interpreting
-frequentist confidence intervals as Bayesian credible regions when they are not
-equivalent [@morey2016]. Frequentist confidence intervals and Bayesian credible
-regions are based on fundamentally different philosophical approaches to
-statistics. Frequentist confidence intervals represent the range of values that
-would contain the true population parameter with a certain probability if the
-experiment was repeated many times. The confidence interval does not represent
-the probability that the interval contains the true value. According to a
-specific model and prior information, Bayesian credible regions represent the
-range of values that contain the parameter with a certain posterior probability.
-The credible region directly represents the probability that the region contains
-the true value. So, frequentist confidence intervals cannot be interpreted as
-Bayesian credible regions because they have fundamentally different meanings.
-Treating an $X\%$ confidence interval like an $X\%$  credible region is
-fallacious since confidence intervals do not represent probabilistic coverage of
-the true value like credible regions. The intervals are generated through
-entirely different procedures.
+observed experimental data.
+<span style="color: #912E27;">
+Bayesian methods avoid common misconceptions associated with the construction of
+frequentists confidence intervals [@morey2016].
+</span>
 

### 2023-12-06 doc/_logistic_sim.qmd
update links to file to be hyperlinks

--- 2023-09-19 doc/_logistic_sim.qmd
@@
 noise**](./figs/figSIX_logistic_growth_04){#fig-SIX_logistic_growth_04}
+
+### Simulation details
+
+This section lists the parameters used to generate the synthetic data used
+throughout the main text. All the code to reproduce every simulation and figure
+in the paper can be found in [the GitHub
+repository](https://github.com/mrazomej/bayesian_fitness/tree/master). Unless
+stated otherwise, the simulations were initialized following these steps:
+
+1. The initial number of cells were defined as:
+- Ancestor lineage: `n_init` $\times$ `frac_anc`.
+- Neutral lineages: $n_0^{(n)} \sim \text{Poisson} 
+\left( 
+    \texttt{n\_init} \times \frac{\texttt{frac\_neutral}}{\texttt{n\_neutral}} 
+\right)$ 
+- Mutant lineages: $n_0^{(m)} \sim \text{LogNormal} 
+\left( 
+    \mu = \log(\texttt{n\_init} \times \frac{\texttt{frac\_mut}}{\texttt{n\_mut}}),
+    \sigma = 2 
+\right)$
+
+where
+
+- `n_init` is the initial number of cells.
+- `frac_anc` $\in [0, 1]$ is the initial fraction of the culture that is
+ancestor.
+- `frac_neutral` $\in [0, 1]$ is the initial fraction of the culture that is
+barcoded neutral lineages.
+- `frac_mut` $\in [0, 1]$ is the initial fraction of the culture that is
+barcoded lineages whose relative fitness we want to infer--hereafter "mutat"
+barcodes.
+- `frac_anc` + `frac_neutral` + `frac_mut` = 1
+- `n_neutral` is the number of unique neutral lineage barcodes.
+- `n_mut` is the number of unique mutant barcodes.
+
+2. The distribution of growth rates $\lambda$ were defined as
+- Ancestor lineage \& neutral lineages: $\lambda_a =$ `λ_a`. See details below
+for specific values for each simulation.
+- Mutant lineages: $\lambda_m \sim 
+\text{truncated}-\mathcal{N}(
+    \mu =$ `λ_mut_mean`,
+    $\sigma =$ `λ_bc_std`,
+    $\text{truncation limits} =$ `λ_trunc`
+$)$. See details below for specific values for each simulation.
+
+#### Fig 2.
+
+The following is a `README.md` file automatically generated when simulating the
+data for Fig. 2 in the main text:
+
+- File location [link to
+file](`https://github.com/mrazomej/bayesian_fitness/code/processing/mcmc_logistic_data001/sim_data001.jl`)
+- Number of mutant barcodes
+`n_mut = 225`
+- Number of neutral barcodes
+`n_neutral = 25`
+- Ancestral strain growth rate
+`λ_a = 1.0`
+- Carrying capacity
+`κ = 1.0e10`
+- Number of generations
+`n_gen = 8`
+- Initial number of cells
+`n_init = κ / (2^(n_gen)) = 3.90625e7`
+- Initial fraction of culture that is ancestor
+`frac_anc = 0.93`
+- Initial fraction of culture that is neutrals
+`frac_neutral = 0.02`
+- Initial fraction of culture that is mutants
+`frac_mut = 0.04999999999999995`
+- Mutant fitness distribution mean
+`λ_mut_mean = λ_a * 1.005 = 1.005`
+- Mutant fitness distribution standard deviation to sample growth rates
+`λ_mut_std = 0.1`
+- Mutant fitness distribution truncation ranges for growth rates
+`λ_trunc = [λ_a .* 0.9999, λ_a * 1.5] = [0.9999, 1.5]`
+- Gaussian noise distribution standard deviation
+`σ_lognormal = 0.3`
+
+#### Fig 3.
+
+The following is a `README.md` file automatically generated when simulating the
+data for Fig. 3 in the main text:
+
+- File location [link to
+file](`https://github.com/mrazomej/bayesian_fitness/code/processing/data003_logistic_1000bc_03env_01rep/sim_multienv_data003.jl`)
+- Number of environments
+`n_env = 3`
+- Environment transitions
+`env_idx = [1, 2, 3, 1, 2, 3]`
+- Number of mutant barcodes
+`n_mut = 900`
+- Number of neutral barcodes
+`n_neutral = 100`
+- Ancestral strain growth rate on each environment
+`λ_a = [1.0, 2.0, 0.5]`
+- Carrying capacity
+`κ = 1.0e10`
+- Number of generations
+`n_gen = 8`
+- Initial number of cells
+`n_init = κ / (2^(n_gen)) = 3.90625e7`
+- Initial fraction of culture that is ancestor
+`frac_anc = 0.93`
+- Initial fraction of culture that is neutrals
+`frac_neutral = 0.02`
+- Initial fraction of culture that is mutants
+`frac_mut = 0.04999999999999995`
+- Mutant fitness distribution mean
+`λ_mut_mean = λ_a * 1.005 = [1.005, 2.01, 0.5025]
+- Mutant fitness distribution standard deviation to sample growth rates
+`λ_mut_std = 0.1`
+- Mutant fitness distribution truncation ranges for growth rates
+`λ_trunc = [λ_a .* 0.9995, λ_a .* 1.5] = [[[0.995, 1.99, 0.4975]], [[1.5, 3.0, 0.75]]]`
+- Gaussian noise distribution standard deviation
+`σ_lognormal = 0.125`
+
+The variable `env_idx` lists the order of the environment cycles used in the
+simulation. In this case, transitions were of the form `1 -> 2`, `2 -> 3`, and
+`3 -> 1`.
+
+#### Fig 4.
+
+The following is a `README.md` file automatically generated when simulating the
+data for Fig. 4 in the main text:
+
+- File location [link to
+file](`https://github.com/mrazomej/bayesian_fitness/code/processing/data002_logistic_1000bc_01env_02rep/sim_hierarchical_data002.jl`)
+- Number of experimental replicates
+`n_data = 2`
+- Number of mutant barcodes
+`n_mut = 900`
+- Number of neutral barcodes
+`n_neutral = 100`
+- Ancestral strain growth rate
+`λ_a = 1.0`
+- Carrying capacity
+`κ = 1.0e10`
+- Number of generations
+`n_gen = 8`
+- Initial number of cells
+`n_init = κ / (2^(n_gen)) = 3.90625e7`
+- Initial fraction of culture that is ancestor
+`frac_anc = 0.93`
+- Initial fraction of culture that is neutrals
+`frac_neutral = 0.02`
+- Initial fraction of culture that is mutants
+`frac_mut = 0.04999999999999995`
+- Mutant fitness distribution mean
+`λ_bc_mean = λ_a * 1.005 = 1.005`
+- Mutant fitness distribution standard deviation to sample growth rates
+`λ_bc_std = 0.1`
+- Mutant fitness distribution truncation ranges for growth rates
+`λ_trunc = [λ_a .* 0.9999, λ_a * 1.5] = [0.9999, 1.5]`
+- Gaussian noise distribution standard deviation
+`σ_lognormal = 0.3`
+- Standard deviation for between-experiment fitness distribution variability
+`σ_exp = 0.015`
+
+The variable `σ_exp` defines the variance used to sample the variability between
+simulated experimental replicates. In other words, given a "hyper-fitness" value
+$\lambda$, for each experimental replicate, the growth rate was defined as
+$\lambda_j \sim \mathcal{N}(\lambda,$ `σ_exp` $)$.
+
+#### Fig 6.
+
+The following is a `README.md` file automatically generated when simulating the
+data for Fig. 6 in the main text:
+
+- File location [link to
+file](`https://github.com/mrazomej/bayesian_fitness/code/processing/data007_logistic_hierarchicalgenotype_1200bc_01env_01rep/sim_data007.jl`)
+- Number of mutant barcodes
+`n_mut = 900`
+- Number of neutral barcodes
+`n_neutral = 100`
+- Ancestral strain growth rate
+`λ_a = 1.0`
+- Carrying capacity
+`κ = 1.0e10`
+- Number of generations
+`n_gen = 8`
+- Initial number of cells
+`n_init = κ / (2^(n_gen)) = 3.90625e7`
+- Initial fraction of culture that is ancestor
+`frac_anc = 0.93`
+- Initial fraction of culture that is neutrals
+`frac_neutral = 0.02`
+- Initial fraction of culture that is mutants
+`frac_mut = 0.04999999999999995`
+- Mutant fitness distribution mean
+`λ_bc_mean = λ_a * 1.005 = 1.005`
+- Mutant fitness distribution standard deviation to sample growth rates
+`λ_bc_std = 0.1`
+- Mutant fitness distribution truncation ranges for growth rates
+`λ_trunc = [λ_a .* 0.9999, λ_a * 1.5] = [0.9999, 1.5]`
+- Gaussian noise distribution standard deviation
+`σ_lognormal = 0.3`
+- Number of unique genotypes
+`n_geno = Int64(n_mut / 10) = 90`
+
+The variable `n_geno` defines the number of unique genotypes in the simulation.
+In this simulation, we aimed to have ten unique barcodes on average for each
+genotype. The actual number of genotypes per barcode was sampled from a
+multinomial distribution with probability `1 / n_geno` for each genotype.

### 2023-12-06 doc/_logistic_sim.qmd
fix links to scripts

--- 2023-09-19 doc/_logistic_sim.qmd
@@
 noise**](./figs/figSIX_logistic_growth_04){#fig-SIX_logistic_growth_04}
+
+### Simulation details
+
+This section lists the parameters used to generate the synthetic data used
+throughout the main text. All the code to reproduce every simulation and figure
+in the paper can be found in [the GitHub
+repository](https://github.com/mrazomej/bayesian_fitness/tree/master). Unless
+stated otherwise, the simulations were initialized following these steps:
+
+1. The initial number of cells were defined as:
+- Ancestor lineage: `n_init` $\times$ `frac_anc`.
+- Neutral lineages: $n_0^{(n)} \sim \text{Poisson} 
+\left( 
+    \texttt{n\_init} \times \frac{\texttt{frac\_neutral}}{\texttt{n\_neutral}} 
+\right)$ 
+- Mutant lineages: $n_0^{(m)} \sim \text{LogNormal} 
+\left( 
+    \mu = \log(\texttt{n\_init} \times \frac{\texttt{frac\_mut}}{\texttt{n\_mut}}),
+    \sigma = 2 
+\right)$
+
+where
+
+- `n_init` is the initial number of cells.
+- `frac_anc` $\in [0, 1]$ is the initial fraction of the culture that is
+ancestor.
+- `frac_neutral` $\in [0, 1]$ is the initial fraction of the culture that is
+barcoded neutral lineages.
+- `frac_mut` $\in [0, 1]$ is the initial fraction of the culture that is
+barcoded lineages whose relative fitness we want to infer--hereafter "mutat"
+barcodes.
+- `frac_anc` + `frac_neutral` + `frac_mut` = 1
+- `n_neutral` is the number of unique neutral lineage barcodes.
+- `n_mut` is the number of unique mutant barcodes.
+
+2. The distribution of growth rates $\lambda$ were defined as
+- Ancestor lineage \& neutral lineages: $\lambda_a =$ `λ_a`. See details below
+for specific values for each simulation.
+- Mutant lineages: $\lambda_m \sim 
+\text{truncated}-\mathcal{N}(
+    \mu =$ `λ_mut_mean`,
+    $\sigma =$ `λ_bc_std`,
+    $\text{truncation limits} =$ `λ_trunc`
+$)$. See details below for specific values for each simulation.
+
+#### Fig 2.
+
+The following is a `README.md` file automatically generated when simulating the
+data for Fig. 2 in the main text:
+
+- File location [`link to
+file`](https://github.com/mrazomej/bayesian_fitness/blob/master/code/processing/data001_logistic_250bc_01env_01rep/sim_data001.jl)
+- Number of mutant barcodes
+`n_mut = 225`
+- Number of neutral barcodes
+`n_neutral = 25`
+- Ancestral strain growth rate
+`λ_a = 1.0`
+- Carrying capacity
+`κ = 1.0e10`
+- Number of generations
+`n_gen = 8`
+- Initial number of cells
+`n_init = κ / (2^(n_gen)) = 3.90625e7`
+- Initial fraction of culture that is ancestor
+`frac_anc = 0.93`
+- Initial fraction of culture that is neutrals
+`frac_neutral = 0.02`
+- Initial fraction of culture that is mutants
+`frac_mut = 0.04999999999999995`
+- Mutant fitness distribution mean
+`λ_mut_mean = λ_a * 1.005 = 1.005`
+- Mutant fitness distribution standard deviation to sample growth rates
+`λ_mut_std = 0.1`
+- Mutant fitness distribution truncation ranges for growth rates
+`λ_trunc = [λ_a .* 0.9999, λ_a * 1.5] = [0.9999, 1.5]`
+- Gaussian noise distribution standard deviation
+`σ_lognormal = 0.3`
+
+#### Fig 3.
+
+The following is a `README.md` file automatically generated when simulating the
+data for Fig. 3 in the main text:
+
+- File location [`link to
+file`](https://github.com/mrazomej/bayesian_fitness/blob/master/code/processing/data003_logistic_1000bc_03env_01rep/sim_multienv_data003.jl)
+- Number of environments
+`n_env = 3`
+- Environment transitions
+`env_idx = [1, 2, 3, 1, 2, 3]`
+- Number of mutant barcodes
+`n_mut = 900`
+- Number of neutral barcodes
+`n_neutral = 100`
+- Ancestral strain growth rate on each environment
+`λ_a = [1.0, 2.0, 0.5]`
+- Carrying capacity
+`κ = 1.0e10`
+- Number of generations
+`n_gen = 8`
+- Initial number of cells
+`n_init = κ / (2^(n_gen)) = 3.90625e7`
+- Initial fraction of culture that is ancestor
+`frac_anc = 0.93`
+- Initial fraction of culture that is neutrals
+`frac_neutral = 0.02`
+- Initial fraction of culture that is mutants
+`frac_mut = 0.04999999999999995`
+- Mutant fitness distribution mean
+`λ_mut_mean = λ_a * 1.005 = [1.005, 2.01, 0.5025]
+- Mutant fitness distribution standard deviation to sample growth rates
+`λ_mut_std = 0.1`
+- Mutant fitness distribution truncation ranges for growth rates
+`λ_trunc = [λ_a .* 0.9995, λ_a .* 1.5] = [[[0.995, 1.99, 0.4975]], [[1.5, 3.0, 0.75]]]`
+- Gaussian noise distribution standard deviation
+`σ_lognormal = 0.125`
+
+The variable `env_idx` lists the order of the environment cycles used in the
+simulation. In this case, transitions were of the form `1 -> 2`, `2 -> 3`, and
+`3 -> 1`.
+
+#### Fig 4.
+
+The following is a `README.md` file automatically generated when simulating the
+data for Fig. 4 in the main text:
+
+- File location [`link to
+file`](https://github.com/mrazomej/bayesian_fitness/blob/master/code/processing/data002_logistic_1000bc_01env_02rep/sim_hierarchical_data002.jl)
+- Number of experimental replicates
+`n_data = 2`
+- Number of mutant barcodes
+`n_mut = 900`
+- Number of neutral barcodes
+`n_neutral = 100`
+- Ancestral strain growth rate
+`λ_a = 1.0`
+- Carrying capacity
+`κ = 1.0e10`
+- Number of generations
+`n_gen = 8`
+- Initial number of cells
+`n_init = κ / (2^(n_gen)) = 3.90625e7`
+- Initial fraction of culture that is ancestor
+`frac_anc = 0.93`
+- Initial fraction of culture that is neutrals
+`frac_neutral = 0.02`
+- Initial fraction of culture that is mutants
+`frac_mut = 0.04999999999999995`
+- Mutant fitness distribution mean
+`λ_bc_mean = λ_a * 1.005 = 1.005`
+- Mutant fitness distribution standard deviation to sample growth rates
+`λ_bc_std = 0.1`
+- Mutant fitness distribution truncation ranges for growth rates
+`λ_trunc = [λ_a .* 0.9999, λ_a * 1.5] = [0.9999, 1.5]`
+- Gaussian noise distribution standard deviation
+`σ_lognormal = 0.3`
+- Standard deviation for between-experiment fitness distribution variability
+`σ_exp = 0.015`
+
+The variable `σ_exp` defines the variance used to sample the variability between
+simulated experimental replicates. In other words, given a "hyper-fitness" value
+$\lambda$, for each experimental replicate, the growth rate was defined as
+$\lambda_j \sim \mathcal{N}(\lambda,$ `σ_exp` $)$.
+
+#### Fig 6.
+
+The following is a `README.md` file automatically generated when simulating the
+data for Fig. 6 in the main text:
+
+- File location [`link to
+file`](https://github.com/mrazomej/bayesian_fitness/blob/master/code/processing/data007_logistic_hierarchicalgenotype_1000bc_01env_01rep/sim_data007.jl)
+- Number of mutant barcodes
+`n_mut = 900`
+- Number of neutral barcodes
+`n_neutral = 100`
+- Ancestral strain growth rate
+`λ_a = 1.0`
+- Carrying capacity
+`κ = 1.0e10`
+- Number of generations
+`n_gen = 8`
+- Initial number of cells
+`n_init = κ / (2^(n_gen)) = 3.90625e7`
+- Initial fraction of culture that is ancestor
+`frac_anc = 0.93`
+- Initial fraction of culture that is neutrals
+`frac_neutral = 0.02`
+- Initial fraction of culture that is mutants
+`frac_mut = 0.04999999999999995`
+- Mutant fitness distribution mean
+`λ_bc_mean = λ_a * 1.005 = 1.005`
+- Mutant fitness distribution standard deviation to sample growth rates
+`λ_bc_std = 0.1`
+- Mutant fitness distribution truncation ranges for growth rates
+`λ_trunc = [λ_a .* 0.9999, λ_a * 1.5] = [0.9999, 1.5]`
+- Gaussian noise distribution standard deviation
+`σ_lognormal = 0.3`
+- Number of unique genotypes
+`n_geno = Int64(n_mut / 10) = 90`
+
+The variable `n_geno` defines the number of unique genotypes in the simulation.
+In this simulation, we aimed to have ten unique barcodes on average for each
+genotype. The actual number of genotypes per barcode was sampled from a
+multinomial distribution with probability `1 / n_geno` for each genotype.

### 2023-12-11 doc/_kinsler.qmd
first draft of Kinsler et al SI section

---
editor:
    render-on-save: true
bibliography: references.bib
csl: ieee.csl
---

## Reanalyzing Kinsler et al., 2020 {#sec-kinsler}

The main text of this paper focuses only on simulation-based assessments of the
performance of our inference algorithm. This supplementary section shows how the
model performs on experimental data. In particular, we focus on a subset of the
data generated for [@kinsler2020]. In this paper, the fitness of a set of
mutants originally evolved in minimal glucose media was determined in multiple
environments using an experimental design described here. @fig-SIX_kinsler_data
shows the raw barcode frequency trajectories for four of such environments
measured in two biological replicates.

![**Kinsler et al., 2020 sample data**. Barcode frequency trajectories for
minimal glucose evolved mutants in four different environments. Each environment
was measured in two biological replicates (R1 and R2 on each panel). For
experimental details, see
[@kinsler2020].](./figs/figSIX_kinsler_data){#fig-SIX_kinsler_data}

### Analysis of individual replicates

As a first analysis, we perform inference assuming all replicates are
independent. @fig-SIX_kinsler_ppc_single shows the posterior predictive checks
for the neutral lineages (upper left plot on each panel) and some sample mutants
for both replicates. We can see that the inference works exceptionally well with
the data, capturing the variability. Although the main text inferences were made
on logistic growth simulations, as described in @sec-logistic, the results
in @sec-fitseq2 and this section show that our model does not depend on the
assumptions underlying such simulations. The only assumption that must be valid
is that the changes in the relative frequencies of each lineage follow an
exponential function.

![**Posterior predictive checks for Kinsler et al., 2020 independent
replicates**. Examples of the posterior predictive checks for all neutral
lineages (upper left panels) and a subset of representative mutant lineages for
all four environments shown in @fig-SIX_kinsler_data. Shaded regions from light
to dark represent the 95\%, 68\%, and 5\% credible regions. Listed fitness
values for mutant lineages represent the 68\% credible
region.](./figs/figSIX_kinsler_ppc_single){#fig-SIX_kinsler_ppc_single}

Furthermore, we can compare the replicate-to-replicate correlation for these
four environments with two biological replicates. The result of this comparison
is shown in @fig-SIX_kinsler_rep_single.

![**Replicate-to-replicate correlation**. Comparison of inferred fitness values
between experimental replicates when fitting each replicate individually for all
environments shown in @fig-SIX_kinsler_data. Gray error bars show the 68\%
credible region for this
inference.](./figs/figSIX_kinsler_rep_single){#fig-SIX_kinsler_rep_single}

### Hierarchical analysis of multiple replicates

As discussed in @sec-replicates, most experimental designs---including that of
[@kinsler2020]---include two or more biological replicate measurements.
Therefore, a natural way to analyze the data is by jointly inferring the fitness
value using a hierarchical model. @fig-SIX_kinsler_ppc_hier shows the posterior
predictive checks when fitting a hierarchical model to each of the four
environments shown in @fig-SIX_kinsler_data. Comparing this figure with
@fig-SIX_kinsler_ppc_single shows that the hierarchical model does not lose the
ability to capture the variability in the data compared to fitting each
replicate individually. However, as shown in @fig-SIX_kinsler_rep_hier, when
comparing the replicate-to-replicate correlation, as expected, the hierarchical
model does a better job.

![**Posterior predictive checks for Kinsler et al., 2020 hierarchical model**.
Examples of the posterior predictive checks for all neutral lineages (upper left
panels) and a subset of representative mutant lineages for all four environments
shown in @fig-SIX_kinsler_data. Shaded regions from light to dark represent the
95\%, 68\%, and 5\% credible regions. Listed fitness values for mutant lineages
represent the 68\% credible
region.](./figs/figSIX_kinsler_ppc_hier){#fig-SIX_kinsler_ppc_hier}

![**Replicate-to-replicate correlation**. Comparison of inferred fitness values
between experimental replicates when fitting a hierarchical model on the
biological replicates for all environments shown in @fig-SIX_kinsler_data. Gray
error bars show the 68\% credible region for this
inference.](./figs/figSIX_kinsler_rep_hier){#fig-SIX_kinsler_rep_hier}

### 2023-12-11 doc/_fitseq2_sim.qmd
small update to text

@@
 parameters for all mutant lineages. @fig-SIX_fitseq2 shows the comparison
-between the inference with the software provided in @li2023 and our inference
+between the inference with the software provided in [@li2023] and our inference
 pipeline. We note that although the values are highly correlated, the scale of
-the axis is very different.
+the axis is very different. This scale difference comes from the differences in
+models used to infer the relative fitness value. While our model assumes an
+exponential change in the relative frequency of the barcode, @li2023 works
+directly with the number of cells rather than the relative frequency, assuming a
+Malthusian fitness form. In other words, while our model is agnostic to the
+underlying type of growth mechanism, as long as the condition of exponential
+change in relative frequency is satisfied, @li2023 assumes exponential growth of
+the number of cells, resulting in a discrepancy between the obtained numbers.
 

### 2023-12-13 doc/_kinsler.qmd
extended analysis significantly

@@
 mutants originally evolved in minimal glucose media was determined in multiple
-environments using an experimental design described here. @fig-SIX_kinsler_data
-shows the raw barcode frequency trajectories for four of such environments
-measured in two biological replicates.
+environments using an experimental design described in @sec-experiment.
+@fig-SIX_kinsler_data shows the raw barcode frequency trajectories for four of
+such environments measured in two biological replicates.
 
-![**Kinsler et al., 2020 sample data**. Barcode frequency trajectories for
-minimal glucose evolved mutants in four different environments. Each environment
-was measured in two biological replicates (R1 and R2 on each panel). For
-experimental details, see
+![**Kinsler et al., 2020 sample data**. Barcode frequency trajectories in four
+different environments for mutants evolved in minimal glucose media. Each
+environment was measured in two biological replicates (R1 and R2 on each panel).
+For experimental details, see
 [@kinsler2020].](./figs/figSIX_kinsler_data){#fig-SIX_kinsler_data}
@@
 four environments with two biological replicates. The result of this comparison
-is shown in @fig-SIX_kinsler_rep_single.
+is shown in @fig-SIX_kinsler_rep_single. We can see that there is a good degree
+of correlation between replicates.
 
-![**Replicate-to-replicate correlation**. Comparison of inferred fitness values
-between experimental replicates when fitting each replicate individually for all
-environments shown in @fig-SIX_kinsler_data. Gray error bars show the 68\%
-credible region for this
+![**Replicate-to-replicate correlation for single-dataset model**. Comparison of
+inferred relative fitness values for mutant strains between experimental
+replicates when fitting each replicate individually for all environments shown
+in @fig-SIX_kinsler_data. Gray error bars show the 68\% credible region for this
 inference.](./figs/figSIX_kinsler_rep_single){#fig-SIX_kinsler_rep_single}
+
+Given that we have two biological replicates of the same experiment, we can ask
+how the difference between each replicate's inferred fitness values compares
+with the individual estimates' uncertainty. In other words, we want to compare
+how different the $x$- and $y$-axis values in @fig-SIX_kinsler_rep_single are
+with the size of the shown error bars. @fig-SIX_kinsler_meandiff_single compares
+the distribution of these magnitudes for all four environments. In each plot,
+the blue line shows the empirical cumulative distribution function (ECDF) of the
+distance between the mean fitness estimate for each replicate---computed as
+$\sqrt{\left( \left\langle s_1^{(m)}\right\rangle - \left\langle
+s_2^{(m)}\right\rangle \right)^2}$, where $\langle s_i^{(m)}\rangle$ is the mean
+fitness estimate of mutant $m$ in replicate $i$---meaning the difference between
+$x$- and $y$-axis in @fig-SIX_kinsler_rep_single. The orange and green lines
+represent the ECDF for each replicate's posterior distribution standard
+deviation, i.e., the ECDF of the size of the error bars in
+@fig-SIX_kinsler_rep_single. From these plots, we can see that the distribution
+of standard deviations is shifted to the left compared to the difference between
+replicates. This implies that the uncertainty in the estimate is generally
+larger than the difference between replicates one would obtain with a naive
+point estimate.
+
+![**Comparing the replicate-to-replicate deviations with within-batch noise
+estimates for single dataset analysis**. Empirical cumulative distribution
+functions of the distance between the mean fitness estimate for each replicate
+(blue lines) and each replicate's posterior distribution standard deviation
+(orange and green
+line).](./figs/figSIX_kinsler_meandiff_single){#fig-SIX_kinsler_meandiff_single}
 
@@
 As discussed in @sec-replicates, most experimental designs---including that of
-[@kinsler2020]---include two or more biological replicate measurements.
-Therefore, a natural way to analyze the data is by jointly inferring the fitness
-value using a hierarchical model. @fig-SIX_kinsler_ppc_hier shows the posterior
+@kinsler2020---include two or more biological replicate measurements. Therefore,
+a natural way to analyze the data is by jointly inferring the fitness value
+using a hierarchical model. @fig-SIX_kinsler_ppc_hier shows the posterior
 predictive checks when fitting a hierarchical model to each of the four
@@
 ability to capture the variability in the data compared to fitting each
-replicate individually. However, as shown in @fig-SIX_kinsler_rep_hier, when
-comparing the replicate-to-replicate correlation, as expected, the hierarchical
-model does a better job.
+replicate individually. However, as shown in @fig-SIX_kinsler_rep_hier, the
+hierarchical model does a better job when comparing the replicate-to-replicate
+correlation, as expected. To emphasize this last point,
+@fig-SIX_kinsler_rep_comb shows both inferences on the same plot---only the mean
+of the posterior distribution is shown for visual clarity.
 
@@
 
-![**Replicate-to-replicate correlation**. Comparison of inferred fitness values
-between experimental replicates when fitting a hierarchical model on the
-biological replicates for all environments shown in @fig-SIX_kinsler_data. Gray
-error bars show the 68\% credible region for this
+![**Replicate-to-replicate correlation for hierarchical model**. Comparison of
+inferred fitness values between experimental replicates when fitting a
+hierarchical model on the biological replicates for all environments shown in
+@fig-SIX_kinsler_data. Gray error bars show the 68\% credible region for this
 inference.](./figs/figSIX_kinsler_rep_hier){#fig-SIX_kinsler_rep_hier}
+
+![**Replicate-to-replicate correlation for both methods**. Comparison of
+**mean** inferred fitness values between experimental replicates when fitting a
+single-dataset model (blue points) vs. a hierarchical model on the biological
+replicates (orange points) for all environments shown in @fig-SIX_kinsler_data.
+Error bars are not included in the plot for visual
+clarity](./figs/figSIX_kinsler_rep_comb){#fig-SIX_kinsler_rep_comb}
+
+We can again compare the magnitude of the difference between replicates'
+inferred mean fitness with the size of the corresponding error bars, as we did
+with @fig-SIX_kinsler_meandiff_single. @fig-SIX_kinsler_meandiff_hier shows the
+ECDF for the resulting difference between replicates mean fitness in blue as
+well as the resulting standard deviation for the inferred hyper-fitness
+parameter that connects both replicates in orange (see @sec-replicates for
+description of this hyper-fitness parameter). Here, we see that the standard
+deviation ECDF (orange line) is shifted more to the right than the mean
+difference between replicates (blue line).
+
+![**Comparing the replicate-to-replicate deviations with within-batch noise
+estimates for hierarchical model analysis**. Empirical cumulative distribution
+functions of the distance between the mean fitness estimate for each replicate
+(blue lines) and each replicate's posterior distribution standard deviation
+(orange and green
+line).](./figs/figSIX_kinsler_meandiff_hier){#fig-SIX_kinsler_meandiff_hier}
+
+Even though the difference between replicates is reduced when using a
+hierarchical model, this does not come at the expense of higher uncertainty in
+the estimate of the relative fitness. To show this is the case,
+@fig-SIX_kinsler_std_comp compares the reported posterior distribution standard
+deviations for the relative fitness when inferring each dataset individually
+($x$-axis) with the standard deviation of the hyper-fitness parameter
+($y$-axis). Most of the points fall below the identity line, demonstrating that
+the uncertainty in the hyper-fitness parameter obtained by utilizing information
+from both replicates is smaller than the equivalent uncertainty obtained with
+one replicate at a time. In conclusion, the hierarchical model not only can
+equally describe the data, as shown in @fig-SIX_kinsler_ppc_hier, but the
+difference between replicates is significantly decreased, as shown in
+@fig-SIX_kinsler_rep_comb, at the same time, the uncertainty in the parameters'
+estimate is decreased, as shown in @fig-SIX_kinsler_std_comp.
+
+![**Comparing estimates uncertainties between single-dataset and hierarchical
+model**. Comparison of the reported standard deviations for the mutants'
+relative fitness when inferring the relative fitness using one dataset at a time
+($x$-axis) and the corresponding standard deviation for the mutants'
+hyper-fitness parameter when utilizing both replicates simultaneously via a
+hierarchical model ($y$-axis). Dashed line shows identity
+line.](./figs/figSIX_kinsler_std_comp){#fig-SIX_kinsler_std_comp}

### 2023-12-13 code/processing/data014_kinsler2020/viz_inference.jl
update analysis plots for Kinsler et al. data

@@ -147,7 +147,7 @@ advidir = "./output/advi_meanfield_joint_inference"
 files = Glob.glob("$(advidir)/*csv")
 
 # Initialize dataframe
-df_advi = DF.DataFrame()
+df_advi_single = DF.DataFrame()
 
 # Loop through files
 for f in files
@@ -165,11 +165,11 @@ for f in files
     df_tmp[!, :env] .= env
     df_tmp[!, :rep] .= rep
 
-    append!(df_advi, df_tmp)
+    append!(df_advi_single, df_tmp)
 end # for
 
 # Keep only fitness values
-df_fitness_single = df_advi[df_advi.vartype.=="bc_fitness", :]
+df_fitness_single = df_advi_single[df_advi_single.vartype.=="bc_fitness", :]
 
 ## %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 # Plot replicates comparisons
@@ -260,7 +260,7 @@ for (i, data) in enumerate(df_group)
 end # for
 
 save("$(git_root())/doc/figs/figSIX_kinsler_rep_single.pdf", fig)
-save("$(git_root())/doc/figs/figSIX_kinsler_rep_signle.png", fig)
+save("$(git_root())/doc/figs/figSIX_kinsler_rep_single.png", fig)
 
 fig
 
@@ -315,7 +315,7 @@ for (i, env) in enumerate(envs)
     df_counts = df[df.env.==env, :]
 
     # Extract ADVI inference
-    df_advi_env = df_advi[df_advi.env.==env, :]
+    df_advi_env = df_advi_single[df_advi_single.env.==env, :]
 
     # Sample from posterior MvNormal
     df_samples = DF.DataFrame(
@@ -514,7 +514,7 @@ advidir = "./output/advi_meanfield_hierarchicalreplicate_inference"
 files = Glob.glob("$(advidir)/*csv")
 
 # Initialize dataframe
-df_advi = DF.DataFrame()
+df_advi_hier = DF.DataFrame()
 
 # Loop through files
 for f in files
@@ -529,11 +529,11 @@ for f in files
     # Add env and rep column
     df_tmp[!, :env] .= env
 
-    append!(df_advi, df_tmp)
+    append!(df_advi_hier, df_tmp)
 end # for
 
 # Keep only fitness values
[…]

### 2023-12-20 doc/_fitness_model.qmd
Add schematic diagram showing sources of uncertainty

@@
 
+<span style="color: #912E27;">
+@fig-01(C) shows schematically the sources of uncertainty accounted for by our
+model. The first three sources---population composition, barcode read counts,
+and barcode frequencies---all contribute to uncertainty source 1. Uncertainty
+source 2 and 3 are depicted by the last two sources, respectively.
+</span>
 To account for all these sources of uncertainty in a principled way, in the next

### 2023-12-20 doc/_discussion.qmd
Add explanation of fitness model dynamics and future directions

@@
 facilitates the inference of the population mean fitness since most of the
-culture---the unlabeled reference genotype---is not tracked. Finally, the
-presented statistical model assumes that relative fitness is solely a constant
-of the environment and the genotype. Future directions of this work could extend
-the fitness model to properly analyze data with time-varying or
+culture---the unlabeled reference genotype---is not tracked. 
+<span style="color: #912E27;">
+Fourth, the fitness model in @eq-fitness does not assume any functional form for
+the cellular growth dynamics. This description of the changes in the relative
+frequency of the genotypes can approximate the dynamics for multiple growth
+forms. For example, in @sec-fitseq2, in the supplementary materials, we apply
+our inference pipeline to simulations like those presented in [@li2023] with
+non-logistic growth models. Moreover, @sec-kinsler, in the supplementary
+materials, reanalyzes experimental data from [@kinsler2020], where previous work
+showed that the fitness advantage from some of the mutants results from a
+shorter lag phase at the beginning of the growth cycle [@li2018]. Thus, our
+method's inferred relative fitness is a coarse-grained quantity that integrates
+the differences in fitness across the entire growth cycle as long as frequencies
+and frequency changes are small enough to be approximated by @eq-fitness.
+</span>
+Finally, the presented statistical model assumes that relative fitness is solely
+a constant of the environment and the genotype. Future directions of this work
+could extend the fitness model to properly analyze data with time-varying or
 frequency-dependent fitness values.

### 2023-12-20 doc/_experiment.qmd
Add explanation of uncertainty sources in figure description

@@
 relative fitness. Darker gray lines define the typical trajectory of neutral
-lineages.](./figs/fig01){#fig-01}
+lineages. 
+<span style="color: #912E27;">
+(C) Sources of uncertainty accounted for by our method. The Bayesian
+model fit to the data propagates uncertainties from the categories schematically
+depicted to all parameters in the inference.
+</span>](./figs/fig01){#fig-01}

### 2023-12-31 doc/_time_scaling.qmd
Add time scaling section to paper

---
editor:
    render-on-save: true
bibliography: references.bib
csl: ieee.csl
---

## Computation time scaling {#sec-scaling}

The biggest advantage of the variational inference approach presented in this
work is the computation speedup compared to MCMC-based Bayesian approaches. This
allows for applying our principled Bayesian pipeline to datasets with many
unique barcodes or multiple datasets analyzed hierarchically.
@fig-SIX_bc_vs_time shows a linear scaling in the amount of time taken for the
ADVI algorithm to take 7,500 steps for a varying number of unique barcodes. This
figure shows that the time to fit larger models increases linearly with the
number of barcodes.

![**Scaling of computation time as a function of number of unique barcodes**.
Fitting a single-dataset model for 500 to 100,000 unique barcodes by taking
7,500 steps.](./figs/figSIX_bc_vs_time){#fig-SIX_bc_vs_time}

### 2024-01-20 doc/paper.tex
update corresponding author
TeX file with all journal requirements

@@
 \setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\setcounter{secnumdepth}{5}
 % Make \paragraph and \subparagraph free-standing
@@
 \usepackage{authblk}
+
+% Allow to define personalized colors
+\usepackage{xcolor}
 \makeatletter
@@
 University}
-\affil[4]{Department of Molecular Biosciences, Northwestern University}
+\affil[4]{Department of Engineering Sciences and Applied Mathematics,
+Northwestern University}
 \affil[5]{Chan Zuckerberg Biohub}
@@
 \maketitle
-\begin{abstract}
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable]}{\end{tcolorbox}}\fi
+
+% Remove main text from the table of contents by specifying not to include
+% any section or subsection
+\addtocontents{toc}{\protect\setcounter{tocdepth}{-1}}
+
+% Define reference segment for main text
+\begin{refsegment}
+% Generate filter to not include references from main text in the
+% supplemental references
+\defbibfilter{notother}{not segment=\therefsegment}
+
+\hypertarget{abstract}{%
+\section*{Abstract}\label{abstract}}
+\addcontentsline{toc}{section}{Abstract}
+
 The tracking of lineage frequencies via DNA barcode sequencing enables
@@
 principled statistical methods in experimental evolution.
-\end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden]}{\end{tcolorbox}}\fi
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Remove main text from the table of contents by specifying not to include
-% any section or subsection
-\addtocontents{toc}{\protect\setcounter{tocdepth}{-1}}
-
-% Define reference segment for main text
-\begin{refsegment}
-% Generate filter to not include references from main text in the
-% supplemental references
-\defbibfilter{notother}{not segment=\therefsegment}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\hypertarget{author-summary}{%
+\section*{Author Summary}\label{author-summary}}
+\addcontentsline{toc}{section}{Author Summary}
+
+In this study, we present a novel Bayesian pipeline for analyzing DNA
+barcode tracking sequencing data, addressing the challenge of accurately
+quantifying competitive microbial fitness in the presence of
+experimental noise. Our method uniquely contributes to understanding
+microbial evolutionary dynamics by enabling reliable inference of the
+relative fitness of diverse microbial strains from high-throughput
+lineage tracking assays. Our approach is distinct in its ability to
+systematically account for and propagate uncertainties from various
+noise sources throughout all inferred parameters. Furthermore, the
+error-propagation quality of our Bayesian method allows us to extend the
+inference pipeline to common dataset structures, such as jointly
+analyzing multiple experimental replicates or accounting for multiple
+unique barcodes mapping the equivalent genotypes. This comprehensive
+treatment of uncertainties is crucial in experimental settings where
+noise can significantly influence the results. Furthermore, we have
+optimized our pipeline for scalability, allowing it to handle large
+numbers of unique barcodes effectively. This scalability is essential
+for analyzing complex datasets typical in microbial fitness studies.
 
@@
 (\(\approx 3-5\%\)). The rest of the culture is left for the ensemble of
-genotypes of interest.
+genotypes of interest. This experimental design in which the barcodes of
+interest represent a small fraction of the culture serves three main
+purposes: First, the model used to infer the relative fitness
+differences between genotypes is valid in the regime where the genotype
+frequency is significantly smaller than one (see
+Section~\ref{sec-fitness_model}). Second, potential problems with
+frequency-dependent selection are minimized as long as the frequency of
+each genotype remains small. Third, having a single genotype dominating
+the culture standardizes the environment experienced by all genotypes.
+This is because variations in the chemistry of the environment are
+effectively dictated by the dominating genotype, allowing for
+reproducible growth-dilution cycles.
 
@@
 fitness. Darker gray lines define the typical trajectory of neutral
-lineages.}
+lineages. (C) Sources of uncertainty accounted for by our method. The
+Bayesian model fit to the data propagates uncertainties from the
+categories schematically depicted to all parameters in the inference. }
 
@@
 
-To account for all these sources of uncertainty in a principled way, in
-the next section, we develop a Bayesian inference pipeline.
+Figure~\ref{fig-01}(C) shows schematically the sources of uncertainty
+accounted for by our model. The first three sources---population
+composition, barcode read counts, and barcode frequencies---all
+contribute to uncertainty source 1. Uncertainty source 2 and 3 are
+depicted by the last two sources, respectively. To account for all these
+sources of uncertainty in a principled way, in the next section, we
+develop a Bayesian inference pipeline.
 
@@
 
-The main advantage of our method is this natural interpretability of
+The main advantage of our method is the natural interpretability of
 these credible regions where an \(X\%\) credible region indeed captures
@@
 the actual value of the parameter lies given our statistical model, our
-prior information, and the observed experimental data. A common mistake
-in the literature is interpreting frequentist confidence intervals as
-Bayesian credible regions when they are not equivalent
-\autocite{morey2016}. Frequentist confidence intervals and Bayesian
-credible regions are based on fundamentally different philosophical
-approaches to statistics. Frequentist confidence intervals represent the
-range of values that would contain the true population parameter with a
-certain probability if the experiment was repeated many times. The
-confidence interval does not represent the probability that the interval
-contains the true value. According to a specific model and prior
-information, Bayesian credible regions represent the range of values
-that contain the parameter with a certain posterior probability. The
-credible region directly represents the probability that the region
-contains the true value. So, frequentist confidence intervals cannot be
-interpreted as Bayesian credible regions because they have fundamentally
-different meanings. Treating an \(X\%\) confidence interval like an
-\(X\%\) credible region is fallacious since confidence intervals do not
-represent probabilistic coverage of the true value like credible
-regions. The intervals are generated through entirely different
-procedures.
+prior information, and the observed experimental data. Bayesian methods
+avoid common misconceptions associated with the construction of
+frequentists confidence intervals \autocite{morey2016}.
 
@@
 together and treat our different experiments as a single measurement
-with higher coverage. This loses the subtle differences due to biotic
-and abiotic batch effects, effectively halving the data that goes into
-our inference problem.
+with higher coverage. However, by doing so, we lose all information
+about experiment-to-experiment variability due to intrinsic biological
+variability and environmental fluctuations---such as small temperature
+fluctuations---that we cannot control. In this sense, combining the
+datasets defeats the purpose of performing multiple measurements to
+account for this variability when extracting the relevant information
+from the observations.
 
@@
 \(\underline{\theta}^G\), where \(G\) is the number of genotypes in the
-dataset.
+dataset. This analysis pipeline assumes a known barcode-to-genotype
+mapping to assign each barcode to the corresponding hyper-fitness
+parameter uniquely.
 
@@
 population mean fitness since most of the culture---the unlabeled
-reference genotype---is not tracked. Finally, the presented statistical
-model assumes that relative fitness is solely a constant of the
-environment and the genotype. Future directions of this work could
-extend the fitness model to properly analyze data with time-varying or
+reference genotype---is not tracked. Fourth, the fitness model in
+Equation~\ref{eq-fitness} does not assume any functional form for the
+cellular growth dynamics. This description of the changes in the
+relative frequency of the genotypes can approximate the dynamics for
+multiple growth forms. For example, in Section~\ref{sec-fitseq2}, in the
+supplementary materials, we apply our inference pipeline to simulations
+like those presented in \autocite{li2023} with non-logistic growth
+models. Moreover, Section~\ref{sec-kinsler}, in the supplementary
+materials, reanalyzes experimental data from \autocite{kinsler2020},
+where previous work showed that the fitness advantage from some of the
+mutants results from a shorter lag phase at the beginning of the growth
+cycle \autocite{li2018}. Thus, our method's inferred relative fitness is
+a coarse-grained quantity that integrates the differences in fitness
+across the entire growth cycle as long as frequencies and frequency
+changes are small enough to be approximated by
+Equation~\ref{eq-fitness}. Finally, the presented statistical model
+assumes that relative fitness is solely a constant of the environment
+and the genotype. Future directions of this work could extend the
+fitness model to properly analyze data with time-varying or
 frequency-dependent fitness values.
@@
 
+\hypertarget{materials-and-methods}{%
+\section*{Materials and Methods}\label{materials-and-methods}}
+\addcontentsline{toc}{section}{Materials and Methods}
+
+All synthetic data generation and custom scripts used in this work were
+stored using Git version control. Code for analysis and figure
+generation can be found on the GitHub repository
+(https://github.com/mrazomej/bayesian\_fitness). The accompanying
+software package \texttt{BarBay.jl} can be directly installed from the
+\texttt{Julia} package repository, or via cloning the corresponding
+GitHub repository (https://github.com/mrazomej/BarBay.jl)
+
 \hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
+\section*{Acknowledgements}\label{acknowledgements}}
+\addcontentsline{toc}{section}{Acknowledgements}
 
@@
 Spencer Farrell, and Shaili Mathur for critical observations on the
-manuscript. This work was supported by the NIH/NIGMS, Genomics of rapid
-adaptation in the lab and in the wild R35GM11816506 (MIRA grant), the
-NIH, Unravelling mechanisms of tumor suppression in lung cancer
-R01CA23434903, the NIH (PQ4), Quantitative and multiplexed analysis of
-gene function in cancer in vivo R01CA23125303, the NIH, Genetic
-Determinants of Tumor Growth and Drug Sensitivity in EGFR Mutant Lung
-Cancer R01CA263715, the NIH, Dissecting the interplay between aging,
-genotype, and the microenvironment in lung cancer U01AG077922, the NIH,
-Genetic dissection of oncogenic Kras signaling R01CA230025, and the CZ
-Biohub investigator program. MRM was supported by the Schmidt Science
-Fellowship. MM was supported by The National Science Foundation-Simons
-Center for Quantitative Biology at Northwestern University and the
-Simons Foundation grant 597491. MM is a Simons Investigator. DP is a CZ
-Biohub investigator.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+manuscript. We are especially thankful to Grant Kinsler for kindly
+providing raw experimental data as well as lengthy discussions about the
+state-of-the-art inference method. This work was supported by the
+NIH/NIGMS, Genomics of rapid adaptation in the lab and in the wild
+R35GM11816506 (MIRA grant), the NIH, Unravelling mechanisms of tumor
+suppression in lung cancer R01CA23434903, the NIH (PQ4), Quantitative
+and multiplexed analysis of gene function in cancer in vivo
+R01CA23125303, the NIH, Genetic Determinants of Tumor Growth and Drug
+Sensitivity in EGFR Mutant Lung Cancer R01CA263715, the NIH, Dissecting
+the interplay between aging, genotype, and the microenvironment in lung
+cancer U01AG077922, the NIH, Genetic dissection of oncogenic Kras
+signaling R01CA230025, and the CZ Biohub investigator program. MRM was
+supported by the Schmidt Science Fellowship. MM was supported by The
+National Science Foundation-Simons Center for Quantitative Biology at
+Northwestern University and the Simons Foundation grant 597491. MM is a
+Simons Investigator. DP is a CZ Biohub investigator.
+
 % Print main text references
@@
 
+\hypertarget{supp_figs}{%
+\section*{Supplemental Materials}\label{supp_figs}}
+\addcontentsline{toc}{section}{Supplemental Materials}
+
+\textbf{Figure S1. High-dimensional numerical quadrature does not scale with
+dimensionality.} Schematic depiction of the problem with naive numerical
+quadrature to integrate over an unknown density. While the density is
+concentrated on the dark peak, most of the evaluations over the \(x_1-x_2\) grid
+do not contribute to the value of the integral.
+
+\textbf{Figure S2. Logistic growth simulation over single growth cycle}. The
+dashed line represents the neutral lineages, with the upper curve being the
+unlabeled neutral strain. Color curves represent the genotypes of interest
+colored by growth rate relative to the neutral lineage.
+
+\textbf{Figure S3. Growth-dilution cycles for logistic growth simulation}. Each
+point represents the final number of cells after a growth cycle for each
+lineage. Colors are the same as in Figure~\ref{fig-SIX_logistic_growth_01}.
+
+\textbf{Figure S4. Log frequency ratio for logistic growth simulations}. The
+relative distance of the color curves from the black dashed line determines the
+relative fitness of each linage.
+
+\textbf{Figure S5. Logistic growth-dilution simulations with and without noise}
+
+\textbf{Figure S6. FitSeq2.0 simulation comparison}.(A) Frequency trajectories
+that represent the raw data going into the inference. (B) Log frequency ratio
+between two adjacent time points used by the inference pipeline. Darker lines
+represent the neutral barcodes. These transformed data are much more noisy than
+the seemingly smooth frequency trajectories. (C) Examples of the posterior
+predictive checks for all neutral lineages (upper left panel) and a subset of
+representative mutant lineages. Shaded regions represent the 95\%, 68\%, and 5\%
+credible regions for the data. The reported errors above the plot represent the
+68\% credible region on the mutant relative fitness marginal distribution. (D)
+Comparison between the ground truth fitness value from the logistic-growth
+simulation and the inferred fitness value. Gray error bars represent the 68\%
+posterior credible region for the relative fitness values. (E) Comparison
+between fitness inferences as obtained using \texttt{FitSeq2.0} from
+\textcite{li2023} and our Bayesian method. Vertical gray error bars represent
+the 68\% posterior credible region for the relative fitness values.
+
+\textbf{Figure S7. Kinsler et al., 2020 sample data}. Barcode frequency
+trajectories in four different environments for mutants evolved in minimal
+glucose media. Each environment was measured in two biological replicates (R1
+and R2 on each panel). For experimental details, see \autocite{kinsler2020}.
+
+\textbf{Figure S8. Posterior predictive checks for Kinsler et al., 2020
+independent replicates}. Examples of the posterior predictive checks for all
+neutral lineages (upper left panels) and a subset of representative mutant
+lineages for all four environments shown in Figure~\ref{fig-SIX_kinsler_data}.
+Shaded regions from light to dark represent the 95\%, 68\%, and 5\% credible
+regions. Listed fitness values for mutant lineages represent the 68\% credible
+region.
+
+\textbf{Figure S9. Replicate-to-replicate correlation for single-dataset model}.
+Comparison of inferred relative fitness values for mutant strains between
+experimental replicates when fitting each replicate individually for all
+environments shown in Figure~\ref{fig-SIX_kinsler_data}. Gray error bars show
+the 68\% credible region for this inference.
+
+\textbf{Figure S10. Comparing the replicate-to-replicate deviations with
+within-batch noise estimates for single dataset analysis}. Empirical cumulative
+distribution functions of the distance between the mean fitness estimate for
+each replicate (blue lines) and each replicate's posterior distribution standard
+deviation (orange and green line).
+
+\textbf{Figure S11. Posterior predictive checks for Kinsler et al., 2020
+hierarchical model}. Examples of the posterior predictive checks for all neutral
+lineages (upper left panels) and a subset of representative mutant lineages for
+all four environments shown in Figure~\ref{fig-SIX_kinsler_data}. Shaded regions
+from light to dark represent the 95\%, 68\%, and 5\% credible regions. Listed
+fitness values for mutant lineages represent the 68\% credible region.
+
+\textbf{Figure S12. Replicate-to-replicate correlation for hierarchical model}.
+Comparison of inferred fitness values between experimental replicates when
+fitting a hierarchical model on the biological replicates for all environments
+shown in Figure~\ref{fig-SIX_kinsler_data}. Gray error bars show the 68\%
+credible region for this inference.
+
+\textbf{Figure S13. Replicate-to-replicate correlation for both methods}.
+Comparison of \textbf{mean} inferred fitness values between experimental
+replicates when fitting a single-dataset model (blue points) vs.~a hierarchical
+model on the biological replicates (orange points) for all environments shown in
+Figure~\ref{fig-SIX_kinsler_data}. Error bars are not included in the plot for
+visual clarity.
+
+\textbf{Figure S14. Comparing the replicate-to-replicate deviations with
+within-batch noise estimates for hierarchical model analysis}. Empirical
+cumulative distribution functions of the distance between the mean fitness
+estimate for each replicate (blue lines) and each replicate's posterior
+distribution standard deviation (orange and green line).
+
+\textbf{Figure S15. Comparing estimates uncertainties between single-dataset and
+hierarchical model}. Comparison of the reported standard deviations for the
+mutants' relative fitness when inferring the relative fitness using one dataset
+at a time (\(x\)-axis) and the corresponding standard deviation for the mutants'
+hyper-fitness parameter when utilizing both replicates simultaneously via a
+hierarchical model (\(y\)-axis). Dashed line shows identity line.
+
+\textbf{Figure S16. Scaling of computation time as a function of number of
+unique barcodes}. Fitting a single-dataset model for 500 to 100,000 unique
+barcodes by taking 7,500 steps.
+
 \clearpage
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 % SUPPLEMENTAL MATERIAL
@@
 \beginsupplement
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% Add table of contents
+\tableofcontents
 
@@
 \section{Supplementary Materials}\label{supplementary-materials}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Add table of contents
-\tableofcontents
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \(q_\phi^*\) such that
-\begin{equation}\protect\hypertarget{eq-vi_objective}{}{
+\begin{equation}\protect\hypertarget{eq-vi_objective_si}{}{
 q_\phi^*(\underline{\theta}) =
@@
 \right),
-}\label{eq-vi_objective}\end{equation} where \(D_{KL}\) is the KL
+}\label{eq-vi_objective_si}\end{equation} where \(D_{KL}\) is the KL
 divergence. Furthermore, we highlight that the KL divergence is a
@@
 
-At first sight, Equation~\ref{eq-vi_objective} does not improve the
+At first sight, Equation~\ref{eq-vi_objective_si} does not improve the
 situation but only introduces further technical complications. After
@@
 fitness values depend on the global hyper-fitness value, obtaining
-\begin{equation}\protect\hypertarget{eq-hier_bayes_exp_indep}{}{
+\begin{equation}\protect\hypertarget{eq-hier_bayes_exp_indep_02}{}{
 \pi(
@@
 ).
-}\label{eq-hier_bayes_exp_indep}\end{equation} Notice we removed the
+}\label{eq-hier_bayes_exp_indep_02}\end{equation} Notice we removed the
 conditioning on the population mean fitness as our prior expectations of
@@
 The first term on the right-hand side of
-Equation~\ref{eq-hier_bayes_exp_indep} takes the same functional form as
-the one derived in Section~\ref{sec-bayes_mutfit}. Therefore, all we are
-left with is to determine the functional forms for the hyper--prior
+Equation~\ref{eq-hier_bayes_exp_indep_02} takes the same functional form
+as the one derived in Section~\ref{sec-bayes_mutfit}. Therefore, all we
+are left with is to determine the functional forms for the hyper--prior
 \(\pi(\underline{\theta}^M)\), and the conditional probability
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% HERE IS WHERE THE SI TEXT GOES
+\hypertarget{simulation-details}{%
+\subsubsection{Simulation details}\label{simulation-details}}
+
+This section lists the parameters used to generate the synthetic data
+used throughout the main text. All the code to reproduce every
+simulation and figure in the paper can be found in
+\href{https://github.com/mrazomej/bayesian_fitness/tree/master}{the
+GitHub repository}. Unless stated otherwise, the simulations were
+initialized following these steps:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  The initial number of cells were defined as:
+\end{enumerate}
+
+\begin{itemize}
+\tightlist
+\item
+  Ancestor lineage: \texttt{n\_init} \(\times\) \texttt{frac\_anc}.
+\item
+  Neutral lineages:
+  \(n_0^{(n)} \sim \text{Poisson} \left(  \texttt{n\_init} \times \frac{\texttt{frac\_neutral}}{\texttt{n\_neutral}} \right)\)
+\item
+  Mutant lineages:
+  \(n_0^{(m)} \sim \text{LogNormal} \left(  \mu = \log(\texttt{n\_init} \times \frac{\texttt{frac\_mut}}{\texttt{n\_mut}}),  \sigma = 2 \right)\)
+\end{itemize}
+
+where
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{n\_init} is the initial number of cells.
+\item
+  \texttt{frac\_anc} \(\in [0, 1]\) is the initial fraction of the
+  culture that is ancestor.
+\item
+  \texttt{frac\_neutral} \(\in [0, 1]\) is the initial fraction of the
+  culture that is barcoded neutral lineages.
+\item
+  \texttt{frac\_mut} \(\in [0, 1]\) is the initial fraction of the
+  culture that is barcoded lineages whose relative fitness we want to
+  infer--hereafter ``mutat'' barcodes.
+\item
+  \texttt{frac\_anc} + \texttt{frac\_neutral} + \texttt{frac\_mut} = 1
+\item
+  \texttt{n\_neutral} is the number of unique neutral lineage barcodes.
+\item
+  \texttt{n\_mut} is the number of unique mutant barcodes.
+\end{itemize}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{1}
+\tightlist
+\item
+  The distribution of growth rates \(\lambda\) were defined as
+\end{enumerate}
+
+\begin{itemize}
+\tightlist
+\item
+  Ancestor lineage \& neutral lineages: \(\lambda_a =\) \texttt{λ\_a}.
+  See details below for specific values for each simulation.
+\item
+  Mutant lineages:
+  \(\lambda_m \sim \text{truncated}-\mathcal{N}(  \mu =\)
+  \texttt{λ\_mut\_mean}, \(\sigma =\) \texttt{λ\_bc\_std},
+  \(\text{truncation limits} =\) \texttt{λ\_trunc} \()\). See details
+  below for specific values for each simulation.
+\end{itemize}
+
+\hypertarget{fig-2.}{%
+\paragraph{Fig 2.}\label{fig-2.}}
+
+The following is a \texttt{README.md} file automatically generated when
+simulating the data for Fig. 2 in the main text:
+
+\begin{itemize}
+\tightlist
+\item
+  File location
+  \href{https://github.com/mrazomej/bayesian_fitness/blob/master/code/processing/data001_logistic_250bc_01env_01rep/sim_data001.jl}{\texttt{link\ to\ file}}
+\item
+  Number of mutant barcodes \texttt{n\_mut\ =\ 225}
+\item
+  Number of neutral barcodes \texttt{n\_neutral\ =\ 25}
+\item
+  Ancestral strain growth rate \texttt{λ\_a\ =\ 1.0}
+\item
+  Carrying capacity \texttt{κ\ =\ 1.0e10}
+\item
+  Number of generations \texttt{n\_gen\ =\ 8}
+\item
+  Initial number of cells
+  \texttt{n\_init\ =\ κ\ /\ (2\^{}(n\_gen))\ =\ 3.90625e7}
+\item
+  Initial fraction of culture that is ancestor
+  \texttt{frac\_anc\ =\ 0.93}
+\item
+  Initial fraction of culture that is neutrals
+  \texttt{frac\_neutral\ =\ 0.02}
+\item
+  Initial fraction of culture that is mutants
+  \texttt{frac\_mut\ =\ 0.04999999999999995}
+\item
+  Mutant fitness distribution mean
+  \texttt{λ\_mut\_mean\ =\ λ\_a\ *\ 1.005\ =\ 1.005}
+\item
+  Mutant fitness distribution standard deviation to sample growth rates
+  \texttt{λ\_mut\_std\ =\ 0.1}
+\item
+  Mutant fitness distribution truncation ranges for growth rates
+  \texttt{λ\_trunc\ =\ {[}λ\_a\ .*\ 0.9999,\ λ\_a\ *\ 1.5{]}\ =\ {[}0.9999,\ 1.5{]}}
+\item
+  Gaussian noise distribution standard deviation
+  \texttt{σ\_lognormal\ =\ 0.3}
+\end{itemize}
+
+\hypertarget{fig-3.}{%
+\paragraph{Fig 3.}\label{fig-3.}}
+
+The following is a \texttt{README.md} file automatically generated when
+simulating the data for Fig. 3 in the main text:
+
+\begin{itemize}
+\tightlist
+\item
+  File location
+  \href{https://github.com/mrazomej/bayesian_fitness/blob/master/code/processing/data003_logistic_1000bc_03env_01rep/sim_multienv_data003.jl}{\texttt{link\ to\ file}}
+\item
+  Number of environments \texttt{n\_env\ =\ 3}
+\item
+  Environment transitions
+  \texttt{env\_idx\ =\ {[}1,\ 2,\ 3,\ 1,\ 2,\ 3{]}}
+\item
+  Number of mutant barcodes \texttt{n\_mut\ =\ 900}
+\item
+  Number of neutral barcodes \texttt{n\_neutral\ =\ 100}
+\item
+  Ancestral strain growth rate on each environment
+  \texttt{λ\_a\ =\ {[}1.0,\ 2.0,\ 0.5{]}}
+\item
+  Carrying capacity \texttt{κ\ =\ 1.0e10}
+\item
+  Number of generations \texttt{n\_gen\ =\ 8}
+\item
+  Initial number of cells
+  \texttt{n\_init\ =\ κ\ /\ (2\^{}(n\_gen))\ =\ 3.90625e7}
+\item
+  Initial fraction of culture that is ancestor
+  \texttt{frac\_anc\ =\ 0.93}
+\item
+  Initial fraction of culture that is neutrals
+  \texttt{frac\_neutral\ =\ 0.02}
+\item
+  Initial fraction of culture that is mutants
+  \texttt{frac\_mut\ =\ 0.04999999999999995}
+\item
+  Mutant fitness distribution mean `λ\_mut\_mean = λ\_a * 1.005 =
+  {[}1.005, 2.01, 0.5025{]}
+\item
+  Mutant fitness distribution standard deviation to sample growth rates
+  \texttt{λ\_mut\_std\ =\ 0.1}
+\item
+  Mutant fitness distribution truncation ranges for growth rates
+  \texttt{λ\_trunc\ =\ {[}λ\_a\ .*\ 0.9995,\ λ\_a\ .*\ 1.5{]}\ =\ {[}{[}{[}0.995,\ 1.99,\ 0.4975{]}{]},\ {[}{[}1.5,\ 3.0,\ 0.75{]}{]}{]}}
+\item
+  Gaussian noise distribution standard deviation
+  \texttt{σ\_lognormal\ =\ 0.125}
+\end{itemize}
+
+The variable \texttt{env\_idx} lists the order of the environment cycles
+used in the simulation. In this case, transitions were of the form
+\texttt{1\ -\textgreater{}\ 2}, \texttt{2\ -\textgreater{}\ 3}, and
+\texttt{3\ -\textgreater{}\ 1}.
+
+\hypertarget{fig-4.}{%
+\paragraph{Fig 4.}\label{fig-4.}}
+
+The following is a \texttt{README.md} file automatically generated when
+simulating the data for Fig. 4 in the main text:
+
+\begin{itemize}
+\tightlist
+\item
+  File location
+  \href{https://github.com/mrazomej/bayesian_fitness/blob/master/code/processing/data002_logistic_1000bc_01env_02rep/sim_hierarchical_data002.jl}{\texttt{link\ to\ file}}
+\item
+  Number of experimental replicates \texttt{n\_data\ =\ 2}
+\item
+  Number of mutant barcodes \texttt{n\_mut\ =\ 900}
+\item
+  Number of neutral barcodes \texttt{n\_neutral\ =\ 100}
+\item
+  Ancestral strain growth rate \texttt{λ\_a\ =\ 1.0}
+\item
+  Carrying capacity \texttt{κ\ =\ 1.0e10}
+\item
+  Number of generations \texttt{n\_gen\ =\ 8}
+\item
+  Initial number of cells
+  \texttt{n\_init\ =\ κ\ /\ (2\^{}(n\_gen))\ =\ 3.90625e7}
+\item
+  Initial fraction of culture that is ancestor
+  \texttt{frac\_anc\ =\ 0.93}
+\item
+  Initial fraction of culture that is neutrals
+  \texttt{frac\_neutral\ =\ 0.02}
+\item
+  Initial fraction of culture that is mutants
+  \texttt{frac\_mut\ =\ 0.04999999999999995}
+\item
+  Mutant fitness distribution mean
+  \texttt{λ\_bc\_mean\ =\ λ\_a\ *\ 1.005\ =\ 1.005}
+\item
+  Mutant fitness distribution standard deviation to sample growth rates
+  \texttt{λ\_bc\_std\ =\ 0.1}
+\item
+  Mutant fitness distribution truncation ranges for growth rates
+  \texttt{λ\_trunc\ =\ {[}λ\_a\ .*\ 0.9999,\ λ\_a\ *\ 1.5{]}\ =\ {[}0.9999,\ 1.5{]}}
+\item
+  Gaussian noise distribution standard deviation
+  \texttt{σ\_lognormal\ =\ 0.3}
+\item
+  Standard deviation for between-experiment fitness distribution
+  variability \texttt{σ\_exp\ =\ 0.015}
+\end{itemize}
+
+The variable \texttt{σ\_exp} defines the variance used to sample the
+variability between simulated experimental replicates. In other words,
+given a ``hyper-fitness'' value \(\lambda\), for each experimental
+replicate, the growth rate was defined as
+\(\lambda_j \sim \mathcal{N}(\lambda,\) \texttt{σ\_exp} \()\).
+
+\hypertarget{fig-6.}{%
+\paragraph{Fig 6.}\label{fig-6.}}
+
+The following is a \texttt{README.md} file automatically generated when
+simulating the data for Fig. 6 in the main text:
+
+\begin{itemize}
+\tightlist
+\item
+  File location
+  \href{https://github.com/mrazomej/bayesian_fitness/blob/master/code/processing/data007_logistic_hierarchicalgenotype_1000bc_01env_01rep/sim_data007.jl}{\texttt{link\ to\ file}}
+\item
+  Number of mutant barcodes \texttt{n\_mut\ =\ 900}
+\item
+  Number of neutral barcodes \texttt{n\_neutral\ =\ 100}
+\item
+  Ancestral strain growth rate \texttt{λ\_a\ =\ 1.0}
+\item
+  Carrying capacity \texttt{κ\ =\ 1.0e10}
+\item
+  Number of generations \texttt{n\_gen\ =\ 8}
+\item
+  Initial number of cells
+  \texttt{n\_init\ =\ κ\ /\ (2\^{}(n\_gen))\ =\ 3.90625e7}
+\item
+  Initial fraction of culture that is ancestor
+  \texttt{frac\_anc\ =\ 0.93}
+\item
+  Initial fraction of culture that is neutrals
+  \texttt{frac\_neutral\ =\ 0.02}
+\item
+  Initial fraction of culture that is mutants
+  \texttt{frac\_mut\ =\ 0.04999999999999995}
+\item
+  Mutant fitness distribution mean
+  \texttt{λ\_bc\_mean\ =\ λ\_a\ *\ 1.005\ =\ 1.005}
+\item
+  Mutant fitness distribution standard deviation to sample growth rates
+  \texttt{λ\_bc\_std\ =\ 0.1}
+\item
+  Mutant fitness distribution truncation ranges for growth rates
+  \texttt{λ\_trunc\ =\ {[}λ\_a\ .*\ 0.9999,\ λ\_a\ *\ 1.5{]}\ =\ {[}0.9999,\ 1.5{]}}
+\item
+  Gaussian noise distribution standard deviation
+  \texttt{σ\_lognormal\ =\ 0.3}
+\item
+  Number of unique genotypes
+  \texttt{n\_geno\ =\ Int64(n\_mut\ /\ 10)\ =\ 90}
+\end{itemize}
+
+The variable \texttt{n\_geno} defines the number of unique genotypes in
+the simulation. In this simulation, we aimed to have ten unique barcodes
+on average for each genotype. The actual number of genotypes per barcode
+was sampled from a multinomial distribution with probability
+\texttt{1\ /\ n\_geno} for each genotype.
+
+\hypertarget{sec-fitseq2}{%
+\subsection{\texorpdfstring{Comparison with
+\texttt{FitSeq2.0}}{Comparison with FitSeq2.0}}\label{sec-fitseq2}}
+
+For a complete comparison, we apply our inference pipeline to a
+simulated dataset closely following the simulation pipeline from
+\textcite{li2023}. This simulation explicitly accounts for multiple
+noise sources captured in the added Gaussian noise in
+Section~\ref{sec-logistic}. Here, we list all the steps involved in this
+explicit simulation, highlighting that the overall results are
+qualitatively very similar to that of our simple logistic growth plus
+added Gaussian noise when computing the relevant quantity of the log
+frequency ratio.
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  We initialized a population consisting of 1000 unique barcodes---50
+  neutral barcodes and 950 mutant barcodes---where the initial number of
+  cells with barcode \(i\) in the culture was sampled from a Gamma
+  distribution as
+  \begin{equation}\protect\hypertarget{eq-fitseq_gamma_pop}{}{
+  n_o^{(i)} \sim \text{Gamma}(20, 0.2)
+  }\label{eq-fitseq_gamma_pop}\end{equation}
+\item
+  The Malthusian fitness (related but not exactly the relative fitness
+  to be inferred from the data) for non-neutral barcodes was then
+  sampled from a skewed normal distribution of the form
+  \begin{equation}\protect\hypertarget{eq-fitseq_skewnormal}{}{
+  \lambda^{(j)} \sim \text{Skew}\mathcal{N}(0, 0.225, 3)
+  }\label{eq-fitseq_skewnormal}\end{equation}
+\item
+  To simulate the noisy growth, the number of cells underwent the
+  following steps:
+\end{enumerate}
+
+\textbf{for} \(t=2:T\) \textbf{do}
+
+~~~~
+\(n_{t,c}^{(i)} = \text{Poisson}(n_{t-1,c}^{(i)} e^{\lambda^{(i)}})\)
+
+\textbf{end for},
+
+where \(T\) is the number of generations, and \(c\) indexes the
+growth-dilution cycle. We use \(T=4\) for our simulation.
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{3}
+\item
+  After the \(T\) generations, cells were stochastically diluted back
+  for the next growth-dilution cycle as
+  \begin{equation}\protect\hypertarget{eq-fitseq_dilution}{}{
+  n_{0,c+1}^{(i)} \sim \text{Poisson}
+  \left(\frac{n_{T,c}^{(i)}}{2^{T}}\right).
+  }\label{eq-fitseq_dilution}\end{equation} This process was repeated
+  \(C=4\) times for a total of four growth-dilution cycles.
+\item
+  To simulate the sampling noise when taking cells that will then be
+  subsequently processed to obtain barcode reads, we sampled
+  \(n_{\text{sample},c}^{(i)}\) cells with barcode \(i\) as
+  \begin{equation}\protect\hypertarget{eq-fitseq_sample}{}{
+  n_{\text{sample},c}^{(i)} \sim 
+  \text{Poisson}(L \times N \times f_{c}^{(i)}),
+  }\label{eq-fitseq_sample}\end{equation} where \(c\) indexes the growth
+  cycle, \(N\) is the total number of unique barcodes, \(f_{j}^{(i)}\)
+  is the fraction of cells in the culture with barcode \(i\) at the end
+  of cycle \(j\), i.e.,
+  \begin{equation}\protect\hypertarget{eq-fitseq_frac_cell}{}{
+  f_{c}^{(i)} = \frac{n_{c}^{(i)}}
+  {\sum_{j=1}^N n_{c}^{(j)}},
+  }\label{eq-fitseq_frac_cell}\end{equation} and \(L\) is the expected
+  average number of cells per lineage. We used \(L = 500\).
+\item
+  To simulate the noise due to PCR amplification, we iterate
+  \(c_{\text{PCR}} = 25\) times, stochastically doubling the number of
+  barcodes with a Poisson sample of the form
+\end{enumerate}
+
+\textbf{for} \(a=1:A\) \textbf{do}
+
+~~~~
+\(n_{\text{PCR},(a+1),c}^{(i)} \sim \text{Poisson}(2 n_{\text{PCR},(a),c}^{(i)})\)
+
+\textbf{end for},
+
+where \((a)\) indexes the amplification cycle, \(A\) is the total number
+of cycles, and \(c\) indexes the growth-dilution cycle. We use \(A=25\)
+for our simulation. For \(a = 1\), we use the number of cells sampled in
+Equation~\ref{eq-fitseq_sample}.
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{6}
+\tightlist
+\item
+  To simulate the noise due to the sampling step involved during the
+  sequencing of the amplicon library, the number of reads mapping to
+  barcode \(i\) at the end of cycle \(c\) used in the inference was
+  sampled as \begin{equation}\protect\hypertarget{eq-fitseq_seq}{}{
+  n_{\text{seq},c}^{(i)} \sim 
+  \text{Poisson}(R \times N \times f_{\text{PCR},c}^{(i)}),
+  }\label{eq-fitseq_seq}\end{equation} where \(R\) is the expected
+  number of reads per barcode (R=100 for our simulation), \(N\) is the
+  number of total barcodes, and \(f_{\text{PCR},c}^{(i)}\) is the
+  fraction of barcodes of type \(i\) in the final amplification cycle
+  from step 4.
+\end{enumerate}
+
+Figure~\ref{fig-SIX_fitseq2}(A) and (B) show the simulation's frequency
+and log frequency ratio trajectories, respectively. Qualitatively, these
+results are very similar to those shown in Fig. 1 of \textcite{li2023}.
+As done throughout the main text, Figure~\ref{fig-SIX_fitseq2}(C) shows
+examples of the posterior predictive checks for the neutral lineages and
+a few of the mutant lineages. We can see that our inference pipeline can
+fit this simulated data for the examples shown. Furthermore,
+Figure~\ref{fig-SIX_fitseq2}(D) compares the ground truth fitness values
+and the inferred parameters for all mutant lineages.
+Figure~\ref{fig-SIX_fitseq2} shows the comparison between the inference
+with the software provided in \autocite{li2023} and our inference
+pipeline. We note that although the values are highly correlated, the
+scale of the axis is very different. This scale difference comes from
+the differences in models used to infer the relative fitness value.
+While our model assumes an exponential change in the relative frequency
+of the barcode, \textcite{li2023} works directly with the number of
+cells rather than the relative frequency, assuming a Malthusian fitness
+form. In other words, while our model is agnostic to the underlying type
+of growth mechanism, as long as the condition of exponential change in
+relative frequency is satisfied, \textcite{li2023} assumes exponential
+growth of the number of cells, resulting in a discrepancy between the
+obtained numbers.
+
+\begin{figure}
+
+{\centering \includegraphics{./figs/figSIX_fitseq2.pdf}
+
+}
+
+\caption{\label{fig-SIX_fitseq2}\textbf{FitSeq2.0 simulation
+comparison}.(A) Frequency trajectories that represent the raw data going
+into the inference. (B) Log frequency ratio between two adjacent time
+points used by the inference pipeline. Darker lines represent the
+neutral barcodes. These transformed data are much more noisy than the
+seemingly smooth frequency trajectories. (C) Examples of the posterior
+predictive checks for all neutral lineages (upper left panel) and a
+subset of representative mutant lineages. Shaded regions represent the
+95\%, 68\%, and 5\% credible regions for the data. The reported errors
+above the plot represent the 68\% credible region on the mutant relative
+fitness marginal distribution. (D) Comparison between the ground truth
+fitness value from the logistic-growth simulation and the inferred
+fitness value. Gray error bars represent the 68\% posterior credible
+region for the relative fitness values. (E) Comparison between fitness
+inferences as obtained using \texttt{FitSeq2.0} from \textcite{li2023}
+and our Bayesian method. Vertical gray error bars represent the 68\%
+posterior credible region for the relative fitness values.}
+
+\end{figure}
+
+\hypertarget{sec-kinsler}{%
+\subsection{Reanalyzing Kinsler et al., 2020}\label{sec-kinsler}}
+
+The main text of this paper focuses only on simulation-based assessments
+of the performance of our inference algorithm. This supplementary
+section shows how the model performs on experimental data. In
+particular, we focus on a subset of the data generated for
+\autocite{kinsler2020}. In this paper, the fitness of a set of mutants
+originally evolved in minimal glucose media was determined in multiple
+environments using an experimental design described in
+Section~\ref{sec-experiment}. Figure~\ref{fig-SIX_kinsler_data} shows
+the raw barcode frequency trajectories for four of such environments
+measured in two biological replicates.
+
+\begin{figure}
+
+{\centering \includegraphics{./figs/figSIX_kinsler_data.pdf}
+
+}
+
+\caption{\label{fig-SIX_kinsler_data}\textbf{Kinsler et al., 2020 sample
+data}. Barcode frequency trajectories in four different environments for
+mutants evolved in minimal glucose media. Each environment was measured
+in two biological replicates (R1 and R2 on each panel). For experimental
+details, see \autocite{kinsler2020}.}
+
+\end{figure}
+
+\hypertarget{analysis-of-individual-replicates}{%
+\subsubsection{Analysis of individual
+replicates}\label{analysis-of-individual-replicates}}
+
+As a first analysis, we perform inference assuming all replicates are
+independent. Figure~\ref{fig-SIX_kinsler_ppc_single} shows the posterior
+predictive checks for the neutral lineages (upper left plot on each
+panel) and some sample mutants for both replicates. We can see that the
+inference works exceptionally well with the data, capturing the
+variability. Although the main text inferences were made on logistic
+growth simulations, as described in Section~\ref{sec-logistic}, the
+results in Section~\ref{sec-fitseq2} and this section show that our
+model does not depend on the assumptions underlying such simulations.
+The only assumption that must be valid is that the changes in the
+relative frequencies of each lineage follow an exponential function.
+
+\begin{figure}
+
+{\centering \includegraphics{./figs/figSIX_kinsler_ppc_single.pdf}
+
+}
+
+\caption{\label{fig-SIX_kinsler_ppc_single}\textbf{Posterior predictive
+checks for Kinsler et al., 2020 independent replicates}. Examples of the
+posterior predictive checks for all neutral lineages (upper left panels)
+and a subset of representative mutant lineages for all four environments
+shown in Figure~\ref{fig-SIX_kinsler_data}. Shaded regions from light to
+dark represent the 95\%, 68\%, and 5\% credible regions. Listed fitness
+values for mutant lineages represent the 68\% credible region.}
+
+\end{figure}
+
+Furthermore, we can compare the replicate-to-replicate correlation for
+these four environments with two biological replicates. The result of
+this comparison is shown in Figure~\ref{fig-SIX_kinsler_rep_single}. We
+can see that there is a good degree of correlation between replicates.
+
+\begin{figure}
+
+{\centering \includegraphics{./figs/figSIX_kinsler_rep_single.pdf}
+
+}
+
+\caption{\label{fig-SIX_kinsler_rep_single}\textbf{Replicate-to-replicate
+correlation for single-dataset model}. Comparison of inferred relative
+fitness values for mutant strains between experimental replicates when
+fitting each replicate individually for all environments shown in
+Figure~\ref{fig-SIX_kinsler_data}. Gray error bars show the 68\%
+credible region for this inference.}
+
+\end{figure}
+
+Given that we have two biological replicates of the same experiment, we
+can ask how the difference between each replicate's inferred fitness
+values compares with the individual estimates' uncertainty. In other
+words, we want to compare how different the \(x\)- and \(y\)-axis values
+in Figure~\ref{fig-SIX_kinsler_rep_single} are with the size of the
+shown error bars. Figure~\ref{fig-SIX_kinsler_meandiff_single} compares
+the distribution of these magnitudes for all four environments. In each
+plot, the blue line shows the empirical cumulative distribution function
+(ECDF) of the distance between the mean fitness estimate for each
+replicate---computed as
+\(\sqrt{\left( \left\langle s_1^{(m)}\right\rangle - \left\langle s_2^{(m)}\right\rangle \right)^2}\),
+where \(\langle s_i^{(m)}\rangle\) is the mean fitness estimate of
+mutant \(m\) in replicate \(i\)---meaning the difference between \(x\)-
+and \(y\)-axis in Figure~\ref{fig-SIX_kinsler_rep_single}. The orange
+and green lines represent the ECDF for each replicate's posterior
+distribution standard deviation, i.e., the ECDF of the size of the error
+bars in Figure~\ref{fig-SIX_kinsler_rep_single}. From these plots, we
+can see that the distribution of standard deviations is shifted to the
+left compared to the difference between replicates. This implies that
+the uncertainty in the estimate is generally larger than the difference
+between replicates one would obtain with a naive point estimate.
+
+\begin{figure}
+
+{\centering \includegraphics{./figs/figSIX_kinsler_meandiff_single.pdf}
+
+}
+
+\caption{\label{fig-SIX_kinsler_meandiff_single}\textbf{Comparing the
+replicate-to-replicate deviations with within-batch noise estimates for
+single dataset analysis}. Empirical cumulative distribution functions of
+the distance between the mean fitness estimate for each replicate (blue
+lines) and each replicate's posterior distribution standard deviation
+(orange and green line).}
+
+\end{figure}
+
+\hypertarget{hierarchical-analysis-of-multiple-replicates}{%
+\subsubsection{Hierarchical analysis of multiple
+replicates}\label{hierarchical-analysis-of-multiple-replicates}}
+
+As discussed in Section~\ref{sec-replicates}, most experimental
+designs---including that of \textcite{kinsler2020}---include two or more
+biological replicate measurements. Therefore, a natural way to analyze
+the data is by jointly inferring the fitness value using a hierarchical
+model. Figure~\ref{fig-SIX_kinsler_ppc_hier} shows the posterior
+predictive checks when fitting a hierarchical model to each of the four
+environments shown in Figure~\ref{fig-SIX_kinsler_data}. Comparing this
+figure with Figure~\ref{fig-SIX_kinsler_ppc_single} shows that the
+hierarchical model does not lose the ability to capture the variability
+in the data compared to fitting each replicate individually. However, as
+shown in Figure~\ref{fig-SIX_kinsler_rep_hier}, the hierarchical model
+does a better job when comparing the replicate-to-replicate correlation,
+as expected. To emphasize this last point,
+Figure~\ref{fig-SIX_kinsler_rep_comb} shows both inferences on the same
+plot---only the mean of the posterior distribution is shown for visual
+clarity.
+
+\begin{figure}
+
+{\centering \includegraphics{./figs/figSIX_kinsler_ppc_hier.pdf}
+
+}
+
+\caption{\label{fig-SIX_kinsler_ppc_hier}\textbf{Posterior predictive
+checks for Kinsler et al., 2020 hierarchical model}. Examples of the
+posterior predictive checks for all neutral lineages (upper left panels)
+and a subset of representative mutant lineages for all four environments
+shown in Figure~\ref{fig-SIX_kinsler_data}. Shaded regions from light to
+dark represent the 95\%, 68\%, and 5\% credible regions. Listed fitness
+values for mutant lineages represent the 68\% credible region.}
+
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics{./figs/figSIX_kinsler_rep_hier.pdf}
+
+}
+
+\caption{\label{fig-SIX_kinsler_rep_hier}\textbf{Replicate-to-replicate
+correlation for hierarchical model}. Comparison of inferred fitness
+values between experimental replicates when fitting a hierarchical model
+on the biological replicates for all environments shown in
+Figure~\ref{fig-SIX_kinsler_data}. Gray error bars show the 68\%
+credible region for this inference.}
+
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics{./figs/figSIX_kinsler_rep_comb.pdf}
+
+}
+
+\caption{\label{fig-SIX_kinsler_rep_comb}\textbf{Replicate-to-replicate
+correlation for both methods}. Comparison of \textbf{mean} inferred
+fitness values between experimental replicates when fitting a
+single-dataset model (blue points) vs.~a hierarchical model on the
+biological replicates (orange points) for all environments shown in
+Figure~\ref{fig-SIX_kinsler_data}. Error bars are not included in the
+plot for visual clarity}
+
+\end{figure}
+
+We can again compare the magnitude of the difference between replicates'
+inferred mean fitness with the size of the corresponding error bars, as
+we did with Figure~\ref{fig-SIX_kinsler_meandiff_single}.
+Figure~\ref{fig-SIX_kinsler_meandiff_hier} shows the ECDF for the
+resulting difference between replicates mean fitness in blue as well as
+the resulting standard deviation for the inferred hyper-fitness
+parameter that connects both replicates in orange (see
+Section~\ref{sec-replicates} for description of this hyper-fitness
+parameter). Here, we see that the standard deviation ECDF (orange line)
+is shifted more to the right than the mean difference between replicates
+(blue line).
+
+\begin{figure}
+
+{\centering \includegraphics{./figs/figSIX_kinsler_meandiff_hier.pdf}
+
+}
+
+\caption{\label{fig-SIX_kinsler_meandiff_hier}\textbf{Comparing the
+replicate-to-replicate deviations with within-batch noise estimates for
+hierarchical model analysis}. Empirical cumulative distribution
+functions of the distance between the mean fitness estimate for each
+replicate (blue lines) and each replicate's posterior distribution
+standard deviation (orange and green line).}
+
+\end{figure}
+
+Even though the difference between replicates is reduced when using a
+hierarchical model, this does not come at the expense of higher
+uncertainty in the estimate of the relative fitness. To show this is the
+case, Figure~\ref{fig-SIX_kinsler_std_comp} compares the reported
+posterior distribution standard deviations for the relative fitness when
+inferring each dataset individually (\(x\)-axis) with the standard
+deviation of the hyper-fitness parameter (\(y\)-axis). Most of the
+points fall below the identity line, demonstrating that the uncertainty
+in the hyper-fitness parameter obtained by utilizing information from
+both replicates is smaller than the equivalent uncertainty obtained with
+one replicate at a time. In conclusion, the hierarchical model not only
+can equally describe the data, as shown in
+Figure~\ref{fig-SIX_kinsler_ppc_hier}, but the difference between
+replicates is significantly decreased, as shown in
+Figure~\ref{fig-SIX_kinsler_rep_comb}, at the same time, the uncertainty
+in the parameters' estimate is decreased, as shown in
+Figure~\ref{fig-SIX_kinsler_std_comp}.
+
+\begin{figure}
+
+{\centering \includegraphics{./figs/figSIX_kinsler_std_comp.pdf}
+
+}
+
+\caption{\label{fig-SIX_kinsler_std_comp}\textbf{Comparing estimates
+uncertainties between single-dataset and hierarchical model}. Comparison
+of the reported standard deviations for the mutants' relative fitness
+when inferring the relative fitness using one dataset at a time
+(\(x\)-axis) and the corresponding standard deviation for the mutants'
+hyper-fitness parameter when utilizing both replicates simultaneously
+via a hierarchical model (\(y\)-axis). Dashed line shows identity line.}
+
+\end{figure}
+
+\hypertarget{sec-scaling}{%
+\subsection{Computation time scaling}\label{sec-scaling}}
+
+The biggest advantage of the variational inference approach presented in
+this work is the computation speedup compared to MCMC-based Bayesian
+approaches. This allows for applying our principled Bayesian pipeline to
+datasets with many unique barcodes or multiple datasets analyzed
+hierarchically. Figure~\ref{fig-SIX_bc_vs_time} shows a linear scaling
+in the amount of time taken for the ADVI algorithm to take 7,500 steps
+for a varying number of unique barcodes. This figure shows that the time
+to fit larger models increases linearly with the number of barcodes.
+
+\begin{figure}
+
+{\centering \includegraphics{./figs/figSIX_bc_vs_time.pdf}
+
+}
+
+\caption{\label{fig-SIX_bc_vs_time}\textbf{Scaling of computation time
+as a function of number of unique barcodes}. Fitting a single-dataset
+model for 500 to 100,000 unique barcodes by taking 7,500 steps.}
+
+\end{figure}
+
 % Print supplemental references changing the title
@@
 \end{refsegment}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2024-02-26 code/figs/web01.jl
update figures

@@ -41,7 +41,7 @@ data = CSV.read(
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 
 # Initialize figure
-fig = Figure(resolution=(350, 300))
+fig = Figure(size=(350, 300))
 
 # Add axis
 ax = Axis(

### 2024-02-26 code/figs/web02.jl

@@ -41,7 +41,7 @@ data = CSV.read(
 # %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
 
 # Initialize figure
-fig = Figure(resolution=(350, 300))
+fig = Figure(size=(350, 300))
 
 # Add axis
 ax = Axis(fig[1, 1], xlabel="time [dilution cycles]", ylabel="ln(fₜ₊₁/fₜ)")

### 2024-02-26 code/processing/data013_fullsim_1000bc_01env_01rep/viz_inference.jl
update visualization of simulation

@@ -1,8 +1,6 @@
 ##
 println("Loading packages...")
 
-
-
 # Import project package
 import BayesFitUtils
 

### 2024-02-26 code/processing/data015_fitseq2sim_1000bc_01env_01rep/fitseq2_inference.jl
update figures for fitseq comparison

@@ -1,0 +1,126 @@
+println("Loading packages...")
+
+
+
+# Import project package
+import BayesFitUtils
+
+# Import library package
+import BarBay
+
+# Import libraries to manipulate data
+import DataFrames as DF
+import CSV
+
+# Import library to list files
+import Glob
+
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
+# Define number of generations per cycle
+n_gen = 4
+
+# Define number of neutral lineages
+n_neutral = 50
+# Define total number of lineages
+n_lin = 1_000
+# Define number of mutants
+n_mut = n_lin - n_neutral
+
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Loading the data
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
+println("Loading data...")
+
+# Import data
+data = CSV.read(
+    "$(git_root())/data/logistic_growth/data_015/tidy_data.csv", DF.DataFrame
+)
+
+# Convert data to arrays for FitSeq2 -i input
+bc_seq_mat = BarBay.utils.data_to_arrays(data)[:bc_count]
+
+# Read table with number of cells
+bc_cell = Matrix(
+    CSV.read(
+        "./output/fitseq_sim_output_EvoSimulation_Bottleneck_Cell_Number.csv",
+        DF.DataFrame,
+        header=false
+    )
+)
+
+# Initialize empty dataframe for CSV FitSeq2 -t input
+bc_cell_mat = DF.DataFrame()
+# Sum number of cells pert time point
+bc_cell_mat[!, :count] = vec(sum(bc_cell, dims=1))
+
+# Transform first column to the number of generations in experiment
+bc_cell_mat[!, :time] = (collect(1:size(bc_cell_mat, 1)) .* n_gen) .- n_gen
+
[…]

### 2024-02-26 code/processing/data015_fitseq2sim_1000bc_01env_01rep/fitseq2_sim.jl

@@ -1,0 +1,192 @@
+println("Loading packages...")
+
+# Import project package
+import BayesFitUtils
+
+# Import libraries to manipulate data
+import DataFrames as DF
+import CSV
+
+# Import library to list files
+import Glob
+
+# Import basic statistical functions
+import StatsBase
+import Distributions
+import Random
+
+
+# Set random seed
+Random.seed!(42)
+## %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Define simulation metadata
+## %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
+# Define number of neutral lineages
+n_neutral = 50
+# Define total number of lineages
+n_lin = 1_000
+# Define number of mutants
+n_mut = n_lin - n_neutral
+
+# Define number of generations per cycle
+n_gen = 4
+
+# Define number of cycles
+n_growth_cycle = 5
+
+# Define average number of reads per barcode
+reads = 1_000
+
+# Define parameters for initial number of cells according to
+# n₀ ~ Gamma(α, β)
+α = 20
+β = 0.2
+
+# Define parameters for fitness distribution according to
+# λ ~ SkewNormal(µ, σ, skew)
+µ = 0.0
+σ = 0.225
+skew = 3
+
+## %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Generate output directory
+## %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
+# Generate output directory if it doesn't exist
+if !isdir("./output/")
+    mkdir("./output/")
+end # if
+
+## %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Generate `--t_seq` csv file
+## %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
+# Define output path
[…]

### 2024-02-26 code/processing/data015_fitseq2sim_1000bc_01env_01rep/viz_inference.jl

@@ -1,0 +1,417 @@
+##
+println("Loading packages...")
+
+# Import project package
+import BayesFitUtils
+
+# Import library package
+import BarBay
+
+# Import basic math
+import LinearAlgebra
+import StatsBase
+import Distributions
+import Random
+
+# Import libraries to manipulate data
+import DataFrames as DF
+import CSV
+
+# Import library to list files
+import Glob
+
+# Import plotting libraries
+using CairoMakie
+import ColorSchemes
+
+# Activate backend
+CairoMakie.activate!()
+
+# Set PBoC Plotting style
+BayesFitUtils.viz.theme_makie!()
+
+Random.seed!(42)
+
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Loading data
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
+# Generate output directory 
+if !isdir("./output/")
+    mkdir("./output/")
+end # if
+
+# Generate figure dictionary
+if !isdir("./output/figs/")
+    mkdir("./output/figs/")
+end # if
+
+println("Loading data...")
+
+# Import data
+data = CSV.read(
+    "$(git_root())/data/logistic_growth/data_015/tidy_data.csv", DF.DataFrame
+)
+
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+# Load variational inference
+# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% #
+
+println("Loading ADVI results...")
+# Define file
+file = first(Glob.glob("./output/advi_meanfield*5000*csv"))
+
+# Convert results to tidy dataframe
+df_advi = CSV.read(file, DF.DataFrame)
+
+# Define number of samples
+n_samples = 10_000
+
+# Sample from posterior MvNormal
+df_samples = DF.DataFrame(
+    Random.rand(
+        Distributions.MvNormal(
[…]

### 2024-02-26 doc/_author_summary.qmd
upload missing document modifications

--- 2023-10-11 doc/paper.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
-%
-\documentclass[
-  letterpaper,
-  DIV=11,
-  numbers=noendperiod]{scrartcl}
+---
+editor:
+    render-on-save: true
+bibliography: references.bib
+csl: ieee.csl
+---
 
-\usepackage{amsmath,amssymb}
-\usepackage{iftex}
-\ifPDFTeX
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-\usepackage{lmodern}
-\ifPDFTeX\else  
-    % xetex/luatex font selection
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-
-\usepackage{lineno}
-\linenumbers
-\usepackage[font={footnotesize, sf}, labelfont=bf, margin=0.05\linewidth]{caption}
-\renewcommand{\familydefault}{\sfdefault}
-% Bold the 'Figure #' in the caption and separate it from the title/caption
-% with a period
-% Captions will be left justified
-\usepackage[
-	aboveskip=1pt,
-	labelfont=bf,
-	labelsep=period,
-	justification=raggedright,
-	singlelinecheck=off
-]{caption}
-
-% Add numbered lines
-\usepackage{lineno}
-\linenumbers
-
-% Package to include multiple title pages
-% This will allow me to add a tile to the main text and to the SI
-\usepackage{titling}
-
-% Define command to begin the supplementary section
-\newcommand{\beginsupplement}{
-\setcounter{section}{0} % Restart section counter
-\renewcommand{\thesection}{S\arabic{section}}%
-\setcounter{table}{0} % Restart table counter
-\renewcommand{\thetable}{S\arabic{table}}%
-\setcounter{figure}{0} % Restart figure counter
-\renewcommand{\thefigure}{S\arabic{figure}}%
-\setcounter{equation}{0} % Restart equation counter
-\renewcommand{\theequation}{S\arabic{equation}}%
-}
-
-% Add author affiliations
-\usepackage{authblk}
-
-\KOMAoption{captions}{tableheading}
-\makeatletter
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\AtBeginDocument{%
-\ifdefined\contentsname
-  \renewcommand*\contentsname{Table of contents}
-\else
-  \newcommand\contentsname{Table of contents}
-\fi
-\ifdefined\listfigurename
-  \renewcommand*\listfigurename{List of Figures}
-\else
-  \newcommand\listfigurename{List of Figures}
-\fi
-\ifdefined\listtablename
-  \renewcommand*\listtablename{List of Tables}
-\else
-  \newcommand\listtablename{List of Tables}
-\fi
-\ifdefined\figurename
-  \renewcommand*\figurename{Figure}
-\else
-  \newcommand\figurename{Figure}
-\fi
-\ifdefined\tablename
-  \renewcommand*\tablename{Table}
-\else
-  \newcommand\tablename{Table}
-\fi
-}
-\@ifpackageloaded{float}{}{\usepackage{float}}
-\floatstyle{ruled}
-\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
-\floatname{codelisting}{Listing}
-\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
-\makeatother
-\makeatletter
-\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
-\makeatother
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-
-% References
-% \usepackage[]{biblatex}
-\usepackage[
-	backend=bibtex,
-	style=phys, 
-	sorting=none,				% Do not sort bibliography
-	url=false, 					% Do not show url in reference
-	doi=false, 					% Do not show doi in reference
-	isbn=false, 				% Do not show isbn link in reference
-	eprint=false, 			% Do not show eprint link in reference
-	maxbibnames=9, 			% Include up to 9 names in citation
-	firstinits=true,
-]{biblatex}
-\addbibresource{references.bib}
-
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
-\hypersetup{
-  pdfauthor={Manuel Razo-Mejia; Madhav Mani; Dmitri Petrov},
-  pdfkeywords={bayesian inference, variational inference, relative
-fitness, open-source, reproducible research},
-  colorlinks=true,
-  linkcolor={blue},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
-  pdfcreator={LaTeX via pandoc}}
-
-% Title
-\title{Bayesian inference of relative fitness on high-throughput pooled competition assays}
-
-% Authors
-\author[1,*]{Manuel Razo-Mejia}
-\author[2,3]{Madhav Mani}
-\author[1,4,5]{Dmitri Petrov}
-
-% Affiliations
-\affil[1]{Department of Biology, Stanford University, Stanford, CA, USA}
-\affil[2]{Department of Engineering Sciences and Applied Mathematics, Northwestern University, Chicago, IL, USA}
-\affil[3]{NSF-Simons Center for Quantitative Biology, Northwestern University, Chicago, IL, USA}
-\affil[4]{Stanford Cancer Institute, Stanford University School of Medicine, Stanford, CA, USA}
-\affil[5]{Chan Zuckerberg Biohub}
-\affil[*]{Correspondence: mrazo@stanford.edu}
-
-\setcounter{Maxaffil}{0}
-% Set affiliations in small font
-\renewcommand\Affilfont{\itshape\small}
-
-\date{}
-
-\begin{document}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden, interior hidden, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi
-
-% Remove main text from the table of contents by specifying not to include
-% any section or subsection
-\addtocontents{toc}{\protect\setcounter{tocdepth}{-1}}
-
-% Define reference segment for main text
-\begin{refsegment}
-% Generate filter to not include references from main text in the
-% supplemental references
-\defbibfilter{notother}{not segment=\therefsegment}
-
-\maketitle
-
-\begin{abstract}
-    The tracking of lineage frequencies via DNA barcode sequencing enables the
-  quantification of microbial fitness. However, experimental noise coming from
-  biotic and abiotic sources complicates the computation of a reliable
-  inference. We present a Bayesian pipeline to infer relative microbial fitness
-  from high-throughput lineage tracking assays. Our model accounts for multiple
-  sources of noise and propagates uncertainties throughout all parameters in a
-  systematic way. Furthermore, using modern variational inference methods based
-  on automatic differentiation, we are able to scale the inference to a large
-  number of unique barcodes. We extend this core model to analyze
-  multi-environment assays, replicate experiments, and barcodes linked to
-  genotypes. On simulations, our method recovers known parameters within
-  posterior credible intervals. This work provides a generalizable Bayesian
-  framework to analyze lineage tracking experiments. The accompanying
-  open-source software library enables the adoption of principled statistical
-  methods in experimental evolution.
-\end{abstract}
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-The advent of DNA barcoding---the ability to uniquely identify cell
-lineages with DNA sequences integrated at a specific locus---and
-high-throughput sequencing has opened new venues for understanding
-microbial evolutionary dynamics with an unprecedented level of temporal
-resolution \autocite{levy2015,nguyenba2019a,ascensao2023}. These
-experimental efforts rely on our ability to reliably infer the relative
-fitness of an ensemble of diverse genotypes. Moreover, inferring these
-fitness values over an ensemble of environmental conditions can help us
-determine the phenotypic diversity of a rapid adaptation process
-\autocite{kinsler2020}.
-
-As with any other sequencing-based quantification, tracking lineages via
-DNA barcode sequencing is inexorably accompanied by noise sources coming
-from experimental manipulation of the microbial cultures, DNA
-extraction, and sequencing library preparation that involves multiple
-rounds of PCR amplification, and the sequencing process itself. Thus,
-accounting for the uncertainty when inferring the relevant parameters
-from the data is a crucial step to draw reliable conclusions. Bayesian
-statistics presents a paradigm by which one can account for all known
-sources of uncertainty in a principled way \autocite{eddy2004a}. This,
-combined with the development of modern Markov Chain Monte Carlo
-sampling algorithms \autocite{betancourt2017} and approximate
-variational approaches \autocite{kucukelbir2016} have boosted a
-resurgence of Bayesian methods in different fields
-\autocite{efron2013a}.
-
-We present a Bayesian inference pipeline to quantify the uncertainty
-about the parametric information we can extract from high-throughput
-competitive fitness assays given a model of the data generation process
-and experimental data. In these assays, the fitness of an ensemble of
-genotypes is determined relative to a reference genotype
-\autocite{kinsler2020,ascensao2023}. Figure~\ref{fig-01}(A) shows a
-schematic of the experimental procedure in which an initial pool of
-barcoded strains are mixed with a reference strain and inoculated into
-fresh media. After some time---usually, enough time for the culture to
-saturate---an aliquot is transferred to fresh media, while the remaining
-culture is used for DNA sequencing of the lineage barcodes. The
-time-series information of the relative abundance of each lineage, i.e.,
-the barcode frequency depicted in Figure~\ref{fig-01}(B), is used to
-infer the relative fitness---the growth advantage on a per-cycle
-basis---for each lineage with respect to the reference strain. The
-proposed statistical model accounts for multiple sources of uncertainty
-when inferring the lineages' relative fitness values (see
-Section~\ref{sec-experiment} for details on sources of uncertainty
-accounted for by the model). Furthermore, minor changes to the core
-statistical model allow us to account for relevant experimental
-variations of these competition assays. More specifically, in
-Section~\ref{sec-multienv}, we present a variation of the statistical
-model to infer fitness on growth dilution cycles in multiple
-environments with proper error propagation. Furthermore, as described in
-Section~\ref{sec-replicates}, our statistical model can account for
-batch-to-batch differences when jointly analyzing multiple experimental
-replicates using a Bayesian hierarchical model. Finally, a variant of
-these hierarchical models, presented in Section~\ref{sec-genotypes}, can
-account for variability within multiple barcodes mapping to equivalent
-genotypes within the same experiment.
-
-For all the model variations presented in this paper, we benchmark the
-ability of our pipeline to infer relative fitness parameters against
-synthetic data generated from logistic growth simulations with added
-random noise. A \texttt{Julia} package accompanies the present method to
-readily implement the inference pipeline with state-of-the-art
-scientific computing software.
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-\hypertarget{sec-experiment}{%
-\subsection{Experimental setup}\label{sec-experiment}}
-
-The present work is designed to analyze time-series data of relative
-abundance of multiple microbial lineages uniquely identified by a DNA
-barcode \autocite{kinsler2020,ascensao2023}. In these competition
-assays, an ensemble of genotypes is pooled together with an unlabeled
-reference strain that, initially, represents the vast majority
-(\(\geq 90\%\)) of the cells in the culture (see schematic in
-Figure~\ref{fig-01}(A)). Furthermore, a fraction of labeled genotypes
-equivalent to the unlabeled reference strain---hereafter defined as
-\emph{neutral} lineages---are spiked in at a relatively high abundance
-(\(\approx 3-5\%\)). The rest of the culture is left for the ensemble of
-genotypes of interest.
-
-To determine the relative fitness of the ensemble of genotypes, a series
-of growth-dilution cycles are performed on either a single or multiple
-environments. In other words, the cultures are grown for some time;
-then, an aliquot is inoculated into fresh media for the next growth
-cycle. This process is repeated for roughly 4-7 cycles, depending on the
-initial abundances of the mutants and their relative growth rates. The
-DNA barcodes are sequenced at the end of each growth cycle to quantify
-the relative abundance of each of the barcodes. We point the reader to
-\textcite{kinsler2020} for specific details on these assays for \emph{S.
-cerevisiae} and to \textcite{ascensao2023} for equivalent assays for
-\emph{E. coli}. Figure~\ref{fig-01}(B) presents a typical barcode
-trajectory where the black trajectories represent the so-called
-\emph{neutral lineages}, genetically equivalent to the untagged ancestor
-strain that initially dominates the culture. These spiked-in neutral
-lineages simplify the inference problem since the fitness metric of all
-relevant barcodes is quantified with respect to these barcodes---thus
-referred to as \emph{relative fitness}.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/fig01.pdf}
-
-}
-
-\caption{\label{fig-01}\textbf{Typical competitive fitness experiment}.
-(A) Schematic of the typical experimental design to determine the
-competitive fitness of an ensemble of barcoded genotypes. Genotypes are
-pooled together and grown over multiple growth-dilution cycles. At the
-end of each cycle, a sample is processed to generate a library for
-amplicon sequencing. (B) Typical barcode trajectory dataset. From each
-time point, the relative frequency of each barcode is determined from
-the total number of reads. Shades of blue represent different relative
-fitness. Darker gray lines define the typical trajectory of neutral
-lineages.}
-
-\end{figure}
-
-\hypertarget{preliminaries-on-mathematical-notation}{%
-\subsection{Preliminaries on mathematical
-notation}\label{preliminaries-on-mathematical-notation}}
-
-Before jumping directly into the Bayesian inference pipeline, let us
-establish the mathematical notation used throughout this paper. We
-define (column) vectors as underlined lowercase symbols such as
-\begin{equation}\protect\hypertarget{eq-vec_def}{}{
-\underline{x} = \begin{bmatrix}
-    x_1\\
-    x_2\\
-    \vdots\\
-    x_N
-\end{bmatrix}.
-}\label{eq-vec_def}\end{equation} In the same way, we define matrices as
-double-underline uppercase symbols such as
-\begin{equation}\protect\hypertarget{eq-mat_def}{}{
-\underline{\underline{A}} =
-\begin{bmatrix}
-    A_{11} & A_{12} & \cdots & A_{1N}\\
-    A_{21} & A_{22} & \cdots & A_{2N}\\
-    \vdots & \vdots & \ddots & \vdots\\
-    A_{M1} & A_{M2} & \cdots & A_{MN}\\
-\end{bmatrix}.
-}\label{eq-mat_def}\end{equation}
-
-\hypertarget{sec-fitness_model}{%
-\subsection{Fitness model}\label{sec-fitness_model}}
-
-Empirically, each barcode frequency trajectory follows an exponential
-function of the form \autocite{levy2015,kinsler2020,ascensao2023}
-\begin{equation}\protect\hypertarget{eq-fitness}{}{
-f_{t+1}^{(b)} = f_{t}^{(b)} \mathrm{e}^{(s^{(b)} - \bar{s}_t)\tau},
-}\label{eq-fitness}\end{equation} where \(f_{t}^{(b)}\) is the frequency
-of barcode \(b\) at the end of cycle number \(t\), \(s^{(b)}\) is the
-relative fitness with respect to the reference strain---the quantity we
-want to infer from the data---\(\bar{s}_t\) is the mean fitness of the
-culture at the end of cycle number \(t\), and \(\tau\) is the time pass
-between cycle \(t\) and \(t+1\). We can rewrite
-Equation~\ref{eq-fitness} as
-\begin{equation}\protect\hypertarget{eq-logfreq}{}{
-\frac{1}{\tau}\ln \frac{f_{t+1}^{(b)}}{f_{t}^{(b)}} = (s^{(b)} - \bar{s}_t).
-}\label{eq-logfreq}\end{equation} Equation~\ref{eq-logfreq} separates
-the measurements---the barcode frequencies---from the unobserved
-(sometimes referred to as latent) parameters we want to infer from the
-data---the population mean fitness and the barcode relative fitness.
-This is ultimately the functional form used in our inference pipeline.
-Therefore, the relative fitness is computed by knowing the log frequency
-ratio of each barcode throughout the growth-dilution cycles.
-
-The presence of the neutral lineages facilitates the determination of
-the population mean fitness value \(\bar{s}_t\). Since every relative
-fitness is determined relative to the neutral lineage that dominates the
-culture, we define their fitness to be \(s^{(n)} = 0\), where the
-superscript \((n)\) specifies their neutrality. This means that
-Equation~\ref{eq-logfreq} for a neutral lineage takes the simpler form
-\begin{equation}\protect\hypertarget{eq-logfreq_neutral}{}{
-\frac{1}{\tau}\ln \frac{f_{t+1}^{(n)}}{f_{t}^{(n)}} = - \bar{s}_t.
-}\label{eq-logfreq_neutral}\end{equation} Therefore, we can use the data
-from these reference barcodes to directly infer the value of the
-population mean fitness.
-
-It is important to notice that the frequencies \(f_{t}^{(b)}\) are not
-the allele frequencies in the population (most of the culture is not
-sequenced since the reference strain is not barcoded), but rather the
-relative frequencies in the total number of sequencing reads. A way to
-conceptualize this subtle but important point is to assume exponential
-growth in the \emph{number of cells} \(N_t^{(b)}\) of the form
-\begin{equation}\protect\hypertarget{eq-ncells}{}{
-N_{t+1}^{(b)} = N_{t}^{(b)} \mathrm{e}^{\lambda^{(b)}\tau},
-}\label{eq-ncells}\end{equation} for every barcode \(b\) with growth
-rate \(\lambda^{(b)}\). However, when we sequence barcodes, we do not
-directly measure the number of cells, but some number of reads
-\(r_t^{(b)}\) that map to barcode \(b\). In the simplest possible
-scenario, we assume \begin{equation}\protect\hypertarget{eq-r_to_n}{}{
-r_{t}^{(b)} \propto N_{t}^{(b)},
-}\label{eq-r_to_n}\end{equation} where, importantly, the proportionality
-constant depends on the total number of reads for the library for cycle
-\(t\), which might vary from library to library. Therefore, to compare
-the number of reads between libraries at different time points, we must
-normalize the number of reads to the same scale. The simplest form is to
-define a relative abundance, i.e., a frequency with respect to the total
-number of reads, \begin{equation}\protect\hypertarget{eq-r_to_n_norm}{}{
-f_{t}^{(b)} \equiv \frac{r_{t}^{(b)}}{\sum_{b'} r_{t}^{(b')}}.
-}\label{eq-r_to_n_norm}\end{equation} This is the frequency
-Equation~\ref{eq-fitness} describes.
-
-Our ultimate objective is to infer the relative fitness \(s^{(b)}\) for
-each of the \(M\) relevant barcodes in the experiment---hereafter
-referred to as \(s^{(m)}\) to distinguish from the general \(s^{(b)}\)
-and the neutral lineages \(s^{(n)}\) relative fitness. To do so, we
-account for the three primary sources of uncertainty in our model:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Uncertainty in the determination of frequencies. Our model relates
-  frequencies between adjacent growth-dilution cycles to the fitness of
-  the corresponding strain. However, we do not directly measure
-  frequencies. Instead, our data for each barcode consists of a length
-  \(T\) vector of counts \(\underline{r}^{(b)}\) for each of the \(T\)
-  cycles in which the measurements were taken.
-\item
-  Uncertainty in the value of the population mean fitness. We define
-  neutral lineages to have fitness \(s^{(n)} = 0\), helping us anchor
-  the value of the population mean fitness \(\bar{s}_t\) for each pair
-  of adjacent growth cycles. Moreover, we take this parameter as an
-  empirical parameter to be obtained from the data, meaning that we do
-  not impose a functional form that relates \(\bar{s}_t\) to
-  \(\bar{s}_{t+1}\). Thus, we must infer the \(T-1\) values of this
-  population mean fitness with their uncertainty that must be propagated
-  to the value of the mutants' relative fitness.
-\item
-  Uncertainty in each of the mutants' fitness values.
-\end{enumerate}
-
-To account for all these sources of uncertainty in a principled way, in
-the next section, we develop a Bayesian inference pipeline.
-
-\hypertarget{sec-bayesian_inference}{%
-\subsection{Bayesian inference}\label{sec-bayesian_inference}}
-
-As defined in Section~\ref{sec-fitness_model}, our ultimate objective is
-to infer the vector of relative fitness values
-\begin{equation}\protect\hypertarget{eq-fitness_mut_vec}{}{
-\underline{s}^M = (s^{(1)}, s^{(2)}, \ldots, s^{(M)})^\dagger,
-}\label{eq-fitness_mut_vec}\end{equation} where \(^\dagger\) indicates
-the transpose. Our data consists of an \(T \times B\) matrix
-\(\underline{\underline{R}}\), where \(B = M + N\) is the number of
-unique barcodes given by the sum of the number of unique, relevant
-barcodes we care about, \(M\), and the number of unique neutral
-barcodes, \(N\), and \(T\) is the number of growth cycles where
-measurements were taken. The data matrix is then of the form
-\begin{equation}\protect\hypertarget{eq-R_mat}{}{
-\underline{\underline{R}} = \begin{bmatrix}
-- & \underline{r}_1 & - \\
-- & \underline{r}_2 & - \\
- & \vdots & \\
-- & \underline{r}_T & - \\
-\end{bmatrix},
-}\label{eq-R_mat}\end{equation} where each row \(\underline{r}_t\) is a
-\(B\)-dimensional array containing the raw barcode counts at cycle
-\(t\). We can further split each vector \(\underline{r}_t\) into two
-vectors of the form
-\begin{equation}\protect\hypertarget{eq-r-vec_split}{}{
-\underline{r}_t = \begin{bmatrix}
-\underline{r}_t^{N} \\
-\underline{r}_t^{M}
-\end{bmatrix},
-}\label{eq-r-vec_split}\end{equation} i.e., the vector containing the
-neutral lineage barcode counts \(\underline{r}_t^{N}\) and the
-corresponding vector containing the mutant barcode counts
-\(\underline{r}_t^{M}\). Following the same logic, matrix
-\(\underline{\underline{R}}\) can be split into two matrices as
-\begin{equation}\protect\hypertarget{eq-R-mat_split}{}{
-\underline{\underline{R}} = \left[ 
-\underline{\underline{R}}^N \; \underline{\underline{R}}^M
-\right],
-}\label{eq-R-mat_split}\end{equation} where
-\(\underline{\underline{R}}^N\) is a \(T \times N\) matrix with the
-barcode reads time series for each neutral lineage and
-\(\underline{\underline{R}}^M\) is the equivalent \(T \times M\) matrix
-for the non-neutral lineages.
-
-Our objective is to compute the joint probability distribution for all
-relative fitness values given our data. We can express this joint
-posterior distribution using Bayes theorem as
-\begin{equation}\protect\hypertarget{eq-bayes_obj}{}{
-\pi(\underline{s}^M \mid \underline{\underline{R}}) = \frac{
-\pi(\underline{\underline{R}} \mid \underline{s}^M) 
-\pi(\underline{s}^M)}
-{\pi(\underline{\underline{R}})},
-}\label{eq-bayes_obj}\end{equation} where hereafter \(\pi(\cdot)\)
-defines a probability density function. When defining our statistical
-model, we need not to focus on the denominator on the right-hand side of
-Equation~\ref{eq-bayes_obj}. Thus, we can write
-\begin{equation}\protect\hypertarget{eq-bayes_obj_propto}{}{
-\pi(\underline{s}^M \mid \underline{\underline{R}}) \propto
-\pi(\underline{\underline{R}} \mid \underline{s}^M) 
-\pi(\underline{s}^M).
-}\label{eq-bayes_obj_propto}\end{equation} However, when implementing
-the model computationally, the normalization constant on the right-hand
-side of Equation~\ref{eq-bayes_obj} must be computed. This can be done
-from the definition of the model via an integral of the form
-\begin{equation}\protect\hypertarget{eq-bayes_evidence}{}{
-\pi(\underline{\underline{R}}) = \int d^M \underline{s}^M
-\pi(\underline{\underline{R}} \mid \underline{s}^M) 
-\pi(\underline{s}^M),
-}\label{eq-bayes_evidence}\end{equation} also known as a marginalization
-integral. Hereafter, differentials of the form \(d^n\) imply a
-\(n\)-dimensional integral.
-
-Although Equation~\ref{eq-bayes_obj} and
-Equation~\ref{eq-bayes_obj_propto} seem simple enough, recall that
-Equation~\ref{eq-fitness} relates barcode frequency values and the
-population mean fitness to the mutant relative fitness. Therefore, we
-must include these nuisance parameters as part of our inference problem.
-We direct the reader to the supplementary materials for the exact
-definitions of these parameters. Here, it suffices to say that the
-inference problem must include the vector \(\underline{\bar{s}}_T\) of
-all population mean fitness values and the matrix
-\(\underline{\underline{F}}\) of all barcode frequencies within the
-sequencing data. With these nuisance variables in hand, the full
-inference problem we must solve takes the form
-\begin{equation}\protect\hypertarget{eq-bayes_full}{}{
-\pi(
-    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
-    \underline{\underline{R}}
-) \propto
-\pi(
-    \underline{\underline{R}} \mid
-    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
-)
-\pi(
-    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
-).
-}\label{eq-bayes_full}\end{equation} To recover the marginal
-distribution over the non-neutral barcodes relative fitness values, we
-can numerically integrate out all nuisance parameters, i.e.,
-\begin{equation}\protect\hypertarget{eq-bayes_full_marginal}{}{
-\pi(\underline{s}^M \mid \underline{\underline{R}}) =
-\int d^{T-1}\underline{\bar{s}}_T
-\int d^{B}\underline{f}_1 \cdots
-\int d^{B}\underline{f}_T
-\;
-\pi(
-    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
-    \underline{\underline{R}}
-).
-}\label{eq-bayes_full_marginal}\end{equation}
-
-\hypertarget{seq-split_posterior}{%
-\subsubsection{Factorizing the posterior
-distribution}\label{seq-split_posterior}}
-
-The left-hand side of Equation~\ref{eq-bayes_full} is extremely
-difficult to work with. However, we can take advantage of the structure
-of our inference problem to rewrite it in a more manageable form.
-Specifically, the statistical dependencies of our observations and
-latent variables allow us to factorize the joint distribution into the
-product of multiple conditional distributions. To gain some intuition
-about this factorization, let us focus on the inference of the
-population mean fitness values \(\underline{\bar{s}}_T\).
-Equation~\ref{eq-logfreq_neutral} relates the value of the population
-mean fitness to the neutral lineage frequencies and nothing else. This
-suggests that when writing the posterior for these population mean
-fitness parameters, we should be able to condition it only on the
-neutral lineage frequency values, i.e.,
-\(\pi(\underline{\bar{s}}_T \mid \underline{\underline{F}}^N)\). We
-point the reader to Section~\ref{sec-bayes_def} for the full
-mathematical details on this factorization. For our purpose here, it
-suffices to say we can rewrite the joint probability distribution as a
-product of conditional distributions of the form
-\begin{equation}\protect\hypertarget{eq-split_posterior}{}{
-\pi(
-    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
-    \underline{\underline{R}}
-) =
-\pi(
-    \underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}}^M
-)
-\pi(
-    \underline{\bar{s}}_T \mid \underline{\underline{F}}^N
-)
-\pi(\underline{\underline{F}} \mid \underline{\underline{R}}).
-}\label{eq-split_posterior}\end{equation} Written in this form,
-Equation~\ref{eq-split_posterior} captures the three sources of
-uncertainty listed in Section~\ref{sec-fitness_model} in each term.
-Starting from right to left, the first term on the right-hand side of
-Equation~\ref{eq-split_posterior} accounts for the uncertainty when
-inferring the frequency values given the barcode reads. The second term
-accounts for the uncertainty in the values of the mean population
-fitness at different time points. The last term accounts for the
-uncertainty in the parameter we care about---the mutants' relative
-fitnesses. We refer the reader to Section~\ref{sec-bayes_def} for an
-extended description of the model with specific functional forms for
-each term on the left-hand side of Equation~\ref{eq-split_posterior} as
-well as the extension of the model to account for multiple experimental
-replicates or hierarchical genotypes.
-
-\hypertarget{variational-inference}{%
-\subsubsection{Variational Inference}\label{variational-inference}}
-
-One of the technical challenges to the adoption of Bayesian methods is
-the analytical intractability of integrals such as that of
-Equation~\ref{eq-bayes_full_marginal}. Furthermore, even though
-efficient Markov Chain Monte Carlo (MCMC) algorithms such as Hamiltonian
-Montecarlo can numerically perform this integration
-\autocite{betancourt2017}, the dimensionality of the problem in
-Equation~\ref{eq-split_posterior} makes an MCMC-based approach
-prohibitively slow.
-
-To overcome this computational limitation, we rely on the recent
-development of the automatic differentiation variational inference
-algorithm (ADVI) \autocite{kucukelbir2016}. Briefly, when performing
-ADVI, our target posterior distribution
-\(\pi(\theta \mid \underline{\underline{R}})\), where
-\(\theta = (\underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}})\),
-is replaced by an approximate posterior distribution \(q_\phi(\theta)\),
-where \(\phi\) fully parametrizes the approximate distribution. As
-further explained in Section~\ref{sec-vi_primer}, the numerical
-integration problem is replaced by an optimization problem of the form
-\begin{equation}\protect\hypertarget{eq-vi_objective}{}{
-q^*_\phi(\theta) = \min _\phi
-D_{KL}(
-    q_\phi(\theta) \lvert \lvert
-    \pi(\theta \mid \underline{\underline{R}})
-),
-}\label{eq-vi_objective}\end{equation} where \(D_{KL}\) is the
-Kulback-Leibler divergence. In other words, the complicated
-high-dimensional numerical integration problem is transformed into a
-much simpler problem of finding the value of the parameters \(\phi\)
-such that Equation~\ref{eq-vi_objective} is satisfied as best as
-possible within some finite computation time. Although to compute
-Equation~\ref{eq-vi_objective}, we require the posterior distribution we
-are trying to approximate
-\(\pi(\theta \mid \underline{\underline{R}})\), it can be shown that
-maximizing the so-called evidence lower bound (ELBO)
-\autocite{kingma2014}---equivalent to minimizing the variational free
-energy \autocite{gottwald2020}---is mathematically equivalent to
-performing the optimization prescribed by
-Equation~\ref{eq-vi_objective}. We direct the reader to
-Section~\ref{sec-vi_primer} for a short primer on variational inference.
-
-This work is accompanied by the Julia library \texttt{BarBay.jl} that
-makes use of the implementation of both MCMC-based integration as well
-as ADVI optimization to numerically approximate the solution of
-Equation~\ref{eq-bayes_full_marginal} within the Julia ecosystem
-\autocite{ge2018}.
-
-\hypertarget{inference-on-a-single-dataset}{%
-\subsection{Inference on a single
-dataset}\label{inference-on-a-single-dataset}}
-
-To assess the inference pipeline performance, we applied it to a
-simulated dataset with known ground truth relative fitness values (See
-Section~\ref{sec-logistic} for details on simulation).
-Figure~\ref{fig-02}(A) shows the structure of the synthetic dataset. The
-majority of barcodes of interest (faint color lines) are adaptive
-compared to the neutral barcodes (\(s^{(m)} > 0\)). Although the barcode
-frequency trajectories look relatively smooth, our fitness model
-requires the computation of the log frequency ratio between adjacent
-time points as derived in Equation~\ref{eq-logfreq}.
-Figure~\ref{fig-02}(B) shows such data transformation where we can
-better appreciate the observational noise input into our statistical
-model. This noise is evident for the darker lines representing the
-neutral barcodes since all of these lineages are assumed to be
-identically distributed.
-
-To visualize the performance of our inference pipeline in fitting our
-fitness model to the observed data, we compute the so-called posterior
-predictive checks (PPC). In short, the PPC consists of repeatedly
-generating synthetic datasets in agreement with the results from the
-inference results. In other words, we use the resulting parameter values
-from the ADVI inference to generate possible datasets in agreement with
-the inferred values (See Section~\ref{sec-ppc} for further details on
-these computations). Figure~\ref{fig-02}(C) shows these results for all
-neutral lineages (upper left corner plot) and a few representative
-non-neutral barcodes. The different color shades represent the 95\%,
-68\%, and 5\% credible regions, i.e., the regions where we expect to
-find the data with the corresponding probability---or in terms of our
-parameter, the \(X\%\) credible region is the interval where we expect
-the true parameter value to lie with \(X\%\) probability.
-
-The main advantage of our method is this natural interpretability of
-these credible regions where an \(X\%\) credible region indeed captures
-the region of parameter space where we expect with \(X\%\) probability
-the actual value of the parameter lies given our statistical model, our
-prior information, and the observed experimental data. A common mistake
-in the literature is interpreting frequentist confidence intervals as
-Bayesian credible regions when they are not equivalent
-\autocite{morey2016}. Frequentist confidence intervals and Bayesian
-credible regions are based on fundamentally different philosophical
-approaches to statistics. Frequentist confidence intervals represent the
-range of values that would contain the true population parameter with a
-certain probability if the experiment was repeated many times. The
-confidence interval does not represent the probability that the interval
-contains the true value. According to a specific model and prior
-information, Bayesian credible regions represent the range of values
-that contain the parameter with a certain posterior probability. The
-credible region directly represents the probability that the region
-contains the true value. So, frequentist confidence intervals cannot be
-interpreted as Bayesian credible regions because they have fundamentally
-different meanings. Treating an \(X\%\) confidence interval like an
-\(X\%\) credible region is fallacious since confidence intervals do not
-represent probabilistic coverage of the true value like credible
-regions. The intervals are generated through entirely different
-procedures.
-
-To capture the global performance of the model, Figure~\ref{fig-02}(D)
-compares the known ground truth with the inferred relative fitness value
-for all barcodes of interest. There is an excellent degree of
-correspondence between these values, with the error bars representing
-the 68\% credible region for the parameter value crossing the identity
-line for most barcodes. This latter point is made clear with
-Figure~\ref{fig-02}(E) where \(\approx 90\%\) of ground truth fitness
-values fall within one standard deviation of the mean in the inferred
-posterior distributions.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/fig02.pdf}
-
-}
-
-\caption{\label{fig-02}\textbf{Single dataset inference}. (A) Frequency
-trajectories that represent the raw data going into the inference. (B)
-Log frequency ratio between two adjacent time points used by the
-inference pipeline. Darker lines represent the neutral barcodes. These
-transformed data are much more noisy than the seemingly smooth frequency
-trajectories. (C) Examples of the posterior predictive checks for all
-neutral lineages (upper left panel) and a subset of representative
-mutant lineages. Shaded regions represent the 95\%, 68\%, and 5\%
-credible regions for the data. The reported errors above the plot
-represent the 68\% credible region on the mutant relative fitness
-marginal distribution. (D) Comparison between the ground truth fitness
-value from the logistic-growth simulation and the inferred fitness
-value. Gray error bars represent the 68\% posterior credible region for
-the relative fitness values. (E) The empirical cumulative distribution
-function (ECDF) for the absolute z-score value of the ground truth
-parameter value within the inferred fitness posterior distribution.}
-
-\end{figure}
-
-\hypertarget{sec-multienv}{%
-\subsection{Fitness inference on multiple
-environments}\label{sec-multienv}}
-
-The fitness model in Equation~\ref{eq-fitness} relates nuisance
-parameters such as the population mean fitness and the barcode
-frequencies to the relative fitness parameter we want to infer from the
-data. These dependencies imply that uncertainty on the estimates of
-these nuisance parameters influences the inference of the relevant
-parameters. For example, imagine a scenario where the neutral lineages
-data were incredibly noisy, leading to poor estimates of the population
-mean fitness values \(\underline{\bar{s}}_T\). Since the relative
-fitness of any non-neutral barcode \(s^{(m)}\) is determined with
-respect to these neutral barcodes, not accounting for the lack of
-precision in the value of the population mean fitness would result in
-misleading estimates of the accuracy with which we determine the value
-of the parameter we care about. Thus, propagating these sources of
-uncertainty in nuisance parameters is vital to generate an unbiased
-estimate of the relevant information we want to extract from the data.
-One of the benefits of Bayesian methods is the intrinsic error
-propagation embedded in the mathematical framework. For our previous
-example, the uncertainty on the value of the population mean fitness
-values is propagated to the relative fitness of a non-neutral barcode
-since we defined a joint posterior distribution over all parameters as
-fully expressed in Equation~\ref{eq-bayes_full}.
-
-This natural error propagation can help us with the experimental design
-schematized in Figure~\ref{fig-03}(A). Here, rather than performing
-growth-dilution cycles in the same environment, the cells are diluted
-into a different environment. Thus, the uncertainty on the fitness
-estimate for the previous environment must be propagated to that of the
-next one. To validate the extension of our statistical model to this
-scenario, Figure~\ref{fig-03}(B) shows the trajectory of the log
-frequency ratios between adjacent time points. The different colored
-regions correspond to the different environments. For this simulation,
-the growth rate of Environment 2 was set to be, on average, half of the
-average growth rate in Environment 1. Equivalently, the growth rate in
-Environment 3 was set to be, on average, twice the average growth rate
-in Environment 1. Figure~\ref{fig-03}(C-E) show the correspondence
-between the simulation ground truth and the inferred fitness values,
-where the error bars represent the 68\% credible region.
-Figure~\ref{fig-03}(F) summarizes the performance of our inference
-pipeline by showing the empirical cumulative distribution functions for
-the absolute value of the ground truth fitness value z-score within the
-posterior distribution. This plot shows that, overall, \(\approx 75\%\)
-of inferred mean values fall within one standard deviation of the ground
-truth. For completeness, Figure~\ref{fig-03}(G) shows the posterior
-predictive checks for a few example barcodes.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/fig03.pdf}
-
-}
-
-\caption{\label{fig-03}\textbf{Multi-environment fitness inference.} (A)
-Schematic of the simulated experimental design where growth-dilution
-cycles are performed into different environments for each cycle. (B) log
-frequency rations between adjacent time points. Darker lines represent
-the neutral barcodes. The colors in the background demark the
-corresponding environment, matching colors in (A). Environment 2 is set
-to have, on average, half the growth rate of environment 1. Likewise,
-environment 3 is set to have, on average, twice the growth rate of
-environment 1. (C-E) Comparison between the ground truth fitness value
-from the logistic-growth simulation and the inferred fitness value for
-each environment. Gray error bars represent the 68\% posterior credible
-region. (F) The empirical cumulative distribution function (ECDF) for
-the absolute z-score value of the ground truth parameter value within
-the inferred fitness posterior distribution for all fitness values
-(black line) and each environment individually (color lines). (G)
-Examples of the posterior predictive checks for all neutral lineages
-(upper left panel) and a subset of representative mutant lineages.
-Shaded regions surrounding the data represent the 95\%, 68\%, and 5\%
-credible regions for the data. The reported errors above the plot
-represent the 68\% credible region on the mutant relative fitness
-marginal distribution. Background colors match those of (A).}
-
-\end{figure}
-
-\hypertarget{sec-replicates}{%
-\subsection{Accounting for experimental replicates via hierarchical
-models}\label{sec-replicates}}
-
-Our inference pipeline can be extended to account for multiple
-experimental replicates via Bayesian hierarchical models
-\autocite{betancourt2013}. Briefly, when accounting for multiple
-repeated measurements of the same phenomena, there are two extreme cases
-one can use to perform the data analysis: On the one hand, we can treat
-each measurement as entirely independent, losing the power to utilize
-multiple measurements when trying to learn a single parameter. This can
-negatively impact the inference since, in principle, the value of our
-parameter of interest should not depend on the particular experimental
-replicate in question. However, this approach does not allow us to
-properly ``combine'' the uncertainties in both experiments when
-performing the inference. On the other hand, we can pool all data
-together and treat our different experiments as a single measurement
-with higher coverage. This loses the subtle differences due to biotic
-and abiotic batch effects, effectively halving the data that goes into
-our inference problem.
-
-Hierarchical models present a middle ground between these extremes.
-First, hierarchical models rely on the definition of so-called
-\emph{hyper-parameters}, that capture the parametric inference we are
-interested in---for this inference problem, we have a hyper-fitness
-value \(\theta^{(m)}\) for each non-neutral barcode. Second, each
-experiment draws randomly from the distribution of this hyper-parameter,
-allowing for subtle variability between experiments to be accounted
-for---in the present inference pipeline, each experimental replicate
-gets assigned a \emph{local} fitness value \(s^{(m)}_i\), where the
-extra sub-index indicates the \(i\)-th experimental replicate.
-Conceptually, we can think of the local fitness for replicate \(i\) as
-being sampled from a distribution that depends on the value of the
-global hyper-fitness value, i.e., \(s^{(m)}_i \sim \pi_{\theta^{(m)}}\),
-where the subindex \(\theta^{(m)}\) indicates the distribution's
-parametric dependence on the hyper-fitness value. This way of
-interpreting the connection between the distribution
-\(\pi_{\theta^{(m)}}\) and the local fitness implies that a large
-replicate-to-replicate variability would lead to a broad hyper-fitness
-distribution---implying a large uncertainty when determining the
-parameter that characterizes the overall relative fitness. We point the
-reader to Section~\ref{sec-hierarchical_model} for the full definition
-of the hierarchical model used in this section. Importantly, as
-schematized in Figure~\ref{fig-04}(A), the influence between different
-experimental replicates runs both ways. First, the data from one
-experimental replicate (\(\underline{\underline{R}}^M_k\) in the
-diagram) informs all local fitness values via the global hyper-fitness
-(upper panel in Figure~\ref{fig-04}(A)). Second, the local fitness value
-is informed by the data from all experimental replicates via the same
-global hyper-fitness parameter (lower panel in Figure~\ref{fig-04}(A)).
-
-To test the performance of this model, we simulated two experimental
-replicates with 1000 unique barcodes (see Figure~\ref{fig-04}(B-C))
-where we randomly sampled a ground truth hyper-fitness value
-\(\theta^{(m)}\) for each barcode. We sampled a variation from this
-hyper-fitness value for each experimental replicate \(s^{(m)}_i\) to
-capture experimental batch effects. Figure~\ref{fig-04}(D) shows the
-relationship between hyper-fitness and replicate fitness values for this
-simulation. The spread around the identity line represents the expected
-batch-to-batch variation. The posterior predictive checks examples in
-Figure~\ref{fig-04}(E) show that the hierarchical model can correctly
-fit the data for each experimental replicate. Furthermore,
-Figure~\ref{fig-04}(F-G) show a high correlation between the ground
-truth and the inferred fitness values. The empirical cumulative
-distribution functions shown in Figure~\ref{fig-04}(H-I) reveal that for
-\(\approx 75\%\) of the non-neutral barcodes, the ground truth
-hyper-fitness values fall within one standard deviation from the mean
-value in the posterior distributions.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/fig04.pdf}
-
-}
-
-\caption{\label{fig-04}\textbf{Hierarchical model on experimental
-replicates}. (A) Schematic depiction of the interactions between local
-fitness values \(\underline{s}_k^M\) through the global hyper-fitness
-value \(\underline{\theta}^M\) for \(K\) hypothetical experimental
-replicates. The upper diagram shows how the data from replicate \(k\)
-informs all local fitness values via the hyper-fitness parameter. The
-lower panel shows the reverse, where all other datasets inform the local
-fitness value. (B-C) Simulated replicate datasets with 900 barcodes of
-interest and 100 neutral lineages. (D) Comparison between the simulation
-ground truth hyper-fitness and each replicate ground truth fitness. The
-scatter between parameters captures experimental batch effects. (E)
-Examples of the posterior predictive checks for all neutral lineages
-(upper left panels) and a subset of representative mutant lineages.
-Shaded regions from light to dark represent the 95\%, 68\%, and 5\%
-credible regions. (F-G) Comparison between the simulation's ground truth
-hyper-fitness (F) and replicate fitness (G) values and the inferred
-parameters. Gray error bars represent the 68\% posterior credible
-region. (H-I) The empirical cumulative distribution function (ECDF) for
-the absolute z-score value of the ground truth parameter value within
-the inferred hyper-fitness posterior distribution (H) and replicate
-fitness (I).}
-
-\end{figure}
-
-As shown in Figure~\ref{fig-05}, the structure imposed by the
-hierarchical model schematized in Figure~\ref{fig-04}(A), where we
-explicitly account for the connection between experimental replicates
-can improve the quality of the inference. Inferred fitness values
-between experimental replicates exhibit a stronger degree of correlation
-in the hierarchical model (Figure~\ref{fig-05}(A)) compared to
-conducting inference on each replicate independently
-(Figure~\ref{fig-05}(B)). Moreover, when comparing the inferred
-hyper-fitness values---the objective parameter when performing multiple
-experimental measurements---the hierarchical model outperforms averaging
-the independent experimental replicates as shown in
-Figure~\ref{fig-05}(C) and (D).
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/fig05.pdf}
-
-}
-
-\caption{\label{fig-05}\textbf{Comparison between hierarchical model and
-single dataset model}. (A-B) comparison of inferred fitness values
-between experimental replicates when fitting a hierarchical model (A) or
-independently fitting each dataset (B). Gray error bars represent the
-68\% credible regions. (C) Comparison between the ground truth
-hyper-fitness value and the inferred parameters. The blue dots show the
-inferred hyper-fitness values when assuming a hierarchical model. Gray
-error bars show the 68\% credible region for this inference. The yellow
-dots show the average of the mean inferred fitness values for the two
-experimental replicates. No error bars are shown for these, as it is
-inappropriate to compute one with two data points per non-neutral
-barcode. (D) Empirical cumulative distribution function (ECDF) of the
-absolute difference between the inferred mean and the ground truth
-hyper-fitness.}
-
-\end{figure}
-
-\hypertarget{sec-genotypes}{%
-\subsection{Accounting for multiple barcodes per genotype via
-hierarchical models}\label{sec-genotypes}}
-
-Hierarchical models can also capture another experimental design in
-which multiple barcodes map to the same or an equivalent genotype. As we
-will show, this many-to-one mapping can improve the inference compared
-to the extreme cases of inferring the fitness of each barcode
-independently or pooling the data of all barcodes mapping to a single
-genotype. As schematized in Figure~\ref{fig-06}(A), a small modification
-of the base model allows us to map the structure of our original model
-to that of a hierarchical model with a fitness hyperparameter vector
-\(\underline{\theta}^G\), where \(G\) is the number of genotypes in the
-dataset.
-
-Figure~\ref{fig-06}(B) shows a single experimental replicate in which 90
-genotypes were assigned a random number of barcodes (a multinomial
-distribution with a mean of ten barcodes per genotype) for a total of
-900 non-neutral barcodes. To assess the performance of the hierarchical
-model proposed in Figure~\ref{fig-06}(A), we performed inference using
-this hierarchical model, as well as the two extreme cases of ignoring
-the connection between the barcodes belonging to the same
-genotype---equivalent to performing inference using the model presented
-in Figure~\ref{fig-02}(A) over the barcodes---or pooling the data of all
-barcodes belonging to the same genotype into a single count---equivalent
-to performing inference using the model presented in
-Figure~\ref{fig-02}(A) over the pooled barcodes.
-Figure~\ref{fig-06}(C-D) shows the comparison between the simulation
-ground truth and the inferred values for these three cases. Not only do
-the hierarchical model results show higher degrees of correlation with
-the ground truth, but the error bars (representing the 68\% credible
-regions) are smaller, meaning that the uncertainty in the estimate of
-the parameter we care about decreases when using the hierarchical model.
-The improvement in the prediction can be seen in Figure~\ref{fig-06}(F)
-where the empirical cumulative distribution function of the absolute
-difference between the mean inferred value and the simulation ground
-truth is shown for all three inference models. The hierarchical model's
-curve ascends more rapidly, showing that, in general, the inferred
-values are closer to the ground truth. For completeness,
-Figure~\ref{fig-06}(G) shows some examples of how the hierarchical model
-can capture the raw log-frequency count observations.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/fig06.pdf}
-
-}
-
-\caption{\label{fig-06}\textbf{Hierarchical model for multiple barcodes
-per genotype.} (A) Schematic depiction of the hierarchical structure for
-multiple barcodes mapping to a single genotype. A set of barcodes
-mapping to an equivalent genotype map to ``local'' fitness values
-\(s^{(b)}\) that are connected via a hyper-fitness parameter for the
-genotype \(\theta^{(g)}\). (B) Simulated dataset with 100 neutral
-lineages and 900 barcodes of interest distributed among 90 genotypes.
-(C-E) Comparison between the inferred and ground truth fitness values
-for a hierarchical model (C), a model where each barcode is inferred
-independently (D), and a model where barcodes mapping to the same
-genotype are pooled together (E). Gray error bars represent the 68\%
-credible regions. (F) Empirical cumulative distribution function (ECDF)
-of the absolute difference between the inferred mean and the ground
-truth fitness values for all three models. (G) Examples of the posterior
-predictive checks for all neutral lineages (upper left panels) and a
-subset of representative mutant lineages. Shaded regions from light to
-dark represent the 95\%, 68\%, and 5\% credible regions.}
-
-\end{figure}
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-Experimental evolution of microbial systems has dramatically advanced
-our understanding of the basic principles of biological evolution
-\autocite{kussell2013}. From questions related to the optimal
-fine-tuning of gene expression programs \autocite{Dekel2005}, to the
-dimensionality, geometry, and accessibility of the adaptive fitness
-landscape explored by these rapidly adapting populations
-\autocite{kinsler2020,maeda2020}, to the emergence of eco-evolutionary
-dynamics in a long-term evolution experiment \autocite{good2017}; for
-all of these and other cases, the microbial experimental platform
-combined with high-throughput sequencing has been essential to tackling
-these questions with empirical data. This exciting research area
-promises to improve as new culturing technologies \autocite{jagdish2022}
-as well as more complex lineage barcoding schemes
-\autocite{nguyenba2019a,yang2022}, are adopted.
-
-For this data-heavy field, properly accounting for the uncertainty in
-parameters inferred from experiments is vital to ensure the conclusions
-drawn are reliable. Bayesian statistics presents a principled way to
-quantify this uncertainty systematically \autocite{gelman2010}.
-Moreover, Bayesian analysis offers a more natural way to interpret the
-role that probability theory plays when performing data analysis
-compared to the often-misinterpreted frequentist methods
-\autocite{vanderplas2014}. Nevertheless, the technical challenges
-associated with Bayesian analysis has limited its application. This is
-set to change as recognition of the misuse of frequentist concepts such
-as the p-value is receiving more attention \autocite{nuzzo2014}.
-Moreover, advances in numerical methods such as Hamiltonian Monte Carlo
-\autocite{betancourt2017} and variational inference
-\autocite{kucukelbir2016} allows for complex Bayesian models to be fit
-to empirical data.
-
-In this paper, we present a computational pipeline to analyze
-lineage-tracking time-series data for massive-parallel competition
-assays. More specifically, we fit a Bayesian model to infer the fitness
-of multiple genotypes relative to a reference
-\autocite{kinsler2020,ascensao2023}. The proposed model accounts for
-multiple sources of uncertainty with proper error propagation intrinsic
-to Bayesian methods. To scale the inference pipeline to large datasets
-with \(> 10,000\) barcodes, we use the ADVI algorithm
-\autocite{kucukelbir2016} to fit a variational posterior distribution.
-The main difference between our method and previous inference pipelines,
-such as \textcite{li2023}, is that the present analysis provides
-interpretable errors on the inferred fitness values. The reported
-uncertainty intervals---known as credible regions---can be formally
-interpreted as capturing the corresponding probability mass of finding
-the true value of the parameter given the model, the prior information,
-and the data. Furthermore, minor modifications to the structure of the
-statistical model presented in this work allow for the analysis of
-different experimental designs, such as growth-dilution cycles in
-different environments, joint analysis of multiple experimental
-replicates of the same experiment via hierarchical models, and a
-hierarchical model for multiple barcodes mapping to equivalent
-genotypes. We validate our analysis pipeline on simulated datasets with
-known ground truth, showing that the model fits the data adequately,
-capturing the ground truth parameters within the posterior distribution.
-
-It is important to highlight some of the consequences of the general
-experimental design and the implicit assumptions within the proposed
-statistical model to analyze the resulting data. First, the composition
-of the population is such that the initial fraction of the population
-occupied by the barcoded genotypes is small---usually \textgreater90\%
-of the initial population is the non-labeled reference strain. This
-constraint is important as the fitness model used to fit the time series
-data assumes that the tracked frequencies are \(\ll 1\). Second, when
-computing log frequency ratios, we can run into the issue of dividing by
-zero. This is a common problem when dealing with molecular count data
-\autocite{lovell2020}. Our model gets around this issue by assuming that
-the frequency of any barcode cannot be, but still can get arbitrarily
-close to, zero. Therefore, we implicitly assume that no lineage goes
-extinct during the experiment. Moreover, the statistical model directly
-accounts for the uncertainty associated with having zero barcode counts,
-increasing the corresponding uncertainty. Third, the models presented in
-this paper require the existence of a labeled sub-population of barcoded
-reference strains. These barcodes help determine the fitness baseline,
-as every fitness is quantified with respect to this reference genotype.
-This experimental design constraint facilitates the inference of the
-population mean fitness since most of the culture---the unlabeled
-reference genotype---is not tracked. Finally, the presented statistical
-model assumes that relative fitness is solely a constant of the
-environment and the genotype. Future directions of this work could
-extend the fitness model to properly analyze data with time-varying or
-frequency-dependent fitness values.
-
-In total, the statistical model presented in this work and the software
-package accompanying the paper allow for a principled way of quantifying
-the accuracy with which we can extract relevant parametric information
-from large-scale multiplexed fitness competition assays. Furthermore,
-the implementation of Bayesian models and their fitting via automatic
-differentiation approaches opens the gate to extend this type of formal
-analysis to the data-rich literature in experimental evolution and other
-high-throughput technologies applications.
-
-\hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
-
-We would like to thank Griffin Chure and Michael Betancourt for their helpful
-advice and discussion. We would like to thank Karna Gowda, Spencer Farrell, and
-Shaili Mathur for critical observations on the manuscript. This work was
-supported by the NIH/NIGMS, Genomics of rapid adaptation in the lab and in the
-wild R35GM11816506 (MIRA grant), the NIH, Unravelling mechanisms of tumor
-suppression in lung cancer R01CA23434903, the NIH (PQ4), Quantitative and
-multiplexed analysis of gene function in cancer in vivo R01CA23125303, the NIH,
-Genetic Determinants of Tumor Growth and Drug Sensitivity in EGFR Mutant Lung
-Cancer R01CA263715, the NIH, Dissecting the interplay between aging, genotype,
-and the microenvironment in lung cancer U01AG077922, the NIH, Genetic dissection
-of oncogenic Kras signaling R01CA230025, and the CZ Biohub investigator program.
-MRM was supported by the Schmidt Science Fellowship. MM was supported by The
-National Science Foundation-Simons Center for Quantitative Biology at
-Northwestern University and the Simons Foundation grant 597491. MM is a Simons
-Investigator. DP is a CZ Biohub investigator.
-
-% Print main text references
-\printbibliography[segment=\therefsegment]
-% Close reference segment
-\end{refsegment}
-
-\clearpage
-
-% SUPPLEMENTAL MATERIAL
-
-% Indicate that now all sections and subsections should be included in the
-% table of contents so that only the SI is included.
-\addtocontents{toc}{\protect\setcounter{tocdepth}{3}}
-
-% Define reference section for the supplemental material
-\begin{refsegment}
-% Set equation, table and figure counters to begin with "S"
-\beginsupplement
-
-\hypertarget{supplementary-materials}{%
-\section{Supplementary Materials}\label{supplementary-materials}}
-
-\tableofcontents
-
-\hypertarget{sec-vi_primer}{%
-\subsection{Primer on Variational Inference}\label{sec-vi_primer}}
-
-In this section, we will briefly introduce the idea behind variational
-inference. Recall that any Bayesian inference problem deals with the
-joint distribution between observations \(\underline{x}\) and unobserved
-latent variables \(\underline{\theta}\). This joint distribution can be
-written as the product of a distribution of the observations
-\(\underline{x}\) conditioned on the \(\underline{\theta}\) and the
-marginal distribution of these latent variables, i.e.,
-\begin{equation}\protect\hypertarget{eq-vi_joint_dist}{}{
-\pi(\underline{x}, \underline{\theta}) =
-\pi(\underline{x} \mid \underline{\theta}) \pi(\underline{\theta}).
-}\label{eq-vi_joint_dist}\end{equation} A Bayesian inference pipeline's
-objective is to compute the latent variables' posterior probability
-given a set of observations. This computation is equivalent to updating
-our prior beliefs about the set of values that the latent variables take
-after taking in new data. We write this as Bayes theorem
-\begin{equation}\protect\hypertarget{eq-vi_bayes_theorem}{}{
-\pi(\underline{\theta} \mid \underline{x}) = 
-\frac{
-        \pi(\underline{x} \mid \underline{\theta})\pi(\underline{\theta})
-    }{
-        \pi(\underline{x})
-    }.
-}\label{eq-vi_bayes_theorem}\end{equation} The main technical challenge
-for working with Equation~\ref{eq-vi_bayes_theorem} comes from the
-computation of the denominator, also known as the \emph{evidence} or the
-\emph{marginalized likelihood}. The reason computing this term is
-challenging is because it involves a (potentially) high-dimensional
-integral of the form
-\begin{equation}\protect\hypertarget{eq-vi_evidence_int}{}{
-\pi(\underline{x}) = 
-\int\cdots\int d^K\underline{\theta}\; \pi(\underline{x}, \underline{\theta}) = 
-\int\cdots\int d^K\underline{\theta}\; 
-\pi(\underline{x} \mid \underline{\theta})
-\pi(\underline{\theta}),
-}\label{eq-vi_evidence_int}\end{equation} where \(K\) is the
-dimesionality of the \(\underline{\theta}\) vector. Here, the integrals
-are taken over the support---the set of values valid for the
-distribution---of \(\pi(\underline{\theta})\). However, only a few
-selected distributions have a closed analytical form; thus, in most
-cases Equation~\ref{eq-vi_evidence_int} must be solved numerically.
-
-Integration in high-dimensional spaces can be computationally extremely
-challenging. For a naive numerical quadrature procedure, integrating
-over a grid of values for each dimension of \(\underline{\theta}\) comes
-with an exponential explosion of the number of required grid point
-evaluations, most of which do not contribute significantly to the
-integration. To gain visual intuition about this challenge, imagine
-integrating the function depicted in Figure~\ref{fig-SI01}. If the
-location of the high-density region (dark peak) is unknown, numerical
-quadrature requires many grid points to ensure we capture this peak.
-However, most of the numerical evaluations of the function on the grid
-points do not contribute significantly to the integral. Therefore, our
-computational resources are wasted on insignificant evaluations. This
-only gets worse as the number of dimensions increases since the number
-of grid point evaluation scales exponentially.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/figSIX_gaussian_peak.png}
-
-}
-
-\caption{\label{fig-SI01}\textbf{High-dimensional numerical quadrature
-does not scale with dimensionality.} Schematic depiction of the problem
-with naive numerical quadrature to integrate over an unknown density.
-While the density is concentrated on the dark peak, most of the
-evaluations over the \(x_1-x_2\) grid do not contribute to the value of
-the integral}
-
-\end{figure}
-
-Modern Markov Chain Monte Carlo algorithms, such as Hamiltonian Monte
-Carlo, can efficiently perform this high-dimensional integration by
-utilizing gradient information from the target density
-\textcite{betancourt2017}. Nevertheless, these sampling-based methods
-become prohibitively slow for the number of dimensions our present
-inference problem presents. Thus, there is a need to find scalable
-methods for the inference problem in Equation~\ref{eq-vi_bayes_theorem}.
-
-Variational inference circumvents these technical challenges by
-proposing an approximate solution to the problem. Instead of working
-with the posterior distribution in its full glory
-\(\pi(\underline{\theta} \mid \underline{x})\), let us propose an
-approximate posterior distribution \(q_\phi\) that belongs to a
-distribution family fully parametrized by \(\phi\). For example, let us
-say that the distribution \(q_\phi\) belongs to the family of
-multivariate Normal distributions such that
-\(\phi = (\underline{\mu}, \underline{\underline{\Sigma}})\), where
-\(\underline{\mu}\) is the vector of means and
-\(\underline{\underline{\Sigma}}\) is the covariance matrix. If we
-replace \(\pi\) by \(q_\phi\), we want \(q_\phi\) to resemble the
-original posterior as much as possible. Mathematically, this can be
-expressed as minimizing a ``\emph{distance metric}''---the
-Kullback-Leibler (KL) divergence, for example---between the
-distributions. Note that we use quotation marks because, formally, the
-KL divergence is not a distance metric since it is not symmetric.
-Nevertheless, the variational objective is set to find a distribution
-\(q_\phi^*\) such that
-\begin{equation}\protect\hypertarget{eq-vi_objective}{}{
-q_\phi^*(\underline{\theta}) =
-\min_\phi D_{KL}\left(
-    q_\phi(\underline{\theta}) \vert\vert 
-    \pi(\underline{\theta} \mid \underline{x})
-\right),
-}\label{eq-vi_objective}\end{equation} where \(D_{KL}\) is the KL
-divergence. Furthermore, we highlight that the KL divergence is a
-strictly positive number, i.e.,
-\begin{equation}\protect\hypertarget{eq-vi_kl_pos}{}{
-D_{KL}\left(
-    q_\phi(\underline{\theta}) \vert\vert 
-    \pi(\underline{\theta} \mid \underline{x})
-\right) \geq 0,
-}\label{eq-vi_kl_pos}\end{equation} as this property will become
-important later on.
-
-At first sight, Equation~\ref{eq-vi_objective} does not improve the
-situation but only introduces further technical complications. After
-all, the definition of the KL divergence
-\begin{equation}\protect\hypertarget{eq-vi_kl_div}{}{
-D_{KL}\left(
-    q_\phi(\underline{\theta}) \vert\vert 
-    \pi(\underline{\theta} \mid \underline{x})
-\right) \equiv 
-\int \cdots \int d^K\underline{\theta}\;
-q_\phi(\underline{\theta})
-\ln \frac{
-    q_\phi(\underline{\theta})
-}{
-    \pi(\underline{\theta} \mid \underline{x})
-},
-}\label{eq-vi_kl_div}\end{equation} includes the posterior distribution
-\(\pi(\underline{\theta} \mid \underline{x})\) we are trying to get
-around. However, let us manipulate Equation~\ref{eq-vi_kl_div} to beat
-it to a more reasonable form. First, we can use the properties of the
-logarithms to write
-\begin{equation}\protect\hypertarget{eq-vi_step01}{}{
-D_{KL}\left(
-    q_\phi(\underline{\theta}) \vert\vert 
-    \pi(\underline{\theta} \mid \underline{x})
-\right) = 
-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln q_\phi(\underline{\theta}) -
-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln \pi(\underline{\theta} \mid \underline{x}),
-}\label{eq-vi_step01}\end{equation} where, for convenience, we write a
-single integration sign (\(d^K\underline{\theta}\;\) still represents a
-multi-dimensional differential). For the second term in
-Equation~\ref{eq-vi_step01}, we can substitute the term inside the
-logarithm using Equation~\ref{eq-vi_bayes_theorem}. This results in
-\begin{equation}\protect\hypertarget{eq-vi_step02}{}{
-\begin{aligned}
-D_{KL}\left(
-    q_\phi(\underline{\theta}) \vert\vert 
-    \pi(\underline{\theta} \mid \underline{x})
-\right) &= 
-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln q_\phi(\underline{\theta}) \\
-&- \int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln \left( 
-    \frac{
-        \pi(\underline{x} \mid \underline{\theta})\pi(\underline{\theta})
-    }{
-        \pi(\underline{x})
-    }
-\right).
-\end{aligned}
-}\label{eq-vi_step02}\end{equation} Again, using the properties of
-logarithms, we can split Equation~\ref{eq-vi_step02}, obtaining
-\begin{equation}\protect\hypertarget{eq-vi_step03}{}{
-\begin{aligned}
-D_{KL}\left(
-    q_\phi(\underline{\theta}) \vert\vert 
-    \pi(\underline{\theta} \mid \underline{x})
-\right) &= 
-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln q_\phi(\underline{\theta}) \\
-&-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln \pi(\underline{x} \mid \underline{\theta}) \\
-&-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln \pi(\underline{\theta}) \\
-&+\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln \pi(\underline{x}).
-\end{aligned}
-}\label{eq-vi_step03}\end{equation} It is convenient to write
-Equation~\ref{eq-vi_step03} as
-\begin{equation}\protect\hypertarget{eq-vi_step04}{}{
-\begin{aligned}
-D_{KL}\left(
-    q_\phi(\underline{\theta}) \vert\vert 
-    \pi(\underline{\theta} \mid \underline{x})
-\right) &= 
-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln \frac{
-    q_\phi(\underline{\theta})
-    }{
-        \pi(\underline{\theta})
-    } \\
-&-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-\ln \pi(\underline{x} \mid \underline{\theta}) \\
-&+ \ln \pi(\underline{x}) 
-\int d^K\underline{\theta}\; q_\phi(\underline{\theta}),
-\end{aligned}
-}\label{eq-vi_step04}\end{equation} where for the last term, we can take
-\(\ln \pi(\underline{x})\) out of the integral since it does not depend
-on \(\underline{\theta}\). Lastly, we utilize two properties:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  The proposed approximate distribution must be normalized, i.e.,
-  \begin{equation}\protect\hypertarget{eq-vi_q_norm}{}{
-  \int d^K\underline{\theta}\; q_\phi(\underline{\theta}) = 1.
-  }\label{eq-vi_q_norm}\end{equation}
-\item
-  The law of the unconscious statistician (LOTUS) establishes that for
-  any probability density function, it must be true that
-  \begin{equation}\protect\hypertarget{eq-vi_lotus}{}{
-  \int d^K\underline{\theta}\; q_\phi(\underline{\theta})
-  f(\underline{\theta}) = \left\langle 
-   f(\underline{\theta}) 
-  \right\rangle_{q_\phi},
-  }\label{eq-vi_lotus}\end{equation} where
-  \(\left\langle\cdot\right\rangle_{q_\phi}\) is the expected value over
-  the \(q_\phi\) distribution.
-\end{enumerate}
-
-Using these two properties, the positivity constraint on the KL
-divergence in Equation~\ref{eq-vi_kl_pos}, and the definition of the KL
-divergence in Equation~\ref{eq-vi_kl_div} we can rewrite
-Equation~\ref{eq-vi_step04} as
-\begin{equation}\protect\hypertarget{eq-vi_step05}{}{
-D_{KL}\left( 
-    q_\phi(\underline{\theta}) \vert \vert
-    \pi(\underline{\theta}) 
-\right) -
-\left\langle
-    \ln \pi(\underline{x} \mid \underline{\theta})
-\right\rangle_{q_\phi}
-\geq - \ln \pi(\underline{x}).
-}\label{eq-vi_step05}\end{equation} Multiplying by a minus one, we have
-the functional form of the so-called evidence lower bound (ELBO)
-\textcite{kingma2014},
-\begin{equation}\protect\hypertarget{eq-vi_elbo}{}{
-\underbrace{
-    \ln \pi(\underline{x})
-}_{\text{log evidence}} \geq
-\underbrace{
-    \left\langle
-        \ln \pi(\underline{x} \mid \underline{\theta})
-    \right\rangle_{q_\phi} -
-    D_{KL}\left( 
-        q_\phi(\underline{\theta}) \vert \vert
-        \pi(\underline{\theta}) 
-    \right)
-}_{\text{ELBO}}.
-}\label{eq-vi_elbo}\end{equation}
-
-Let us recapitulate where we are. We started by presenting the challenge
-of working with Bayes' theorem, as it requires a high-dimensional
-integral of the form in Equation~\ref{eq-vi_evidence_int}. As an
-alternative, variational inference posits to approximate the posterior
-distribution \(\pi(\underline{\theta} \mid \underline{x})\) with a
-parametric distribution \(q_\phi(\underline{\theta})\). By minimizing
-the KL divergence between these distributions, we arrive at the result
-in Equation~\ref{eq-vi_elbo}, where the left-hand side---the log
-marginalized likelihood or log evidence---we cannot compute for
-technical/computational reasons. However, the right-hand side is
-composed of things we can easily evaluate. We can easily evaluate the
-log-likelihood \(\ln \pi(\underline{x} \mid \underline{\theta})\) and
-the KL divergence between our proposed approximate distribution
-\(q_\phi(\underline{\theta})\) and the prior distribution
-\(\pi(\underline{\theta})\). Moreover, we can compute the gradients of
-these functions with respect to the parameters of our proposed
-distribution. This last point implies that we can change the parameters
-of the proposed distribution to maximize the ELBO. And, although we
-cannot compute the left-hand side of Equation~\ref{eq-vi_elbo}, we know
-that however large we make the ELBO, it will always be smaller than (or
-equal) the log-marginal likelihood. Therefore, the larger we can make
-the ELBO by modifying the parameters \(\phi\), the closer it gets to the
-log-marginal likelihood, and, as a consequence, the better our proposed
-distribution \(q_\phi(\underline{\theta})\) gets to the true posterior
-distribution \(\pi(\underline{\theta} \mid \underline{x})\).
-
-In this sense, variational inference turns the intractable numerical
-integration problem to an optimization routine, for which there are
-several algorithms available.
-
-\hypertarget{advi-algorithm}{%
-\subsubsection{ADVI algorithm}\label{advi-algorithm}}
-
-To maximize the right-hand side of Equation~\ref{eq-vi_elbo}, the
-Automatic Differentiation Variational Inference (ADVI) algorithm
-developed in \autocite{kucukelbir2016} takes advantage of advances in
-probabilistic programming languages to generate a robust method to
-perform this optimization. Without going into the details of the
-algorithm implementation, for our purposes, it suffices to say that we
-define our joint distribution \(\pi(\underline{\theta}, \underline{x})\)
-as the product defined in Equation~\ref{eq-vi_joint_dist}. ADVI then
-proposes an approximate variational distribution \(q_\phi\) that can
-either be a multivariate Normal distribution with a diagonal covariance
-matrix, i.e., \begin{equation}\protect\hypertarget{eq-vi_meanfield}{}{
-\phi = (\underline{\mu}, \underline{\underline{D}}),
-}\label{eq-vi_meanfield}\end{equation} where
-\(\underline{\underline{D}}\) is the identity matrix, with the diagonal
-elements given by the vector of variances \(\underline{\sigma}^2\) for
-each variable or a full-rank multivariate Normal distribution
-\begin{equation}\protect\hypertarget{eq-vi_full}{}{
-\phi = (\underline{\mu}, \underline{\underline{\Sigma}}).
-}\label{eq-vi_full}\end{equation}
-
-Then, the parameters are initialized in some value \(\phi_o\). These
-parameters are iteratively updated by computing the gradient of the ELBO
-(right-hand side of Equation~\ref{eq-vi_elbo}), hereafter defined as
-\(\mathcal{L}\), with respect to the parameters,
-\begin{equation}\protect\hypertarget{eq-vi_elbo_gradient}{}{
-\nabla_\phi \mathcal{L} = \nabla_{\underline{\mu}} \mathcal{L} + 
-\nabla_{\underline{\sigma}}\mathcal{L},
-}\label{eq-vi_elbo_gradient}\end{equation} and then computing \[
-\phi_{t+1} = \phi_{t} + \eta \nabla_\phi \mathcal{L},
-\] where \(\eta\) defines the step size.
-
-This short explanation behind the ADVI algorithm is intended only to
-gain intuition on how the optimal variational distribution \(q_\phi\) be
-computed. There are many nuances in the implementation of the ADVI
-algorithm. We invite the reader to look at the original reference for
-further details.
-
-\hypertarget{sec-bayes_def}{%
-\subsection{Defining the Bayesian model}\label{sec-bayes_def}}
-
-In the main text, we specify the inference problem we must solve as
-being of the form
-\begin{equation}\protect\hypertarget{eq-full_inference_SI}{}{
-\pi(
-    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
-    \underline{\underline{R}}
-) \propto
-\pi(
-    \underline{\underline{R}} \mid
-    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
-)
-\pi(
-    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
-).
-}\label{eq-full_inference_SI}\end{equation} Here, we briefly define the
-missing nuisance parameters. Let
-\begin{equation}\protect\hypertarget{eq-pop_fitness_vec}{}{
-\underline{\bar{s}}_T = (\bar{s}_1, \bar{s}_2, \ldots, \bar{s}_{T-1})^\dagger,
-}\label{eq-pop_fitness_vec}\end{equation} be the vector containing the
-\(T-1\) population mean fitness we compute from the \(T\) time points
-where measurements were taken. We have \(T-1\) since the value of any
-\(\bar{s}_t\) requires cycle numbers \(t\) and \(t+1\). Furthermore, let
-the matrix \(\underline{\underline{F}}\) be a \(T \times B\) matrix
-containing all frequency values. As with Equation~\ref{eq-R-mat_split}
-in the main text, we can split \(\underline{\underline{F}}\) into two
-matrices of the form
-\begin{equation}\protect\hypertarget{eq-F-mat_split}{}{
-\underline{\underline{F}} = \left[ 
-\underline{\underline{F}}^N \; \underline{\underline{F}}^M
-\right],
-}\label{eq-F-mat_split}\end{equation} to separate the corresponding
-neutral and non-neutral barcode frequencies.
-
-Let us now define each of the terms in Equation~\ref{eq-split_posterior}
-described in Section~\ref{sec-bayesian_inference} of the main text. The
-following sections will specify the functional form each of these terms
-takes.
-
-\hypertarget{sec-bayes_freq}{%
-\subsubsection{\texorpdfstring{Frequency uncertainty
-\(\pi(\underline{\underline{F}} \mid \underline{\underline{R}})\)}{Frequency uncertainty \textbackslash pi(\textbackslash underline\{\textbackslash underline\{F\}\} \textbackslash mid \textbackslash underline\{\textbackslash underline\{R\}\})}}\label{sec-bayes_freq}}
-
-We begin with the probability of the frequency values given the raw
-barcode reads. The first assumption is that the inference of the
-frequency values for time \(t\) is independent of any other time.
-Therefore, we can write the joint probability distribution as a product
-of independent distributions of the form
-\begin{equation}\protect\hypertarget{eq-freq_indep}{}{
-\pi(\underline{\underline{F}} \mid \underline{\underline{R}}) =
-\prod_{t=1}^T \pi(\underline{f}_t \mid \underline{r}_t),
-}\label{eq-freq_indep}\end{equation} where \(\underline{f}_t\) and
-\(\underline{r}_t\) are the \(t\)-th row of the matrix containing all of
-the measurements for time \(t\). We imagine that when the barcode reads
-are obtained via sequencing, the quantified number of reads is a Poisson
-sample from the ``true'' underlying number of barcodes within the pool.
-This translates to assuming that the number of reads for each barcode at
-any time point \(r^{(b)}_t\) is an independent Poisson random variable,
-i.e., \begin{equation}\protect\hypertarget{eq-freq_poisson_reads}{}{
-r^{(b)}_t \sim \operatorname{Poiss}(\lambda^{(b)}_t),
-}\label{eq-freq_poisson_reads}\end{equation} where the symbol
-``\(\sim\)'' is read ``distributed as.'' Furthermore, for a Poisson
-distribution, we have that
-\begin{equation}\protect\hypertarget{eq-freq_poisson_lambda}{}{
-\lambda^{(b)}_t = \left\langle r^{(b)}_t \right\rangle = 
-\left\langle 
-    \left( r^{(b)}_t - \left\langle r^{(b)}_t \right\rangle \right)^2
-\right\rangle,
-}\label{eq-freq_poisson_lambda}\end{equation} where
-\(\left\langle \cdot \right\rangle\) is the expected value. In other
-words the Poisson parameter is equal to the mean and variance of the
-distribution. The Poisson distribution has the convenient property that
-for two Poisson distributed random variables
-\(X \sim \operatorname{Poiss}(\lambda_x)\) and
-\(Y \sim \operatorname{Poiss}(\lambda_y)\), we have that
-\begin{equation}\protect\hypertarget{eq-freq_additivity}{}{
-Z \equiv X + Y \sim \operatorname{Poiss}(\lambda_x + \lambda_y).
-}\label{eq-freq_additivity}\end{equation} This additivity allows us to
-write the total number of reads at time \(t\) \(n_t\) also as a
-Poisson-distributed random variable of the form
-\begin{equation}\protect\hypertarget{eq-freq_poisson_total}{}{
-n_t \sim \operatorname{Poiss}\left( \sum_{b=1}^B \lambda^{(b)}_t \right),
-}\label{eq-freq_poisson_total}\end{equation} where the sum is taken over
-all \(B\) barcodes.
-
-If the total number of reads is given by
-Equation~\ref{eq-freq_poisson_total}, the array with the number of reads
-for each barcode at time \(t\), \(\underline{r}_t\) is then distributed
-as \begin{equation}\protect\hypertarget{eq-freq_multinomial}{}{
-\underline{r}_t \sim \operatorname{Multinomial}(n_t, \underline{f}_t),
-}\label{eq-freq_multinomial}\end{equation} where each of the \(B\)
-entries of the frequency vector \(\underline{f}_t\) is a function of the
-\(\underline{\lambda}_t\) vector, given by
-\begin{equation}\protect\hypertarget{eq-freq_freq_lambda}{}{
-f_t^{(b)} \equiv f_t^{(b)}(\underline{\lambda}_t) = 
-\frac{\lambda_t^{(b)}}{\sum_{b'=1}^B \lambda_t^{(b')}}.
-}\label{eq-freq_freq_lambda}\end{equation} In other words, we can think
-of the \(B\) barcode counts as independent Poisson samples or as a
-single multinomial draw with a random number of total draws, \(n_t\),
-and the frequency vector \(\underline{f}_t\) we are interested in.
-Notice that Equation~\ref{eq-freq_freq_lambda} is a deterministic
-function that connects the Poisson parameters to the frequencies.
-Therefore, we have the equivalence that
-\begin{equation}\protect\hypertarget{eq-freq_f_lambda_equiv}{}{
-\pi(\underline{f}_t \mid \underline{r}_t) = 
-\pi(\underline{\lambda}_t \mid \underline{r}_t),
-}\label{eq-freq_f_lambda_equiv}\end{equation} meaning that the
-uncertainty comes from the \(\underline{\lambda}_t\) vector. By Bayes
-theorem, we therefore write
-\begin{equation}\protect\hypertarget{eq-freq_lambda_bayes}{}{
-\pi(\underline{\lambda}_t \mid n_t, \underline{r}_t) \propto
-\pi(n_t, \underline{r}_t \mid \underline{\lambda}_t) \pi(\underline{\lambda}_t),
-}\label{eq-freq_lambda_bayes}\end{equation} where we explicitly include
-the dependence on \(n_t\). This does not affect the distribution or
-brings more uncertainty because \(\underline{r}_t\) already contains all
-the information to compute \(n_t\) since
-\begin{equation}\protect\hypertarget{eq-freq_n_sum_r}{}{
-n_t = \sum_{b=1}^B r_t^{(b)}.
-}\label{eq-freq_n_sum_r}\end{equation} But adding the variable allows us
-to factorize Equation~\ref{eq-freq_lambda_bayes} as
-\begin{equation}\protect\hypertarget{eq-freq_lambda_bayes_factorized}{}{
-\pi(\underline{\lambda}_t \mid n_t, \underline{r}_t) \propto
-\pi(\underline{r}_t \mid n_t, \underline{\lambda}_t)
-\pi(n_t \mid \underline{\lambda}_t)
-\pi(\underline{\lambda}_t)
-}\label{eq-freq_lambda_bayes_factorized}\end{equation} We then have
-\begin{equation}\protect\hypertarget{eq-freq_r_bayes}{}{
-\underline{r}_t \mid n_t, \underline{\lambda}_t \sim
-\operatorname{Multinomial}(n_t, \underline{f}_t(\underline{\lambda}_t)).
-}\label{eq-freq_r_bayes}\end{equation} Furthermore, we have \[
-n_t \mid \underline{\lambda}_t \sim 
-\operatorname{Poiss}\left(\sum_{b=1}^B \lambda_t^{(b)}\right).
-\]\{\#eq=freq\_n\_bayes\} Finally, for our prior
-\(\pi(\underline{\lambda}_t)\), we first assume each parameter is
-independent, i.e., \[
-\pi(\underline{\lambda}_t) = \prod_{b=1}^B \pi(\lambda_t^{(b)}).
-\] A reasonable prior for each \(\lambda_t^{(b)}\) representing the
-expected number of reads for barcode \(b\) should span several orders of
-magnitude. Furthermore, we assume that no barcode in the dataset ever
-goes extinct. Thus, no frequency can equal zero, facilitating the
-computation of the log frequency ratios needed to infer the relative
-fitness. The log-normal distribution satisfies these constraints;
-therefore, for the prior, we assume
-\begin{equation}\protect\hypertarget{eq-freq_lambda_prior}{}{
-\lambda_t^{(b)} \sim 
-\log\mathcal{N}(\mu_{\lambda_t^{(b)}}, \sigma_{\lambda_t^{(b)}}),
-}\label{eq-freq_lambda_prior}\end{equation} with
-\(\mu_{\lambda_t^{(b)}}, \sigma_{\lambda_t^{(b)}}\) as the user-defined
-parameters that characterize the prior distribution.
-
-\hypertarget{summary}{%
-\paragraph{Summary}\label{summary}}
-
-Putting all the pieces developed in this section together gives a term
-for our inference of the form
-\begin{equation}\protect\hypertarget{eq-freq_final_01}{}{
-\pi(\underline{\underline{F}} \mid \underline{\underline{R}}) \propto
-\prod_{t=1}^T\left\{
-    \pi(\underline{r}_t \mid n_t, \underline{\lambda}_t)
-    \pi(n_t \mid \underline{\lambda}_t)
-    \left[ 
-        \prod_{b=1}^B \pi(\lambda_t^{(b)})
-    \right]
-\right\}
-}\label{eq-freq_final_01}\end{equation} where
-\begin{equation}\protect\hypertarget{eq-freq_final_02}{}{
-\underline{r}_t \mid n_t, \underline{\lambda}_t \sim
-\operatorname{Multinomial}(n_t, \underline{f}_t(\underline{\lambda}_t)),
-}\label{eq-freq_final_02}\end{equation}
-\begin{equation}\protect\hypertarget{eq-freq_final_03}{}{
-n_t \mid \underline{\lambda}_t \sim 
-\operatorname{Poiss}\left(\sum_{b=1}^B \lambda_t^{(b)}\right).
-}\label{eq-freq_final_03}\end{equation} and
-\begin{equation}\protect\hypertarget{eq-freq_final_04}{}{
-\lambda_t^{(b)} \sim 
-\log\mathcal{N}(\mu_{\lambda_t^{(b)}}, \sigma_{\lambda_t^{(b)}}),
-}\label{eq-freq_final_04}\end{equation}
-
-\hypertarget{sec-bayes_meanfit}{%
-\subsubsection{\texorpdfstring{Population mean fitness uncertainty
-\(\pi(\underline{\bar{s}}_T \mid \underline{\underline{F}}, \underline{\underline{R}})\)}{Population mean fitness uncertainty \textbackslash pi(\textbackslash underline\{\textbackslash bar\{s\}\}\_T \textbackslash mid \textbackslash underline\{\textbackslash underline\{F\}\}, \textbackslash underline\{\textbackslash underline\{R\}\})}}\label{sec-bayes_meanfit}}
-
-Next, we turn our attention to the problem of determining the population
-mean fitnesses \(\underline{\bar{s}}_T\). First, we notice that our
-fitness model in Equation~\ref{eq-fitness} does not include the value of
-the raw reads. They enter the calculation indirectly through the
-inference of the frequency values we developed in
-Section~\ref{sec-bayes_freq}. This means that we can remove the
-conditioning of the value of \(\underline{\bar{s}}_T\) on the number of
-reads, obtaining a simpler probability function
-\begin{equation}\protect\hypertarget{eq-meanfit_noreads}{}{
-\pi(
-    \underline{\bar{s}}_T \mid 
-    \underline{\underline{F}}, \underline{\underline{R}}
-) = 
-\pi(
-    \underline{\bar{s}}_T \mid 
-    \underline{\underline{F}}
-).
-}\label{eq-meanfit_noreads}\end{equation} Moreover, our fitness model
-does not directly explain how the population mean fitness evolves over
-time. In other words, our model cannot explicitly compute the population
-mean fitness at time \(t+1\) from the information we have about time
-\(t\). Given this model limitation, we are led to assume that we must
-infer each \(\bar{s}_t\) independently. Expressing this for our
-inference results in
-\begin{equation}\protect\hypertarget{eq-meanfit_indep}{}{
-\pi(
-    \underline{\bar{s}}_T \mid 
-    \underline{\underline{F}}
-) =
-\prod_{t=1}^{T-1} \pi(\bar{s}_t \mid \underline{f}_t, \underline{f}_{t+1}),
-}\label{eq-meanfit_indep}\end{equation} where we split our matrix
-\(\underline{\underline{F}}\) for each time point and only kept the
-conditioning on the relevant frequencies needed to compute the mean
-fitness at time \(t\).
-
-Although our fitness model in Equation~\ref{eq-fitness} also includes
-the relative fitness \(s^{(m)}\), to infer the population mean fitness
-we only utilize data from the neutral lineages that, by definition, have
-a relative fitness \(s^{(n)} = 0\). Therefore, the conditioning on
-Equation~\ref{eq-meanfit_indep} can be further simplified by only
-keeping the frequencies of the neutral lineages, i.e.,
-\begin{equation}\protect\hypertarget{eq-meanfit_indep_neutral}{}{
-\pi(\bar{s}_t \mid \underline{f}_t, \underline{f}_{t+1}) =
-\pi(\bar{s}_t \mid \underline{f}_t^N, \underline{f}_{t+1}^N).
-}\label{eq-meanfit_indep_neutral}\end{equation}
-
-Recall that in Section~\ref{sec-fitness_model} we emphasized that the
-frequencies \(f_t^{(n)}\) do not represent the true frequency of a
-particular lineage in the population but rather a ``normalized number of
-cells.'' Therefore, it is safe to assume each of the \(N\) neutral
-lineages' frequencies is changing independently. The correlation of how
-increasing the frequency of one lineage will decrease the frequency of
-others is already captured in the model presented in
-Section~\ref{sec-bayes_freq}. Thus, we write
-\begin{equation}\protect\hypertarget{eq-meanfit_indep_lineages}{}{
-\pi(\bar{s}_t \mid \underline{f}_t^N, \underline{f}_{t+1}^N) =
-\prod_{n=1}^N \pi(\bar{s}_t \mid f_t^{(n)}, f_{t+1}^{(n)}).
-}\label{eq-meanfit_indep_lineages}\end{equation}
-
-Now, we can focus on one of the terms on the right-hand side of
-Equation~\ref{eq-meanfit_indep_lineages}. Writing Bayes theorem results
-in \begin{equation}\protect\hypertarget{eq-meanfit_bayes}{}{
-\pi(\bar{s}_t \mid f_t^{(n)}, f_{t+1}^{(n)}) \propto
-\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t) \pi(\bar{s}_t).
-}\label{eq-meanfit_bayes}\end{equation} Notice the likelihood defines
-the joint distribution of neutral barcode frequencies conditioned on the
-population mean fitness. However, rewriting our fitness model in
-Equation~\ref{eq-fitness} for a neutral lineage to leave frequencies on
-one side and fitness on the other results in
-\begin{equation}\protect\hypertarget{eq-fitness_ratio_neutral}{}{
-\frac{f_{t+1}^{(n)}}{f_t^{(n)}} = \mathrm{e}^{- \bar{s}_t\tau}.
-}\label{eq-fitness_ratio_neutral}\end{equation}
-Equation~\ref{eq-fitness_ratio_neutral} implies that our fitness model
-only relates \textbf{the ratio} of frequencies and not the individual
-values. To get around this complication, we define
-\begin{equation}\protect\hypertarget{eq-gamma_def}{}{
-\gamma_t^{(b)} \equiv \frac{f_{t+1}^{(b)}}{f_t^{(b)}},
-}\label{eq-gamma_def}\end{equation} as the ratio of frequencies between
-two adjacent time points for any barcode \(b\). This allows us to
-rewrite the joint distribution
-\(\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t)\) as
-\begin{equation}\protect\hypertarget{eq-joint_freq_gamma}{}{
-\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t) =
-\pi(f_t^{(n)}, \gamma_{t}^{(n)} \mid \bar{s}_t).
-}\label{eq-joint_freq_gamma}\end{equation} Let us rephrase this subtle
-but necessary change of variables since it is a key part of the
-inference problem: our series of independence assumptions lead us to
-Equation~\ref{eq-meanfit_bayes} that relates the value of the population
-mean fitness \(\bar{s}_t\) to the frequency of a neutral barcode at
-times \(t\) and \(t+1\). However, as shown in
-Equation~\ref{eq-fitness_ratio_neutral}, our model functionally relates
-the ratio of frequencies---that we defined as \(\gamma_t^{(n)}\)---and
-not the independent frequencies to the mean fitness. Therefore, instead
-of writing for the likelihood the joint distribution of the frequency
-values at times \(t\) and \(t+1\) conditioned on the mean fitness, we
-write the joint distribution of the barcode frequency at time \(t\) and
-the ratio of the frequencies. These \textbf{must be} equivalent joint
-distributions since there is a one-to-one mapping between
-\(\gamma_t^{(n)}\) and \(f_{t+1}^{(n)}\) for a given value of
-\(f_t^{(n)}\). Another way to phrase this is to say that knowing the
-frequency at time \(t\) and at time \(t+1\) provides the same amount of
-information as knowing the frequency at time \(t\) and the ratio of the
-frequencies. This is because if we want to obtain \(f_{t+1}^{(n)}\)
-given this information, we simply compute
-\begin{equation}\protect\hypertarget{eq-meanfit_ft+1_gamma}{}{
-f_{t+1}^{(n)} = \gamma_t^{(n)} f_t^{(n)}.
-}\label{eq-meanfit_ft+1_gamma}\end{equation}
-
-The real advantage of rewriting the joint distribution as in
-Equation~\ref{eq-joint_freq_gamma} comes from splitting this joint
-distribution as a product of conditional distributions of the form
-\begin{equation}\protect\hypertarget{eq-joint_to_prod_gamma}{}{
-\pi(f_t^{(n)}, \gamma_{t}^{(n)} \mid \bar{s}_t) =
-\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t)
-\pi(\gamma_{t}^{(n)} \mid \bar{s}_t).
-}\label{eq-joint_to_prod_gamma}\end{equation} Written in this form, we
-can finally propose a probabilistic model for how the mean fitness
-relates to the frequency ratios we determine in our experiments. The
-second term on the right-hand side of
-Equation~\ref{eq-joint_to_prod_gamma} relates how the determined
-frequency ratio \(\gamma_t^{(b)}\) relates to the mean fitness
-\(\bar{s}_t\). From Equation~\ref{eq-fitness_ratio_neutral} and
-Equation~\ref{eq-gamma_def}, we can write
-\begin{equation}\protect\hypertarget{eq-log_ratio_neutral}{}{
-\ln \gamma_t^{(n)} = - \bar{s}_t + \varepsilon_t^{(n)},
-}\label{eq-log_ratio_neutral}\end{equation} where, for simplicity, we
-set \(\tau = 1\). Note that we added an extra term,
-\(\varepsilon_t^{(n)}\), characterizing the deviations of the
-measurements from the theoretical model. We assume these errors are
-normally distributed with mean zero and some standard deviation
-\(\sigma_t\), implying that
-\begin{equation}\protect\hypertarget{eq-log_gamma_normal}{}{
-\ln \gamma_t^{(n)} \mid \bar{s}_t, \sigma_t  \sim 
-\mathcal{N}\left(-\bar{s}_t, \sigma_t \right),
-}\label{eq-log_gamma_normal}\end{equation} where we include the nuisance
-parameter \(\sigma_t\) to be determined. If we assume the log frequency
-ratio is normally distributed, this implies the frequency ratio itself
-is distributed log-normal. This means that
-\begin{equation}\protect\hypertarget{eq-meanfit_likelihood}{}{
-\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t  \sim 
-\log \mathcal{N}\left(-\bar{s}_t, \sigma_t \right).
-}\label{eq-meanfit_likelihood}\end{equation} Having added the nuisance
-parameter \(\sigma_t\) implies that we must update
-Equation~\ref{eq-meanfit_bayes} to
-\begin{equation}\protect\hypertarget{eq-meanfit_bayes_full}{}{
-\pi(\bar{s}_t, \sigma_t \mid f_t^{(n)}, f_{t+1}^{(n)}) \propto
-\pi(f_t^{(n)}, \gamma_t^{(n)} \mid \bar{s}_t, \sigma_t) 
-\pi(\bar{s}_t) \pi(\sigma_t),
-}\label{eq-meanfit_bayes_full}\end{equation} where we assume the prior
-for each parameter is independent, i.e.,
-\begin{equation}\protect\hypertarget{eq-meanfit_indep_prior}{}{
-\pi(\bar{s}_t, \sigma_t) = \pi(\bar{s}_t) \pi(\sigma_t).
-}\label{eq-meanfit_indep_prior}\end{equation} For numerical stability,
-we will select weakly-informative priors for both of these parameters.
-In the case of the nuisance parameter \(\sigma_t\), the prior must be
-restricted to positive values only, since standard deviations cannot be
-negative.
-
-For the first term on the right-hand side of
-Equation~\ref{eq-joint_to_prod_gamma},
-\(\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t)\), we remove the
-conditioning on the population mean fitness since it does not add any
-information on top of what the frequency ratio \(\gamma_t^{(n)}\)
-already gives. Therefore, we have
-\begin{equation}\protect\hypertarget{eq-freq_cond_gamma}{}{
-\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t) =
-\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}).
-}\label{eq-freq_cond_gamma}\end{equation} The right-hand side of
-Equation~\ref{eq-freq_cond_gamma} asks us to compute the probability of
-observing a frequency value \(f_t^{(n)}\) given that we get to observe
-the ratio \(\gamma_{t}^{(n)}\). If the ratio happened to be
-\(\gamma_{t}^{(n)} = 2\), we could have \(f_{t+1}^{(n)} = 1\) and
-\(f_{t+1}^{(n)} = 0.5\), for example. Although, it would be equally
-likely that \(f_{t+1}^{(n)} = 0.6\) and \(f_{t+1}^{(n)} = 0.3\) or
-\(f_{t+1}^{(n)} = 0.1\) and \(f_{t+1}^{(n)} = 0.05\) for that matter. If
-we only get to observe the frequency ratio \(\gamma_t^{(n)}\), we know
-that the numerator \(f_{t+1}^{(n)}\) can only take values between zero
-and one, all of them being equally likely given only the information on
-the ratio. As a consequence, the value of the frequency in the
-denominator \(f_{t}^{(n)}\) is restricted to fall in the range
-\begin{equation}\protect\hypertarget{eq-freq_given_gamma_range}{}{
-f_{t}^{(n)} \in \left(0, \frac{1}{\gamma_t^{(n)}} \right].
-}\label{eq-freq_given_gamma_range}\end{equation} A priori, we do not
-have any reason to favor any value over any other, therefore it is
-natural to write
-\begin{equation}\protect\hypertarget{eq-f_given_gamma_uniform}{}{
-f_t^{(n)} \mid \gamma_t^{(n)} \sim 
-\operatorname{Uniform}\left( 0, \frac{1}{\gamma_t^{(n)}} \right).
-}\label{eq-f_given_gamma_uniform}\end{equation}
-
-\hypertarget{summary-1}{%
-\paragraph{Summary}\label{summary-1}}
-
-Putting all the pieces we have developed in this section together
-results in an inference for the population mean fitness values of the
-form \begin{equation}\protect\hypertarget{eq-meanfit_full_inference}{}{
-\pi(
-    \underline{\bar{s}}_T, \underline{\sigma}_T \mid \underline{\underline{F}}
-) \propto
-\prod_{t=1}^{T-1} \left\{
-    \prod_{n=1}^N \left[
-        \pi(f_t^{(n)} \mid \gamma_t^{(n)}) 
-        \pi(\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t)
-    \right]
-    \pi(\bar{s}_t) \pi(\sigma_t)
-\right\},
-}\label{eq-meanfit_full_inference}\end{equation} where we have
-\begin{equation}\protect\hypertarget{eq-meanfit_eq1}{}{
-f_t^{(n)} \mid \gamma_t^{(n)} \sim 
-\operatorname{Uniform} \left(0, \frac{1}{\gamma_t^{(n)}} \right),
-}\label{eq-meanfit_eq1}\end{equation}
-\begin{equation}\protect\hypertarget{eq-meanfit_eq2}{}{
-\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t \sim 
-\log\mathcal{N}(\bar{s}_t, \sigma_t),
-}\label{eq-meanfit_eq2}\end{equation}
-\begin{equation}\protect\hypertarget{eq-meanfit_eq3}{}{
-\bar{s}_t \sim \mathcal{N}(0, \sigma_{\bar{s}_t}),
-}\label{eq-meanfit_eq3}\end{equation} and
-\begin{equation}\protect\hypertarget{eq-meanfit_eq4}{}{
-\sigma_t \sim \log\mathcal{N}(\mu_{\sigma_t}, \sigma_{\sigma_t}),
-}\label{eq-meanfit_eq4}\end{equation} where \(\sigma_{\bar{s}_t}\),
-\(\mu_{\sigma_t}\), and \(\sigma_{\sigma_t}\) are user-defined
-parameters.
-
-\hypertarget{sec-bayes_mutfit}{%
-\subsubsection{\texorpdfstring{Mutant relative fitness uncertainty
-\(\pi(\underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}}, \underline{\underline{R}})\)}{Mutant relative fitness uncertainty \textbackslash pi(\textbackslash underline\{s\}\^{}M \textbackslash mid \textbackslash underline\{\textbackslash bar\{s\}\}\_T, \textbackslash underline\{\textbackslash underline\{F\}\}, \textbackslash underline\{\textbackslash underline\{R\}\})}}\label{sec-bayes_mutfit}}
-
-The last piece of our inference is the piece that we care about the
-most: the probability distribution of all the mutants' relative fitness,
-given the inferred population mean fitness and the frequencies. First,
-we assume that all fitness values are independent of each other. This
-allows us to write
-\begin{equation}\protect\hypertarget{eq-mutfit_indep}{}{
-\pi(
-    \underline{s}^M \mid 
-    \underline{\bar{s}}_T, \underline{\underline{F}}, \underline{\underline{R}}
-) = 
-\prod_{m=1}^M \pi(
-    s^{(m)} \mid
-    \underline{\bar{s}}_T, \underline{\underline{F}}, \underline{\underline{R}}
-).
-}\label{eq-mutfit_indep}\end{equation} Furthermore, as was the case with
-the population mean fitness, our fitness model relates frequencies, not
-raw reads. Moreover, the fitness value of mutant \(m\) only depends on
-the frequencies of such mutant. Therefore, we can simplify the
-conditioning to
-\begin{equation}\protect\hypertarget{eq-mutfit_cond_simple}{}{
-\pi(
-    s^{(m)} \mid
-    \underline{\bar{s}}_T, \underline{\underline{F}}, \underline{\underline{R}}
-) = 
-\pi(s^{(m)} \mid \underline{\bar{s}}_T, \underline{f}^{(m)}),
-}\label{eq-mutfit_cond_simple}\end{equation} where
-\begin{equation}\protect\hypertarget{eq-mutfit_timeseries_def}{}{
-\underline{f}^{(m)} = (f_0^{(m)}, f_1^{(m)}, \ldots, f_T^{(m)})^\dagger,
-}\label{eq-mutfit_timeseries_def}\end{equation} is the vector containing
-the frequency time series for mutant \(m\). Writing Bayes' theorem for
-the right-hand side of Equation~\ref{eq-mutfit_cond_simple} results in
-\begin{equation}\protect\hypertarget{eq-mutfit_bayes}{}{
-\pi(s^{(m)} \mid \underline{\bar{s}}_T, \underline{f}^{(m)}) \propto
-\pi(\underline{f}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)})
-\pi(s^{(m)} \mid \underline{\bar{s}}_T).
-}\label{eq-mutfit_bayes}\end{equation} Notice the conditioning on the
-mean fitness values \(\underline{\bar{s}}_T\) is not inverted since we
-already inferred these values.
-
-Following the logic used in Section~\ref{sec-bayes_meanfit}, let us
-define \begin{equation}\protect\hypertarget{eq-gamma_vec_def}{}{
-\underline{\gamma}^{(m)} = 
-(\gamma_0^{(m)}, \gamma_1^{(m)}, \ldots, \gamma_{T-1}^{m})^\dagger,
-}\label{eq-gamma_vec_def}\end{equation} where each entry
-\(\gamma_t^{(m)}\) is defined by Equation~\ref{eq-gamma_def}. In the
-same way we rewrote the joint distribution between two adjacent time
-point frequencies to the joint distribution between one of the
-frequencies and the ratio of both frequencies in
-Equation~\ref{eq-joint_freq_gamma}, we can rewrite the joint
-distribution of the frequency time series for mutant \(m\) as
-\begin{equation}\protect\hypertarget{eq-joint_freq_gammas_mutfit}{}{
-\pi(\underline{f}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}) =
-\pi(f_0^{(m)}, \underline{\gamma}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}).
-}\label{eq-joint_freq_gammas_mutfit}\end{equation} One can think about
-Equation~\ref{eq-joint_freq_gammas_mutfit} as saying that knowing the
-individual frequencies at each time point contain equivalent information
-as knowing the initial frequency and the subsequent ratios of
-frequencies. This is because if we want to know the value of
-\(f_1^{(m)}\) given the ratios, we only need to compute
-\begin{equation}\protect\hypertarget{eq-f1_from_ratio}{}{
-f_1^{(m)} = \gamma_0^{(m)} f_0^{(m)}.
-}\label{eq-f1_from_ratio}\end{equation} Moreover, if we want to know
-\(f_2^{(m)}\), we have
-\begin{equation}\protect\hypertarget{eq-f2_from_ratio}{}{
-f_2^{(m)} = \gamma_1^{(m)} f_1^{(m)} =
-\gamma_1^{(m)} \left(\gamma_0^{(m)} f_0^{(m)}\right),
-}\label{eq-f2_from_ratio}\end{equation} and so on. We can then write the
-joint distribution on the right-hand side of
-Equation~\ref{eq-joint_freq_gammas_mutfit} as a product of conditional
-distributions of the form
-\begin{equation}\protect\hypertarget{eq-mutfit_joint_to_cond}{}{
-\begin{aligned}
-\pi(f_0^{(m)}, \underline{\gamma}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}) =
-&\pi(
-    f_0^{(m)} \mid 
-    \gamma_0^{(m)}, \ldots, \gamma_{T-1}^{(m)}, \underline{\bar{s}}_T, s^{(m)}
-) \times \\
-&\pi(
-    \gamma_0^{(m)} \mid 
-    \gamma_1^{(m)}, \ldots, \gamma_{T-1}^{(m)}, \underline{\bar{s}}_T, s^{(m)}
-) \times \\
-&\pi(
-    \gamma_1^{(m)} \mid 
-    \gamma_2^{(m)}, \ldots, \gamma_{T-1}^{(m)}, \underline{\bar{s}}_T, s^{(m)}
-) \times \\
-&\vdots \\
-&\pi(
-    \gamma_{T-2}^{(m)} \mid \gamma_{T-1}^{(m)}, \underline{\bar{s}}_T, s^{(m)}
-) \times \\
-&\pi(\gamma_{T-1}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}).
-\end{aligned}
-}\label{eq-mutfit_joint_to_cond}\end{equation} Writing the fitness model
-in Equation~\ref{eq-fitness} as \[
-\gamma_t^{(m)} = \frac{f_{t+1}^{(m)}}{f_t^{(m)}} = 
-\mathrm{e}^{(s^{(m)} - s_t)\tau},
-\] reveals that the value of each of the ratios \(\gamma_t^{(m)}\) only
-depends on the corresponding fitness value \(\bar{s}_t\) and the
-relative fitness \(s^{(m)}\). Therefore, we can remove most of the
-conditioning on the right-hand side of
-Equation~\ref{eq-mutfit_joint_to_cond}, resulting in a much simpler
-joint distribution of the form
-\begin{equation}\protect\hypertarget{eq-mutfit_joint_to_cond_simple}{}{
-\begin{aligned}
-\pi(f_0^{(m)}, \underline{\gamma}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}) =
-&\pi(f_0^{(m)} \mid \gamma_0^{(m)}) \times \\
-&\pi(\gamma_0^{(m)} \mid \bar{s}_0, s^{(m)}) \times \\
-&\pi(\gamma_1^{(m)} \mid \bar{s}_1, s^{(m)}) \times \\
-&\vdots \\
-&\pi(\gamma_{T-2}^{(m)} \mid \bar{s}_{T-2}, s^{(m)}) \times \\
-&\pi(\gamma_{T-1}^{(m)} \mid \bar{s}_{T-1}, s^{(m)}),
-\end{aligned}
-}\label{eq-mutfit_joint_to_cond_simple}\end{equation} where for the
-first term on the right-hand side of
-Equation~\ref{eq-mutfit_joint_to_cond_simple} we apply the same logic as
-in Equation~\ref{eq-freq_cond_gamma} to remove all other dependencies.
-We emphasize that although Equation~\ref{eq-mutfit_joint_to_cond_simple}
-looks like a series of independent inferences, the value of the relative
-fitness \(s^{(m)}\) is shared among all of them. This means that the
-parameter is not inferred individually for each time point, resulting in
-different estimates of the parameter, but each time point contributes
-independently to the inference of a single estimate of \(s^{(m)}\).
-
-Using equivalent arguments to those in Section~\ref{sec-bayes_meanfit},
-we assume \[
-f_0^{(m)} \mid \gamma_0^{(m)} \sim 
-\operatorname{Uniform}\left(0, \frac{1}{\gamma_0^{(m)}} \right),
-\] and \begin{equation}\protect\hypertarget{eq-mutfit_lognormal}{}{
-\gamma_t^{(m)} \mid \bar{s}_t, s^{(m)}, \sigma^{(m)} \sim 
-\log\mathcal{N}\left(s^{(m)} - \bar{s}_t, \sigma^{(m)} \right),
-}\label{eq-mutfit_lognormal}\end{equation} where we add the nuisance
-parameter \(\sigma^{(m)}\) to the inference. Notice that this parameter
-is not indexed by time. This means that we assume the deviations from
-the theoretical prediction do not depend on time, but only on the
-mutant. Adding the nuisance parameter demands us to update
-Equation~\ref{eq-mutfit_bayes} to
-\begin{equation}\protect\hypertarget{eq-mutfit_bayes_full}{}{
-\pi(
-    s^{(m)}, \sigma^{(m)} \mid \underline{\bar{s}}_T, \underline{f}^{(m)}
-) \propto
-\pi(\underline{f}^{(m)} \mid \underline{\bar{s}}_T, s^{(m)}, \sigma^{(m)})
-\pi(s^{(m)}) \pi(\sigma^{(m)}),
-}\label{eq-mutfit_bayes_full}\end{equation} where we assume independent
-priors for both parameters. We also removed the conditioning on the
-values of the mean fitness as knowing such values does not change our
-prior information about the possible range of values that the parameters
-can take. As with the priors on Section~\ref{sec-bayes_meanfit}, we will
-assign weakly-informative priors to these parameters.
-
-\hypertarget{summary-2}{%
-\paragraph{Summary}\label{summary-2}}
-
-With all pieces in place, we write the full inference of the relative
-fitness values as
-\begin{equation}\protect\hypertarget{eq-mutfit_full_inference}{}{
-\pi(
-    \underline{s}^M ,\underline{\sigma}^M \mid 
-    \underline{\bar{s}}_T, \underline{\underline{F}}
-) \propto
-\prod_{m=1}^M \left\{ 
-    \pi(f_0^{(m)} \mid \gamma_0^{(m)})
-    \prod_{t=0}^{T-1} \left[
-        \pi(\gamma_t^{(m)} \mid \bar{s}_t, s^{(m)}, \sigma^{(m)})
-    \right]
-    \pi(s^{(m)}) \pi(\sigma^{(m)})
-\right\},
-}\label{eq-mutfit_full_inference}\end{equation} where
-\begin{equation}\protect\hypertarget{eq-mutfit_eq1}{}{
-f_0^{(m)} \mid \gamma_0^{(m)} \sim 
-\operatorname{Uniform}\left(0, \frac{1}{\gamma_0^{(m)}} \right),
-}\label{eq-mutfit_eq1}\end{equation}
-\begin{equation}\protect\hypertarget{eq-mutfit_eq2}{}{
-\gamma_t^{(m)} \mid \bar{s}_t, s^{(m)}, \sigma^{(m)} \sim 
-\log\mathcal{N}\left(s^{(m)} - \bar{s}_t, \sigma^{(m)} \right),
-}\label{eq-mutfit_eq2}\end{equation}
-\begin{equation}\protect\hypertarget{eq-mutfit_eq3}{}{
-s^{(m)} \sim \mathcal{N}(0, \sigma_{s^{(m)}}),
-}\label{eq-mutfit_eq3}\end{equation} and
-\begin{equation}\protect\hypertarget{eq-mutfit_eq4}{}{
-\sigma^{(m)} \sim \log\mathcal{N}(\mu_{\sigma^{(m)}}, \sigma_{\sigma^{(m)}}),
-}\label{eq-mutfit_eq4}\end{equation} where \(\sigma_{s^{(m)}}\),
-\(\mu_{\sigma^{(m)}}\), and \(\sigma_{\sigma^{(m)}}\) are user-defined
-parameters.
-
-\hypertarget{sec-hierarchical_model}{%
-\subsubsection{Hierarchical models for multiple experimental
-replicates}\label{sec-hierarchical_model}}
-
-As detailed in Section~\ref{sec-replicates} of the main text, we define
-a Bayesian hierarchical model to analyze data from multiple experimental
-replicates. The implementation requires only slightly modifying the base
-model detailed in the previous sections. The hierarchical model defines
-a hyper-fitness parameter \(\theta^{(m)}\) for every non-neutral
-barcode. We can thus collect all of the \(M\) hyperparameters in an
-array of the form
-\begin{equation}\protect\hypertarget{eq-hier_hyper_vector}{}{
-\underline{\theta}^M = (\theta^{(1)}, \ldots, \theta^{(M)})^\dagger.
-}\label{eq-hier_hyper_vector}\end{equation} Our data now consists of a
-series of matrices \(\underline{\underline{R}}_{[j]}\), where the
-subindex \([j]\) refers to the \(j\)-th experimental replicate. These
-matrices need not have the same number of rows, as the time points
-measured for each replicate can vary. The statistical model we must
-define is then of the form
-\begin{equation}\protect\hypertarget{eq-hier_bayes_full}{}{
-\begin{aligned}
-\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\},
-    \{\underline{\underline{F}}_{[j]}\} \mid
-    \{\underline{\underline{R}}_{[j]}\}
-) \propto \;
-&\pi(
-    \{\underline{\underline{R}}_{[j]}\} \mid
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\},
-    \{\underline{\underline{F}}_{[j]}\} 
-) \times \\\
-&\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\},
-    \{\underline{\underline{F}}_{[j]}\} 
-)
-\end{aligned}
-}\label{eq-hier_bayes_full}\end{equation} where the parameters within
-curly braces with subindex \([j]\) indicate one set of parameters per
-experimental replicate. For example,
-\begin{equation}\protect\hypertarget{eq-hier_array}{}{
-\{\underline{s}^M_{[j]}\} = \{
-    \underline{s}^M_{[1]}, \underline{s}^M_{[2]}, \ldots \underline{s}^M_{[E]}
-\},
-}\label{eq-hier_array}\end{equation} where \(E\) is the number of
-experimental replicates.
-
-Given the dependencies between the variables, we can factorize
-Equation~\ref{eq-hier_bayes_full} to be of the form
-\begin{equation}\protect\hypertarget{eq-hier_factorized}{}{
-\begin{aligned}
-\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\},
-    \{\underline{\underline{F}}_{[j]}\} \mid
-    \{\underline{\underline{R}}_{[j]}\}
-) =
-&\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\} \mid
-    \{\underline{\bar{s}}_{T[j]}\},
-    \{\underline{\underline{F}}_{[j]}\} 
-) \times \\
-&\pi(
-    \{\underline{\bar{s}}_{T[j]}\} \mid
-    \{\underline{\underline{F}}_{[j]}\}
-) \times \\
-&\pi(
-    \{\underline{\underline{F}}_{[j]}\} \mid
-    \{\underline{\underline{R}}_{[j]}\}
-)
-\end{aligned}
-}\label{eq-hier_factorized}\end{equation} Furthermore, the hierarchical
-structure only connects the replicates via the relative fitness
-parameters. This means that the population mean fitness values and the
-frequencies can be independently inferred for each dataset. This allows
-us to rewrite the right-hand side of Equation~\ref{eq-hier_factorized}
-as
-\begin{equation}\protect\hypertarget{eq-hier_bayes_factorized_replicate}{}{
-\begin{aligned}
-\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\},
-    \{\underline{\underline{F}}_{[j]}\} \mid
-    \{\underline{\underline{R}}_{[j]}\}
-) =
-&\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\} \mid
-    \{\underline{\bar{s}}_{T[j]}\},
-    \{\underline{\underline{F}}_{[j]}\} 
-) \times \\
-&\prod_{j=1}^E \left[
-    \pi(
-        \underline{\bar{s}}_{T[j]} \mid
-        \underline{\underline{F}}_{[j]}
-    )
-    \pi(
-        \underline{\underline{F}}_{[j]} \mid
-        \underline{\underline{R}}_{[j]}
-    )
-\right].
-\end{aligned}
-}\label{eq-hier_bayes_factorized_replicate}\end{equation} The terms
-inside the square brackets in
-Equation~\ref{eq-hier_bayes_factorized_replicate} take the same
-functional form as those derived in Section~\ref{sec-bayes_meanfit} and
-Section~\ref{sec-bayes_freq}. Therefore, to implement the desired
-hierarchical model, we only need to focus on the first term on the
-right-hand side of Equation~\ref{eq-hier_bayes_factorized_replicate}. A
-way to think about the structure of the hierarchical model is as
-follows: imagine each genotype as a ``\emph{true}'' relative fitness
-value. However, every time we perform an experiment, small variations in
-the biotic and abiotic conditions---also known as batch effects---might
-result in small deviations from this value. We model this by defining a
-distribution for the hyper-fitness parameter---the ground truth we are
-interested in---and having each experimental replicate sample from this
-hyper-parameter distribution to determine the ``\emph{local}'' fitness
-value. The wider the hyper-parameter distribution is the more
-variability between experimental replicates.
-
-Writing Bayes' theorem for the first term in
-Equation~\ref{eq-hier_bayes_factorized_replicate} results in
-\begin{equation}\protect\hypertarget{eq-hier_bayes}{}{
-\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\} \mid
-    \{\underline{\underline{F}}_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\}
-) \propto
-\pi(
-    \{\underline{\underline{F}}_{[j]}\} \mid
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\}
-)
-\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\} \mid
-    \{\underline{\bar{s}}_{T[j]}\}
-),
-}\label{eq-hier_bayes}\end{equation} where we leave the conditioning on
-the population mean fitness as we did in Section~\ref{sec-bayes_mutfit}.
-This expression can be simplified in two ways. First, the frequency
-values for each experimental replicate depend directly on the local
-fitness values and the corresponding population mean fitness, as the
-relationship between experimental replicates only occurs through the
-relative fitness values. Therefore, we can write
-\begin{equation}\protect\hypertarget{eq-hier_bayes_exp_indep}{}{
-\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\} \mid
-    \{\underline{\underline{F}}_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\}
-) \propto
-\prod_{j=1}^E\left[
-    \pi(
-        \underline{\underline{F}}_{[j]} \mid
-        \underline{s}^M_{[j]},
-        \underline{\bar{s}}_{T[j]}
-    )
-\right]
-\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\} \mid
-    \{\underline{\bar{s}}_{T[j]}\}
-).
-}\label{eq-hier_bayes_exp_indep}\end{equation} Second, the relationship
-between the hyper-fitness and the local fitness values allows us to
-write their joint distribution as a conditional distribution where local
-fitness values depend on the global hyper-fitness value, obtaining
-\begin{equation}\protect\hypertarget{eq-hier_bayes_exp_indep}{}{
-\pi(
-    \underline{\theta}^M, 
-    \{\underline{s}^M_{[j]}\} \mid
-    \{\underline{\underline{F}}_{[j]}\},
-    \{\underline{\bar{s}}_{T[j]}\}
-) \propto
-\prod_{j=1}^E\left[
-    \pi(
-        \underline{\underline{F}}_{[j]} \mid
-        \underline{s}^M_{[j]},
-        \underline{\bar{s}}_{T[j]}
-    )
-    \pi(
-        \underline{s}^M_{[j]} \mid
-        \underline{\theta}^M
-    )
-\right] 
-\pi(
-    \underline{\theta}^M
-).
-}\label{eq-hier_bayes_exp_indep}\end{equation} Notice we removed the
-conditioning on the population mean fitness as our prior expectations of
-what the global hyper-fitness or local fitness value might be do not
-depend on these nuisance parameters.
-
-The first term on the right-hand side of
-Equation~\ref{eq-hier_bayes_exp_indep} takes the same functional form as
-the one derived in Section~\ref{sec-bayes_mutfit}. Therefore, all we are
-left with is to determine the functional forms for the hyper--prior
-\(\pi(\underline{\theta}^M)\), and the conditional probability
-\(\pi(\underline{s}^M_{[j]} \mid \underline{\theta}^M)\). In analogy to
-the assumptions used for the fitness values in
-Section~\ref{sec-bayes_mutfit}, we define the value of each
-hyper-fitness as independent. This means that we have
-\begin{equation}\protect\hypertarget{eq-hier_hyper_prior}{}{
-\pi(\underline{\theta}^M) = \prod_{m=1}^M \pi(\theta^{(m)}).
-}\label{eq-hier_hyper_prior}\end{equation} Furthermore, we assume this
-prior is of the form
-\begin{equation}\protect\hypertarget{eq-hier_hyper_prior_02}{}{
-\theta^{(m)} \sim \mathcal{N}(\mu_{\theta^{(m)}}, \sigma_{\theta^{(m)}}),
-}\label{eq-hier_hyper_prior_02}\end{equation} where
-\(\mu_{\theta^{(m)}}\) and \(\sigma_{\theta^{(m)}}\) are user-defined
-parameters encoding the prior expectations on the fitness values.
-
-For the conditional distribution
-\(\pi(\underline{s}^M_{[j]} \mid \underline{\theta}^M)\), we use the
-so-called non-centered parametrization that avoids some of the intrinsic
-degeneracies associated with hierarchical models
-\autocite{betancourt2013}. We invite the reader to check
-\href{https://betanalpha.github.io/assets/case_studies/hierarchical_modeling.html}{this
-excellent blog} explaining the difficulties of working with hierarchical
-models. This non-centered parameterization implies that we introduce two
-nuisance parameters such that the local fitness \(s_{[j]}^{(m)}\) is
-computed as \begin{equation}\protect\hypertarget{eq-hier_noncentered}{}{
-s_{[j]}^{(m)} = \theta^{(m)} + 
-(\tau_{[j]}^{(m)} \times \xi_{[j]}^{(m)}),
-}\label{eq-hier_noncentered}\end{equation} where \(\theta^{(m)}\) is the
-corresponding genotype hyper-fitness value, \(\xi_{[j]}^{(m)}\) is a
-standard normal random variable, i.e.,
-\begin{equation}\protect\hypertarget{eq-hier_stdnormal}{}{
-\xi_{[j]}^{(m)} \sim \mathcal{N}(0, 1),
-}\label{eq-hier_stdnormal}\end{equation} that allows deviations from the
-hyper-fitness value to be either positive or negative, and
-\(\tau_{[j]}^{(m)}\) is a strictly positive random variable that
-characterizes the deviation of the local fitness value from the global
-hyper-fitness. We assume
-\begin{equation}\protect\hypertarget{eq-hier_tau}{}{
-\tau_{[j]}^{(m)} \sim \log\mathcal{N}(
-    \mu_{\tau_{[j]}^{(m)}}, \sigma_{\tau_{[j]}^{(m)}}
-)
-}\label{eq-hier_tau}\end{equation} where \(\mu_{\tau_{[j]}^{(m)}}\) and
-\(\sigma_{\tau_{[j]}^{(m)}}\) are user-defined parameters capturing the
-expected magnitude of the batch effects.
-
-\hypertarget{defining-prior-probabilities}{%
-\subsubsection{Defining prior
-probabilities}\label{defining-prior-probabilities}}
-
-One aspect commonly associated---in both positive and negative ways---to
-Bayesian analysis is the definition of prior probabilities. On the one
-hand, the naive textbook version of Bayesian analysis defines the prior
-as encoding the information we have about the inference in question
-before acquiring any data. This is the ``ideal'' use of priors that,
-whenever possible, should be implemented. On the other hand, for most
-practitioners of Bayesian statistics in the age of big data, the
-definition of prior becomes a tool to ensure the convergence of sampling
-algorithms such as MCMC \autocite{gelman2017}. However, for our
-particular problem, although we deal with large amounts of data
-(inferences can be made for \textgreater{} 10K barcodes over multiple
-time points, resulting in \textgreater{} 100K parameters), each barcode
-has very little data, as they are measured only once per time point over
-\textless{} 10 growth-dilution cycles. Furthermore, it is incredibly
-challenging to understand the noise sources related to culturing
-conditions, DNA extraction, library preparation, etc., and encode them
-into reasonable prior distributions.
-
-Empirically, our approach for this work defined the priors based solely
-on the neutral lineage data, as they represent the only repeated
-measurements of a single genotype in our experimental design. We
-acknowledge that defining the priors after observing the data might be
-considered an incoherent inference. However, as expressed by
-\textcite{gelman2017}
-
-\begin{quote}
-Incoherence is an unavoidable aspect of much real-world data analysis;
-and, indeed, one might argue that as scientists we learn the most from
-the anomalies and reassessments associated with episodes of incoherence.
-\end{quote}
-
-With this in mind, we leave it to the reader to judge the selection of
-priors. Furthermore, the software package associated with this work,
-\texttt{BarBay.jl}, is written so that users can experiment with
-different prior selection criteria that fit their needs. We strongly
-advocate that statistics should not be done in a black-box fit-all tool
-mindset but rather as a formal way to encode the assumptions behind the
-analysis, subject to constructive criticism. With this philosophical
-baggage behind us, let us now focus on how the priors used for this work
-were selected.
-
-\hypertarget{naive-neutral-lineage-based-priors}{%
-\paragraph{Naive neutral lineage-based
-priors}\label{naive-neutral-lineage-based-priors}}
-
-For the base model presented in this work, the user-defined prior
-parameters include the following:
-
-\begin{itemize}
-\item
-  Prior on population mean fitness (one per pair of adjacent time
-  points) \begin{equation}\protect\hypertarget{eq-prior_meanfit_mean}{}{
-  \bar{s}_t \sim \mathcal{N}(\mu_{\bar{s}_t}, \sigma_{{\bar{s}_t}}).
-  }\label{eq-prior_meanfit_mean}\end{equation}
-\item
-  Prior on standard deviation associated with neutral lineages
-  likelihood function (one per pair of adjacent time points)
-  \begin{equation}\protect\hypertarget{eq-prior_meanfit_var}{}{
-  \sigma_t \sim \log\mathcal{N}(\mu_{\sigma_t}, \sigma_{\sigma_t}).
-  }\label{eq-prior_meanfit_var}\end{equation}
-\item
-  Prior on relative fitness (one per non-neutral barcode)
-  \begin{equation}\protect\hypertarget{eq-prior_fit_mean}{}{
-  s^{(m)} \sim \mathcal{N}(\mu_{s^{(m)}}, \sigma_{s^{(m)}}).
-  }\label{eq-prior_fit_mean}\end{equation}
-\item
-  Prior on standard deviation associated with non-neutral lineages
-  likelihood function (one per non-neutral barcode)
-  \begin{equation}\protect\hypertarget{eq-prior_fit_var}{}{
-  \sigma^{(m)} \sim \log\mathcal{N}(\mu_{\sigma^{(m)}}, \sigma_{\sigma^{(m)}})
-  }\label{eq-prior_fit_var}\end{equation}
-\end{itemize}
-
-The \texttt{BarBay.jl} package includes a function \texttt{naive\_prior}
-within the \texttt{stats} module. This function utilizes the data from
-the neutral lineages to determine some of the prior parameters to
-facilitate the inference algorithm's numerical convergence. In
-particular, it defines the population mean fitness parameter
-\(\mu_{\bar{s}_t}\) as
-\begin{equation}\protect\hypertarget{eq-prior_meanfit_mean_data}{}{
-\mu_{\bar{s}_t} = \frac{1}{N} \sum_{n=1}^N 
--\ln\left( \frac{r_{t+1}^{(n)}}{r_t^{(n)}} \right),
-}\label{eq-prior_meanfit_mean_data}\end{equation} where \(N\) is the
-number of neutral lineages and \(r_t^{(n)}\) is the number of neutral
-lineages. In other words, it defines the mean of the prior distribution
-as the mean of what one naively would compute from the neutral lineages,
-discarding cases where the ratio diverges because the denominator
-\(r_t^{(n)} = 0\). For the variance parameter, we chose a value
-\(\sigma_{\bar{s}_t} = 0.05\).
-
-Furthermore, the \texttt{naive\_prior} function defines the mean of the
-variance parameter as the standard deviation of the log frequency ratios
-for the neutral lineages, i.e.,
-\begin{equation}\protect\hypertarget{eq-prior_meanfit_var_data}{}{
-\mu_{\sigma_t} = \sqrt{
-    \operatorname{Var}\left(\frac{r_{t+1}^{(n)}}{r_t^{(n)}}\right)
-},
-}\label{eq-prior_meanfit_var_data}\end{equation} where
-\(\operatorname{Var}\) is the sample variance. This same value was
-utilized for the mean of the non-neutral barcode variance
-\(\mu_{\sigma^{(m)}}\). While we assigne the corresponding variances to
-be \(\sigma_{\sigma_t} = \sigma_{\sigma^{(m)}} = 1\).
-
-\hypertarget{sec-ppc}{%
-\subsection{Posterior predictive checks}\label{sec-ppc}}
-
-Throughout the main text, we allude to the concept of posterior
-predictive checks as a formal way to assess the accuracy of our
-inference pipeline. Here, we explain the mechanics behind the
-computation of these credible regions, given the output of the
-inference.
-
-Bayesian models encode what is known as a \emph{generative model}. This
-statement means that in our definition of the likelihood function and
-the prior distribution, we, as modelers, propose a mathematical function
-that captures all relevant relationships between unobserved (latent)
-variables. Therefore, when these latent variables are input into the
-mathematical model, this function \emph{generates} data that should be,
-in principle, indistinguishable from the real observations if the model
-is a good account of the underlying processes involved in the phenomena
-of interest. This generative model implies that once we run the
-inference process and update our posterior beliefs about the state of
-the latent variables, we can input back the inferred values to our model
-and generate synthetic data. Furthermore, we can repeat this process
-multiple times to compute the range where we expect to observe our data
-conditioned on the accuracy of the model.
-
-For our specific scenario, recall that our objective is to infer the
-relative fitness of a non-neutral lineage \(s^{(m)}\) along with
-nuisance parameters related to the population mean fitness at each
-point, \(\underline{\bar{s}}_t\), and the barcode frequency time series
-\(\underline{f}^{(m)}\). All these variables are related through our
-fitness model (see Section~\ref{sec-fitness_model} in the main text)
-\begin{equation}\protect\hypertarget{eq-ppc_model}{}{
-f_{t+1}^{(m)} = f_{t}^{(m)} \mathrm{e}^{(s^{(m)} - s_t)\tau}.
-}\label{eq-ppc_model}\end{equation} As we saw, it is convenient to
-rewrite Equation~\ref{eq-ppc_model} as
-\begin{equation}\protect\hypertarget{eq-ppc_logratio}{}{
-\frac{1}{\tau}\ln\frac{f_{t+1}^{(m)}}{f_{t}^{(m)}} = (s^{(m)} - s_t).
-}\label{eq-ppc_logratio}\end{equation} Written in this way, we separate
-the quantities we can compute from the experimental observations---the
-left-hand side of Equation~\ref{eq-ppc_logratio} can be computed from
-the barcode reads---from the latent variables.
-
-Although we perform the joint inference over all barcodes in the present
-work, let us focus on the inference task for a single barcode as if it
-were computed independently. For a non-neutral barcode, our task
-consists of computing the posterior probability
-\begin{equation}\protect\hypertarget{eq-ppc_posterior}{}{
-\pi(\theta \mid \underline{r}^{(m)}) =
-\pi(
-    s^{(m)}, \sigma^{(m)}, \underline{s}_t, \underline{f}^{(m)}
-    \mid \underline{r}^{(m)}
-),
-}\label{eq-ppc_posterior}\end{equation} where \(\theta\) represents all
-parameters to be inferred and \(\underline{r}^{(m)}\) is the vector with
-the barcode raw counts time series. The list of parameters are
-
-\begin{itemize}
-\tightlist
-\item
-  \(s^{(m)}\): The barcode's relative fitness.
-\item
-  \(\sigma^{(m)}\): A nuisance parameter used in the likelihood to
-  generate the data. This captures the expected deviation from
-  Equation~\ref{eq-ppc_logratio}
-\item
-  \(\underline{s}_t\): The vector with all population mean fitness for
-  each pair of adjacent time points.
-\item
-  \(\underline{f}^{(m)}\): The vector with the barcode frequency time
-  series.
-\end{itemize}
-
-Furthermore, let us define a naive estimate of the barcode frequency at
-time \(t\) as \begin{equation}\protect\hypertarget{eq-ppc_naive_freq}{}{
-\hat{f}_t^{(m)} = \frac{r^{(m)}_t}{\sum_{b=1}^B r^{(b)}_t}.
-}\label{eq-ppc_naive_freq}\end{equation} We can compute this quantity
-from the data by normalizing the raw barcode counts by the sum of all
-barcode counts. Furthermore, we can compute a naive estimate of the log
-frequency ratio from the raw barcode counts as
-\begin{equation}\protect\hypertarget{eq-ppc_naive_logfreq}{}{
-\ln \hat{\gamma}_t^{(m)} = \ln \frac{\hat{f}_{t+1}^{(m)}}{\hat{f}_t^{(m)}}
-}\label{eq-ppc_naive_logfreq}\end{equation}
-
-In our generative model, we assumed
-\begin{equation}\protect\hypertarget{eq-ppc_logfreq_likelihood}{}{
-\ln\gamma^{(m)}_t \mid \theta \sim \mathcal{N}(s^{(m)} - s_t, \sigma^{(m)}).
-}\label{eq-ppc_logfreq_likelihood}\end{equation} This implies that once
-we determine the posterior distribution of our parameters, we can
-generate synthetic values of \(\ln \gamma^{(m)}_t\) that we can then
-compare with the values obtained from applying
-Equation~\ref{eq-ppc_naive_logfreq} and
-Equation~\ref{eq-ppc_naive_logfreq} to the raw data.
-
-In practice, to compute the posterior predictive checks, we generate
-multiple samples from the posterior distribution
-\(\pi(\theta \mid \underline{r}^{(m)})\)
-\begin{equation}\protect\hypertarget{eq-ppc_theta_samples}{}{
-\underline{\theta} = (\theta_1, \theta_2, \ldots, \theta_N).
-}\label{eq-ppc_theta_samples}\end{equation} With these samples in hand,
-the \texttt{BarBay.jl} package includes the function
-\texttt{logfreq\_ratio\_bc\_ppc} for non-neutral barcodes that uses this
-set of posterior parameter samples to generate samples from the
-distribution defined in Equation~\ref{eq-ppc_logfreq_likelihood}. For a
-large-enough number of samples, we can then compute the desired
-percentiles---5, 68, and 95 percentiles in all figures in the main
-text---that are equivalent to the corresponding credible regions. In
-other words, the range of values of \(\ln\gamma_t^{(m)}\) generated by
-this bootstrap process can be used to compute the region where we expect
-to find our raw estimates \(\ln\hat{\gamma}_t^{(m)}\) with the desired
-probability. The package \texttt{BarBay.jl} includes an equivalent
-function, \texttt{logfreq\_ratio\_popmean\_ppc}, for neutral lineages.
-
-\hypertarget{sec-logistic}{%
-\subsection{Logistic growth simulation}\label{sec-logistic}}
-
-In this section, we explain the simulations used to assess the validity
-of our inference pipeline. Let us begin by assuming that, since the
-strains are grown for two full days in the experiment, having left
-behind the exponential phase for almost an entire day, a simple
-exponential growth of the form
-\begin{equation}\protect\hypertarget{eq-logistic_expo}{}{
-\frac{dn_i}{dt} = \lambda_i n_i,
-}\label{eq-logistic_expo}\end{equation} where \(n_i\) is the number of
-cells of strain \(i\), and \(\lambda_i\) is the corresponding growth
-rate is not enough. Instead, we will assume that the cells follow the
-logistic growth equation of the form
-\begin{equation}\protect\hypertarget{eq-logistic_fn}{}{
-\frac{dn_i}{dt} = \lambda_i n_i 
-\left( 1 - \frac{\sum_{j=1}^{N}n_i}{\kappa}\right),
-}\label{eq-logistic_fn}\end{equation} where \(\kappa\) is the carrying
-capacity, and \(N\) is the total number of strains in the culture.
-
-The inference method is based on the model that assumes that the time
-passed between dilutions \(\tau \approx 8\) generations, the change in
-frequency for a mutant barcode can be approximated from cycle \(t\) to
-the next cycle \(t + 1\) as
-\begin{equation}\protect\hypertarget{eq-logistic_fitness_fn}{}{
-f_{t+1}^{(m)} = f_t^{(m)} \mathrm{e}^{(s^{(m)} - \bar{s}_t)\tau},
-}\label{eq-logistic_fitness_fn}\end{equation} where \(s^{(m)}\) is the
-relative fitness for strain \(i\) compared to the ancestral strain and
-\(\bar{s}_t\) is the mean fitness of the population at cycle \(t\). To
-test this assumption, we implemented a numerical experiment following
-the logistic growth model described in Equation~\ref{eq-logistic_fn}.
-Figure~\ref{fig-SIX_logistic_growth_01} shows an example of the
-deterministic trajectories for 50 labeled neutral lineages and 1000
-lineages of interest. The upper red curve that dominates the culture
-represents the unlabeled ancestral strain included in the experimental
-design described in Section~\ref{sec-experiment}.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/figSIX_logistic_growth_01.pdf}
-
-}
-
-\caption{\label{fig-SIX_logistic_growth_01}\textbf{Logistic growth
-simulation over single growth cycle}. The dashed line represents the
-neutral lineages, with the upper curve being the unlabeled neutral
-strain. Color curves represent the genotypes of interest colored by
-growth rate relative to the neutral lineage.}
-
-\end{figure}
-
-To simulate multiple growth-dilution cycles, we take the population
-composition at the final time point and use it to initialize a new
-logistic growth simulation. Figure~\ref{fig-SIX_logistic_growth_02}
-shows the resulting number of cells at the last time point of a cycle
-over multiple growth-dilution cycles for the genotypes in
-@Figure~\ref{fig-SIX_logistic_growth_01}. We can see that the adaptive
-lineages (blue curves) increase in abundance, while detrimental lineages
-(red curves) decrease.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/figSIX_logistic_growth_02.pdf}
-
-}
-
-\caption{\label{fig-SIX_logistic_growth_02}\textbf{Growth-dilution
-cycles for logistic growth simulation}. Each point represents the final
-number of cells after a growth cycle for each lineage. Colors are the
-same as in Figure~\ref{fig-SIX_logistic_growth_01}.}
-
-\end{figure}
-
-In Section~\ref{sec-fitness_model}, we derive the functional form to
-infer the relative fitness of each lineage as
-\begin{equation}\protect\hypertarget{eq-logistic_logfreq}{}{
-\frac{1}{\tau}\ln \frac{f_{t+1}^{(b)}}{f_{t}^{(b)}} = (s^{(b)} - \bar{s}_t).
-}\label{eq-logistic_logfreq}\end{equation}
-Figure~\ref{fig-SIX_logistic_growth_03} shows the corresponding log
-frequency ratio curves for the logistic growth simulation. The
-displacement of these curves with respect to the neutral lineages
-determines the ground truth relative fitness value for these
-simulations.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/figSIX_logistic_growth_03.pdf}
-
-}
-
-\caption{\label{fig-SIX_logistic_growth_03}\textbf{Log frequency ratio
-for logistic growth simulations}. The relative distance of the color
-curves from the black dashed line determines the relative fitness of
-each linage.}
-
-\end{figure}
-
-To simulate the experimental noise, we add two types of noise:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Poisson noise between dilutions. For this, we take the final point of
-  the logistic growth simulation and sample a random Poisson number
-  based on this last point to set the initial condition for the next
-  cycle.
-\item
-  Gaussian noise when performing the measurements. When translating the
-  underlying population composition to the number of reads, we can add a
-  custom amount of Gaussian noise.
-\end{enumerate}
-
-Figure~\ref{fig-SIX_logistic_growth_04} shows the frequency trajectories
-(left panels) and log frequency ratios (right panels) for a noiseless
-simulation (upper panels) and a simulation with added noise (lower
-panels). The noiseless simulation is used to determine the relative
-fitness for each of the lineages, which serves as the ground truth to be
-compared with the resulting inference.
-
-\begin{figure}
-
-{\centering \includegraphics{./figs/figSIX_logistic_growth_04.pdf}
-
-}
-
-\caption{\label{fig-SIX_logistic_growth_04}\textbf{Logistic
-growth-dilution simulations with and without noise}}
-
-\end{figure}
-
-% Print supplemental references changing the title
-\printbibliography[title={Supplemental References},
-segment=\therefsegment, filter=notother]
-\end{refsegment}
-
-% \printbibliography
-
-\end{document}
+In this study, we present a novel Bayesian pipeline for analyzing DNA barcode
+tracking sequencing data, addressing the challenge of accurately quantifying
+competitive microbial fitness in the presence of experimental noise. Our method
+uniquely contributes to understanding microbial evolutionary dynamics by
+enabling reliable inference of the relative fitness of diverse microbial strains
+from high-throughput lineage tracking assays. Our approach is distinct in its
+ability to systematically account for and propagate uncertainties from various
+noise sources throughout all inferred parameters. Furthermore, the
+error-propagation quality of our Bayesian method allows us to extend the
+inference pipeline to common dataset structures, such as jointly analyzing
+multiple experimental replicates or accounting for multiple unique barcodes
+mapping the equivalent genotypes. This comprehensive treatment of uncertainties
+is crucial in experimental settings where noise can significantly influence the
+results. Furthermore, we have optimized our pipeline for scalability, allowing
+it to handle large numbers of unique barcodes effectively. This scalability is
+essential for analyzing complex datasets typical in microbial fitness studies.

### 2024-02-26 doc/paper.qmd
working template. Everything looks okay.
prepare paper for submission
add lua filters to split main text and SI text references in the PDF automatically
add new SI section
add kinsler section to paper
Add time scaling section to paper
upload missing document modifications

@@
-# Abstract
+::: {.maintext}
 
-{{< include _abstract.qmd >}}
+# Abstract {.unnumbered}
+
+The tracking of lineage frequencies via DNA barcode sequencing enables the
+quantification of microbial fitness. However, experimental noise coming from
+biotic and abiotic sources complicates the computation of a reliable inference.
+We present a Bayesian pipeline to infer relative microbial fitness from
+high-throughput lineage tracking assays. Our model accounts for multiple sources
+of noise and propagates uncertainties throughout all parameters in a systematic
+way. Furthermore, using modern variational inference methods based on automatic
+differentiation, we are able to scale the inference to a large number of unique
+barcodes. We extend this core model to analyze multi-environment assays,
+replicate experiments, and barcodes linked to genotypes. On simulations, our
+method recovers known parameters within posterior credible intervals. This work
+provides a generalizable Bayesian framework to analyze lineage tracking
+experiments. The accompanying open-source software library enables the adoption
+of principled statistical methods in experimental evolution.
+
+# Author Summary {.unnumbered}
+
+{{< include _author_summary.qmd >}}
 
@@
 
-# Acknowledgements
+# Materials and Methods {.unnumbered}
+
+All synthetic data generation and custom scripts used in this work were stored
+using Git version control. Code for analysis and figure generation can be found
+on the GitHub repository (https://github.com/mrazomej/bayesian_fitness). The
+accompanying software package `BarBay.jl` can be directly installed from the
+`Julia` package repository, or via cloning the corresponding GitHub repository
+(https://github.com/mrazomej/BarBay.jl)
+
+# Acknowledgements {.unnumbered}
 
 {{< include _acknowledgements.qmd >}}
+:::
 
+::: {.sitext}
 # Supplementary Materials
@@
 {{< include _logistic_sim.qmd >}}
+
+<!-- Stochastic growth simulations -->
+{{< include _fitseq2_sim.qmd >}}
+
+<!-- Kinsler et al. 2020 -->
+{{< include _kinsler.qmd >}}
+
+<!-- Scaling as a function of number of barcodes -->
+{{< include _time_scaling.qmd >}}
+:::
